### 2020-11-24 main.tex
Initial Overleaf Import

\documentclass{article}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{hyperref}       % hyperlinks
\usepackage{tikz}           % define for loops

\hypersetup{ % SLJ: my standard paper setup...
	pdftitle={MAP convergence},
	pdfkeywords={},
	pdfborder=0 0 0,
	pdfpagemode=UseNone,
	colorlinks=true,
	linkcolor=blue, %mydarkblue,
	citecolor=blue, %mydarkblue,
	filecolor=blue, %mydarkblue,
	urlcolor=blue, %mydarkblue,
	pdfview=FitH,
	pdfauthor={Anonymous},
}

\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsubsection*{\bibname}}

\newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
\newcommand{\violet}[1]{\textcolor{violet}{#1}}

% my packages
\usepackage{math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpartition}{A}
\newcommand{\conj}{\logpartition^*}
\newcommand{\bregman}{\cB_\logpartition}
\newcommand{\bregmanconj}{\cB_{\logpartition^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\Dir}{\mathrm{Dir}} % Dirichlet
\newcommand{\linear}{\ell} % linearization of a function

\title{Convergence rate of MAP estimates \\
for the exponential family}
\author{R\'emi Le Priol}
\date{October 2020}

\begin{document}

\maketitle

\section{Background}
It's hard to find general convergence rate on the KL for the maximum likelihood estimates of the exponential family.  We want the simplest one.
We hope  to get a new result by combining tools from statistics and optimization.

The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model
\begin{align}
    p(X|\natp) &= \exp( \natp^\top T(X) - \logpartition(\natp)) \; ,
\end{align}
with log-partition function 
\begin{align}
    \logpartition(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
\end{align}
Recall that $\logpartition$ verifies the two following identities 
\begin{align}
    \nabla\logpartition(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpartition(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
The second identity entails that $\logpartition$ is strictly-convex.
The conjugate prior for this distribution is
\begin{align}
    \pi(\natp) &\propto \exp( - n_0 \bregman(\natp || \natp_0) )
\end{align}
where $n_0$ is a number of fictional points observed from a distribution with parameter $\natp_0$.
$\bregman(\natp || \natp_0)$ is the Bregman divergence induced by $\logpartition$ between $\natp$ and $\natp_0$
\begin{align}
    \bregman (\natp || \natp_0)
    & = \logpartition(\natp) - \logpartition(\natp_0) 
    - \langle \meanp_0 , \natp - \natp_0 \rangle
\end{align}
with $\meanp_0 = \nabla \logpartition(\natp_0) = \expect[\natp_0]{T(X)}$ the mean parameter associated to the natural parameter $\natp_0$.
The negative log-likelihood of the prior is then
\begin{align*}
    -\log \pi(\natp) = n_0 (\logpartition(\natp)  - \natp^\top \meanp_0 ) + \cst
\end{align*}
Thus the joint NLL of $(x_1,\dots,x_n,\natp)$ is
\begin{align}
    -\log p(X|\natp)\pi(\natp) 
    = (n_0+n) \logpartition (\natp) 
    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(x_i) \right) \; .
\end{align}
Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
\begin{align}
    \hat \natp = \argmin_\natp -\log p(X|\natp) + n_0 \bregman(\natp || \natp_0)
\end{align}
such that the MAP is
\begin{align}
    \nabla \logpartition(\hat \natp) = \hat \meanp
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
\end{align}
The MAP estimate is a random quantity.

\section{Stochastic Proximal Bregman Point}
We see that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp || \natp_0)$. This is a stochastic proximal Bregman step with step-size $\inv{n_0}$. This can also be seen at each step since
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp -\log p(x_{n+1}|\natp) + (n_0 + n) \bregman(\natp || \hat \natp_n) \\
    &= \argmin_\natp f(\natp) +  \inv{\gamma_n}\bregman(\natp || \hat \natp_n) \; .
\end{align}
Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.

This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.

\section{Stochastic Mirror Descent}
It turns out that the MAP can also be seen as the trajectory of stochastic mirror descent
\begin{align}
    \hat \natp_{n+1} 
    &= \argmin_\natp 
    - \langle T(x_{t+1}), \natp \rangle  + \logpartition(\natp) + (n_0 + n) \bregman(\natp || \natp_n) \\
    %&(n_0+n)\left(\logpartition(\natp) - \logpartition(\natp_n) - \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle \right)  \\
    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpartition(\natp_n) + \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp || \natp_n) \\
    &= \argmin_\natp \linear_f(\natp;\natp_n) + (n_0 + n + 1) \bregman(\natp || \natp_n)
\end{align}
where $\linear_f(\natp;\natp_n)$ is the linearization of $f$ at $\natp_n$ evaluated at $\natp$. 
This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpartition$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.

In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpartition$ \RLP{cite {bubeck}}.
In our setting this is not always true -- eg gaussians.
However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity. 
For mirror descent, there are well linear and sublinear rate, even when the function is not smooth or strongly convex ! Beautiful. Cite nesterov, and teboulle, and the first birnbaum.

For stochastic mirror descent, there is  a sublinear rate (cite richtarik), but under a weird assumption on gradient's variance. Still need to check “Relative Continuity” for Non-Lipschitz Nonsmooth Convex Optimization Using Stochastic (or Deterministic) Mirror Descent".

Also cite "The Information Geometry of Mirror Descent", saying they only give asymptotic results.

\section{Straightforward Convergence Rate}
In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
\begin{align}
    \expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}
    &= \KL( p(.|\natp^*) || p(.|\natp)) \\
    &= \bregman (\natp || \natp^*) \\
    &= \bregmanconj ( \meanp^* || \meanp)
    %& = \logpartition(\natp) - \logpartition(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
\end{align}
where $\conj$ is the entropy, the convex conjugate of the log-partition.
The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. 
The question is: how does this quantity behave when $\natp$ is the MAP estimate ? Can we get bounds -- in expectation or high-probability ?

\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpartition$ is defined within the $\ell^2$-ball of radius $L$), then
\begin{align}
    \bregmanconj(\mu^* || \mu) 
    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
    \leq 2L \norm{\mu^* - \mu}
\end{align}
so $\bregmanconj$ is $2L$-Lipschitz.
Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this is the convergence rate of the log-likelihood. \RLP{find relevant inequalities}.

\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpartition$ is $L^{-1}$-strongly convex), then
\begin{align}
    \bregmanconj(\mu^* || \mu) 
    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
\end{align}
so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.

\section{Self-Concordance}

A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
\begin{align}
    \logpartition(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
\end{align}
which is defined on $\eta \in \real^d$ and $\Lambda\in \real^{d\times d}$ symmetric positive definite. 
It is not strongly convex, so $\conj$ is not smooth.

Another hypothesis that may be more suitable is self-concordance. $f: \real \rightarrow \real$ is SC if 
\begin{align}
    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
\end{align}
The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
Negative logarithm $-\log(x)$ and entropy $x\log(x)$ are both self-concordant function. This is good news for us since log-partition function may include SC logarithmic barriers.
In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.

\subsection{Fenchel conjugate motivation to self-concordance}

In the most regular case, when $f(x)$ is a convex function, continuously differentiable on its domain, then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ verifies
\begin{align}
    \nabla f \circ \nabla f^*  &= \Id \\
     \nabla f^* \circ \nabla f &= \Id
\end{align}
where $\Id$ is the identity function on the relevant domain. In words, the gradients of $f$ and $f^*$ are reciprocal.
Deriving this equality yields
\begin{align}
    \nabla^2 f(x) \nabla^2 f^*(x^*) = I_n
\end{align}
where $x,x^*$ are conjugate points -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
Now, it gets interesting to us when we derive again this equality. Let's tackle the 1D case first
\begin{align}
    &f''(x) f^{* \prime \prime}(f'(x)) = 1, \forall x \\
    \implies
    &f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))  = 0 \\
    \implies
    &\frac{f'''(x)}{f''(x){\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*){\frac{3}{2}}}  = 0
\end{align}
where to get to the last line we used the first line, and we divided the second line by $f''(x)^{\frac{1}{2}}$. 
We see that for a pair of conjugate functions, the self-concordance ratio is preserved, modulo the sign.
This gives another rational, beyond dimensional analysis, for using this ratio as a regularity assumption for convex analysis.

It is also very helpful for us, since we are looking at pairs $\logpartition, \conj$, and their associated Bregman divergences. 
If $\logpartition$ is SC, then so is 
$f(\natp)= \bregman(\natp || \natp^*) = \logpartition(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
And $\conj$ is SC as well, thus  $h(\meanp) = \bregmanconj ( \meanp || \meanp^*)$ is SC.
But there is no reason for 
$g(\meanp) = \bregmanconj ( \meanp^* || \meanp) =  \cst - \conj(\meanp) - \langle \nabla \conj (\meanp), \meanp^* - \meanp \rangle $ to be SC.

The multivariate generalization of this formula is a third order tensor equality
\begin{align}
    \nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1} + \nabla^2 f^{* -\half} \nabla^3 f^* \nabla^2 f^{* -1} = 0
\end{align}
where we omit  multiplication axis and functions take relevant argument $x$ or $x^*$. Consequently, a multivariate definition of self-concordance might take the form of an inequality on the 3d tensor $\nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1}$.

\subsection{Suboptimality and Newton Decrement}
An important property of self-concordant functions (cite Boyd's book, although Nesterov's may be better) is that their suboptimality  may be upper bounded by the Newton Decrement 
\begin{align}
    \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
\end{align}
In general, subtracting the minimum $f^*$ of $f$ , we have
\begin{align}
    f(x) - f* \leq - \decrement(x) - \log( 1- \decrement(x)) \; .
\end{align}
Note that this bound is vacuous for $\decrement(x)\geq 1$. For $y = \decrement(x) \leq 0.68$, we have $ - y - \log(1-y) \leq y^2$, so we get the bound
\begin{align}
    f(x) - f^* \leq \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
\end{align}

Our functions of interest is $f(\natp)= \bregman (\natp || \natp^*)= \bregmanconj ( \meanp^* || \meanp) = g(\meanp)$, with minimum $f^*=0$. 
If $\logpartition$ is self-concordant, then so is $f$, but not necessarily $g$.
The gradient and Hessian of $f$ are
\begin{align}
    f(\natp) 
    &= \logpartition(\natp) - \logpartition(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
    \nabla f(\natp) 
    &= \meanp - \meanp^* = \expect[p(X|\natp)]{T(X)} - \expect[p(X|\natp^*)]{T(X)} \\
    \nabla^2 f(\natp) 
    &= \Sigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
\end{align}
so that we get the bound.
\begin{align}
    \bregmanconj(\mu^* || \mu) 
    \leq \decrement(\natp)^2 
    = \norm{\mu^* - \mu(\theta)}^2_{\Sigma(\theta)^{-1}} 
    \leq 0.46
\end{align}
% If we were to use $g(\meanp)$ instead, we would get dirty third order derivatives
% \begin{align}
%     g(\meanp) 
%     &= \conj(\meanp^*) - \conj(\meanp) - \langle \natp, \meanp^* - \meanp \rangle \\
%     \nabla g(\meanp) 
%     &= -\natp - \nabla^2 \conj(\meanp) (\meanp^* - \meanp) + \natp \\
%     &= \nabla^2 \conj(\meanp) (\meanp - \meanp^*) \\
%     \nabla^2 g(\meanp) 
%     &= \nabla^2 \conj(\meanp) + \nabla^3 \conj(\meanp ) (\meanp - \meanp^*)
% \end{align}
% where the last expression involves a third order tensor.

Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp || \meanp^*)$, then we would get
\begin{align}
    \nabla h(\meanp ) 
    &= \natp - \natp^* \\
    \nabla^2 h(\meanp ) 
    &= \nabla^2 \conj(\meanp) 
    = \nabla^2 \logpartition(\natp) ^{-1}
    = \Cov_{p(X|\natp)}[T(X)]^{-1}\\
    \implies 
    \decrement(\meanp)
    &= \Var_{p(X|\natp)}[(\natp - \natp^*)^T T(X)] \; .
\end{align}
This is just a remark. I don't think it can help us to get anywhere.

\bibsection
\bibliography{}

\newpage

\section{MAP on Graphs}
Assume that the variable $X$ factors along some graph $G$. 
We write $G(i)$ the parents of $X_i$ in $G$. 
Then we model the conditional distribution of $X$ given parameter vector $\theta$ factors as
\begin{align}
    p(X|\natp) = \prod_i p(X_i | X_{G(i)};\natp_i)
\end{align}
where $\natp_i$ is the parameter associated to the mechanism $X_{G(i)} \rightarrow X_i$. 
Embracing the Bayesian viewpoint, the independent mechanism principle is embodied as independence between parameters
\begin{align}
    p(\natp) = \prod_i p(\natp_i) \; .
\end{align}
% In other words, we have the following assumptions on $(\natp,X)$
% \begin{align}
%     \indep_{i=1}^d \natp_i \\
%     \natp_i \indep X_{G(i)}
% \end{align}
Following these equations, the joint distribution on $(\natp,X)$ factors along a larger graph $G'$ which augments $G$ by adding nodes $\natp_i$ with arrows pointing to $X_i$, as illustrated in Figure~\ref{fig:joint_graph}.
\begin{figure}[h]
    \centering
    \begin{tikzpicture}[shorten >=1pt,->]
      \tikzstyle{vertex}=[draw, circle,minimum size=24pt,inner sep=0pt]
      \foreach \name/\txt/\x/\y in {x1/X_1/0/0, x2/X_2/3/0, x3/X_3/2/-1.5, t1/\theta_1/-1.5/-0.5, t2/\theta_2/4.5/-0.5, t3/\theta_3/0.5/-2}
        \node[vertex] (G-\name) at (\x,\y) {$\txt$};

      \foreach \from/\to in {x1/x2, x2/x3, x1/x3, t1/x1, t2/x2, t3/x3}
        \draw (G-\from) -- (G-\to);
    \end{tikzpicture}
    \caption{A graph $G'$ factorizing $(\natp,X)$. Although the graph restricted on $X$ does not encode any conditional independence, $G'$ does on the joint distribution.}
    \label{fig:joint_graph}
\end{figure}
With such a graph, the Bayesian posterior can be factorized as well
\begin{align}
    p(\natp|X) 
    &\propto  p(X|\natp) p(\natp)& \\
    & = \prod_i p(X_i | X_{G(i)};\natp_i) p(\natp_i)
    &(\indep_i \natp_i) \\
    & = \prod_i p(X_i | X_{G(i)};\natp_i) p(\natp_i\violet{|X_{G(i)}})
    &(\natp_i \indep X_{G(i)}) \\
    & = \prod_i p(X_i, \violet{\natp_i} | X_{G(i)})    &  \\
    & = \prod_i p(\natp_i| \violet{X_i}, X_{G(i)})p(X_i|X_{G(i)})    &  \\
    \implies p(\natp|X) 
    &= \prod_i p(\natp_i| \violet{X_i}, X_{G(i)}) \; .
\end{align}
In words, a consequence of the independence mechanism principle is that the posterior distribution of $\natp_i$ can be inferred solely from $X_i$ and its parents.

\subsection{Equality of directions for 2 categorical variables}
In my paper on the analysis of causal speed, I proved the equivalence between sampling a joint distribution $\omega = p(A,B) \in \simplex_{K\times K}$ on $(A,B)\in \{1,\dots,K\}^2$ from a Dirichlet with parameter $\gamma\in \real_+^{K\times K}$ and sampling independently the marginal distribution $\mu = p(A)\in \simplex_K$ and the conditional distributions $\nu_i = p(B|A=i)\in\simplex_K$ from Dirichlets with respective parameters $\sum_{j=1}^K \gamma_{:,j} = \gamma \ones$ (matrix vector product) and $\gamma_{i,:}$
\begin{align}
    \underbrace{\Dir_{K^2}((\gamma_{i,j})_{i,j=1}^K) }_{p(\omega)}
    \equiv 
    \underbrace{\Dir_K\left( \gamma \ones \right)}_{p(\mu)} 
    \otimes \left (\bigotimes_{i=1}^K 
    \underbrace{\Dir_K((\gamma_{i,j})_j)}_{p(\nu_i)} 
    \right ) 
    \label{eq:dirichlet_factors}
\end{align}
Seeing data samples $(\cA,\cB) = (A_i,B_i)_{i=1}^n$ as one-hot encodings in $\real^K\times \real^K$, the posterior reads
\begin{align}
p(\mu|\cA) &= \Dir(\gamma\ones + \sum_i A_i)\\
p(\nu_k|\cA,\cB) &= \Dir(\gamma_{k,:} + \sum_i  A_{i,k} B_i ) \\
p(\omega | \cA,\cB) & = \Dir( \gamma + \sum_i A_i B_i^\top )  
\end{align}
where $A_i B_i^\top$ is the one hot matrix encoding of $A,B$.
These three posteriors are obtained independently of each other following rules of calculus for Dirichlet distributions. Yet they happen to define the same distribution on distributions, as we verify below with the two equalities from equation \eqref{eq:dirichlet_factors}.
\begin{align}
    \left(\gamma + \sum_i A_i B_i^\top \right)_{k,l} 
    &= \left(\gamma_{k,:} + \sum_i  A_{i,k} B_i \right)_l \\
    \left(\gamma + \sum_i A_i B_i^\top \right) \ones 
    &= \gamma \ones + \sum_i A_i \; .
\end{align}
The interpretation of this result is that \emph{taking the posterior with the decomposition $A\rightarrow B$ or $B\rightarrow A$ give the same result}. As a corollary the MAP is also the same
\begin{align}
\hat \omega^\text{MAP} 
= \frac{\gamma + \sum_i A_i B_i^\top}{\ones^\top (\gamma + \sum_i A_i B_i^\top) \ones}
=\frac{\gamma + \sum_i A_i B_i^\top}{n_0 + n}
\end{align}
Using Bayesian statistics with this prior, there is no distinction between directions.

Is this bound to happen with a symmetric prior ? Let's give a name to the change of variable $f(\omega) = \mu,\nu$. 
Remark that $f(\omega^\top) = \mu_\leftarrow, \nu_\leftarrow$, eg in the categorical special case transposing omega and changing variables give the anticausal direction. 
For sure  $p(X|\mu,\nu) = p(X|f(\omega)) = p(X|\omega)$. 
Using the change of variable formula  we get something.
But which equality am I looking for exactly ? 
\end{document}

### 2020-11-24 main.tex
Comment on the prior work with relative smoothness

@@
 \renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsubsection*{\bibname}}
+\renewcommand{\bibsection}{\subsection*{\bibname}}
 
@@
     &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpartition(\natp_n) + \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp || \natp_n) \\
-    &= \argmin_\natp \linear_f(\natp;\natp_n) + (n_0 + n + 1) \bregman(\natp || \natp_n)
+    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp || \natp_n)
 \end{align}
-where $\linear_f(\natp;\natp_n)$ is the linearization of $f$ at $\natp_n$ evaluated at $\natp$. 
+where $\linear_f(\natp;\natp_n, x_{t+1})$ is the linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$. 
 This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpartition$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
 
-In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpartition$ \RLP{cite {bubeck}}.
+In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpartition$ \citep{bubeck2015convex}.
 In our setting this is not always true -- eg gaussians.
-However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity. 
-For mirror descent, there are well linear and sublinear rate, even when the function is not smooth or strongly convex ! Beautiful. Cite nesterov, and teboulle, and the first birnbaum.
+However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity. $f$ is $\mu$ strongly-convex and $L$-smooth relative to $h$ if
+\begin{align}
+	f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y||x)
+	\leq f(y) \\
+	\text{and} \quad 
+	f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y||x), \forall x,y \\
+	\iff \mu\cB_h(y||x) \leq \cB_f(y||x) \leq L \cB_h(y||x), \forall x,y \\
+	\iff \mu \nabla^2 h(x) \leq \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
+\end{align} 
+where the last equivalence holds only when $f$ and $h$ are twice differentiable. 
+This sweet generalization of smoothness and strong-convexity transfers the Loewner partial order between matrices to functions, via the Hessian. 
+As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
+As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
+More precisely, he proved the simple finite rate
+\begin{align}
+	f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* || x_0) }{t} \; .
+\end{align}
+These notions were rediscovered and expanded by \citet{bauschke2017descent} and \citet{lu2018relatively}.
+If you need to read one, pick \citet{lu2018relatively} -- I found it much much easier and more enjoyable to read.
+This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.
 
-For stochastic mirror descent, there is  a sublinear rate (cite richtarik), but under a weird assumption on gradient's variance.
-Still need to check “Relative Continuity” for Non-Lipschitz Nonsmooth Convex Optimization Using Stochastic (or Deterministic) Mirror Descent".
+Now our setting is Stochastic Mirror Descent (SMD), meaning at each step we observe a random unbiased estimate gradient.
+This setting was studied by \citet{hanzely2018fastest}, who proved in the smooth strongly-convex case with tail averaging :  with constant step-size, linear convergence down to a variance ball, and with step-size $\gamma_t = n_0 + t$ a rate $\tilde O(\inv{t})$.
+These results match the rates for standard SGD.
+This is very interesting for us, but the variance hyper-parameter is oddly defined.
+Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
+Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1}$.
+\begin{align}
+	\Cov(g_t,\natp_{t+1}) 
+	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \gamma_t \sigma^2, \forall t \\
+	 &= \expect{\langle g_t - \nabla f(\natp_t), \natp_{t+1}  - \natp \rangle}, (\forall \natp)
+\end{align}
+where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$.
+Still need to check €œRelative Continuity€ for Non-Lipschitz Nonsmooth Convex Optimization Using Stochastic (or Deterministic) Mirror Descent, which makes different hypothesis and is well-written.
 
@@
 \bibsection
-\bibliography{}
+\bibliographystyle{apalike}
+\bibliography{references.bib}
 

### 2020-11-26 main.tex
Add details on mirror descent

@@
 \newcommand{\linear}{\ell} % linearization of a function
+\newcommand{\lr}{\gamma} % learning rate, or step-size
 
@@
     &= \argmin_\natp -\log p(x_{n+1}|\natp) + (n_0 + n) \bregman(\natp || \hat \natp_n) \\
-    &= \argmin_\natp f(\natp) +  \inv{\gamma_n}\bregman(\natp || \hat \natp_n) \; .
+    &= \argmin_\natp f(\natp; x_{n+1}) +  \inv{\gamma_n}\bregman(\natp || \hat \natp_n) \; .
 \end{align}
@@
 
-\section{Stochastic Mirror Descent}
-It turns out that the MAP can also be seen as the trajectory of stochastic mirror descent
+\section{Stochastic Mirror Descent (SMD)}
+\subsection{MAP as SMD}
+Let $f(\natp) := \expect[x]{-\log p(x|\natp)} = - \langle \mu_*, \natp\rangle + \logpartition(\natp)$.
+In words, $f$ is linear modification of a convex function $\logpartition$, which we can access only through noisy estimates of $\mu_*$.
+It turns out that the MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpartition$.
+First let's recall mirror descent iteration
+\begin{align}
+	\natp_{t+1} 
+	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp || \natp_t)  \\
+	& = \nabla \conj (\nabla \logpartition(\natp_t) - \lr \nabla f(\natp_t))
+\end{align}
+where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$.
+Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpartition(\natp)$, eg computing the convex conjugate of $\logpartition$.
+Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
+\begin{align}
+	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
+	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
+\end{align}
+which shows exponential convergence and 1-step convergence when $\lr =1$.
+Back to our sheep, the MAP iteration can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1})$ a stochastic estimate  of $\nabla f(\natp_t)$
 \begin{align}
@@
 \end{align}
-where $\linear_f(\natp;\natp_n, x_{t+1})$ is the linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$. 
+where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$.
 This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpartition$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
 
+\subsection{Relative Smoothness for Mirror Descent}
 In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpartition$ \citep{bubeck2015convex}.
@@
 \begin{align}
-	f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y||x)
-	\leq f(y) \\
-	\text{and} \quad 
-	f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y||x), \forall x,y \\
-	\iff \mu\cB_h(y||x) \leq \cB_f(y||x) \leq L \cB_h(y||x), \forall x,y \\
-	\iff \mu \nabla^2 h(x) \leq \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
+	& f-\mu h \ \text{ convex and } Lh - f \ \text{convex} \\
+	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y||x) \leq f(y) \\
+	&\text{and} \quad f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y||x), \forall x,y \\
+	\iff & \mu\cB_h(y||x) \leq \cB_f(y||x) \leq L \cB_h(y||x), \forall x,y \\
+	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
 \end{align} 
 where the last equivalence holds only when $f$ and $h$ are twice differentiable. 
-This sweet generalization of smoothness and strong-convexity transfers the Loewner partial order between matrices to functions, via the Hessian. 
+This sweet generalization of smoothness and strong-convexity transfers the Loewner partial order on symmetric matrices to functions, via the Hessian. 
 As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
 As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
-More precisely, he proved the simple finite rate
+More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality is upper bounded by the simple formula
 \begin{align}
-	f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* || x_0) }{t} \; .
+	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x || x_t) \\
+	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* || x_0) }{t} \; .
 \end{align}
@@
 
+\subsubsection{Three point lemma}
+These proofs rely on another basic property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
+\begin{align}
+	\min	_{x\in C} \overbrace{f(x) + \cB_h(x || y)}^{\phi(x)}
+\end{align}
+where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector.
+Then
+\begin{align}
+	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
+\end{align}
+This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).
+
+\begin{proof}
+	The first order optimality condition is
+	\begin{align}
+		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
+		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
+	\end{align}
+	This proof relies on another property called three point property 
+	\begin{align}
+		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
+	\end{align}
+	which can be proved by expanding the right hand side.
+	By convexity of $f$ we also have 
+	\begin{align}
+		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
+	\end{align} 
+	Putting it all together we get
+	\begin{align}
+		0 
+		&\leq  \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
+		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
+	\end{align}
+	which concludes the proof.
+\end{proof}
+
+\subsection{Relative smoothness for SMD}
 Now our setting is Stochastic Mirror Descent (SMD), meaning at each step we observe a random unbiased estimate gradient.
@@
 These results match the rates for standard SGD.
+
 This is very interesting for us, but the variance hyper-parameter is oddly defined.
 Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
-Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1}$.
+Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1} = \nabla \logpartition^*(\nabla \logpartition(\natp_t) - \gamma_t g_t )$.
 \begin{align}
@@
 \end{align}
-where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$.
-Still need to check €œRelative Continuity€ for Non-Lipschitz Nonsmooth Convex Optimization Using Stochastic (or Deterministic) Mirror Descent, which makes different hypothesis and is well-written.
+where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$. In words, $g_t$ and $\natp_{t+1}$ should not be too negatively correlated. In our case, 
+\begin{align}
+	g_t - \nabla f(\natp_t) = \E[T(X)] - T(x_{t+1})	
+\end{align}
+so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable.
 
-Also cite "The Information Geometry of Mirror Descent", saying they only give asymptotic results.
+The other article \citet{lu2019relative} derives another convergence rate for SMD under another gradient variance assumption.
+
+Also \citet{raskutti2015information} only gives asymptotic results.
 
@@
 The question is: how does this quantity behave when $\natp$ is the MAP estimate ? Can we get bounds -- in expectation or high-probability ? 
+\begin{align}
+	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \\
+	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right )
+\end{align}
 

### 2020-11-27 main.tex
Expand discussion on variance assumption

@@
 \newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
-\newcommand{\logpartition}{A}
-\newcommand{\conj}{\logpartition^*}
-\newcommand{\bregman}{\cB_\logpartition}
-\newcommand{\bregmanconj}{\cB_{\logpartition^*}}
+\newcommand{\logpart}{A}
+\newcommand{\conj}{\logpart^*}
+\newcommand{\bregman}{\cB_\logpart}
+\newcommand{\bregmanconj}{\cB_{\logpart^*}}
 \newcommand{\natp}{\theta}
@@
 \begin{align}
-    p(X|\natp) &= \exp( \natp^\top T(X) - \logpartition(\natp)) \; ,
+    p(X|\natp) &= \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
 \end{align}
@@
 \begin{align}
-    \logpartition(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
+    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
 \end{align}
-Recall that $\logpartition$ verifies the two following identities 
+Recall that $\logpart$ verifies the two following identities 
 \begin{align}
-    \nabla\logpartition(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
-    \nabla^2 \logpartition(\natp) &= \Cov_\natp[T(X)] > 0
+    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
+    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
 \end{align}
 where $\meanp$ is called the mean parameter.
-The second identity entails that $\logpartition$ is strictly-convex.
+The second identity entails that $\logpart$ is strictly-convex.
 The conjugate prior for this distribution is
@@
 where $n_0$ is a number of fictional points observed from a distribution with parameter $\natp_0$.
-$\bregman(\natp || \natp_0)$ is the Bregman divergence induced by $\logpartition$ between $\natp$ and $\natp_0$
+$\bregman(\natp || \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
 \begin{align}
     \bregman (\natp || \natp_0)
-    & = \logpartition(\natp) - \logpartition(\natp_0) 
+    & = \logpart(\natp) - \logpart(\natp_0) 
     - \langle \meanp_0 , \natp - \natp_0 \rangle
 \end{align}
-with $\meanp_0 = \nabla \logpartition(\natp_0) = \expect[\natp_0]{T(X)}$ the mean parameter associated to the natural parameter $\natp_0$.
+with $\meanp_0 = \nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)}$ the mean parameter associated to the natural parameter $\natp_0$.
 The negative log-likelihood of the prior is then
 \begin{align*}
-    -\log \pi(\natp) = n_0 (\logpartition(\natp)  - \natp^\top \meanp_0 ) + \cst
+    -\log \pi(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
 \end{align*}
@@
     -\log p(X|\natp)\pi(\natp) 
-    = (n_0+n) \logpartition (\natp) 
+    = (n_0+n) \logpart (\natp) 
     - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(x_i) \right) \; .
@@
 \begin{align}
-    \nabla \logpartition(\hat \natp) = \hat \meanp
+    \nabla \logpart(\hat \natp) = \hat \meanp
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
@@
 
-\section{Stochastic Proximal Bregman Point}
+\section{Stochastic Bregman Proximal Point}
 We see that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp || \natp_0)$.
-This is a stochastic proximal Bregman step with step-size $\inv{n_0}$.
+This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
 This can also be seen at each step since
@@
 \subsection{MAP as SMD}
-Let $f(\natp) := \expect[x]{-\log p(x|\natp)} = - \langle \mu_*, \natp\rangle + \logpartition(\natp)$.
-In words, $f$ is linear modification of a convex function $\logpartition$, which we can access only through noisy estimates of $\mu_*$.
-It turns out that the MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpartition$.
+Let $f(\natp) := \expect[x]{-\log p(x|\natp)} = - \langle \mu_*, \natp\rangle + \logpart(\natp)$.
+In words, $f$ is linear modification of a convex function $\logpart$, which we can access only through noisy estimates of $\mu_*$.
+It turns out that the MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpart$.
 First let's recall mirror descent iteration
@@
 	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp || \natp_t)  \\
-	& = \nabla \conj (\nabla \logpartition(\natp_t) - \lr \nabla f(\natp_t))
+	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
 \end{align}
 where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$.
-Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpartition(\natp)$, eg computing the convex conjugate of $\logpartition$.
+Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$.
 Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
@@
     &= \argmin_\natp 
-    - \langle T(x_{t+1}), \natp \rangle  + \logpartition(\natp) + (n_0 + n) \bregman(\natp || \natp_n) \\
-    %&(n_0+n)\left(\logpartition(\natp) - \logpartition(\natp_n) - \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle \right)  \\
-    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpartition(\natp_n) + \langle \nabla \logpartition(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp || \natp_n) \\
+    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp || \natp_n) \\
+    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
+    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp || \natp_n) \\
     &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp || \natp_n)
@@
 where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$.
-This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpartition$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
+This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
 
 \subsection{Relative Smoothness for Mirror Descent}
-In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpartition$ \citep{bubeck2015convex}.
+In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpart$ \citep{bubeck2015convex}.
 In our setting this is not always true -- eg gaussians.
@@
 As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
-More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality is upper bounded by the simple formula
+More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality of the sequence 
 \begin{align}
-	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x || x_t) \\
+	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x || x_t)
+\end{align}
+is upper bounded by the simple formula
+\begin{align}
 	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* || x_0) }{t} \; .
@@
 		0 
-		&\leq  \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
+		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
 		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
@@
 Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
-Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1} = \nabla \logpartition^*(\nabla \logpartition(\natp_t) - \gamma_t g_t )$.
+Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$.
 \begin{align}
 	\Cov(g_t,\natp_{t+1}) 
-	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \gamma_t \sigma^2, \forall t \\
+	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \lr_t \sigma^2, \forall t \\
 	 &= \expect{\langle g_t - \nabla f(\natp_t), \natp_{t+1}  - \natp \rangle}, (\forall \natp)
@@
 \begin{align}
-	g_t - \nabla f(\natp_t) = \E[T(X)] - T(x_{t+1})	
+	g_t - \nabla f(\natp_t) = \E[T(X)] - T(X_{t+1})	
 \end{align}
-so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable.
+so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
+\begin{align}
+	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
+	\leq \lr_t\sigma^2 \\
+	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
+\end{align}
+In other words, we need for all $\lr\in(0,1)$, and all $\mu$ in the interior of the marginal polytope,
+\begin{align}
+\label{eq:hypvariance}
+	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 
+	\leq \lr \sigma^2 \; .
+\end{align}
+When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanconj} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal poytope, we get
+\begin{multline}
+	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
+	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
+	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
+\end{multline}
+where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}.
+In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality.
+The question remains of : when is such a bound valid?
+For which values of $\sigma^2$.
+To answer such questions, we need to turn to examples.
 
@@
 \begin{align}
-    \expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}
-    &= \KL( p(.|\natp^*) || p(.|\natp)) \\
-    &= \bregman (\natp || \natp^*) \\
-    &= \bregmanconj ( \meanp^* || \meanp)
-    %& = \logpartition(\natp) - \logpartition(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
+    &\expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}\\
+    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
+    =& \bregman (\natp || \natp^*) \\
+    \label{eq:bregmanconj}
+    =& \bregmanconj ( \meanp^* || \meanp) \\
+    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
+    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
 \end{align}
 where $\conj$ is the entropy, the convex conjugate of the log-partition.
-The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. 
+The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. \RLP{add citations} 
 The question is: how does this quantity behave when $\natp$ is the MAP estimate ? Can we get bounds -- in expectation or high-probability ? 
@@
 
-\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpartition$ is defined within the $\ell^2$-ball of radius $L$), then
+\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
 \begin{align}
@@
 
-\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpartition$ is $L^{-1}$-strongly convex), then
+\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
 \begin{align}
@@
 \begin{align}
-    \logpartition(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
+    \logpart(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
 \end{align}
@@
 
-It is also very helpful for us, since we are looking at pairs $\logpartition, \conj$, and their associated Bregman divergences. 
-If $\logpartition$ is SC, then so is 
-$f(\natp)= \bregman(\natp || \natp^*) = \logpartition(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
+It is also very helpful for us, since we are looking at pairs $\logpart, \conj$, and their associated Bregman divergences. 
+If $\logpart$ is SC, then so is 
+$f(\natp)= \bregman(\natp || \natp^*) = \logpart(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
 And $\conj$ is SC as well, thus  $h(\meanp) = \bregmanconj ( \meanp || \meanp^*)$ is SC.
@@
 Our functions of interest is $f(\natp)= \bregman (\natp || \natp^*)= \bregmanconj ( \meanp^* || \meanp) = g(\meanp)$, with minimum $f^*=0$. 
-If $\logpartition$ is self-concordant, then so is $f$, but not necessarily $g$.
+If $\logpart$ is self-concordant, then so is $f$, but not necessarily $g$.
 The gradient and Hessian of $f$ are
@@
     f(\natp) 
-    &= \logpartition(\natp) - \logpartition(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
+    &= \logpart(\natp) - \logpart(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
     \nabla f(\natp) 
@@
     &= \nabla^2 \conj(\meanp) 
-    = \nabla^2 \logpartition(\natp) ^{-1}
+    = \nabla^2 \logpart(\natp) ^{-1}
     = \Cov_{p(X|\natp)}[T(X)]^{-1}\\

### 2020-11-27 main.tex
Move mirror descent forward and stuff to appendix

@@
 
+\section{Straightforward Convergence Rate}
+In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
+\begin{align}
+    &\expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}\\
+    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
+    =& \bregman (\natp || \natp^*) \\
+    =& \bregmanconj ( \meanp^* || \meanp) \\
+    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
+    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
+\end{align}
+where $\conj$ is the entropy, the convex conjugate of the log-partition.
+The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. \RLP{add citations} 
+The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability ?
+\begin{align}
+	\label{eq:bregmanMLE}
+	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \\
+	\label{eq:bregmanMAP}
+	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
+\end{align}
+
+\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
+\begin{align}
+    \bregmanconj(\mu^* || \mu) 
+    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
+    \leq 2L \norm{\mu^* - \mu}
+\end{align}
+so $\bregmanconj$ is $2L$-Lipschitz.
+Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this is the convergence rate of the log-likelihood.
+\RLP{find relevant inequalities}.
+
+\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
+\begin{align}
+    \bregmanconj(\mu^* || \mu) 
+    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
+\end{align}
+so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.
+
 \section{Stochastic Bregman Proximal Point}
@@
 
-\subsubsection{Three point lemma}
-These proofs rely on another basic property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
-\begin{align}
-	\min	_{x\in C} \overbrace{f(x) + \cB_h(x || y)}^{\phi(x)}
-\end{align}
-where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector.
-Then
-\begin{align}
-	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
-\end{align}
-This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).
-
-\begin{proof}
-	The first order optimality condition is
-	\begin{align}
-		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
-		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
-	\end{align}
-	This proof relies on another property called three point property 
-	\begin{align}
-		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
-	\end{align}
-	which can be proved by expanding the right hand side.
-	By convexity of $f$ we also have 
-	\begin{align}
-		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
-	\end{align} 
-	Putting it all together we get
-	\begin{align}
-		0 
-		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
-		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
-	\end{align}
-	which concludes the proof.
-\end{proof}
-
 \subsection{Relative smoothness for SMD}
@@
 \end{align}
-When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanconj} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal poytope, we get
+When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal poytope, we get
 \begin{multline}
@@
 Also \citet{raskutti2015information} only gives asymptotic results.
-
-\section{Straightforward Convergence Rate}
-In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
-\begin{align}
-    &\expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}\\
-    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
-    =& \bregman (\natp || \natp^*) \\
-    \label{eq:bregmanconj}
-    =& \bregmanconj ( \meanp^* || \meanp) \\
-    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
-    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
-\end{align}
-where $\conj$ is the entropy, the convex conjugate of the log-partition.
-The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. \RLP{add citations} 
-The question is: how does this quantity behave when $\natp$ is the MAP estimate ? Can we get bounds -- in expectation or high-probability ? 
-\begin{align}
-	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \\
-	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right )
-\end{align}
-
-\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
-\begin{align}
-    \bregmanconj(\mu^* || \mu) 
-    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
-    \leq 2L \norm{\mu^* - \mu}
-\end{align}
-so $\bregmanconj$ is $2L$-Lipschitz.
-Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this is the convergence rate of the log-likelihood.
-\RLP{find relevant inequalities}.
-
-\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
-\begin{align}
-    \bregmanconj(\mu^* || \mu) 
-    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
-\end{align}
-so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.
 
@@
 In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
-
-\subsection{Fenchel conjugate motivation to self-concordance}
-
-In the most regular case, when $f(x)$ is a convex function, continuously differentiable on its domain, then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ verifies
-\begin{align}
-    \nabla f \circ \nabla f^*  &= \Id \\
-     \nabla f^* \circ \nabla f &= \Id
-\end{align}
-where $\Id$ is the identity function on the relevant domain. In words, the gradients of $f$ and $f^*$ are reciprocal.
-Deriving this equality yields
-\begin{align}
-    \nabla^2 f(x) \nabla^2 f^*(x^*) = I_n
-\end{align}
-where $x,x^*$ are conjugate points -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
-Now, it gets interesting to us when we derive again this equality. Let's tackle the 1D case first
-\begin{align}
-    &f''(x) f^{* \prime \prime}(f'(x)) = 1, \forall x \\
-    \implies
-    &f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))  = 0 \\
-    \implies
-    &\frac{f'''(x)}{f''(x){\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*){\frac{3}{2}}}  = 0
-\end{align}
-where to get to the last line we used the first line, and we divided the second line by $f''(x)^{\frac{1}{2}}$. 
-We see that for a pair of conjugate functions, the self-concordance ratio is preserved, modulo the sign.
-This gives another rational, beyond dimensional analysis, for using this ratio as a regularity assumption for convex analysis.
-
-It is also very helpful for us, since we are looking at pairs $\logpart, \conj$, and their associated Bregman divergences. 
-If $\logpart$ is SC, then so is 
-$f(\natp)= \bregman(\natp || \natp^*) = \logpart(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
-And $\conj$ is SC as well, thus  $h(\meanp) = \bregmanconj ( \meanp || \meanp^*)$ is SC.
-But there is no reason for 
-$g(\meanp) = \bregmanconj ( \meanp^* || \meanp) =  \cst - \conj(\meanp) - \langle \nabla \conj (\meanp), \meanp^* - \meanp \rangle $ to be SC.
-
-The multivariate generalization of this formula is a third order tensor equality
-\begin{align}
-    \nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1} + \nabla^2 f^{* -\half} \nabla^3 f^* \nabla^2 f^{* -1} = 0
-\end{align}
-where we omit  multiplication axis and functions take relevant argument $x$ or $x^*$.
-Consequently, a multivariate definition of self-concordance might take the form of an inequality on the 3d tensor $\nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1}$.
 
@@
 \newpage
+\appendix
+
+\section{Three point lemma}
+Proofs of convergence for mirror descent with relative smoothness rely on a specific property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
+\begin{align}
+	\min	_{x\in C} \overbrace{f(x) + \cB_h(x || y)}^{\phi(x)}
+\end{align}
+where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector.
+Then
+\begin{align}
+	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
+\end{align}
+This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).
+
+\begin{proof}
+	The first order optimality condition is
+	\begin{align}
+		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
+		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
+	\end{align}
+	This proof relies on another property called three point property 
+	\begin{align}
+		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
+	\end{align}
+	which can be proved by expanding the right hand side.
+	By convexity of $f$ we also have 
+	\begin{align}
+		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
+	\end{align} 
+	Putting it all together we get
+	\begin{align}
+		0 
+		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
+		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
+	\end{align}
+	which concludes the proof.
+\end{proof}
+
+\section{Fenchel conjugate motivation to self-concordance}
+
+In the most regular case, when $f(x)$ is a convex function, continuously differentiable on its domain, then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ verifies
+\begin{align}
+    \nabla f \circ \nabla f^*  &= \Id \\
+     \nabla f^* \circ \nabla f &= \Id
+\end{align}
+where $\Id$ is the identity function on the relevant domain. In words, the gradients of $f$ and $f^*$ are reciprocal.
+Deriving this equality yields
+\begin{align}
+    \nabla^2 f(x) \nabla^2 f^*(x^*) = I_n
+\end{align}
+where $x,x^*$ are conjugate points -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
+Now, it gets interesting to us when we derive again this equality. Let's tackle the 1D case first
+\begin{align}
+    &f''(x) f^{* \prime \prime}(f'(x)) = 1, \forall x \\
+    \implies
+    &f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))  = 0 \\
+    \implies
+    &\frac{f'''(x)}{f''(x){\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*){\frac{3}{2}}}  = 0
+\end{align}
+where to get to the last line we used the first line, and we divided the second line by $f''(x)^{\frac{1}{2}}$. 
+We see that for a pair of conjugate functions, the self-concordance ratio is preserved, modulo the sign.
+This gives another rational, beyond dimensional analysis, for using this ratio as a regularity assumption for convex analysis.
+
+It is also very helpful for us, since we are looking at pairs $\logpart, \conj$, and their associated Bregman divergences. 
+If $\logpart$ is SC, then so is 
+$f(\natp)= \bregman(\natp || \natp^*) = \logpart(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
+And $\conj$ is SC as well, thus  $h(\meanp) = \bregmanconj ( \meanp || \meanp^*)$ is SC.
+But there is no reason for 
+$g(\meanp) = \bregmanconj ( \meanp^* || \meanp) =  \cst - \conj(\meanp) - \langle \nabla \conj (\meanp), \meanp^* - \meanp \rangle $ to be SC.
+
+The multivariate generalization of this formula is a third order tensor equality
+\begin{align}
+    \nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1} + \nabla^2 f^{* -\half} \nabla^3 f^* \nabla^2 f^{* -1} = 0
+\end{align}
+where we omit  multiplication axis and functions take relevant argument $x$ or $x^*$.
+Consequently, a multivariate definition of self-concordance might take the form of an inequality on the 3d tensor $\nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1}$.
 

### 2020-11-27 main.tex
Discuss ell 2 distance
Add bias-variance decomposition

@@
 so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.
+
+\paragraph{$\ell^2$-norm analysis}
+Let us make these statements more precise. 
+If $\Var T(X) = \sigma^2$, then 
+\begin{align}
+	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
+\end{align}
+eg the variance of the mean is $n$ times smaller than the variance of the samples. We can extend this result to a Manahalobis distance with any matrix $\mM$
+\begin{align}
+	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
+\end{align}
+where $\mSigma$ is the covariance of $T(X)$. This  bound is equal to $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$. 
+Adding a reference mean $\mu_0$ to get the MAP,
+\begin{align}
+	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
+	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
+	&= O\left(\frac{\sigma^2}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
+\end{align}
+so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !
 
@@
     \nabla^2 f(\natp) 
-    &= \Sigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
+    &= \mSigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
 \end{align}
@@
     \leq \decrement(\natp)^2 
-    = \norm{\mu^* - \mu(\theta)}^2_{\Sigma(\theta)^{-1}} 
+    = \norm{\mu^* - \mu(\theta)}^2_{\mSigma(\theta)^{-1}} 
     \leq 0.46
@@
 Following these equations, the joint distribution on $(\natp,X)$ factors along a larger graph $G'$ which augments $G$ by adding nodes $\natp_i$ with arrows pointing to $X_i$, as illustrated in Figure~\ref{fig:joint_graph}.
-\begin{figure}[h]
+\begin{figure}[ht]
     \centering

### 2020-11-27 main.tex
Change description of variance bound

@@
 Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
-Then the variance bound $\sigma^2$ is a lower bound on the covariance between $g_t$ and $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$.
+Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps
 \begin{align}
-	\Cov(g_t,\natp_{t+1}) 
-	&= \expect{\langle g_t, \natp_{t+1} \rangle}  - \langle \E[g_t], \E[\natp_{t+1}] \rangle \geq -  \lr_t \sigma^2, \forall t \\
-	 &= \expect{\langle g_t - \nabla f(\natp_t), \natp_{t+1}  - \natp \rangle}, (\forall \natp)
+	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t}) \leq \lr_t \sigma^2 \\
+	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
+	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
 \end{align}
-where $\gamma_t$ is the step-size and expectations are conditional on $\natp_t$. In words, $g_t$ and $\natp_{t+1}$ should not be too negatively correlated. In our case, 
+where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD. In our case, 
 \begin{align}
-	g_t - \nabla f(\natp_t) = \E[T(X)] - T(X_{t+1})	
+	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
 \end{align}

### 2020-11-27 main.tex
Bias-variance for Bregman

@@
 where $\conj$ is the entropy, the convex conjugate of the log-partition.
-The relationship between Bregman divergences and Fenchel conjugacy is well explained in Wainwright's book, and the article geometry of exponential family. \RLP{add citations} 
+The relationship between Bregman divergences and Fenchel conjugacy is well explained in \citet{wainwright2008graphical}, and \citet{agarwal2010geometric}. 
 The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability ?
@@
 Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this is the convergence rate of the log-likelihood.
-\RLP{find relevant inequalities}.
 
@@
 \end{align}
-When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal poytope, we get
+When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal polytope, we get
 \begin{multline}
@@
 In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality.
-The question remains of : when is such a bound valid?
-For which values of $\sigma^2$.
-To answer such questions, we need to turn to examples.
+We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias.
+Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
+\begin{align}
+	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) } 
+	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
+	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
+\end{align}
+so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
+\begin{align}
+	 \bregmanconj (\mu_* ; \tilde \mu)  
+	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
+\end{align}
+where $\mSigma_* = \Cov_{\mu_*}(T(X))$. \RLP{beware of the order, this might be inaccurate}
+However the scalar product is harder to bound.
+Hopefully it could be positive.
+If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$.
+Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.
+
+\paragraph{TODO}
+The question remains of : when is such a bound valid? For which values of $\sigma^2$. To answer such questions, we need to turn to examples.
 

### 2020-12-14 main.tex
Add  Bregman locally euclidean discussion

@@
 \end{align}
-eg the variance of the mean is $n$ times smaller than the variance of the samples. We can extend this result to a Manahalobis distance with any matrix $\mM$
-\begin{align}
-	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
-\end{align}
-where $\mSigma$ is the covariance of $T(X)$. This  bound is equal to $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$. 
-Adding a reference mean $\mu_0$ to get the MAP,
+eg the variance of the mean is $n$ times smaller than the variance of the samples.
+Adding a reference mean $\mu_0$ to get the MAP yields
 \begin{align}
@@
 so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !
+
+In the limit, the MAP estimates reach the Cramer-Rao lower bound. This can be seen by approximating the Bregman divergence using a second order Taylor expansion
+\begin{align}
+    \bregmanconj(\mu^* || \mu) 
+    &= (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
+    + O(\norm{\mu - \mu^*}^3)\\
+    &= \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
+    + O(\norm{\mu - \mu^*}^3)
+\end{align}
+where we introduced the Mahalanobis distance induced by the matrix 
+\begin{align}
+    \nabla^2\conj(\mu^*) 
+    = \nabla^2\logpart(\natp^*)^{-1} 
+    = \Cov_{\natp^*}[T(X)]^{-1} 
+    =: \mSigma^{-1}  \; .
+\end{align}
+To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
+\begin{align}
+	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
+\end{align}
+where $\mSigma$ is the covariance of $T(X)$.
+This is equal to $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.
+It is worth $\nicefrac{d}{n}$ when $\mM = \mSigma^{-1}$.
+\RLP{Maybe that is Cramer-Rao, but check again.} And for the MAP we  get
+
+\begin{align}
+	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
+	&= \frac{n}{(n+n_0)^2} \Tr(\mSigma^{-1}\mSigma) +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
+	&= O\left(\frac{d}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
+\end{align}
+Remark how the variance does not even appear, only the dimension divided by $n$ really matters.
+I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
+Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?
 

### 2020-12-14 main.tex
Minor changes

--- 2020-11-27 main.tex
@@
     = (n_0+n) \logpart (\natp) 
-    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(x_i) \right) \; .
+    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(x_i) \right) + \cst \; .
 \end{align}
@@
 \begin{align}
-    &\expect[x\sim p(.|\natp^*)]{-\log p(x|\natp) + \log p(x|\natp^*)}\\
+    &\expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*)}\\
     =& \KL( p(.|\natp^*) || p(.|\natp)) \\
@@
 The relationship between Bregman divergences and Fenchel conjugacy is well explained in \citet{wainwright2008graphical}, and \citet{agarwal2010geometric}. 
-The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability ?
+The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability -- on the following quantities ?
 \begin{align}
 	\label{eq:bregmanMLE}
-	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \\
+	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \; , \\
 	\label{eq:bregmanMAP}
-	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
+	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) \; .
 \end{align}
@@
 \end{align}
-In other words, we need for all $\lr\in(0,1)$, and all $\mu$ in the interior of the marginal polytope,
+In other words, we need for all $\lr\in(0,\inv{n})$, and all $\mu$ in the interior of the marginal polytope,
 \begin{align}
@@
 Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.
+
+This article \url{http://davidpfau.com/assets/generalized_bvd_proof.pdf} gives a bias variance decomposition which is applicable here.
 

### 2020-12-15 main.tex
Clarify intro and discuss proximal point

@@
 \section{Background}
-It's hard to find general convergence rate on the KL for the maximum likelihood estimates of the exponential family.  We want the simplest one.
+
+\paragraph{Motivation.} 
+We do not know general convergence rates on the KL for maximum likelihood estimates of the exponential family.  We want the simplest one.
 We hope  to get a new result by combining tools from statistics and optimization.
 
+\paragraph{Exponential Family.}
 The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model
@@
 The second identity entails that $\logpart$ is strictly-convex.
-The conjugate prior for this distribution is
+
+\paragraph{Conjugate Prior}
+The conjugate prior for $p(X|\natp)$ is
 \begin{align}
@@
 \end{align}
-where $n_0$ is a number of fictional points observed from a distribution with parameter $\natp_0$.
-$\bregman(\natp || \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
+where $n_0$ and $\natp_0$ are (hyper)parameters of the prior.
+Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
+Finally, $\bregman(\natp || \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
 \begin{align}
@@
     & = \logpart(\natp) - \logpart(\natp_0) 
-    - \langle \meanp_0 , \natp - \natp_0 \rangle
+    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
 \end{align}
-with $\meanp_0 = \nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)}$ the mean parameter associated to the natural parameter $\natp_0$.
+with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$.
+
+\paragraph{MAP.}
 The negative log-likelihood of the prior is then
@@
 \end{align}
-The MAP estimate is a random quantity.
+The MAP estimate is a random quantity. We wish to bound its deviation the optimum $\natp^*$ or $\meanp^*$.
 
@@
 so $\bregmanconj$ is $2L$-Lipschitz.
-Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this is the convergence rate of the log-likelihood.
+Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this bound applies to the log-likelihood.
 
@@
 \paragraph{$\ell^2$-norm analysis}
-Let us make these statements more precise. 
-If $\Var T(X) = \sigma^2$, then 
+Let us make these statements more precise. A bound on the variance becomes a bound on the variance of the average
 \begin{align}
-	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
+	\Var T(X) = \E \norm{T(X) - \mu^*}^2 = \sigma^2 \\
+	\implies \E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
 \end{align}
@@
     \bregmanconj(\mu^* || \mu) 
-    &= (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
+    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
     + O(\norm{\mu - \mu^*}^3)\\
-    &= \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
+    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
     + O(\norm{\mu - \mu^*}^3)
@@
     \hat \natp_{n+1} 
-    &= \argmin_\natp -\log p(x_{n+1}|\natp) + (n_0 + n) \bregman(\natp || \hat \natp_n) \\
-    &= \argmin_\natp f(\natp; x_{n+1}) +  \inv{\gamma_n}\bregman(\natp || \hat \natp_n) \; .
+    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp || \hat \natp_n) \\
+    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp || \hat \natp_n) \; .
 \end{align}
@@
 This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
+
+The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
+First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
+One needs an assumption on the quality of the stochastic estimates $f(.
+; x)$.
+\RLP{Write down the formulas going with this text.}
+Given the proof, the most straightforward such assumption is 
+\begin{align}
+\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
+\end{align}
+which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
+This is exactly the same assumption as given by the SMD analysis.
+However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
+It does not immediately apply to the last iterate.
 
@@
 
-\section{Three point lemma}
+\section{Three points lemma}
+\label{app:3points}
+
 Proofs of convergence for mirror descent with relative smoothness rely on a specific property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 

### 2020-12-15 main.tex
Add centered gaussians

@@
 \end{align}
-where $\mSigma_* = \Cov_{\mu_*}(T(X))$. \RLP{beware of the order, this might be inaccurate}
-However the scalar product is harder to bound.
+where $\mSigma_* = \Cov_{\mu_*}(T(X))$. 
+The order of terms $\bregmanconj (\mu_* ; \tilde \mu)  \neq \bregmanconj (\tilde \mu ; \mu_* ) $ only affects third order terms.
+That's for the bregman, but the scalar product is harder to bound.
 Hopefully it could be positive.
@@
 
-This article \url{http://davidpfau.com/assets/generalized_bvd_proof.pdf} gives a bias variance decomposition which is applicable here.
+This article \url{http://davidpfau.com/assets/generalized_bvd_proof.pdf} gives a bias variance decomposition which applies here. However it does not give interesting conclusions.
 
@@
 Also \citet{raskutti2015information} only gives asymptotic results.
+
+\section{Variance bound for centered gaussians}
+Sections on proximal point and mirror descent gave us the variance assumption 
+\begin{align}
+	\Cov(T(x_{n}) , \natp_{n+1}) = \Cov_X \bigg(T(X), \nabla \conj (\lr_n T(X) + (1-\lr_n)\mu_n)\bigg	) 
+	\leq \lr_n M \; .
+\end{align}
+on the sufficient statistic and $\conj$.
+Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$ with natural parameter $\lambda > 0$ the precision and sufficient statistic $T(X)= - \half x^2$.
+The mean parameter is $- \half \sigma^2$ and  the conjugate map is $\nabla \conj (- \half \sigma^2) = \inv{\sigma^2} = \lambda$.
+\begin{align}
+	\logpart (\lambda) = -\half \log(\lambda)  + \half \log \pi \\
+	\conj(-\half \sigma^2) = -\half \log(\sigma^2) + \half (\log\frac{\pi}{2} - 1) \\
+	\KL(\hat \sigma ; \sigma_*) = \half \log \frac{\hat \sigma^2}{\sigma_*^{2}} + \frac{\sigma_*^2 - \hat \sigma^2}{2 \sigma^2}
+\end{align}
+
+We want to bound
+\begin{align}
+	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
+	\approx \lr (1-\lr)^{-2} \left (\frac{\sigma_*}{\sigma_0} \right)^4
+\end{align}
+using a first order approximation of the inverse , eg a first order approximation of $\nabla \conj$. This can explode for small values of $\sigma_0$. In fact it does as illustrated experimentally.
 

### 2021-01-12 main.tex
Updates

@@
 	\label{eq:bregmanMLE}
-	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(x_i) \right ) \; , \\
+	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) \; , \\
 	\label{eq:bregmanMAP}
-	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) \; .
+	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right ) \; .
 \end{align}
@@
 \end{align}
-where $\mSigma$ is the covariance of $T(X)$.
-This is equal to $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.
-It is worth $\nicefrac{d}{n}$ when $\mM = \mSigma^{-1}$.
+where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
+\begin{align}
+	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
+\end{align}
 \RLP{Maybe that is Cramer-Rao, but check again.} And for the MAP we  get 
-
 \begin{align}
@@
 \end{align}
-In other words, we need for all $\lr\in(0,\inv{n})$, and all $\mu$ in the interior of the marginal polytope,
+In other words, we need for all $\lr\in(0,\inv{n})$, and $ \mu$ in the interior of the marginal polytope,
 \begin{align}
 \label{eq:hypvariance}
-	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr)\mu)\bigg	) 
+	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr) \mu)\bigg	) 
 	\leq \lr \sigma^2 \; .
@@
 	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
-	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
+	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}
+	 	\; \text{with}\; \mu=\inv{n-1}\smallsum_{j\neq i} T(x_j)  } ]  \leq \frac{\sigma^2}{n}
 \end{multline}
@@
 \begin{align}
-	\logpart (\lambda) = -\half \log(\lambda)  + \half \log \pi \\
-	\conj(-\half \sigma^2) = -\half \log(\sigma^2) + \half (\log\frac{\pi}{2} - 1) \\
-	\KL(\hat \sigma ; \sigma_*) = \half \log \frac{\hat \sigma^2}{\sigma_*^{2}} + \frac{\sigma_*^2 - \hat \sigma^2}{2 \sigma^2}
+	X \sim \cN(0,\sigma^2) \ ; \quad T(X)= - \half x^2 \ ; \quad  \mu = - \half \sigma^2  \\
+	\natp = \lambda = \inv{\sigma^2} =  \nabla \conj (- \half \sigma^2) >0 \\
+	\logpart (\lambda) = -\half \log(\lambda)  + \cst \\% \half \log \pi \\
+	\conj(-\half \sigma^2) = -\half \log(\sigma^2) + \cst \\%\half (\log\frac{\pi}{2} - 1) \\
+	\KL(\hat \sigma^2 ; \sigma_*^2) = \half \log \frac{\hat \sigma^2}{\sigma_*^{2}} + \frac{\sigma_*^2 - \hat \sigma^2}{2 \hat \sigma^2}
 \end{align}
@@
 
-\bibsection
 \bibliographystyle{apalike}

### 2021-01-21 main.tex
Updates from Overleaf
Clarify section on gaussian variance

@@
 	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
-	&= \frac{n}{(n+n_0)^2} \Tr(\mSigma^{-1}\mSigma) +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
+	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
 	&= O\left(\frac{d}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
@@
 In our setting this is not always true -- eg gaussians.
-However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity. $f$ is $\mu$ strongly-convex and $L$-smooth relative to $h$ if
+However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity.
 \begin{align}
-	& f-\mu h \ \text{ convex and } Lh - f \ \text{convex} \\
-	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y||x) \leq f(y) \\
-	&\text{and} \quad f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y||x), \forall x,y \\
-	\iff & \mu\cB_h(y||x) \leq \cB_f(y||x) \leq L \cB_h(y||x), \forall x,y \\
-	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
+	& \text{$f$ is $L$-smooth relative to $h$} \\
+	\iff & Lh - f \ \text{convex} \\
+	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y||x), \forall x,y \\
+	\iff & \cB_f(y||x) \leq L \cB_h(y||x), \forall x,y \\
+	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
 \end{align} 
 where the last equivalence holds only when $f$ and $h$ are twice differentiable.
-This sweet generalization of smoothness and strong-convexity transfers the Loewner partial order on symmetric matrices to functions, via the Hessian. 
+In words, $f$ is upper bounded by it linear approximation plus the $h$-Bregman divergence, which can also be seen as a bound between divergences, or more locally as a bound between Hessians.
+Similarly, $f$ is $\mu$ strongly-convex relative to $h$ if
+\begin{align}
+	& f-\mu h \ \text{ convex} \\
+	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y||x) \leq f(y)\forall x,y \\
+	\iff & \mu\cB_h(y||x) \leq \cB_f(y||x) \forall x,y \\
+	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
+\end{align}
+
+Another way to view this elegant generalization of smoothness and strong-convexity is as a transfer of the Loewner partial order on symmetric matrices to functions, via the Hessian. 
 As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
@@
 \end{multline}
+\textcolor{red}{what about:}
+\begin{multline}
+	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
+	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
+	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) \textcolor{red}{- \hat \mu_{n-1} + \hat \mu_{n-1}} - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
+\end{multline}
 where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. 
@@
 on the sufficient statistic and $\conj$.
-Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$ with natural parameter $\lambda > 0$ the precision and sufficient statistic $T(X)= - \half x^2$.
-The mean parameter is $- \half \sigma^2$ and  the conjugate map is $\nabla \conj (- \half \sigma^2) = \inv{\sigma^2} = \lambda$.
+Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
+The density of a centered normal variable is
 \begin{align}
-	X \sim \cN(0,\sigma^2) \ ; \quad T(X)= - \half x^2 \ ; \quad  \mu = - \half \sigma^2  \\
-	\natp = \lambda = \inv{\sigma^2} =  \nabla \conj (- \half \sigma^2) >0 \\
-	\logpart (\lambda) = -\half \log(\lambda)  + \cst \\% \half \log \pi \\
-	\conj(-\half \sigma^2) = -\half \log(\sigma^2) + \cst \\%\half (\log\frac{\pi}{2} - 1) \\
-	\KL(\hat \sigma^2 ; \sigma_*^2) = \half \log \frac{\hat \sigma^2}{\sigma_*^{2}} + \frac{\sigma_*^2 - \hat \sigma^2}{2 \hat \sigma^2}
+	X \sim \cN(0,\sigma^2) \implies p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
+\end{align}
+Using the $X^2$ as a sufficient statistic we get the variance as a mean parameter and minus half the precision as a natural parameter
+\begin{align}
+	\quad T(X)= x^2 \ ; \quad  \mu = \sigma^2 \ ; \quad   \natp = -\inv{2 \sigma^2} = -\half \lambda \; .
+\end{align}
+Nw we can match the log-likelihood with the exponential family template to get the log-partition function.
+\begin{align}
+	-\log p(x) = \frac{x^2}{2 \sigma^2} - \half \log(2\inv{2\sigma^2}) + \half \log(2 \pi) 
+	= -x^2 \lambda + \logpart(\lambda) \\
+	\implies \logpart (\lambda) = -\half \log(-\lambda)  + \half \log(\pi)
+\end{align}
+We can then use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
+\begin{align}
+	\conj(\sigma^2) = -\half \log(\sigma^2) + \half \log\frac{\pi}{2} - \half \\
+\end{align}
+Remark that both the entropy and the log-partition have the same shape $z\mapsto -\inv{2z}$. Which yields the same Bregman divergence
+\begin{align}
+	\KL(\hat \sigma^2 ; \sigma_*^2) = \half \log \frac{\hat \sigma^2}{\sigma_*^{2}} + \frac{\sigma_*^2 - \hat \sigma^2}{2 \hat \sigma^2} \; .
 \end{align}

### 2021-01-25 main.tex
Update section on centered gaussian

@@
 \newcommand{\violet}[1]{\textcolor{violet}{#1}}
+\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}
 
@@
 \begin{align}
-	\quad T(X)= x^2 \ ; \quad  \mu = \sigma^2 \ ; \quad   \natp = -\inv{2 \sigma^2} = -\half \lambda \; .
+	\quad T(X)= x^2 \ ; \quad  \mu = \sigma^2 \ ; \quad   \natp = -\inv{2 \sigma^2} = -\inv{2 \mu}\; .
 \end{align}
-Nw we can match the log-likelihood with the exponential family template to get the log-partition function.
+Now we can match the log-likelihood with the exponential family template to get the log-partition function.
 \begin{align}
 	-\log p(x) = \frac{x^2}{2 \sigma^2} - \half \log(2\inv{2\sigma^2}) + \half \log(2 \pi) 
-	= -x^2 \lambda + \logpart(\lambda) \\
-	\implies \logpart (\lambda) = -\half \log(-\lambda)  + \half \log(\pi)
+	= -x^2 \natp + \logpart(\natp) \\
+	\implies \logpart (\natp) = -\half \log(-\natp)  + \half \log(\pi)
 \end{align}
@@
 \begin{align}
-	\conj(\sigma^2) = -\half \log(\sigma^2) + \half \log\frac{\pi}{2} - \half \\
+	\conj(\mu) = -\half \log(\mu) + \half \log\frac{\pi}{2} - \half \\
 \end{align}
-Remark that both the entropy and the log-partition have the same shape $z\mapsto -\inv{2z}$. Which yields the same Bregman divergence
+Remark that both the entropy and the log-partition have the same shape $z\mapsto -\inv{2z}$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
 \begin{align}
-	\KL(\hat \sigma^2 ; \sigma_*^2) = \half \log \frac{\hat \sigma^2}{\sigma_*^{2}} + \frac{\sigma_*^2 - \hat \sigma^2}{2 \hat \sigma^2} \; .
+	\KL( \sigma_n^2 ; \sigma_*^2) 
+	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
+	\bregmanconj( \mu_*; \mu_n) 
+	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
+	\bregman( \natp_n; \natp_* ) 
+	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
 \end{align}
-
-We want to bound
+In other words, this is divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi(z)= \half (z - 1 - \log(z))$.
+\begin{align}
+	\phi(z) := \half (z - 1 - \log(z)) 
+	\quad \text{ and } \quad
+	\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  \\
+	\implies \bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
+\end{align}
+The covariance  \eqref{eq:hypvariance} we want to bound is equal to
 \begin{align}
@@
 \end{align}
-using a first order approximation of the inverse , eg a first order approximation of $\nabla \conj$. This can explode for small values of $\sigma_0$. In fact it does as illustrated experimentally.
+using a first order approximation of the inverse, eg a first order approximation of $\nabla \conj$ that holds when $\lr X^2 \ll (1-\lr) \sigma_0^2$.
+This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.
+
+\paragraph{Prior} A conjugate prior over the natural parameter $\natp$ is 
+\begin{align}
+	-\log \pi(\natp)
+	&= n_0 \bregman(\natp ; \natp_0) + \cst \\
+	&= n_0 \logpart(\natp) - n_0\mu_0 \natp + \cst \\
+	&= -\half[n_0] \log(-\natp) + n_0\mu_0 (-\natp) + \cst \; . \\
+	\pi(\natp) &= (-\natp)^{1 + \half[n_0] - 1} e^{- n_0 \mu_0 (-\natp)} / Z
+\end{align}
+In words, $-\natp$ is a random variable from the exponential family with sufficient statistics $(-\natp, \log(-\natp))$ and with natural parameters $(-n_0 \mu_0, \half[n_0])$.
+This is the definition of a gamma distribution with shape parameter $1 + \half[n_0]$ and with rate parameter $n_0 \mu_0$.
 

### 2021-01-26 main.tex
Try fixing UTF8 error
Fix UTF8 error
Use semi colon instead of || in all divergences

@@
 \begin{align}
-    \pi(\natp) &\propto \exp( - n_0 \bregman(\natp || \natp_0) )
+    \pi(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
 \end{align}
@@
 Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
-Finally, $\bregman(\natp || \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
+Finally, $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
 \begin{align}
-    \bregman (\natp || \natp_0)
+    \bregman (\natp ; \natp_0)
     & = \logpart(\natp) - \logpart(\natp_0) 
@@
 \begin{align}
-    \hat \natp = \argmin_\natp -\log p(X|\natp) + n_0 \bregman(\natp || \natp_0)
+    \hat \natp = \argmin_\natp -\log p(X|\natp) + n_0 \bregman(\natp ; \natp_0)
 \end{align}
@@
     &\expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*)}\\
-    =& \KL( p(.|\natp^*) || p(.|\natp)) \\
-    =& \bregman (\natp || \natp^*) \\
-    =& \bregmanconj ( \meanp^* || \meanp) \\
+    =& \KL( p(.|\natp^*) ; p(.|\natp)) \\
+    =& \bregman (\natp ; \natp^*) \\
+    =& \bregmanconj ( \meanp^* ; \meanp) \\
     =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
@@
 \begin{align}
-    \bregmanconj(\mu^* || \mu) 
+    \bregmanconj(\mu^* ; \mu) 
     \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
@@
 \begin{align}
-    \bregmanconj(\mu^* || \mu) 
+    \bregmanconj(\mu^* ; \mu) 
     \leq \frac{L}{2} \norm{\mu^* - \mu}^2
@@
 \begin{align}
-    \bregmanconj(\mu^* || \mu) 
+    \bregmanconj(\mu^* ; \mu) 
     &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
@@
 
+\subsection{Posterior Expected Loss}
+
+A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is 
+\begin{align}
+    \pi(\natp) 
+    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
+    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle)
+\end{align}
+so we see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
+When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
+This is well detailed in \citet{agarwal2010geometric}.
+
+Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
+\begin{align}
+    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+\end{align}
+
 \section{Stochastic Bregman Proximal Point}
-We see that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp || \natp_0)$.
+We see that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
 This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
@@
     \hat \natp_{n+1} 
-    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp || \hat \natp_n) \\
-    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp || \hat \natp_n) \; .
+    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
+    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
 \end{align}
@@
 	\natp_{t+1} 
-	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp || \natp_t)  \\
+	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
 	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
@@
     &= \argmin_\natp 
-    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp || \natp_n) \\
+    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
     %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
-    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp || \natp_n) \\
-    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp || \natp_n)
+    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
+    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
 \end{align}
@@
 	\iff & Lh - f \ \text{convex} \\
-	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y||x), \forall x,y \\
-	\iff & \cB_f(y||x) \leq L \cB_h(y||x), \forall x,y \\
+	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y;x), \forall x,y \\
+	\iff & \cB_f(y;x) \leq L \cB_h(y;x), \forall x,y \\
 	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
@@
 	& f-\mu h \ \text{ convex} \\
-	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y||x) \leq f(y)\forall x,y \\
-	\iff & \mu\cB_h(y||x) \leq \cB_f(y||x) \forall x,y \\
+	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y;x) \leq f(y)\forall x,y \\
+	\iff & \mu\cB_h(y;x) \leq \cB_f(y;x) \forall x,y \\
 	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
@@
 \begin{align}
-	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x || x_t)
+	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x ; x_t)
 \end{align}
@@
 \begin{align}
-	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* || x_0) }{t} \; .
+	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* ; x_0) }{t} \; .
 \end{align}
@@
 \begin{multline}
-	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) }
+	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) }
 	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
@@
 \begin{align}
-	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ||\hat \mu) } 
+	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) } 
 	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
@@
 \begin{align}
-	\quad T(X)= x^2 \ ; \quad  \mu = \sigma^2 \ ; \quad   \natp = -\inv{2 \sigma^2} = -\inv{2 \mu}\; .
+	\quad T(X)= X^2 \ ; \quad  \mu = \sigma^2 \ ; \quad   \natp = -\inv{2 \sigma^2} = -\inv{2 \mu}\; .
 \end{align}
@@
 	&= -\half[n_0] \log(-\natp) + n_0\mu_0 (-\natp) + \cst \; . \\
-	\pi(\natp) &= (-\natp)^{1 + \half[n_0] - 1} e^{- n_0 \mu_0 (-\natp)} / Z
+	\pi (\natp) &= (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
 \end{align}
@@
 
-Our functions of interest is $f(\natp)= \bregman (\natp || \natp^*)= \bregmanconj ( \meanp^* || \meanp) = g(\meanp)$, with minimum $f^*=0$. 
+Our functions of interest is $f(\natp)= \bregman (\natp ; \natp^*)= \bregmanconj ( \meanp^* ; \meanp) = g(\meanp)$, with minimum $f^*=0$. 
 If $\logpart$ is self-concordant, then so is $f$, but not necessarily $g$.
@@
 \begin{align}
-    \bregmanconj(\mu^* || \mu) 
+    \bregmanconj(\mu^* ; \mu) 
     \leq \decrement(\natp)^2 
@@
 
-Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp || \meanp^*)$, then we would get
+Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$, then we would get
 \begin{align}
@@
 \begin{align}
-	\min	_{x\in C} \overbrace{f(x) + \cB_h(x || y)}^{\phi(x)}
+	\min	_{x\in C} \overbrace{f(x) + \cB_h(x ; y)}^{\phi(x)}
 \end{align}
@@
 If $\logpart$ is SC, then so is 
-$f(\natp)= \bregman(\natp || \natp^*) = \logpart(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
-And $\conj$ is SC as well, thus  $h(\meanp) = \bregmanconj ( \meanp || \meanp^*)$ is SC.
+$f(\natp)= \bregman(\natp ; \natp^*) = \logpart(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
+And $\conj$ is SC as well, thus  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$ is SC.
 But there is no reason for 
-$g(\meanp) = \bregmanconj ( \meanp^* || \meanp) =  \cst - \conj(\meanp) - \langle \nabla \conj (\meanp), \meanp^* - \meanp \rangle $ to be SC.
+$g(\meanp) = \bregmanconj ( \meanp^* ; \meanp) =  \cst - \conj(\meanp) - \langle \nabla \conj (\meanp), \meanp^* - \meanp \rangle $ to be SC.
 

### 2021-01-26 main.tex
Add Cramer-Rao discussion

@@
 
+\usepackage[top=2cm, bottom=2cm]{geometry}
 \usepackage[utf8]{inputenc} % allow utf-8 input
@@
 
-In the limit, the MAP estimates reach the Cramer-Rao lower bound. This can be seen by approximating the Bregman divergence using a second order Taylor expansion
+In the limit, the MAP estimates reach a rate $O(1/2n)$. This is shown by approximating the Bregman divergence using a second order Taylor expansion
 \begin{align}
@@
 \end{align}
-\RLP{Maybe that is Cramer-Rao, but check again.} And for the MAP we  get 
+and for the MAP we  get 
 \begin{align}
@@
 I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
+Indeed this is because it is hidden within the Bregman divergence loss. 
 Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?
+
+\subsection{Cramer-Rao Lower Bound}
+In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
+In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
+\begin{align}
+    \Cov(\hat \natp_n)  
+    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
+    \iff
+    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
+    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
+\end{align}
+This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance.
+In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"
 
@@
 \end{align}
+This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral.
+We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
+This is exactly twice the expected loss of the MAP.
 
@@
 In words, $-\natp$ is a random variable from the exponential family with sufficient statistics $(-\natp, \log(-\natp))$ and with natural parameters $(-n_0 \mu_0, \half[n_0])$.
-This is the definition of a gamma distribution with shape parameter $1 + \half[n_0]$ and with rate parameter $n_0 \mu_0$.
+This is the definition of a Gamma distribution with shape parameter $1 + \half[n_0]$ and with rate parameter $n_0 \mu_0$.
 

### 2021-01-28 main.tex
Add related work section

@@
 
+Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
+For the estimator $\hat \natp_n$, this lower bound is not trivial.
+
 \subsection{Posterior Expected Loss}
@@
 
-Also \citet{raskutti2015information} only gives asymptotic results.
-
 \section{Variance bound for centered gaussians}
@@
 This is just a remark. I don't think it can help us to get anywhere.
+
+\section{Related Work}
+Exponential families became mainstream in the machine learning community to generalize frameworks such as linear regression \citep{mccullagh1989generalized} or PCA \citep{collins2001generalization} to diverse data types and distributions.
+\citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure. 
+\citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem.
+They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
+Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors. 
+\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.
+Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
+\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
+Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
 

### 2021-01-28 main.tex
Expand related work

@@
 
-\paragraph{TODO}
-The question remains of : when is such a bound valid? For which values of $\sigma^2$. To answer such questions, we need to turn to examples.
+\paragraph{Open questions}
+The question remains of : when is such a bound valid?  what does $\sigma^2$ depend on ? To answer these questions and more, we need to turn to examples, see  Section~\ref{sec:gaussian-variance}.
 
-The other article \citet{lu2019relative} derives another convergence rate for SMD under another gradient variance assumption.
+\TODO{This bound might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
+\citet{lu2019relative} derive another convergence rate for SMD under another relative continuity and gradient variance assumption. Can we exploit it ?}
 
 \section{Variance bound for centered gaussians}
+\label{sec:gaussian-variance}
 Sections on proximal point and mirror descent gave us the variance assumption 
@@
 \section{Related Work}
-Exponential families became mainstream in the machine learning community to generalize frameworks such as linear regression \citep{mccullagh1989generalized} or PCA \citep{collins2001generalization} to diverse data types and distributions.
+Exponential families are a mainstream tool in machine learning.
+They are used  to generalize linear regression \citep{mccullagh1989generalized}, PCA \citep{collins2001generalization} or k-means \citep{banerjee2005clustering} to diverse data types and distributions.
 \citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure. 
@@
 They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
-Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors. 
+Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors yet critical as it does not hold for multivariate normals with unknown covariance.
 \citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.

### 2021-02-09 main.tex
Add discussion of relative continuity

@@
 \end{align}
+\RLP{Add details for 11 12}
 where $\conj$ is the entropy, the convex conjugate of the log-partition.
@@
 
-This is very interesting for us, but the variance hyper-parameter is oddly defined.
+This is very interesting to us.
+Let us have a look at their variance assumption \citep[Assumption 5.1]{hanzely2018fastest}.
 Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
@@
 
+\paragraph{Relative Continuity}
+\citet{lu2019relative} define relative continuity to tackle non-differentiable functions -- eg Hinge loss.
+These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
+Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and $\ell^2$ distance.
+Although this hypothesis is far from intuitive, it does yield convergence rates very similar to typical SGD analysis.
+In this work we are first concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
+
 \paragraph{Open questions}
@@
 
-\TODO{This bound might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
-\citet{lu2019relative} derive another convergence rate for SMD under another relative continuity and gradient variance assumption. Can we exploit it ?}
+\TODO{This bound might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?}
 

### 2021-02-09 main.tex
Updates from Overleaf

--- 2021-01-28 main.tex
@@
 \end{align}
+We study this assumption in Section \ref{sec:gaussian_variance} where we show that it cannot hold in general for the $\mu$ we care about. 
 When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal polytope, we get
@@
 \section{Variance bound for centered gaussians}
+\label{sec:gaussian_variance}
+
 Sections on proximal point and mirror descent gave us the variance assumption 
@@
 This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.
+This is a problem, because we need this bound to hold  for all values of $\sigma_0$ in the support of $X^2$, meaning on the whole $[0,+\infty[$, which is impossible because $1/X^2$ diverges, so we need to process this integral differently.
 

### 2021-02-23 main.tex
Add alternative research tracks

--- 2021-02-09 main.tex
@@
 \end{align}
+We study this assumption in Section \ref{sec:gaussian_variance} where we show that it cannot hold in general for the $\mu$ we care about. 
+\RLP{should we get a bound that depends on mu instead ? or split between $\mu$ smaller or larger than $\varepsilon$}
 When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal polytope, we get
@@
 \section{Variance bound for centered gaussians}
-\label{sec:gaussian-variance}
+\label{sec:gaussian_variance}
 Sections on proximal point and mirror descent gave us the variance assumption 
@@
 This is just a remark. I don't think it can help us to get anywhere.
+
+\section{Alternative Research Track}
+Weakly supervised : missing part of $x_i$ for each $i$.
+
+Structured prediction : graphical model on $X$.
 

### 2021-02-23 main.tex
new stuff

--- 2021-02-09 main.tex
@@
 \renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsection*{\bibname}}
+\renewcommand{\bibsection}{\subsection*{\bibname}}ni
 
@@
 \end{align}
-Recall that $\logpart$ verifies the two following identities 
+Recall that $\logpart$ verifies the two following identities.
 \begin{align}
@@
 
+\subsection{A big calculus in this specific setting}
+
+Indeed we know the distribution of $\mu$ it is linked with the chi-square distribution.
+We also know the expression of the excess of risk.
+So we could do a big calculus to get the exact expression of the averaged excess of risk.
+
+Recall the expression of the Gamma integral
+\begin{align}
+  \Gamma(\alpha):=\int_0^\infty t^{\alpha-1} \exp(-t) d t
+\end{align}
+Recall also that
+\begin{align}
+    \int x^q \exp(-nx^p) dx
+    &= n^{-(q+1)/p} \int (n^{1/p}x)^q \exp(-(n^{1/p}x)^p) d(n^{1/p}x)\\
+    &= n^{-(q+1)/p} \int u^q \exp(-u^p) du\\
+    &= n^{-(q+1)/p} \int t^{q/p} \exp(-t) d(t^{1/p})\\
+    &= n^{-(q+1)/p} \int p^{-1} t^{(q+1)/p - 1} \exp(-t) dt\\
+    &= n^{-(q+1)/p} p^{-1} \Gamma((q+1)/p).
+\end{align}
+Finally if there is problem in zero with the logarithm appearing in Eq. (75, at time of writing), we can have a Taylor expansion of the log 
+\[
+  \log(x) = \sum_{i=1}^\infty \frac{(-1)^{i+1}(x-1)^i}{i},
+\]
+and show convergence with some Fubini's theorem on series-integral.
+
+\subsection{Generic approach}
+
+In this subsection, I detailed a generic approach to get non-asymptotic rates.
+It is based on the calibration inequality Eq. (18, at time of writing).
+The next step would be to detailed those derivations and have precise results for the specific case of only looking at the variance.
+
+\paragraph{Calibration inequality.}
+One of the main difficulty of the proof is that $\theta_n = (\nabla A)^{-1}(\mu_n)$ can be really big, meaning that we can not apply directly ``calibration inequality'' between the Bregman divergence ${\cal B}_{A^*}(\mu^*; \mu)$ and $\norm{\mu - \mu^*}^2$.
+However, if we restrict ourselves to the domain $\theta'\in {\cal B}(0, R)$ with $R = \max(\norm{\theta}, \norm{\theta^*})$, we have that $A$ is $2^{-1}R^{-2}$-strongly convex.
+Therefore using Eq.~(18, at the time of writing), we have
+\begin{align}
+    {\cal B}_{A^*}(\mu^*; \mu) 
+    &\leq 2^{-1}R^{-2} \norm{\mu^* - \mu}^2\\
+    &= 2 \max(\norm{\theta}, \norm{\theta^*})^2\norm{\mu^* - \mu}^2\\
+    &= 2 \max(\norm{(\nabla A)^{-1}(\mu)}, \norm{(\nabla A)^{-1}(\mu^*)})^2\norm{\mu^* - \mu}^2.
+\end{align}
+With will proceed with this equation, we might get better equations.
+For example by considering a constant of strong convexity on the segment $[\norm{\theta}, \norm{\theta^*}]$ rather than the ball ${\cal B}(0, R)$.
+The behavior of the max, will be similar to the behavior of the sum of the two elements since $\max(a, b) \leq a + b \leq 2\max(a, b)$, therefore, we proceed
+with
+\begin{align}
+    {\cal B}_{A^*}(\mu^*; \mu) 
+    &\leq 2 \norm{(\nabla A)^{-1}(\mu^*)}^2 \norm{\mu^* - \mu}^2
+    + 2 \norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu^* - \mu}^2.
+\end{align}
+The first term is easy to control in expectation, see Eqs. (19-20, at time of writing), it is $\Var{T(X)} / n$. It is really the second part that we want to study. 
+Because of the structure $\norm{(\nabla A)^{-1}(\mu)} \simeq \norm{\mu}^{-1}$
+We have to take care of the event ``$\mu = n^{-1} \sum_{i\leq n} T(X_i)$ too small''.
+We know that $\mu$ concentrate towards $\mu^*$ so this event should not happen too often.
+So we want to study $\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2$.
+
+\paragraph{Tail bound or average excess of risk?}
+If we think with concentration inequality, we will get tail bound on the excess of risk, and not an average excess of risk.
+However, recall that
+\begin{align}
+  \E[X] = \E[\int_{0}^{\infty} 1_{X > t} dt]
+  = \int_{0}^\infty \E[1_{X>t}] dt
+  = \int_0^\infty \mathbb{P}(X > t) dt.
+\end{align}
+So we can translate tail bound into average.
+
+\paragraph{Concentration of a product.}
+We want to study $\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2$, it is a product of two term, we can proceed with the fact that if $A$ and $B$ are two random variables from a sample space $(\Omega, \mu)$ to $\mathbb{R}$, we have
+\[
+    \{ \omega \in \Omega \vert A(\omega) B(\omega) > t\}
+    \subset 
+    \{ \omega \in \Omega \vert A(\omega) > s\}
+    \cup
+    \{ \omega \in \Omega \vert A(\omega) > t/s\}
+\]
+Which results from the fact that if $ab > t$ than either $a > s$, either $b > t/s$.
+In our case, it means that for any $t > 0$, we have the deviation bound
+\begin{align}
+  \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2 > t)
+  &\leq \inf_{s>0} \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)}^2 > 1/s)
+  + \mathbb{P}(\norm{\mu - \mu^*}^2 > st)\\
+  &= \inf_{s>0} \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > s^{1/2})
+  + \mathbb{P}(\norm{\mu - \mu^*} > (st)^{1/2}).
+\end{align}
+So we are left with the study of 
+\begin{align}
+  \mathbb{P}(\norm{\mu - \mu^*} > t)
+  \qquad\text{and}\qquad 
+  \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > t).
+\end{align}
+
+\paragraph{Concentration on $\mu - \mu^*$.}
+Most likely $T(X)$ is sub-Gaussian (do hand-derivations or look at works like \url{https://arxiv.org/pdf/1011.3027.pdf} or \url{https://arxiv.org/pdf/1902.03736.pdf}) and Bernstein concentration inequality leads to a tail bound of the type, with $\sigma^2$ a variance parameter
+($\sigma^2 = \Var_{\theta^*}(T(X))$)
+\begin{align}
+    \mathbb{P}_{(X_i)_{i\leq n}}(\norm{\mu - \mu^*} > t) \leq \exp(-n t^2 / \sigma^2).
+\end{align}
+Normally, in Bernstein inequality there is an additional term in $n$, but the convergence rates will not depend on this additional term, so I am omitting it for simplicity.
+Note that when we will integrate that we will get
+\begin{align}
+    \int_0^\infty \mathbb{P}_{(X_i)_{i\leq n}}(\norm{\mu - \mu^*} > t^{-1/2}) dt
+    &=\int_0^\infty \exp(-n t / \sigma^2) dt\\
+    &= \sigma^2 n^{-1}\int_0^\infty \exp(- nt / \sigma^2) d(nt\sigma^{-2})\\
+    &= \sigma^2 n^{-1} \int_0^\infty \exp(-x) dx =\sigma^2 n^{-1}.
+\end{align}
+
+\paragraph{Probability that $\mu$ is close to zero.}
+Intuitively, we should study $\mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > t)$, which should relates to $\mathbb{P}(\norm{\mu} < t)$, which can be express exactly as we know the cumulative distribution function of the chi-square and that $\mu$ is a variant of the chi-square.
+It seems to me that here we will not get optimal rates, because the calibration inequality is too harsh when $\mu$ goes to $0$, as I develop in the next paragraph.
+
+\paragraph{Big picture.}
+Intuitively, we have summarize in the part in $\mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} < t)$, the problem that arise with ${\cal B}_{A^*}(\mu^*; \mu)$ when $\mu$ is small. 
+We could make this intuition more formal with
+\begin{align}
+   {\cal B}_{A^*}(\mu^*; \mu)
+   &= 1_{\norm{\mu} \geq s} {\cal B}_{A^*}(\mu^*; \mu)
+   + 1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu)\\
+   &\leq \sup_{\norm{\mu'} \geq s}(\norm{(\nabla A)^{-1}(\mu')})^2 \norm{\mu - \mu^*}^2
+   + 1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu).
+\end{align}
+Suppose the $\sup$ is finite, the first term is easy to control.
+The second term is what we should take care of. We have
+\begin{align}
+    1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu) 
+    &=  1_{\norm{\mu} < s} (A^*(\mu^*) - A^*(\mu) - <\nabla A^*(\mu), \mu^* - \mu>)\\
+    &\leq  1_{\norm{\mu} < s} (A^*(\mu^*) - \inf_{\norm{\mu'} < s} A^*(\mu') + \norm{\nabla A^*(\mu)} \sup_{\norm{\mu''} < s} \norm{\mu^* - \mu''}).
+\end{align}
+We know that, under some strict convexity condition, $\nabla A^* = (\nabla A)^{-1}$ (similar to Eq. (25) at time of writing). Therefore we really have to study, with harsh notation
+\begin{align}
+    \int_{0}^s \norm{(\nabla A)^{-1}(\mu)} d\mathbb{P}(\norm{\mu} < t)
+    =\int_{0}^s \norm{(\nabla A)^{-1}(\mu)} \mathbb{P}(\norm{\mu} \in [t, t+dt]).
+\end{align}
+It seems to me that with the precedent development we were implicitly studying
+\begin{align}
+    \int_{0}^s \norm{(\nabla A)^{-1}(\mu)}^2 d\mathbb{P}(\norm{\mu} < t),
+\end{align}
+which of course is worse.
+The study can be done precisely as we know the distribution of $\mu$.
+In terms of rates, in my current view, this path is tight if and only if $<\nabla A^*(\mu), \mu^* - \mu>$ do behave like $\norm{(\nabla A)^{-1}(\mu)}$ towards $0$.
+In terms of constants, we have done pretty harsh majoration in order to get back to $c\norm{\mu-\mu^*}$ with constants $c$ that are clearly suboptimal.
+
+\paragraph{Tail bound or average risk?}
+Note that, if $\mu$ is sub-Gaussian concentrating towards $\mu^*$,
+we can give a simpler result than average risk, 
+which is that we can that that 
+with proba $1 - \mathbb{P}(\norm{\mu} < 1/2\norm{\mu^*}) = 1 - \exp(-c\norm{\mu^*}^2 n)$, 
+we have a tail bound $\mathbb{P}(\norm{\mu}^{-2}\norm{\mu - \mu^*}^2 > t) = \exp(-cnt / \norm{\mu^*})$.
+This means that we can say that 
+with proba $1 - \exp(-c\norm{\mu^*}^2 n) - \delta$ for any $\delta >0$, 
+we have that $B_{A^*}(\mu^*; \mu) < cn^{-1}\log(1/\delta) / \norm{\mu^*}$.
+
 \section{Self-Concordance}

### 2021-02-23 main.tex

@@
 We study this assumption in Section \ref{sec:gaussian_variance} where we show that it cannot hold in general for the $\mu$ we care about. 
+\RLP{should we get a bound that depends on mu instead ? or split between $\mu$ smaller or larger than $\varepsilon$}
 When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal polytope, we get
@@
 
-\paragraph{TODO}
-The question remains of : when is such a bound valid? For which values of $\sigma^2$. To answer such questions, we need to turn to examples.
+\paragraph{Relative Continuity}
+\citet{lu2019relative} define relative continuity to tackle non-differentiable functions -- eg Hinge loss.
+These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
+Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and $\ell^2$ distance.
+Although this hypothesis is far from intuitive, it does yield convergence rates very similar to typical SGD analysis.
+In this work we are first concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
 
-The other article \citet{lu2019relative} derives another convergence rate for SMD under another gradient variance assumption.
+\paragraph{Open questions}
+The question remains of : when is such a bound valid?  what does $\sigma^2$ depend on ? To answer these questions and more, we need to turn to examples, see  Section~\ref{sec:gaussian-variance}.
+
+\TODO{This bound might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?}
 
@@
 \label{sec:gaussian_variance}
-
 Sections on proximal point and mirror descent gave us the variance assumption 
@@
 
+\section{Alternative Research Track}
+Weakly supervised : missing part of $x_i$ for each $i$.
+
+Structured prediction : graphical model on $X$.
+
 \section{Related Work}
-Exponential families became mainstream in the machine learning community to generalize frameworks such as linear regression \citep{mccullagh1989generalized} or PCA \citep{collins2001generalization} to diverse data types and distributions.
+Exponential families are a mainstream tool in machine learning.
+They are used  to generalize linear regression \citep{mccullagh1989generalized}, PCA \citep{collins2001generalization} or k-means \citep{banerjee2005clustering} to diverse data types and distributions.
 \citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure. 

### 2021-02-23 main.tex
Change section layouts

@@
 \renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsection*{\bibname}}ni
+\renewcommand{\bibsection}{\subsection*{\bibname}}
 
@@
 We do not know general convergence rates on the KL for maximum likelihood estimates of the exponential family.  We want the simplest one.
-We hope  to get a new result by combining tools from statistics and optimization.
+We hope  to get a new result by combining tools from statistics and optimization. \TODO{Introduce Gaussian variance right from the start.}
 
@@
 where $\meanp$ is called the mean parameter.
+If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
 The second identity entails that $\logpart$ is strictly-convex.
+For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}. \TODO{Add standard illustration}
 
@@
 \end{align}
-with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$.
+with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. We can re-write this prior as 
+\begin{align}
+    \pi(\natp) \propto 
+    \exp( -n_0 \logpart (\natp) 
+    + \langle n_0 \mu_0, \natp \rangle ) \; ,
+\end{align}
+which is the formula for the exponential family with sufficient statistics $(\natp , \logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
 
@@
 \end{align}
-The MAP estimate is a random quantity. We wish to bound its deviation the optimum $\natp^*$ or $\meanp^*$.
+The MAP estimate is a random quantity. We wish to bound its deviation from the optimum $\natp^*$ or $\meanp^*$.
 
-\section{Straightforward Convergence Rate}
+\section{Problem}
 In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
@@
 \end{align}
-\RLP{Add details for 11 12}
+\TODO{Add details for 11 12}
 where $\conj$ is the entropy, the convex conjugate of the log-partition.
-The relationship between Bregman divergences and Fenchel conjugacy is well explained in \citet{wainwright2008graphical}, and \citet{agarwal2010geometric}. 
 The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability -- on the following quantities ?
@@
 	\label{eq:bregmanMAP}
-	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right ) \; .
+	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right ) \; ,
 \end{align}
+where the randomness is coming from the dataset $X_1, \dots, X_n$.
+
+\section{Comparison with  Euclidean Distance}
+Often we can relate this Bregman suboptimality with the $\ell^2$ distance in mean parameter space.
 
@@
 
-\paragraph{$\ell^2$-norm analysis}
+\paragraph{$\ell^2$-norm Analysis}
 Let us make these statements more precise. A bound on the variance becomes a bound on the variance of the average
@@
 
+\section{Asymptotic Behaviour}
 In the limit, the MAP estimates reach a rate $O(1/2n)$. This is shown by approximating the Bregman divergence using a second order Taylor expansion
@@
 Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?
-
-\subsection{Cramer-Rao Lower Bound}
-In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
-In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
-\begin{align}
-    \Cov(\hat \natp_n)  
-    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
-    \iff
-    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
-    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
-\end{align}
-This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance.
-In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"
-
-Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
-For the estimator $\hat \natp_n$, this lower bound is not trivial.
-
-\subsection{Posterior Expected Loss}
-
-A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is 
-\begin{align}
-    \pi(\natp) 
-    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
-    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle)
-\end{align}
-so we see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
-When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
-This is well detailed in \citet{agarwal2010geometric}.
-
-Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
-\begin{align}
-    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-\end{align}
-This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral.
-We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
-This is exactly twice the expected loss of the MAP.
-
-\section{Stochastic Bregman Proximal Point}
-We see that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
-This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
-This can also be seen at each step since
-\begin{align}
-    \hat \natp_{n+1} 
-    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
-    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
-\end{align}
-Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
-
-This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
-
-The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
-First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
-One needs an assumption on the quality of the stochastic estimates $f(.
-; x)$.
-\RLP{Write down the formulas going with this text.}
-Given the proof, the most straightforward such assumption is 
-\begin{align}
-\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
-\end{align}
-which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
-This is exactly the same assumption as given by the SMD analysis.
-However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
-It does not immediately apply to the last iterate.
 
@@
 \end{align}
-where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD. In our case, 
+where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD.
+
+\subsection{Transferring SMD's covariance assumption to MAP estimates}
+In our setting, 
 \begin{align}
@@
 \RLP{should we get a bound that depends on mu instead ? or split between $\mu$ smaller or larger than $\varepsilon$}
-When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(x)$ belong to the marginal polytope, we get
+When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(X)$ belong to the marginal polytope, we get
 \begin{multline}
@@
 \paragraph{Open questions}
-The question remains of : when is such a bound valid?  what does $\sigma^2$ depend on ? To answer these questions and more, we need to turn to examples, see  Section~\ref{sec:gaussian-variance}.
+The question remains of : when is such a bound valid?  what does $\sigma^2$ depend on ? To answer these questions and more, we need to turn to examples, see  Section~\ref{sec:gaussian_variance}.
 
@@
 
+\section{Cramer-Rao Lower Bound}
+In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
+In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
+\begin{align}
+    \Cov(\hat \natp_n)  
+    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
+    \iff
+    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
+    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
+\end{align}
+This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance.
+In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"
+
+Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
+For the estimator $\hat \natp_n$, this lower bound is not trivial.
+
+\section{Posterior Expected Loss}
+
+A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is 
+\begin{align}
+    \pi(\natp) 
+    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
+    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle)
+\end{align}
+so we see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
+When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
+This is well detailed in \citet{agarwal2010geometric}.
+
+Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
+\begin{align}
+    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+\end{align}
+This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral.
+We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
+This is exactly twice the expected loss of the MAP.
+
 \section{Three points lemma}
@@
 \end{proof}
+
+\section{Stochastic Bregman Proximal Point}
+Remark that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
+This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
+This can also be seen at each step since
+\begin{align}
+    \hat \natp_{n+1} 
+    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
+    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
+\end{align}
+Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
+
+This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
+
+The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
+First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
+One needs an assumption on the quality of the stochastic estimates $f(.
+; x)$.
+Given the proof, the most straightforward such assumption (see paper notes) is 
+\begin{align}
+\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
+\end{align}
+which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
+This is exactly the same assumption as given by the SMD analysis.
+However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
+It does not immediately apply to the last iterate.
 

### 2021-02-23 main.tex
Add table of contents

@@
 where the randomness is coming from the dataset $X_1, \dots, X_n$.
+
+\tableofcontents
 
@@
 \end{align}
-In other words, we need for all $\lr\in(0,\inv{n})$, and $ \mu$ in the interior of the marginal polytope,
+In other words, we need for all $\lr\in(0,\inv{n})$, and a certain subset of $\mu$,
 \begin{align}
 \label{eq:hypvariance}
+	\boxed{
 	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr) \mu)\bigg	) 
 	\leq \lr \sigma^2 \; .
+	}
 \end{align}
-We study this assumption in Section \ref{sec:gaussian_variance} where we show that it cannot hold in general for the $\mu$ we care about. 
-\RLP{should we get a bound that depends on mu instead ? or split between $\mu$ smaller or larger than $\varepsilon$}
-When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, assuming all $T(X)$ belong to the marginal polytope, we get
+
+\paragraph{MLE Analysis.}
+When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, we get
 \begin{multline}
@@
 	 	\; \text{with}\; \mu=\inv{n-1}\smallsum_{j\neq i} T(x_j)  } ]  \leq \frac{\sigma^2}{n}
-\end{multline}
-\textcolor{red}{what about:}
-\begin{multline}
-	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) }
-	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
-	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) \textcolor{red}{- \hat \mu_{n-1} + \hat \mu_{n-1}} - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance} } ]  \leq \frac{\sigma^2}{n}
+\label{eq:MLE_bound}
 \end{multline}
@@
 In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality.
+
+This is too simple to be true.
+Indeed we study assumption \eqref{eq:hypvariance} for the special case of the variance of a Gaussian in Section \ref{sec:gaussian_variance}.
+In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
+The mean parameter $\mu$ take values arbitrarily close from $0$, which makes the covariance explode.
+\RLP{should we get a functional bound $\sigma^2(\mu)$ instead ? or split between $\mu$ smaller or larger than $\varepsilon$.}
+
+\paragraph{MAP Analysis.}
 We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
@@
 
-\paragraph{Open questions}
-The question remains of : when is such a bound valid?  what does $\sigma^2$ depend on ? To answer these questions and more, we need to turn to examples, see  Section~\ref{sec:gaussian_variance}.
+\paragraph{Bound on the trajectory.}
+Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
+No we cannot because $\mu$ takes every possible values as a random variable within the expectation.
 
-\TODO{This bound might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?}
+\section{Expected Relative Smoothness}
+To avoid gradient variance assumption, we can take inspiration from expected smoothness, which is a consequence of convexity and smoothness, and which yields a general analysis of SGD.
+Can we get something similar with relativity and SMD?
+The analog of gradient norm is the Bregman between iterates, similar to the Homeomorphic EM rate.
 
-\section{Variance bound for centered gaussians}
+\section{Finding the variance of a Gaussian}
 \label{sec:gaussian_variance}
@@
 
-\paragraph{Big picture.}
+\subsection{Big picture.}
 Intuitively, we have summarize in the part in $\mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} < t)$, the problem that arise with ${\cal B}_{A^*}(\mu^*; \mu)$ when $\mu$ is small. 
@@
 \section{Self-Concordance}
-
+\subsection{Motivation}
 A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 

### 2021-02-23 main.tex
Clean alternative tracks

@@
 
-This article \url{http://davidpfau.com/assets/generalized_bvd_proof.pdf} gives a bias variance decomposition which applies here. However it does not give interesting conclusions.
-
 \paragraph{Relative Continuity}
@@
 
-\section{Alternative Research Track}
-Weakly supervised : missing part of $x_i$ for each $i$.
-
-Structured prediction : graphical model on $X$.
-
 \section{Related Work}
@@
 \citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure.
+
 \citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem.
@@
 Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
+
 \citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
 Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
+
+\citet{pfau2013generalized} gives a bias variance decomposition for Bregman divergences. However it does not give interesting conclusions for our problem of interest.
+
+\subsection{Alternative Research Tracks}
+\begin{itemize}
+	\item Weakly supervised : missing part of $x_i$ for each $i$.
+	\item Structured prediction : graphical model on $X$ or other kind of structure.
+\end{itemize}
 

### 2021-03-02 main.tex
Interleave the gaussian example

@@
 \usepackage{hyperref}       % hyperlinks
+\usepackage[dvipsnames]{xcolor}
 \usepackage{tikz}           % define for loops
@@
 \newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}
+
+\definecolor{light-gray}{gray}{0.9}
+\usepackage[framemethod=TikZ]{mdframed} % frames with colored background
+\newenvironment{example}{
+	\begin{mdframed}[backgroundcolor=light-gray, roundcorner=5pt]
+}{		
+	\end{mdframed}
+}
+\definecolor{myblue}{RGB}{199, 235, 255}
+\newenvironment{important}{
+	\begin{mdframed}[backgroundcolor=myblue, roundcorner=5pt]
+}{		
+	\end{mdframed}
+}
 
@@
 \paragraph{Motivation.} 
-We do not know general convergence rates on the KL for maximum likelihood estimates of the exponential family.  We want the simplest one.
-We hope  to get a new result by combining tools from statistics and optimization. \TODO{Introduce Gaussian variance right from the start.}
-
+We do not know general convergence rates on the KL for maximum likelihood estimates of the exponential family.
+We want the simplest one.
+We hope  to get a new result by combining tools from statistics and optimization. 
 \paragraph{Exponential Family.}
@@
 \end{align}
-with log-partition function 
+where $\logpart$ is the log-partition function (aka normalization factor) 
 \begin{align}
@@
 \end{align}
-Recall that $\logpart$ verifies the two following identities.
+
+\begin{example}
+	These frames contain the trailing example of this paper: a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
+	The density of a centered normal variable is
+\begin{align}
+	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
+\end{align}
+Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
+Mean and natural parameters are roughly inverse of each other
+\begin{align}
+	\natp = -\inv{2 \mu}\; .
+\end{align}
+Now we can match the log-likelihood with the exponential family template to get the log-partition function.
+\begin{align}
+	\log p(x) = \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
+	= x^2 \natp - \logpart(\natp) \\
+	\implies \logpart (\natp) = \half \left( -\log(-\natp)  + \log(\pi) \right)
+\end{align}
+\end{example}
+
+\paragraph{Duality}
+The logpartition function $\logpart$ verifies the two following identities
 \begin{align}
@@
 The second identity entails that $\logpart$ is strictly-convex. 
-For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}. \TODO{Add standard illustration}
+At this point it is useful to introduce the \href{https://en.wikipedia.org/wiki/Convex_conjugate}{convex conjugate} (aka Fenchel-Legendre transform) of the logpartition function
+\begin{align}
+	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
+\end{align}
+It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
+If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
+For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
+\begin{figure}[h]
+	\centering
+	\includegraphics{figs/duality}
+	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$.
+Figure reproduced from \citet{kunstner2020homeomorphic}.}
+	\label{fig:duality}
+\end{figure}
+\begin{example}
+		For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
+\begin{align}
+	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
+\end{align}
+\end{example}
 
@@
 \end{align}
-where $n_0$ and $\natp_0$ are (hyper)parameters of the prior.
-Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
-Finally, $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
+where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
 \begin{align}
@@
 \end{align}
-with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. We can re-write this prior as 
+with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
+Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
+We can re-write this prior as 
 \begin{align}
@@
 \end{align}
-which is the formula for the exponential family with sufficient statistics $(\natp , \logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
+which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
+
+\begin{example}
+In the case of $\cN(0,\sigma^2)$, the logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
+In particular,
+\begin{align}
+	\pi(\natp) 
+	&\propto 
+    \exp( -n_0 \logpart (\natp) 
+    + \langle n_0 \mu_0, \natp \rangle ) \\
+    &\propto \exp( \half[n_0] \log(-\natp) + n_0\mu_0 \natp ) \\
+	&\propto (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
+\end{align}
+from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$.
+\end{example}
 
 \paragraph{MAP.}
-The negative log-likelihood of the prior is then
+The negative log-likelihood of the prior is
 \begin{align*}
@@
 \end{align*}
-Thus the joint NLL of $(x_1,\dots,x_n,\natp)$ is
+Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
 \begin{align}
-    -\log p(X|\natp)\pi(\natp) 
+    -\log p(\cD|\natp)\pi(\natp) 
     = (n_0+n) \logpart (\natp) 
-    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(x_i) \right) + \cst \; .
+    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
 \end{align}
@@
 \begin{align}
-    \hat \natp = \argmin_\natp -\log p(X|\natp) + n_0 \bregman(\natp ; \natp_0)
+    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
 \end{align}
@@
 \end{align}
-The MAP estimate is a random quantity. We wish to bound its deviation from the optimum $\natp^*$ or $\meanp^*$.
+When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
+\begin{align}
+	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
+\end{align}
+The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
 
 \section{Problem}
-In the realizable case, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
+For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
 \begin{align}
-    &\expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*)}\\
-    =& \KL( p(.|\natp^*) ; p(.|\natp)) \\
-    =& \bregman (\natp ; \natp^*) \\
-    =& \bregmanconj ( \meanp^* ; \meanp) \\
-    =& \conj(\mu^*) + \logpart(\natp) - \langle \natp , \mu^* \rangle
-    %& = \logpart(\natp) - \logpart(\natp^*) - \meanp^{*\top}(\natp - \natp^*) \; .
+    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
+	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
 \end{align}
-\TODO{Add details for 11 12}
-where $\conj$ is the entropy, the convex conjugate of the log-partition.
+For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
+\begin{align}
+	\KL( p(.|\natp^*) ; p(.|\natp)) 
+	= \bregman (\natp ; \natp^*)  \; .
+\end{align}
+There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
+\begin{align}
+	\bregman (\natp ; \natp^*)
+    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
+    = \bregmanconj ( \meanp^* ; \meanp)
+\end{align}
+so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
+\begin{align}
+\boxed{
+	\KL( p(.|\natp^*) ; p(.|\natp))
+    = \bregman (\natp ; \natp^*)
+    = \bregmanconj ( \meanp^* ; \meanp) \; .
+}
+\end{align}
+\begin{important}
 The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability -- on the following quantities ?
@@
 where the randomness is coming from the dataset $X_1, \dots, X_n$.
+\end{important}
+\begin{example}
+	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$.
+Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
+\begin{align}
+	\KL( \sigma_n^2 ; \sigma_*^2) 
+	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
+	\bregmanconj( \mu_*; \mu_n) 
+	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
+	\bregman( \natp_n; \natp_* ) 
+	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
+\end{align}
+In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
+\begin{align}
+	\phi(z) := \half (z - 1 - \log(z)) \\
+	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
+\end{align}
+as illustrated in Figure~\ref{fig:phi}.
+\end{example}
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=.4\textwidth]{figs/bregmandef.pdf}
+	\includegraphics[width=.4\textwidth]{figs/phi.pdf}
+	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$, so it is poorly approximated by quadratics.}
+	\label{fig:phi}
+\end{figure}
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=.54\textwidth]{figs/asymptote.pdf}
+	\includegraphics[width=.43\textwidth]{figs/fewsamples.pdf}
+	\caption{Suboptimality of a gaussian variance MLE against number of samples $n$.
+Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, ligth shade is min-max interval.
+	\textbf{Left:} as $n$ increases, the suboptimality matches the $1/2N$ asymptote.
+	\textbf{Right:} the first few samples significantly deviate from this behavior, can we bound this deviation ? 
+	}
+	\label{fig:curves}
+\end{figure}
+
+\begin{example}
+	\paragraph{Exact computation}
+	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \sum_i X_i^2 /n$.
+The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$.
+Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
+	\begin{align}
+		\expect{\frac{\mu_*}{ \mu_n} } =
+		\begin{cases}
+			\frac{n}{n-2} \ \text{ if } n>2, \\
+			+\infty \  \text{ otherwise. }
+		\end{cases}
+	\end{align}
+	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
+	\begin{align}
+		\expect{\log \frac{\mu_n}{\mu_*}} = \log(2) + \psi(\half[n]) - \log(n)
+	\end{align}
+	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
+	Consequently the suboptimality of the MLE has a closed form solution
+	\begin{align}
+	\expect{\bregmanconj( \mu_*; \mu_n) }
+	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log(\frac{\mu_n}{\mu_*} } \\
+	& =
+	\begin{cases}
+		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
+			+\infty \  \text{ otherwise. }
+	\end{cases}
+	\end{align}
+	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
+	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
+	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
+	\begin{align}
+		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
+	\end{align}
+	  For $n\geq 3$,
+	\begin{align}
+		\inv{n-2} - \inv{n}
+		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
+		\leq \inv{n-2} - \inv{2 n} \\
+		\iff
+			\frac{2}{n(n-2)}
+			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
+			\leq \inv{2n} +\frac{2}{n(n-2)}
+			\label{eq:tight_bounds}
+	\end{align}
+	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
+	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
+	\begin{align}
+		\expect{\bregmanconj( \mu_*; \mu_n) }
+		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
+	\end{align}
+	Unfortunately, formulas for the MAP seem to be much less nice.
+There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}.
+Again the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension.
+We do not perform these calculations for now as we want to focus on the simplest possible examples.
+\end{example}
 
@@
 
+\begin{example}
+Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
+The covariance  \eqref{eq:hypvariance} we want to bound is equal to
+\begin{align}
+	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
+	\approx \lr (1-\lr)^{-2} \left (\frac{\sigma_*}{\sigma_0} \right)^4
+\end{align}
+using a first order approximation of the inverse, eg a first order approximation of $\nabla \conj$ that holds when $\lr X^2 \ll (1-\lr) \sigma_0^2$.
+This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.
+This is a problem, because we need this bound to hold  for all values of $\sigma_0$ in the support of $X^2$, meaning on the whole $[0,+\infty[$, which is impossible because $1/X^2$ diverges, so we need to process this integral differently.
+\end{example}
+
 \paragraph{MLE Analysis.}
@@
 The analog of gradient norm is the Bregman between iterates, similar to the Homeomorphic EM rate.
-
-\section{Finding the variance of a Gaussian}
-\label{sec:gaussian_variance}
-Sections on proximal point and mirror descent gave us the variance assumption 
-\begin{align}
-	\Cov(T(x_{n}) , \natp_{n+1}) = \Cov_X \bigg(T(X), \nabla \conj (\lr_n T(X) + (1-\lr_n)\mu_n)\bigg	) 
-	\leq \lr_n M \; .
-\end{align}
-on the sufficient statistic and $\conj$.
-Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
-The density of a centered normal variable is
-\begin{align}
-	X \sim \cN(0,\sigma^2) \implies p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
-\end{align}
-Using the $X^2$ as a sufficient statistic we get the variance as a mean parameter and minus half the precision as a natural parameter
-\begin{align}
-	\quad T(X)= X^2 \ ; \quad  \mu = \sigma^2 \ ; \quad   \natp = -\inv{2 \sigma^2} = -\inv{2 \mu}\; .
-\end{align}
-Now we can match the log-likelihood with the exponential family template to get the log-partition function.
-\begin{align}
-	-\log p(x) = \frac{x^2}{2 \sigma^2} - \half \log(2\inv{2\sigma^2}) + \half \log(2 \pi) 
-	= -x^2 \natp + \logpart(\natp) \\
-	\implies \logpart (\natp) = -\half \log(-\natp)  + \half \log(\pi)
-\end{align}
-We can then use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
-\begin{align}
-	\conj(\mu) = -\half \log(\mu) + \half \log\frac{\pi}{2} - \half \\
-\end{align}
-Remark that both the entropy and the log-partition have the same shape $z\mapsto -\inv{2z}$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
-\begin{align}
-	\KL( \sigma_n^2 ; \sigma_*^2) 
-	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
-	\bregmanconj( \mu_*; \mu_n) 
-	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
-	\bregman( \natp_n; \natp_* ) 
-	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
-\end{align}
-In other words, this is divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi(z)= \half (z - 1 - \log(z))$.
-\begin{align}
-	\phi(z) := \half (z - 1 - \log(z)) 
-	\quad \text{ and } \quad
-	\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  \\
-	\implies \bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
-\end{align}
-The covariance  \eqref{eq:hypvariance} we want to bound is equal to
-\begin{align}
-	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
-	\approx \lr (1-\lr)^{-2} \left (\frac{\sigma_*}{\sigma_0} \right)^4
-\end{align}
-using a first order approximation of the inverse, eg a first order approximation of $\nabla \conj$ that holds when $\lr X^2 \ll (1-\lr) \sigma_0^2$.
-This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.
-This is a problem, because we need this bound to hold  for all values of $\sigma_0$ in the support of $X^2$, meaning on the whole $[0,+\infty[$, which is impossible because $1/X^2$ diverges, so we need to process this integral differently.
-
-\paragraph{Prior} A conjugate prior over the natural parameter $\natp$ is 
-\begin{align}
-	-\log \pi(\natp)
-	&= n_0 \bregman(\natp ; \natp_0) + \cst \\
-	&= n_0 \logpart(\natp) - n_0\mu_0 \natp + \cst \\
-	&= -\half[n_0] \log(-\natp) + n_0\mu_0 (-\natp) + \cst \; . \\
-	\pi (\natp) &= (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
-\end{align}
-In words, $-\natp$ is a random variable from the exponential family with sufficient statistics $(-\natp, \log(-\natp))$ and with natural parameters $(-n_0 \mu_0, \half[n_0])$.
-This is the definition of a Gamma distribution with shape parameter $1 + \half[n_0]$ and with rate parameter $n_0 \mu_0$.
 
@@
 We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
-This is exactly twice the expected loss of the MAP.
+This is exactly twice the expected loss of the MLE.
+
+\begin{example}
+	We can also get a closed form for the posterior of a centered gaussian variance 
+	\begin{align}
+		\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
+	\end{align}
+	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma.
+The using results from the MLE, we get the exact formula
+	\begin{align}
+		\expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
+	\end{align}
+	and we bound it to 
+	\begin{align}
+		\inv{2n} + \frac{4}{n(n-2)}
+		\leq \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		\leq 
+		\inv{n} + \frac{4}{n(n-2)}
+	\end{align}
+	which is about twice the MLE bounds~\eqref{eq:tight_bounds}, exactly like as we observed for the gaussian mean.
+\end{example}
 

### 2021-03-03 main.tex

@@
 which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
+The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.
 
@@
 \end{align}
-from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$.
+from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\natp \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$.
+After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.
 \end{example}
 
-\paragraph{MAP.}
+\paragraph{Maximum A Posteriori (MAP).}
 The negative log-likelihood of the prior is
@@
 \begin{align}
-    \nabla \logpart(\hat \natp) = \hat \meanp
+    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
@@
 \end{important}
+
+\paragraph{Remark.}
+What we are looking for is really akin to concentration inequality, expressed with a Bregman divergence instead of a norm.
+A key difference though, is that the random variable $T(X)$ is connected to the metric $\logpart$.
+Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
+For instance, if we plug in $\conj(\mu)= -\log(\mu)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
+
 \begin{example}
@@
 
+\section{Conjecture}
+
+We know that asymptotically, the convergence rate is dominated by $O(n^{-1})$ variance term. To this we may add a bias term in $O(n^{-2})$. 
+We also know from the $\cN(0,\mu_*)$ example that the MLE can have infinite loss when there is not enough samples.
+This is not the case of the MAP which always has a bounded loss, since it can never reach the boundary of the marginal polytope (the domain of $\mu$).
+Thus we may have a bias term which is infinite for $n_0=0$ (eg no prior, MLE), and that decreases as $n_0$ increases, except that the convergence speed gets worse and worse for larger $n_0$, eg
+\begin{align}
+    \frac{d}{2n} +  \log \left (\inv{n_0} \right) \left(1+ \frac{n}{n_0} \right)^{-2} \bregman(\natp_*; \natp_0)
+\end{align}
+We also know that this expression should dominate the asymptote.
+Only the variance is responsible for the infinite behavior, except it gets controlled when $n_0>0$.
+
+I really want to get a bound for the MAP of a Gaussian variance, to get inspiration from its shape.
+
 \section{Stochastic Mirror Descent (SMD)}
@@
 
-This is too simple to be true.
-Indeed we study assumption \eqref{eq:hypvariance} for the special case of the variance of a Gaussian in Section \ref{sec:gaussian_variance}.
+This is too simple to be true, as shown on our trailing example.
 In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
@@
 
+% todo: idee: adapter la preuve de stochastic gradient descent a stochastic mirror descent.
+Exploiter un parralele entre ces deux preuves pour obtenir une preuve pour stochastic mirror descent dans le cas ou f est relatively smooth w.r.t.
+A*.
+ce qu'on veut, c'est une hypothese analogue a expected smoothness (see paper of nico loizou) pour le relatively smooth.
+% SGD: General Analysis and Improved Rates gower et al -- section 2 and 3 see equation (9) in particular (expected smoothnesS)
+% Relatively-Smooth Convex Optimization by First-Order Methods, and Applications,  -- Section 3
+
+\section{Gaussian Variance}
 \subsection{A big calculus in this specific setting}

### 2021-03-09 main.tex
No idea

@@
 	\includegraphics[width=.43\textwidth]{figs/fewsamples.pdf}
-	\caption{Suboptimality of a gaussian variance MLE against number of samples $n$.
-Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, ligth shade is min-max interval.
+	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$.
+Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval.
 	\textbf{Left:} as $n$ increases, the suboptimality matches the $1/2N$ asymptote.
-	\textbf{Right:} the first few samples significantly deviate from this behavior, can we bound this deviation ? 
+	\textbf{Right:} the first few samples significantly deviate from this behavior.
+In fact, for $n=1$ and $n=2$, the expected value is infinite, but we have a closed form solution and a simple upper-bound for $n>2$.
 	}
@@
 	\expect{\bregmanconj( \mu_*; \mu_n) }
-	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log(\frac{\mu_n}{\mu_*} } \\
+	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
 	& =

### 2021-03-18 main.tex
Lots of modifications - add bias variance section

@@
 
+%% Faster processing
+\renewcommand\appendix{\end{document}}
+
 \begin{document}
@@
 \begin{align}
-	\log p(x) = \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
+	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
 	= x^2 \natp - \logpart(\natp) \\
-	\implies \logpart (\natp) = \half \left( -\log(-\natp)  + \log(\pi) \right)
+	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
 \end{align}
@@
 \end{align}
+We can also take gradient and derivative of $\logpart(\natp)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
+\begin{align}
+	\nabla\logpart(\natp) &= \frac{-1}{2 \natp} = \sigma^2 = \mu = \E[X^2] \\
+	\nabla^2\logpart(\natp) &= \frac{1}{2 \natp^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
+\end{align}
+which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.
 \end{example}
@@
 \begin{important}
-The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds -- in expectation or high-probability -- on the following quantities ?
+The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
 \begin{align}
 	\label{eq:bregmanMLE}
-	\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) \; , \\
+	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
 	\label{eq:bregmanMAP}
-	\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right ) \; ,
-\end{align}
-where the randomness is coming from the dataset $X_1, \dots, X_n$.
+	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
+\end{align}
+where the outer expectation is on the dataset $X_1, \dots, X_n$?
 \end{important}
@@
 	\begin{align}
-		\expect{\frac{\mu_*}{ \mu_n} } =
+		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
+		= \expect{\frac{n}{\chi^2(n)} - 1} =
 		\begin{cases}
-			\frac{n}{n-2} \ \text{ if } n>2, \\
+			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
 			+\infty \  \text{ otherwise. }
@@
 	\begin{align}
-		\expect{\log \frac{\mu_n}{\mu_*}} = \log(2) + \psi(\half[n]) - \log(n)
+		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
 	\end{align}
@@
 
+\section{Bias-Variance Decomposition}
+Using Jensen inequality, we have the following analog of the bias-variance decomposition:
+
+\newcommand{\bias}[1]{\textcolor{Red}{\underbrace{\textcolor{black}{#1}}_{\text{bias}}}}
+\newcommand{\variance}[1]{\textcolor{Green}{\underbrace{\textcolor{black}{#1}}_{\text{variance}}}}
+
+\begin{align}
+	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} \\
+	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
+	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle 
+	+ \frac{1}{n+n_0} \smallsum_i \variance{\expect{\langle \hat \natp_n ; T(X_i) - \mu^* \rangle } }
+\end{align}
+where $\hat \mu_n$ is the MAP estimate,
+$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$,
+and $\theta_n = \nabla \conj(\mu_n)$.
+
 \section{Conjecture}
@@
 This is not the case of the MAP which always has a bounded loss, since it can never reach the boundary of the marginal polytope (the domain of $\mu$).
+I really want to get a bound for the MAP of a Gaussian variance, to get inspiration from its shape. 
 Thus we may have a bias term which is infinite for $n_0=0$ (eg no prior, MLE), and that decreases as $n_0$ increases, except that the convergence speed gets worse and worse for larger $n_0$, eg
@@
 \end{align}
-We also know that this expression should dominate the asymptote.
-Only the variance is responsible for the infinite behavior, except it gets controlled when $n_0>0$.
-
-I really want to get a bound for the MAP of a Gaussian variance, to get inspiration from its shape.
+Besides this bias term should be visible from a simple mirror descent, with no noise or variance. Yet it is not trivial to get a $O(n^{-2})$ rate in this setting. See note by Frederik.
 
@@
 
-\section{Gaussian Variance}
+\subsection{Convergence of Mirror Descent}
+To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$.
+Note that the strong convexity constant $m$ may be zero.
+Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
+\begin{align}
+	f(\natp^{t+1}) 
+	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
+	& \text{(Relative $L$-smoothness)} \\
+	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
+	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \text{(Relative $m$-strong convexity)}
+\end{align}
+Applying the 3 points lemma with \TODO{f}
+
+\section{Self-Concordance}
+\subsection{Motivation}
+A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
+\begin{align}
+    \logpart(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
+\end{align}
+which is defined on $\eta \in \real^d$ and $\Lambda\in \real^{d\times d}$ symmetric positive definite. 
+It is not strongly convex, so $\conj$ is not smooth.
+
+Another hypothesis that may be more suitable is self-concordance. $f: \real \rightarrow \real$ is SC if 
+\begin{align}
+    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
+\end{align}
+The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
+A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
+Negative logarithm $-\log(x)$ and entropy $x\log(x)$ are both self-concordant function. This is good news for us since log-partition function may include SC logarithmic barriers.
+In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
+
+\subsection{Suboptimality and Newton Decrement}
+An important property of self-concordant functions (cite Boyd's book, although Nesterov's may be better) is that their suboptimality  may be upper bounded by the Newton Decrement 
+\begin{align}
+    \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
+\end{align}
+In general, subtracting the minimum $f^*$ of $f$ , we have
+\begin{align}
+    f(x) - f* \leq - \decrement(x) - \log( 1- \decrement(x)) \; .
+\end{align}
+Note that this bound is vacuous for $\decrement(x)\geq 1$. For $y = \decrement(x) \leq 0.68$, we have $ - y - \log(1-y) \leq y^2$, so we get the bound
+\begin{align}
+    f(x) - f^* \leq \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
+\end{align}
+
+Our functions of interest is $f(\natp)= \bregman (\natp ; \natp^*)= \bregmanconj ( \meanp^* ; \meanp) = g(\meanp)$, with minimum $f^*=0$. 
+If $\logpart$ is self-concordant, then so is $f$, but not necessarily $g$.
+The gradient and Hessian of $f$ are
+\begin{align}
+    f(\natp) 
+    &= \logpart(\natp) - \logpart(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
+    \nabla f(\natp) 
+    &= \meanp - \meanp^* = \expect[p(X|\natp)]{T(X)} - \expect[p(X|\natp^*)]{T(X)} \\
+    \nabla^2 f(\natp) 
+    &= \mSigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
+\end{align}
+so that we get the bound.
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    \leq \decrement(\natp)^2 
+    = \norm{\mu^* - \mu(\theta)}^2_{\mSigma(\theta)^{-1}} 
+    \leq 0.46
+\end{align}
+% If we were to use $g(\meanp)$ instead, we would get dirty third order derivatives
+% \begin{align}
+%     g(\meanp) 
+%     &= \conj(\meanp^*) - \conj(\meanp) - \langle \natp, \meanp^* - \meanp \rangle \\
+%     \nabla g(\meanp) 
+%     &= -\natp - \nabla^2 \conj(\meanp) (\meanp^* - \meanp) + \natp \\
+%     &= \nabla^2 \conj(\meanp) (\meanp - \meanp^*) \\
+%     \nabla^2 g(\meanp) 
+%     &= \nabla^2 \conj(\meanp) + \nabla^3 \conj(\meanp ) (\meanp - \meanp^*)
+% \end{align}
+% where the last expression involves a third order tensor.
+
+Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$, then we would get
+\begin{align}
+    \nabla h(\meanp ) 
+    &= \natp - \natp^* \\
+    \nabla^2 h(\meanp ) 
+    &= \nabla^2 \conj(\meanp) 
+    = \nabla^2 \logpart(\natp) ^{-1}
+    = \Cov_{p(X|\natp)}[T(X)]^{-1}\\
+    \implies 
+    \decrement(\meanp)
+    &= \Var_{p(X|\natp)}[(\natp - \natp^*)^T T(X)] \; .
+\end{align}
+This is just a remark. I don't think it can help us to get anywhere.
+
+\section{Related Work}
+Exponential families are a mainstream tool in machine learning.
+They are used  to generalize linear regression \citep{mccullagh1989generalized}, PCA \citep{collins2001generalization} or k-means \citep{banerjee2005clustering} to diverse data types and distributions.
+\citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure.
+
+\citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem.
+They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
+Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors yet critical as it does not hold for multivariate normals with unknown covariance.
+\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.
+Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
+
+\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
+Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
+
+\citet{pfau2013generalized} gives a bias variance decomposition for Bregman divergences. However it does not give interesting conclusions for our problem of interest.
+
+\subsection{Alternative Research Tracks}
+\begin{itemize}
+	\item Weakly supervised : missing part of $x_i$ for each $i$.
+	\item Structured prediction : graphical model on $X$ or other kind of structure.
+\end{itemize}
+
+\bibliographystyle{apalike}
+\bibliography{references.bib}
+
+\newpage
+\appendix
+
+\section{Cramer-Rao Lower Bound}
+In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
+In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
+\begin{align}
+    \Cov(\hat \natp_n)  
+    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
+    \iff
+    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
+    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
+\end{align}
+This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance.
+In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"
+
+Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
+For the estimator $\hat \natp_n$, this lower bound is not trivial.
+
+\section{Posterior Expected Loss}
+
+A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is 
+\begin{align}
+    \pi(\natp) 
+    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
+    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle)
+\end{align}
+so we see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
+When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
+This is well detailed in \citet{agarwal2010geometric}.
+
+Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
+\begin{align}
+    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+\end{align}
+This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral.
+We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
+This is exactly twice the expected loss of the MLE.
+
+\begin{example}
+	We can also get a closed form for the posterior of a centered gaussian variance 
+	\begin{align}
+		\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
+	\end{align}
+	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma.
+The using results from the MLE, we get the exact formula
+	\begin{align}
+		\expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
+	\end{align}
+	and we bound it to 
+	\begin{align}
+		\inv{2n} + \frac{4}{n(n-2)}
+		\leq \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		\leq 
+		\inv{n} + \frac{4}{n(n-2)}
+	\end{align}
+	which is about twice the MLE bounds~\eqref{eq:tight_bounds}, exactly like as we observed for the gaussian mean.
+\end{example}
+
+\section{Three points lemma}
+\label{app:3points}
+
+Proofs of convergence for mirror descent with relative smoothness rely on a specific property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
+\begin{align}
+	\min	_{x\in C} \overbrace{f(x) + \cB_h(x ; y)}^{\phi(x)}
+\end{align}
+where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector.
+Then
+\begin{align}
+	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
+\end{align}
+This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).
+
+\begin{proof}
+	The first order optimality condition is
+	\begin{align}
+		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
+		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
+	\end{align}
+	This proof relies on another property called three point property 
+	\begin{align}
+		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
+	\end{align}
+	which can be proved by expanding the right hand side.
+	By convexity of $f$ we also have 
+	\begin{align}
+		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
+	\end{align} 
+	Putting it all together we get
+	\begin{align}
+		0 
+		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
+		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
+	\end{align}
+	which concludes the proof.
+\end{proof}
+
+\section{Stochastic Bregman Proximal Point}
+Remark that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
+This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
+This can also be seen at each step since
+\begin{align}
+    \hat \natp_{n+1} 
+    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
+    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
+\end{align}
+Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
+
+This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
+
+The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
+First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
+One needs an assumption on the quality of the stochastic estimates $f(.
+; x)$.
+Given the proof, the most straightforward such assumption (see paper notes) is 
+\begin{align}
+\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
+\end{align}
+which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
+This is exactly the same assumption as given by the SMD analysis.
+However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
+It does not immediately apply to the last iterate.
+
+\section{Vivien's approach}
 \subsection{A big calculus in this specific setting}
@@
 we have that $B_{A^*}(\mu^*; \mu) < cn^{-1}\log(1/\delta) / \norm{\mu^*}$.
-
-\section{Self-Concordance}
-\subsection{Motivation}
-A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
-\begin{align}
-    \logpart(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
-\end{align}
-which is defined on $\eta \in \real^d$ and $\Lambda\in \real^{d\times d}$ symmetric positive definite. 
-It is not strongly convex, so $\conj$ is not smooth.
-
-Another hypothesis that may be more suitable is self-concordance. $f: \real \rightarrow \real$ is SC if 
-\begin{align}
-    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
-\end{align}
-The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
-A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
-Negative logarithm $-\log(x)$ and entropy $x\log(x)$ are both self-concordant function. This is good news for us since log-partition function may include SC logarithmic barriers.
-In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
-
-\subsection{Suboptimality and Newton Decrement}
-An important property of self-concordant functions (cite Boyd's book, although Nesterov's may be better) is that their suboptimality  may be upper bounded by the Newton Decrement 
-\begin{align}
-    \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
-\end{align}
-In general, subtracting the minimum $f^*$ of $f$ , we have
-\begin{align}
-    f(x) - f* \leq - \decrement(x) - \log( 1- \decrement(x)) \; .
-\end{align}
-Note that this bound is vacuous for $\decrement(x)\geq 1$. For $y = \decrement(x) \leq 0.68$, we have $ - y - \log(1-y) \leq y^2$, so we get the bound
-\begin{align}
-    f(x) - f^* \leq \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
-\end{align}
-
-Our functions of interest is $f(\natp)= \bregman (\natp ; \natp^*)= \bregmanconj ( \meanp^* ; \meanp) = g(\meanp)$, with minimum $f^*=0$. 
-If $\logpart$ is self-concordant, then so is $f$, but not necessarily $g$.
-The gradient and Hessian of $f$ are
-\begin{align}
-    f(\natp) 
-    &= \logpart(\natp) - \logpart(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
-    \nabla f(\natp) 
-    &= \meanp - \meanp^* = \expect[p(X|\natp)]{T(X)} - \expect[p(X|\natp^*)]{T(X)} \\
-    \nabla^2 f(\natp) 
-    &= \mSigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
-\end{align}
-so that we get the bound.
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq \decrement(\natp)^2 
-    = \norm{\mu^* - \mu(\theta)}^2_{\mSigma(\theta)^{-1}} 
-    \leq 0.46
-\end{align}
-% If we were to use $g(\meanp)$ instead, we would get dirty third order derivatives
-% \begin{align}
-%     g(\meanp) 
-%     &= \conj(\meanp^*) - \conj(\meanp) - \langle \natp, \meanp^* - \meanp \rangle \\
-%     \nabla g(\meanp) 
-%     &= -\natp - \nabla^2 \conj(\meanp) (\meanp^* - \meanp) + \natp \\
-%     &= \nabla^2 \conj(\meanp) (\meanp - \meanp^*) \\
-%     \nabla^2 g(\meanp) 
-%     &= \nabla^2 \conj(\meanp) + \nabla^3 \conj(\meanp ) (\meanp - \meanp^*)
-% \end{align}
-% where the last expression involves a third order tensor.
-
-Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$, then we would get
-\begin{align}
-    \nabla h(\meanp ) 
-    &= \natp - \natp^* \\
-    \nabla^2 h(\meanp ) 
-    &= \nabla^2 \conj(\meanp) 
-    = \nabla^2 \logpart(\natp) ^{-1}
-    = \Cov_{p(X|\natp)}[T(X)]^{-1}\\
-    \implies 
-    \decrement(\meanp)
-    &= \Var_{p(X|\natp)}[(\natp - \natp^*)^T T(X)] \; .
-\end{align}
-This is just a remark. I don't think it can help us to get anywhere.
-
-\section{Related Work}
-Exponential families are a mainstream tool in machine learning.
-They are used  to generalize linear regression \citep{mccullagh1989generalized}, PCA \citep{collins2001generalization} or k-means \citep{banerjee2005clustering} to diverse data types and distributions.
-\citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure.
-
-\citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem.
-They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
-Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors yet critical as it does not hold for multivariate normals with unknown covariance.
-\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.
-Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
-
-\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
-Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
-
-\citet{pfau2013generalized} gives a bias variance decomposition for Bregman divergences. However it does not give interesting conclusions for our problem of interest.
-
-\subsection{Alternative Research Tracks}
-\begin{itemize}
-	\item Weakly supervised : missing part of $x_i$ for each $i$.
-	\item Structured prediction : graphical model on $X$ or other kind of structure.
-\end{itemize}
-
-\bibliographystyle{apalike}
-\bibliography{references.bib}
-
-\newpage
-\appendix
-
-\section{Cramer-Rao Lower Bound}
-In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
-In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
-\begin{align}
-    \Cov(\hat \natp_n)  
-    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
-    \iff
-    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
-    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
-\end{align}
-This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance.
-In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"
-
-Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
-For the estimator $\hat \natp_n$, this lower bound is not trivial.
-
-\section{Posterior Expected Loss}
-
-A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is 
-\begin{align}
-    \pi(\natp) 
-    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
-    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle)
-\end{align}
-so we see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
-When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
-This is well detailed in \citet{agarwal2010geometric}.
-
-Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
-\begin{align}
-    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-\end{align}
-This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral.
-We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
-This is exactly twice the expected loss of the MLE.
-
-\begin{example}
-	We can also get a closed form for the posterior of a centered gaussian variance 
-	\begin{align}
-		\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
-	\end{align}
-	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma.
-The using results from the MLE, we get the exact formula
-	\begin{align}
-		\expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
-	\end{align}
-	and we bound it to 
-	\begin{align}
-		\inv{2n} + \frac{4}{n(n-2)}
-		\leq \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-		\leq 
-		\inv{n} + \frac{4}{n(n-2)}
-	\end{align}
-	which is about twice the MLE bounds~\eqref{eq:tight_bounds}, exactly like as we observed for the gaussian mean.
-\end{example}
-
-\section{Three points lemma}
-\label{app:3points}
-
-Proofs of convergence for mirror descent with relative smoothness rely on a specific property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
-\begin{align}
-	\min	_{x\in C} \overbrace{f(x) + \cB_h(x ; y)}^{\phi(x)}
-\end{align}
-where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector.
-Then
-\begin{align}
-	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
-\end{align}
-This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).
-
-\begin{proof}
-	The first order optimality condition is
-	\begin{align}
-		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
-		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
-	\end{align}
-	This proof relies on another property called three point property 
-	\begin{align}
-		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
-	\end{align}
-	which can be proved by expanding the right hand side.
-	By convexity of $f$ we also have 
-	\begin{align}
-		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
-	\end{align} 
-	Putting it all together we get
-	\begin{align}
-		0 
-		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
-		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
-	\end{align}
-	which concludes the proof.
-\end{proof}
-
-\section{Stochastic Bregman Proximal Point}
-Remark that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
-This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
-This can also be seen at each step since
-\begin{align}
-    \hat \natp_{n+1} 
-    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
-    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
-\end{align}
-Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
-
-This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
-
-The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
-First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
-One needs an assumption on the quality of the stochastic estimates $f(.
-; x)$.
-Given the proof, the most straightforward such assumption (see paper notes) is 
-\begin{align}
-\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
-\end{align}
-which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
-This is exactly the same assumption as given by the SMD analysis.
-However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
-It does not immediately apply to the last iterate.
 

### 2021-03-23 main.tex
Add bias variance

@@
 \begin{align}
-	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} \\
+	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	= \conj(\mu^* ) 
+	- \underbrace{\expect{\conj(\hat \mu_n)}}_{\leq \conj(\mu_n) \text{ (Jensen)} } 
+	\pm \langle \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
+	- \E[\langle \hat \natp_n ; \mu^* - \hat \mu_n  \rangle ] \\
+	& \leq \conj(\mu^* ) 
+	- \conj(\mu_n) 
+	- \langle \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
+	+ \frac{n_0}{n+n_0} \langle \natp_n ; \mu^* - \mu_0 \rangle \nonumber\\
+	& \quad \quad \quad 
+	- \frac{n_0}{n+n_0} \langle \E[\hat \natp_n] ; \mu^*- \mu_0  \rangle
+	- \inv{n+n_0} \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
 	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
@@
 \end{align}
-where $\hat \mu_n$ is the MAP estimate,
-$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$,
-and $\theta_n = \nabla \conj(\mu_n)$.
+where $\hat \mu_n$ is the mean MAP estimate, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ is the natural MAP estimate, 
+$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ is the expected MAP estimate,
+and $\theta_n = \nabla \conj(\mu_n)$ is the natural parameter of this expectation.
+
+Note that unless $\conj$ is $\ell^2$, or degenerate 0 variance cases, $\theta_n \neq \expect{\hat \theta_n}$.
+In fact, $\theta_n - \expect{\hat \theta_n} = \nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]$
 

### 2021-03-29 main.tex
Bias Variance discussion

@@
 	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
-	&= O\left(\frac{d}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
+	&= \frac{d}{n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
+\end{align}
+or 
+\begin{align}
+	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
 \end{align}
@@
 \section{Bias-Variance Decomposition}
-Using Jensen inequality, we have the following analog of the bias-variance decomposition:
+Using Jensen inequality and inserting the right quantities, we obtain the following analog of the bias-variance decomposition:
 
@@
 	- \frac{n_0}{n+n_0} \langle \E[\hat \natp_n] ; \mu^*- \mu_0  \rangle
-	- \inv{n+n_0} \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
+	- \inv{n+n_0} \smallsum_i \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
 	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
 	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle 
-	+ \frac{1}{n+n_0} \smallsum_i \variance{\expect{\langle \hat \natp_n ; T(X_i) - \mu^* \rangle } }
+	+ \frac{n}{n+n_0} \variance{\expect{\langle \hat \natp_n  ; T(X_0) - \mu^* \rangle } }
 \end{align}
@@
 
+The interpretation of the bias term should be clear.
+The variance term is called as such because it is the covariance between $\hat \natp_n$ and $T(X)$
+\begin{align}
+	\Cov \bigg(T(X_0), \nabla \conj \big(\frac{\smallsum_i T(X_i) +n_0 \mu_0}{n+n_0} \big)\bigg	) \; .
+\end{align}
+The mixed bias variance term is might be negative. 
+\TODO{add figure.}
 Note that unless $\conj$ is $\ell^2$, or degenerate 0 variance cases, $\theta_n \neq \expect{\hat \theta_n}$.
-In fact, $\theta_n - \expect{\hat \theta_n} = \nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]$
+In fact, $\theta_n - \expect{\hat \theta_n} = \nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]$ is sort of a commuting bracket.
+
+\subsection{Asymptotes}
+Using the quadratic approximation of the Bregman divergence near $\mu^*$, we get a $O(n^{-2})$ rate for the pure bias term
+\begin{align}
+	\bregmanconj(\mu^*; \mu_n) 
+	= \frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}}}{2(1 + \frac{n}{n_0})^{-2} } + O(n^{-3}) \; .
+\end{align}
+Using a linear approximation of the mirror map, we also get a $O(n^{-1})$ rate for the variance term
+\begin{align}
+	&\nabla \conj(\mu)  
+	= \nabla \conj(\mu^*) + \nabla^2\conj(\mu^*)(\mu - \mu^*) + O(\norm{\mu - \mu^*}^2) \\
+	\implies &\hat \natp_n 
+	= \natp^* + \Sigma^{-1} (\hat \mu_n - \mu^*)  O(\norm{\hat \mu_n - \mu^*}^2) \\
+	\implies &\Cov(\hat \natp_n ; T(X) ) 
+	= \inv{n+n_0} \E[ \norm{T(X) - \mu^*}^2_{\Sigma^{-1}}]   + O(n^{-2}) \\
+	\implies &\Cov(\hat \natp_n ; T(X) ) 
+	 = \frac{d}{n+n_0} + O(n^{-2}) \; .
+\end{align}
+where we lost the factor $2$ compared to the straightforward asymptotic analysis, probably because of the Jensen upper inequality.
+
+Finally, we can use this same linear approximation to bound the mixed bias-variance term
 

### 2021-04-03 main.tex
Asymptotic discussion of bias-variance decomposition

@@
 \begin{align}
-	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
+\label{eq:MAP_asymptote}
+	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
+	= \frac{d}{2n} + O(n^{- \frac{3}{2}})
 \end{align}
@@
 \section{Bias-Variance Decomposition}
-Using Jensen inequality and inserting the right quantities, we obtain the following analog of the bias-variance decomposition:
-
+Using Jensen inequality and inserting appropriate quantities, we obtain the following analog of the bias-variance decomposition:
 \newcommand{\bias}[1]{\textcolor{Red}{\underbrace{\textcolor{black}{#1}}_{\text{bias}}}}
 \newcommand{\variance}[1]{\textcolor{Green}{\underbrace{\textcolor{black}{#1}}_{\text{variance}}}}
-
 \begin{align}
@@
 \end{align}
-where $\hat \mu_n$ is the mean MAP estimate, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ is the natural MAP estimate, 
+where $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ is the MAP estimate of the mean parameter, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ is the corresponding natural parameter, 
 $\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ is the expected MAP estimate,
 and $\theta_n = \nabla \conj(\mu_n)$ is the natural parameter of this expectation.
-
-The interpretation of the bias term should be clear.
-The variance term is called as such because it is the covariance between $\hat \natp_n$ and $T(X)$
-\begin{align}
-	\Cov \bigg(T(X_0), \nabla \conj \big(\frac{\smallsum_i T(X_i) +n_0 \mu_0}{n+n_0} \big)\bigg	) \; .
-\end{align}
-The mixed bias variance term is might be negative. 
+Note that unless $\conj$ is $\ell^2$, or we are in a degenerate case with $\hat \mu_n$ a Dirac, $\theta_n \neq \expect{\hat \theta_n}$.
+At the last line, we used the property that all data points play a symmetric role to remove the sum over $i$.
+
 \TODO{add figure.}
-Note that unless $\conj$ is $\ell^2$, or degenerate 0 variance cases, $\theta_n \neq \expect{\hat \theta_n}$.
-In fact, $\theta_n - \expect{\hat \theta_n} = \nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]$ is sort of a commuting bracket.
-
-\subsection{Asymptotes}
+
+\subsection{Pure Bias Term}
+The pure bias term measures the divergence between the optimum and the expected MAP estimate $\bregmanconj(\mu^*; \mu_n)$.
+
+\paragraph{Aymptote.}
 Using the quadratic approximation of the Bregman divergence near $\mu^*$, we get a $O(n^{-2})$ rate for the pure bias term
@@
 \end{align}
-Using a linear approximation of the mirror map, we also get a $O(n^{-1})$ rate for the variance term
+
+\subsection{Pure Variance Term}
+The pure variance term $\expect{\langle \hat \natp_n  ; T(X_0) - \mu^* \rangle }$ is the covariance between $\hat \natp_n$ and $T(X_0)$. 
+In a way it is a modified variance of $T(X)$, with the $\nabla \conj$ operator in between, and a $\inv{n+n_0}$ multiplicative factor within $\hat \natp_n = \nabla \conj( \frac{T(X_0)}{n+n_0} + \dots) $.
+
+\paragraph{Positivity.}
+It is positive thanks to the monotonicity of $\nabla \conj$
+\begin{align}
+&\langle \hat \natp_n - \natp_n ; \hat \mu_n - \mu_n \rangle 
+	= \langle \nabla \conj (\hat \mu_n) - \nabla \conj(\mu_n) ;  \hat \mu_n - \mu_n \rangle 
+	\geq 0
+\end{align}
+Taking the expectation, removing the constant $\theta_n$ from the covariance, and expanding $\hat \mu_n - \mu_n$ we get
+\begin{align}
+	\implies 
+	0 \leq \expect{\langle \hat \natp_n - \natp_n ; \hat \mu_n - \mu_n \rangle}
+	&= \expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle} \\
+	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
+	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
+\end{align}
+which concludes the proof of positivity.
+
+\paragraph{Asymptote.}
+Using a linear approximation of the mirror map, we get a $O(n^{-1})$ rate for the variance term
 \begin{align}
@@
 	\implies &\hat \natp_n 
-	= \natp^* + \Sigma^{-1} (\hat \mu_n - \mu^*)  O(\norm{\hat \mu_n - \mu^*}^2) \\
+	= \natp^* + \Sigma^{-1} (\hat \mu_n - \mu^*)  + O(\E[\norm{\hat \mu_n - \mu^*}^2]) \\
 	\implies &\Cov(\hat \natp_n ; T(X) ) 
@@
 \end{align}
-where we lost the factor $2$ compared to the straightforward asymptotic analysis, probably because of the Jensen upper inequality.
-
-Finally, we can use this same linear approximation to bound the mixed bias-variance term
+Note that we lost a factor $\half$ compared to the straightforward asymptotic analysis~\eqref{eq:MAP_asymptote}, probably because of the Jensen upper bound.
+
+\subsection{Mixed Bias-Variance Term}
+The mixed bias variance term is $ \frac{n_0}{n+n_0} \langle  {\mu_* - \mu_0} ; {\natp_n - \E[\hat \natp_n]} \rangle $.
+Note that it might be negative !
+Let's focus on the right factor
+\begin{align}
+\theta_n - \expect{\hat \theta_n} = \nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]
+\end{align}
+is sort of a commuting bracket between $\nabla \conj$ and the expectation over $\hat \mu_n$. It will be large if $\nabla \conj$ is highly non linear over the distribution of $\hat \mu_n$.
+
+\paragraph{Asymptote.}
+A linear approximation of $\nabla \conj$ shows that the mixed term is in $O(n^{-3})$, since the linear function commutes with the expectation.
+We need to use a quadratic approximation of $\nabla \conj$ to get a non-trivial approximation
+\begin{align}
+	&\nabla \conj(\mu)  
+	= \nabla \conj(\mu^*) + \nabla^2\conj(\mu^*)(\mu - \mu^*) +  \half \nabla^3 \conj(\mu^*) [ (\mu - \mu^*)  ; (\mu - \mu^*) ]
+	+ O(\norm{\mu - \mu^*}^3) \\
+	&\implies 
+	\nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]
+	= \half \nabla^3 \conj (\mu^*) \Cov(\hat \mu_n) 
+	+ O(\E[\norm{\hat \mu_n - \mu^*}^3]) \\
+	&\implies 
+	\nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]
+	= \frac{\nabla^3 \conj (\mu^*) \Sigma}{2(n+n_0)^2} 
+	+ O(n^{-3}) \\
+\end{align}
+where the axis of the tensor product between third derivative and vector / matrix are all kept implicit for simplicity.
+Consequently the bias-variance term is 
+\begin{align}
+	\frac{n_0}{n+n_0} \langle  \bias{\mu^* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle
+	 = \frac{n_0}{(n+n_0)^3} (\mu^* - \mu_0)^\top \nabla^3 \conj (\mu^*) \Sigma + O(n^{-4}) = O(n^{-3}) \; .
+\end{align}
 

### 2021-04-06 main.tex
Illustrate the bias-variance decomposition

@@
 For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
-\begin{figure}[h]
+\begin{figure}[ht]
 	\centering
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 	\centering
@@
 
-\TODO{add figure.}
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=.8\textwidth]{figs/bias-variance.png}
+	\caption{A schematic illustration of the different characters featured in the bias-variance decomposition.}
+	\label{fig:bias-variance}
+\end{figure}
 
@@
 Note that we lost a factor $\half$ compared to the straightforward asymptotic analysis~\eqref{eq:MAP_asymptote}, probably because of the Jensen upper bound.
+
+\begin{example}
+	For the gaussian variance MAP, the co-variance term is
+	\begin{align}
+		\frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X) )  
+		&= -\frac{n}{2} \expect{ \frac{X_0^2 - \mu^*}{\sum_i T(X_i)  + n_0 \mu_0} } \\
+		&=  \half \expect{ \frac{n - \chi^2(n)}{\chi^2(n) + n_0 \frac{\mu_0}{\mu^*}}} \\
+		& = \half[n] \expect{\inv{\chi^2(n) + a}} - \half \expect{\inv{1 + \frac{a}{\chi^2(n)}}}
+	\end{align}
+	where $a=n_0 \frac{\mu_0}{\mu^*}$, and at the second line we re-introduced the $X_i$ to form a chi-square. 
+	These expectations are not simple to solve, but using wolfram alpha, we get  some formulas :
+	\begin{align}
+		\expect{\inv{\chi^2(n) + a}}  = \half e^{\frac{a}{2}} \text{ExpIntegralE}[\half[n],\half[a]]
+	\end{align}
+	where the \href{https://reference.wolfram.com/language/ref/ExpIntegralE.html}{exponential integral function} is defined as 
+	\begin{align}
+		\text{ExpIntegralE}[n , z] = E_n(z) = \int_1^\infty \frac{e^{-z t} }{t^n} dt = z^{n-1} \Gamma(1-n , z)
+	\end{align}
+	with the upper incomplete gamma function.
+	Now our goal is to find a simple upper bound on this function.
+\end{example}
 

### 2021-04-07 main.tex
Frame covariance as a nonlinear variance

@@
 	- \underbrace{\expect{\conj(\hat \mu_n)}}_{\leq \conj(\mu_n) \text{ (Jensen)} } 
-	\pm \langle \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
-	- \E[\langle \hat \natp_n ; \mu^* - \hat \mu_n  \rangle ] \\
+	\pm \langle \underbrace{\nabla \conj( \mu_n)}_{=:\natp_n} ; \mu^* - \mu_n \rangle
+	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \\
 	& \leq \conj(\mu^* ) 
@@
 	- \langle  \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
-	+ \frac{n_0}{n+n_0} \langle \natp_n ; \mu^* - \mu_0 \rangle \nonumber\\
-	& \quad \quad \quad 
-	- \frac{n_0}{n+n_0} \langle \E[\hat \natp_n] ; \mu^*- \mu_0  \rangle
-	- \inv{n+n_0} \smallsum_i \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
+	+ \langle \E[\hat \natp_n] - \natp_n ;  \mu_n - \mu^* \rangle
+	+ \E[\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle ]
+	\nonumber\\
+%	& \quad \quad \quad 
+%	- \inv{n+n_0} \smallsum_i \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
 	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
-	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle 
-	+ \frac{n}{n+n_0} \variance{\expect{\langle \hat \natp_n  ; T(X_0) - \mu^* \rangle } }
+	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
+	+ \variance{\Cov(\hat \natp_n ; \hat \mu_n ) }
 \end{align}
@@
 Note that unless $\conj$ is $\ell^2$, or we are in a degenerate case with $\hat \mu_n$ a Dirac, $\theta_n \neq \expect{\hat \theta_n}$.
-At the last line, we used the property that all data points play a symmetric role to remove the sum over $i$.
 
@@
 \subsection{Pure Variance Term}
-The pure variance term $\expect{\langle \hat \natp_n  ; T(X_0) - \mu^* \rangle }$ is the covariance between $\hat \natp_n$ and $T(X_0)$. 
-In a way it is a modified variance of $T(X)$, with the $\nabla \conj$ operator in between, and a $\inv{n+n_0}$ multiplicative factor within $\hat \natp_n = \nabla \conj( \frac{T(X_0)}{n+n_0} + \dots) $.
-
-\paragraph{Positivity.}
-It is positive thanks to the monotonicity of $\nabla \conj$
+The pure variance term $\Cov(\hat \natp_n ; \hat \mu_n )$   can be seen as a non-linear variance of $\hat \mu_n$.
+Indeed, it is positive thanks to the monotonicity of $\nabla \conj$
 \begin{align}
@@
 \end{align}
-Taking the expectation, removing the constant $\theta_n$ from the covariance, and expanding $\hat \mu_n - \mu_n$ we get
-\begin{align}
-	\implies 
-	0 \leq \expect{\langle \hat \natp_n - \natp_n ; \hat \mu_n - \mu_n \rangle}
-	&= \expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle} \\
+Taking the expectation, (which allows removing the constant $\theta_n$) 
+\begin{align}
+	\expect{\langle \hat \natp_n - \natp_n ; \hat \mu_n - \mu_n \rangle} 
+	= \Cov(\hat \natp_n ; \hat \mu_n ) \geq 0
+\end{align}
+Consequently, we will now write about this covariance as a variance, with the monotone operator $\nabla \conj$ (or equivalently $\nabla \logpart$) plugged in the middle
+\begin{align}
+	\Var_{\nabla \conj}(\hat \mu_n) := \Cov(\nabla\conj(\hat \mu_n) ; \hat \mu_n ) \geq 0 \; .
+\end{align}
+
+\paragraph{Expansion}
+We can expand  $\hat \mu_n - \mu_n$ to get another formulation of this variance
+\begin{align}
+	\expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle}
 	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
 	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
-\end{align}
-which concludes the proof of positivity.
+	= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
+\end{align}
+where we used the property that all data points play a symmetric role to remove the sum over $i$.
 

### 2021-04-14 main.tex
Upper bound for variance and mixed terms of the example

@@
 \newcommand{\lr}{\gamma} % learning rate, or step-size
+
+\newcommand{\MAPm}{\hat \mu_n}
+\newcommand{\MAPt}{\hat \natp_n}
 
@@
 
+\begin{example}
+		We need a formula  for $\E[\MAPt]$ the expectation of the natural MAP estimate
+	\begin{align}
+		\E[\MAPt] &= -\frac{n+n_0}{2}\expect{\inv{n_0 \mu_0 + \sum_i X_i^2}}  \\
+		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
+		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{a + \chi^2(n)}}
+	\end{align}
+	where $\sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
+	\begin{align}
+			a=n_0 \frac{\mu_0}{\mu^*} \; .
+	\end{align}
+	This expectations is not simple to solve, but using wolfram, we get  the formula :
+	\begin{align}
+		\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])
+	\end{align}
+	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
+	\begin{align}
+		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
+	\end{align}
+	Now our goal is to bound this function with simpler functions.
+	Fortunately, mathematicians have been working on these integrals for decades.
+	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8#E1}{verifies the bound}
+	\begin{align}
+		\half \log(1 + \frac{2}{x}) 
+		\leq e^x E_1(x) 
+		\leq \log(1+\inv{x}) \leq \inv{x} \; .
+	\end{align}
+	For larger $n$, we have \href{https://dlmf.nist.gov/8.19#E21}{the general bound}
+	\begin{align}
+		\inv{x+n}
+		\leq e^x E_k(x) 
+		\leq \inv{x + k - 1}
+	\end{align}
+	so that 
+	\begin{align}
+		\inv{a+n}
+		\leq \expect{\inv{a + \chi^2(n)}}
+		\leq \inv{a + n - 2}
+	\end{align}
+	so that
+	\begin{align}
+		\frac{n +n_0}{2 \mu^* (a + n)}	
+		&\leq - \E[\MAPt]
+		\leq \frac{n +n_0}{2 \mu^* (a + n - 2)}	\\
+		\iff \inv{2 \mu_n}
+		&\leq - \E[\MAPt]
+		\leq \inv{2 \mu_n - \frac{4 \mu^*}{n + n_0}}
+		\label{eq:theta_expectation_bound}
+	\end{align}
+	because $\frac{\mu^* (a + n)}{n +n_0} = \mu_n$. Remark how we almost bounded the deviation between $\E[\MAPt]$ and $\natp_n$.
+\end{example}
+
+	\begin{figure}[ht]
+    	\centering
+	    \includegraphics[width=.7\textwidth]{figs/wolfram.png}
+    	\label{fig:wolfram}
+	\end{figure}
+
 \subsection{Pure Bias Term}
@@
 \end{align}
+
+\begin{example}
+	\begin{align}
+		\bregmanconj(\mu^*; \mu_n)
+		 &= \frac{\mu^*}{\mu_n} -1  - \log \frac{\mu^*}{\mu_n}  \\
+		 & = \frac{n+n_0}{a + n} - 1 - \log \frac{n+n_0}{a +n} \\
+		 & = \frac{n_0 - a}{a + n} - \log \frac{n+n_0}{a +n} \\		 
+	\end{align}
+\end{example}
 
@@
 \begin{example}
-	For the gaussian variance MAP, the co-variance term is
-	\begin{align}
-		\frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X) )  
-		&= -\frac{n}{2} \expect{ \frac{X_0^2 - \mu^*}{\sum_i T(X_i)  + n_0 \mu_0} } \\
-		&=  \half \expect{ \frac{n - \chi^2(n)}{\chi^2(n) + n_0 \frac{\mu_0}{\mu^*}}} \\
-		& = \half[n] \expect{\inv{\chi^2(n) + a}} - \half \expect{\inv{1 + \frac{a}{\chi^2(n)}}}
-	\end{align}
-	where $a=n_0 \frac{\mu_0}{\mu^*}$, and at the second line we re-introduced the $X_i$ to form a chi-square. 
-	These expectations are not simple to solve, but using wolfram alpha, we get  some formulas :
-	\begin{align}
-		\expect{\inv{\chi^2(n) + a}}  = \half e^{\frac{a}{2}} \text{ExpIntegralE}[\half[n],\half[a]]
-	\end{align}
-	where the \href{https://reference.wolfram.com/language/ref/ExpIntegralE.html}{exponential integral function} is defined as 
-	\begin{align}
-		\text{ExpIntegralE}[n , z] = E_n(z) = \int_1^\infty \frac{e^{-z t} }{t^n} dt = z^{n-1} \Gamma(1-n , z)
-	\end{align}
-	with the upper incomplete gamma function.
-	Now our goal is to find a simple upper bound on this function.
+	For the gaussian variance MAP, the variance term is
+		\begin{align}
+			\Cov(\hat \natp_n ; \hat \mu_n )  
+			= \E[\MAPt \MAPm] - \E[\MAPt] \E[\MAPm] 
+			=  - \half - \E[\MAPt] \mu_n
+	\end{align}
+	where we used the fact that $\natp \mu = -\half$ for all valid pairs $(\natp, \mu)$. 
+	Plugging-in \eqref{eq:theta_expectation_bound} yield
+	\begin{align}
+		-\half + \frac{\mu_n}{2 \mu_n}
+		&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
+		\leq -\half + \half \frac{a + n}{a + n - 2} \\
+		\iff 0&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
+		\leq \half \frac{a+n - (a + n -2)}{a + n - 2} \\
+		\iff 0&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
+		\leq \inv{a +n - 2}
+	\end{align}
+	which is exactly the kind of upper bounds we were looking for !
 \end{example}
@@
 
+\begin{example}
+	As we have the tight bounds \eqref{eq:theta_expectation_bound} for the expectation of the natural parameter of the MAP, we can bound above and below this mixed term
+	\begin{align}
+		0
+		\leq \natp_n - \E[\MAPt]
+		\leq \frac{n_0 +n}{2 \mu^*} \left ( \inv{a+n-2} - \inv{a+n} \right)
+		 = \frac{n_0+n}{\mu^*(a+n)(a+n-2)}	 \in O(n^{-1}) \; .
+	\end{align} 
+	Remark how this asymptotic class differs from the $O(n^{-2})$ I expected. Is this bound less tight than I thought, or are my calculus wrong ? 
+	Also remark that $\natp_n$ is always greater than $\E[\MAPt]$, which is a good sanity-check because $\nabla \conj(\mu) = - \inv{2\mu}$ is a concave function.
+
+	The rate on the mixed term depends on the sign of $\mu^* - \mu_0 $
+	\begin{align}
+		\frac{n_0}{n+n_0} \langle  \mu^* - \mu_0 ; \natp_n - \E[\hat \natp_n] \rangle
+		\leq \begin{cases}
+			0 \ \text{ if } \ \mu^*\leq \mu_0, \\
+			\frac{n_0 - a}{(a+n)(a+n-2)} \ \text{ otherwise.}
+		\end{cases}
+	\end{align}
+\end{example}
+
 \section{Conjecture}
@@
 Besides this bias term should be visible from a simple mirror descent, with no noise or variance. Yet it is not trivial to get a $O(n^{-2})$ rate in this setting. See note by Frederik.
+
+\section{Stochastic Bregman Proximal Point}
+Remark that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
+This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
+This can also be seen at each step since
+\begin{align}
+    \hat \natp_{n+1} 
+    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
+    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
+\end{align}
+Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
+
+This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
+
+The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
+First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
+One needs an assumption on the quality of the stochastic estimates $f(.
+; x)$.
+Given the proof, the most straightforward such assumption (see paper notes) is 
+\begin{align}
+\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
+\end{align}
+which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
+This is exactly the same assumption as given by the SMD analysis.
+However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
+It does not immediately apply to the last iterate.
 
@@
 
+\citet{ostrovskii2021finite}  shows a convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant.
+This convergence rate holds only after a given number of samples have been seen.
+This is because they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.
+
 \subsection{Alternative Research Tracks}
@@
 \end{proof}
-
-\section{Stochastic Bregman Proximal Point}
-Remark that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
-This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
-This can also be seen at each step since
-\begin{align}
-    \hat \natp_{n+1} 
-    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
-    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
-\end{align}
-Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
-
-This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
-
-The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
-First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
-One needs an assumption on the quality of the stochastic estimates $f(.
-; x)$.
-Given the proof, the most straightforward such assumption (see paper notes) is 
-\begin{align}
-\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
-\end{align}
-which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
-This is exactly the same assumption as given by the SMD analysis.
-However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
-It does not immediately apply to the last iterate.
 

### 2021-04-14 main.tex
New lower bound on the logarithm

@@
 \end{align}
+as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
+\begin{align}
+	1 - \inv{z} \leq \log(z) 
+	\label{eq:log_bound}\\
+	\implies \phi(z) &\leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z}
+\end{align}
 as illustrated in Figure~\ref{fig:phi}.
@@
 \begin{example}
-	\paragraph{Exact computation}
+	\paragraph{MLE exact computation}
 	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \sum_i X_i^2 /n$.
@@
 \end{example}
+
+\begin{example}
+	\paragraph{MAP upper bound.}
+	Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ but it does for $\expect{\frac{\mu^*}{\MAPm}} = \expect{\frac{\MAPt}{\natp^*}}$. 
+	To start, let us plug in the definition of $\MAPm$
+	\begin{align}
+		\E[\MAPt] &= -\frac{n+n_0}{2}\expect{\inv{n_0 \mu_0 + \sum_i X_i^2}}  \\
+		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
+		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{a + \chi^2(n)}}
+	\end{align}
+	where $\sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
+	\begin{align}
+			a=n_0 \frac{\mu_0}{\mu^*} \; .
+	\end{align}
+	This expectations is not simple to solve, but using wolfram, we get  the formula :
+	\begin{align}
+		\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])
+		\label{eq:wolfram_input}
+	\end{align}
+	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
+	\begin{align}
+		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
+	\end{align}
+	Now our goal is to bound this function with simpler functions.
+	Fortunately, mathematicians have been working on these integrals for decades.
+	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8#E1}{verifies the bound}
+	\begin{align}
+		\half \log(1 + \frac{2}{x}) 
+		\leq e^x E_1(x) 
+		\leq \log(1+\inv{x}) \leq \inv{x} \; .
+	\end{align}
+	For larger $n$, we have \href{https://dlmf.nist.gov/8.19#E21}{the general bound}
+	\begin{align}
+		\inv{x+n}
+		\leq e^x E_k(x) 
+		\leq \inv{x + k - 1}
+	\end{align}
+	so that 
+	\begin{align}
+		\inv{a+n}
+		\leq \expect{\inv{a + \chi^2(n)}}
+		\leq \inv{a + n - 2}
+	\end{align}
+	so that
+	\begin{align}
+		\frac{n +n_0}{2 \mu^* (a + n)}	
+		&\leq - \E[\MAPt]
+		\leq \frac{n +n_0}{2 \mu^* (a + n - 2)}	\\
+		\iff \inv{2 \mu_n}
+		&\leq - \E[\MAPt]
+		\leq \inv{2 \mu_n - \frac{4 \mu^*}{n + n_0}}	\\
+		\iff \frac{\mu^*}{\mu_n}
+		&\leq \expect{\frac{\mu^*}{\MAPm}} 
+		\leq \frac{\mu^*}{\mu_n - \frac{4 \mu^*}{n + n_0}}  
+		\label{eq:theta_expectation_bound}
+	\end{align}
+	because $\frac{\mu^* (a + n)}{n +n_0} = \mu_n$.
+Now using the bound \eqref{eq:log_bound} on the logarithm, we get
+\begin{align}
+	2\expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq \expect{\frac{\mu^*}{\MAPm}}  + \expect{\frac{\MAPm}{\mu^*}} - 1 \\
+	&\leq \frac{\mu^*}{\mu_n - \frac{4 \mu^*}{n + n_0}}   + \frac{\mu_n}{\mu^*} - 1 \\
+	&= \frac{n_0 + n}{a +n - 2} + \frac{a+n}{n_0 + n} - 2 \\
+	&= \frac{n_0 - a + 2}{a +n - 2} + \frac{a - n_0}{n_0 + n} \\
+	&= \frac{2}{a+n-2} + \frac{(n_0 - a)^2}{(a+n-2)(n_0 + n)}
+\end{align}
+where we remind you that $a=n_0\frac{\mu_0}{\mu^*}$. This inequality highlights a clear variance-bias decomposition, to compare with the MLE.
+\end{example}
+
+	\begin{figure}[ht]
+    	\centering
+	    \includegraphics[width=.7\textwidth]{figs/wolfram.png}
+	    \caption{Output of Wolfram when queried for the integral \eqref{eq:wolfram_input}}
+    	\label{fig:wolfram}
+	\end{figure}
 
@@
 
-\begin{example}
-		We need a formula  for $\E[\MAPt]$ the expectation of the natural MAP estimate
-	\begin{align}
-		\E[\MAPt] &= -\frac{n+n_0}{2}\expect{\inv{n_0 \mu_0 + \sum_i X_i^2}}  \\
-		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
-		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{a + \chi^2(n)}}
-	\end{align}
-	where $\sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
-	\begin{align}
-			a=n_0 \frac{\mu_0}{\mu^*} \; .
-	\end{align}
-	This expectations is not simple to solve, but using wolfram, we get  the formula :
-	\begin{align}
-		\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])
-	\end{align}
-	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
-	\begin{align}
-		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
-	\end{align}
-	Now our goal is to bound this function with simpler functions.
-	Fortunately, mathematicians have been working on these integrals for decades.
-	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8#E1}{verifies the bound}
-	\begin{align}
-		\half \log(1 + \frac{2}{x}) 
-		\leq e^x E_1(x) 
-		\leq \log(1+\inv{x}) \leq \inv{x} \; .
-	\end{align}
-	For larger $n$, we have \href{https://dlmf.nist.gov/8.19#E21}{the general bound}
-	\begin{align}
-		\inv{x+n}
-		\leq e^x E_k(x) 
-		\leq \inv{x + k - 1}
-	\end{align}
-	so that 
-	\begin{align}
-		\inv{a+n}
-		\leq \expect{\inv{a + \chi^2(n)}}
-		\leq \inv{a + n - 2}
-	\end{align}
-	so that
-	\begin{align}
-		\frac{n +n_0}{2 \mu^* (a + n)}	
-		&\leq - \E[\MAPt]
-		\leq \frac{n +n_0}{2 \mu^* (a + n - 2)}	\\
-		\iff \inv{2 \mu_n}
-		&\leq - \E[\MAPt]
-		\leq \inv{2 \mu_n - \frac{4 \mu^*}{n + n_0}}
-		\label{eq:theta_expectation_bound}
-	\end{align}
-	because $\frac{\mu^* (a + n)}{n +n_0} = \mu_n$. Remark how we almost bounded the deviation between $\E[\MAPt]$ and $\natp_n$.
-\end{example}
-
-	\begin{figure}[ht]
-    	\centering
-	    \includegraphics[width=.7\textwidth]{figs/wolfram.png}
-    	\label{fig:wolfram}
-	\end{figure}
-
 \subsection{Pure Bias Term}
@@
 \begin{example}
+We can use the classical lower bound on the logarithm to upper bound the pure bias term 
 	\begin{align}
@@
 		 &= \frac{\mu^*}{\mu_n} -1  - \log \frac{\mu^*}{\mu_n}  \\
-		 & = \frac{n+n_0}{a + n} - 1 - \log \frac{n+n_0}{a +n} \\
-		 & = \frac{n_0 - a}{a + n} - \log \frac{n+n_0}{a +n} \\		 
-	\end{align}
+		 &\leq \frac{(\mu^*- \mu_n)^2}{2 \mu^* \mu_n} 
+		 = \half ( \frac{\mu^*}{\mu_n} + \frac{\mu_n}{\mu^*	}) - 1 \\		 
+		 & =  \half (\frac{n_0 + n}{a + n} -1 + \frac{a +n}{n_0+n} -1) \\
+		 & =  \half (\frac{n_0 - a}{a+n} + \frac{a - n_0}{n_0 + n}) \\
+		 & = \half[n_0-a] (\inv{a+n} - \inv{n_0+n}) \\
+		 &= \half[n_0-a] \frac{n_0 + n - a- n}{(a+n)(n_0+n)} \\
+		 &= \frac{(n_0-a)^2}{2(a+n)(n_0+n)}
+	\end{align}
+	which is a $O(n^{-2})$ term, worth $0$ when $n_0=a$, eg $\mu_0=\mu^*$, exactly as we expected.
 \end{example}
@@
 		\end{cases}
+	\end{align}
+\end{example}
+
+\subsection{Putting it all together}
+\begin{example}
+\begin{align}
+	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	\leq \bias{\frac{(n_0-a)^2}{2(a+n)(n_0+n)}}
+	+ \frac{\max(n_0 - a , 0)}{(a+n)(a+n-2)}
+	+ \variance{\frac{1}{a+n-2}}
 \end{align}

### 2021-04-15 main.tex
Apply the log lower bound to the MLE

@@
 	&= \frac{n_0 - a + 2}{a +n - 2} + \frac{a - n_0}{n_0 + n} \\
-	&= \frac{2}{a+n-2} + \frac{(n_0 - a)^2}{(a+n-2)(n_0 + n)}
-\end{align}
-where we remind you that $a=n_0\frac{\mu_0}{\mu^*}$. This inequality highlights a clear variance-bias decomposition, to compare with the MLE.
+	&= \frac{2}{a+n-2} + \frac{(n_0 - a)^2}{(a+n-2)(n_0 + n)} \\
+	\implies \expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a)^2}{2(a+n-2)(n_0 + n)} 
+\end{align}
+where recall that $a=n_0\frac{\mu_0}{\mu^*}$. 
+This inequality highlights a clear variance-bias decomposition, to compare with the MLE.
+Unfortunately it is not tight asymptotically as the inequality \eqref{eq:log_bound} is far from tight.
+We are basically losing a factor 2 compared to $1/2n$.
+
+\paragraph{MLE Loose Upper Bound.}
+If we apply \eqref{eq:log_bound} to the MLE instead, we get the upper bound
+\begin{align}
+	\expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})}
+	&\leq  \half \left( \expect{\frac{n}{\chi^2(n)}} + \expect{\frac{\chi^2(n)}{n}} - 2 \right ) \\
+	&= \half \left( \frac{n}{n-2} + 1 - 2 \right)  
+	= \frac{1}{n-2}
+\end{align}
+which is exactly $1/2n$ larger than \eqref{eq:tight_bounds} which we got thanks to an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.
 \end{example}

### 2021-04-15 main.tex
Remove obsolete conjecture

@@
 
-\section{Conjecture}
-
-We know that asymptotically, the convergence rate is dominated by $O(n^{-1})$ variance term. To this we may add a bias term in $O(n^{-2})$. 
-We also know from the $\cN(0,\mu_*)$ example that the MLE can have infinite loss when there is not enough samples.
-This is not the case of the MAP which always has a bounded loss, since it can never reach the boundary of the marginal polytope (the domain of $\mu$).
-I really want to get a bound for the MAP of a Gaussian variance, to get inspiration from its shape. 
-Thus we may have a bias term which is infinite for $n_0=0$ (eg no prior, MLE), and that decreases as $n_0$ increases, except that the convergence speed gets worse and worse for larger $n_0$, eg
-\begin{align}
-    \frac{d}{2n} +  \log \left (\inv{n_0} \right) \left(1+ \frac{n}{n_0} \right)^{-2} \bregman(\natp_*; \natp_0)
-\end{align}
-Besides this bias term should be visible from a simple mirror descent, with no noise or variance. Yet it is not trivial to get a $O(n^{-2})$ rate in this setting. See note by Frederik.
-
 \section{Stochastic Bregman Proximal Point}

### 2021-04-15 main.tex
Present clearly the optimization problem for Loizou

@@
 
+\section{Optimization Perspective}
+\paragraph{Problem.}
+Our goal is to solve the following problem
+\begin{align}
+	\min_\natp f(\natp) := -\langle \mu^* ; \natp \rangle + \logpart(\natp)
+\end{align}
+where $\logpart$ is a convex function and we only have access to stochastic oracles $X \sim p(X | \natp^*)$, meaning we can compute stochastic values and gradients 
+\begin{align}
+	f(\natp ; X) &= -\langle T(X) ; \natp \rangle + \logpart(\natp) \\
+	\nabla f(\natp ; X) &= A(\natp) - T(X)
+\end{align}
+such that $T(X) = \mu^*$.
+
+\paragraph{Algorithms.}
+We observe $n$ data points $X_i$. 
+We are studying two algorithms to solve this problem : the MLE and the MAP
+\begin{align}
+	\MAPm^\text{MLE} &= \inv{n}\sum_{i=1}^n T(X_i) 
+	&\MAPt^\text{MLE} &= \nabla\logpart^{-1}(\MAPm^\text{MLE}) \\
+	\MAPm^\text{MAP} &= \frac{n_0\mu_0 + \smallsum_{i=1}^n T(X_i)}{n_0 + n}
+	&\MAPt^\text{MAP} &= \nabla\logpart^{-1}(\MAPm^\text{MAP})	
+\end{align}
+
+\paragraph{Interpretation.}
+As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms. 
+\begin{itemize}
+	\item stochastic bregman proximal point with divergence $\bregman$, initialization $\natp_0$, and learning rate $\inv{n_0+n}$,
+	\item stochastic mirror descent with divergence $\bregman$, initialization $\natp_0$, and learning rate  $\inv{n_0+n +1}$.
+\end{itemize}
+In both cases, $f(\natp)$ is convex and 1-smooth / 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.
+
+\paragraph{Other assumptions.}
+An assumption pointed out by Nicolas Loizou is 
+\begin{align}
+	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
+\end{align}
+for which a sufficient condition is 
+\begin{align}
+\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
+\end{align}
+which holds in a lot of settings but not all.
+For instance, a gaussian $\cN(\mu,\sigma^2)$ can overfit on a single data point and give it infinite density -- eg converge to a Dirac.
+However we can overcome this issue by grouping data points : a gaussian can not overfit on two points.
+Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$.
+So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)
+
 \section{Stochastic Bregman Proximal Point}

### 2021-05-10 main.tex
Separate theorem and proof for gaussian variance

@@
 \begin{align}
-	\KL( \sigma_n^2 ; \sigma_*^2) 
+	\KL( \sigma_*^2 ; \sigma_n^2 ) 
 	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
@@
 \begin{align}
-	1 - \inv{z} \leq \log(z) 
+	1 - \inv{z} &\leq \log(z) 
 	\label{eq:log_bound}\\
@@
 \begin{example}
-	\paragraph{MLE exact computation}
-	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \sum_i X_i^2 /n$.
-The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$.
-Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
-	\begin{align}
-		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
-		= \expect{\frac{n}{\chi^2(n)} - 1} =
-		\begin{cases}
-			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
-			+\infty \  \text{ otherwise. }
+	\begin{theorem}[MLE tight upper bound]
+	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
+	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
+	\begin{align}
+		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
+			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
+			\label{eq:tight_bounds}
+	\end{align}
+\end{theorem}
+Note that this upper bound is asymptotically tight.
+ There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}.
+The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension.
+We do not perform these calculations for now as we want to focus on the simplest possible examples.
+
+As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
+To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
+Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
+We now present a lemma on its expected inverse.
+\begin{lemma}[Expected MAP natural parameter]
+	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+	\begin{align}
+		\frac{\mu^*}{\mu_n}
+		\leq \expect{\frac{\mu^*}{\MAPm}} 
+		= \expect{\frac{\MAPt}{\natp^*}} 
+		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
+	\end{align}
+\end{lemma}
+Using this lemma, along with~\eqref{eq:log_bound}, we get the following result.
+\begin{theorem}[MAP Upper Bound]
+ The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
+\begin{equation}
+	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
+	\leq \begin{cases}
+		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
+		\frac{1+n_0}{4} e^{\half[a]} \sqrt{\frac{2 \pi}{a}} \mathrm{erfc}(\sqrt{\half[a]})  - \half  + \frac{a - n_0}{2(1+n_0)} \ \text{if}\ n=1,\\
+		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} if n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
 	\end{cases}
-	\end{align}
-	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
-	\begin{align}
-		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
-	\end{align}
-	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
-	Consequently the suboptimality of the MLE has a closed form solution
-	\begin{align}
-	\expect{\bregmanconj( \mu_*; \mu_n) }
-	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
-	& =
-	\begin{cases}
-		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
-			+\infty \  \text{ otherwise. }
-	\end{cases}
-	\end{align}
-	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
-	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
-	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
-	\begin{align}
-		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
-	\end{align}
-	  For $n\geq 3$,
-	\begin{align}
-		\inv{n-2} - \inv{n}
-		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
-		\leq \inv{n-2} - \inv{2 n} \\
-		\iff
-			\frac{2}{n(n-2)}
-			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
-			\leq \inv{2n} +\frac{2}{n(n-2)}
-			\label{eq:tight_bounds}
-	\end{align}
-	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
-	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
-	\begin{align}
-		\expect{\bregmanconj( \mu_*; \mu_n) }
-		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
-	\end{align}
-	Unfortunately, formulas for the MAP seem to be much less nice.
-There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}.
-Again the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension.
-We do not perform these calculations for now as we want to focus on the simplest possible examples.
-\end{example}
-
-\begin{example}
-	\paragraph{MAP upper bound.}
+\end{equation}
+\end{theorem}
+
+\begin{proof}
 	Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ but it does for $\expect{\frac{\mu^*}{\MAPm}} = \expect{\frac{\MAPt}{\natp^*}}$. 
@@
 	\begin{align}
-		\E[\MAPt] &= -\frac{n+n_0}{2}\expect{\inv{n_0 \mu_0 + \sum_i X_i^2}}  \\
-		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
-		&= -\frac{n+n_0}{2 \mu^*}\expect{\inv{a + \chi^2(n)}}
-	\end{align}
-	where $\sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
+		\expect{\frac{\mu^*}{\MAPm}} 
+		= \expect{\frac{\MAPt}{\natp^*}} 
+		&= expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  \\
+		&= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
+		&= (n_0 +n) \expect{\inv{a + \chi^2(n)}}
+	\end{align}
+	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
 	\begin{align}
@@
 	\end{align}
-	This expectations is not simple to solve, but using wolfram, we get  the formula :
-	\begin{align}
-		\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])
+	This expectations is not simple to solve, but using the trick
+	\begin{align}
+		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} 
+		\ \implies \ \inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt
+		\\
+		\implies 
+		&\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,
+	\end{align} 
+	we get  the formula :
+	\begin{align}
+		(n_0 +n) \expect{\inv{a + \chi^2(n)}}  = \half[n_0+n] e^{\frac{a}{2}} E_{\half[n]}(\half[a])
 		\label{eq:wolfram_input}
@@
 	Fortunately, mathematicians have been working on these integrals for decades.
-	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8#E1}{verifies the bound}
+	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
 	\begin{align}
@@
 	\end{align}
-	For larger $n$, we have \href{https://dlmf.nist.gov/8.19#E21}{the general bound}
-	\begin{align}
-		\inv{x+n}
+	or 
+	\begin{align}
+		E_{\half[1]}(\half[a]) = \sqrt{\frac{a}{2\pi}}
+	\end{align}
+	For larger $n$, we have \href{https://dlmf.nist.gov/8.19\#E21}{the general bound}
+	\begin{align}
+		\inv{x+k}
 		\leq e^x E_k(x) 
@@
 		&\leq \expect{\frac{\mu^*}{\MAPm}} 
-		\leq \frac{\mu^*}{\mu_n - \frac{4 \mu^*}{n + n_0}}  
+		\leq \frac{\mu^*}{\mu_n - \frac{2 \mu^*}{n + n_0}}  
 		\label{eq:theta_expectation_bound}
@@
 	&\leq \expect{\frac{\mu^*}{\MAPm}}  + \expect{\frac{\MAPm}{\mu^*}} - 1 \\
-	&\leq \frac{\mu^*}{\mu_n - \frac{4 \mu^*}{n + n_0}}   + \frac{\mu_n}{\mu^*} - 1 \\
+	&\leq \frac{\mu^*}{\mu_n - \frac{2 \mu^*}{n + n_0}}   + \frac{\mu_n}{\mu^*} - 1 \\
 	&= \frac{n_0 + n}{a +n - 2} + \frac{a+n}{n_0 + n} - 2 \\
@@
 	\implies \expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a)^2}{2(a+n-2)(n_0 + n)} 
+	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a)^2}{2(a+n-2)(n_0 + n)} \\
+	&\leq \frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} 
 \end{align}
@@
 This inequality highlights a clear variance-bias decomposition, to compare with the MLE.
-Unfortunately it is not tight asymptotically as the inequality \eqref{eq:log_bound} is far from tight.
+Unfortunately it is not tight asymptotically as the inequality \eqref{eq:log_bound} is not quadratically tight around $1$.
 We are basically losing a factor 2 compared to $1/2n$.
-
-\paragraph{MLE Loose Upper Bound.}
-If we apply \eqref{eq:log_bound} to the MLE instead, we get the upper bound
+\end{proof}
+
+\end{example}
+
+\section{Proofs for Gaussian Variance}
+\subsection{MLE Tight Upper Bound}
+	\begin{theorem}[MLE tight upper bound]
+	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
+	Its expected suboptimality is infinite when $n\leq 2$ and 
+	\begin{align}
+		 \expect{\bregmanconj( \mu_*; \mu_n) }
+			\leq \inv{2n} +\frac{2}{n(n-2)}
+	\end{align}
+	otherwise. Note that this upper bound is asymptotically tight.
+	\end{theorem}
+
+	\begin{proof}
+	The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$.
+Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
+	\begin{align}
+		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
+		= \expect{\frac{n}{\chi^2(n)} - 1} =
+		\begin{cases}
+			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
+			+\infty \  \text{ otherwise. }
+		\end{cases}
+	\end{align}
+	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
+	\begin{align}
+		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
+	\end{align}
+	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
+	Consequently the suboptimality of the MLE has a closed form solution
+	\begin{align}
+	\expect{\bregmanconj( \mu_*; \mu_n) }
+	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
+	& =
+	\begin{cases}
+		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
+			+\infty \  \text{ otherwise. }
+	\end{cases}
+	\end{align}
+	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
+	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
+	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
+	\begin{align}
+		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
+	\end{align}
+	  For $n\geq 3$,
+	\begin{align}
+		\inv{n-2} - \inv{n}
+		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
+		\leq \inv{n-2} - \inv{2 n} \\
+		\iff
+			\frac{2}{n(n-2)}
+			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
+			\leq \inv{2n} +\frac{2}{n(n-2)}
+	\end{align}
+	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
+	\end{proof}
+
+	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
+	\begin{align}
+		\expect{\bregmanconj( \mu_*; \mu_n) }
+		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
+	\end{align}
+
+\subsection{MLE Loose Upper Bound}
+If we apply \eqref{eq:log_bound} to the MLE, we get the upper bound
 \begin{align}
@@
 \end{align}
-which is exactly $1/2n$ larger than \eqref{eq:tight_bounds} which we got thanks to an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.
-\end{example}
-
-	\begin{figure}[ht]
-    	\centering
-	    \includegraphics[width=.7\textwidth]{figs/wolfram.png}
-	    \caption{Output of Wolfram when queried for the integral \eqref{eq:wolfram_input}}
-    	\label{fig:wolfram}
-	\end{figure}
+which is exactly  an additive term $1/2n$ larger than \eqref{eq:tight_bounds}.
+This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.
 
@@
 \begin{align}
-	\min_\natp f(\natp) := -\langle \mu^* ; \natp \rangle + \logpart(\natp)
+	\min_\natp \E[f(\natp ; X)] := -\langle \E[T(X)] ; \natp \rangle + \logpart(\natp)
 \end{align}
@@
 	f(\natp ; X) &= -\langle T(X) ; \natp \rangle + \logpart(\natp) \\
-	\nabla f(\natp ; X) &= A(\natp) - T(X)
-\end{align}
-such that $T(X) = \mu^*$.
+	\nabla f(\natp ; X) &= \nabla \logpart(\natp) - T(X)
+\end{align}
+such that $\E[T(X)] = \mu^*$.
 

### 2021-05-11 main.tex
Start typing proof of integral

@@
 \begin{lemma}[Expected MAP natural parameter]
-	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+	For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
 	\begin{align}
@@
 \end{lemma}
-Using this lemma, along with~\eqref{eq:log_bound}, we get the following result.
+Not that the lower bound is a simple consequence of the convexity of the inverse function. Using this lemma, along with~\eqref{eq:log_bound}, we get the following result.
+\begin{theorem}[MAP Upper Bound]
+ Introducing $a=n_0 \frac{\mu_0}{\mu^*}$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
+\begin{equation}
+	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
+	\leq \begin{cases}
+		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
+		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
+		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
+	\end{cases}
+\end{equation}
+\end{theorem}
+
+\end{example}
+
+\section{Proofs for Gaussian Variance}
+\subsection{MLE Tight Upper Bound}
+	\begin{theorem}[MLE tight upper bound]
+	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
+	Its expected suboptimality is infinite when $n\leq 2$ and otherwise
+	\begin{align}
+		 \expect{\bregmanconj( \mu_*; \mu_n) }
+			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
+	\end{align}
+	\end{theorem}
+
+	\begin{proof}
+	The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$.
+Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
+	\begin{align}
+		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
+		= \expect{\frac{n}{\chi^2(n)} - 1} =
+		\begin{cases}
+			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
+			+\infty \  \text{ otherwise. }
+		\end{cases}
+	\end{align}
+	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
+	\begin{align}
+		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
+	\end{align}
+	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
+	Consequently the suboptimality of the MLE has a closed form solution
+	\begin{align}
+	\expect{\bregmanconj( \mu_*; \mu_n) }
+	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
+	& =
+	\begin{cases}
+		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
+			+\infty \  \text{ otherwise. }
+	\end{cases}
+	\end{align}
+	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
+	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
+	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
+	\begin{align}
+		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
+	\end{align}
+	  For $n\geq 3$,
+	\begin{align}
+		\inv{n-2} - \inv{n}
+		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
+		\leq \inv{n-2} - \inv{2 n} \\
+		\iff
+			\frac{2}{n(n-2)}
+			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
+			\leq \inv{2n} +\frac{2}{n(n-2)}
+	\end{align}
+	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
+	\end{proof}
+	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
+	\begin{align}
+		\expect{\bregmanconj( \mu_*; \mu_n) }
+		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
+	\end{align}
+
+\subsection{MLE Loose Upper Bound}
+If we apply \eqref{eq:log_bound} to the MLE, we get the upper bound
+\begin{align}
+	\expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})}
+	&\leq  \half \left( \expect{\frac{n}{\chi^2(n)}} + \expect{\frac{\chi^2(n)}{n}} - 2 \right ) \\
+	&= \half \left( \frac{n}{n-2} + 1 - 2 \right)  
+	= \frac{1}{n-2}
+\end{align}
+which is exactly  an additive term $1/2n$ larger than \eqref{eq:tight_bounds}.
+This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.
+
+\subsection{Expected MAP Natural Parameter}
+\begin{lemma}[Expected MAP natural parameter]
+	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+	\begin{align}
+		\frac{\mu^*}{\mu_n}
+		\leq \expect{\frac{\mu^*}{\MAPm}} 
+		= \expect{\frac{\MAPt}{\natp^*}} 
+		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
+	\end{align}
+\end{lemma}
+
+\begin{proof}
+	To start, let us plug in the definition of $\MAPm$
+	\begin{align}
+		\expect{\frac{\mu^*}{\MAPm}} 
+		= \expect{\frac{\MAPt}{\natp^*}} 
+		&= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  \\
+		&= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
+		&= (n_0 +n) \expect{\inv{a + \chi^2(n)}}
+	\end{align}
+	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
+	\begin{align}
+			a=n_0 \frac{\mu_0}{\mu^*} \; .
+	\end{align}
+	We reformulate this expectation with the trick
+	\begin{align}
+		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
+		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
+		\implies &\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,
+	\end{align} 
+	to get two integrals instead of one
+	\begin{align}
+		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
+		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
+		&= \int_0^\infty dx x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt  e^{-\half[(a+x)t]}  \\
+		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
+	\end{align}
+	where we used Fubini to switch integrals, without further justification.
+	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes \TODO{finish}.
+	we get  the formula :
+	\begin{align}
+		(n_0 +n) \expect{\inv{a + \chi^2(n)}}  = \half[n_0+n] e^{\frac{a}{2}} E_{\half[n]}(\half[a])
+		\label{eq:wolfram_input}
+	\end{align}
+	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
+	\begin{align}
+		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
+	\end{align}
+	Now our goal is to bound this function with simpler functions.
+	Fortunately, mathematicians have been working on these integrals for decades.
+	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
+	\begin{align}
+		\half \log(1 + \frac{2}{x}) 
+		\leq e^x E_1(x) 
+		\leq \log(1+\inv{x}) \leq \inv{x} \; .
+	\end{align}
+	When $n=1$, the integral can be written with the complementary error function
+	\begin{align}
+		E_{\half[1]}(\half[a]) = \sqrt{\frac{2\pi}{a}} \mathrm{erfc}(\sqrt{\half[a]}) \; .
+	\end{align}
+	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19\#E21}{the general bound}
+	\begin{align}
+		\inv{x+k}
+		\leq e^x E_k(x) 
+		\leq \inv{x + k - 1}
+	\end{align}
+	so that 
+	\begin{align}
+		\inv{a+n}
+		&\leq \expect{\inv{a + \chi^2(n)}}
+		\leq \inv{a + n - 2} \\
+		\iff \frac{n +n_0}{a + n}	
+		&\leq  \expect{\frac{\mu^*}{\MAPm}} 
+		\leq \frac{n +n_0}{a + n - 2}	\\
+		\iff \frac{\mu^*}{\mu_n}
+		&\leq \expect{\frac{\mu^*}{\MAPm}} 
+		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n + n_0}}  
+		\label{eq:theta_expectation_bound}
+	\end{align}
+	where we used $\frac{a + n}{n +n_0} = \mu_n$.
+\end{proof}
+
+\subsection{MAP Upper Bound}
 \begin{theorem}[MAP Upper Bound]
@@
 		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
-		\frac{1+n_0}{4} e^{\half[a]} \sqrt{\frac{2 \pi}{a}} \mathrm{erfc}(\sqrt{\half[a]})  - \half  + \frac{a - n_0}{2(1+n_0)} \ \text{if}\ n=1,\\
+		\bregmanconj( \mu_*; \mu_0) +\inv{a} \ \text{if}\ n=1,\\
 		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} if n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
@@
 
+$\frac{1+n_0}{4} e^{\half[a]} \sqrt{\frac{2 \pi}{a}} \mathrm{erfc}(\sqrt{\half[a]})  - \half  + \frac{a - n_0}{2(1+n_0)} $
+
 \begin{proof}
-	Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ but it does for $\expect{\frac{\mu^*}{\MAPm}} = \expect{\frac{\MAPt}{\natp^*}}$. 
-	To start, let us plug in the definition of $\MAPm$
-	\begin{align}
-		\expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\natp^*}} 
-		&= expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  \\
-		&= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
-		&= (n_0 +n) \expect{\inv{a + \chi^2(n)}}
-	\end{align}
-	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
-	\begin{align}
-			a=n_0 \frac{\mu_0}{\mu^*} \; .
-	\end{align}
-	This expectations is not simple to solve, but using the trick
-	\begin{align}
-		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} 
-		\ \implies \ \inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt
-		\\
-		\implies 
-		&\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,
-	\end{align} 
-	we get  the formula :
-	\begin{align}
-		(n_0 +n) \expect{\inv{a + \chi^2(n)}}  = \half[n_0+n] e^{\frac{a}{2}} E_{\half[n]}(\half[a])
-		\label{eq:wolfram_input}
-	\end{align}
-	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
-	\begin{align}
-		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
-	\end{align}
-	Now our goal is to bound this function with simpler functions.
-	Fortunately, mathematicians have been working on these integrals for decades.
-	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
-	\begin{align}
-		\half \log(1 + \frac{2}{x}) 
-		\leq e^x E_1(x) 
-		\leq \log(1+\inv{x}) \leq \inv{x} \; .
-	\end{align}
-	or 
-	\begin{align}
-		E_{\half[1]}(\half[a]) = \sqrt{\frac{a}{2\pi}}
-	\end{align}
-	For larger $n$, we have \href{https://dlmf.nist.gov/8.19\#E21}{the general bound}
-	\begin{align}
-		\inv{x+k}
-		\leq e^x E_k(x) 
-		\leq \inv{x + k - 1}
-	\end{align}
-	so that 
-	\begin{align}
-		\inv{a+n}
-		\leq \expect{\inv{a + \chi^2(n)}}
-		\leq \inv{a + n - 2}
-	\end{align}
-	so that
-	\begin{align}
-		\frac{n +n_0}{2 \mu^* (a + n)}	
-		&\leq - \E[\MAPt]
-		\leq \frac{n +n_0}{2 \mu^* (a + n - 2)}	\\
-		\iff \inv{2 \mu_n}
-		&\leq - \E[\MAPt]
-		\leq \inv{2 \mu_n - \frac{4 \mu^*}{n + n_0}}	\\
-		\iff \frac{\mu^*}{\mu_n}
-		&\leq \expect{\frac{\mu^*}{\MAPm}} 
-		\leq \frac{\mu^*}{\mu_n - \frac{2 \mu^*}{n + n_0}}  
-		\label{eq:theta_expectation_bound}
-	\end{align}
-	because $\frac{\mu^* (a + n)}{n +n_0} = \mu_n$.
-Now using the bound \eqref{eq:log_bound} on the logarithm, we get
+		Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus solely on the simpler upper bound given Now using the bound \eqref{eq:log_bound} on the logarithm, we get
 \begin{align}
@@
 We are basically losing a factor 2 compared to $1/2n$.
+by~\eqref{eq:log_bound}.
 \end{proof}
-
-\end{example}
-
-\section{Proofs for Gaussian Variance}
-\subsection{MLE Tight Upper Bound}
-	\begin{theorem}[MLE tight upper bound]
-	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
-	Its expected suboptimality is infinite when $n\leq 2$ and 
-	\begin{align}
-		 \expect{\bregmanconj( \mu_*; \mu_n) }
-			\leq \inv{2n} +\frac{2}{n(n-2)}
-	\end{align}
-	otherwise. Note that this upper bound is asymptotically tight.
-	\end{theorem}
-
-	\begin{proof}
-	The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$.
-Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
-	\begin{align}
-		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
-		= \expect{\frac{n}{\chi^2(n)} - 1} =
-		\begin{cases}
-			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
-			+\infty \  \text{ otherwise. }
-		\end{cases}
-	\end{align}
-	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
-	\begin{align}
-		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
-	\end{align}
-	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
-	Consequently the suboptimality of the MLE has a closed form solution
-	\begin{align}
-	\expect{\bregmanconj( \mu_*; \mu_n) }
-	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
-	& =
-	\begin{cases}
-		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
-			+\infty \  \text{ otherwise. }
-	\end{cases}
-	\end{align}
-	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
-	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
-	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
-	\begin{align}
-		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
-	\end{align}
-	  For $n\geq 3$,
-	\begin{align}
-		\inv{n-2} - \inv{n}
-		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
-		\leq \inv{n-2} - \inv{2 n} \\
-		\iff
-			\frac{2}{n(n-2)}
-			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
-			\leq \inv{2n} +\frac{2}{n(n-2)}
-	\end{align}
-	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
-	\end{proof}
-
-	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
-	\begin{align}
-		\expect{\bregmanconj( \mu_*; \mu_n) }
-		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
-	\end{align}
-
-\subsection{MLE Loose Upper Bound}
-If we apply \eqref{eq:log_bound} to the MLE, we get the upper bound
-\begin{align}
-	\expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})}
-	&\leq  \half \left( \expect{\frac{n}{\chi^2(n)}} + \expect{\frac{\chi^2(n)}{n}} - 2 \right ) \\
-	&= \half \left( \frac{n}{n-2} + 1 - 2 \right)  
-	= \frac{1}{n-2}
-\end{align}
-which is exactly  an additive term $1/2n$ larger than \eqref{eq:tight_bounds}.
-This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.
 

### 2021-05-18 main.tex
Updates from Overleaf
Use shadow instead of border for frames

@@
 
-\definecolor{light-gray}{gray}{0.9}
-\usepackage[framemethod=TikZ]{mdframed} % frames with colored background
-\newenvironment{example}{
-	\begin{mdframed}[backgroundcolor=light-gray, roundcorner=5pt]
-}{		
-	\end{mdframed}
-}
+% frames with colored background
+\definecolor{light-gray}{gray}{0.85}
 \definecolor{myblue}{RGB}{199, 235, 255}
-\newenvironment{important}{
-	\begin{mdframed}[backgroundcolor=myblue, roundcorner=5pt]
-}{		
-	\end{mdframed}
-}
+\usepackage[framemethod=TikZ]{mdframed} 
+\usetikzlibrary{shadows}
+
+\newmdenv[% box with grey background
+    backgroundcolor=gray!15,
+    linewidth=0pt,
+    roundcorner=5pt,
+    shadow=true,
+	shadowsize=4pt
+]{example}
+\newmdenv[
+	backgroundcolor=myblue,
+	roundcorner=5pt, 
+	linewidth=0pt, 
+	shadow=true,
+	shadowsize=4pt
+]{important}
 
@@
 The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model 
-\begin{align}
-    p(X|\natp) &= \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
-\end{align}
+\begin{equation}
+	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
+\end{equation}
 where $\logpart$ is the log-partition function (aka normalization factor) 
@@
 	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
+\begin{important}
 	\begin{align}
@@
 	\end{align}
+\end{important}
 	We reformulate this expectation with the trick
@@
 		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
-		&= \int_0^\infty dx x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt  e^{-\half[(a+x)t]}  \\
+		&= \int_0^\infty dx\  x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt\  e^{-\half[(a+x)t]}  \\
 		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
@@
 	where we used Fubini to switch integrals, without further justification.
-	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes \TODO{finish}.
-	we get  the formula :
-	\begin{align}
-		(n_0 +n) \expect{\inv{a + \chi^2(n)}}  = \half[n_0+n] e^{\frac{a}{2}} E_{\half[n]}(\half[a])
-		\label{eq:wolfram_input}
+	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
+	\begin{align}
+		\expect{\inv{a + \chi^2(n)}}  
+		&= \frac{e^{\half[a]} }{2^{\half[n]+1}\Gamma(\half[n]) }\int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty 2\frac{dy}{t} \  (2\frac{y}{t})^{\frac{n}{2}-1} e^{-y} \\
+		&= \frac{e^{\half[a]} }{2\Gamma(\half[n]) } 
+		\underbrace{\int_1^\infty  \frac{e^{-\half[a t]}}{t^{\frac{n}{2}}}  dt}_{E_{\half[n]}(\half[a])  } \ 
+		\underbrace{\int_0^\infty  y^{\frac{n}{2}-1} e^{-y} dy}_{\Gamma(\half[n])}
+	\end{align}
+	so we finally get  the formula valid for all $n\geq 1$
+	\begin{align}
+		\boxed{\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])}
+		\label{eq:hard_expectation}
 	\end{align}
@@
 	Fortunately, mathematicians have been working on these integrals for decades.
-	When $n=2$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
+	When $n=2$, eg $k=1$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
 	\begin{align}
@@
 		\leq \log(1+\inv{x}) \leq \inv{x} \; .
-	\end{align}
-	When $n=1$, the integral can be written with the complementary error function
-	\begin{align}
-		E_{\half[1]}(\half[a]) = \sqrt{\frac{2\pi}{a}} \mathrm{erfc}(\sqrt{\half[a]}) \; .
 	\end{align}
@@
 	\end{align}
-	where we used $\frac{a + n}{n +n_0} = \mu_n$.
+	where we used $\frac{a + n}{n +n_0} = \frac{\mu_n}{\mu^*}$.
 \end{proof}
+
+We are left with a special case when  $n=1$, eg $k=\half$.
+Then the integral can be written with the complementary error function, or \href{https://dlmf.nist.gov/7.8#E1}{Mill's ratio} $M(x) = e^{x^2} \int_x^\infty e^{-t^2} dt$.
+Indeed, taking $t= x \sqrt{u}$
+\begin{align}
+	\int_x^\infty e^{-t^2} dt = x \int_1^\infty \frac{e^{-x^2 u}}{\sqrt{u}} du = x E_{\half}(x^2) \\
+	\implies M(x) = x e^{x^2} E_{\half}(x^2)
+\end{align}
+so that 
+\begin{align}
+		\half e^{\half[a]} E_{\half[1]}(\half[a]) 
+		= \half \sqrt{\frac{2}{a}} M(\sqrt{\half[a]})
+		= \inv{a} \sqrt{\half[a]} M(\sqrt{\half[a]}) \; .
+\end{align}
+and \href{https://dlmf.nist.gov/7.8#E5}{we known that} $x M(x) < \half$ so we get 
+\begin{align}
+	\expect{\inv{a + X^2}} < \inv{2 a}
+\end{align}
+FUCK That's  trivial as  it's true for 1/a.
 

### 2021-05-19 main.tex
Add empirical illustration of bias variance decomposition

@@
 			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
-			\label{eq:tight_bounds}
+			\label{eq:MLE_rate}
 	\end{align}
@@
 \end{lemma}
-Not that the lower bound is a simple consequence of the convexity of the inverse function. Using this lemma, along with~\eqref{eq:log_bound}, we get the following result.
+Not that the lower bound is a simple consequence of the convexity of the inverse function.
+Using this lemma, along withthe los upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP.
+We relate this rate convergence rate with the rate of MLE in Figure~\ref{fig:MAP_rate}.
 \begin{theorem}[MAP Upper Bound]
@@
 	\end{cases}
+	\label{eq:MAP_rate}
 \end{equation}
 \end{theorem}
+Using the same bound on the log~\eqref{eq:log_bound}, we can also get a loose bound on the MLE, which is useful for comparison purposes.
+\begin{equation}
+	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
+	\leq \inv{n-2} \; .
+	\label{eq:MLE_loose_rate}
+\end{equation}
 
 \end{example}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=.7\textwidth]{figs/MAP_rates/MAP_rate_n0=5.pdf}
+	\caption{
+	Convergence rate of MAP~\eqref{eq:MAP_rate} 
+	\label{fig:MAP_rate} against the ratio between initialization and optimum $ \frac{\mu_0}{\mu^*}$. 
+	We also report the tight convergence rate of MLE~\eqref{eq:MLE_rate} (dash-dotted horizontal lines), and the loose upper bound~\eqref{eq:MLE_loose_rate} (horizontal lines). 
+	We report these curves for various values of $n$, and $n_0=5$. 
+	We observe that the rate of MAP is significantly better (lower) than the rate of MLE, only when we guessed $\mu_0$ right, within an order of magnitude from $\mu^*$, and for a small number of sample ($n<10$).
+	This observation hints towards the optimality of MLE for large $n$. 
+	It makes me feel very interested in convergence rates of MAP in the overparametrized regime $n<\dim(T(X))$.
+	}
+\end{figure}
 
@@
 \end{align}
-which is exactly  an additive term $1/2n$ larger than \eqref{eq:tight_bounds}.
+which is exactly  an additive term $1/2n$ larger than \eqref{eq:MLE_rate}.
 This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.
@@
 \section{Bias-Variance Decomposition}
+\paragraph{Covariance.}
 Using Jensen inequality and inserting appropriate quantities, we obtain the following analog of the bias-variance decomposition:
@@
 	\pm \langle \underbrace{\nabla \conj( \mu_n)}_{=:\natp_n} ; \mu^* - \mu_n \rangle
-	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \\
+	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \nonumber \\
 	& \leq \conj(\mu^* ) 
@@
 	+ \E[\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle ]
-	\nonumber\\
-%	& \quad \quad \quad 
-%	- \inv{n+n_0} \smallsum_i \E[\langle \hat \natp_n ; \mu^* - T(X_i)  \rangle]\\
+	\nonumber \\
 	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
@@
 	\label{fig:bias-variance}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{figs/numerical_schema.pdf}
+	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
+	\label{fig:bias-variance-numerical}
+\end{figure}
+
+\paragraph{No Jensen.}
+In fact this previous decomposition is highly suboptimal. 
+We do not need to use Jensen inequality.
+\begin{align}
+	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	&= \conj(\mu^* ) 
+	\pm \conj(\mu_n)
+	-  \expect{\conj(\hat \mu_n)}
+	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \\
+	&= \conj(\mu^* ) 
+	- \conj(\mu_n)  + \E[\bregmanconj(\mu_n, \MAPm)]
+	+ \langle \E[\hat \natp_n] \pm \nabla \conj(\mu_n) ;  \mu_n - \mu^* \rangle
+	\nonumber\\
+	& = \bias{\bregmanconj(\mu^*; \mu_n)}
+	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
+	+ \variance{\E[\bregmanconj(\mu_n, \MAPm)]}
+\end{align}
+With Jensen, we were losing a Bregman term, as per the 3 point property, the covariance is equal to the symmetrized Bregman.
+\begin{align}
+	\Cov(\hat \natp_n ; \hat \mu_n )  
+	=\expect{\bregmanconj(\MAPm ; \mu_n)
+	+ \bregmanconj(\mu_n ; \MAPm)}
+\end{align}
+
+\paragraph{No Mixed Term.}
+In fact, it is possible to have no mixed term using the central point $\tilde \mu_n = \nabla \logpart(\E[\MAPt])$ instead of $\mu_n = \E[\MAPm]$,  exactly as described by \citet[Theorem 0.1]{pfau2013generalized}.
+\begin{align}
+	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
+	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
+\end{align}
+Now the big question, is : what kind of results do I get with such quantities ?
+In general, it seems that $\tilde \mu_n$ is a very central and interesting quantity to consider.
+We show these decompositions in Figure~\ref{fig:experimental_decomposition}.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/biasvariance/ratio=0.8_n0=10}.pdf}
+	\caption{
+	\textbf{Left:} training curves and analytic upper bound. 
+	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
+	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
+	}
+	\label{fig:experimental_decomposition}
 \end{figure}
@@
 	\end{align}
-	which is about twice the MLE bounds~\eqref{eq:tight_bounds}, exactly like as we observed for the gaussian mean.
+	which is about twice the MLE bounds~\eqref{eq:MLE_rate}, exactly like as we observed for the gaussian mean.
 \end{example}

### 2021-05-26 main.tex
Add more bias variance illustrations

@@
 	\centering
-	\includegraphics[width=\textwidth]{figs/numerical_schema.pdf}
+	\includegraphics[width=\textwidth]{figs/numerical_schema_n=3.pdf}
 	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
@@
 	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
-	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
+	&= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
+	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
+	\expect{\bregman(\MAPt; \natp^*)} 
+	&= \bias{\bregman(\E[\MAPt]; \natp^*)}
+	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}} 
 \end{align}
@@
 In general, it seems that $\tilde \mu_n$ is a very central and interesting quantity to consider.
-We show these decompositions in Figure~\ref{fig:experimental_decomposition}.
+We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
 
@@
 	\centering
-	\includegraphics[width=\textwidth]{{figs/biasvariance/ratio=0.8_n0=10}.pdf}
+	\includegraphics[width=\textwidth]{{figs/biasvariance/linear_ratio=0.8_n0=10}.pdf}
 	\caption{
-	\textbf{Left:} training curves and analytic upper bound. 
+	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
 	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
@@
 	}
-	\label{fig:experimental_decomposition}
+	\label{fig:variance_decomposition}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/gaussians/log_n0=1}.pdf}
+	\caption{
+	\textbf{Full gaussian $\cN(\mu, \sigma^2)$ example. Left:} training curves and analytic upper bound. 
+	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
+	\textbf{Right:} bias-variance decomposition, using the generalized mean.
+	}
+	\label{fig:gaussian_decomposition}
 \end{figure}

### 2021-05-31 main.tex
Add shadow boxes and clean figure links

@@
 \usetikzlibrary{shadows}
+% Using the pgf-blur package instead 
+% https://ctan.math.illinois.edu/graphics/pgf/contrib/pgf-blur/pgf-blur.pdf
+\usetikzlibrary{shadows.blur}
+
+\mdfdefinestyle{blurry-shadow}{
+	linewidth=0pt,
+    roundcorner=0pt,
+	shadow=true,
+ 	apptotikzsetting={\tikzset{mdfshadow/.style={blur shadow={%
+ 		shadow xshift=1pt,
+ 		shadow yshift=-1pt,
+ 		shadow scale=1.00,
+ 		shadow opacity = 50,
+ 		%shadow blur extra rounding,
+ 		shadow blur steps = 10,
+ 		shadow blur radius = 5pt,
+ 		every shadow
+ 	}}}},
+	linewidth=0pt,
+ }
 
 \newmdenv[% box with grey background
+	style=blurry-shadow,
     backgroundcolor=gray!15,
-    linewidth=0pt,
-    roundcorner=5pt,
-    shadow=true,
-	shadowsize=4pt
 ]{example}
 \newmdenv[
+	style=blurry-shadow,
 	backgroundcolor=myblue,
-	roundcorner=5pt, 
-	linewidth=0pt, 
-	shadow=true,
-	shadowsize=4pt
 ]{important}
@@
 
-\begin{figure}[ht]
+\begin{figure}[p]
 	\centering
-	\includegraphics[width=\textwidth]{figs/numerical_schema_n=3.pdf}
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=1.pdf}
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=2.pdf}
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=10.pdf}
 	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}

### 2021-06-18 main.tex
Move 3 point lemma to main text
Declutter gaussian example and bias-variance decomposition sections by moving to appendix

@@
 	style=blurry-shadow,
-	backgroundcolor=myblue,
+	backgroundcolor=Red!30,
 ]{important}
@@
 %% Faster processing
-\renewcommand\appendix{\end{document}}
+%\renewcommand\appendix{\end{document}}
 
@@
 \paragraph{Conjugate Prior}
+\TODO{Under-parametrized conjugate prior ?}
 The conjugate prior for $p(X|\natp)$ is
@@
 
-\section{Problem}
+\section{Open Problem}
 For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
@@
 For instance, if we plug in $\conj(\mu)= -\log(\mu)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
+
+\paragraph{Remark 2.}
+The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$.
+Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.
+
+Below we review all the attempts we made on this problem.
+
+\tableofcontents
+
+\section{Gaussian Variance Example}
 
@@
 \end{align}
+\end{example}
 In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
@@
 \begin{align}
-	1 - \inv{z} &\leq \log(z) 
-	\label{eq:log_bound}\\
-	\implies \phi(z) &\leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z}
-\end{align}
-as illustrated in Figure~\ref{fig:phi}.
-\end{example}
+	1 - \inv{z} \leq \log(z) 
+	\label{eq:log_bound} 
+	\implies \phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
+\end{align}
 
@@
 	\includegraphics[width=.4\textwidth]{figs/phi.pdf}
-	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$, so it is poorly approximated by quadratics.}
+	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics.}
 	\label{fig:phi}
@@
 \begin{example}
-	\begin{theorem}[MLE tight upper bound]
+	\begin{theorem}[MLE Tight Bound]
 	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
@@
 \end{theorem}
-Note that this upper bound is asymptotically tight.
+\end{example}
+This upper bound is asymptotically tight.
+We illustrate its behavior against empirical data in Figure~\ref{fig:curves}.
  There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}.
@@
 We now present a lemma on its expected inverse.
-\begin{lemma}[Expected MAP natural parameter]
+\begin{example}
+\begin{lemma}[Expected MAP Natural Parameter]
 	For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
@@
 		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
+	\label{eq:theta_expectation_bound}
 	\end{align}
 \end{lemma}
-Not that the lower bound is a simple consequence of the convexity of the inverse function.
-Using this lemma, along withthe los upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP.
-We relate this rate convergence rate with the rate of MLE in Figure~\ref{fig:MAP_rate}.
-\begin{theorem}[MAP Upper Bound]
+\end{example}
+Note that the lower bound is a simple consequence of the convexity of the inverse function. 
+Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
+\begin{example}
+\begin{theorem}[MAP Bound]
  Introducing $a=n_0 \frac{\mu_0}{\mu^*}$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
@@
 \end{theorem}
-Using the same bound on the log~\eqref{eq:log_bound}, we can also get a loose bound on the MLE, which is useful for comparison purposes.
+\end{example}
+Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
+We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
+\begin{example}
+\begin{corollary}[MLE Loose Bound]
 \begin{equation}
@@
 \end{equation}
-
+\end{corollary}
 \end{example}
@@
 
-\section{Proofs for Gaussian Variance}
-\subsection{MLE Tight Upper Bound}
+\section{Euclidean Behaviour}
+Often we can relate this Bregman suboptimality with the $\ell^2$ distance in mean parameter space. 
+\TODO{Add base measure to add here a discussion of the euclidean case -- the mean of a gaussian}
+\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
+    \leq 2L \norm{\mu^* - \mu}
+\end{align}
+so $\bregmanconj$ is $2L$-Lipschitz.
+Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this bound applies to the log-likelihood.
+\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
+\end{align}
+so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.
+
+\paragraph{$\ell^2$-norm Analysis}
+Let us make these statements more precise. A bound on the variance becomes a bound on the variance of the average
+\begin{align}
+	\Var T(X) = \E \norm{T(X) - \mu^*}^2 = \sigma^2 \\
+	\implies \E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
+\end{align}
+eg the variance of the mean is $n$ times smaller than the variance of the samples.
+Adding a reference mean $\mu_0$ to get the MAP yields
+\begin{align}
+	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
+	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
+	&= O\left(\frac{\sigma^2}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
+\end{align}
+so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !
+
+\section{Asymptotic Behaviour}
+In the limit, the MAP estimates reach a rate $O(1/2n)$. This is shown by approximating the Bregman divergence using a second order Taylor expansion
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
+    + O(\norm{\mu - \mu^*}^3)\\
+    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
+    + O(\norm{\mu - \mu^*}^3)
+\end{align}
+where we introduced the Mahalanobis distance induced by the matrix 
+\begin{align}
+    \nabla^2\conj(\mu^*) 
+    = \nabla^2\logpart(\natp^*)^{-1} 
+    = \Cov_{\natp^*}[T(X)]^{-1} 
+    =: \mSigma^{-1}  \; .
+\end{align}
+To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
+\begin{align}
+	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
+\end{align}
+where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
+\begin{align}
+	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
+\end{align}
+and for the MAP we  get 
+\begin{align}
+	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
+	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
+	&= \frac{d}{n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
+\end{align}
+or 
+\begin{align}
+\label{eq:MAP_asymptote}
+	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
+	= \frac{d}{2n} + O(n^{- \frac{3}{2}})
+\end{align}
+Remark how the variance does not even appear, only the dimension divided by $n$ really matters.
+I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
+Indeed this is because it is hidden within the Bregman divergence loss. 
+Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?
+
+\section{Bias-Variance Decomposition}
+\newcommand{\bias}[1]{\textcolor{Red}{\underbrace{\textcolor{black}{#1}}_{\text{bias}}}}
+\newcommand{\variance}[1]{\textcolor{Green}{\underbrace{\textcolor{black}{#1}}_{\text{variance}}}}
+
+\paragraph{Dual Expectation Pivot.}
+Let $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ be the MAP estimate of the mean (dual) parameter, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ the corresponding natural (primal) parameter, and
+$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ the expected MAP dual estimate
+Taking $\mu_n$ as a pivot, we can decompose the expected suboptimality in 3 terms -- e.g. 
+\begin{align}
+	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	&= \conj(\mu^* ) 
+	\pm \conj(\mu_n)
+	-  \expect{\conj(\hat \mu_n)}
+	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] 
+	\nonumber \\
+	&= \conj(\mu^* ) 
+	- \conj(\mu_n)  + \E[\bregmanconj(\mu_n, \MAPm)]
+	+ \langle \E[\hat \natp_n] \pm \nabla \conj(\mu_n) ;  \mu_n - \mu^* \rangle
+	\nonumber\\
+	& = \bias{\bregmanconj(\mu^*; \mu_n)}
+	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
+	+ \variance{\E[\bregmanconj(\mu_n, \MAPm)]} \; .
+	\label{eq:bias-mixed-variance}
+\end{align}
+Remark that unless $\conj$ is $\ell^2$, or unless we are in a degenerate case with $\hat \mu_n$ a Dirac, the dual and primal expectation represent different distributions $\theta_n \neq \expect{\hat \theta_n}$, which is why the mixed term (bias-variance scalar product) is non-zero.
+Asymptotically, these terms behave like : bias $O(n^{-2}$, mixed $O(n^{-3})$ and variance $O(n^{-1})$, but the story is different for finite samples.
+We know for sure that the bias and the variance term are positive, but the mixed bias-variance term may be negative, which removes some appeal of this decomposition. 
+Fortunately there is another decomposition revolving around the primal expectation.
+
+%\begin{figure}[ht]
+%	\centering
+%	\includegraphics[width=.8\textwidth]{figs/bias-variance.png}
+%	\caption{A schematic illustration of the different characters featured in the bias-variance decomposition.}
+%	\label{fig:bias-variance}
+%\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
+	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
+	\label{fig:bias-variance-numerical}
+\end{figure}
+
+\paragraph{Primal Expectation Pivot}
+Let $\tilde \theta_n := \expect{\hat \theta_n}$ and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the primal and dual parameters of the primal expectation of the MAP.
+As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
+\begin{align}
+	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	&= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
+	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
+	\expect{\bregman(\MAPt; \natp^*)} 
+	&= \bias{\bregman(\E[\MAPt]; \natp^*)}
+	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
+\end{align}
+
+\paragraph{Illustrations.}
+We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
+In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in both of these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
+
+We review each of these terms independently below.
+
+\begin{important}
+Goal = match variance assumptions with these bounds	
+\end{important}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/biasvariance/linear_ratio=0.8_n0=10}.pdf}
+	\caption{
+	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
+	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
+	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
+	}
+	\label{fig:variance_decomposition}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/gaussians/log_n0=1}.pdf}
+	\caption{
+	\textbf{Full gaussian $\cN(\mu, \sigma^2)$ example. Left:} training curves and analytic upper bound. 
+	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
+	\textbf{Right:} bias-variance decomposition, using the generalized mean.
+	}
+	\label{fig:gaussian_decomposition}
+\end{figure}
+
+\subsection{About Variance and Covariance}
+
+\begin{lemma}
+Taking  $\MAPt$ and $\MAPm$ as the primal and dual parameters of the MAP estimate, and $\tilde \mu_n$ the primal expectation and $\mu_n$ the dual expectation, the following identities hold
+	\begin{align}
+	\Cov(\hat \natp_n ; \MAPm )  
+	&=\expect{\bregmanconj(\MAPm ; \mu_n)
+	+ \bregmanconj(\mu_n ; \MAPm)} \\
+	&=\expect{\bregmanconj(\MAPm ; \tilde \mu_n)
+	+ \bregmanconj(\tilde \mu_n ; \MAPm)} \; .
+\end{align}
+\end{lemma}
+\begin{proof}
+By the three point identity (or plain derivation), we have 
+\begin{align}
+	\bregmanconj(\MAPm ; \mu_n)
+	+ \bregmanconj(\mu_n ; \MAPm)
+	= \langle \nabla \conj (\hat \mu_n) - \nabla \conj(\mu_n) ;  \hat \mu_n - \mu_n \rangle 
+	= \langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle 
+\end{align}
+Taking the expectation,  the constant $\theta_n$ disappears since $\expect{\hat \mu_n - \mu_n}=0$.
+\begin{align}
+	\expect{\langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle } 
+	 = \expect{\langle \MAPt; \MAPm \rangle} - \langle \expect{\MAPt} ; \MAPm \rangle
+	= \Cov(\hat \natp_n ; \hat \mu_n ) 
+\end{align}
+which is the first identity. We get the second identity by symmetry between the primal and the dual.
+\end{proof}
+
+Note that, as a corollary, the covariance between primal and dual MAP
+$\Cov(\hat \natp_n ; \hat \mu_n )$ is positive, and it can be seen as a non-linear variance of $\hat \mu_n$ -- e.g.
+we can define a variance operator with a non-linear monotone operator $\nabla\conj$ plugged in the middle $\Var_{\nabla \conj}(\hat \mu_n) := \Cov(\nabla\conj(\hat \mu_n) ; \hat \mu_n ) \geq 0$.
+
+\paragraph{Expansion}
+We can expand  $\hat \mu_n - \mu_n$ to get another formulation of this covariance
+\begin{align}
+	\expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle}
+	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
+	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
+	= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
+\end{align}
+where we used the property that all data points play a symmetric role to remove the sum over $i$.
+
+\clearpage
+
+\section{Optimization Perspective}
+\paragraph{Problem.}
+Our goal is to solve the following problem
+\begin{align}
+	\min_\natp \E[f(\natp ; X)] := -\langle \E[T(X)] ; \natp \rangle + \logpart(\natp)
+\end{align}
+where $\logpart$ is a convex function and we only have access to stochastic oracles $X \sim p(X | \natp^*)$, meaning we can compute stochastic values and gradients 
+\begin{align}
+	f(\natp ; X) &= -\langle T(X) ; \natp \rangle + \logpart(\natp) \\
+	\nabla f(\natp ; X) &= \nabla \logpart(\natp) - T(X)
+\end{align}
+such that $\E[T(X)] = \mu^*$.
+
+\paragraph{Algorithms.}
+We observe $n$ data points $X_i$. 
+We are studying two algorithms to solve this problem : the MLE and the MAP
+\begin{align}
+	\MAPm^\text{MLE} &= \inv{n}\sum_{i=1}^n T(X_i) 
+	&\MAPt^\text{MLE} &= \nabla\logpart^{-1}(\MAPm^\text{MLE}) \\
+	\MAPm^\text{MAP} &= \frac{n_0\mu_0 + \smallsum_{i=1}^n T(X_i)}{n_0 + n}
+	&\MAPt^\text{MAP} &= \nabla\logpart^{-1}(\MAPm^\text{MAP})	
+\end{align}
+
+\paragraph{Interpretation.}
+As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms. 
+\begin{itemize}
+	\item stochastic bregman proximal point with divergence $\bregman$, initialization $\natp_0$, and learning rate $\inv{n_0+n}$,
+	\item stochastic mirror descent with divergence $\bregman$, initialization $\natp_0$, and learning rate  $\inv{n_0+n +1}$.
+\end{itemize}
+In both cases, $f(\natp)$ is convex and 1-smooth / 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.
+
+\paragraph{Goal.} We want to bound the suboptimality (in function value) as a function of the number of samples.
+
+\paragraph{Other assumptions.}
+An assumption pointed out by Nicolas Loizou is 
+\begin{align}
+	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
+\end{align}
+for which a sufficient condition is 
+\begin{align}
+\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
+\end{align}
+which holds in a lot of settings but not all.
+For instance, a gaussian $\cN(\mu,\sigma^2)$ can overfit on a single data point and give it infinite density -- eg converge to a Dirac.
+However we can overcome this issue by grouping data points : a gaussian can not overfit on two points.
+Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$.
+So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)
+
+\section{Stochastic Bregman Proximal Point}
+Remark that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
+This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
+This can also be seen at each step since
+\begin{align}
+    \hat \natp_{n+1} 
+    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
+    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
+\end{align}
+Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
+
+This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
+
+The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
+First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
+One needs an assumption on the quality of the stochastic estimates $f(.
+; x)$.
+Given the proof, the most straightforward such assumption (see paper notes) is 
+\begin{align}
+\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
+\end{align}
+which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
+This is exactly the same assumption as given by the SMD analysis.
+However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
+It does not immediately apply to the last iterate.
+
+\section{Stochastic Mirror Descent (SMD)}
+\subsection{MAP as SMD}
+Let $f(\natp) := \expect[x]{-\log p(x|\natp)} = - \langle \mu_*, \natp\rangle + \logpart(\natp)$.
+In words, $f$ is linear modification of a convex function $\logpart$, which we can access only through noisy estimates of $\mu_*$.
+It turns out that the MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpart$.
+First let's recall mirror descent iteration
+\begin{align}
+	\natp_{t+1} 
+	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
+	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
+\end{align}
+where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$.
+Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$.
+Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
+\begin{align}
+	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
+	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
+\end{align}
+which shows exponential convergence and 1-step convergence when $\lr =1$.
+Back to our sheep, the MAP iteration can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1})$ a stochastic estimate  of $\nabla f(\natp_t)$
+\begin{align}
+    \hat \natp_{n+1} 
+    &= \argmin_\natp 
+    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
+    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
+    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
+    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
+\end{align}
+where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$.
+This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
+
+\subsection{Relative Smoothness for Mirror Descent}
+In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpart$ \citep{bubeck2015convex}.
+In our setting this is not always true -- eg gaussians.
+However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity.
+\begin{align}
+	& \text{$f$ is $L$-smooth relative to $h$} \\
+	\iff & Lh - f \ \text{convex} \\
+	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y;x), \forall x,y \\
+	\iff & \cB_f(y;x) \leq L \cB_h(y;x), \forall x,y \\
+	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
+\end{align} 
+where the last equivalence holds only when $f$ and $h$ are twice differentiable.
+In words, $f$ is upper bounded by it linear approximation plus the $h$-Bregman divergence, which can also be seen as a bound between divergences, or more locally as a bound between Hessians.
+Similarly, $f$ is $\mu$ strongly-convex relative to $h$ if
+\begin{align}
+	& f-\mu h \ \text{ convex} \\
+	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y;x) \leq f(y)\forall x,y \\
+	\iff & \mu\cB_h(y;x) \leq \cB_f(y;x) \forall x,y \\
+	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
+\end{align}
+
+Another way to view this elegant generalization of smoothness and strong-convexity is as a transfer of the Loewner partial order on symmetric matrices to functions, via the Hessian. 
+As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
+As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
+More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality of the sequence 
+\begin{align}
+	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x ; x_t)
+\end{align}
+is upper bounded by the simple formula
+\begin{align}
+	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* ; x_0) }{t} \; .
+\end{align}
+These notions were rediscovered and expanded by \citet{bauschke2017descent} and \citet{lu2018relatively}.
+If you need to read one, pick \citet{lu2018relatively} -- I found it much much easier and more enjoyable to read.
+This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.
+
+\subsection{Relative smoothness for SMD}
+Now our setting is Stochastic Mirror Descent (SMD), meaning at each step we observe a random unbiased estimate gradient.
+This setting was studied by \citet{hanzely2018fastest}, who proved in the smooth strongly-convex case with tail averaging :  with constant step-size, linear convergence down to a variance ball, and with step-size $\gamma_t = n_0 + t$ a rate $\tilde O(\inv{t})$.
+These results match the rates for standard SGD.
+
+This is very interesting to us.
+Let us have a look at their variance assumption \citep[Assumption 5.1]{hanzely2018fastest}.
+Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
+Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps
+\begin{align}
+	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t}) \leq \lr_t \sigma^2 \\
+	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
+	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
+\end{align}
+where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD.
+
+\subsection{Transferring SMD's covariance assumption to MAP estimates}
+In our setting, 
+\begin{align}
+	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
+\end{align}
+so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
+\begin{align}
+	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
+	\leq \lr_t\sigma^2 \\
+	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
+\end{align}
+In other words, we need for all $\lr\in(0,\inv{n})$, and a certain subset of $\mu$,
+\begin{align}
+\label{eq:hypvariance}
+	\boxed{
+	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr) \mu)\bigg	) 
+	\leq \lr \sigma^2 \; .
+	}
+\end{align}
+
+\begin{example}
+Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
+The covariance  \eqref{eq:hypvariance} we want to bound is equal to
+\begin{align}
+	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
+	\approx \lr (1-\lr)^{-2} \left (\frac{\sigma_*}{\sigma_0} \right)^4
+\end{align}
+using a first order approximation of the inverse, eg a first order approximation of $\nabla \conj$ that holds when $\lr X^2 \ll (1-\lr) \sigma_0^2$.
+This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.
+This is a problem, because we need this bound to hold  for all values of $\sigma_0$ in the support of $X^2$, meaning on the whole $[0,+\infty[$, which is impossible because $1/X^2$ diverges, so we need to process this integral differently.
+\end{example}
+
+\paragraph{MLE Analysis.}
+When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, we get
+\begin{multline}
+	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) }
+	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
+	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}
+	 	\; \text{with}\; \mu=\inv{n-1}\smallsum_{j\neq i} T(x_j)  } ]  \leq \frac{\sigma^2}{n}
+\label{eq:MLE_bound}
+\end{multline}
+where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. 
+In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality.
+
+This is too simple to be true, as shown on our trailing example.
+In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
+The mean parameter $\mu$ take values arbitrarily close from $0$, which makes the covariance explode.
+\RLP{should we get a functional bound $\sigma^2(\mu)$ instead ? or split between $\mu$ smaller or larger than $\varepsilon$.}
+
+\paragraph{MAP Analysis.}
+We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
+Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
+\begin{align}
+	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) } 
+	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
+	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
+\end{align}
+so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
+\begin{align}
+	 \bregmanconj (\mu_* ; \tilde \mu)  
+	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
+\end{align}
+where $\mSigma_* = \Cov_{\mu_*}(T(X))$. 
+The order of terms $\bregmanconj (\mu_* ; \tilde \mu)  \neq \bregmanconj (\tilde \mu ; \mu_* ) $ only affects third order terms.
+That's for the bregman, but the scalar product is harder to bound.
+Hopefully it could be positive.
+If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$.
+Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.
+
+\paragraph{Relative Continuity}
+\citet{lu2019relative} define relative continuity to tackle non-differentiable functions -- eg Hinge loss.
+These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
+Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and $\ell^2$ distance.
+Although this hypothesis is far from intuitive, it does yield convergence rates very similar to typical SGD analysis.
+In this work we are first concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
+
+\paragraph{Bound on the trajectory.}
+Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
+No we cannot because $\mu$ takes every possible values as a random variable within the expectation.
+
+\section{Expected Relative Smoothness}
+To avoid gradient variance assumption, we can take inspiration from expected smoothness, which is a consequence of convexity and smoothness, and which yields a general analysis of SGD.
+Can we get something similar with relativity and SMD?
+The analog of gradient norm is the Bregman between iterates, similar to the Homeomorphic EM rate.
+
+% todo: idee: adapter la preuve de stochastic gradient descent a stochastic mirror descent.
+Exploiter un parralele entre ces deux preuves pour obtenir une preuve pour stochastic mirror descent dans le cas ou f est relatively smooth w.r.t.
+A*.
+ce qu'on veut, c'est une hypothese analogue a expected smoothness (see paper of nico loizou) pour le relatively smooth.
+% SGD: General Analysis and Improved Rates gower et al -- section 2 and 3 see equation (9) in particular (expected smoothnesS)
+% Relatively-Smooth Convex Optimization by First-Order Methods, and Applications,  -- Section 3
+
+\subsection{Three points lemma}
+\label{app:3points}
+
+Proofs of convergence for mirror descent with relative smoothness rely on a specific property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
+\begin{align}
+	x_+ \in \argmin	_{x\in C} \overbrace{f(x) + \cB_h(x ; y)}^{\phi(x)}
+\end{align}
+where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector.
+Then
+\begin{align}
+	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
+\end{align}
+This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).
+
+\begin{proof}
+	The first order optimality condition is
+	\begin{align}
+		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
+		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
+	\end{align}
+	This proof relies on another property called three point property 
+	\begin{align}
+		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
+	\end{align}
+	which can be proved by expanding the right hand side.
+	By convexity of $f$ we also have 
+	\begin{align}
+		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
+	\end{align} 
+	Putting it all together we get
+	\begin{align}
+		0 
+		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
+		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
+	\end{align}
+	which concludes the proof.
+\end{proof}
+
+\subsection{Convergence of Mirror Descent}
+To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$.
+Note that the strong convexity constant $m$ may be zero.
+Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
+\begin{align}
+	f(\natp^{t+1}) 
+	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
+	& \text{(Relative $L$-smoothness)} \\
+	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
+	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \text{(Relative $m$-strong convexity)}
+\end{align}
+Applying the 3 points lemma with \TODO{f}
+
+\section{Self-Concordance}
+\subsection{Motivation}
+A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
+\begin{align}
+    \logpart(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
+\end{align}
+which is defined on $\eta \in \real^d$ and $\Lambda\in \real^{d\times d}$ symmetric positive definite. 
+It is not strongly convex, so $\conj$ is not smooth.
+
+Another hypothesis that may be more suitable is self-concordance. $f: \real \rightarrow \real$ is SC if 
+\begin{align}
+    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
+\end{align}
+The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
+A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
+Negative logarithm $-\log(x)$ and entropy $x\log(x)$ are both self-concordant function. This is good news for us since log-partition function may include SC logarithmic barriers.
+In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
+
+\subsection{Suboptimality and Newton Decrement}
+An important property of self-concordant functions (cite Boyd's book, although Nesterov's may be better) is that their suboptimality  may be upper bounded by the Newton Decrement 
+\begin{align}
+    \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
+\end{align}
+In general, subtracting the minimum $f^*$ of $f$ , we have
+\begin{align}
+    f(x) - f* \leq - \decrement(x) - \log( 1- \decrement(x)) \; .
+\end{align}
+Note that this bound is vacuous for $\decrement(x)\geq 1$. For $y = \decrement(x) \leq 0.68$, we have $ - y - \log(1-y) \leq y^2$, so we get the bound
+\begin{align}
+    f(x) - f^* \leq \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
+\end{align}
+
+Our functions of interest is $f(\natp)= \bregman (\natp ; \natp^*)= \bregmanconj ( \meanp^* ; \meanp) = g(\meanp)$, with minimum $f^*=0$. 
+If $\logpart$ is self-concordant, then so is $f$, but not necessarily $g$.
+The gradient and Hessian of $f$ are
+\begin{align}
+    f(\natp) 
+    &= \logpart(\natp) - \logpart(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
+    \nabla f(\natp) 
+    &= \meanp - \meanp^* = \expect[p(X|\natp)]{T(X)} - \expect[p(X|\natp^*)]{T(X)} \\
+    \nabla^2 f(\natp) 
+    &= \mSigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
+\end{align}
+so that we get the bound.
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    \leq \decrement(\natp)^2 
+    = \norm{\mu^* - \mu(\theta)}^2_{\mSigma(\theta)^{-1}} 
+    \leq 0.46
+\end{align}
+% If we were to use $g(\meanp)$ instead, we would get dirty third order derivatives
+% \begin{align}
+%     g(\meanp) 
+%     &= \conj(\meanp^*) - \conj(\meanp) - \langle \natp, \meanp^* - \meanp \rangle \\
+%     \nabla g(\meanp) 
+%     &= -\natp - \nabla^2 \conj(\meanp) (\meanp^* - \meanp) + \natp \\
+%     &= \nabla^2 \conj(\meanp) (\meanp - \meanp^*) \\
+%     \nabla^2 g(\meanp) 
+%     &= \nabla^2 \conj(\meanp) + \nabla^3 \conj(\meanp ) (\meanp - \meanp^*)
+% \end{align}
+% where the last expression involves a third order tensor.
+
+Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$, then we would get
+\begin{align}
+    \nabla h(\meanp ) 
+    &= \natp - \natp^* \\
+    \nabla^2 h(\meanp ) 
+    &= \nabla^2 \conj(\meanp) 
+    = \nabla^2 \logpart(\natp) ^{-1}
+    = \Cov_{p(X|\natp)}[T(X)]^{-1}\\
+    \implies 
+    \decrement(\meanp)
+    &= \Var_{p(X|\natp)}[(\natp - \natp^*)^T T(X)] \; .
+\end{align}
+This is just a remark. I don't think it can help us to get anywhere.
+
+\section{Related Work}
+Exponential families are a mainstream tool in machine learning.
+They are used  to generalize linear regression \citep{mccullagh1989generalized}, PCA \citep{collins2001generalization} or k-means \citep{banerjee2005clustering} to diverse data types and distributions.
+\citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure.
+
+\citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem.
+They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
+Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors yet critical as it does not hold for multivariate normals with unknown covariance.
+\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.
+Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
+
+\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
+Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
+
+\citet{pfau2013generalized} gives a bias variance decomposition for Bregman divergences. However it does not give interesting conclusions for our problem of interest.
+
+\citet{ostrovskii2021finite}  shows a convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant.
+This convergence rate holds only after a given number of samples have been seen.
+This is because they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.
+
+\subsection{Alternative Research Tracks}
+\begin{itemize}
+	\item Weakly supervised : missing part of $x_i$ for each $i$.
+	\item Structured prediction : graphical model on $X$ or other kind of structure.
+\end{itemize}
+
+\bibliographystyle{apalike}
+\bibliography{references.bib}
+
+\newpage
+\appendix
+
+\section{Cramer-Rao Lower Bound}
+In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
+In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
+\begin{align}
+    \Cov(\hat \natp_n)  
+    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
+    \iff
+    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
+    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
+\end{align}
+This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance.
+In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"
+
+Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
+For the estimator $\hat \natp_n$, this lower bound is not trivial.
+
+\section{Posterior Expected Loss}
+
+A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is 
+\begin{align}
+    \pi(\natp) 
+    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
+    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle)
+\end{align}
+so we see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
+When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
+This is well detailed in \citet{agarwal2010geometric}.
+
+Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
+\begin{align}
+    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+\end{align}
+This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral.
+We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
+This is exactly twice the expected loss of the MLE.
+
+\begin{example}
+	We can also get a closed form for the posterior of a centered gaussian variance 
+	\begin{align}
+		\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
+	\end{align}
+	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma.
+The using results from the MLE, we get the exact formula
+	\begin{align}
+		\expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
+	\end{align}
+	and we bound it to 
+	\begin{align}
+		\inv{2n} + \frac{4}{n(n-2)}
+		\leq \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		\leq 
+		\inv{n} + \frac{4}{n(n-2)}
+	\end{align}
+	which is about twice the MLE bounds~\eqref{eq:MLE_rate}, exactly like as we observed for the gaussian mean.
+\end{example}
+
+\section{Gaussian Variance}
+\subsection{Proof of MLE Tight Bound}
 	\begin{theorem}[MLE tight upper bound]
@@
 
-\subsection{MLE Loose Upper Bound}
+\subsection{Proof of MLE Loose Upper Bound}
 If we apply \eqref{eq:log_bound} to the MLE, we get the upper bound
@@
 
-\subsection{Expected MAP Natural Parameter}
+\subsection{Proof of Expected MAP Natural Parameter}
 \begin{lemma}[Expected MAP natural parameter]
@@
 		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n + n_0}}  
-		\label{eq:theta_expectation_bound}
 	\end{align}
@@
 
-\subsection{MAP Upper Bound}
+\subsection{Proof of MAP Bound}
 \begin{theorem}[MAP Upper Bound]
@@
 \begin{proof}
-		Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus solely on the simpler upper bound given Now using the bound \eqref{eq:log_bound} on the logarithm, we get
+		Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound}. We get
 \begin{align}
@@
 
-\tableofcontents
-
-\section{Comparison with  Euclidean Distance}
-Often we can relate this Bregman suboptimality with the $\ell^2$ distance in mean parameter space.
-
-\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
-    \leq 2L \norm{\mu^* - \mu}
-\end{align}
-so $\bregmanconj$ is $2L$-Lipschitz.
-Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this bound applies to the log-likelihood.
-
-\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
-\end{align}
-so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.
-
-\paragraph{$\ell^2$-norm Analysis}
-Let us make these statements more precise. A bound on the variance becomes a bound on the variance of the average
-\begin{align}
-	\Var T(X) = \E \norm{T(X) - \mu^*}^2 = \sigma^2 \\
-	\implies \E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
-\end{align}
-eg the variance of the mean is $n$ times smaller than the variance of the samples.
-Adding a reference mean $\mu_0$ to get the MAP yields
-\begin{align}
-	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
-	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
-	&= O\left(\frac{\sigma^2}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
-\end{align}
-so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !
-
-\section{Asymptotic Behaviour}
-In the limit, the MAP estimates reach a rate $O(1/2n)$. This is shown by approximating the Bregman divergence using a second order Taylor expansion
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
-    + O(\norm{\mu - \mu^*}^3)\\
-    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
-    + O(\norm{\mu - \mu^*}^3)
-\end{align}
-where we introduced the Mahalanobis distance induced by the matrix 
-\begin{align}
-    \nabla^2\conj(\mu^*) 
-    = \nabla^2\logpart(\natp^*)^{-1} 
-    = \Cov_{\natp^*}[T(X)]^{-1} 
-    =: \mSigma^{-1}  \; .
-\end{align}
-To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
-\begin{align}
-	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
-\end{align}
-where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
-\begin{align}
-	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
-\end{align}
-and for the MAP we  get 
-\begin{align}
-	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
-	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
-	&= \frac{d}{n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
-\end{align}
-or 
-\begin{align}
-\label{eq:MAP_asymptote}
-	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
-	= \frac{d}{2n} + O(n^{- \frac{3}{2}})
-\end{align}
-Remark how the variance does not even appear, only the dimension divided by $n$ really matters.
-I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
-Indeed this is because it is hidden within the Bregman divergence loss. 
-Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?
-
-\section{Bias-Variance Decomposition}
-\paragraph{Covariance.}
-Using Jensen inequality and inserting appropriate quantities, we obtain the following analog of the bias-variance decomposition:
-\newcommand{\bias}[1]{\textcolor{Red}{\underbrace{\textcolor{black}{#1}}_{\text{bias}}}}
-\newcommand{\variance}[1]{\textcolor{Green}{\underbrace{\textcolor{black}{#1}}_{\text{variance}}}}
-\begin{align}
-	& \expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	= \conj(\mu^* ) 
-	- \underbrace{\expect{\conj(\hat \mu_n)}}_{\leq \conj(\mu_n) \text{ (Jensen)} } 
-	\pm \langle \underbrace{\nabla \conj( \mu_n)}_{=:\natp_n} ; \mu^* - \mu_n \rangle
-	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \nonumber \\
-	& \leq \conj(\mu^* ) 
-	- \conj(\mu_n) 
-	- \langle  \nabla \conj( \mu_n) ; \mu^* - \mu_n \rangle
-	+ \langle \E[\hat \natp_n] - \natp_n ;  \mu_n - \mu^* \rangle
-	+ \E[\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle ]
-	\nonumber \\
-	& \leq \bias{\bregmanconj(\mu^*; \mu_n)}
-	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
-	+ \variance{\Cov(\hat \natp_n ; \hat \mu_n ) }
-\end{align}
-where $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ is the MAP estimate of the mean parameter, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ is the corresponding natural parameter, 
-$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ is the expected MAP estimate,
-and $\theta_n = \nabla \conj(\mu_n)$ is the natural parameter of this expectation.
-Note that unless $\conj$ is $\ell^2$, or we are in a degenerate case with $\hat \mu_n$ a Dirac, $\theta_n \neq \expect{\hat \theta_n}$.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.8\textwidth]{figs/bias-variance.png}
-	\caption{A schematic illustration of the different characters featured in the bias-variance decomposition.}
-	\label{fig:bias-variance}
-\end{figure}
-
-\begin{figure}[p]
-	\centering
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=1.pdf}
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=2.pdf}
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=10.pdf}
-	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
-	\label{fig:bias-variance-numerical}
-\end{figure}
-
-\paragraph{No Jensen.}
-In fact this previous decomposition is highly suboptimal. 
-We do not need to use Jensen inequality.
-\begin{align}
-	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	&= \conj(\mu^* ) 
-	\pm \conj(\mu_n)
-	-  \expect{\conj(\hat \mu_n)}
-	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] \\
-	&= \conj(\mu^* ) 
-	- \conj(\mu_n)  + \E[\bregmanconj(\mu_n, \MAPm)]
-	+ \langle \E[\hat \natp_n] \pm \nabla \conj(\mu_n) ;  \mu_n - \mu^* \rangle
-	\nonumber\\
-	& = \bias{\bregmanconj(\mu^*; \mu_n)}
-	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
-	+ \variance{\E[\bregmanconj(\mu_n, \MAPm)]}
-\end{align}
-With Jensen, we were losing a Bregman term, as per the 3 point property, the covariance is equal to the symmetrized Bregman.
-\begin{align}
-	\Cov(\hat \natp_n ; \hat \mu_n )  
-	=\expect{\bregmanconj(\MAPm ; \mu_n)
-	+ \bregmanconj(\mu_n ; \MAPm)}
-\end{align}
-
-\paragraph{No Mixed Term.}
-In fact, it is possible to have no mixed term using the central point $\tilde \mu_n = \nabla \logpart(\E[\MAPt])$ instead of $\mu_n = \E[\MAPm]$,  exactly as described by \citet[Theorem 0.1]{pfau2013generalized}.
-\begin{align}
-	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	&= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
-	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
-	\expect{\bregman(\MAPt; \natp^*)} 
-	&= \bias{\bregman(\E[\MAPt]; \natp^*)}
-	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}} 
-\end{align}
-Now the big question, is : what kind of results do I get with such quantities ?
-In general, it seems that $\tilde \mu_n$ is a very central and interesting quantity to consider.
-We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=\textwidth]{{figs/biasvariance/linear_ratio=0.8_n0=10}.pdf}
-	\caption{
-	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
-	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
-	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
-	}
-	\label{fig:variance_decomposition}
-\end{figure}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=\textwidth]{{figs/gaussians/log_n0=1}.pdf}
-	\caption{
-	\textbf{Full gaussian $\cN(\mu, \sigma^2)$ example. Left:} training curves and analytic upper bound. 
-	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
-	\textbf{Right:} bias-variance decomposition, using the generalized mean.
-	}
-	\label{fig:gaussian_decomposition}
-\end{figure}
-
-\subsection{Pure Bias Term}
+\section{More on the Bias-Variance Decomposition}
+
+\subsection{Bias Term}
 The pure bias term measures the divergence between the optimum and the expected MAP estimate $\bregmanconj(\mu^*; \mu_n)$.
 
-\paragraph{Aymptote.}
+\paragraph{Asymptote.}
 Using the quadratic approximation of the Bregman divergence near $\mu^*$, we get a $O(n^{-2})$ rate for the pure bias term
@@
 
-\subsection{Pure Variance Term}
-The pure variance term $\Cov(\hat \natp_n ; \hat \mu_n )$   can be seen as a non-linear variance of $\hat \mu_n$.
-Indeed, it is positive thanks to the monotonicity of $\nabla \conj$
-\begin{align}
-&\langle \hat \natp_n - \natp_n ; \hat \mu_n - \mu_n \rangle 
-	= \langle \nabla \conj (\hat \mu_n) - \nabla \conj(\mu_n) ;  \hat \mu_n - \mu_n \rangle 
-	\geq 0
-\end{align}
-Taking the expectation, (which allows removing the constant $\theta_n$) 
-\begin{align}
-	\expect{\langle \hat \natp_n - \natp_n ; \hat \mu_n - \mu_n \rangle} 
-	= \Cov(\hat \natp_n ; \hat \mu_n ) \geq 0
-\end{align}
-Consequently, we will now write about this covariance as a variance, with the monotone operator $\nabla \conj$ (or equivalently $\nabla \logpart$) plugged in the middle
-\begin{align}
-	\Var_{\nabla \conj}(\hat \mu_n) := \Cov(\nabla\conj(\hat \mu_n) ; \hat \mu_n ) \geq 0 \; .
-\end{align}
-
-\paragraph{Expansion}
-We can expand  $\hat \mu_n - \mu_n$ to get another formulation of this variance
-\begin{align}
-	\expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle}
-	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
-	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
-	= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
-\end{align}
-where we used the property that all data points play a symmetric role to remove the sum over $i$.
+\subsection{Variance Term}
 
 \paragraph{Asymptote.}
-Using a linear approximation of the mirror map, we get a $O(n^{-1})$ rate for the variance term
+Using a linear approximation of the mirror map, we get a $O(n^{-1})$ rate for the covariance
 \begin{align}
@@
 
-\section{Optimization Perspective}
-\paragraph{Problem.}
-Our goal is to solve the following problem
-\begin{align}
-	\min_\natp \E[f(\natp ; X)] := -\langle \E[T(X)] ; \natp \rangle + \logpart(\natp)
-\end{align}
-where $\logpart$ is a convex function and we only have access to stochastic oracles $X \sim p(X | \natp^*)$, meaning we can compute stochastic values and gradients 
-\begin{align}
-	f(\natp ; X) &= -\langle T(X) ; \natp \rangle + \logpart(\natp) \\
-	\nabla f(\natp ; X) &= \nabla \logpart(\natp) - T(X)
-\end{align}
-such that $\E[T(X)] = \mu^*$.
-
-\paragraph{Algorithms.}
-We observe $n$ data points $X_i$. 
-We are studying two algorithms to solve this problem : the MLE and the MAP
-\begin{align}
-	\MAPm^\text{MLE} &= \inv{n}\sum_{i=1}^n T(X_i) 
-	&\MAPt^\text{MLE} &= \nabla\logpart^{-1}(\MAPm^\text{MLE}) \\
-	\MAPm^\text{MAP} &= \frac{n_0\mu_0 + \smallsum_{i=1}^n T(X_i)}{n_0 + n}
-	&\MAPt^\text{MAP} &= \nabla\logpart^{-1}(\MAPm^\text{MAP})	
-\end{align}
-
-\paragraph{Interpretation.}
-As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms. 
-\begin{itemize}
-	\item stochastic bregman proximal point with divergence $\bregman$, initialization $\natp_0$, and learning rate $\inv{n_0+n}$,
-	\item stochastic mirror descent with divergence $\bregman$, initialization $\natp_0$, and learning rate  $\inv{n_0+n +1}$.
-\end{itemize}
-In both cases, $f(\natp)$ is convex and 1-smooth / 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.
-
-\paragraph{Other assumptions.}
-An assumption pointed out by Nicolas Loizou is 
-\begin{align}
-	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
-\end{align}
-for which a sufficient condition is 
-\begin{align}
-\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
-\end{align}
-which holds in a lot of settings but not all.
-For instance, a gaussian $\cN(\mu,\sigma^2)$ can overfit on a single data point and give it infinite density -- eg converge to a Dirac.
-However we can overcome this issue by grouping data points : a gaussian can not overfit on two points.
-Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$.
-So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)
-
-\section{Stochastic Bregman Proximal Point}
-Remark that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
-This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
-This can also be seen at each step since
-\begin{align}
-    \hat \natp_{n+1} 
-    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
-    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
-\end{align}
-Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
-
-This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
-
-The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
-First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
-One needs an assumption on the quality of the stochastic estimates $f(.
-; x)$.
-Given the proof, the most straightforward such assumption (see paper notes) is 
-\begin{align}
-\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
-\end{align}
-which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
-This is exactly the same assumption as given by the SMD analysis.
-However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
-It does not immediately apply to the last iterate.
-
-\section{Stochastic Mirror Descent (SMD)}
-\subsection{MAP as SMD}
-Let $f(\natp) := \expect[x]{-\log p(x|\natp)} = - \langle \mu_*, \natp\rangle + \logpart(\natp)$.
-In words, $f$ is linear modification of a convex function $\logpart$, which we can access only through noisy estimates of $\mu_*$.
-It turns out that the MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpart$.
-First let's recall mirror descent iteration
-\begin{align}
-	\natp_{t+1} 
-	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
-	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
-\end{align}
-where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$.
-Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$.
-Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
-\begin{align}
-	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
-	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
-\end{align}
-which shows exponential convergence and 1-step convergence when $\lr =1$.
-Back to our sheep, the MAP iteration can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1})$ a stochastic estimate  of $\nabla f(\natp_t)$
-\begin{align}
-    \hat \natp_{n+1} 
-    &= \argmin_\natp 
-    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
-    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
-    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
-    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
-\end{align}
-where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$.
-This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
-
-\subsection{Relative Smoothness for Mirror Descent}
-In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpart$ \citep{bubeck2015convex}.
-In our setting this is not always true -- eg gaussians.
-However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity.
-\begin{align}
-	& \text{$f$ is $L$-smooth relative to $h$} \\
-	\iff & Lh - f \ \text{convex} \\
-	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y;x), \forall x,y \\
-	\iff & \cB_f(y;x) \leq L \cB_h(y;x), \forall x,y \\
-	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
-\end{align} 
-where the last equivalence holds only when $f$ and $h$ are twice differentiable.
-In words, $f$ is upper bounded by it linear approximation plus the $h$-Bregman divergence, which can also be seen as a bound between divergences, or more locally as a bound between Hessians.
-Similarly, $f$ is $\mu$ strongly-convex relative to $h$ if
-\begin{align}
-	& f-\mu h \ \text{ convex} \\
-	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y;x) \leq f(y)\forall x,y \\
-	\iff & \mu\cB_h(y;x) \leq \cB_f(y;x) \forall x,y \\
-	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
-\end{align}
-
-Another way to view this elegant generalization of smoothness and strong-convexity is as a transfer of the Loewner partial order on symmetric matrices to functions, via the Hessian. 
-As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
-As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
-More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality of the sequence 
-\begin{align}
-	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x ; x_t)
-\end{align}
-is upper bounded by the simple formula
-\begin{align}
-	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* ; x_0) }{t} \; .
-\end{align}
-These notions were rediscovered and expanded by \citet{bauschke2017descent} and \citet{lu2018relatively}.
-If you need to read one, pick \citet{lu2018relatively} -- I found it much much easier and more enjoyable to read.
-This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.
-
-\subsection{Relative smoothness for SMD}
-Now our setting is Stochastic Mirror Descent (SMD), meaning at each step we observe a random unbiased estimate gradient.
-This setting was studied by \citet{hanzely2018fastest}, who proved in the smooth strongly-convex case with tail averaging :  with constant step-size, linear convergence down to a variance ball, and with step-size $\gamma_t = n_0 + t$ a rate $\tilde O(\inv{t})$.
-These results match the rates for standard SGD.
-
-This is very interesting to us.
-Let us have a look at their variance assumption \citep[Assumption 5.1]{hanzely2018fastest}.
-Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
-Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps
-\begin{align}
-	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t}) \leq \lr_t \sigma^2 \\
-	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
-	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
-\end{align}
-where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD.
-
-\subsection{Transferring SMD's covariance assumption to MAP estimates}
-In our setting, 
-\begin{align}
-	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
-\end{align}
-so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
-\begin{align}
-	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
-	\leq \lr_t\sigma^2 \\
-	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
-\end{align}
-In other words, we need for all $\lr\in(0,\inv{n})$, and a certain subset of $\mu$,
-\begin{align}
-\label{eq:hypvariance}
-	\boxed{
-	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr) \mu)\bigg	) 
-	\leq \lr \sigma^2 \; .
-	}
-\end{align}
-
-\begin{example}
-Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
-The covariance  \eqref{eq:hypvariance} we want to bound is equal to
-\begin{align}
-	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
-	\approx \lr (1-\lr)^{-2} \left (\frac{\sigma_*}{\sigma_0} \right)^4
-\end{align}
-using a first order approximation of the inverse, eg a first order approximation of $\nabla \conj$ that holds when $\lr X^2 \ll (1-\lr) \sigma_0^2$.
-This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.
-This is a problem, because we need this bound to hold  for all values of $\sigma_0$ in the support of $X^2$, meaning on the whole $[0,+\infty[$, which is impossible because $1/X^2$ diverges, so we need to process this integral differently.
-\end{example}
-
-\paragraph{MLE Analysis.}
-When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, we get
-\begin{multline}
-	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) }
-	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
-	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}
-	 	\; \text{with}\; \mu=\inv{n-1}\smallsum_{j\neq i} T(x_j)  } ]  \leq \frac{\sigma^2}{n}
-\label{eq:MLE_bound}
-\end{multline}
-where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. 
-In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality.
-
-This is too simple to be true, as shown on our trailing example.
-In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
-The mean parameter $\mu$ take values arbitrarily close from $0$, which makes the covariance explode.
-\RLP{should we get a functional bound $\sigma^2(\mu)$ instead ? or split between $\mu$ smaller or larger than $\varepsilon$.}
-
-\paragraph{MAP Analysis.}
-We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
-Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
-\begin{align}
-	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) } 
-	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
-	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
-\end{align}
-so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
-\begin{align}
-	 \bregmanconj (\mu_* ; \tilde \mu)  
-	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
-\end{align}
-where $\mSigma_* = \Cov_{\mu_*}(T(X))$. 
-The order of terms $\bregmanconj (\mu_* ; \tilde \mu)  \neq \bregmanconj (\tilde \mu ; \mu_* ) $ only affects third order terms.
-That's for the bregman, but the scalar product is harder to bound.
-Hopefully it could be positive.
-If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$.
-Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.
-
-\paragraph{Relative Continuity}
-\citet{lu2019relative} define relative continuity to tackle non-differentiable functions -- eg Hinge loss.
-These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
-Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and $\ell^2$ distance.
-Although this hypothesis is far from intuitive, it does yield convergence rates very similar to typical SGD analysis.
-In this work we are first concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
-
-\paragraph{Bound on the trajectory.}
-Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
-No we cannot because $\mu$ takes every possible values as a random variable within the expectation.
-
-\section{Expected Relative Smoothness}
-To avoid gradient variance assumption, we can take inspiration from expected smoothness, which is a consequence of convexity and smoothness, and which yields a general analysis of SGD.
-Can we get something similar with relativity and SMD?
-The analog of gradient norm is the Bregman between iterates, similar to the Homeomorphic EM rate.
-
-% todo: idee: adapter la preuve de stochastic gradient descent a stochastic mirror descent.
-Exploiter un parralele entre ces deux preuves pour obtenir une preuve pour stochastic mirror descent dans le cas ou f est relatively smooth w.r.t.
-A*.
-ce qu'on veut, c'est une hypothese analogue a expected smoothness (see paper of nico loizou) pour le relatively smooth.
-% SGD: General Analysis and Improved Rates gower et al -- section 2 and 3 see equation (9) in particular (expected smoothnesS)
-% Relatively-Smooth Convex Optimization by First-Order Methods, and Applications,  -- Section 3
-
-\subsection{Convergence of Mirror Descent}
-To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$.
-Note that the strong convexity constant $m$ may be zero.
-Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
-\begin{align}
-	f(\natp^{t+1}) 
-	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
-	& \text{(Relative $L$-smoothness)} \\
-	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
-	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(Relative $m$-strong convexity)}
-\end{align}
-Applying the 3 points lemma with \TODO{f}
-
-\section{Self-Concordance}
-\subsection{Motivation}
-A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
-\begin{align}
-    \logpart(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
-\end{align}
-which is defined on $\eta \in \real^d$ and $\Lambda\in \real^{d\times d}$ symmetric positive definite. 
-It is not strongly convex, so $\conj$ is not smooth.
-
-Another hypothesis that may be more suitable is self-concordance. $f: \real \rightarrow \real$ is SC if 
-\begin{align}
-    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
-\end{align}
-The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
-A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
-Negative logarithm $-\log(x)$ and entropy $x\log(x)$ are both self-concordant function. This is good news for us since log-partition function may include SC logarithmic barriers.
-In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
-
-\subsection{Suboptimality and Newton Decrement}
-An important property of self-concordant functions (cite Boyd's book, although Nesterov's may be better) is that their suboptimality  may be upper bounded by the Newton Decrement 
-\begin{align}
-    \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
-\end{align}
-In general, subtracting the minimum $f^*$ of $f$ , we have
-\begin{align}
-    f(x) - f* \leq - \decrement(x) - \log( 1- \decrement(x)) \; .
-\end{align}
-Note that this bound is vacuous for $\decrement(x)\geq 1$. For $y = \decrement(x) \leq 0.68$, we have $ - y - \log(1-y) \leq y^2$, so we get the bound
-\begin{align}
-    f(x) - f^* \leq \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
-\end{align}
-
-Our functions of interest is $f(\natp)= \bregman (\natp ; \natp^*)= \bregmanconj ( \meanp^* ; \meanp) = g(\meanp)$, with minimum $f^*=0$. 
-If $\logpart$ is self-concordant, then so is $f$, but not necessarily $g$.
-The gradient and Hessian of $f$ are
-\begin{align}
-    f(\natp) 
-    &= \logpart(\natp) - \logpart(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
-    \nabla f(\natp) 
-    &= \meanp - \meanp^* = \expect[p(X|\natp)]{T(X)} - \expect[p(X|\natp^*)]{T(X)} \\
-    \nabla^2 f(\natp) 
-    &= \mSigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
-\end{align}
-so that we get the bound.
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq \decrement(\natp)^2 
-    = \norm{\mu^* - \mu(\theta)}^2_{\mSigma(\theta)^{-1}} 
-    \leq 0.46
-\end{align}
-% If we were to use $g(\meanp)$ instead, we would get dirty third order derivatives
-% \begin{align}
-%     g(\meanp) 
-%     &= \conj(\meanp^*) - \conj(\meanp) - \langle \natp, \meanp^* - \meanp \rangle \\
-%     \nabla g(\meanp) 
-%     &= -\natp - \nabla^2 \conj(\meanp) (\meanp^* - \meanp) + \natp \\
-%     &= \nabla^2 \conj(\meanp) (\meanp - \meanp^*) \\
-%     \nabla^2 g(\meanp) 
-%     &= \nabla^2 \conj(\meanp) + \nabla^3 \conj(\meanp ) (\meanp - \meanp^*)
-% \end{align}
-% where the last expression involves a third order tensor.
-
-Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$, then we would get
-\begin{align}
-    \nabla h(\meanp ) 
-    &= \natp - \natp^* \\
-    \nabla^2 h(\meanp ) 
-    &= \nabla^2 \conj(\meanp) 
-    = \nabla^2 \logpart(\natp) ^{-1}
-    = \Cov_{p(X|\natp)}[T(X)]^{-1}\\
-    \implies 
-    \decrement(\meanp)
-    &= \Var_{p(X|\natp)}[(\natp - \natp^*)^T T(X)] \; .
-\end{align}
-This is just a remark. I don't think it can help us to get anywhere.
-
-\section{Related Work}
-Exponential families are a mainstream tool in machine learning.
-They are used  to generalize linear regression \citep{mccullagh1989generalized}, PCA \citep{collins2001generalization} or k-means \citep{banerjee2005clustering} to diverse data types and distributions.
-\citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure.
-
-\citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem.
-They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
-Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors yet critical as it does not hold for multivariate normals with unknown covariance.
-\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.
-Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
-
-\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
-Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
-
-\citet{pfau2013generalized} gives a bias variance decomposition for Bregman divergences. However it does not give interesting conclusions for our problem of interest.
-
-\citet{ostrovskii2021finite}  shows a convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant.
-This convergence rate holds only after a given number of samples have been seen.
-This is because they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.
-
-\subsection{Alternative Research Tracks}
-\begin{itemize}
-	\item Weakly supervised : missing part of $x_i$ for each $i$.
-	\item Structured prediction : graphical model on $X$ or other kind of structure.
-\end{itemize}
-
-\bibliographystyle{apalike}
-\bibliography{references.bib}
-
-\newpage
-\appendix
-
-\section{Cramer-Rao Lower Bound}
-In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
-In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
-\begin{align}
-    \Cov(\hat \natp_n)  
-    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
-    \iff
-    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
-    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
-\end{align}
-This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance.
-In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"
-
-Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
-For the estimator $\hat \natp_n$, this lower bound is not trivial.
-
-\section{Posterior Expected Loss}
-
-A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is 
-\begin{align}
-    \pi(\natp) 
-    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
-    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle)
-\end{align}
-so we see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
-When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
-This is well detailed in \citet{agarwal2010geometric}.
-
-Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
-\begin{align}
-    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-\end{align}
-This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral.
-We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
-This is exactly twice the expected loss of the MLE.
-
-\begin{example}
-	We can also get a closed form for the posterior of a centered gaussian variance 
-	\begin{align}
-		\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
-	\end{align}
-	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma.
-The using results from the MLE, we get the exact formula
-	\begin{align}
-		\expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
-	\end{align}
-	and we bound it to 
-	\begin{align}
-		\inv{2n} + \frac{4}{n(n-2)}
-		\leq \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-		\leq 
-		\inv{n} + \frac{4}{n(n-2)}
-	\end{align}
-	which is about twice the MLE bounds~\eqref{eq:MLE_rate}, exactly like as we observed for the gaussian mean.
-\end{example}
-
-\section{Three points lemma}
-\label{app:3points}
-
-Proofs of convergence for mirror descent with relative smoothness rely on a specific property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
-\begin{align}
-	\min	_{x\in C} \overbrace{f(x) + \cB_h(x ; y)}^{\phi(x)}
-\end{align}
-where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector.
-Then
-\begin{align}
-	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
-\end{align}
-This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).
-
-\begin{proof}
-	The first order optimality condition is
-	\begin{align}
-		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
-		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
-	\end{align}
-	This proof relies on another property called three point property 
-	\begin{align}
-		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
-	\end{align}
-	which can be proved by expanding the right hand side.
-	By convexity of $f$ we also have 
-	\begin{align}
-		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
-	\end{align} 
-	Putting it all together we get
-	\begin{align}
-		0 
-		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
-		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
-	\end{align}
-	which concludes the proof.
-\end{proof}
+\subsection{More Illustrations of the Bias-Variance Decomposition}
+See Figure~\ref{app:fig:bias-variance-numerical}.
+
+\begin{figure}[p]
+	\centering
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=1.pdf}
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=2.pdf}
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=10.pdf}
+	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
+	\label{app:fig:bias-variance-numerical}
+\end{figure}
 

### 2021-06-18 main.tex
Minor changes to Cramer Rao and Posterior appendix

@@
 
+In our case, the bias variance decomposition tells us that we are interested in the covariance between primal and dual parameters, rather than the covariance of only of these -- $\Cov(\MAPt;\MAPm)$ rather that $\Var(\MAPm)$ or $\Var(\MAPt)$.
+
 \section{Posterior Expected Loss}
 
-A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is 
+A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is written as
 \begin{align}
@@
     \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
-    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle)
-\end{align}
-so we see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
+    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle) \; .
+\end{align}
+We see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
 When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
@@
 	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma.
-The using results from the MLE, we get the exact formula
+	Then using results from the MLE, we get the exact formula
 	\begin{align}
@@
 	\end{align}
-	which is about twice the MLE bounds~\eqref{eq:MLE_rate}, exactly like as we observed for the gaussian mean.
+	which is about twice the MLE bounds~\eqref{eq:MLE_rate}, exactly as we reportred for the gaussian mean.
 \end{example}

### 2021-06-18 main.tex
Remove Vivien's work from appendix

@@
 
-\section{Vivien's approach}
-\subsection{A big calculus in this specific setting}
-
-Indeed we know the distribution of $\mu$ it is linked with the chi-square distribution.
-We also know the expression of the excess of risk.
-So we could do a big calculus to get the exact expression of the averaged excess of risk.
-
-Recall the expression of the Gamma integral
-\begin{align}
-  \Gamma(\alpha):=\int_0^\infty t^{\alpha-1} \exp(-t) d t
-\end{align}
-Recall also that
-\begin{align}
-    \int x^q \exp(-nx^p) dx
-    &= n^{-(q+1)/p} \int (n^{1/p}x)^q \exp(-(n^{1/p}x)^p) d(n^{1/p}x)\\
-    &= n^{-(q+1)/p} \int u^q \exp(-u^p) du\\
-    &= n^{-(q+1)/p} \int t^{q/p} \exp(-t) d(t^{1/p})\\
-    &= n^{-(q+1)/p} \int p^{-1} t^{(q+1)/p - 1} \exp(-t) dt\\
-    &= n^{-(q+1)/p} p^{-1} \Gamma((q+1)/p).
-\end{align}
-Finally if there is problem in zero with the logarithm appearing in Eq. (75, at time of writing), we can have a Taylor expansion of the log 
-\[
-  \log(x) = \sum_{i=1}^\infty \frac{(-1)^{i+1}(x-1)^i}{i},
-\]
-and show convergence with some Fubini's theorem on series-integral.
-
-\subsection{Generic approach}
-
-In this subsection, I detailed a generic approach to get non-asymptotic rates.
-It is based on the calibration inequality Eq. (18, at time of writing).
-The next step would be to detailed those derivations and have precise results for the specific case of only looking at the variance.
-
-\paragraph{Calibration inequality.}
-One of the main difficulty of the proof is that $\theta_n = (\nabla A)^{-1}(\mu_n)$ can be really big, meaning that we can not apply directly ``calibration inequality'' between the Bregman divergence ${\cal B}_{A^*}(\mu^*; \mu)$ and $\norm{\mu - \mu^*}^2$.
-However, if we restrict ourselves to the domain $\theta'\in {\cal B}(0, R)$ with $R = \max(\norm{\theta}, \norm{\theta^*})$, we have that $A$ is $2^{-1}R^{-2}$-strongly convex.
-Therefore using Eq.~(18, at the time of writing), we have
-\begin{align}
-    {\cal B}_{A^*}(\mu^*; \mu) 
-    &\leq 2^{-1}R^{-2} \norm{\mu^* - \mu}^2\\
-    &= 2 \max(\norm{\theta}, \norm{\theta^*})^2\norm{\mu^* - \mu}^2\\
-    &= 2 \max(\norm{(\nabla A)^{-1}(\mu)}, \norm{(\nabla A)^{-1}(\mu^*)})^2\norm{\mu^* - \mu}^2.
-\end{align}
-With will proceed with this equation, we might get better equations.
-For example by considering a constant of strong convexity on the segment $[\norm{\theta}, \norm{\theta^*}]$ rather than the ball ${\cal B}(0, R)$.
-The behavior of the max, will be similar to the behavior of the sum of the two elements since $\max(a, b) \leq a + b \leq 2\max(a, b)$, therefore, we proceed
-with
-\begin{align}
-    {\cal B}_{A^*}(\mu^*; \mu) 
-    &\leq 2 \norm{(\nabla A)^{-1}(\mu^*)}^2 \norm{\mu^* - \mu}^2
-    + 2 \norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu^* - \mu}^2.
-\end{align}
-The first term is easy to control in expectation, see Eqs. (19-20, at time of writing), it is $\Var{T(X)} / n$. It is really the second part that we want to study. 
-Because of the structure $\norm{(\nabla A)^{-1}(\mu)} \simeq \norm{\mu}^{-1}$
-We have to take care of the event ``$\mu = n^{-1} \sum_{i\leq n} T(X_i)$ too small''.
-We know that $\mu$ concentrate towards $\mu^*$ so this event should not happen too often.
-So we want to study $\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2$.
-
-\paragraph{Tail bound or average excess of risk?}
-If we think with concentration inequality, we will get tail bound on the excess of risk, and not an average excess of risk.
-However, recall that
-\begin{align}
-  \E[X] = \E[\int_{0}^{\infty} 1_{X > t} dt]
-  = \int_{0}^\infty \E[1_{X>t}] dt
-  = \int_0^\infty \mathbb{P}(X > t) dt.
-\end{align}
-So we can translate tail bound into average.
-
-\paragraph{Concentration of a product.}
-We want to study $\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2$, it is a product of two term, we can proceed with the fact that if $A$ and $B$ are two random variables from a sample space $(\Omega, \mu)$ to $\mathbb{R}$, we have
-\[
-    \{ \omega \in \Omega \vert A(\omega) B(\omega) > t\}
-    \subset 
-    \{ \omega \in \Omega \vert A(\omega) > s\}
-    \cup
-    \{ \omega \in \Omega \vert A(\omega) > t/s\}
-\]
-Which results from the fact that if $ab > t$ than either $a > s$, either $b > t/s$.
-In our case, it means that for any $t > 0$, we have the deviation bound
-\begin{align}
-  \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2 > t)
-  &\leq \inf_{s>0} \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)}^2 > 1/s)
-  + \mathbb{P}(\norm{\mu - \mu^*}^2 > st)\\
-  &= \inf_{s>0} \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > s^{1/2})
-  + \mathbb{P}(\norm{\mu - \mu^*} > (st)^{1/2}).
-\end{align}
-So we are left with the study of 
-\begin{align}
-  \mathbb{P}(\norm{\mu - \mu^*} > t)
-  \qquad\text{and}\qquad 
-  \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > t).
-\end{align}
-
-\paragraph{Concentration on $\mu - \mu^*$.}
-Most likely $T(X)$ is sub-Gaussian (do hand-derivations or look at works like \url{https://arxiv.org/pdf/1011.3027.pdf} or \url{https://arxiv.org/pdf/1902.03736.pdf}) and Bernstein concentration inequality leads to a tail bound of the type, with $\sigma^2$ a variance parameter
-($\sigma^2 = \Var_{\theta^*}(T(X))$)
-\begin{align}
-    \mathbb{P}_{(X_i)_{i\leq n}}(\norm{\mu - \mu^*} > t) \leq \exp(-n t^2 / \sigma^2).
-\end{align}
-Normally, in Bernstein inequality there is an additional term in $n$, but the convergence rates will not depend on this additional term, so I am omitting it for simplicity.
-Note that when we will integrate that we will get
-\begin{align}
-    \int_0^\infty \mathbb{P}_{(X_i)_{i\leq n}}(\norm{\mu - \mu^*} > t^{-1/2}) dt
-    &=\int_0^\infty \exp(-n t / \sigma^2) dt\\
-    &= \sigma^2 n^{-1}\int_0^\infty \exp(- nt / \sigma^2) d(nt\sigma^{-2})\\
-    &= \sigma^2 n^{-1} \int_0^\infty \exp(-x) dx =\sigma^2 n^{-1}.
-\end{align}
-
-\paragraph{Probability that $\mu$ is close to zero.}
-Intuitively, we should study $\mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > t)$, which should relates to $\mathbb{P}(\norm{\mu} < t)$, which can be express exactly as we know the cumulative distribution function of the chi-square and that $\mu$ is a variant of the chi-square.
-It seems to me that here we will not get optimal rates, because the calibration inequality is too harsh when $\mu$ goes to $0$, as I develop in the next paragraph.
-
-\subsection{Big picture.}
-Intuitively, we have summarize in the part in $\mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} < t)$, the problem that arise with ${\cal B}_{A^*}(\mu^*; \mu)$ when $\mu$ is small. 
-We could make this intuition more formal with
-\begin{align}
-   {\cal B}_{A^*}(\mu^*; \mu)
-   &= 1_{\norm{\mu} \geq s} {\cal B}_{A^*}(\mu^*; \mu)
-   + 1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu)\\
-   &\leq \sup_{\norm{\mu'} \geq s}(\norm{(\nabla A)^{-1}(\mu')})^2 \norm{\mu - \mu^*}^2
-   + 1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu).
-\end{align}
-Suppose the $\sup$ is finite, the first term is easy to control.
-The second term is what we should take care of. We have
-\begin{align}
-    1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu) 
-    &=  1_{\norm{\mu} < s} (A^*(\mu^*) - A^*(\mu) - <\nabla A^*(\mu), \mu^* - \mu>)\\
-    &\leq  1_{\norm{\mu} < s} (A^*(\mu^*) - \inf_{\norm{\mu'} < s} A^*(\mu') + \norm{\nabla A^*(\mu)} \sup_{\norm{\mu''} < s} \norm{\mu^* - \mu''}).
-\end{align}
-We know that, under some strict convexity condition, $\nabla A^* = (\nabla A)^{-1}$ (similar to Eq. (25) at time of writing). Therefore we really have to study, with harsh notation
-\begin{align}
-    \int_{0}^s \norm{(\nabla A)^{-1}(\mu)} d\mathbb{P}(\norm{\mu} < t)
-    =\int_{0}^s \norm{(\nabla A)^{-1}(\mu)} \mathbb{P}(\norm{\mu} \in [t, t+dt]).
-\end{align}
-It seems to me that with the precedent development we were implicitly studying
-\begin{align}
-    \int_{0}^s \norm{(\nabla A)^{-1}(\mu)}^2 d\mathbb{P}(\norm{\mu} < t),
-\end{align}
-which of course is worse.
-The study can be done precisely as we know the distribution of $\mu$.
-In terms of rates, in my current view, this path is tight if and only if $<\nabla A^*(\mu), \mu^* - \mu>$ do behave like $\norm{(\nabla A)^{-1}(\mu)}$ towards $0$.
-In terms of constants, we have done pretty harsh majoration in order to get back to $c\norm{\mu-\mu^*}$ with constants $c$ that are clearly suboptimal.
-
-\paragraph{Tail bound or average risk?}
-Note that, if $\mu$ is sub-Gaussian concentrating towards $\mu^*$,
-we can give a simpler result than average risk, 
-which is that we can that that 
-with proba $1 - \mathbb{P}(\norm{\mu} < 1/2\norm{\mu^*}) = 1 - \exp(-c\norm{\mu^*}^2 n)$, 
-we have a tail bound $\mathbb{P}(\norm{\mu}^{-2}\norm{\mu - \mu^*}^2 > t) = \exp(-cnt / \norm{\mu^*})$.
-This means that we can say that 
-with proba $1 - \exp(-c\norm{\mu^*}^2 n) - \delta$ for any $\delta >0$, 
-we have that $B_{A^*}(\mu^*; \mu) < cn^{-1}\log(1/\delta) / \norm{\mu^*}$.
-
 \section{Fenchel conjugate motivation to self-concordance}

### 2021-06-18 main.tex
Move irrelevant sections on self concordance and MAP on graphs to another document

@@
 \newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
-\newcommand{\violet}[1]{\textcolor{violet}{#1}}
 \newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}
@@
 \newcommand{\decrement}{D}
-\newcommand{\Dir}{\mathrm{Dir}} % Dirichlet
 \newcommand{\linear}{\ell} % linearization of a function
@@
 \section{Self-Concordance}
-\subsection{Motivation}
+
 A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
@@
 In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
-
-\subsection{Suboptimality and Newton Decrement}
-An important property of self-concordant functions (cite Boyd's book, although Nesterov's may be better) is that their suboptimality  may be upper bounded by the Newton Decrement 
-\begin{align}
-    \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
-\end{align}
-In general, subtracting the minimum $f^*$ of $f$ , we have
-\begin{align}
-    f(x) - f* \leq - \decrement(x) - \log( 1- \decrement(x)) \; .
-\end{align}
-Note that this bound is vacuous for $\decrement(x)\geq 1$. For $y = \decrement(x) \leq 0.68$, we have $ - y - \log(1-y) \leq y^2$, so we get the bound
-\begin{align}
-    f(x) - f^* \leq \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
-\end{align}
-
-Our functions of interest is $f(\natp)= \bregman (\natp ; \natp^*)= \bregmanconj ( \meanp^* ; \meanp) = g(\meanp)$, with minimum $f^*=0$. 
-If $\logpart$ is self-concordant, then so is $f$, but not necessarily $g$.
-The gradient and Hessian of $f$ are
-\begin{align}
-    f(\natp) 
-    &= \logpart(\natp) - \logpart(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
-    \nabla f(\natp) 
-    &= \meanp - \meanp^* = \expect[p(X|\natp)]{T(X)} - \expect[p(X|\natp^*)]{T(X)} \\
-    \nabla^2 f(\natp) 
-    &= \mSigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
-\end{align}
-so that we get the bound.
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq \decrement(\natp)^2 
-    = \norm{\mu^* - \mu(\theta)}^2_{\mSigma(\theta)^{-1}} 
-    \leq 0.46
-\end{align}
-% If we were to use $g(\meanp)$ instead, we would get dirty third order derivatives
-% \begin{align}
-%     g(\meanp) 
-%     &= \conj(\meanp^*) - \conj(\meanp) - \langle \natp, \meanp^* - \meanp \rangle \\
-%     \nabla g(\meanp) 
-%     &= -\natp - \nabla^2 \conj(\meanp) (\meanp^* - \meanp) + \natp \\
-%     &= \nabla^2 \conj(\meanp) (\meanp - \meanp^*) \\
-%     \nabla^2 g(\meanp) 
-%     &= \nabla^2 \conj(\meanp) + \nabla^3 \conj(\meanp ) (\meanp - \meanp^*)
-% \end{align}
-% where the last expression involves a third order tensor.
-
-Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$, then we would get
-\begin{align}
-    \nabla h(\meanp ) 
-    &= \natp - \natp^* \\
-    \nabla^2 h(\meanp ) 
-    &= \nabla^2 \conj(\meanp) 
-    = \nabla^2 \logpart(\natp) ^{-1}
-    = \Cov_{p(X|\natp)}[T(X)]^{-1}\\
-    \implies 
-    \decrement(\meanp)
-    &= \Var_{p(X|\natp)}[(\natp - \natp^*)^T T(X)] \; .
-\end{align}
-This is just a remark. I don't think it can help us to get anywhere.
 
@@
 
-\section{Fenchel conjugate motivation to self-concordance}
-
-In the most regular case, when $f(x)$ is a convex function, continuously differentiable on its domain, then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ verifies
-\begin{align}
-    \nabla f \circ \nabla f^*  &= \Id \\
-     \nabla f^* \circ \nabla f &= \Id
-\end{align}
-where $\Id$ is the identity function on the relevant domain. In words, the gradients of $f$ and $f^*$ are reciprocal.
-Deriving this equality yields
-\begin{align}
-    \nabla^2 f(x) \nabla^2 f^*(x^*) = I_n
-\end{align}
-where $x,x^*$ are conjugate points -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
-Now, it gets interesting to us when we derive again this equality. Let's tackle the 1D case first
-\begin{align}
-    &f''(x) f^{* \prime \prime}(f'(x)) = 1, \forall x \\
-    \implies
-    &f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))  = 0 \\
-    \implies
-    &\frac{f'''(x)}{f''(x){\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*){\frac{3}{2}}}  = 0
-\end{align}
-where to get to the last line we used the first line, and we divided the second line by $f''(x)^{\frac{1}{2}}$. 
-We see that for a pair of conjugate functions, the self-concordance ratio is preserved, modulo the sign.
-This gives another rational, beyond dimensional analysis, for using this ratio as a regularity assumption for convex analysis.
-
-It is also very helpful for us, since we are looking at pairs $\logpart, \conj$, and their associated Bregman divergences. 
-If $\logpart$ is SC, then so is 
-$f(\natp)= \bregman(\natp ; \natp^*) = \logpart(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
-And $\conj$ is SC as well, thus  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$ is SC.
-But there is no reason for 
-$g(\meanp) = \bregmanconj ( \meanp^* ; \meanp) =  \cst - \conj(\meanp) - \langle \nabla \conj (\meanp), \meanp^* - \meanp \rangle $ to be SC.
-
-The multivariate generalization of this formula is a third order tensor equality
-\begin{align}
-    \nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1} + \nabla^2 f^{* -\half} \nabla^3 f^* \nabla^2 f^{* -1} = 0
-\end{align}
-where we omit  multiplication axis and functions take relevant argument $x$ or $x^*$.
-Consequently, a multivariate definition of self-concordance might take the form of an inequality on the 3d tensor $\nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1}$.
-
-\section{MAP on Graphs}
-Assume that the variable $X$ factors along some graph $G$. 
-We write $G(i)$ the parents of $X_i$ in $G$. 
-Then we model the conditional distribution of $X$ given parameter vector $\theta$ factors as
-\begin{align}
-    p(X|\natp) = \prod_i p(X_i | X_{G(i)};\natp_i)
-\end{align}
-where $\natp_i$ is the parameter associated to the mechanism $X_{G(i)} \rightarrow X_i$. 
-Embracing the Bayesian viewpoint, the independent mechanism principle is embodied as independence between parameters
-\begin{align}
-    p(\natp) = \prod_i p(\natp_i) \; .
-\end{align}
-% In other words, we have the following assumptions on $(\natp,X)$
-% \begin{align}
-%     \indep_{i=1}^d \natp_i \\
-%     \natp_i \indep X_{G(i)}
-% \end{align}
-Following these equations, the joint distribution on $(\natp,X)$ factors along a larger graph $G'$ which augments $G$ by adding nodes $\natp_i$ with arrows pointing to $X_i$, as illustrated in Figure~\ref{fig:joint_graph}.
-\begin{figure}[ht]
-    \centering
-    \begin{tikzpicture}[shorten >=1pt,->]
-      \tikzstyle{vertex}=[draw, circle,minimum size=24pt,inner sep=0pt]
-      \foreach \name/\txt/\x/\y in {x1/X_1/0/0, x2/X_2/3/0, x3/X_3/2/-1.5, t1/\theta_1/-1.5/-0.5, t2/\theta_2/4.5/-0.5, t3/\theta_3/0.5/-2}
-        \node[vertex] (G-\name) at (\x,\y) {$\txt$};
-
-      \foreach \from/\to in {x1/x2, x2/x3, x1/x3, t1/x1, t2/x2, t3/x3}
-        \draw (G-\from) -- (G-\to);
-    \end{tikzpicture}
-    \caption{A graph $G'$ factorizing $(\natp,X)$. Although the graph restricted on $X$ does not encode any conditional independence, $G'$ does on the joint distribution.}
-    \label{fig:joint_graph}
-\end{figure}
-With such a graph, the Bayesian posterior can be factorized as well
-\begin{align}
-    p(\natp|X) 
-    &\propto  p(X|\natp) p(\natp)& \\
-    & = \prod_i p(X_i | X_{G(i)};\natp_i) p(\natp_i)
-    &(\indep_i \natp_i) \\
-    & = \prod_i p(X_i | X_{G(i)};\natp_i) p(\natp_i\violet{|X_{G(i)}})
-    &(\natp_i \indep X_{G(i)}) \\
-    & = \prod_i p(X_i, \violet{\natp_i} | X_{G(i)})    &  \\
-    & = \prod_i p(\natp_i| \violet{X_i}, X_{G(i)})p(X_i|X_{G(i)})    &  \\
-    \implies p(\natp|X) 
-    &= \prod_i p(\natp_i| \violet{X_i}, X_{G(i)}) \; .
-\end{align}
-In words, a consequence of the independence mechanism principle is that the posterior distribution of $\natp_i$ can be inferred solely from $X_i$ and its parents.
-
-\subsection{Equality of directions for 2 categorical variables}
-In my paper on the analysis of causal speed, I proved the equivalence between sampling a joint distribution $\omega = p(A,B) \in \simplex_{K\times K}$ on $(A,B)\in \{1,\dots,K\}^2$ from a Dirichlet with parameter $\gamma\in \real_+^{K\times K}$ and sampling independently the marginal distribution $\mu = p(A)\in \simplex_K$ and the conditional distributions $\nu_i = p(B|A=i)\in\simplex_K$ from Dirichlets with respective parameters $\sum_{j=1}^K \gamma_{:,j} = \gamma \ones$ (matrix vector product) and $\gamma_{i,:}$
-\begin{align}
-    \underbrace{\Dir_{K^2}((\gamma_{i,j})_{i,j=1}^K) }_{p(\omega)}
-    \equiv 
-    \underbrace{\Dir_K\left( \gamma \ones \right)}_{p(\mu)} 
-    \otimes \left (\bigotimes_{i=1}^K 
-    \underbrace{\Dir_K((\gamma_{i,j})_j)}_{p(\nu_i)} 
-    \right ) 
-    \label{eq:dirichlet_factors}
-\end{align}
-Seeing data samples $(\cA,\cB) = (A_i,B_i)_{i=1}^n$ as one-hot encodings in $\real^K\times \real^K$, the posterior reads
-\begin{align}
-p(\mu|\cA) &= \Dir(\gamma\ones + \sum_i A_i)\\
-p(\nu_k|\cA,\cB) &= \Dir(\gamma_{k,:} + \sum_i  A_{i,k} B_i ) \\
-p(\omega | \cA,\cB) & = \Dir( \gamma + \sum_i A_i B_i^\top )  
-\end{align}
-where $A_i B_i^\top$ is the one hot matrix encoding of $A,B$.
-These three posteriors are obtained independently of each other following rules of calculus for Dirichlet distributions.
-Yet they happen to define the same distribution on distributions, as we verify below with the two equalities from equation \eqref{eq:dirichlet_factors}.
-\begin{align}
-    \left(\gamma + \sum_i A_i B_i^\top \right)_{k,l} 
-    &= \left(\gamma_{k,:} + \sum_i  A_{i,k} B_i \right)_l \\
-    \left(\gamma + \sum_i A_i B_i^\top \right) \ones 
-    &= \gamma \ones + \sum_i A_i \; .
-\end{align}
-The interpretation of this result is that \emph{taking the posterior with the decomposition $A\rightarrow B$ or $B\rightarrow A$ give the same result}. As a corollary the MAP is also the same
-\begin{align}
-\hat \omega^\text{MAP} 
-= \frac{\gamma + \sum_i A_i B_i^\top}{\ones^\top (\gamma + \sum_i A_i B_i^\top) \ones}
-=\frac{\gamma + \sum_i A_i B_i^\top}{n_0 + n}
-\end{align}
-Using Bayesian statistics with this prior, there is no distinction between directions.
-
-Is this bound to happen with a symmetric prior ? Let's give a name to the change of variable $f(\omega) = \mu,\nu$. 
-Remark that $f(\omega^\top) = \mu_\leftarrow, \nu_\leftarrow$, eg in the categorical special case transposing omega and changing variables give the anticausal direction. 
-For sure  $p(X|\mu,\nu) = p(X|f(\omega)) = p(X|\omega)$. 
-Using the change of variable formula  we get something.
-But which equality am I looking for exactly ? 
 \end{document}

### 2021-06-18 other.tex
Move irrelevant sections on self concordance and MAP on graphs to another document

\documentclass{article}

\usepackage[top=2cm, bottom=2cm]{geometry}
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
%\usepackage{subfigure}
\usepackage{subcaption}
\usepackage{hyperref}       % hyperlinks
\usepackage[dvipsnames]{xcolor}
\usepackage{tikz}           % define for loops

\usepackage{math_commands}

\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size

\newcommand{\MAPm}{\hat \mu_n}
\newcommand{\MAPt}{\hat \natp_n}

\newcommand{\Dir}{\mathrm{Dir}} % Dirichlet
\newcommand{\violet}[1]{\textcolor{violet}{#1}}

\begin{document}

\section{Self-Concordance}

\subsection{Fenchel conjugate motivation to self-concordance}

In the most regular case, when $f(x)$ is a convex function, continuously differentiable on its domain, then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ verifies
\begin{align}
    \nabla f \circ \nabla f^*  &= \Id \\
     \nabla f^* \circ \nabla f &= \Id
\end{align}
where $\Id$ is the identity function on the relevant domain. In words, the gradients of $f$ and $f^*$ are reciprocal.
Deriving this equality yields
\begin{align}
    \nabla^2 f(x) \nabla^2 f^*(x^*) = I_n
\end{align}
where $x,x^*$ are conjugate points -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
Now, it gets interesting to us when we derive again this equality. Let's tackle the 1D case first
\begin{align}
    &f''(x) f^{* \prime \prime}(f'(x)) = 1, \forall x \\
    \implies
    &f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))  = 0 \\
    \implies
    &\frac{f'''(x)}{f''(x){\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*){\frac{3}{2}}}  = 0
\end{align}
where to get to the last line we used the first line, and we divided the second line by $f''(x)^{\frac{1}{2}}$. 
We see that for a pair of conjugate functions, the self-concordance ratio is preserved, modulo the sign.
This gives another rational, beyond dimensional analysis, for using this ratio as a regularity assumption for convex analysis.

It is also very helpful for us, since we are looking at pairs $\logpart, \conj$, and their associated Bregman divergences. 
If $\logpart$ is SC, then so is 
$f(\natp)= \bregman(\natp ; \natp^*) = \logpart(\natp) - \langle \meanp^*, \natp \rangle + \cst$. 
And $\conj$ is SC as well, thus  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$ is SC.
But there is no reason for 
$g(\meanp) = \bregmanconj ( \meanp^* ; \meanp) =  \cst - \conj(\meanp) - \langle \nabla \conj (\meanp), \meanp^* - \meanp \rangle $ to be SC.

The multivariate generalization of this formula is a third order tensor equality
\begin{align}
    \nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1} + \nabla^2 f^{* -\half} \nabla^3 f^* \nabla^2 f^{* -1} = 0
\end{align}
where we omit  multiplication axis and functions take relevant argument $x$ or $x^*$. Consequently, a multivariate definition of self-concordance might take the form of an inequality on the 3d tensor $\nabla^2 f^{-\half} \nabla^3 f \nabla^2 f^{-1}$.

\subsection{Suboptimality and Newton Decrement}
An important property of self-concordant functions (cite Boyd's book, although Nesterov's may be better) is that their suboptimality  may be upper bounded by the Newton Decrement 
\begin{align}
    \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
\end{align}
In general, subtracting the minimum $f^*$ of $f$ , we have
\begin{align}
    f(x) - f* \leq - \decrement(x) - \log( 1- \decrement(x)) \; .
\end{align}
Note that this bound is vacuous for $\decrement(x)\geq 1$. For $y = \decrement(x) \leq 0.68$, we have $ - y - \log(1-y) \leq y^2$, so we get the bound
\begin{align}
    f(x) - f^* \leq \decrement(x)^2 = \nabla f(x)^T \nabla^2 f(x)^{-1} \nabla f(x) \; .
\end{align}

Our functions of interest is $f(\natp)= \bregman (\natp ; \natp^*)= \bregmanconj ( \meanp^* ; \meanp) = g(\meanp)$, with minimum $f^*=0$. 
If $\logpart$ is self-concordant, then so is $f$, but not necessarily $g$.
The gradient and Hessian of $f$ are
\begin{align}
    f(\natp) 
    &= \logpart(\natp) - \logpart(\natp^*) - \langle \meanp^*, \natp - \natp^* \rangle \\
    \nabla f(\natp) 
    &= \meanp - \meanp^* = \expect[p(X|\natp)]{T(X)} - \expect[p(X|\natp^*)]{T(X)} \\
    \nabla^2 f(\natp) 
    &= \mSigma(\natp) = \Cov_{p(X|\natp)}[T(X)]
\end{align}
so that we get the bound.
\begin{align}
    \bregmanconj(\mu^* ; \mu) 
    \leq \decrement(\natp)^2 
    = \norm{\mu^* - \mu(\theta)}^2_{\mSigma(\theta)^{-1}} 
    \leq 0.46
\end{align}
% If we were to use $g(\meanp)$ instead, we would get dirty third order derivatives
% \begin{align}
%     g(\meanp) 
%     &= \conj(\meanp^*) - \conj(\meanp) - \langle \natp, \meanp^* - \meanp \rangle \\
%     \nabla g(\meanp) 
%     &= -\natp - \nabla^2 \conj(\meanp) (\meanp^* - \meanp) + \natp \\
%     &= \nabla^2 \conj(\meanp) (\meanp - \meanp^*) \\
%     \nabla^2 g(\meanp) 
%     &= \nabla^2 \conj(\meanp) + \nabla^3 \conj(\meanp ) (\meanp - \meanp^*)
% \end{align}
% where the last expression involves a third order tensor.

Finally, if instead we were looking at a different function switching the role of $\meanp$ and $\meanp^*$,  $h(\meanp) = \bregmanconj ( \meanp ; \meanp^*)$, then we would get
\begin{align}
    \nabla h(\meanp ) 
    &= \natp - \natp^* \\
    \nabla^2 h(\meanp ) 
    &= \nabla^2 \conj(\meanp) 
    = \nabla^2 \logpart(\natp) ^{-1}
    = \Cov_{p(X|\natp)}[T(X)]^{-1}\\
    \implies 
    \decrement(\meanp)
    &= \Var_{p(X|\natp)}[(\natp - \natp^*)^T T(X)] \; .
\end{align}
This is just a remark. I don't think it can help us to get anywhere.

\section{MAP on Graphs}
Assume that the variable $X$ factors along some graph $G$. 
We write $G(i)$ the parents of $X_i$ in $G$. 
Then we model the conditional distribution of $X$ given parameter vector $\theta$ factors as
\begin{align}
    p(X|\natp) = \prod_i p(X_i | X_{G(i)};\natp_i)
\end{align}
where $\natp_i$ is the parameter associated to the mechanism $X_{G(i)} \rightarrow X_i$. 
Embracing the Bayesian viewpoint, the independent mechanism principle is embodied as independence between parameters
\begin{align}
    p(\natp) = \prod_i p(\natp_i) \; .
\end{align}
% In other words, we have the following assumptions on $(\natp,X)$
% \begin{align}
%     \indep_{i=1}^d \natp_i \\
%     \natp_i \indep X_{G(i)}
% \end{align}
Following these equations, the joint distribution on $(\natp,X)$ factors along a larger graph $G'$ which augments $G$ by adding nodes $\natp_i$ with arrows pointing to $X_i$, as illustrated in Figure~\ref{fig:joint_graph}.
\begin{figure}[ht]
    \centering
    \begin{tikzpicture}[shorten >=1pt,->]
      \tikzstyle{vertex}=[draw, circle,minimum size=24pt,inner sep=0pt]
      \foreach \name/\txt/\x/\y in {x1/X_1/0/0, x2/X_2/3/0, x3/X_3/2/-1.5, t1/\theta_1/-1.5/-0.5, t2/\theta_2/4.5/-0.5, t3/\theta_3/0.5/-2}
        \node[vertex] (G-\name) at (\x,\y) {$\txt$};

      \foreach \from/\to in {x1/x2, x2/x3, x1/x3, t1/x1, t2/x2, t3/x3}
        \draw (G-\from) -- (G-\to);
    \end{tikzpicture}
    \caption{A graph $G'$ factorizing $(\natp,X)$. Although the graph restricted on $X$ does not encode any conditional independence, $G'$ does on the joint distribution.}
    \label{fig:joint_graph}
\end{figure}
With such a graph, the Bayesian posterior can be factorized as well
\begin{align}
    p(\natp|X) 
    &\propto  p(X|\natp) p(\natp)& \\
    & = \prod_i p(X_i | X_{G(i)};\natp_i) p(\natp_i)
    &(\indep_i \natp_i) \\
    & = \prod_i p(X_i | X_{G(i)};\natp_i) p(\natp_i\violet{|X_{G(i)}})
    &(\natp_i \indep X_{G(i)}) \\
    & = \prod_i p(X_i, \violet{\natp_i} | X_{G(i)})    &  \\
    & = \prod_i p(\natp_i| \violet{X_i}, X_{G(i)})p(X_i|X_{G(i)})    &  \\
    \implies p(\natp|X) 
    &= \prod_i p(\natp_i| \violet{X_i}, X_{G(i)}) \; .
\end{align}
In words, a consequence of the independence mechanism principle is that the posterior distribution of $\natp_i$ can be inferred solely from $X_i$ and its parents.

\subsection{Equality of directions for 2 categorical variables}
In my paper on the analysis of causal speed, I proved the equivalence between sampling a joint distribution $\omega = p(A,B) \in \simplex_{K\times K}$ on $(A,B)\in \{1,\dots,K\}^2$ from a Dirichlet with parameter $\gamma\in \real_+^{K\times K}$ and sampling independently the marginal distribution $\mu = p(A)\in \simplex_K$ and the conditional distributions $\nu_i = p(B|A=i)\in\simplex_K$ from Dirichlets with respective parameters $\sum_{j=1}^K \gamma_{:,j} = \gamma \ones$ (matrix vector product) and $\gamma_{i,:}$
\begin{align}
    \underbrace{\Dir_{K^2}((\gamma_{i,j})_{i,j=1}^K) }_{p(\omega)}
    \equiv 
    \underbrace{\Dir_K\left( \gamma \ones \right)}_{p(\mu)} 
    \otimes \left (\bigotimes_{i=1}^K 
    \underbrace{\Dir_K((\gamma_{i,j})_j)}_{p(\nu_i)} 
    \right ) 
    \label{eq:dirichlet_factors}
\end{align}
Seeing data samples $(\cA,\cB) = (A_i,B_i)_{i=1}^n$ as one-hot encodings in $\real^K\times \real^K$, the posterior reads
\begin{align}
p(\mu|\cA) &= \Dir(\gamma\ones + \sum_i A_i)\\
p(\nu_k|\cA,\cB) &= \Dir(\gamma_{k,:} + \sum_i  A_{i,k} B_i ) \\
p(\omega | \cA,\cB) & = \Dir( \gamma + \sum_i A_i B_i^\top )  
\end{align}
where $A_i B_i^\top$ is the one hot matrix encoding of $A,B$.
These three posteriors are obtained independently of each other following rules of calculus for Dirichlet distributions. Yet they happen to define the same distribution on distributions, as we verify below with the two equalities from equation \eqref{eq:dirichlet_factors}.
\begin{align}
    \left(\gamma + \sum_i A_i B_i^\top \right)_{k,l} 
    &= \left(\gamma_{k,:} + \sum_i  A_{i,k} B_i \right)_l \\
    \left(\gamma + \sum_i A_i B_i^\top \right) \ones 
    &= \gamma \ones + \sum_i A_i \; .
\end{align}
The interpretation of this result is that \emph{taking the posterior with the decomposition $A\rightarrow B$ or $B\rightarrow A$ give the same result}. As a corollary the MAP is also the same
\begin{align}
\hat \omega^\text{MAP} 
= \frac{\gamma + \sum_i A_i B_i^\top}{\ones^\top (\gamma + \sum_i A_i B_i^\top) \ones}
=\frac{\gamma + \sum_i A_i B_i^\top}{n_0 + n}
\end{align}
Using Bayesian statistics with this prior, there is no distinction between directions.

Is this bound to happen with a symmetric prior ? Let's give a name to the change of variable $f(\omega) = \mu,\nu$. 
Remark that $f(\omega^\top) = \mu_\leftarrow, \nu_\leftarrow$, eg in the categorical special case transposing omega and changing variables give the anticausal direction. 
For sure  $p(X|\mu,\nu) = p(X|f(\omega)) = p(X|\omega)$. 
Using the change of variable formula  we get something.
But which equality am I looking for exactly ? 
\end{document}

### 2021-06-22 main.tex
Start clarifying SBPP

@@
 \TODO{Under-parametrized conjugate prior ?}
-The conjugate prior for $p(X|\natp)$ is
-\begin{align}
-    \pi(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
+One conjugate prior for $p(X|\natp)$ is
+\begin{align}
+    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
 \end{align}
@@
 \begin{align}
-    \pi(\natp) \propto 
+    p(\natp) \propto 
     \exp( -n_0 \logpart (\natp) 
@@
 \begin{align}
-	\pi(\natp) 
+	p(\natp) 
 	&\propto 
@@
 \begin{align*}
-    -\log \pi(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
+    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
 \end{align*}
@@
 \begin{align}
-    -\log p(\cD|\natp)\pi(\natp) 
+    -\log p(\cD|\natp)p(\natp) 
     = (n_0+n) \logpart (\natp) 
@@
 \begin{theorem}[MAP Bound]
- Introducing $a=n_0 \frac{\mu_0}{\mu^*}$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
+ Introducing $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
  \begin{equation}
@@
 
-\paragraph{Dual Expectation Pivot.}
+\subsection{Dual Expectation Pivot.}
 Let $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ be the MAP estimate of the mean (dual) parameter, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ the corresponding natural (primal) parameter, and
@@
 
-\paragraph{Primal Expectation Pivot}
+\subsection{Primal Expectation Pivot}
 Let $\tilde \theta_n := \expect{\hat \theta_n}$ and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the primal and dual parameters of the primal expectation of the MAP.
@@
 In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in both of these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
-
-We review each of these terms independently below.
-
-\begin{important}
-Goal = match variance assumptions with these bounds	
-\end{important}
 
@@
 \paragraph{Problem.}
-Our goal is to solve the following problem
-\begin{align}
-	\min_\natp \E[f(\natp ; X)] := -\langle \E[T(X)] ; \natp \rangle + \logpart(\natp)
-\end{align}
-where $\logpart$ is a convex function and we only have access to stochastic oracles $X \sim p(X | \natp^*)$, meaning we can compute stochastic values and gradients 
+We want  to solve the following problem
+\begin{align}
+	f^* = \min_\natp \E[f(\natp ; X)] := -\langle \E[T(X)] ; \natp \rangle + \logpart(\natp)
+\end{align}
+where $\logpart$ is a convex function.
+We assume access to stochastic oracles $X \sim p(X | \natp^*)$ -- e.g. we observe $X$ such that $\E[T(X)] = \mu^*$ and we can compute  stochastic function values and gradients 
 \begin{align}
 	f(\natp ; X) &= -\langle T(X) ; \natp \rangle + \logpart(\natp) \\
-	\nabla f(\natp ; X) &= \nabla \logpart(\natp) - T(X)
-\end{align}
-such that $\E[T(X)] = \mu^*$.
+	\nabla f(\natp ; X) &= \nabla \logpart(\natp) - T(X)\; .
+\end{align}
+It is important to note that our problem is merely a stochastic linear tilt of the convex function $\logpart$, yet none of the convergence rates we studied applied to this setting.
 
@@
 We observe $n$ data points $X_i$. 
-We are studying two algorithms to solve this problem : the MLE and the MAP
+We are studying two algorithms solving this problem : MLE and MAP.
 \begin{align}
@@
 \paragraph{Interpretation.}
-As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms. 
+As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms with divergence $\bregman$ and initialization $\natp_0$ : 
 \begin{itemize}
-	\item stochastic bregman proximal point with divergence $\bregman$, initialization $\natp_0$, and learning rate $\inv{n_0+n}$,
-	\item stochastic mirror descent with divergence $\bregman$, initialization $\natp_0$, and learning rate  $\inv{n_0+n +1}$.
+	\item Stochastic Bregman Proximal Point with learning rate $\inv{n_0+n}$,
+	\item Stochastic Mirror Sescent with learning rate  $\inv{n_0+n +1}$.
 \end{itemize}
-In both cases, $f(\natp)$ is convex and 1-smooth / 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.
-
-\paragraph{Goal.} We want to bound the suboptimality (in function value) as a function of the number of samples.
-
-\paragraph{Other assumptions.}
-An assumption pointed out by Nicolas Loizou is 
-\begin{align}
-	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
-\end{align}
-for which a sufficient condition is 
-\begin{align}
-\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
-\end{align}
-which holds in a lot of settings but not all.
-For instance, a gaussian $\cN(\mu,\sigma^2)$ can overfit on a single data point and give it infinite density -- eg converge to a Dirac.
-However we can overcome this issue by grouping data points : a gaussian can not overfit on two points.
-Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$.
-So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)
+In both cases, $f(\natp)$ is convex and 1-smooth  and 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.
+
+\paragraph{Goal.} We want to bound the suboptimality $f(\natp) - f^*$ as a function of the number of samples.
 
 \section{Stochastic Bregman Proximal Point}
-Remark that the MAP estimate minimizes the sum of a stochastic loss $ -\log p(X|\natp)$ and a deterministic divergence to an initial point $n_0 \bregman(\natp ; \natp_0)$.
-This is a stochastic Bregman proximal step with step-size $\inv{n_0}$.
-This can also be seen at each step since
-\begin{align}
-    \hat \natp_{n+1} 
-    &= \argmin_\natp -\log p(x_{n}|\natp) + (n_0 + n) \bregman(\natp ; \hat \natp_n) \\
-    &= \argmin_\natp f(\natp; x_{n}) +  \inv{\gamma_{n}}\bregman(\natp ; \hat \natp_n) \; .
-\end{align}
-Hence the MAP estimate can also be seen as the result of a stochastic proximal Bregman point algorithm with step-size $\gamma_n = \inv{n_0 + n}$ at step $n$.
-
-This is similar to the online learning setup, and it may be possible to bound the regret, with approaches similar to Adagrad.
-
+\subsection{Correspondence with MAP}
+For a dataset of independent samples $\cD_n = \{X_1,\dots,X_n\}$, the posterior can be seen iteratively
+\begin{align}
+	p(\natp |\cD_n)
+	=p(\natp|X_n,\cD_{n-1})
+	\propto p(X_n|\natp,\cD_{n-1}) p(\natp|\cD_{n-1})
+	=p(X_n|\natp)p(\natp |\cD_{n-1}) \; .
+\end{align}
+For the MAP of the exponential family with a Bregman conjugate prior, the posterior takes the form
+\begin{align}
+	- \log p(\natp | \cD_n) 
+	&= \sum_{i=1}^n -\log p(X_i | \natp) - \log p(\natp) + \cst\\
+	&= \sum_i - \langle T(X_i) ; \natp \rangle  + n \logpart(\natp) + n_0 \bregman(\natp ; \natp_0) +\cst \\
+	&= (n+n_0)\logpart(\natp)  - \langle n_0 \mu_0 + \sum_i T(X_i) ; \natp \rangle + \cst \\
+	&= (n+n_0) \bregman(\natp; \hat \natp_n^\text{MAP}) + \cst \; ,
+\end{align}
+so we immediately get the recursion
+\begin{align}
+    \hat \natp_n^\text{MAP}
+    &= \argmin_\natp - \log p(\natp | \cD_n) \\
+    &= \argmin_\natp - \log p(X_n|\natp) - \log p(\natp | \cD_{n-1}) \\
+    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
+\end{align} 
+Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0}$.
+
+\subsection{Straightforward Analysis}
 The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
@@
 This is exactly the same assumption as given by the SMD analysis.
-However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which gives more weight to the first iterates.
+However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which is an average in the primal space, and gives more weight to the first iterates.
 It does not immediately apply to the last iterate.
@@
 
+\subsection{Work by Loizou}
+An assumption pointed out by Nicolas Loizou is 
+\begin{align}
+	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
+\end{align}
+for which a sufficient condition is 
+\begin{align}
+\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
+\end{align}
+which holds in a lot of settings but not all.
+For instance, a gaussian $\cN(\mu,\sigma^2)$ can overfit on a single data point and give it infinite density -- eg converge to a Dirac.
+However we can overcome this issue by grouping data points : a gaussian can not overfit on two points.
+Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$.
+So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)
+
 \section{Expected Relative Smoothness}
@@
 % Relatively-Smooth Convex Optimization by First-Order Methods, and Applications,  -- Section 3
-
-\subsection{Three points lemma}
-\label{app:3points}
-
-Proofs of convergence for mirror descent with relative smoothness rely on a specific property of Bregman divergences, familiar to information geometry folks. If $x_+$ is solution to 
-\begin{align}
-	x_+ \in \argmin	_{x\in C} \overbrace{f(x) + \cB_h(x ; y)}^{\phi(x)}
-\end{align}
-where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$ (we will drop the index, and use a semicolon instead of double bars to lighten the notation), and $y$ is some reference vector.
-Then
-\begin{align}
-	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
-\end{align}
-This property is an analog of the Pythagorean theorem for generalized projections (setting $f=0$ and $h=\|.\|^2$).
-
-\begin{proof}
-	The first order optimality condition is
-	\begin{align}
-		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
-		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
-	\end{align}
-	This proof relies on another property called three point property 
-	\begin{align}
-		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
-	\end{align}
-	which can be proved by expanding the right hand side.
-	By convexity of $f$ we also have 
-	\begin{align}
-		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
-	\end{align} 
-	Putting it all together we get
-	\begin{align}
-		0 
-		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
-		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
-	\end{align}
-	which concludes the proof.
-\end{proof}
 
@@
 
+\paragraph{Bregman proximal point}
+
+\paragraph{high probability bounds}
+
 \subsection{Alternative Research Tracks}
@@
 
+\section{Three points lemma}
+\label{app:3points}
+
+The typical descent lemma of updates involving minimization of a Bregman divergence is called the three point lemma. 
+It corresponds to first order optimality conditions of the update.
+With its hep, one can prove convergence for mirror descent with relative smoothness, or for Bregman proximal point algorithms.
+
+\begin{lemma}
+Let $x_+$ be a solution to 
+\begin{align}
+	x_+ \in \argmin	_{x\in C} \overbrace{f(x) + \cB_h(x ; y)}^{\phi(x)}
+\end{align}
+where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$, and $y$ is some reference vector. Then 
+\begin{align}
+	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
+\end{align}
+\end{lemma}
+
+Taking an information geometry lens, this property is analog to the Pythagorean theorem for generalized projections.
+We recover euclidean projections setting $f=0$ and $h=\|.\|^2$.
+
+\begin{proof}
+	The first order optimality condition is
+	\begin{align}
+		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
+		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
+	\end{align}
+	This proof relies on another property called three point property 
+	\begin{align}
+		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
+	\end{align}
+	which can be proved by expanding the right hand side.
+	By convexity of $f$ we also have 
+	\begin{align}
+		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
+	\end{align} 
+	Putting it all together we get
+	\begin{align}
+		0 
+		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
+		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
+	\end{align}
+	which concludes the proof.
+\end{proof}
+
 \section{Cramer-Rao Lower Bound}
@@
 \begin{align}
-    \pi(\natp) 
+    p(\natp) 
     \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
@@
 \end{align}
-We see that $\pi$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
-When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $\pi(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
+We see that $p(\natp)$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
+When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $p(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
 This is well detailed in \citet{agarwal2010geometric}.
@@
 \begin{align}
-    \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+    \expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
 \end{align}
@@
 	\begin{align}
-		\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
+		\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
 	\end{align}
@@
 	\begin{align}
-		\expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		\expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
 		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
@@
 		\inv{2n} + \frac{4}{n(n-2)}
-		\leq \expect[\cD\sim\natp_*]{\expect[\pi(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		\leq \expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
 		\leq 

### 2021-06-22 main.tex
Fix example bug
Add proofs for Stochastic BPP

@@
 We now present a lemma on its expected inverse.
-\begin{example}
 \begin{lemma}[Expected MAP Natural Parameter]
@@
 \end{lemma}
-\end{example}
 Note that the lower bound is a simple consequence of the convexity of the inverse function. 
@@
 \subsection{Straightforward Analysis}
-The analysis of deterministic Bregman proximal point relies on the three points lemma (Appendix~\ref{app:3points}).
-First one can prove descent, then one can prove that the total path length and the sum of suboptimalities are bounded, but this does not transpose immediately to the stochastic setting.
-One needs an assumption on the quality of the stochastic estimates $f(.
-; x)$.
-Given the proof, the most straightforward such assumption (see paper notes) is 
-\begin{align}
-\expect{f(\natp_{n+1}) - f(\natp_{n+1} ; x_{n})} \leq \gamma_n \sigma^2
-\end{align}
-which translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in our case.
-This is exactly the same assumption as given by the SMD analysis.
+
+\paragraph{Bregman Proximal Point}
+The analysis of deterministic Bregman proximal \citep{chen1993convergence} point relies on the three points lemma (Appendix~\ref{app:3points})
+\begin{align}
+	 \forall \natp, \lr_t f(\natp) + \cB(\natp ; \natp_t) \geq \lr_t f(\natp_{t+1}) + \cB( \natp; \natp_{t+1}) +\cB(\natp_{t+1} ; \natp_{t}) \; .
+\end{align}
+Applying it with $\natp=\natp_t$, we prove that function values are decreasing
+\begin{align}
+	f(\natp_{t+1})  \leq f(\natp_{t})  - \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \; .
+\end{align}
+Then applying it with $\natp=\natp^*$ (assuming such a minimizer exists), we obtain a telescopic sum
+\begin{align}
+	\lr_t(f(\natp_{t+1}) - f(\natp^*)) +  \cB(\natp_{t+1} ; \natp_{t}) 
+	&\leq  \cB( \natp^*; \natp_{t})  - \cB( \natp^*; \natp_{t+1}) \\
+	\implies \sum_{t=0}^{T-1} \lr_t (f(\natp_{t+1}) - f(\natp^*)) 
+	+ \sum_{t=0}^{T-1} \cB(\natp_{t+1} ; \natp_{t}) 
+	&\leq \cB( \natp^*; \natp_{0})  - \cB( \natp^*; \natp_{T})  
+	\leq \cB( \natp^*; \natp_{0})  \; ,
+\end{align}
+which proves on the way that the total path length $ \sum_{t=0}^{t-1} \cB(\natp_{t+1} ; \natp_{t}) $ is bounded.
+Using the decreasing property of $f(\natp_t)$, we get
+\begin{align}
+	f(\natp_{T}) - f(\natp^*) \leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t} \; .
+\end{align}
+This upper bound converges to $0$ whenever the sum of step-sizes converges to $0$. 
+This rate can be arbitrarily fast because the oracle is strong.
+Indeed, taking an inifinite step-size solves the problem with one iteration, but the update is (theoretically) harder to solve for larger step-sizes.
+
+\paragraph{Stochastic Setting.}
+This proof does not transpose immediately to the stochastic setting.
+First off, we do not have a decreasing sequence anymore
+\begin{align}
+	f(\natp_{t+1}, X_t)  \leq f(\natp_{t}, X_t)  
+	- \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
+	\implies \expect{f(\natp_{t+1}, X_t)} \leq f(\natp_{t}) 
+	- \inv{\lr_t}\expect{\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}))}  \; .
+\end{align}
+Indeed the inequality bears on $\expect{f(\natp_{t+1}, X_t)}$, which features a correlation between $\natp_{t+1}$ and $X_t$, rather than $f(\natp_{t+1})$.
+Similarly, applying the three points lemma with $\natp^*$, taking the tower expectation, and telescoping the sum, we get
+\begin{align}
+	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_t) - f(\natp^*)}
+	\leq \cB( \natp^*; \natp_{0})  \; .
+\end{align}
+To pursue further, we need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
+The most straightforward such assumption is 
+\begin{align}
+\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t})} \leq \lr_t \sigma^2 \; ,
+\label{eq:proximal_variance}
+\end{align}
+e.g. we relate the stochastic function value on the next iterate to the function value.
+Then, using convexity of $f$, we get the rate
+\begin{align}
+	 \expect{f(\frac{\sum_{t=1}^{T} \lr_t \natp_{t+1}}{\sum_{t=1}^{T} \lr_t } )} - f(\natp^*)
+	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
+	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
+\end{align}
+which is very analogous to standard rates for SGD.
+It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
+When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{t^{\half[3]}})$.
+But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
+In addition the convergence bears on average iterates, which is far from perfect.
+
+\paragraph{Application to the MAP.}
+The variance assumption \eqref{eq:proximal_variance} translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in the exponential family MAP setting -- e.g.
+exactly the same assumption as in \citet{hanzely2018fastest}.
 However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which is an average in the primal space, and gives more weight to the first iterates.

### 2021-09-06 main.tex
Updates from june

@@
 
+\subsection{Triangle Scaling Exponents}
+
+\TODO{cite accelerated Bregman gradient paper} 
+defines the Triangle Scaling Exponent, to characterize an acceleration Bregman methods. The exponent $\gamma$ of a Bregman divergence $\cB$ is defined as the largest coefficient such that $
+	\forall x_0, x_1, x_2, \forall \lambda\in[0,1],$
+\begin{align}
+	\cB((1-\lambda)x_0 + \lambda x_1 ; (1-\lambda)x_0 + \lambda x_2)
+	\leq \lambda^\gamma \cB(x_1 ; x_2) \; .
+\end{align}
+Geometrically, for any triangle, it tells us a shape within which the geodesic triangle is contained (to clarify with a picture.)
+It is related to Thales triangle.
+We precisely have a Thales triangle with the dual expectation pivot, meaning we can relate the MAP variance term with the MLE expected suboptimality. Writing $\lambda = \frac{n}{n+n_0}$, we observe
+\begin{align}
+	\MAPm^\text{MAP} &= (1-\lambda)\mu_0 + \lambda \MAPm^\text{MLE} \\
+	\mu_n &= (1-\lambda)\mu_0 + \lambda \mu^*
+\end{align}
+thus
+\begin{align}
+	\expect{\bregmanconj(\mu_n; \MAPm^\text{MAP})}
+	\leq \left (\frac{n}{n+n_0} \right)^\gamma \expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})} \; .
+\end{align}
+This relationship between the MAP variance and the MLE may be void, as the MLE expected loss may be infinite. But in some settings it might be helpful to characterize the MAP convergence.
+For instance, this same paper says that for a gaussian variance, $\gamma=1$.
+
 \clearpage
@@
 	\item Stochastic Bregman Proximal Point with learning rate $\inv{n_0+n}$,
-	\item Stochastic Mirror Sescent with learning rate  $\inv{n_0+n +1}$.
+	\item Stochastic Mirror Descent with learning rate  $\inv{n_0+n +1}$.
 \end{itemize}
@@
 \paragraph{Goal.} We want to bound the suboptimality $f(\natp) - f^*$ as a function of the number of samples.
+
+\paragraph{\TODO{Discussion with Radu}}
+PROBLEM. Even in the Euclidean setting, the decreasing step-size analysis with SGD (or SPP) gives a rate $O(1/\log(n))$, which is far from the asymptotic $O(1/n)$.
+
+If we change the step-size, for $1/\sqrt{n}$ for instance, the algorithm changes.
+It becomes a weighted average of the sufficient statistics.
+It might be helpful for online learning (changing target), but it is not relevant when the target is fixed.
 
@@
     &= \argmin_\natp f(\natp; X_{n}) + (n+n_0) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
+    \label{eq:SBPP}
 \end{align} 
@@
 e.g. we relate the stochastic function value on the next iterate to the function value.
-Then, using convexity of $f$, we get the rate
-\begin{align}
-	 \expect{f(\frac{\sum_{t=1}^{T} \lr_t \natp_{t+1}}{\sum_{t=1}^{T} \lr_t } )} - f(\natp^*)
+Then, taking the iterate with the least expected loss, we get the rate
+\begin{align}
+	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
 	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
@@
 which is very analogous to standard rates for SGD.
-It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
+It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
 When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{t^{\half[3]}})$.
 But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
-In addition the convergence bears on average iterates, which is far from perfect.
 
@@
 \subsection{MAP as SMD}
-Let $f(\natp) := \expect[x]{-\log p(x|\natp)} = - \langle \mu_*, \natp\rangle + \logpart(\natp)$.
-In words, $f$ is linear modification of a convex function $\logpart$, which we can access only through noisy estimates of $\mu_*$.
-It turns out that the MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpart$.
-First let's recall mirror descent iteration
+In this section, we show that MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpart$.
+
+\paragraph{Mirror Descent.}
+Let us recall mirror descent iteration
 \begin{align}
@@
 which shows exponential convergence and 1-step convergence when $\lr =1$.
-Back to our sheep, the MAP iteration can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1})$ a stochastic estimate  of $\nabla f(\natp_t)$
+
+\paragraph{Stochastic Mirror Descent}
+Let $f(\natp, x) = -\log p(x|\natp)$.
+MAP iterations can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1}) = \logpart(\natp_t) - T(x_{t+1})$ as a stochastic estimate  of $\nabla f(\natp_t)$.
+Resuming from the Stochastic Bregman Proximal Point formula~\eqref{eq:SBPP} and manipulating the Bregman yields
 \begin{align}
@@
 This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
-
-\subsection{Relative Smoothness for Mirror Descent}
-In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpart$ \citep{bubeck2015convex}.
-In our setting this is not always true -- eg gaussians.
-However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity.
-\begin{align}
-	& \text{$f$ is $L$-smooth relative to $h$} \\
-	\iff & Lh - f \ \text{convex} \\
-	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y;x), \forall x,y \\
-	\iff & \cB_f(y;x) \leq L \cB_h(y;x), \forall x,y \\
-	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
-\end{align} 
-where the last equivalence holds only when $f$ and $h$ are twice differentiable.
-In words, $f$ is upper bounded by it linear approximation plus the $h$-Bregman divergence, which can also be seen as a bound between divergences, or more locally as a bound between Hessians.
-Similarly, $f$ is $\mu$ strongly-convex relative to $h$ if
-\begin{align}
-	& f-\mu h \ \text{ convex} \\
-	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y;x) \leq f(y)\forall x,y \\
-	\iff & \mu\cB_h(y;x) \leq \cB_f(y;x) \forall x,y \\
-	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
-\end{align}
-
-Another way to view this elegant generalization of smoothness and strong-convexity is as a transfer of the Loewner partial order on symmetric matrices to functions, via the Hessian. 
-As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
-As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
-More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality of the sequence 
-\begin{align}
-	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x ; x_t)
-\end{align}
-is upper bounded by the simple formula
-\begin{align}
-	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* ; x_0) }{t} \; .
-\end{align}
-These notions were rediscovered and expanded by \citet{bauschke2017descent} and \citet{lu2018relatively}.
-If you need to read one, pick \citet{lu2018relatively} -- I found it much much easier and more enjoyable to read.
-This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.
 
@@
 
-% todo: idee: adapter la preuve de stochastic gradient descent a stochastic mirror descent.
-Exploiter un parralele entre ces deux preuves pour obtenir une preuve pour stochastic mirror descent dans le cas ou f est relatively smooth w.r.t.
-A*.
-ce qu'on veut, c'est une hypothese analogue a expected smoothness (see paper of nico loizou) pour le relatively smooth.
-% SGD: General Analysis and Improved Rates gower et al -- section 2 and 3 see equation (9) in particular (expected smoothnesS)
-% Relatively-Smooth Convex Optimization by First-Order Methods, and Applications,  -- Section 3
-
 \subsection{Convergence of Mirror Descent}
@@
 This is because they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.
+\TODO{Sanity check literature search with Bach's references.}
+
+\TODO{3 threads of literature search: optimization, statistics, learning theory.
+This last one is not covered yet.
+It is more interested in finite sample results than statistics.
+Contact Alexander Rakhlin}
 
@@
 \begin{itemize}
-	\item Weakly supervised : missing part of $x_i$ for each $i$.
-	\item Structured prediction : graphical model on $X$ or other kind of structure.
+	\item A picture / geometric understanding of the equality between ELBO (Jensen) and relative smoothness for EM.
+	\item Bregman is about general geometry, can we bring it to the non-convex realm via local analysis ?
+	\item Can we design an algorithm with higher order Bregman divergences ?
+Does it mean anything ?
+The Bregman is a quadratic approximation of a function, yielding a geometry.
+Higher order does not carry such meaning.
+But we would like to involve some quadratic function information within mirror descent !
 \end{itemize}
@@
 \end{proof}
+
+\section{Relative Smoothness for Mirror Descent}
+In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpart$ \citep{bubeck2015convex}.
+In our setting this is not always true -- eg gaussians.
+However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity.
+\begin{align}
+	& \text{$f$ is $L$-smooth relative to $h$} \\
+	\iff & Lh - f \ \text{convex} \\
+	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y;x), \forall x,y \\
+	\iff & \cB_f(y;x) \leq L \cB_h(y;x), \forall x,y \\
+	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
+\end{align} 
+where the last equivalence holds only when $f$ and $h$ are twice differentiable.
+In words, $f$ is upper bounded by it linear approximation plus the $h$-Bregman divergence, which can also be seen as a bound between divergences, or more locally as a bound between Hessians.
+Similarly, $f$ is $\mu$ strongly-convex relative to $h$ if
+\begin{align}
+	& f-\mu h \ \text{ convex} \\
+	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y;x) \leq f(y)\forall x,y \\
+	\iff & \mu\cB_h(y;x) \leq \cB_f(y;x) \forall x,y \\
+	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
+\end{align}
+
+Another way to view this elegant generalization of smoothness and strong-convexity is as a transfer of the Loewner partial order on symmetric matrices to functions, via the Hessian. 
+As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
+As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
+More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality of the sequence 
+\begin{align}
+	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x ; x_t)
+\end{align}
+is upper bounded by the simple formula
+\begin{align}
+	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* ; x_0) }{t} \; .
+\end{align}
+These notions were rediscovered and expanded by \citet{bauschke2017descent} and \citet{lu2018relatively}.
+If you need to read one, pick \citet{lu2018relatively} -- I found it much much easier and more enjoyable to read.
+This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.
 

### 2021-09-06 main.tex
Updates from Overleaf

--- 2021-06-22 main.tex
@@
 
+\clearpage
+
+\section{Tentative with self-concordant function}
+
+We want to have a rate of convergence on
+\[
+    \mathbb{E}_{X_i} \left[ \mathcal{B}_{A^*}(\mu_*,\mu_n) \right],
+\]
+where
+\[
+    \mu_* = \mathbb{E}_{X} T(X), \quad \mu_n =  \frac{\sum_{i=1}^n T(X_i) + n_0 \mu_0}{n_0+n}.
+\]
+We first assume that $A^*$ is self-concordant.
+
+\subsection{Notations}
+\begin{itemize}
+    \item $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$
+    \item $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$
+    \item $\lambda(x) = \|(A^*)'(x)\|_x^*$ (Newton decrements)
+    \item $\omega = $
+    \item $\omega^* = -t-\ln(1-t)$
+\end{itemize}
+Note that $\omega^*$ is convex and monotonne.
+
+\begin{proposition}
+    (Conversion of norms) We have
+    \[
+        \frac{\|y-x\|_x}{1+\|y-x\|_x} \leq \|y-x\|_y \leq \frac{\|y-x\|_x}{1-\|y-x\|_x}
+    \]
+\end{proposition}
+
+\begin{proposition}
+    (Bounded Hessian change) We have
+    \[
+        (1-\|y-x\|_x)^2 \nabla^2 A^*(x) \leq \nabla^2 A^*(y) \leq \frac{1}{(1-\|y-x\|_x)^2} \nabla^2 A^*(x)
+    \]
+\end{proposition}
+
+\begin{proposition}
+    (Function bound) We have, if $\|y-x\|_x < 1$,
+    \[
+        A^*(y) \leq A^*(x) + \nabla A^*(x)(y-x) + \omega^*(\|y-x\|_x)
+    \]
+\end{proposition}
+
+The last proposition implies that, when $\|y-x\|_x < 1$,
+\[
+    \mathcal{B}_{A^*}(y,x) \leq \omega^*(\|y-x\|_x).
+\]
+Therefore,
+\[
+    \mathcal{B}_{A^*}(\mu^*,\mu_n) \leq \omega^*(\|\mu^*-\mu_n\|_{\mu_n}) .
+\]
+then, if $\|\mu^*-\mu_n\|_{\mu^*} < 1$, we have
+\[
+    \mathcal{B}_{A^*}(\mu^*,\mu_n) \leq \omega^*\left(\frac{\|\mu^*-\mu_n\|_{\mu^*}}{1-\|\mu^*-\mu_n\|_{\mu^*}}\right) .
+\]
+Assume $\|\mu^*-\mu_n\|_{\mu^*} < c$,
+\[
+    \mathcal{B}_{A^*}(\mu^*,\mu_n) \leq \omega^*\left(\frac{\|\mu^*-\mu_n\|_{\mu^*}}{1-c}\right) .
+\]
+
+\subsection{not good}
+However,
+\begin{align*}
+    \|\mu^*-\mu_n\|^2_{\mu_n} & = \left\|\frac{n+n_0-1}{n+n_0}(\mu^*-\mu_{n-1}) + \frac{1}{n+n_0}(\mu^*-T_n)\right\|^2_{\mu_n},\\
+    & = \left\|\frac{n+n_0-1}{n+n_0}(\mu^*-\mu_{n-1})\right\|_{\mu_n}^2 +  \left\|\frac{1}{n+n_0}(\mu^*-T_n)\right\|^2_{\mu_n} + \frac{2(n+n_0+1)}{(n+n_0)^2}\langle \nabla^2 [A(\mu_n)](\mu^*-\mu_{n-1}),\,(\mu^*-T_n) \rangle,\\
+\end{align*}
+and
+\begin{align*}
+    \|\mu^*-\mu_n\|_{\mu_n} & \leq  \frac{n+n_0-1}{n+n_0}\left\|\mu^*-\mu_{n-1}\right\|_{\mu_n} + \frac{1}{n+n_0}\left\|\mu^*-T_n\right\|_{\mu_n} .
+\end{align*}
+Moreover, by the bounded Hessian change,
+\[
+    \left\|\mu^*-\mu_{n-1}\right\|_{\mu_n} \leq \frac{1}{1-\left\|\mu_n-\mu_{n-1}\right\|_{\mu_{n-1}}}\left\|\mu^*-\mu_{n-1}\right\|_{\mu_{n-1}}
+\]
+
 \end{document}

### 2021-09-10 main.tex
Add discussion of the entropic barrier paper

@@
 \begin{lemma}[Expected MAP Natural Parameter]
-	For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
-	\begin{align}
-		\frac{\mu^*}{\mu_n}
+	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+	\begin{align}
+		\frac{\mu^*}{\mu_n} = \frac{n_0+n}{a+n}
 		\leq \expect{\frac{\mu^*}{\MAPm}} 
 		= \expect{\frac{\MAPt}{\natp^*}} 
-		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
+		\leq \frac{n_0+n}{a+n-2} \; .
 	\label{eq:theta_expectation_bound}
@@
 \begin{theorem}[MAP Bound]
- Introducing $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
+ Defining $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
  \begin{equation}
@@
 		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
-		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
+		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{\frac{a+n-2}{n_0}(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
 	\end{cases}
@@
 It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
-When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{t^{\half[3]}})$.
+When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{\sqrt{t}})$.
 But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
@@
 MAP iterations can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1}) = \logpart(\natp_t) - T(x_{t+1})$ as a stochastic estimate  of $\nabla f(\natp_t)$.
-Resuming from the Stochastic Bregman Proximal Point formula~\eqref{eq:SBPP} and manipulating the Bregman yields
+Resuming from the Stochastic Bregman Proximal Point formula~\eqref{eq:SBPP}, we write $\MAPt$ for the n-th MAP iterate, and do some Bregman manipulations
 \begin{align}
@@
 
+\paragraph{Questions.}
+Is there already an SC analysis of proximal point methods. (that’s not affine invariant)?
+If the log-partition is SC, then what does it mean for its gradient, the mapping from natural to mean parameters ?
+
 \section{Related Work}
@@
 
-\citet{ostrovskii2021finite}  shows a convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant.
-This convergence rate holds only after a given number of samples have been seen.
-This is because they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.
-\TODO{Sanity check literature search with Bach's references.}
+\citet{bubeck2015entropic} shows that the entropy of an exponential family defined on a compact support $\cX$ is a self-concordant barrier on this compact set.
+They extend the definition of self-concordance to $\nu$-self-concordance, stating that the gradient should be bounded by the square root of the hessian along any direction.
+In 1 dimension, their result also means that the log-partition is self-concordant as well.
+In general, does this mean that the log-partition is self-concordant  (e.g.
+is there a link between the dual function on a line and the line of the primal , probably not)?
+
+In the realm of self-concordance, \citet{ostrovskii2021finite}  shows a finite sample convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant.
+This convergence rate holds only after a given number of samples have been seen, and they give an innteresting formula for this number.
+To get this results, they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.
 

### 2021-09-10 main.tex
Add Bregman proximal point references

@@
 \paragraph{Bregman Proximal Point}
-The analysis of deterministic Bregman proximal \citep{chen1993convergence} point relies on the three points lemma (Appendix~\ref{app:3points})
+This algorithm was first published in \citet{eckstein1993nonlinear} with an asymptotic convergence property.
+\citet{chen1993convergence} then provided a finite sample convergence rate for deterministic Bregman proximal point.
+Their analysis relies on the three points lemma (Appendix~\ref{app:3points})
 \begin{align}
@@
 \paragraph{Stochastic Setting.}
-This proof does not transpose immediately to the stochastic setting.
+While \citet{eckstein1998approximate} shows some asymptotic convergence of the Bregman proximal point algorithm when each iteration is only solved approximately, we are not aware of published analysis when iterations are stochastic.
+The deterministic proof does not transpose immediately to the stochastic setting.
 First off, we do not have a decreasing sequence anymore
@@
 \label{app:3points}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics{figs/3pointslemma}
+	\caption{Illustration of the other three points lemma -- eg the identity between Bregmans -- reproduced from \citet{eckstein1998approximate}.}
+	\label{fig:3pointslemma}
+\end{figure}
 

### 2021-09-15 main.tex
Add discussion of french musketeers

@@
 
-\TODO{cite accelerated Bregman gradient paper} 
+\citet{hanzely2021accelerated} 
 defines the Triangle Scaling Exponent, to characterize an acceleration Bregman methods. The exponent $\gamma$ of a Bregman divergence $\cB$ is defined as the largest coefficient such that $
@@
 For instance, this same paper says that for a gaussian variance, $\gamma=1$.
+Then they proceed to define an intrinsic scaling exponent, with a local scaling factor.
+This intrinsic exponent is always 2 for Bregman divergences induced by a convex and $\cC^2$ function.
+They use this fact to derive an accelerated algorithm.
 
@@
 
-\subsection{Relative smoothness for SMD}
-Now our setting is Stochastic Mirror Descent (SMD), meaning at each step we observe a random unbiased estimate gradient.
-This setting was studied by \citet{hanzely2018fastest}, who proved in the smooth strongly-convex case with tail averaging :  with constant step-size, linear convergence down to a variance ball, and with step-size $\gamma_t = n_0 + t$ a rate $\tilde O(\inv{t})$.
+Stochastic Mirror Descent (SMD) was studied under relative smoothness assumptions in 2 papers : \citet{hanzely2018fastest} and \citet{dragomir2021fast}.
+
+\subsection{Work by Hanzely}
+
+\citet{hanzely2018fastest} got convergence rates in the smooth strongly-convex case with tail averaging.
+With constant step-size, they proved linear convergence down to a variance ball.
+With step-size $\gamma_t = n_0 + t$, they proved a rate $\tilde O(\inv{t})$.
 These results match the rates for standard SGD.
-
-This is very interesting to us.
 Let us have a look at their variance assumption \citep[Assumption 5.1]{hanzely2018fastest}.
@@
 \begin{align}
-	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t}) \leq \lr_t \sigma^2 \\
+	\lr_t \sigma^2 \geq 
+	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
 	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
-	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
-\end{align}
-where $\gamma_t$ is the step-size and expectations are conditional on the past. Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD.
-
-\subsection{Transferring SMD's covariance assumption to MAP estimates}
+	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}\\
+	 &= \expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }
+\end{align}
+where $\gamma_t$ is the step-size, expectations are conditional on the past, and $\bar \natp_{t+1}$ is the theoretical output of a deterministic Bregman gradient step.
+Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD.
+
 In our setting, 
@@
 	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
-	\approx \lr (1-\lr)^{-2} \left (\frac{\sigma_*}{\sigma_0} \right)^4
-\end{align}
-using a first order approximation of the inverse, eg a first order approximation of $\nabla \conj$ that holds when $\lr X^2 \ll (1-\lr) \sigma_0^2$.
-This does not hold for $\sigma_0\ll 1$, which is precisely when the covariance  explodes, as illustrated in Figure \TODO{}.
-This is a problem, because we need this bound to hold  for all values of $\sigma_0$ in the support of $X^2$, meaning on the whole $[0,+\infty[$, which is impossible because $1/X^2$ diverges, so we need to process this integral differently.
+\end{align}
+When $\sigma_0 \rightarrow 0$, as might happen, this value explodes.
 \end{example}
+
+\subsection{Work by the French Musketeers}
+
+\citet{dragomir2021fast} use a quite different assumption, bearing on a Bregman measure of the stochastic gradients norm at the optimum
+\begin{align}
+	2 \lr^2 \sigma^2
+	\geq \expect[X]{\bregmanconj(\mu_t - 2 \lr_t \nabla f(\natp^*, X) ; \mu_t ) } \; .
+	\label{eq:dragomir_hypothesis}
+\end{align}
+When $\logpart$ is strongly convex, this hypothesis is akin to bounding the variance of the gradient at the optimum, whereas \eqref{eq:hanzely_hypothesis} is akin to bounding gradient variance everywhere.
+This comparison makes this assumption more interesting, as it might apply to unbounded strongly convex function, like modern analysis of SGD.
+
+However it does use a metric which depends on the trajectory $\mu_t$.
+It is the flaw of both these variance assumptions \eqref{eq:hanzely_hypothesis} and \eqref{eq:dragomir_hypothesis}, as well the proximal point variance assumption \eqref{eq:proximal_variance}, that they depend on the stochastic trajectory.
+Indeed this trajectory has a non-zero probability of being arbitrarily close from the border of the domain, where $\bregman$ may get extremely ill-conditioned.
+
+\subsection{Transferring SMD's covariance assumption to MAP estimates}
 

### 2021-09-22 main.tex
Precise variance n=1 case

@@
 \begin{lemma}[Expected MAP Natural Parameter]
-	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
-	\begin{align}
-		\frac{\mu^*}{\mu_n} = \frac{n_0+n}{a+n}
+	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+	\begin{equation}
+		\frac{\mu^*}{\mu_n}
 		\leq \expect{\frac{\mu^*}{\MAPm}} 
 		= \expect{\frac{\MAPt}{\natp^*}} 
-		\leq \frac{n_0+n}{a+n-2} \; .
+		\leq \frac{n_0 +n}{a+ (n-2)_+}
 		\label{eq:theta_expectation_bound}
-	\end{align}
+	\end{equation}
+	where $(x)_+ = \max(0,x)$.
 \end{lemma}
@@
 \begin{theorem}[MAP Bound]
- Defining $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
+ Let us  define
+ \begin{align}
+	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
+ \end{align}
+The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
  \begin{equation}
@@
 	\leq \begin{cases}
-		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
-		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
-		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{\frac{a+n-2}{n_0}(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
+		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
+		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
 	\end{cases}
@@
 \end{example}
+This inequality highlights a clear variance-bias decomposition.
+In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
+For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
+Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$.
+We are basically losing a factor 2 compared to $1/2n$.
+
 Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
@@
 \begin{example}
-\begin{corollary}[MLE Loose Bound]
+\begin{corollary}[MLE Loose Bound] When $n\geq 2$, the MLE expected loss is upper bounded as
 \begin{equation}
@@
 \end{figure}
+
+\TODO{Tonight ! add equations for the full gaussian. I can dig them ou of my previous notebook.}
+
+\TODO{add equations for other families, eg gaussian with mean and variance, or gamma or Dirichlet, or Categorical..}
 
@@
 	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
+	\label{eq:primal_pivot}
 \end{align}
@@
 \begin{lemma}
+\label{lem:covariance}
 Taking  $\MAPt$ and $\MAPm$ as the primal and dual parameters of the MAP estimate, and $\tilde \mu_n$ the primal expectation and $\mu_n$ the dual expectation, the following identities hold
@@
 	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
-	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
-	= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
+	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle} \\
+	&= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
+	\label{eq:covariance_expansion}
 \end{align}
@@
 Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$.
-Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
+Note that finding $\natp_*$ is done with 1 step of mirror descent with step-size 1.
+Indeed plugging in definitions of $f$ and $\mu$, and assuming constant step-size $\lr$ yields
 \begin{align}
@@
 
-Stochastic Mirror Descent (SMD) was studied under relative smoothness assumptions in 2 papers : \citet{hanzely2018fastest} and \citet{dragomir2021fast}.
+We know of 2 papers that study Stochastic Mirror Descent (SMD) under relative smoothness assumptions -- \citet{hanzely2018fastest} and \citet{dragomir2021fast}. We are now going to review these works.
 
@@
 
+\paragraph{General Approach.} 
 \citet{hanzely2018fastest} got convergence rates in the smooth strongly-convex case with tail averaging.
@@
 Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
-Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps
+Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the next iterate $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$ that should hold for all time steps
 \begin{align}
 	\lr_t \sigma^2 \geq 
-	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
-	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
-	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}\\
-	 &= \expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }
-\end{align}
-where $\gamma_t$ is the step-size, expectations are conditional on the past, and $\bar \natp_{t+1}$ is the theoretical output of a deterministic Bregman gradient step.
+	&\Cov(- g_t,\natp_{t+1} | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
+	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
+\end{align}
+where $\gamma_t$ is the step-size, and expectations are conditional on all of the past. 
 Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD. 
-
+If we introduce $\bar \natp_{t+1}=\nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t  \nabla f(\natp_t) )$ the theoretical output of a deterministic Bregman gradient step, then following Lemma~\ref{lem:covariance} we can rewrite this covariance as the expectation of symmetrized Bregman between stochastic and deterministic iterates $\expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }$.
+This view is useful to understand that we are measuring the variance of each iteration, which is necessarily positive.
+
+\paragraph{MAP Special Case.}
 In our setting, 
@@
 \end{align}
-so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
-\begin{align}
-	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
-	\leq \lr_t\sigma^2 \\
+so that $\sigma$ is really a bound on the covariance of the sufficient statistics $T(X_{t+1})$ with the next iterate $\natp_{t+1}$, which itself depends on $T(X_{t+1})$ via a coefficient $\lr_t$ and a non-linearity $\nabla\conj$.
+\begin{align}
+	&\Cov( T(X_{t+1}), \natp_{t+1} | X_{1\dots t})
+	\leq \lr_t\sigma^2 
+	\label{eq:covariance_assumption} \\
 	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
 \end{align}
-In other words, we need for all $\lr\in(0,\inv{n})$, and a certain subset of $\mu$,
+In other words, we need for all $\lr_t \in [0; \lr_0]$, and the subset of $\mu$ that can be reached by on the trajectory $\mu_t$,
 \begin{align}
@@
 \begin{example}
-Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
-The covariance  \eqref{eq:hypvariance} we want to bound is equal to
-\begin{align}
-	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
-\end{align}
+Unfortunately this bound is not trivially satisfied, even for simple exponential family members.
+For instance, centered gaussians variance estimation $\cN(0,\mu^*)$ does not verify \eqref{eq:hypvariance}.
+Indeed the covariance  \eqref{eq:hypvariance} we want to bound is equal to
+\begin{align}
+	\half \expect[X\sim \cN(0,\mu^*)]{\frac{\mu*^2 - X^2 }{ X^2 \lr + (1-\lr) \mu^2 }}
+	= \inv{\lr} (1+a) \expect[X \sim \cN(0,1)]{\inv{a + X^2}} - \inv{\lr} 
+\end{align}
+where $a = \frac{1-\lr}{\lr} \frac{\mu}{\mu^*}$
 When $\sigma_0 \rightarrow 0$, as might happen, this value explodes.
@@
 
-\subsection{Work by the French Musketeers}
-
-\citet{dragomir2021fast} use a quite different assumption, bearing on a Bregman measure of the stochastic gradients norm at the optimum
-\begin{align}
-	2 \lr^2 \sigma^2
-	\geq \expect[X]{\bregmanconj(\mu_t - 2 \lr_t \nabla f(\natp^*, X) ; \mu_t ) } \; .
-	\label{eq:dragomir_hypothesis}
-\end{align}
-When $\logpart$ is strongly convex, this hypothesis is akin to bounding the variance of the gradient at the optimum, whereas \eqref{eq:hanzely_hypothesis} is akin to bounding gradient variance everywhere.
-This comparison makes this assumption more interesting, as it might apply to unbounded strongly convex function, like modern analysis of SGD.
-
-However it does use a metric which depends on the trajectory $\mu_t$.
-It is the flaw of both these variance assumptions \eqref{eq:hanzely_hypothesis} and \eqref{eq:dragomir_hypothesis}, as well the proximal point variance assumption \eqref{eq:proximal_variance}, that they depend on the stochastic trajectory.
-Indeed this trajectory has a non-zero probability of being arbitrarily close from the border of the domain, where $\bregman$ may get extremely ill-conditioned.
-
-\subsection{Transferring SMD's covariance assumption to MAP estimates}
-
-\paragraph{MLE Analysis.}
+\paragraph{Straightforward analysis}
 When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, we get
@@
 In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
-The mean parameter $\mu$ take values arbitrarily close from $0$, which makes the covariance explode.
-\RLP{should we get a functional bound $\sigma^2(\mu)$ instead ? or split between $\mu$ smaller or larger than $\varepsilon$.}
-
-\paragraph{MAP Analysis.}
-We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
-Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
-\begin{align}
-	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) } 
-	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
-	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
-\end{align}
-so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
-\begin{align}
-	 \bregmanconj (\mu_* ; \tilde \mu)  
-	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
-\end{align}
-where $\mSigma_* = \Cov_{\mu_*}(T(X))$. 
-The order of terms $\bregmanconj (\mu_* ; \tilde \mu)  \neq \bregmanconj (\tilde \mu ; \mu_* ) $ only affects third order terms.
-That's for the bregman, but the scalar product is harder to bound.
-Hopefully it could be positive.
-If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$.
-Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.
-
-\paragraph{Relative Continuity}
-\citet{lu2019relative} define relative continuity to tackle non-differentiable functions -- eg Hinge loss.
-These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
-Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and $\ell^2$ distance.
-Although this hypothesis is far from intuitive, it does yield convergence rates very similar to typical SGD analysis.
-In this work we are first concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
-
-\paragraph{Bound on the trajectory.}
-Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
-No we cannot because $\mu$ takes every possible values as a random variable within the expectation.
+The mean parameter $\mu$ takes values arbitrarily close from $0$, which makes the covariance explode.
+
+\paragraph{Bias-Variance Perspective.}
+As shown in equation~\eqref{eq:covariance_expansion}, covariance~\eqref{eq:covariance_assumption} is an upper bound on the variance term in either the primal \eqref{eq:primal_pivot} or dual \eqref{eq:bias-mixed-variance} expectation pivot decompositions
+\begin{align}
+	\Cov(\MAPm, \MAPt)  = \frac{n}{n+n_0} \expect[X_{1:n-1}]{\underbrace{\Cov(\MAPt ; T(X_{n}) | X_{1:n-1} )}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}}}
+	\leq  \frac{n \sigma^2}{(n+n_0)^2}
+\end{align}
+so we recover a $O(n^{-1})$ upper bound on the variance term and we are still looking for the $O(n^{-2})$ rate for the bias term.
+It is simpler to consider the primal pivot, but still hard to say anything about $\bregmanconj(\mu^* ; \tilde \mu_n) = \bregman(\E[\MAPt]; \natp^*)$ without further assumptions.
+
+\subsection{Work by the French Musketeers}
+
+\citet{dragomir2021fast} use a quite different assumption, bearing on a Bregman measure of the stochastic gradients norm at the optimum
+\begin{align}
+	2 \lr^2 \sigma^2
+	\geq \expect[X]{\bregmanconj(\mu_t - 2 \lr_t \nabla f(\natp^*, X) ; \mu_t ) } \; .
+	\label{eq:dragomir_hypothesis}
+\end{align}
+When $\logpart$ is strongly convex, this hypothesis is akin to bounding the variance of the gradient at the optimum, whereas \eqref{eq:hanzely_hypothesis} is akin to bounding gradient variance everywhere.
+This comparison makes this assumption more interesting, as it might apply to unbounded strongly convex function, like modern analysis of SGD.
+
+However it does use a metric which depends on the trajectory $\mu_t$.
+It is the flaw of both these variance assumptions \eqref{eq:hanzely_hypothesis} and \eqref{eq:dragomir_hypothesis}, as well the proximal point variance assumption \eqref{eq:proximal_variance}, that they depend on the stochastic trajectory.
+Indeed this trajectory has a non-zero probability of being arbitrarily close from the border of the domain, where $\bregman$ may get extremely ill-conditioned.
 
@@
 So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)
+
+\subsection{Discussion}
+
+\paragraph{Bound on the trajectory.}
+Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory.
+Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
+In the case of the Gaussian variance, we cannot because $\mu$ takes every possible values as a random variable within the expectation.
 
@@
 \citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
-Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
+Then they use recent mirror descent rates based on the idea of \textit{relative smoothness} \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
+
+Relative smoothness gave rise to sibling hypothesis, such as \textit{relative continuity} \citep{lu2019relative},  which was used to characterize non-differentiable functions -- eg Hinge loss.
+These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
+Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and the $\ell^2$ distance.
+Although this hypothesis is far from intuitive, \citet{lu2019relative} obtained convergence rates for mirror descent that are very similar to typical SGD analysis.
+In this work we are concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
+Consequently, we do not further investigate relative continuity.
 
@@
 
+\section{Other Exponential Families}
+\label{app:expfamily}
+
+\subsection{Full Gaussian $\cN(m, \sigma^2)$}
+Take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
+The log-density is written
+\begin{align}
+	\log p(X) 
+	&= -\frac{(X-m)^2}{2 \sigma^2} - \half \log \sigma^2 + \cst \\
+	&=  \frac{m}{\sigma^2} X -\inv{2 \sigma^2} X^2 - \frac{m^2}{2 \sigma^2} - \half \log \sigma^2 \\
+	&=  \natp_1 X + \natp_2 X^2  - \logpart(\natp) \; .
+\end{align}
+From this formula, we can identify the duality map
+\begin{align}
+	\natp_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
+	\natp_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
+\end{align}
+and we can also explicit the log-partition as a function of the natural parameters
+\begin{align}
+	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
+	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-\natp_2) \; .
+\end{align}
+We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\natp)$
+\begin{align}
+	\mu_1 &= \frac{\natp_1}{-2\natp_2}  \\
+	\mu_2 & = \left( \frac{\natp_1}{-2\natp_2} \right)^2 + \inv{-2 \natp_2} \; .
+\end{align}
+Now let us discover the entropy
+\begin{align}
+	\conj (\mu) &= \mu_1 \natp_1 + \mu_2 \natp_2  - \logpart(\natp(\mu)) \\
+	& = \frac{\mu_1^2}{\mu_2 - \mu_1^2}  - \frac{\mu_2}{2(\mu_2 - \mu_1^2)} - \frac{\mu_1^2}{2(\mu_2 - \mu_1^2)}  - \half \log(\mu_2 - \mu_1^2) \\
+	&= \half \frac{\mu_1^2 - \mu_2}{\mu_2 - \mu_1^2} - \half \log(\mu_2 - \mu_1^2) \\
+	& = - \half \log(\mu_2 - \mu_1^2) + \cst \; .
+\end{align}
+In this case, the Bregman divergences induced by the log-partition or the entropy are ugly, so we will not focus on them.
+
 \section{Three points lemma}
@@
 \subsection{Proof of Expected MAP Natural Parameter}
+Let us introduce a variable characterizing the importance of the prior
+\begin{important}
+	\begin{align}
+			a=n_0 \frac{\mu_0}{\mu^*} \; .
+	\end{align}
+\end{important}
 \begin{lemma}[Expected MAP natural parameter]
 	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
-	\begin{align}
+\begin{equation}
 		\frac{\mu^*}{\mu_n}
@@
 		= \expect{\frac{\MAPt}{\natp^*}} 
-		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
-	\end{align}
+		\leq \begin{cases}
+		\frac{n_0 +n}{a+n-2} \text{ when } n\geq 2,\\
+		\frac{n_0 +1}{a} \text{ when } n=1.
+		\end{cases}
+		\label{eq:lemma_natp_bound}
+	\end{equation}
 \end{lemma}
@@
 \begin{proof}
+	The lower bound can be readily obtained with the Jensen inequality applied on the convex function $x\mapsto\inv{x}$ for $x>0$.
+	The upper bound requires much more work.
 	To start, let us plug in the definition of $\MAPm$
@@
 	\end{align}
-	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
-\begin{important}
-	\begin{align}
-			a=n_0 \frac{\mu_0}{\mu^*} \; .
-	\end{align}
-\end{important}
+	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$.
 	We reformulate this expectation with the trick
@@
 	Fortunately, mathematicians have been working on these integrals for decades.
-	When $n=2$, eg $k=1$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
+	When $n=2$, eg $k=1$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8.E1}{verifies the bound}
 	\begin{align}
@@
 	\end{align}
-	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19\#E21}{the general bound}
+	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound}
 	\begin{align}
@@
 	where we used $\frac{a + n}{n +n_0} = \frac{\mu_n}{\mu^*}$.
+
+We are left with a special case when  $n=1$, eg $k=\half$. 
+Then we can use the trivial bound 
+\begin{align}
+	\expect{\inv{a + X^2}} < \inv{a}
+	\label{eq:trivial_bound}
+\end{align}
+which is what we are using in the lemma,
+or we can get a stronger result with further derivations.
+The integral can be written with \href{https://dlmf.nist.gov/7.8.E1}{Mill's ratio} $M(x) = e^{x^2} \int_x^\infty e^{-t^2} dt$, by doing the change of variable $t= x \sqrt{u}$, $dt=\frac{x}{2} \frac{du}{\sqrt{u}}$
+\begin{align}
+	& \int_x^\infty e^{-t^2} dt 
+	= \frac{x}{2} \int_1^\infty \frac{e^{-x^2 u}}{\sqrt{u}} du 
+	= \frac{x}{2} E_{\half}(x^2) \\
+	\implies & M(x) = \frac{x}{2} e^{x^2} E_{\half}(x^2) \\
+	\implies & e^{y} E_{\half}(y) = 2 \frac{M(\sqrt{y})}{\sqrt{y} }
+\end{align}
+so that 
+\begin{align}
+		\expect{\inv{a + X^2}} 
+		= \half e^{\half[a]} E_{\half[1]}(\half[a]) 
+		= \sqrt{\frac{2}{a}} M(\sqrt{\half[a]})
+		= \frac{2}{a} \sqrt{\half[a]} M(\sqrt{\half[a]}) \; .
+\end{align}
+Then using \href{https://dlmf.nist.gov/7.8.E8}{another bound from the DLMF}, we get that 
+\begin{align}
+	\inv{a+1}
+	\leq \expect{\inv{a + X^2}} 
+	< \inv{a}\left( 1 - \inv{a+3} \right) \; .
+\end{align}
 \end{proof}
-
-We are left with a special case when  $n=1$, eg $k=\half$.
-Then the integral can be written with the complementary error function, or \href{https://dlmf.nist.gov/7.8#E1}{Mill's ratio} $M(x) = e^{x^2} \int_x^\infty e^{-t^2} dt$.
-Indeed, taking $t= x \sqrt{u}$
-\begin{align}
-	\int_x^\infty e^{-t^2} dt = x \int_1^\infty \frac{e^{-x^2 u}}{\sqrt{u}} du = x E_{\half}(x^2) \\
-	\implies M(x) = x e^{x^2} E_{\half}(x^2)
-\end{align}
-so that 
-\begin{align}
-		\half e^{\half[a]} E_{\half[1]}(\half[a]) 
-		= \half \sqrt{\frac{2}{a}} M(\sqrt{\half[a]})
-		= \inv{a} \sqrt{\half[a]} M(\sqrt{\half[a]}) \; .
-\end{align}
-and \href{https://dlmf.nist.gov/7.8#E5}{we known that} $x M(x) < \half$ so we get 
-\begin{align}
-	\expect{\inv{a + X^2}} < \inv{2 a}
-\end{align}
-FUCK That's  trivial as  it's true for 1/a.
 
@@
 		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
-		\bregmanconj( \mu_*; \mu_0) +\inv{a} \ \text{if}\ n=1,\\
-		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} if n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
+		\inv{2(n_0+1)}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 \frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} \ \text{if}\ n=1,\\
+		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\ n\geq 2
 	\end{cases}
@@
 
-$\frac{1+n_0}{4} e^{\half[a]} \sqrt{\frac{2 \pi}{a}} \mathrm{erfc}(\sqrt{\half[a]})  - \half  + \frac{a - n_0}{2(1+n_0)} $
-
 \begin{proof}
-		Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound}. We get
+When $n=0$, the inequality is an equality. 
+Wen $n>0$, Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound},
 \begin{align}
 	2\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \expect{\frac{\mu^*}{\MAPm}}  + \expect{\frac{\MAPm}{\mu^*}} - 1 \\
-	&\leq \frac{\mu^*}{\mu_n - \frac{2 \mu^*}{n + n_0}}   + \frac{\mu_n}{\mu^*} - 1 \\
-	&= \frac{n_0 + n}{a +n - 2} + \frac{a+n}{n_0 + n} - 2 \\
-	&= \frac{n_0 - a + 2}{a +n - 2} + \frac{a - n_0}{n_0 + n} \\
-	&= \frac{2}{a+n-2} + \frac{(n_0 - a)^2}{(a+n-2)(n_0 + n)} \\
-	\implies \expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a)^2}{2(a+n-2)(n_0 + n)} \\
-	&\leq \frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} 
+	&\leq \expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1 \; .
+\end{align}
+We readily get 
+\begin{align}
+	 \expect{\frac{\MAPm}{\mu^*}} - 1 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
 \end{align}
 where recall that $a=n_0\frac{\mu_0}{\mu^*}$. 
-This inequality highlights a clear variance-bias decomposition, to compare with the MLE.
-Unfortunately it is not tight asymptotically as the inequality \eqref{eq:log_bound} is not quadratically tight around $1$.
-We are basically losing a factor 2 compared to $1/2n$.
-by~\eqref{eq:log_bound}.
+There remains the more problematic term with the expectation of the inverse mean parameter.
+Fortunately, we derived the bound \eqref{eq:lemma_natp_bound} for this purpose.
+
+\textbf{When $n\geq 2$}, we get 
+\begin{align}
+	\expect{\frac{\mu^*}{\MAPm}} - 1 
+	\leq \frac{n_0 + n}{a +n - 2} -1
+	= \frac{n_0 - a + 2}{a +n - 2}
+	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
+\end{align}
+so putting it all together we get 
+\begin{align}
+	\expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
+	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \frac{n_0 - a + 2}{(a+n-2)(n_0+n)} \\
+	&= \frac{1}{a+n-2} + \frac{(n_0 - a)^2 + 2(n_0-a) \pm 1}{2(a+n-2)(n_0+n)} \\
+	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
+	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
+	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
+\end{align}
+
+\textbf{When $n=1$} the bound~\eqref{eq:lemma_natp_bound}  on the expected natural parameter gives
+\begin{align}
+	\expect{\frac{\mu^*}{\MAPm}} - 1 
+	\leq \frac{n_0 + 1}{a} -1
+	= \frac{n_0 +1 - a}{a}
+\end{align}
+so putting it all together we get
+\begin{align}
+	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
+	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 - a + 1}{a} \\
+	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
+	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
+	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 \frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
+\end{align}
+so we do recover the same bias variance structure as when $n\geq2$.
 \end{proof}

### 2021-09-27 main.tex
Add proof of convergence for mirror descent

@@
 
-\section{Expected Relative Smoothness}
-To avoid gradient variance assumption, we can take inspiration from expected smoothness, which is a consequence of convexity and smoothness, and which yields a general analysis of SGD.
-Can we get something similar with relativity and SMD?
-The analog of gradient norm is the Bregman between iterates, similar to the Homeomorphic EM rate.
-
-\subsection{Convergence of Mirror Descent}
-To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$.
-Note that the strong convexity constant $m$ may be zero.
-Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
-\begin{align}
-	f(\natp^{t+1}) 
-	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
-	& \text{(Relative $L$-smoothness)} \\
-	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
-	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(Relative $m$-strong convexity)}
-\end{align}
-Applying the 3 points lemma with \TODO{f}
+\section{Optimization Ideas}
 
@@
 
-\subsection{Full Gaussian $\cN(m, \sigma^2)$}
-Take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
+\subsection{Full Gaussian}
+To model $\cN(m,\sigma^2)$, we take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
 The log-density is written
@@
 	&= \half \frac{\mu_1^2 - \mu_2}{\mu_2 - \mu_1^2} - \half \log(\mu_2 - \mu_1^2) \\
-	& = - \half \log(\mu_2 - \mu_1^2) + \cst \; .
+	& = - \half \log (\mu_2 - \mu_1^2) + \cst \; .
 \end{align}
@@
 This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.
+
+\subsection{Convergence of Mirror Descent}
+\paragraph{Descent Lemma.}
+To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$.
+Note that the strong convexity constant $m$ may be zero.
+Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
+\begin{align}
+	f(\natp^{t+1}) 
+	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
+	& \text{(Relative $L$-smoothness)} \\
+	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
+	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \text{(Relative $m$-strong convexity)}
+\end{align}
+
+\paragraph{Monotonicity.}
+Applying the 3 points lemma with $x=\natp_t$, 
+\begin{align}
+	f(\natp^{t+1}) \leq  f(\natp^t) -L (\bregman(\natp^{t}, \natp^{t+1}) + \bregman(\natp^{t+1} , \natp^{t} ))
+\end{align}
+shows the function value is monotonically decreasing, and we the gap at each step is greater than the symmetrized bregman between iterates.
+
+\paragraph{No Strong-Convexity.}
+The simpler case is when $m=0$, then we get a telescopic sum
+\begin{align}
+	\sum_{t=0}^{T-1} (f(\natp^{t+1}) - f(\natp^*)) 
+	&\leq L \sum_t \bregman(\natp^{*}; \natp^{t}) - \bregman(\natp^{*}); \natp^{t+1}) \\
+	&= L \bregman(\natp^{*}; \natp^{0}) - L\bregman(\natp^{*}); \natp^{T})\\
+	&\leq L \bregman(\natp^{*}; \natp^{0}) \; ,
+\end{align}
+from which we can conclude thanks to the monotonicity of $f(\natp^t)$
+\begin{align}
+	f(\natp^{T}) - f(\natp^*)
+	\leq  \frac{L}{T} \bregman(\natp^{*}; \natp^{0}) \; ,
+\end{align}
+
+\paragraph{Strong Convexity.} 
+When $m>0$, we need to do some cooking with a coefficient $\frac{L-m}{L}$ to create the telescopic sum
+\begin{align}
+	\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\natp^{t+1}) - f(\natp^*)) 
+	&\leq \sum_{t=0}^{T-1} \frac{( L-m)^{T-t}}{L^{T-t -1}} \bregman(\natp^{*}; \natp^{t}) 
+	- \frac{( L-m)^{T-t-1}}{L^{T-t-2}} \bregman(\natp^{*}); \natp^{t+1}) \\
+	&= \frac{( L-m)^{T}}{L^{T-1}}  \bregman(\natp^{*}; \natp^{0}) 
+	- L \bregman(\natp^{*}); \natp^{T}) \\
+	&\leq L \left( \frac{L-m}{L} \right)^T \bregman(\natp^{*}; \natp^{0}) \; .
+\end{align}
+Then using the monotonicity of $f(\natp^t)$, we end up with the geometric sum of $\frac{L-m}{L} < 1$
+\begin{align}
+		\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\natp^{t+1}) - f(\natp^*)) 
+		&\geq (f(\natp^{T}) - f(\natp^*)) \sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} \\
+		&\geq (f(\natp^{T}) - f(\natp^*)) \inv{1 - \frac{L-m}{L}} \\
+		&= (f(\natp^{T}) - f(\natp^*)) \frac{L}{m} \; .
+\end{align}
+Finally we end up with the final iterate linear convergence rate
+\begin{align}
+	f(\natp^{T}) - f(\natp^*)
+	\leq m  \left( 1 - \frac{m}{L} \right)^T \bregman(\natp^{*}; \natp^{0}) 
+\end{align}
+featuring $\frac{L}{m} $ the relative condition number.
 

### 2021-09-27 main.tex
Move SC to appendix and start exponent idea

@@
 
-\paragraph{\TODO{Discussion with Radu}}
-PROBLEM. Even in the Euclidean setting, the decreasing step-size analysis with SGD (or SPP) gives a rate $O(1/\log(n))$, which is far from the asymptotic $O(1/n)$.
-
-If we change the step-size, for $1/\sqrt{n}$ for instance, the algorithm changes.
-It becomes a weighted average of the sufficient statistics.
-It might be helpful for online learning (changing target), but it is not relevant when the target is fixed.
+\subsection{Other Idea}
+There are two main bottlenecks when casting MAP as an iterative algorithm. 
+\begin{enumerate}
+	\item The first one is the variance definition which always end up being infinite when asked to be uniform on every possible trajectories, if the log-partition  has some barrier structure.
+	\item The second one is the step-size scheduling.
+$O(\inv{t})$ is decreasing too fast for most proof techniques.
+Even in the Euclidean setting where the variance does not matter, it results in a disappointing rate $O(\inv{\log t})$.
+\end{enumerate}
+Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size.
+Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem.Their descent lemma looks like
+\begin{align}
+	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^2 \sigma^2
+\end{align}
+where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\natp^t$ and $\natp^*$ ; and $\lr_t \leq \inv{2L}$.
+On the other hand, the problem of most Bregman analysis is that they do not get this $\lr^2$ in front the variance.
+Instead they simply get a $\lr$, because they do not benefit from the homogeneity of the norm
+\begin{align}
+	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t \sigma^2 \; .
+\end{align}
+The problem is that this exponent $2$ is crucial to get fast convergence with a $O(\inv{t})$ step-size.
+Otherwise the variance is too large and we need to decrease the step-size more slowly to reach the optimum quickly.
+Now our question is what happens if we get something in-between ? e.g.
+\begin{align}
+	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^\alpha \sigma^2 \; .
+\end{align}
+where $1<\alpha<2$ is some exponent characterizing the geometry of the problem.
+Inspired by \citet{gower2019sgd} we set
+\begin{align}
+	\lr_t := \inv{m} \left ( 1 - \left (\frac{k}{k+1} \right)^\alpha \right ) \; .
+\end{align}
 
@@
 It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
-When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{\sqrt{t}})$.
 But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
+With the sums  $\sum \lr_t$ and $\sum \lr_t^2$, we would need a $O(\inv{\sqrt{t}})$ step-size to ensure a $O(\inv{\sqrt{t}})$ convergence rate, but then the iterate becomes a weighted average of the sufficient statistics, giving more weight to the end of the tail.
+This may be suited for online learning, but not for our stationary setting.
 
@@
 
+Their work can be seen as a Bregman generalization of \citet{gower2019sgd}.
+
 \subsection{Work by Loizou}
@@
 In the case of the Gaussian variance, we cannot because $\mu$ takes every possible values as a random variable within the expectation.
-
-\section{Optimization Ideas}
-
-\section{Self-Concordance}
-
-A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
-\begin{align}
-    \logpart(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
-\end{align}
-which is defined on $\eta \in \real^d$ and $\Lambda\in \real^{d\times d}$ symmetric positive definite. 
-It is not strongly convex, so $\conj$ is not smooth.
-
-Another hypothesis that may be more suitable is self-concordance. $f: \real \rightarrow \real$ is SC if 
-\begin{align}
-    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
-\end{align}
-The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
-A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
-Negative logarithm $-\log(x)$ and entropy $x\log(x)$ are both self-concordant function. This is good news for us since log-partition function may include SC logarithmic barriers.
-In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
-
-\paragraph{Questions.}
-Is there already an SC analysis of proximal point methods. (that’s not affine invariant)?
-If the log-partition is SC, then what does it mean for its gradient, the mapping from natural to mean parameters ?
 
@@
 
-\section{Tentative with self-concordant function}
+\section{Self-Concordance}
+An hypothesis that may be more suitable than smoothness of Lipschitzness is self-concordance. $f: \real \rightarrow \real$ is self-concordant (SC) if 
+\begin{align}
+    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
+\end{align}
+The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
+A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
+Most importantly, the negative logarithm $-\log(x)$ and its matrix generalization $-\log \det(\mX)$ (to verify) are self-concordant functions.
+This is great, but it is missing all the other barrier objectives $x^{-\alpha}, \alpha>0$. 
+For instance, gaussians have a logarithmic term, but hey also have an inverse term which is not self-concordant.
+These other barriers can fit in the framework of  generalized self-concordance \citep{dvurechensky2020selfconcordant} , but the calculus quickly gets very complex, as this generalized self-concordance is not as clean as the standard self-concordance.
+
+\paragraph{Questions.}
+Is there already an SC analysis of proximal point methods. (that’s not affine invariant)?
+If the log-partition is SC, then what does it mean for its gradient, the mapping from natural to mean parameters ?
+
+\subsection{Damien's Trials}
 
@@
 
-\subsection{Notations}
+\paragraph{Notations}
 \begin{itemize}
@@
 
-\subsection{not good}
+\paragraph{Discarded}
 However,

### 2021-09-27 main.tex
Detail the step-size with weird exponent

@@
 Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size.
-Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem.Their descent lemma looks like
+Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem, along with $K=\frac{L}{m}$ the condition number.Their descent lemma looks like
 \begin{align}
@@
 \begin{align}
-	\lr_t := \inv{m} \left ( 1 - \left (\frac{k}{k+1} \right)^\alpha \right ) \; .
-\end{align}
+	\lr_t := \inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \; .
+\end{align}
+Asymptotically, we are in the desired category
+\begin{align}
+	\lr_t := \inv{m} \left ( 1 - \left (1 - \frac{1}{t+1} \right)^\alpha \right ) 
+	\approx \inv{m} \left ( 1 - \left (1 - \frac{\alpha}{t+1} \right) \right ) 
+	\approx \frac{\alpha}{m t} \in O(\inv{t} \; .
+\end{align}
+When does it start to $\lr_t \leq \inv{2L}$ ?
+\begin{align}
+	\lr_t \leq \inv{2L}
+	&\iff 1 - \left (\frac{t}{t+1} \right)^\alpha \leq \inv{2K} \\
+	&\iff \left (1 - \inv{2K}  \right)^{\inv{\alpha}} \leq \frac{t}{t+1}\\
+	&\iff \frac{t+1}{t} \leq \left (1 - \inv{2K}  \right)^{-\inv{\alpha}}
+\end{align}
+and we know by convexity of $(1+x)^\beta$ that $	\left (1 - \inv{2K}  \right)^{-\inv{\alpha}} \geq 1 + \inv{2\alpha K}$.
+This is giving us a sufficient condition on $t$ :
+\begin{align}
+	\frac{t+1}{t}  \leq 1 + \inv{2\alpha K} 
+	&\iff t ( 1 - 1 -  \inv{2\alpha K}) + 1 \leq 0 \\
+	&\iff t \geq 2\alpha K \; .
+\end{align}
+Thus when $t < t_0 = \ceil{2\alpha K}$, we set $\lr_t = \inv{2L}$.
+
+\TODO{Questions for later : finish typing the proof of convergence. Then think about how this exponent could appear?}
 

### 2021-09-30 main.tex
Finish proof with exponent alpha

--- 2021-09-27 main.tex
@@
 where $1<\alpha<2$ is some exponent characterizing the geometry of the problem.
+
+\paragraph{Step-size.}
 Inspired by \citet{gower2019sgd} we set
 \begin{align}
-	\lr_t := \inv{m} \left ( 1 - \left (\frac{k}{k+1} \right)^\alpha \right ) \; .
-\end{align}
+	\lr_t := \inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \; .
+\end{align}
+The function $x\mapsto (1+x)^\beta$ is convex, meaning that 
+\begin{align}
+	\left (\frac{t}{t+1} \right)^\alpha 
+	= \left (1 - \frac{1}{t+1} \right)^\alpha 
+	\gtrsim 1 - \frac{\alpha}{t+1}
+\end{align}
+where the $\gtrsim$ symbol stands for greater or equal $\geq$ and asymptotically equivalent $\sim$.
+Back to the step-size, we get
+\begin{align}
+	\lr_t \lesssim \frac{\alpha}{m(t+1)} \; .
+	\label{eq:lr_asymptote}
+\end{align} 
+From this we draw two conclusions.
+First the step-size is in the right asymptotic category $\lr_t \in \Theta(t^{-1})$.
+Second, we have a sufficient condition for the step-size to be small enough $\lr_t \leq \inv{2L}$ and the descent lemma to apply
+\begin{align}
+	\lr_t \leq \inv{2L}
+	\impliedby \frac{\alpha}{m(t+1)}  \leq \inv{2L} 
+	\iff \frac{2 \alpha L}{m} \leq t +1
+	\impliedby t \geq t_0 := \ceil{2\alpha K} -1
+\end{align}
+where $K= L/m$ is the condition number of the problem.
+Consequently, when $t < t_0 :== \ceil{2\alpha K}$, we set $\lr_t = \inv{2L}$.
+Let us correct our definition to account for this revision
+\begin{equation}
+	\lr_t := \begin{cases}
+		\inv{2L} \ \text{ if } \ t< t_0 = \ceil{2\alpha K} -1, \\
+		\inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \ \text{otherwise.}
+	\end{cases}
+\end{equation}
+
+\paragraph{Telescopic Sum.}
+For $t \geq t_0$, the step-size definition means that the decreasing factor of the descent lemma is
+\begin{align}
+	1- \lr_t m = \left(\frac{t}{t+1} \right)^\alpha  \; .
+\end{align}
+Multiplying both sides of the descent lemma with $(t+1)^\alpha$ gives
+\begin{align}
+	(t+1)^\alpha B_{t+1} 
+	\leq t^\alpha B_t + ((t+1)\lr_t)^\alpha \sigma^2 \; .
+\end{align}
+Thanks to~\eqref{eq:lr_asymptote}, we know that $(t+1)\lr_t \leq \frac{\alpha}{m}$, thus the second term or the upper bound is upper bounded as
+\begin{align}
+	((t+1)\lr_t)^\alpha \sigma^2  
+	\leq \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2 \; .
+\end{align}
+By recursion (or telescopic summing), we get
+\begin{align}
+	t^\alpha B_t \leq t_0^\alpha B_{t_0} 
+	+ (t - t_0) \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2  \; .
+\end{align}
+It is possible to get the constant step-size rate for this $B_{t_0}$, but for now, we already see that
+\begin{align}
+	B_t \leq \left(\frac{t_0}{t} \right)^\alpha B_{t_0} 
+	+ \frac{t - t_0}{t^\alpha} \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2 \in \Theta(t^{1-\alpha}) 
+\end{align}
+where $-1 <1-\alpha < 0$ and this slowly decreasing factor is coming from the variance term.
 

### 2021-09-30 main.tex
Add details to procimal point proof

@@
 Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0}$.
+As a reminder this algorithm is defined with the update
+\begin{align}
+	 \hat \natp_n
+    &= \argmin_\natp \lr_n f(\natp; X_{n}) + \bregman(\natp; \hat \natp_{n-1}) \\
+    \iff \nabla \logpart(\natp_{n-1}) &= \nabla\logpart(\natp_n) + \lr_t \nabla f(\natp_n ; X_t) \; .
+\end{align}
+In words, if it converges to $\natp^*$, it can be seen as a backward mirror ascent starting from $\natp^*$.
 
@@
 
-\paragraph{Bregman Proximal Point}
-This algorithm was first published in \citet{eckstein1993nonlinear} with an asymptotic convergence property.
-\citet{chen1993convergence} then provided a finite sample convergence rate for deterministic Bregman proximal point.
-Their analysis relies on the three points lemma (Appendix~\ref{app:3points})
+\subsubsection{Deterministic Bregman Proximal Point}
+Deterministic Bregman proximal point.
+was first published in \citet{eckstein1993nonlinear} with an asymptotic convergence property, soon completed by a finite sample convergence rate in \citet{chen1993convergence}.
+This convergence rate relies on the three points lemma (Appendix~\ref{app:3points})
 \begin{align}
@@
 \end{align}
+
+\paragraph{Monotonicity of Function Values.}
 Applying it with $\natp=\natp_t$, we prove that function values are decreasing
 \begin{align}
-	f(\natp_{t+1})  \leq f(\natp_{t})  - \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \; .
-\end{align}
-Then applying it with $\natp=\natp^*$ (assuming such a minimizer exists), we obtain a telescopic sum
+	f(\natp_{t+1})  \leq f(\natp_{t})  - \inv{\lr_t} \underbrace{\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) }_{=: \cS(\natp_{t} ; \natp_{t+1})}
+\end{align}
+where we introduced $\cS$ the symmetrized Bregman divergence
+\begin{align}
+	\cS(x; y) = \cB(x ; y) + \cB(y ; x) = \langle\nabla A(x) - \nabla A(y) ; x-y \rangle \; .
+\end{align}
+
+\paragraph{Descent Lemma and Telescopic Sum.}
+Applying the three point lemma with $\natp=\natp^*$ (assuming such a minimizer exists), we obtain a telescopic sum
 \begin{align}
@@
 which proves on the way that the total path length $ \sum_{t=0}^{t-1} \cB(\natp_{t+1} ; \natp_{t}) $ is bounded.
-Using the decreasing property of $f(\natp_t)$, we get
+Using the monotonicity of $f(\natp_t)$, we get
 \begin{align}
@@
 
-\paragraph{Stochastic Setting.}
+\subsubsection{Stochastic Bregman Proximal Point.}
 While \citet{eckstein1998approximate} shows some asymptotic convergence of the Bregman proximal point algorithm when each iteration is only solved approximately, we are not aware of published analysis when iterations are stochastic.
+
+\paragraph{Variance Assumption.}
 The deterministic proof does not transpose immediately to the stochastic setting.
-First off, we do not have a decreasing sequence anymore
-\begin{align}
-	f(\natp_{t+1}, X_t)  \leq f(\natp_{t}, X_t)  
-	- \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
-	\implies \expect{f(\natp_{t+1}, X_t)} \leq f(\natp_{t}) 
-	- \inv{\lr_t}\expect{\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}))}  \; .
-\end{align}
-Indeed the inequality bears on $\expect{f(\natp_{t+1}, X_t)}$, which features a correlation between $\natp_{t+1}$ and $X_t$, rather than $f(\natp_{t+1})$.
-Similarly, applying the three points lemma with $\natp^*$, taking the tower expectation, and telescoping the sum, we get
-\begin{align}
-	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_t) - f(\natp^*)}
+We need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
+As we will see soon, the most straightforward such assumption is 
+\begin{align}
+\boxed{\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t+1})} \leq \lr_t \sigma^2 \; , }
+\label{eq:proximal_variance}
+\end{align}
+e.g. we relate the stochastic function value on the next iterate to the function value.
+Asymptotically, this assumption is equivalent to the one of~\citet{hanzely2018fastest}. 
+To see this, linearize $f(\natp_{t+1})$ and $f(\natp_{t+1} ; X_{t+1}) $ in $\natp_t$
+\begin{align}
+	\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t+1})} 
+	&\approx \expect{f(\natp_{t}) - f(\natp_{t} ; X_{t+1})} + \expect{\langle \nabla f(\natp_t) - \nabla f(\natp_t ; X_{t+1}) ; \natp_{t+1} - \natp_t \rangle } \\
+	&= 0 + \expect{\langle \nabla f(\natp_t) - \nabla f(\natp_t ; X_{t+1}) ; \natp_{t+1} \rangle } \\
+	& = \Cov(- \nabla f(\natp_t ; X_{t+1})  ; \natp_{t+1} ) \; .
+\end{align}
+
+\paragraph{Monotonicity of Expected Function Values.}
+Applying the tree point lemma with $\natp=\natp_t$ gives
+\begin{align}
+	f(\natp_{t+1}, X_{t+1})  
+	&\leq f(\natp_{t}, X_{t+1})  - \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
+	&= f(\natp_{t}, X_{t+1})  - \inv{\lr_t} \cS(\natp_{t+1} ; \natp_{t}) \\
+	\implies \expect{f(\natp_{t+1}, X_{t+1})} 
+	&\leq f(\natp_{t}) - \inv{\lr_t}\expect{\cS(\natp_{t+1} ; \natp_{t})}  \; .
+	\label{eq:SPP_monotone}
+\end{align}
+This inequality bears on $\expect{f(\natp_{t+1}, X_{t+1})}$, which features a correlation between $\natp_{t+1}$ and $X_{t+1}$, and is different from $f(\natp_{t+1})$.
+Combining~\eqref{eq:SPP_monotone} with the variance assumption gives
+\begin{align}
+	\expect{f(\natp_{t+1})} 
+	&\leq \expect{f(\natp_{t+1} ; X_{t+1})}  + \lr_t \sigma^2 \\
+	&\leq f(\natp_t) +  \lr_t \sigma^2 - \inv{\lr_t}\expect{\cS(\natp_{t+1} ; \natp_{t})}
+\end{align}
+from which we see that monotonicity is guaranteed if
+\begin{align}
+	\sigma^2 \leq \inv{\lr_t^2}\expect{\cS(\natp_{t+1} ; \natp_{t})} \; .
+\end{align}
+This is quite annoying, as our variance assumption requires $\sigma^2$ to be large, and this inequality requires $	\sigma^2 $ to be small -- the step we take should be greater than the variance.
+
+\paragraph{Descent Lemma and Telescopic Sum.}
+Similarly, applying the three points lemma with $\natp^*$, and taking te expectation with respect to $X_{t+1}$ yields
+\begin{align}
+	\lr_t(\E[f(\natp_{t+1},X_{t+1})] - f(\natp^*)) +  \E[\cB(\natp_{t+1} ; \natp_{t})] 
+	&\leq  \cB( \natp^*; \natp_{t})  - \E[\cB( \natp^*; \natp_{t+1})] \; .
+\end{align}
+
+taking the tower expectation, and telescoping the sum, we get
+\begin{align}
+	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_{t+1}) - f(\natp^*)}
 	\leq \cB( \natp^*; \natp_{0})  \; .
 \end{align}
-To pursue further, we need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
-The most straightforward such assumption is 
-\begin{align}
-\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t})} \leq \lr_t \sigma^2 \; ,
-\label{eq:proximal_variance}
-\end{align}
-e.g. we relate the stochastic function value on the next iterate to the function value.
 Then, taking the iterate with the least expected loss, we get the rate

### 2021-09-30 main.tex
Add details to exponent idea

@@
 \end{enumerate}
-Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size.
+Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} or \citet{dragomir2021fast} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size.
 Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem, along with $K=\frac{L}{m}$ the condition number.Their descent lemma looks like
@@
 where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\natp^t$ and $\natp^*$ ; and $\lr_t \leq \inv{2L}$.
-On the other hand, the problem of most Bregman analysis is that they do not get this $\lr^2$ in front the variance.
+On the other hand, the problem of most Bregman analysis \citep{hanzely2018fastest,dorazio2021stochastic} is that they do not get this $\lr^2$ in front the variance.
 Instead they simply get a $\lr$, because they do not benefit from the homogeneity of the norm
@@
 \paragraph{Descent Lemma and Telescopic Sum.}
-Similarly, applying the three points lemma with $\natp^*$, and taking te expectation with respect to $X_{t+1}$ yields
-\begin{align}
-	\lr_t(\E[f(\natp_{t+1},X_{t+1})] - f(\natp^*)) +  \E[\cB(\natp_{t+1} ; \natp_{t})] 
-	&\leq  \cB( \natp^*; \natp_{t})  - \E[\cB( \natp^*; \natp_{t+1})] \; .
-\end{align}
-
-taking the tower expectation, and telescoping the sum, we get
+Similarly, applying the three points lemma with $\natp^*$ yields
+\begin{align}
+	\lr_t(f(\natp_{t+1},X_{t+1}) - f(\natp^*)) +  \cB(\natp_{t+1} ; \natp_{t})
+	&\leq  \cB( \natp^*; \natp_{t})  - \cB( \natp^*; \natp_{t+1}) \; .
+\end{align}
+Ignoring the distance between iterates $\cB(\natp_{t+1} ; \natp_{t})
+$, taking the tower expectation, and telescoping the sum, we get
 \begin{align}
@@
 exactly the same assumption as in \citet{hanzely2018fastest}.
-However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which is an average in the primal space, and gives more weight to the first iterates.
-It does not immediately apply to the last iterate.
 
@@
 Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$.
-So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)
+So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$.
+As mentioned earlier, though, this assumption does not convert to a $O(t^{-1})$ rate when the step-size is itself $O(t^{-1})$.
 

### 2021-09-30 main.tex
Add 3 points lemma figure
Updates from Overleaf

--- 2021-09-30 main.tex
@@
 \begin{lemma}[Expected MAP Natural Parameter]
-	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
-	\begin{equation}
-		\frac{\mu^*}{\mu_n}
+	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+	\begin{align}
+		\frac{\mu^*}{\mu_n} = \frac{n_0+n}{a+n}
 		\leq \expect{\frac{\mu^*}{\MAPm}} 
 		= \expect{\frac{\MAPt}{\natp^*}} 
-		\leq \frac{n_0 +n}{a+ (n-2)_+}
+		\leq \frac{n_0+n}{a+n-2} \; .
 	\label{eq:theta_expectation_bound}
-	\end{equation}
-	where $(x)_+ = \max(0,x)$.
+	\end{align}
 \end{lemma}
@@
 \begin{theorem}[MAP Bound]
- Let us  define
- \begin{align}
-	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
- \end{align}
-The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
+ Defining $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
  \begin{equation}
@@
 	\leq \begin{cases}
-		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
-		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
+		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
+		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
+		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{\frac{a+n-2}{n_0}(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
 	\end{cases}
@@
 \end{example}
-This inequality highlights a clear variance-bias decomposition.
-In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
-For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
-Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$.
-We are basically losing a factor 2 compared to $1/2n$.
-
 Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
@@
 \begin{example}
-\begin{corollary}[MLE Loose Bound] When $n\geq 2$, the MLE expected loss is upper bounded as
+\begin{corollary}[MLE Loose Bound]
 \begin{equation}
@@
 \end{figure}
-
-\TODO{Tonight ! add equations for the full gaussian. I can dig them ou of my previous notebook.}
-
-\TODO{add equations for other families, eg gaussian with mean and variance, or gamma or Dirichlet, or Categorical..}
 
@@
 	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
-	\label{eq:primal_pivot}
 \end{align}
@@
 \begin{lemma}
-\label{lem:covariance}
 Taking  $\MAPt$ and $\MAPm$ as the primal and dual parameters of the MAP estimate, and $\tilde \mu_n$ the primal expectation and $\mu_n$ the dual expectation, the following identities hold
@@
 	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
-	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle} \\
-	&= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
-	\label{eq:covariance_expansion}
+	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
+	= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
 \end{align}
@@
 
-\subsection{Other Idea}
-There are two main bottlenecks when casting MAP as an iterative algorithm. 
-\begin{enumerate}
-	\item The first one is the variance definition which always end up being infinite when asked to be uniform on every possible trajectories, if the log-partition  has some barrier structure.
-	\item The second one is the step-size scheduling.
-$O(\inv{t})$ is decreasing too fast for most proof techniques.
-Even in the Euclidean setting where the variance does not matter, it results in a disappointing rate $O(\inv{\log t})$.
-\end{enumerate}
-Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size.
-Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem, along with $K=\frac{L}{m}$ the condition number.Their descent lemma looks like
-\begin{align}
-	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^2 \sigma^2
-\end{align}
-where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\natp^t$ and $\natp^*$ ; and $\lr_t \leq \inv{2L}$.
-On the other hand, the problem of most Bregman analysis is that they do not get this $\lr^2$ in front the variance.
-Instead they simply get a $\lr$, because they do not benefit from the homogeneity of the norm
-\begin{align}
-	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t \sigma^2 \; .
-\end{align}
-The problem is that this exponent $2$ is crucial to get fast convergence with a $O(\inv{t})$ step-size.
-Otherwise the variance is too large and we need to decrease the step-size more slowly to reach the optimum quickly.
-Now our question is what happens if we get something in-between ? e.g.
-\begin{align}
-	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^\alpha \sigma^2 \; .
-\end{align}
-where $1<\alpha<2$ is some exponent characterizing the geometry of the problem.
-
-\paragraph{Step-size.}
-Inspired by \citet{gower2019sgd} we set
-\begin{align}
-	\lr_t := \inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \; .
-\end{align}
-The function $x\mapsto (1+x)^\beta$ is convex, meaning that 
-\begin{align}
-	\left (\frac{t}{t+1} \right)^\alpha 
-	= \left (1 - \frac{1}{t+1} \right)^\alpha 
-	\gtrsim 1 - \frac{\alpha}{t+1}
-\end{align}
-where the $\gtrsim$ symbol stands for greater or equal $\geq$ and asymptotically equivalent $\sim$.
-Back to the step-size, we get
-\begin{align}
-	\lr_t \lesssim \frac{\alpha}{m(t+1)} \; .
-	\label{eq:lr_asymptote}
-\end{align} 
-From this we draw two conclusions.
-First the step-size is in the right asymptotic category $\lr_t \in \Theta(t^{-1})$.
-Second, we have a sufficient condition for the step-size to be small enough $\lr_t \leq \inv{2L}$ and the descent lemma to apply
-\begin{align}
-	\lr_t \leq \inv{2L}
-	\impliedby \frac{\alpha}{m(t+1)}  \leq \inv{2L} 
-	\iff \frac{2 \alpha L}{m} \leq t +1
-	\impliedby t \geq t_0 := \ceil{2\alpha K} -1
-\end{align}
-where $K= L/m$ is the condition number of the problem.
-Consequently, when $t < t_0 :== \ceil{2\alpha K}$, we set $\lr_t = \inv{2L}$.
-Let us correct our definition to account for this revision
-\begin{equation}
-	\lr_t := \begin{cases}
-		\inv{2L} \ \text{ if } \ t< t_0 = \ceil{2\alpha K} -1, \\
-		\inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \ \text{otherwise.}
-	\end{cases}
-\end{equation}
-
-\paragraph{Telescopic Sum.}
-For $t \geq t_0$, the step-size definition means that the decreasing factor of the descent lemma is
-\begin{align}
-	1- \lr_t m = \left(\frac{t}{t+1} \right)^\alpha  \; .
-\end{align}
-Multiplying both sides of the descent lemma with $(t+1)^\alpha$ gives
-\begin{align}
-	(t+1)^\alpha B_{t+1} 
-	\leq t^\alpha B_t + ((t+1)\lr_t)^\alpha \sigma^2 \; .
-\end{align}
-Thanks to~\eqref{eq:lr_asymptote}, we know that $(t+1)\lr_t \leq \frac{\alpha}{m}$, thus the second term or the upper bound is upper bounded as
-\begin{align}
-	((t+1)\lr_t)^\alpha \sigma^2  
-	\leq \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2 \; .
-\end{align}
-By recursion (or telescopic summing), we get
-\begin{align}
-	t^\alpha B_t \leq t_0^\alpha B_{t_0} 
-	+ (t - t_0) \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2  \; .
-\end{align}
-It is possible to get the constant step-size rate for this $B_{t_0}$, but for now, we already see that
-\begin{align}
-	B_t \leq \left(\frac{t_0}{t} \right)^\alpha B_{t_0} 
-	+ \frac{t - t_0}{t^\alpha} \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2 \in \Theta(t^{1-\alpha}) 
-\end{align}
-where $-1 <1-\alpha < 0$ and this slowly decreasing factor is coming from the variance term.
+\paragraph{\TODO{Discussion with Radu}}
+PROBLEM. Even in the Euclidean setting, the decreasing step-size analysis with SGD (or SPP) gives a rate $O(1/\log(n))$, which is far from the asymptotic $O(1/n)$.
+
+If we change the step-size, for $1/\sqrt{n}$ for instance, the algorithm changes.
+It becomes a weighted average of the sufficient statistics.
+It might be helpful for online learning (changing target), but it is not relevant when the target is fixed.
 
@@
 It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
+When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{\sqrt{t}})$.
 But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
-With the sums  $\sum \lr_t$ and $\sum \lr_t^2$, we would need a $O(\inv{\sqrt{t}})$ step-size to ensure a $O(\inv{\sqrt{t}})$ convergence rate, but then the iterate becomes a weighted average of the sufficient statistics, giving more weight to the end of the tail.
-This may be suited for online learning, but not for our stationary setting.
 
@@
 Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$.
-Note that finding $\natp_*$ is done with 1 step of mirror descent with step-size 1.
-Indeed plugging in definitions of $f$ and $\mu$, and assuming constant step-size $\lr$ yields
+Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
 \begin{align}
@@
 
-We know of 2 papers that study Stochastic Mirror Descent (SMD) under relative smoothness assumptions -- \citet{hanzely2018fastest} and \citet{dragomir2021fast}. We are now going to review these works.
+Stochastic Mirror Descent (SMD) was studied under relative smoothness assumptions in 2 papers : \citet{hanzely2018fastest} and \citet{dragomir2021fast}.
 
@@
 
-\paragraph{General Approach.} 
 \citet{hanzely2018fastest} got convergence rates in the smooth strongly-convex case with tail averaging.
@@
 Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
-Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the next iterate $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$ that should hold for all time steps
+Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps
 \begin{align}
 	\lr_t \sigma^2 \geq 
-	&\Cov(- g_t,\natp_{t+1} | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
-	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
-\end{align}
-where $\gamma_t$ is the step-size, and expectations are conditional on all of the past. 
+	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
+	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
+	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}\\
+	 &= \expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }
+\end{align}
+where $\gamma_t$ is the step-size, expectations are conditional on the past, and $\bar \natp_{t+1}$ is the theoretical output of a deterministic Bregman gradient step.
 Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD.
-If we introduce $\bar \natp_{t+1}=\nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t  \nabla f(\natp_t) )$ the theoretical output of a deterministic Bregman gradient step, then following Lemma~\ref{lem:covariance} we can rewrite this covariance as the expectation of symmetrized Bregman between stochastic and deterministic iterates $\expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }$.
-This view is useful to understand that we are measuring the variance of each iteration, which is necessarily positive.
-
-\paragraph{MAP Special Case.}
+
 In our setting, 
@@
 \end{align}
-so that $\sigma$ is really a bound on the covariance of the sufficient statistics $T(X_{t+1})$ with the next iterate $\natp_{t+1}$, which itself depends on $T(X_{t+1})$ via a coefficient $\lr_t$ and a non-linearity $\nabla\conj$.
-\begin{align}
-	&\Cov( T(X_{t+1}), \natp_{t+1} | X_{1\dots t})
-	\leq \lr_t\sigma^2 
-	\label{eq:covariance_assumption} \\
+so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
+\begin{align}
+	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
+	\leq \lr_t\sigma^2 \\
 	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
 \end{align}
-In other words, we need for all $\lr_t \in [0; \lr_0]$, and the subset of $\mu$ that can be reached by on the trajectory $\mu_t$,
+In other words, we need for all $\lr\in(0,\inv{n})$, and a certain subset of $\mu$,
 \begin{align}
@@
 \begin{example}
-Unfortunately this bound is not trivially satisfied, even for simple exponential family members.
-For instance, centered gaussians variance estimation $\cN(0,\mu^*)$ does not verify \eqref{eq:hypvariance}.
-Indeed the covariance  \eqref{eq:hypvariance} we want to bound is equal to
-\begin{align}
-	\half \expect[X\sim \cN(0,\mu^*)]{\frac{\mu*^2 - X^2 }{ X^2 \lr + (1-\lr) \mu^2 }}
-	= \inv{\lr} (1+a) \expect[X \sim \cN(0,1)]{\inv{a + X^2}} - \inv{\lr} 
-\end{align}
-where $a = \frac{1-\lr}{\lr} \frac{\mu}{\mu^*}$
+Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
+The covariance  \eqref{eq:hypvariance} we want to bound is equal to
+\begin{align}
+	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
+\end{align}
 When $\sigma_0 \rightarrow 0$, as might happen, this value explodes.
@@
 
-\paragraph{Straightforward analysis}
+\subsection{Work by the French Musketeers}
+
+\citet{dragomir2021fast} use a quite different assumption, bearing on a Bregman measure of the stochastic gradients norm at the optimum
+\begin{align}
+	2 \lr^2 \sigma^2
+	\geq \expect[X]{\bregmanconj(\mu_t - 2 \lr_t \nabla f(\natp^*, X) ; \mu_t ) } \; .
+	\label{eq:dragomir_hypothesis}
+\end{align}
+When $\logpart$ is strongly convex, this hypothesis is akin to bounding the variance of the gradient at the optimum, whereas \eqref{eq:hanzely_hypothesis} is akin to bounding gradient variance everywhere.
+This comparison makes this assumption more interesting, as it might apply to unbounded strongly convex function, like modern analysis of SGD.
+
+However it does use a metric which depends on the trajectory $\mu_t$.
+It is the flaw of both these variance assumptions \eqref{eq:hanzely_hypothesis} and \eqref{eq:dragomir_hypothesis}, as well the proximal point variance assumption \eqref{eq:proximal_variance}, that they depend on the stochastic trajectory.
+Indeed this trajectory has a non-zero probability of being arbitrarily close from the border of the domain, where $\bregman$ may get extremely ill-conditioned.
+
+\subsection{Transferring SMD's covariance assumption to MAP estimates}
+
+\paragraph{MLE Analysis.}
 When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, we get
@@
 In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
-The mean parameter $\mu$ takes values arbitrarily close from $0$, which makes the covariance explode.
-
-\paragraph{Bias-Variance Perspective.}
-As shown in equation~\eqref{eq:covariance_expansion}, covariance~\eqref{eq:covariance_assumption} is an upper bound on the variance term in either the primal \eqref{eq:primal_pivot} or dual \eqref{eq:bias-mixed-variance} expectation pivot decompositions
-\begin{align}
-	\Cov(\MAPm, \MAPt)  = \frac{n}{n+n_0} \expect[X_{1:n-1}]{\underbrace{\Cov(\MAPt ; T(X_{n}) | X_{1:n-1} )}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}}}
-	\leq  \frac{n \sigma^2}{(n+n_0)^2}
-\end{align}
-so we recover a $O(n^{-1})$ upper bound on the variance term and we are still looking for the $O(n^{-2})$ rate for the bias term.
-It is simpler to consider the primal pivot, but still hard to say anything about $\bregmanconj(\mu^* ; \tilde \mu_n) = \bregman(\E[\MAPt]; \natp^*)$ without further assumptions.
-
-\subsection{Work by the French Musketeers}
-
-\citet{dragomir2021fast} use a quite different assumption, bearing on a Bregman measure of the stochastic gradients norm at the optimum
-\begin{align}
-	2 \lr^2 \sigma^2
-	\geq \expect[X]{\bregmanconj(\mu_t - 2 \lr_t \nabla f(\natp^*, X) ; \mu_t ) } \; .
-	\label{eq:dragomir_hypothesis}
-\end{align}
-When $\logpart$ is strongly convex, this hypothesis is akin to bounding the variance of the gradient at the optimum, whereas \eqref{eq:hanzely_hypothesis} is akin to bounding gradient variance everywhere.
-This comparison makes this assumption more interesting, as it might apply to unbounded strongly convex function, like modern analysis of SGD.
-
-However it does use a metric which depends on the trajectory $\mu_t$.
-It is the flaw of both these variance assumptions \eqref{eq:hanzely_hypothesis} and \eqref{eq:dragomir_hypothesis}, as well the proximal point variance assumption \eqref{eq:proximal_variance}, that they depend on the stochastic trajectory.
-Indeed this trajectory has a non-zero probability of being arbitrarily close from the border of the domain, where $\bregman$ may get extremely ill-conditioned.
-
-Their work can be seen as a Bregman generalization of \citet{gower2019sgd}.
+The mean parameter $\mu$ take values arbitrarily close from $0$, which makes the covariance explode.
+\RLP{should we get a functional bound $\sigma^2(\mu)$ instead ? or split between $\mu$ smaller or larger than $\varepsilon$.}
+
+\paragraph{MAP Analysis.}
+We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
+Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
+\begin{align}
+	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) } 
+	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
+	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
+\end{align}
+so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
+\begin{align}
+	 \bregmanconj (\mu_* ; \tilde \mu)  
+	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
+\end{align}
+where $\mSigma_* = \Cov_{\mu_*}(T(X))$. 
+The order of terms $\bregmanconj (\mu_* ; \tilde \mu)  \neq \bregmanconj (\tilde \mu ; \mu_* ) $ only affects third order terms.
+That's for the bregman, but the scalar product is harder to bound.
+Hopefully it could be positive.
+If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$.
+Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.
+
+\paragraph{Relative Continuity}
+\citet{lu2019relative} define relative continuity to tackle non-differentiable functions -- eg Hinge loss.
+These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
+Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and $\ell^2$ distance.
+Although this hypothesis is far from intuitive, it does yield convergence rates very similar to typical SGD analysis.
+In this work we are first concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
+
+\paragraph{Bound on the trajectory.}
+Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
+No we cannot because $\mu$ takes every possible values as a random variable within the expectation.
 
@@
 
-\subsection{Discussion}
-
-\paragraph{Bound on the trajectory.}
-Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory.
-Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
-In the case of the Gaussian variance, we cannot because $\mu$ takes every possible values as a random variable within the expectation.
+\section{Expected Relative Smoothness}
+To avoid gradient variance assumption, we can take inspiration from expected smoothness, which is a consequence of convexity and smoothness, and which yields a general analysis of SGD.
+Can we get something similar with relativity and SMD?
+The analog of gradient norm is the Bregman between iterates, similar to the Homeomorphic EM rate.
+
+\subsection{Convergence of Mirror Descent}
+To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$.
+Note that the strong convexity constant $m$ may be zero.
+Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
+\begin{align}
+	f(\natp^{t+1}) 
+	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
+	& \text{(Relative $L$-smoothness)} \\
+	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
+	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \text{(Relative $m$-strong convexity)}
+\end{align}
+Applying the 3 points lemma with \TODO{f}
+
+\section{Self-Concordance}
+
+A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
+\begin{align}
+    \logpart(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
+\end{align}
+which is defined on $\eta \in \real^d$ and $\Lambda\in \real^{d\times d}$ symmetric positive definite. 
+It is not strongly convex, so $\conj$ is not smooth.
+
+Another hypothesis that may be more suitable is self-concordance. $f: \real \rightarrow \real$ is SC if 
+\begin{align}
+    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
+\end{align}
+The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
+A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
+Negative logarithm $-\log(x)$ and entropy $x\log(x)$ are both self-concordant function. This is good news for us since log-partition function may include SC logarithmic barriers.
+In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
+
+\paragraph{Questions.}
+Is there already an SC analysis of proximal point methods. (that’s not affine invariant)?
+If the log-partition is SC, then what does it mean for its gradient, the mapping from natural to mean parameters ?
 
@@
 \citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
-Then they use recent mirror descent rates based on the idea of \textit{relative smoothness} \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
-
-Relative smoothness gave rise to sibling hypothesis, such as \textit{relative continuity} \citep{lu2019relative},  which was used to characterize non-differentiable functions -- eg Hinge loss.
-These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
-Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and the $\ell^2$ distance.
-Although this hypothesis is far from intuitive, \citet{lu2019relative} obtained convergence rates for mirror descent that are very similar to typical SGD analysis.
-In this work we are concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
-Consequently, we do not further investigate relative continuity.
+Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
 
@@
 
-\section{Other Exponential Families}
-\label{app:expfamily}
-
-\subsection{Full Gaussian}
-To model $\cN(m,\sigma^2)$, we take the sufficient statistic $T(X) = (X, X^2)$. Then the mean parameter is $\mu = \E[T(X)] = ( m , m^2 + \sigma^2) = (\mu_1, \mu_2) $.
-The log-density is written
-\begin{align}
-	\log p(X) 
-	&= -\frac{(X-m)^2}{2 \sigma^2} - \half \log \sigma^2 + \cst \\
-	&=  \frac{m}{\sigma^2} X -\inv{2 \sigma^2} X^2 - \frac{m^2}{2 \sigma^2} - \half \log \sigma^2 \\
-	&=  \natp_1 X + \natp_2 X^2  - \logpart(\natp) \; .
-\end{align}
-From this formula, we can identify the duality map
-\begin{align}
-	\natp_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
-	\natp_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
-\end{align}
-and we can also explicit the log-partition as a function of the natural parameters
-\begin{align}
-	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
-	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-\natp_2) \; .
-\end{align}
-We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\natp)$
-\begin{align}
-	\mu_1 &= \frac{\natp_1}{-2\natp_2}  \\
-	\mu_2 & = \left( \frac{\natp_1}{-2\natp_2} \right)^2 + \inv{-2 \natp_2} \; .
-\end{align}
-Now let us discover the entropy
-\begin{align}
-	\conj (\mu) &= \mu_1 \natp_1 + \mu_2 \natp_2  - \logpart(\natp(\mu)) \\
-	& = \frac{\mu_1^2}{\mu_2 - \mu_1^2}  - \frac{\mu_2}{2(\mu_2 - \mu_1^2)} - \frac{\mu_1^2}{2(\mu_2 - \mu_1^2)}  - \half \log(\mu_2 - \mu_1^2) \\
-	&= \half \frac{\mu_1^2 - \mu_2}{\mu_2 - \mu_1^2} - \half \log(\mu_2 - \mu_1^2) \\
-	& = - \half \log (\mu_2 - \mu_1^2) + \cst \; .
-\end{align}
-In this case, the Bregman divergences induced by the log-partition or the entropy are ugly, so we will not focus on them.
-
 \section{Three points lemma}
@@
 This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.
-
-\subsection{Convergence of Mirror Descent}
-\paragraph{Descent Lemma.}
-To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$.
-Note that the strong convexity constant $m$ may be zero.
-Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
-\begin{align}
-	f(\natp^{t+1}) 
-	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
-	& \text{(Relative $L$-smoothness)} \\
-	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
-	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(Relative $m$-strong convexity)}
-\end{align}
-
-\paragraph{Monotonicity.}
-Applying the 3 points lemma with $x=\natp_t$, 
-\begin{align}
-	f(\natp^{t+1}) \leq  f(\natp^t) -L (\bregman(\natp^{t}, \natp^{t+1}) + \bregman(\natp^{t+1} , \natp^{t} ))
-\end{align}
-shows the function value is monotonically decreasing, and we the gap at each step is greater than the symmetrized bregman between iterates.
-
-\paragraph{No Strong-Convexity.}
-The simpler case is when $m=0$, then we get a telescopic sum
-\begin{align}
-	\sum_{t=0}^{T-1} (f(\natp^{t+1}) - f(\natp^*)) 
-	&\leq L \sum_t \bregman(\natp^{*}; \natp^{t}) - \bregman(\natp^{*}); \natp^{t+1}) \\
-	&= L \bregman(\natp^{*}; \natp^{0}) - L\bregman(\natp^{*}); \natp^{T})\\
-	&\leq L \bregman(\natp^{*}; \natp^{0}) \; ,
-\end{align}
-from which we can conclude thanks to the monotonicity of $f(\natp^t)$
-\begin{align}
-	f(\natp^{T}) - f(\natp^*)
-	\leq  \frac{L}{T} \bregman(\natp^{*}; \natp^{0}) \; ,
-\end{align}
-
-\paragraph{Strong Convexity.} 
-When $m>0$, we need to do some cooking with a coefficient $\frac{L-m}{L}$ to create the telescopic sum
-\begin{align}
-	\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\natp^{t+1}) - f(\natp^*)) 
-	&\leq \sum_{t=0}^{T-1} \frac{( L-m)^{T-t}}{L^{T-t -1}} \bregman(\natp^{*}; \natp^{t}) 
-	- \frac{( L-m)^{T-t-1}}{L^{T-t-2}} \bregman(\natp^{*}); \natp^{t+1}) \\
-	&= \frac{( L-m)^{T}}{L^{T-1}}  \bregman(\natp^{*}; \natp^{0}) 
-	- L \bregman(\natp^{*}); \natp^{T}) \\
-	&\leq L \left( \frac{L-m}{L} \right)^T \bregman(\natp^{*}; \natp^{0}) \; .
-\end{align}
-Then using the monotonicity of $f(\natp^t)$, we end up with the geometric sum of $\frac{L-m}{L} < 1$
-\begin{align}
-		\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\natp^{t+1}) - f(\natp^*)) 
-		&\geq (f(\natp^{T}) - f(\natp^*)) \sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} \\
-		&\geq (f(\natp^{T}) - f(\natp^*)) \inv{1 - \frac{L-m}{L}} \\
-		&= (f(\natp^{T}) - f(\natp^*)) \frac{L}{m} \; .
-\end{align}
-Finally we end up with the final iterate linear convergence rate
-\begin{align}
-	f(\natp^{T}) - f(\natp^*)
-	\leq m  \left( 1 - \frac{m}{L} \right)^T \bregman(\natp^{*}; \natp^{0}) 
-\end{align}
-featuring $\frac{L}{m} $ the relative condition number.
 
@@
 \subsection{Proof of Expected MAP Natural Parameter}
-Let us introduce a variable characterizing the importance of the prior
-\begin{important}
-	\begin{align}
-			a=n_0 \frac{\mu_0}{\mu^*} \; .
-	\end{align}
-\end{important}
 \begin{lemma}[Expected MAP natural parameter]
 	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
-\begin{equation}
+	\begin{align}
 		\frac{\mu^*}{\mu_n}
@@
 		= \expect{\frac{\MAPt}{\natp^*}} 
-		\leq \begin{cases}
-		\frac{n_0 +n}{a+n-2} \text{ when } n\geq 2,\\
-		\frac{n_0 +1}{a} \text{ when } n=1.
-		\end{cases}
-		\label{eq:lemma_natp_bound}
-	\end{equation}
+		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
+	\end{align}
 \end{lemma}
@@
 \begin{proof}
-	The lower bound can be readily obtained with the Jensen inequality applied on the convex function $x\mapsto\inv{x}$ for $x>0$.
-	The upper bound requires much more work.
 	To start, let us plug in the definition of $\MAPm$
@@
 	\end{align}
-	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$.
+	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
+\begin{important}
+	\begin{align}
+			a=n_0 \frac{\mu_0}{\mu^*} \; .
+	\end{align}
+\end{important}
 	We reformulate this expectation with the trick
@@
 	Fortunately, mathematicians have been working on these integrals for decades.
-	When $n=2$, eg $k=1$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8.E1}{verifies the bound}
+	When $n=2$, eg $k=1$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
 	\begin{align}
@@
 	\end{align}
-	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound}
+	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19\#E21}{the general bound}
 	\begin{align}
@@
 	where we used $\frac{a + n}{n +n_0} = \frac{\mu_n}{\mu^*}$.
+\end{proof}
 
 We are left with a special case when  $n=1$, eg $k=\half$.
-Then we can use the trivial bound 
-\begin{align}
-	\expect{\inv{a + X^2}} < \inv{a}
-	\label{eq:trivial_bound}
-\end{align}
-which is what we are using in the lemma,
-or we can get a stronger result with further derivations.
-The integral can be written with \href{https://dlmf.nist.gov/7.8.E1}{Mill's ratio} $M(x) = e^{x^2} \int_x^\infty e^{-t^2} dt$, by doing the change of variable $t= x \sqrt{u}$, $dt=\frac{x}{2} \frac{du}{\sqrt{u}}$
-\begin{align}
-	& \int_x^\infty e^{-t^2} dt 
-	= \frac{x}{2} \int_1^\infty \frac{e^{-x^2 u}}{\sqrt{u}} du 
-	= \frac{x}{2} E_{\half}(x^2) \\
-	\implies & M(x) = \frac{x}{2} e^{x^2} E_{\half}(x^2) \\
-	\implies & e^{y} E_{\half}(y) = 2 \frac{M(\sqrt{y})}{\sqrt{y} }
+Then the integral can be written with the complementary error function, or \href{https://dlmf.nist.gov/7.8#E1}{Mill's ratio} $M(x) = e^{x^2} \int_x^\infty e^{-t^2} dt$.
+Indeed, taking $t= x \sqrt{u}$
+\begin{align}
+	\int_x^\infty e^{-t^2} dt = x \int_1^\infty \frac{e^{-x^2 u}}{\sqrt{u}} du = x E_{\half}(x^2) \\
+	\implies M(x) = x e^{x^2} E_{\half}(x^2)
 \end{align}
@@
 \begin{align}
-		\expect{\inv{a + X^2}} 
-		= \half e^{\half[a]} E_{\half[1]}(\half[a]) 
-		= \sqrt{\frac{2}{a}} M(\sqrt{\half[a]})
-		= \frac{2}{a} \sqrt{\half[a]} M(\sqrt{\half[a]}) \; .
-\end{align}
-Then using \href{https://dlmf.nist.gov/7.8.E8}{another bound from the DLMF}, we get that 
-\begin{align}
-	\inv{a+1}
-	\leq \expect{\inv{a + X^2}} 
-	< \inv{a}\left( 1 - \inv{a+3} \right) \; .
-\end{align}
-\end{proof}
+		\half e^{\half[a]} E_{\half[1]}(\half[a]) 
+		= \half \sqrt{\frac{2}{a}} M(\sqrt{\half[a]})
+		= \inv{a} \sqrt{\half[a]} M(\sqrt{\half[a]}) \; .
+\end{align}
+and \href{https://dlmf.nist.gov/7.8#E5}{we known that} $x M(x) < \half$ so we get 
+\begin{align}
+	\expect{\inv{a + X^2}} < \inv{2 a}
+\end{align}
+FUCK That's  trivial as  it's true for 1/a.
 
@@
 		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
-		\inv{2(n_0+1)}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 \frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} \ \text{if}\ n=1,\\
-		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \ \text{if}\ n\geq 2
+		\bregmanconj( \mu_*; \mu_0) +\inv{a} \ \text{if}\ n=1,\\
+		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} if n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
 	\end{cases}
@@
 
+$\frac{1+n_0}{4} e^{\half[a]} \sqrt{\frac{2 \pi}{a}} \mathrm{erfc}(\sqrt{\half[a]})  - \half  + \frac{a - n_0}{2(1+n_0)} $
+
 \begin{proof}
-When $n=0$, the inequality is an equality. 
-Wen $n>0$, Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound},
+		Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound}. We get
 \begin{align}
 	2\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1 \; .
-\end{align}
-We readily get 
-\begin{align}
-	 \expect{\frac{\MAPm}{\mu^*}} - 1 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
+	&\leq \expect{\frac{\mu^*}{\MAPm}}  + \expect{\frac{\MAPm}{\mu^*}} - 1 \\
+	&\leq \frac{\mu^*}{\mu_n - \frac{2 \mu^*}{n + n_0}}   + \frac{\mu_n}{\mu^*} - 1 \\
+	&= \frac{n_0 + n}{a +n - 2} + \frac{a+n}{n_0 + n} - 2 \\
+	&= \frac{n_0 - a + 2}{a +n - 2} + \frac{a - n_0}{n_0 + n} \\
+	&= \frac{2}{a+n-2} + \frac{(n_0 - a)^2}{(a+n-2)(n_0 + n)} \\
+	\implies \expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a)^2}{2(a+n-2)(n_0 + n)} \\
+	&\leq \frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} 
 \end{align}
 where recall that $a=n_0\frac{\mu_0}{\mu^*}$. 
-There remains the more problematic term with the expectation of the inverse mean parameter.
-Fortunately, we derived the bound \eqref{eq:lemma_natp_bound} for this purpose.
-
-\textbf{When $n\geq 2$}, we get 
-\begin{align}
-	\expect{\frac{\mu^*}{\MAPm}} - 1 
-	\leq \frac{n_0 + n}{a +n - 2} -1
-	= \frac{n_0 - a + 2}{a +n - 2}
-	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
-\end{align}
-so putting it all together we get 
-\begin{align}
-	\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
-	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \frac{n_0 - a + 2}{(a+n-2)(n_0+n)} \\
-	&= \frac{1}{a+n-2} + \frac{(n_0 - a)^2 + 2(n_0-a) \pm 1}{2(a+n-2)(n_0+n)} \\
-	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
-	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
-	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
-\end{align}
-
-\textbf{When $n=1$} the bound~\eqref{eq:lemma_natp_bound}  on the expected natural parameter gives
-\begin{align}
-	\expect{\frac{\mu^*}{\MAPm}} - 1 
-	\leq \frac{n_0 + 1}{a} -1
-	= \frac{n_0 +1 - a}{a}
-\end{align}
-so putting it all together we get
-\begin{align}
-	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
-	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 - a + 1}{a} \\
-	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
-	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
-	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 \frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
-\end{align}
-so we do recover the same bias variance structure as when $n\geq2$.
+This inequality highlights a clear variance-bias decomposition, to compare with the MLE.
+Unfortunately it is not tight asymptotically as the inequality \eqref{eq:log_bound} is not quadratically tight around $1$.
+We are basically losing a factor 2 compared to $1/2n$.
+by~\eqref{eq:log_bound}.
 \end{proof}
@@
 
-\section{Self-Concordance}
-An hypothesis that may be more suitable than smoothness of Lipschitzness is self-concordance. $f: \real \rightarrow \real$ is self-concordant (SC) if 
-\begin{align}
-    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
-\end{align}
-The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
-A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
-Most importantly, the negative logarithm $-\log(x)$ and its matrix generalization $-\log \det(\mX)$ (to verify) are self-concordant functions.
-This is great, but it is missing all the other barrier objectives $x^{-\alpha}, \alpha>0$. 
-For instance, gaussians have a logarithmic term, but hey also have an inverse term which is not self-concordant.
-These other barriers can fit in the framework of  generalized self-concordance \citep{dvurechensky2020selfconcordant} , but the calculus quickly gets very complex, as this generalized self-concordance is not as clean as the standard self-concordance.
-
-\paragraph{Questions.}
-Is there already an SC analysis of proximal point methods. (that’s not affine invariant)?
-If the log-partition is SC, then what does it mean for its gradient, the mapping from natural to mean parameters ?
-
-\subsection{Damien's Trials}
+\section{Tentative with self-concordant function}
 
@@
 
-\paragraph{Notations}
+\subsection{Notations}
 \begin{itemize}
@@
     \item $\lambda(x) = \|(A^*)'(x)\|_x^*$ (Newton decrements)
-    \item $\omega = $
+    \item $\omega = t-\ln (1+t)$
     \item $\omega^* = -t-\ln(1-t)$
@@
 
-\paragraph{Discarded}
+\subsection{not good}
 However,

### 2021-09-30 main.tex
Move other idea to end of optimization part

--- 2021-09-30 main.tex
@@
 
-\TODO{Tonight ! add equations for the full gaussian. I can dig them ou of my previous notebook.}
-
-\TODO{add equations for other families, eg gaussian with mean and variance, or gamma or Dirichlet, or Categorical..}
+\TODO{add equations for other families, eg gamma or Dirichlet, or Categorical..}
 
@@
 
-\subsection{Other Idea}
-There are two main bottlenecks when casting MAP as an iterative algorithm. 
-\begin{enumerate}
-	\item The first one is the variance definition which always end up being infinite when asked to be uniform on every possible trajectories, if the log-partition  has some barrier structure.
-	\item The second one is the step-size scheduling.
-$O(\inv{t})$ is decreasing too fast for most proof techniques.
-Even in the Euclidean setting where the variance does not matter, it results in a disappointing rate $O(\inv{\log t})$.
-\end{enumerate}
-Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} or \citet{dragomir2021fast} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size.
-Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem, along with $K=\frac{L}{m}$ the condition number.Their descent lemma looks like
-\begin{align}
-	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^2 \sigma^2
-\end{align}
-where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\natp^t$ and $\natp^*$ ; and $\lr_t \leq \inv{2L}$.
-On the other hand, the problem of most Bregman analysis \citep{hanzely2018fastest,dorazio2021stochastic} is that they do not get this $\lr^2$ in front the variance.
-Instead they simply get a $\lr$, because they do not benefit from the homogeneity of the norm
-\begin{align}
-	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t \sigma^2 \; .
-\end{align}
-The problem is that this exponent $2$ is crucial to get fast convergence with a $O(\inv{t})$ step-size.
-Otherwise the variance is too large and we need to decrease the step-size more slowly to reach the optimum quickly.
-Now our question is what happens if we get something in-between ? e.g.
-\begin{align}
-	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^\alpha \sigma^2 \; .
-\end{align}
-where $1<\alpha<2$ is some exponent characterizing the geometry of the problem.
-
-\paragraph{Step-size.}
-Inspired by \citet{gower2019sgd} we set
-\begin{align}
-	\lr_t := \inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \; .
-\end{align}
-The function $x\mapsto (1+x)^\beta$ is convex, meaning that 
-\begin{align}
-	\left (\frac{t}{t+1} \right)^\alpha 
-	= \left (1 - \frac{1}{t+1} \right)^\alpha 
-	\gtrsim 1 - \frac{\alpha}{t+1}
-\end{align}
-where the $\gtrsim$ symbol stands for greater or equal $\geq$ and asymptotically equivalent $\sim$.
-Back to the step-size, we get
-\begin{align}
-	\lr_t \lesssim \frac{\alpha}{m(t+1)} \; .
-	\label{eq:lr_asymptote}
-\end{align} 
-From this we draw two conclusions.
-First the step-size is in the right asymptotic category $\lr_t \in \Theta(t^{-1})$.
-Second, we have a sufficient condition for the step-size to be small enough $\lr_t \leq \inv{2L}$ and the descent lemma to apply
-\begin{align}
-	\lr_t \leq \inv{2L}
-	\impliedby \frac{\alpha}{m(t+1)}  \leq \inv{2L} 
-	\iff \frac{2 \alpha L}{m} \leq t +1
-	\impliedby t \geq t_0 := \ceil{2\alpha K} -1
-\end{align}
-where $K= L/m$ is the condition number of the problem.
-Consequently, when $t < t_0 :== \ceil{2\alpha K}$, we set $\lr_t = \inv{2L}$.
-Let us correct our definition to account for this revision
-\begin{equation}
-	\lr_t := \begin{cases}
-		\inv{2L} \ \text{ if } \ t< t_0 = \ceil{2\alpha K} -1, \\
-		\inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \ \text{otherwise.}
-	\end{cases}
-\end{equation}
-
-\paragraph{Telescopic Sum.}
-For $t \geq t_0$, the step-size definition means that the decreasing factor of the descent lemma is
-\begin{align}
-	1- \lr_t m = \left(\frac{t}{t+1} \right)^\alpha  \; .
-\end{align}
-Multiplying both sides of the descent lemma with $(t+1)^\alpha$ gives
-\begin{align}
-	(t+1)^\alpha B_{t+1} 
-	\leq t^\alpha B_t + ((t+1)\lr_t)^\alpha \sigma^2 \; .
-\end{align}
-Thanks to~\eqref{eq:lr_asymptote}, we know that $(t+1)\lr_t \leq \frac{\alpha}{m}$, thus the second term or the upper bound is upper bounded as
-\begin{align}
-	((t+1)\lr_t)^\alpha \sigma^2  
-	\leq \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2 \; .
-\end{align}
-By recursion (or telescopic summing), we get
-\begin{align}
-	t^\alpha B_t \leq t_0^\alpha B_{t_0} 
-	+ (t - t_0) \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2  \; .
-\end{align}
-It is possible to get the constant step-size rate for this $B_{t_0}$, but for now, we already see that
-\begin{align}
-	B_t \leq \left(\frac{t_0}{t} \right)^\alpha B_{t_0} 
-	+ \frac{t - t_0}{t^\alpha} \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2 \in \Theta(t^{1-\alpha}) 
-\end{align}
-where $-1 <1-\alpha < 0$ and this slowly decreasing factor is coming from the variance term.
-
 \section{Stochastic Bregman Proximal Point}
@@
 
-\subsection{Discussion}
-
-\paragraph{Bound on the trajectory.}
-Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory.
-Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
-In the case of the Gaussian variance, we cannot because $\mu$ takes every possible values as a random variable within the expectation.
+\section{Discussion and other idea}
+There are two main bottlenecks when casting MAP as an iterative algorithm. 
+\begin{enumerate}
+	\item The first one is the variance definition which always end up being infinite when asked to be uniform on every possible trajectories, if the log-partition  has some barrier structure.
+	\item The second one is the step-size scheduling.
+$O(\inv{t})$ is decreasing too fast for most proof techniques.
+Even in the Euclidean setting where the variance does not matter, it results in a disappointing rate $O(\inv{\log t})$.
+\end{enumerate}
+Regarding this second point, it might be possible to do something comparable to \citet{gower2019sgd} or \citet{dragomir2021fast} who get a $O(\inv{t})$ rate with a $O(\inv{t})$ step-size.
+Take $m$ and $L$ as the (relative) strong-convexity and smoothness constants of the problem, along with $K=\frac{L}{m}$ the condition number.Their descent lemma looks like
+\begin{align}
+	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^2 \sigma^2
+\end{align}
+where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\natp^t$ and $\natp^*$ ; and $\lr_t \leq \inv{2L}$.
+On the other hand, the problem of most Bregman analysis \citep{hanzely2018fastest,dorazio2021stochastic} is that they do not get this $\lr^2$ in front the variance.
+Instead they simply get a $\lr$, because they do not benefit from the homogeneity of the norm
+\begin{align}
+	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t \sigma^2 \; .
+\end{align}
+The problem is that this exponent $2$ is crucial to get fast convergence with a $O(\inv{t})$ step-size.
+Otherwise the variance is too large and we need to decrease the step-size more slowly to reach the optimum quickly.
+Now our question is what happens if we get something in-between ? e.g.
+\begin{align}
+	B_{t+1} \leq ( 1 - \lr_t m) B_t + \lr_t^\alpha \sigma^2 \; .
+\end{align}
+where $1<\alpha<2$ is some exponent characterizing the geometry of the problem.
+
+\paragraph{Step-size.}
+Inspired by \citet{gower2019sgd} we set
+\begin{align}
+	\lr_t := \inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \; .
+\end{align}
+The function $x\mapsto (1+x)^\beta$ is convex, meaning that 
+\begin{align}
+	\left (\frac{t}{t+1} \right)^\alpha 
+	= \left (1 - \frac{1}{t+1} \right)^\alpha 
+	\gtrsim 1 - \frac{\alpha}{t+1}
+\end{align}
+where the $\gtrsim$ symbol stands for greater or equal $\geq$ and asymptotically equivalent $\sim$.
+Back to the step-size, we get
+\begin{align}
+	\lr_t \lesssim \frac{\alpha}{m(t+1)} \; .
+	\label{eq:lr_asymptote}
+\end{align} 
+From this we draw two conclusions.
+First the step-size is in the right asymptotic category $\lr_t \in \Theta(t^{-1})$.
+Second, we have a sufficient condition for the step-size to be small enough $\lr_t \leq \inv{2L}$ and the descent lemma to apply
+\begin{align}
+	\lr_t \leq \inv{2L}
+	\impliedby \frac{\alpha}{m(t+1)}  \leq \inv{2L} 
+	\iff \frac{2 \alpha L}{m} \leq t +1
+	\impliedby t \geq t_0 := \ceil{2\alpha K} -1
+\end{align}
+where $K= L/m$ is the condition number of the problem.
+Consequently, when $t < t_0 :== \ceil{2\alpha K}$, we set $\lr_t = \inv{2L}$.
+Let us correct our definition to account for this revision
+\begin{equation}
+	\lr_t := \begin{cases}
+		\inv{2L} \ \text{ if } \ t< t_0 = \ceil{2\alpha K} -1, \\
+		\inv{m} \left ( 1 - \left (\frac{t}{t+1} \right)^\alpha \right ) \ \text{otherwise.}
+	\end{cases}
+\end{equation}
+
+\paragraph{Telescopic Sum.}
+For $t \geq t_0$, the step-size definition means that the decreasing factor of the descent lemma is
+\begin{align}
+	1- \lr_t m = \left(\frac{t}{t+1} \right)^\alpha  \; .
+\end{align}
+Multiplying both sides of the descent lemma with $(t+1)^\alpha$ gives
+\begin{align}
+	(t+1)^\alpha B_{t+1} 
+	\leq t^\alpha B_t + ((t+1)\lr_t)^\alpha \sigma^2 \; .
+\end{align}
+Thanks to~\eqref{eq:lr_asymptote}, we know that $(t+1)\lr_t \leq \frac{\alpha}{m}$, thus the second term or the upper bound is upper bounded as
+\begin{align}
+	((t+1)\lr_t)^\alpha \sigma^2  
+	\leq \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2 \; .
+\end{align}
+By recursion (or telescopic summing), we get
+\begin{align}
+	t^\alpha B_t \leq t_0^\alpha B_{t_0} 
+	+ (t - t_0) \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2  \; .
+\end{align}
+It is possible to get the constant step-size rate for this $B_{t_0}$, but for now, we already see that
+\begin{align}
+	B_t \leq \left(\frac{t_0}{t} \right)^\alpha B_{t_0} 
+	+ \frac{t - t_0}{t^\alpha} \left(\frac{\alpha}{m} \right)^\alpha  \sigma^2 \in \Theta(t^{1-\alpha}) 
+\end{align}
+where $-1 <1-\alpha < 0$ and this slowly decreasing factor is coming from the variance term.
 
@@
 	\centering
-	\includegraphics{figs/3pointslemma}
-	\caption{Illustration of the other three points lemma -- eg the identity between Bregmans -- reproduced from \citet{eckstein1998approximate}.}
+	\includegraphics[width=.7\textwidth]{figs/3pointslemma}
+	\caption{Illustration of the other three points lemma -- eg the identity between Bregmans -- taken from \citet{eckstein1998approximate}.}
 	\label{fig:3pointslemma}

### 2021-10-01 aistats/sample_paper.tex
Add aistats style and open problem aastract

\documentclass[twoside]{article}

\usepackage{aistats2022}
% If your paper is accepted, change the options for the package
% aistats2022 as follows:
%
%\usepackage[accepted]{aistats2022}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

% If you set papersize explicitly, activate the following three lines:
%\special{papersize = 8.5in, 11in}
%\setlength{\pdfpageheight}{11in}
%\setlength{\pdfpagewidth}{8.5in}

% If you use natbib package, activate the following three lines:
%\usepackage[round]{natbib}
%\renewcommand{\bibname}{References}
%\renewcommand{\bibsection}{\subsubsection*{\bibname}}

% If you use BibTeX in apalike style, activate the following line:
%\bibliographystyle{apalike}

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Instructions for Paper Submissions to AISTATS 2022}

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]

\begin{abstract}
We raise the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate, or a conjugate maximum a posteriori for the exponential family. 
Surprisingly, we found no solution to this problem in the literature -- we are not able to tell how many samples we need to fit a gaussian within a few bits of the true distribution.
After displaying some properties and special cases of this problem, 
we show it is a special case of several optimization algorithms, but it falls out of their scopes, thus highlighting range of progress in the analysis of these algorithms.
\end{abstract}

\section{GENERAL FORMATTING INSTRUCTIONS}

The camera-ready versions of the accepted papers are 8 pages,
plus any additional pages needed for references.

Papers are in 2 columns with the overall line width of 6.75~inches (41~picas).
Each column is 3.25~inches wide (19.5~picas).  The space
between the columns is .25~inches wide (1.5~picas).  The left margin is 0.88~inches (5.28~picas).
Use 10~point type with a vertical spacing of
11~points. Please use US Letter size paper instead of A4.

Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.
Top rule is 4~points thick and bottom rule is 1~point thick.
Allow 1/4~inch space above and below title to rules.

Author descriptions are center-justified, initial caps.  The lead
author is to be listed first (left-most), and the Co-authors are set
to follow.  If up to three authors, use a single row of author
descriptions, each one center-justified, and all set side by side;
with more authors or unusually long names or institutions, use more
rows.

Use one-half line space between paragraphs, with no indent.

\section{FIRST LEVEL HEADINGS}

First level headings are all caps, flush left, bold, and in point size
12. Use one line space before the first level heading and one-half line space
after the first level heading.

\subsection{Second Level Heading}

Second level headings are initial caps, flush left, bold, and in point
size 10. Use one line space before the second level heading and one-half line
space after the second level heading.

\subsubsection{Third Level Heading}

Third level headings are flush left, initial caps, bold, and in point
size 10. Use one line space before the third level heading and one-half line
space after the third level heading.

\paragraph{Fourth Level Heading}

Fourth level headings must be flush left, initial caps, bold, and
Roman type.  Use one line space before the fourth level heading, and
place the section text immediately after the heading with no line
break, but an 11 point horizontal space.

%%%
\subsection{Citations, Figure, References}

\subsubsection{Citations in Text}

Citations within the text should include the author's last name and
year, e.g., (Cheesman, 1985). 
%Apart from including the author's last name and year, citations can follow any style, as long as the style is consistent throughout the paper.  
Be sure that the sentence reads
correctly if the citation is deleted: e.g., instead of ``As described
by (Cheesman, 1985), we first frobulate the widgets,'' write ``As
described by Cheesman (1985), we first frobulate the widgets.''

The references listed at the end of the paper can follow any style as long as it is used consistently.

%Be sure to avoid
%accidentally disclosing author identities through citations.

\subsubsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Use 8 point type for footnotes. Place the
footnotes at the bottom of the column in which their markers appear,
continuing to the next column if required. Precede the footnote
section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
long.\footnote{Sample of the second footnote.}

\subsubsection{Figures}

All artwork must be centered, neat, clean, and legible.  All lines
should be very dark for purposes of reproduction, and art work should
not be hand-drawn.  Figures may appear at the top of a column, at the
top of a page spanning multiple columns, inline within a column, or
with text wrapped around them, but the figure number and caption
always appear immediately below the figure.  Leave 2 line spaces
between the figure and the caption. The figure caption is initial caps
and each figure should be numbered consecutively.

Make sure that the figure caption does not get separated from the
figure. Leave extra white space at the bottom of the page rather than
splitting the figure and figure caption.
\begin{figure}[h]
\vspace{.3in}
\centerline{\fbox{This figure intentionally left non-blank}}
\vspace{.3in}
\caption{Sample Figure Caption}
\end{figure}

\subsubsection{Tables}

All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables.
Table number and title always appear above the table.
See Table~\ref{sample-table}.

Use one line space before the table title, one line space after the table title,
and one line space after the table. The table title must be
initial caps and each table numbered consecutively.

\begin{table}[h]
\caption{Sample Table Title} \label{sample-table}
\begin{center}
\begin{tabular}{ll}
\textbf{PART}  &\textbf{DESCRIPTION} \\
\hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{SUPPLEMENTARY MATERIAL}

If you need to include additional appendices during submission, you can include them in the supplementary material file.
You can submit a single file of additional supplementary material which may be either a pdf file (such as proof details) or a zip file for other formats/more files (such as code or videos). 
Note that reviewers are under no obligation to examine your supplementary material. 
If you have only one supplementary pdf file, please upload it as is; otherwise gather everything to the single zip file.

You must use \texttt{aistats2022.sty} as a style file for your supplementary pdf file and follow the same formatting instructions as in the main paper. 
The only difference is that it must be in a \emph{single-column} format.
You can use \texttt{supplement.tex} in our starter pack as a starting point.
Alternatively, you may append the supplementary content to the main paper and split the final PDF into two separate files.

\section{SUBMISSION INSTRUCTIONS}

To submit your paper to AISTATS 2022, please follow these instructions.

\begin{enumerate}
    \item Download \texttt{aistats2022.sty}, \texttt{fancyhdr.sty}, and \texttt{sample\_paper.tex} provided in our starter pack. 
    Please, do not modify the style files as this might result in a formatting violation.

    \item Use \texttt{sample\_paper.tex} as a starting point.
    \item Begin your document with
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage\{aistats2022\}}
    \end{flushleft}
    The \texttt{twoside} option for the class article allows the
    package \texttt{fancyhdr.sty} to include headings for even and odd
    numbered pages.
    \item When you are ready to submit the manuscript, compile the latex file to obtain the pdf file.
    \item Check that the content of your submission, \emph{excluding} references, is limited to \textbf{8 pages}. The number of pages containing references alone is not limited.
    \item Upload the PDF file along with other supplementary material files to the CMT website.
\end{enumerate}

\subsection{Camera-ready Papers}

%For the camera-ready paper, if you are using \LaTeX, please make sure
%that you follow these instructions.  
% (If you are not using \LaTeX,
%please make sure to achieve the same effect using your chosen
%typesetting package.)

If your papers are accepted, you will need to submit the camera-ready version. Please make sure that you follow these instructions:
\begin{enumerate}
    %\item Download \texttt{fancyhdr.sty} -- the
    %\texttt{aistats2022.sty} file will make use of it.
    \item Change the beginning of your document to
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage[accepted]\{aistats2022\}}
    \end{flushleft}
    The option \texttt{accepted} for the package
    \texttt{aistats2022.sty} will write a copyright notice at the end of
    the first column of the first page. This option will also print
    headings for the paper.  For the \emph{even} pages, the title of
    the paper will be used as heading and for \emph{odd} pages the
    author names will be used as heading.  If the title of the paper
    is too long or the number of authors is too large, the style will
    print a warning message as heading. If this happens additional
    commands can be used to place as headings shorter versions of the
    title and the author names. This is explained in the next point.
    \item  If you get warning messages as described above, then
    immediately after $\texttt{\textbackslash
    begin\{document\}}$, write
    \begin{flushleft}
    \texttt{\textbackslash runningtitle\{Provide here an alternative
    shorter version of the title of your paper\}}\\
    \texttt{\textbackslash runningauthor\{Provide here the surnames of
    the authors of your paper, all separated by commas\}}
    \end{flushleft}
    Note that the text that appears as argument in \texttt{\textbackslash
      runningtitle} will be printed as a heading in the \emph{even}
    pages. The text that appears as argument in \texttt{\textbackslash
      runningauthor} will be printed as a heading in the \emph{odd}
    pages.  If even the author surnames do not fit, it is acceptable
    to give a subset of author names followed by ``et al.''

    %\item Use the file sample\_paper.tex as an example.

    \item The camera-ready versions of the accepted papers are 8
      pages, plus any additional pages needed for references.

    \item If you need to include additional appendices,
      you can include them in the supplementary
      material file.

    \item Please, do not change the layout given by the above
      instructions and by the style file.

\end{enumerate}

\subsubsection*{Acknowledgements}
All acknowledgments go at the end of the paper, including thanks to reviewers who gave useful comments, to colleagues who contributed to the ideas, and to funding agencies and corporate sponsors that provided financial support. 
To preserve the anonymity, please include acknowledgments \emph{only} in the camera-ready papers.

\subsubsection*{References}

References follow the acknowledgements.  Use an unnumbered third level
heading for the references section.  Please use the same font
size for references as for the body of the paper---remember that
references do not count against your page length total.

\begin{thebibliography}{}
\setlength{\itemindent}{-\leftmargin}
\makeatletter\renewcommand{\@biblabel}[1]{}\makeatother
\bibitem{} J.~Alspector, B.~Gupta, and R.~B.~Allen (1989).
    \newblock Performance of a stochastic learning microchip.
    \newblock In D. S. Touretzky (ed.),
    \textit{Advances in Neural Information Processing Systems 1}, 748--760.
    San Mateo, Calif.: Morgan Kaufmann.

\bibitem{} F.~Rosenblatt (1962).
    \newblock \textit{Principles of Neurodynamics.}
    \newblock Washington, D.C.: Spartan Books.

\bibitem{} G.~Tesauro (1989).
    \newblock Neurogammon wins computer Olympiad.
    \newblock \textit{Neural Computation} \textbf{1}(3):321--323.
\end{thebibliography}

\end{document}

### 2021-10-02 aistats/paper.tex
Start adding first few sections to paper

--- 2021-09-30 main.tex
@@
-\documentclass{article}
+\documentclass[twoside]{article}
+\usepackage{aistats2022}
+% If your paper is accepted, change the options for the package
+% aistats2022 as follows:
+%\usepackage[accepted]{aistats2022}
 
-\usepackage[top=2cm, bottom=2cm]{geometry}
+% If you set papersize explicitly, activate the following three lines:
+%\special{papersize = 8.5in, 11in}
+%\setlength{\pdfpageheight}{11in}
+%\setlength{\pdfpagewidth}{8.5in}
+
+% If you use natbib package, activate the following three lines:
+\usepackage[round]{natbib}
+\renewcommand{\bibname}{References}
+\renewcommand{\bibsection}{\subsubsection*{\bibname}}
+
 \usepackage[utf8]{inputenc} % allow utf-8 input
@@
 \usepackage{graphicx}
+\graphicspath{{../figs/}}
 %\usepackage{subfigure}
@@
 \usepackage[dvipsnames]{xcolor}
-\usepackage{tikz}           % define for loops
 
@@
 
-\usepackage[round]{natbib}
-\renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsection*{\bibname}}
-
 \newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
@@
 
-% frames with colored background
-\definecolor{light-gray}{gray}{0.85}
-\definecolor{myblue}{RGB}{199, 235, 255}
-\usepackage[framemethod=TikZ]{mdframed} 
-\usetikzlibrary{shadows}
-% Using the pgf-blur package instead 
-% https://ctan.math.illinois.edu/graphics/pgf/contrib/pgf-blur/pgf-blur.pdf
-\usetikzlibrary{shadows.blur}
-
-\mdfdefinestyle{blurry-shadow}{
-	linewidth=0pt,
-    roundcorner=0pt,
-	shadow=true,
- 	apptotikzsetting={\tikzset{mdfshadow/.style={blur shadow={%
- 		shadow xshift=1pt,
- 		shadow yshift=-1pt,
- 		shadow scale=1.00,
- 		shadow opacity = 50,
- 		%shadow blur extra rounding,
- 		shadow blur steps = 10,
- 		shadow blur radius = 5pt,
- 		every shadow
- 	}}}},
-	linewidth=0pt,
- }
-
-\newmdenv[% box with grey background
-	style=blurry-shadow,
-    backgroundcolor=gray!15,
-]{example}
-\newmdenv[
-	style=blurry-shadow,
-	backgroundcolor=Red!30,
-]{important}
-
 % my packages
-\usepackage{math_commands}
+\usepackage{../math_commands}
 % some custom math commands
@@
 
-\title{Convergence rate of MAP estimates \\
-for the exponential family}
-\author{R\'emi Le Priol}
-\date{October 2020}
-
-%% Faster processing
-%\renewcommand\appendix{\end{document}}
-
 \begin{document}
 
-\maketitle
+% If your paper is accepted and the title of your paper is very long,
+% the style will print as headings an error message. Use the following
+% command to supply a shorter title of your paper so that it can be
+% used as headings.
+%
+%\runningtitle{I use this title instead because the last one was very long}
 
-\section{Background}
+% If your paper is accepted and the number of authors is large, the
+% style will print as headings an error message. Use the following
+% command to supply a shorter version of the authors names so that
+% they can be used as headings (for example, use only the surnames)
+%
+%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
 
-\paragraph{Motivation.} 
-We do not know general convergence rates on the KL for maximum likelihood estimates of the exponential family.
-We want the simplest one.
-We hope  to get a new result by combining tools from statistics and optimization. 
-\paragraph{Exponential Family.}
-The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model 
+\twocolumn[
+
+\aistatstitle{Looking for Convergence Rates\\ for the MAP of the Exponential Family}
+
+\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner \And  Damien Scieur \And Simon Lacoste-Julien }
+
+\aistatsaddress{ Mila \And  UBC \And SAIT SAIL \And Mila} 
+]
+
+\begin{abstract}
+We raise the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate, or a conjugate maximum a posteriori for the exponential family. 
+Surprisingly, we found no solution to this problem in the literature -- we are not able to tell how many samples we need to fit a gaussian within a few bits of the true distribution.
+After displaying some properties and special cases of this problem, 
+we show it is a special case of several optimization algorithms, but it falls out of their scopes, thus highlighting range of progress in the analysis of these algorithms.
+\end{abstract}
+
+\section{Plan}
+
+\begin{enumerate}
+	\item intro to exponential family, and density estimation
+	\item the thing we want to bound
+	\item Examples : gaussian mean and gaussian variance (+other examples, just mentioned)
+	\item Insight : Strongly convex case. (+ self-concordance that is not verified either)
+	\item insight : bias-variance decomposition
+	\item Optimization perspective : SBPP or SBG. But no analysis hold, revealing a flaw of all these techniques.
+	\item Discussion : we believe finding a convergence rate would bring new tools useful to deal with common objects such as barrier losses.
+\end{enumerate}
+
+Open questions
+\begin{itemize}
+	\item does a base measure change anything ?
+	\item is there multiple conjugate priors ?
+	\item misspecified case. Do we have a   formula ?
+	\item make sure the gaussian entropy is not SC.
+\end{itemize}
+
+\section{Introduction and Background}
+\RLP{Goal = disseminate related work to make it look nice.}
+Exponential Families are an elegant and lean way to model a wide variety of data : binary, categorical, natural numbers, positive float, long or short tailed... 
+They are literally the linear model of probabilities.
+The exponential family for data $X\in\cX$  with sufficient statistic $T$ and natural parameter $\natp$ is the model 
 \begin{equation}
@@
 \end{equation}
-where $\logpart$ is the log-partition function (aka normalization factor) 
+where $\logpart$ is the log-partition function -- i.e. the normalization factor
 \begin{align}
-    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
+    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx 
 \end{align}
-
-\begin{example}
-	These frames contain the trailing example of this paper: a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
-	The density of a centered normal variable is
-\begin{align}
-	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
-\end{align}
-Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
-Mean and natural parameters are roughly inverse of each other
-\begin{align}
-	\natp = -\inv{2 \mu}\; .
-\end{align}
-Now we can match the log-likelihood with the exponential family template to get the log-partition function.
-\begin{align}
-	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
-	= x^2 \natp - \logpart(\natp) \\
-	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
-\end{align}
-\end{example}
+where the integral stands for a sum if $x$ has discrete support. \RLP{Use base measure instead.}
+Note that an exponential family is entirely specified by its support set $\cX$ and its sufficient statistic $T$.
+This simple model encompasses both categorical distributions $\cX = \{1, \dots, k\}$ with $T(X)$ being the one-hot encoding, and multivariate normal distributions $\cX=\real, T(X)=(X, X^2)$.
 
@@
 	\centering
-	\includegraphics{figs/duality}
+	\includegraphics{duality}
 	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$.
@@
 \end{figure}
-\begin{example}
-For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
+
+\section{Open Problem}
+
+For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
+\begin{multline}
+    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
+	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
+\end{multline}
+For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
+\begin{align}
+	\KL( p(.|\natp^*) ; p(.|\natp)) 
+	= \bregman (\natp ; \natp^*)  \; .
+\end{align}
+There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
+\begin{align}
+	\bregman (\natp ; \natp^*)
+    %= \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
+    = \bregmanconj ( \meanp^* ; \meanp)
+\end{align}
+so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
+\begin{align}
+\boxed{
+	\KL( p(.|\natp^*) ; p(.|\natp))
+    = \bregman (\natp ; \natp^*)
+    = \bregmanconj ( \meanp^* ; \meanp) \; .
+}
+\end{align}
+
+The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
+\begin{align}
+	\label{eq:bregmanMLE}
+	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
+	\label{eq:bregmanMAP}
+	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
+\end{align}
+where the outer expectation is on the dataset $X_1, \dots, X_n$?
+
+\paragraph{Remark.}
+What we are looking for is really akin to concentration inequality, expressed with a Bregman divergence instead of a norm.
+A key difference though, is that the random variable $T(X)$ is connected to the metric $\logpart$.
+Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
+For instance, if we plug in $\conj(\mu)= -\log(\mu)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
+
+\paragraph{Remark 2.}
+The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$.
+Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.
+
+\section{Examples}
+\subsection{Gaussian Mean}
+
+\subsection{Gaussian Variance}
+The trailing example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
+The density of a centered normal variable is
+\begin{align}
+	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
+\end{align}
+Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
+Mean and natural parameters are roughly inverse of each other
+\begin{align}
+	\natp = -\inv{2 \mu}\; .
+\end{align}
+Now we can match the log-likelihood with the exponential family template to get the log-partition function.
+\begin{align}
+	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
+	= x^2 \natp - \logpart(\natp) \\
+	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
+\end{align}
+
+We can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
 \begin{align}
@@
 which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.
-\end{example}
 
-\paragraph{Conjugate Prior}
-\TODO{Under-parametrized conjugate prior ?}
-One conjugate prior for $p(X|\natp)$ is
-\begin{align}
-    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
-\end{align}
-where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
-\begin{align}
-    \bregman (\natp ; \natp_0)
-    & = \logpart(\natp) - \logpart(\natp_0) 
-    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
-\end{align}
-with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
-Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
-We can re-write this prior as 
-\begin{align}
-    p(\natp) \propto 
-    \exp( -n_0 \logpart (\natp) 
-    + \langle n_0 \mu_0, \natp \rangle ) \; ,
-\end{align}
-which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
-The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.
-
-\begin{example}
-In the case of $\cN(0,\sigma^2)$, the logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
+The logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
 In particular,
@@
 After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.
-\end{example}
 
-\paragraph{Maximum A Posteriori (MAP).}
-The negative log-likelihood of the prior is
-\begin{align*}
-    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
-\end{align*}
-Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
-\begin{align}
-    -\log p(\cD|\natp)p(\natp) 
-    = (n_0+n) \logpart (\natp) 
-    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
-\end{align}
-Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
-\begin{align}
-    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
-\end{align}
-such that the MAP is
-\begin{align}
-    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
-    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
-\end{align}
-When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
-\begin{align}
-	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
-\end{align}
-The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
-
-\section{Open Problem}
-For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
-\begin{align}
-    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
-	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
-\end{align}
-For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
-\begin{align}
-	\KL( p(.|\natp^*) ; p(.|\natp)) 
-	= \bregman (\natp ; \natp^*)  \; .
-\end{align}
-There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
-\begin{align}
-	\bregman (\natp ; \natp^*)
-    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
-    = \bregmanconj ( \meanp^* ; \meanp)
-\end{align}
-so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
-\begin{align}
-\boxed{
-	\KL( p(.|\natp^*) ; p(.|\natp))
-    = \bregman (\natp ; \natp^*)
-    = \bregmanconj ( \meanp^* ; \meanp) \; .
-}
-\end{align}
-\begin{important}
-The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
-\begin{align}
-	\label{eq:bregmanMLE}
-	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
-	\label{eq:bregmanMAP}
-	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
-\end{align}
-where the outer expectation is on the dataset $X_1, \dots, X_n$?
-\end{important}
-
-\paragraph{Remark.}
-What we are looking for is really akin to concentration inequality, expressed with a Bregman divergence instead of a norm.
-A key difference though, is that the random variable $T(X)$ is connected to the metric $\logpart$.
-Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
-For instance, if we plug in $\conj(\mu)= -\log(\mu)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
-
-\paragraph{Remark 2.}
-The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$.
-Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.
-
-Below we review all the attempts we made on this problem.
-
-\tableofcontents
-
-\section{Gaussian Variance Example}
-
-\begin{example}
-	For $\cN(0,\sigma^2)$, both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$.
-Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
+Both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
 \begin{align}
@@
 \end{align}
-\end{example}
+
 In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{figs/bregmandef.pdf}
-	\includegraphics[width=.4\textwidth]{figs/phi.pdf}
+	\includegraphics[width=.4\textwidth]{bregmandef.pdf}
+	\includegraphics[width=.4\textwidth]{phi.pdf}
 	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics.}
@@
 	\centering
-	\includegraphics[width=.54\textwidth]{figs/asymptote.pdf}
-	\includegraphics[width=.43\textwidth]{figs/fewsamples.pdf}
+\includegraphics[width=.48\textwidth]{asymptote.pdf}
+\includegraphics[width=.48\textwidth]{fewsamples.pdf}
 	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$.
@@
 
-\begin{example}
-	\begin{theorem}[MLE Tight Bound]
-	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
-	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
-	\begin{align}
-		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
-			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
-			\label{eq:MLE_rate}
-	\end{align}
-\end{theorem}
-\end{example}
-This upper bound is asymptotically tight.
-We illustrate its behavior against empirical data in Figure~\ref{fig:curves}.
- There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}.
-The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension.
-We do not perform these calculations for now as we want to focus on the simplest possible examples.
-
-As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
-To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
-Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
-We now present a lemma on its expected inverse.
-\begin{lemma}[Expected MAP Natural Parameter]
-	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 2$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
-	\begin{align}
-		\frac{\mu^*}{\mu_n} = \frac{n_0+n}{a+n}
-		\leq \expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\natp^*}} 
-		\leq \frac{n_0+n}{a+n-2} \; .
-	\label{eq:theta_expectation_bound}
-	\end{align}
-\end{lemma}
-Note that the lower bound is a simple consequence of the convexity of the inverse function. 
-Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP. 
-\begin{example}
-\begin{theorem}[MAP Bound]
- Defining $a=n_0 \frac {\mu_0 }{ \mu^* }$, the expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
- \begin{equation}
-	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
-	\leq \begin{cases}
-		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
-		\bregmanconj( \mu_*; \mu_0) + \inv{a} \ \text{if}\ n=1,\\
-		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{\frac{a+n-2}{n_0}(1 + \frac{n}{n_0} )} \ \text{if}\  n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
-	\end{cases}
-	\label{eq:MAP_rate}
-\end{equation}
-\end{theorem}
-\end{example}
-Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
-We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
-\begin{example}
-\begin{corollary}[MLE Loose Bound]
-\begin{equation}
-	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
-	\leq \inv{n-2} \; .
-	\label{eq:MLE_loose_rate}
-\end{equation}
-\end{corollary}
-\end{example}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.7\textwidth]{figs/MAP_rates/MAP_rate_n0=5.pdf}
-	\caption{
-	Convergence rate of MAP~\eqref{eq:MAP_rate} 
-	\label{fig:MAP_rate} against the ratio between initialization and optimum $ \frac{\mu_0}{\mu^*}$. 
-	We also report the tight convergence rate of MLE~\eqref{eq:MLE_rate} (dash-dotted horizontal lines), and the loose upper bound~\eqref{eq:MLE_loose_rate} (horizontal lines). 
-	We report these curves for various values of $n$, and $n_0=5$. 
-	We observe that the rate of MAP is significantly better (lower) than the rate of MLE, only when we guessed $\mu_0$ right, within an order of magnitude from $\mu^*$, and for a small number of sample ($n<10$).
-	This observation hints towards the optimality of MLE for large $n$. 
-	It makes me feel very interested in convergence rates of MAP in the overparametrized regime $n<\dim(T(X))$.
-	}
-\end{figure}
-
-\section{Euclidean Behaviour}
-Often we can relate this Bregman suboptimality with the $\ell^2$ distance in mean parameter space. 
-\TODO{Add base measure to add here a discussion of the euclidean case -- the mean of a gaussian}
-\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
-    \leq 2L \norm{\mu^* - \mu}
-\end{align}
-so $\bregmanconj$ is $2L$-Lipschitz.
-Since the empirical average converges in expectation to the population mean at a rate of $1/\sqrt{n}$ in $\ell^2$ norm, we know that this bound applies to the log-likelihood.
-\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
-\end{align}
-so $\bregmanconj$ is upper bounded by a quadratic. In expectation, it should converge at a rate $1/n$.
-
-\paragraph{$\ell^2$-norm Analysis}
-Let us make these statements more precise. A bound on the variance becomes a bound on the variance of the average
-\begin{align}
-	\Var T(X) = \E \norm{T(X) - \mu^*}^2 = \sigma^2 \\
-	\implies \E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
-\end{align}
-eg the variance of the mean is $n$ times smaller than the variance of the samples.
-Adding a reference mean $\mu_0$ to get the MAP yields
-\begin{align}
-	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
-	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
-	&= O\left(\frac{\sigma^2}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
-\end{align}
-so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !
-
-\section{Asymptotic Behaviour}
-In the limit, the MAP estimates reach a rate $O(1/2n)$. This is shown by approximating the Bregman divergence using a second order Taylor expansion
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    &= \half (\mu - \mu^*)^\top \nabla^2\conj(\mu^*)(\mu - \mu^*)  
-    + O(\norm{\mu - \mu^*}^3)\\
-    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
-    + O(\norm{\mu - \mu^*}^3)
-\end{align}
-where we introduced the Mahalanobis distance induced by the matrix 
-\begin{align}
-    \nabla^2\conj(\mu^*) 
-    = \nabla^2\logpart(\natp^*)^{-1} 
-    = \Cov_{\natp^*}[T(X)]^{-1} 
-    =: \mSigma^{-1}  \; .
-\end{align}
-To exploit this approximation, let us extend the $\ell^2$ norm results to a Mahalanobis distance induced by matrix $\mM$
-\begin{align}
-	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
-\end{align}
-where $\mSigma$ is the covariance of $T(X)$. We retrieve $\nicefrac{\sigma^2}{n}$, the variance divided by the number of samples, when the metric is the identity $\mM=\mI$.  For the MLE we get
-\begin{align}
-	\E \bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right ) = \frac{d}{2n} + O(n^{- \frac{3}{2}})
-\end{align}
-and for the MAP we  get 
-\begin{align}
-	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
-	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
-	&= \frac{d}{n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
-\end{align}
-or 
-\begin{align}
-\label{eq:MAP_asymptote}
-	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
-	= \frac{d}{2n} + O(n^{- \frac{3}{2}})
-\end{align}
-Remark how the variance does not even appear, only the dimension divided by $n$ really matters.
-I find this result quite spectacular : the convergence speed of MLE is not affected by the covariance of sufficient statistics, and for MAP it matters only in the $O(n^{-2})$ term.
-Indeed this is because it is hidden within the Bregman divergence loss. 
-Anyway that's all good for asymptotic results, but we are interested in a finite sample analysis. How do we get this ?
-
-\section{Bias-Variance Decomposition}
-\newcommand{\bias}[1]{\textcolor{Red}{\underbrace{\textcolor{black}{#1}}_{\text{bias}}}}
-\newcommand{\variance}[1]{\textcolor{Green}{\underbrace{\textcolor{black}{#1}}_{\text{variance}}}}
-
-\subsection{Dual Expectation Pivot.}
-Let $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ be the MAP estimate of the mean (dual) parameter, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ the corresponding natural (primal) parameter, and
-$\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ the expected MAP dual estimate
-Taking $\mu_n$ as a pivot, we can decompose the expected suboptimality in 3 terms -- e.g. 
-\begin{align}
-	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	&= \conj(\mu^* ) 
-	\pm \conj(\mu_n)
-	-  \expect{\conj(\hat \mu_n)}
-	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] 
-	\nonumber \\
-	&= \conj(\mu^* ) 
-	- \conj(\mu_n)  + \E[\bregmanconj(\mu_n, \MAPm)]
-	+ \langle \E[\hat \natp_n] \pm \nabla \conj(\mu_n) ;  \mu_n - \mu^* \rangle
-	\nonumber\\
-	& = \bias{\bregmanconj(\mu^*; \mu_n)}
-	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
-	+ \variance{\E[\bregmanconj(\mu_n, \MAPm)]} \; .
-	\label{eq:bias-mixed-variance}
-\end{align}
-Remark that unless $\conj$ is $\ell^2$, or unless we are in a degenerate case with $\hat \mu_n$ a Dirac, the dual and primal expectation represent different distributions $\theta_n \neq \expect{\hat \theta_n}$, which is why the mixed term (bias-variance scalar product) is non-zero.
-Asymptotically, these terms behave like : bias $O(n^{-2}$, mixed $O(n^{-3})$ and variance $O(n^{-1})$, but the story is different for finite samples.
-We know for sure that the bias and the variance term are positive, but the mixed bias-variance term may be negative, which removes some appeal of this decomposition. 
-Fortunately there is another decomposition revolving around the primal expectation.
-
-%\begin{figure}[ht]
-%	\centering
-%	\includegraphics[width=.8\textwidth]{figs/bias-variance.png}
-%	\caption{A schematic illustration of the different characters featured in the bias-variance decomposition.}
-%	\label{fig:bias-variance}
-%\end{figure}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
-	\label{fig:bias-variance-numerical}
-\end{figure}
-
-\subsection{Primal Expectation Pivot}
-Let $\tilde \theta_n := \expect{\hat \theta_n}$ and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the primal and dual parameters of the primal expectation of the MAP.
-As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
-\begin{align}
-	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	&= \bias{\bregmanconj(\mu^* ; \tilde \mu_n)}
-	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
-	\expect{\bregman(\MAPt; \natp^*)} 
-	&= \bias{\bregman(\E[\MAPt]; \natp^*)}
-	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
-\end{align}
-
-\paragraph{Illustrations.}
-We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
-In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in both of these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=\textwidth]{{figs/biasvariance/linear_ratio=0.8_n0=10}.pdf}
-	\caption{
-	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
-	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
-	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
-	}
-	\label{fig:variance_decomposition}
-\end{figure}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=\textwidth]{{figs/gaussians/log_n0=1}.pdf}
-	\caption{
-	\textbf{Full gaussian $\cN(\mu, \sigma^2)$ example. Left:} training curves and analytic upper bound. 
-	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
-	\textbf{Right:} bias-variance decomposition, using the generalized mean.
-	}
-	\label{fig:gaussian_decomposition}
-\end{figure}
-
-\subsection{About Variance and Covariance}
-
-\begin{lemma}
-Taking  $\MAPt$ and $\MAPm$ as the primal and dual parameters of the MAP estimate, and $\tilde \mu_n$ the primal expectation and $\mu_n$ the dual expectation, the following identities hold
-	\begin{align}
-	\Cov(\hat \natp_n ; \MAPm )  
-	&=\expect{\bregmanconj(\MAPm ; \mu_n)
-	+ \bregmanconj(\mu_n ; \MAPm)} \\
-	&=\expect{\bregmanconj(\MAPm ; \tilde \mu_n)
-	+ \bregmanconj(\tilde \mu_n ; \MAPm)} \; .
-\end{align}
-\end{lemma}
-\begin{proof}
-By the three point identity (or plain derivation), we have 
-\begin{align}
-	\bregmanconj(\MAPm ; \mu_n)
-	+ \bregmanconj(\mu_n ; \MAPm)
-	= \langle \nabla \conj (\hat \mu_n) - \nabla \conj(\mu_n) ;  \hat \mu_n - \mu_n \rangle 
-	= \langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle 
-\end{align}
-Taking the expectation,  the constant $\theta_n$ disappears since $\expect{\hat \mu_n - \mu_n}=0$.
-\begin{align}
-	\expect{\langle \MAPt - \theta_n;  \MAPm- \mu_n \rangle } 
-	 = \expect{\langle \MAPt; \MAPm \rangle} - \langle \expect{\MAPt} ; \MAPm \rangle
-	= \Cov(\hat \natp_n ; \hat \mu_n ) 
-\end{align}
-which is the first identity. We get the second identity by symmetry between the primal and the dual.
-\end{proof}
-
-Note that, as a corollary, the covariance between primal and dual MAP
-$\Cov(\hat \natp_n ; \hat \mu_n )$ is positive, and it can be seen as a non-linear variance of $\hat \mu_n$ -- e.g.
-we can define a variance operator with a non-linear monotone operator $\nabla\conj$ plugged in the middle $\Var_{\nabla \conj}(\hat \mu_n) := \Cov(\nabla\conj(\hat \mu_n) ; \hat \mu_n ) \geq 0$.
-
-\paragraph{Expansion}
-We can expand  $\hat \mu_n - \mu_n$ to get another formulation of this covariance
-\begin{align}
-	\expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle}
-	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
-	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle}
-	= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
-\end{align}
-where we used the property that all data points play a symmetric role to remove the sum over $i$.
-
-\subsection{Triangle Scaling Exponents}
-
-\citet{hanzely2021accelerated} 
-defines the Triangle Scaling Exponent, to characterize an acceleration Bregman methods. The exponent $\gamma$ of a Bregman divergence $\cB$ is defined as the largest coefficient such that $
-	\forall x_0, x_1, x_2, \forall \lambda\in[0,1],$
-\begin{align}
-	\cB((1-\lambda)x_0 + \lambda x_1 ; (1-\lambda)x_0 + \lambda x_2)
-	\leq \lambda^\gamma \cB(x_1 ; x_2) \; .
-\end{align}
-Geometrically, for any triangle, it tells us a shape within which the geodesic triangle is contained (to clarify with a picture.)
-It is related to Thales triangle.
-We precisely have a Thales triangle with the dual expectation pivot, meaning we can relate the MAP variance term with the MLE expected suboptimality. Writing $\lambda = \frac{n}{n+n_0}$, we observe
-\begin{align}
-	\MAPm^\text{MAP} &= (1-\lambda)\mu_0 + \lambda \MAPm^\text{MLE} \\
-	\mu_n &= (1-\lambda)\mu_0 + \lambda \mu^*
-\end{align}
-thus
-\begin{align}
-	\expect{\bregmanconj(\mu_n; \MAPm^\text{MAP})}
-	\leq \left (\frac{n}{n+n_0} \right)^\gamma \expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})} \; .
-\end{align}
-This relationship between the MAP variance and the MLE may be void, as the MLE expected loss may be infinite. But in some settings it might be helpful to characterize the MAP convergence.
-For instance, this same paper says that for a gaussian variance, $\gamma=1$.
-Then they proceed to define an intrinsic scaling exponent, with a local scaling factor.
-This intrinsic exponent is always 2 for Bregman divergences induced by a convex and $\cC^2$ function.
-They use this fact to derive an accelerated algorithm.
-
-\clearpage
-
-\section{Optimization Perspective}
-\paragraph{Problem.}
-We want  to solve the following problem
-\begin{align}
-	f^* = \min_\natp \E[f(\natp ; X)] := -\langle \E[T(X)] ; \natp \rangle + \logpart(\natp)
-\end{align}
-where $\logpart$ is a convex function.
-We assume access to stochastic oracles $X \sim p(X | \natp^*)$ -- e.g. we observe $X$ such that $\E[T(X)] = \mu^*$ and we can compute  stochastic function values and gradients 
-\begin{align}
-	f(\natp ; X) &= -\langle T(X) ; \natp \rangle + \logpart(\natp) \\
-	\nabla f(\natp ; X) &= \nabla \logpart(\natp) - T(X)\; .
-\end{align}
-It is important to note that our problem is merely a stochastic linear tilt of the convex function $\logpart$, yet none of the convergence rates we studied applied to this setting.
-
-\paragraph{Algorithms.}
-We observe $n$ data points $X_i$. 
-We are studying two algorithms solving this problem : MLE and MAP.
-\begin{align}
-	\MAPm^\text{MLE} &= \inv{n}\sum_{i=1}^n T(X_i) 
-	&\MAPt^\text{MLE} &= \nabla\logpart^{-1}(\MAPm^\text{MLE}) \\
-	\MAPm^\text{MAP} &= \frac{n_0\mu_0 + \smallsum_{i=1}^n T(X_i)}{n_0 + n}
-	&\MAPt^\text{MAP} &= \nabla\logpart^{-1}(\MAPm^\text{MAP})	
-\end{align}
-
-\paragraph{Interpretation.}
-As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms with divergence $\bregman$ and initialization $\natp_0$ : 
-\begin{itemize}
-	\item Stochastic Bregman Proximal Point with learning rate $\inv{n_0+n}$,
-	\item Stochastic Mirror Descent with learning rate  $\inv{n_0+n +1}$.
-\end{itemize}
-In both cases, $f(\natp)$ is convex and 1-smooth  and 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.
-
-\paragraph{Goal.} We want to bound the suboptimality $f(\natp) - f^*$ as a function of the number of samples.
-
-\paragraph{\TODO{Discussion with Radu}}
-PROBLEM. Even in the Euclidean setting, the decreasing step-size analysis with SGD (or SPP) gives a rate $O(1/\log(n))$, which is far from the asymptotic $O(1/n)$.
-
-If we change the step-size, for $1/\sqrt{n}$ for instance, the algorithm changes.
-It becomes a weighted average of the sufficient statistics.
-It might be helpful for online learning (changing target), but it is not relevant when the target is fixed.
-
-\section{Stochastic Bregman Proximal Point}
-\subsection{Correspondence with MAP}
-For a dataset of independent samples $\cD_n = \{X_1,\dots,X_n\}$, the posterior can be seen iteratively
-\begin{align}
-	p(\natp |\cD_n)
-	=p(\natp|X_n,\cD_{n-1})
-	\propto p(X_n|\natp,\cD_{n-1}) p(\natp|\cD_{n-1})
-	=p(X_n|\natp)p(\natp |\cD_{n-1}) \; .
-\end{align}
-For the MAP of the exponential family with a Bregman conjugate prior, the posterior takes the form
-\begin{align}
-	- \log p(\natp | \cD_n) 
-	&= \sum_{i=1}^n -\log p(X_i | \natp) - \log p(\natp) + \cst\\
-	&= \sum_i - \langle T(X_i) ; \natp \rangle  + n \logpart(\natp) + n_0 \bregman(\natp ; \natp_0) +\cst \\
-	&= (n+n_0)\logpart(\natp)  - \langle n_0 \mu_0 + \sum_i T(X_i) ; \natp \rangle + \cst \\
-	&= (n+n_0) \bregman(\natp; \hat \natp_n^\text{MAP}) + \cst \; ,
-\end{align}
-so we immediately get the recursion
-\begin{align}
-    \hat \natp_n^\text{MAP}
-    &= \argmin_\natp - \log p(\natp | \cD_n) \\
-    &= \argmin_\natp - \log p(X_n|\natp) - \log p(\natp | \cD_{n-1}) \\
-    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
-    \label{eq:SBPP}
-\end{align} 
-Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0}$.
-
-\subsection{Straightforward Analysis}
-
-\paragraph{Bregman Proximal Point}
-This algorithm was first published in \citet{eckstein1993nonlinear} with an asymptotic convergence property.
-\citet{chen1993convergence} then provided a finite sample convergence rate for deterministic Bregman proximal point.
-Their analysis relies on the three points lemma (Appendix~\ref{app:3points})
-\begin{align}
-	 \forall \natp, \lr_t f(\natp) + \cB(\natp ; \natp_t) \geq \lr_t f(\natp_{t+1}) + \cB( \natp; \natp_{t+1}) +\cB(\natp_{t+1} ; \natp_{t}) \; .
-\end{align}
-Applying it with $\natp=\natp_t$, we prove that function values are decreasing
-\begin{align}
-	f(\natp_{t+1})  \leq f(\natp_{t})  - \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \; .
-\end{align}
-Then applying it with $\natp=\natp^*$ (assuming such a minimizer exists), we obtain a telescopic sum
-\begin{align}
-	\lr_t(f(\natp_{t+1}) - f(\natp^*)) +  \cB(\natp_{t+1} ; \natp_{t}) 
-	&\leq  \cB( \natp^*; \natp_{t})  - \cB( \natp^*; \natp_{t+1}) \\
-	\implies \sum_{t=0}^{T-1} \lr_t (f(\natp_{t+1}) - f(\natp^*)) 
-	+ \sum_{t=0}^{T-1} \cB(\natp_{t+1} ; \natp_{t}) 
-	&\leq \cB( \natp^*; \natp_{0})  - \cB( \natp^*; \natp_{T})  
-	\leq \cB( \natp^*; \natp_{0})  \; ,
-\end{align}
-which proves on the way that the total path length $ \sum_{t=0}^{t-1} \cB(\natp_{t+1} ; \natp_{t}) $ is bounded.
-Using the decreasing property of $f(\natp_t)$, we get
-\begin{align}
-	f(\natp_{T}) - f(\natp^*) \leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t} \; .
-\end{align}
-This upper bound converges to $0$ whenever the sum of step-sizes converges to $0$. 
-This rate can be arbitrarily fast because the oracle is strong.
-Indeed, taking an inifinite step-size solves the problem with one iteration, but the update is (theoretically) harder to solve for larger step-sizes.
-
-\paragraph{Stochastic Setting.}
-While \citet{eckstein1998approximate} shows some asymptotic convergence of the Bregman proximal point algorithm when each iteration is only solved approximately, we are not aware of published analysis when iterations are stochastic.
-The deterministic proof does not transpose immediately to the stochastic setting.
-First off, we do not have a decreasing sequence anymore
-\begin{align}
-	f(\natp_{t+1}, X_t)  \leq f(\natp_{t}, X_t)  
-	- \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
-	\implies \expect{f(\natp_{t+1}, X_t)} \leq f(\natp_{t}) 
-	- \inv{\lr_t}\expect{\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}))}  \; .
-\end{align}
-Indeed the inequality bears on $\expect{f(\natp_{t+1}, X_t)}$, which features a correlation between $\natp_{t+1}$ and $X_t$, rather than $f(\natp_{t+1})$.
-Similarly, applying the three points lemma with $\natp^*$, taking the tower expectation, and telescoping the sum, we get
-\begin{align}
-	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_t) - f(\natp^*)}
-	\leq \cB( \natp^*; \natp_{0})  \; .
-\end{align}
-To pursue further, we need an assumption on the quality of the stochastic estimates $f(. ; x)$. 
-The most straightforward such assumption is 
-\begin{align}
-\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t})} \leq \lr_t \sigma^2 \; ,
-\label{eq:proximal_variance}
-\end{align}
-e.g. we relate the stochastic function value on the next iterate to the function value.
-Then, taking the iterate with the least expected loss, we get the rate
-\begin{align}
-	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
-	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
-	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
-\end{align}
-which is very analogous to standard rates for SGD.
-It does converge, whenever $\sum_{t=0}^{T-1} \lr_t$ diverges to $+\infty$ much faster than $\sum_{t=0}^{T-1} \lr_t^2$.
-When $\lr_t \propto t^{-\half}$ for instance, we get a rate $O(\frac{\log(t)}{\sqrt{t}})$.
-But in our case, $\lr_t \propto t^{-1}$, and we merely get a $O(\inv{\log(t)})$ rate.
-
-\paragraph{Application to the MAP.}
-The variance assumption \eqref{eq:proximal_variance} translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in the exponential family MAP setting -- e.g.
-exactly the same assumption as in \citet{hanzely2018fastest}.
-However even with this assumption, the convergence rate applies to the iterate $\sum_t \lr_t \natp_{t+1} / \sum_t \lr_t$, which is an average in the primal space, and gives more weight to the first iterates.
-It does not immediately apply to the last iterate.
-
-\section{Stochastic Mirror Descent (SMD)}
-\subsection{MAP as SMD}
-In this section, we show that MAP estimate can also be seen as the iterates of stochastic mirror descent with mirror map $\logpart$.
-
-\paragraph{Mirror Descent.}
-Let us recall mirror descent iteration
-\begin{align}
-	\natp_{t+1} 
-	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
-	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
-\end{align}
-where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$.
-Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$.
-Note that finding $\natp_*$ is done with 1 step of mirror descent. Indeed plugging in definitions of $f$ and $\mu$ yields
-\begin{align}
-	&\mu_{t+1}  = \mu_t - \lr (\mu_t - \mu_*) \\
-	\implies &\mu_t  = \mu_* + (1- \lr)^t (\mu_0 - \mu_*)
-\end{align}
-which shows exponential convergence and 1-step convergence when $\lr =1$.
-
-\paragraph{Stochastic Mirror Descent}
-Let $f(\natp, x) = -\log p(x|\natp)$.
-MAP iterations can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1}) = \logpart(\natp_t) - T(x_{t+1})$ as a stochastic estimate  of $\nabla f(\natp_t)$.
-Resuming from the Stochastic Bregman Proximal Point formula~\eqref{eq:SBPP}, we write $\MAPt$ for the n-th MAP iterate, and do some Bregman manipulations
-\begin{align}
-    \hat \natp_{n+1}
-    &= \argmin_\natp 
-    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
-    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
-    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
-    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
-\end{align}
-where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$.
-This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
-
-Stochastic Mirror Descent (SMD) was studied under relative smoothness assumptions in 2 papers : \citet{hanzely2018fastest} and \citet{dragomir2021fast}.
-
-\subsection{Work by Hanzely}
-
-\citet{hanzely2018fastest} got convergence rates in the smooth strongly-convex case with tail averaging.
-With constant step-size, they proved linear convergence down to a variance ball.
-With step-size $\gamma_t = n_0 + t$, they proved a rate $\tilde O(\inv{t})$.
-These results match the rates for standard SGD.
-Let us have a look at their variance assumption \citep[Assumption 5.1]{hanzely2018fastest}.
-Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
-Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the descent direction $\natp_{t+1} - \natp_t = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t ) - \natp_t $ that should hold for all time steps
-\begin{align}
-	\lr_t \sigma^2 \geq 
-	&\Cov(- g_t,\natp_{t+1} - \natp_t | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
-	&= \expect{\langle - g_t, \natp_{t+1} \rangle}  - \langle \E[-g_t], \E[\natp_{t+1}] \rangle   \\
-	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}\\
-	 &= \expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }
-\end{align}
-where $\gamma_t$ is the step-size, expectations are conditional on the past, and $\bar \natp_{t+1}$ is the theoretical output of a deterministic Bregman gradient step.
-Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD.
-
-In our setting, 
-\begin{align}
-	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
-\end{align}
-so that $\sigma$ is really a bound on the covariance of the sufficient statistics with another variable
-\begin{align}
-	&\Cov( T(X_{t+1}), \natp_{t+1}| X_{1\dots t})
-	\leq \lr_t\sigma^2 \\
-	=& \Cov(T(X_{t+1}), \nabla\conj \left(\lr_t T(X_{t+1}) + (1-\lr_t ) \mu_t \right) | X_{1\dots t}) \;.
-\end{align}
-In other words, we need for all $\lr\in(0,\inv{n})$, and a certain subset of $\mu$,
-\begin{align}
-\label{eq:hypvariance}
-	\boxed{
-	\Cov_X \bigg(T(X), \nabla \conj (\lr T(X) + (1-\lr) \mu)\bigg	) 
-	\leq \lr \sigma^2 \; .
-	}
-\end{align}
-
-\begin{example}
-Unfortunately this bound is not trivially satisfied, even for simple exponential family members, such as centered gaussians $\cN(0,\sigma^2)$, when we are aiming to estimate the variance.
-The covariance  \eqref{eq:hypvariance} we want to bound is equal to
-\begin{align}
-	\half \expect{\frac{\sigma_*^2 - X^2 }{ X^2 \lr + (1-\lr) \sigma_0^2 }}
-\end{align}
-When $\sigma_0 \rightarrow 0$, as might happen, this value explodes.
-\end{example}
-
-\subsection{Work by the French Musketeers}
-
-\citet{dragomir2021fast} use a quite different assumption, bearing on a Bregman measure of the stochastic gradients norm at the optimum
-\begin{align}
-	2 \lr^2 \sigma^2
-	\geq \expect[X]{\bregmanconj(\mu_t - 2 \lr_t \nabla f(\natp^*, X) ; \mu_t ) } \; .
-	\label{eq:dragomir_hypothesis}
-\end{align}
-When $\logpart$ is strongly convex, this hypothesis is akin to bounding the variance of the gradient at the optimum, whereas \eqref{eq:hanzely_hypothesis} is akin to bounding gradient variance everywhere.
-This comparison makes this assumption more interesting, as it might apply to unbounded strongly convex function, like modern analysis of SGD.
-
-However it does use a metric which depends on the trajectory $\mu_t$.
-It is the flaw of both these variance assumptions \eqref{eq:hanzely_hypothesis} and \eqref{eq:dragomir_hypothesis}, as well the proximal point variance assumption \eqref{eq:proximal_variance}, that they depend on the stochastic trajectory.
-Indeed this trajectory has a non-zero probability of being arbitrarily close from the border of the domain, where $\bregman$ may get extremely ill-conditioned.
-
-\subsection{Transferring SMD's covariance assumption to MAP estimates}
-
-\paragraph{MLE Analysis.}
-When we plug this bound into the expected suboptimality formula~\eqref{eq:bregmanMLE} with the maximum likelihood estimate $\hat \mu =  \inv{n}\sum_i T(X_i)$, we get
-\begin{multline}
-	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) }
-	=\conj(\mu_*) \overbrace{- \E[\conj(\hat \mu)]}^{\leq - \conj(\E[\hat \mu])} \\
-	+ \inv{n}\sum_i \E[\underbrace{\E[\langle  T(X_i) - \mu_*; \nabla\conj (\hat \mu) \rangle | X_j, j\neq i]}_{\leq \nicefrac{\sigma^2}{n} \eqref{eq:hypvariance}
-	 	\; \text{with}\; \mu=\inv{n-1}\smallsum_{j\neq i} T(x_j)  } ]  \leq \frac{\sigma^2}{n}
-\label{eq:MLE_bound}
-\end{multline}
-where we used the decomposition $\hat \mu = \inv{n}T(X_i) + (1-\inv{n}) \inv{n-1} \sum_{j\neq i}T(X_j)$ to apply~\eqref{eq:hypvariance}. 
-In words, this variance assumption on $T(X)$ and $\conj$ immediately gives us a bound on the suboptimality.
-
-This is too simple to be true, as shown on our trailing example.
-In particular we show that it cannot hold for all possible values of $\mu$ within the double expectation in~\eqref{eq:MLE_bound}.
-The mean parameter $\mu$ take values arbitrarily close from $0$, which makes the covariance explode.
-\RLP{should we get a functional bound $\sigma^2(\mu)$ instead ? or split between $\mu$ smaller or larger than $\varepsilon$.}
-
-\paragraph{MAP Analysis.}
-We can also apply this result to the MAP estimate $\hat \mu = \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n}$ but I did not manage to reach a satisfying conclusion about the bias. 
-Note $\tilde \mu = \E[\hat \mu] = \frac{n_0 \mu_0 + n \mu_*}{n_0+n}$ and $\gamma_0= \frac{n_0}{n_0+n}$ and the step-size $\gamma = \inv{n_0 + n}$ .
-\begin{align}
-	\expect[X_{1\dots n}]{\bregmanconj ( \mu_* ;\hat \mu) } 
-	&\leq \conj(\mu_*) - \conj(\tilde \mu) + \gamma_0 \langle \mu_0 - \mu_* ; \E[\hat \natp] \rangle + (1 - \gamma_0) \gamma \sigma^2  \\
-	& = \bregmanconj (\mu_* ; \tilde \mu) - \langle \tilde \mu - \mu_* ;    \tilde \natp - \E[\hat \natp] \rangle + \frac{n}{(n_0+n)^2}\sigma^2
-\end{align}
-so we recover the $O(n^{-1})$ for the variance and we are still looking for $O(n^{-2})$ rate for the bias. The Bregman term in the bias should be in this spirit, given a quadratic approximation
-\begin{align}
-	 \bregmanconj (\mu_* ; \tilde \mu)  
-	 = \bregmanconj (\mu_* ; \mu_* + \gamma_0 (\mu_0 - \mu_*) ) \approx \half \gamma_0^2 \norm{\mu_0 - \mu_*}^2_{\mSigma_*^{-1}}
-\end{align}
-where $\mSigma_* = \Cov_{\mu_*}(T(X))$. 
-The order of terms $\bregmanconj (\mu_* ; \tilde \mu)  \neq \bregmanconj (\tilde \mu ; \mu_* ) $ only affects third order terms.
-That's for the bregman, but the scalar product is harder to bound.
-Hopefully it could be positive.
-If not we know that $\tilde \mu - \mu_* = \frac{n_0}{n_0+n} (\mu_0 - \mu_*) = O(n^{-1})$ , and we can hope that the same holds for $\tilde \natp - \E[\hat \natp] = \nabla \conj(\E[\hat \mu]) - \E[\nabla\conj (\hat \mu)]$, which kinda measures the non-linearity, or the curvature of $\conj$.
-Using a simple cauchy-schwartz, and the hessian of the conjugate, we might be able to get something.
-
-\paragraph{Relative Continuity}
-\citet{lu2019relative} define relative continuity to tackle non-differentiable functions -- eg Hinge loss.
-These functions are typically assumed to be Lipschitz, but that is not always true -- eg squared Hinge loss.
-Relative continuity generalizes Lipschitz-ness by upper bounding the gradient norm (or expected norm of stochastic gradients) with the ratio between a Bregman and $\ell^2$ distance.
-Although this hypothesis is far from intuitive, it does yield convergence rates very similar to typical SGD analysis.
-In this work we are first concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
-
-\paragraph{Bound on the trajectory.}
-Bound \eqref{eq:hypvariance} might not hold in general, but it can along the trajectory. Can we prove it using an inductive method such as \citet[Appendix A]{lacoste2012simpler}?
-No we cannot because $\mu$ takes every possible values as a random variable within the expectation.
-
-\subsection{Work by Loizou}
-An assumption pointed out by Nicolas Loizou is 
-\begin{align}
-	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
-\end{align}
-for which a sufficient condition is 
-\begin{align}
-\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
-\end{align}
-which holds in a lot of settings but not all.
-For instance, a gaussian $\cN(\mu,\sigma^2)$ can overfit on a single data point and give it infinite density -- eg converge to a Dirac.
-However we can overcome this issue by grouping data points : a gaussian can not overfit on two points.
-Aggregating points is simple : average sufficient statistics, exactly like in MLE or MAP $T'(X_1,X_2) = \frac{T(X_1) + T(X_2)}{2}$.
-So in general we may say that this hypothesis holds, at the expense of dividing the number of samples by some number $n/k$. What kind of results can it give us my dear Nicolas Loizou ? ;)
-
-\section{Expected Relative Smoothness}
-To avoid gradient variance assumption, we can take inspiration from expected smoothness, which is a consequence of convexity and smoothness, and which yields a general analysis of SGD.
-Can we get something similar with relativity and SMD?
-The analog of gradient norm is the Bregman between iterates, similar to the Homeomorphic EM rate.
-
-\subsection{Convergence of Mirror Descent}
-To get the descent lemma, apply successively relative smoothness, 3 points lemma, and relative strong convexity on $f$.
-Note that the strong convexity constant $m$ may be zero.
-Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
-\begin{align}
-	f(\natp^{t+1}) 
-	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
-	& \text{(Relative $L$-smoothness)} \\
-	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
-	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(Relative $m$-strong convexity)}
-\end{align}
-Applying the 3 points lemma with \TODO{f}
-
-\section{Self-Concordance}
-
-A big problem is that $\conj$ is seldom Lipschitz or smooth. For instance the log-partition function of a multivariate normal is 
-\begin{align}
-    \logpart(\eta, \Lambda) = \half \eta^\top \Lambda^{-1} \eta - \log\det(\Lambda)
-\end{align}
-which is defined on $\eta \in \real^d$ and $\Lambda\in \real^{d\times d}$ symmetric positive definite. 
-It is not strongly convex, so $\conj$ is not smooth.
-
-Another hypothesis that may be more suitable is self-concordance. $f: \real \rightarrow \real$ is SC if 
-\begin{align}
-    \abs{f'''(x)} \leq 2 f''(x)^{\frac{3}{2}} \; .
-\end{align}
-The exponent $\frac{3}{2}$ is motivated by dimensional analysis and the factor $2$ appears to simplify downstream calculus.
-A multidimensional function $f: \real^n \rightarrow \real$ is SC if it's restriction to any line is SC.
-Negative logarithm $-\log(x)$ and entropy $x\log(x)$ are both self-concordant function. This is good news for us since log-partition function may include SC logarithmic barriers.
-In particular, gaussians have a logarithmic term. They also have an inverse term which is not self-concordant, but which is generalized self-concordant.
-
-\paragraph{Questions.}
-Is there already an SC analysis of proximal point methods. (that’s not affine invariant)?
-If the log-partition is SC, then what does it mean for its gradient, the mapping from natural to mean parameters ?
-
-\section{Related Work}
-Exponential families are a mainstream tool in machine learning.
-They are used  to generalize linear regression \citep{mccullagh1989generalized}, PCA \citep{collins2001generalization} or k-means \citep{banerjee2005clustering} to diverse data types and distributions.
-\citet[Chapter 3]{wainwright2008graphical} give a great overview of exponential families and their dual structure.
-
-\citet{agarwal2010geometric} highlight that the MAP problem of exponential families is also a Bregman median problem.
-They then use this equivalence to justify the use of conjugate prior in Bayesian estimation and hybrid generative-discriminative modeling.
-Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors yet critical as it does not hold for multivariate normals with unknown covariance.
-\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.
-Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
-
-\citet{kunstner2020homeomorphic} show with an exquisite elegance that Expectation-Maximization in exponential families can be cast as mirror descent on the negative log-likelihood.
-Then they use recent mirror descent rates based on the idea of relative smoothness \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively} to show convergence of EM in KL-divergence, including for the ubiquitous Gaussian Mixture model, which had resisted bounding attempts for 50 years.
-
-\citet{pfau2013generalized} gives a bias variance decomposition for Bregman divergences. However it does not give interesting conclusions for our problem of interest.
-
-\citet{bubeck2015entropic} shows that the entropy of an exponential family defined on a compact support $\cX$ is a self-concordant barrier on this compact set.
-They extend the definition of self-concordance to $\nu$-self-concordance, stating that the gradient should be bounded by the square root of the hessian along any direction.
-In 1 dimension, their result also means that the log-partition is self-concordant as well.
-In general, does this mean that the log-partition is self-concordant  (e.g.
-is there a link between the dual function on a line and the line of the primal , probably not)?
-
-In the realm of self-concordance, \citet{ostrovskii2021finite}  shows a finite sample convergence rate for the MLE (or other $M$-estimators)   when  the log-likelihood is self-concordant.
-This convergence rate holds only after a given number of samples have been seen, and they give an innteresting formula for this number.
-To get this results, they use the fact that self-concordant losses are upper bounded by a quadratic in a neighborhood of the optimum.
-
-\TODO{3 threads of literature search: optimization, statistics, learning theory.
-This last one is not covered yet.
-It is more interested in finite sample results than statistics.
-Contact Alexander Rakhlin}
-
-\paragraph{Bregman proximal point}
-
-\paragraph{high probability bounds}
-
-\subsection{Alternative Research Tracks}
-\begin{itemize}
-	\item A picture / geometric understanding of the equality between ELBO (Jensen) and relative smoothness for EM.
-	\item Bregman is about general geometry, can we bring it to the non-convex realm via local analysis ?
-	\item Can we design an algorithm with higher order Bregman divergences ?
-Does it mean anything ?
-The Bregman is a quadratic approximation of a function, yielding a geometry.
-Higher order does not carry such meaning.
-But we would like to involve some quadratic function information within mirror descent !
-\end{itemize}
-
 \bibliographystyle{apalike}
-\bibliography{references.bib}
-
-\newpage
-\appendix
-
-\section{Three points lemma}
-\label{app:3points}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics{figs/3pointslemma}
-	\caption{Illustration of the other three points lemma -- eg the identity between Bregmans -- reproduced from \citet{eckstein1998approximate}.}
-	\label{fig:3pointslemma}
-\end{figure}
-
-The typical descent lemma of updates involving minimization of a Bregman divergence is called the three point lemma. 
-It corresponds to first order optimality conditions of the update.
-With its hep, one can prove convergence for mirror descent with relative smoothness, or for Bregman proximal point algorithms.
-
-\begin{lemma}
-Let $x_+$ be a solution to 
-\begin{align}
-	x_+ \in \argmin	_{x\in C} \overbrace{f(x) + \cB_h(x ; y)}^{\phi(x)}
-\end{align}
-where $C$ is a closed convex set, $f$ is a convex function, $\cB_h$ is the Bregman divergence induce by some convex function $h$, and $y$ is some reference vector. Then 
-\begin{align}
-	\forall x, f(x) + \cB(x ;y) \geq f(x_+) + \cB(x ; x_+) +\cB(x_+ ; y) \; .
-\end{align}
-\end{lemma}
-
-Taking an information geometry lens, this property is analog to the Pythagorean theorem for generalized projections.
-We recover euclidean projections setting $f=0$ and $h=\|.\|^2$.
-
-\begin{proof}
-	The first order optimality condition is
-	\begin{align}
-		&\langle \nabla \phi (x_+) , x -  x_+ \rangle \geq 0, \forall x \\
-		=& \langle \nabla f(x_+) + \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  
-	\end{align}
-	This proof relies on another property called three point property 
-	\begin{align}
-		\langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  =\cB(x;y) -\cB(x;x_+) - \cB(x_+;y) 
-	\end{align}
-	which can be proved by expanding the right hand side.
-	By convexity of $f$ we also have 
-	\begin{align}
-		 f(x_+) + \langle \nabla f(x_+), x - x_+ \rangle \leq f(x) \; .
-	\end{align} 
-	Putting it all together we get
-	\begin{align}
-		0 
-		&\leq  \langle \nabla f(x_+), x - x_+ \rangle + \langle \nabla h(x_+) - \nabla h (y), x - x_+ \rangle  \\
-		&\leq f(x) - f(x_+) + \cB(x;y) -\cB(x;x_+) - \cB(x_+;y)
-	\end{align}
-	which concludes the proof.
-\end{proof}
-
-\section{Relative Smoothness for Mirror Descent}
-In the classic setting, SMD is studied under strong-convexity assumption on the mirror map $\logpart$ \citep{bubeck2015convex}.
-In our setting this is not always true -- eg gaussians.
-However a recent and fast-expanding body of work is concerned with a new assumption: relative smoothness and relative strong-convexity.
-\begin{align}
-	& \text{$f$ is $L$-smooth relative to $h$} \\
-	\iff & Lh - f \ \text{convex} \\
-	\iff & f(y) \leq 	f(x) + \langle \nabla f(x) , y-x \rangle + L\cB_h(y;x), \forall x,y \\
-	\iff & \cB_f(y;x) \leq L \cB_h(y;x), \forall x,y \\
-	\iff & \nabla^2 f(x) \leq L \nabla^2 h(x),  \forall x
-\end{align} 
-where the last equivalence holds only when $f$ and $h$ are twice differentiable.
-In words, $f$ is upper bounded by it linear approximation plus the $h$-Bregman divergence, which can also be seen as a bound between divergences, or more locally as a bound between Hessians.
-Similarly, $f$ is $\mu$ strongly-convex relative to $h$ if
-\begin{align}
-	& f-\mu h \ \text{ convex} \\
-	\iff & f(x) + \langle \nabla f(x) , y-x \rangle + \mu \cB_h(y;x) \leq f(y)\forall x,y \\
-	\iff & \mu\cB_h(y;x) \leq \cB_f(y;x) \forall x,y \\
-	\iff & \mu \nabla^2 h(x) \leq \nabla^2 f(x)  \forall x \; .
-\end{align}
-
-Another way to view this elegant generalization of smoothness and strong-convexity is as a transfer of the Loewner partial order on symmetric matrices to functions, via the Hessian. 
-As such it can be applied to many functions that were out of reach for $\ell^2$ norm, by taking the appropriate reference function.  For instance $h(x) = -\log(x)$ or $h(x) = x^4$. 
-As early as 2011, \citet{birnbaum2011distributed} showed $O(\inv{t})$ convergence rate for mirror descent under smoothness assumption relative to the mirror map. 
-More precisely, he proved that when $f$ is $L$-smooth relative to $h$, then the suboptimality of the sequence 
-\begin{align}
-	x_{t+1}  = \argmin_x \langle \nabla f(x_t), x \rangle + \cB_h(x ; x_t)
-\end{align}
-is upper bounded by the simple formula
-\begin{align}
-	\implies f(x_t) - f(x_*) \leq \frac{L \cB_h(x_* ; x_0) }{t} \; .
-\end{align}
-These notions were rediscovered and expanded by \citet{bauschke2017descent} and \citet{lu2018relatively}.
-If you need to read one, pick \citet{lu2018relatively} -- I found it much much easier and more enjoyable to read.
-This latter paper also derived a linear convergence rate for mirror descent under relative smoothness and strong-convexity, with the relative condition number $\frac{L}{\mu}$ appearing.
-
-\section{Cramer-Rao Lower Bound}
-In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
-In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
-\begin{align}
-    \Cov(\hat \natp_n)  
-    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
-    \iff
-    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
-    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
-\end{align}
-This is a typical statistic bound, which is concerned with the $\ell^2$ accuracy of an estimator, via its covariance.
-In contrast, here we are concerned with the quality of the estimator in terms of KL-divergence, eg "how accurately am I modeling this distribution?"
-
-Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
-For the estimator $\hat \natp_n$, this lower bound is not trivial.
-
-In our case, the bias variance decomposition tells us that we are interested in the covariance between primal and dual parameters, rather than the covariance of only of these -- $\Cov(\MAPt;\MAPm)$ rather that $\Var(\MAPm)$ or $\Var(\MAPt)$.
-
-\section{Posterior Expected Loss}
-
-A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is written as
-\begin{align}
-    p(\natp) 
-    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
-    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle) \; .
-\end{align}
-We see that $p(\natp)$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
-When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $p(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
-This is well detailed in \citet{agarwal2010geometric}.
-
-Given this posterior, a natural quantity to consider is the expected loss. Thus we might ask "what is the expectation of the posterior expected suboptimality?"
-\begin{align}
-    \expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-\end{align}
-This quantity is even harder to estimate than the expected MAP suboptimality, because of the double integral.
-We can get a closed form for the age-old $\ell^2$ loss of $\cN(\mu, 1)$, which yields the exact rate $\inv{n}$.
-This is exactly twice the expected loss of the MLE.
-
-\begin{example}
-	We can also get a closed form for the posterior of a centered gaussian variance 
-	\begin{align}
-		\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
-	\end{align}
-	using the mean of the \href{https://en.wikipedia.org/wiki/Inverse-gamma_distribution}{inverse Gamma distribution} and the \href{https://en.wikipedia.org/wiki/Gamma_distribution#Logarithmic_expectation_and_variance}{logarithmic expectation} of a Gamma.
-	Then using results from the MLE, we get the exact formula
-	\begin{align}
-		\expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
-	\end{align}
-	and we bound it to 
-	\begin{align}
-		\inv{2n} + \frac{4}{n(n-2)}
-		\leq \expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
-		\leq 
-		\inv{n} + \frac{4}{n(n-2)}
-	\end{align}
-	which is about twice the MLE bounds~\eqref{eq:MLE_rate}, exactly as we reportred for the gaussian mean.
-\end{example}
-
-\section{Gaussian Variance}
-\subsection{Proof of MLE Tight Bound}
-	\begin{theorem}[MLE tight upper bound]
-	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
-	Its expected suboptimality is infinite when $n\leq 2$ and otherwise
-	\begin{align}
-		 \expect{\bregmanconj( \mu_*; \mu_n) }
-			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
-	\end{align}
-	\end{theorem}
-
-	\begin{proof}
-	The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$.
-Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
-	\begin{align}
-		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
-		= \expect{\frac{n}{\chi^2(n)} - 1} =
-		\begin{cases}
-			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
-			+\infty \  \text{ otherwise. }
-		\end{cases}
-	\end{align}
-	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
-	\begin{align}
-		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
-	\end{align}
-	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
-	Consequently the suboptimality of the MLE has a closed form solution
-	\begin{align}
-	\expect{\bregmanconj( \mu_*; \mu_n) }
-	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
-	& =
-	\begin{cases}
-		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
-			+\infty \  \text{ otherwise. }
-	\end{cases}
-	\end{align}
-	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
-	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
-	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
-	\begin{align}
-		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
-	\end{align}
-	  For $n\geq 3$,
-	\begin{align}
-		\inv{n-2} - \inv{n}
-		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
-		\leq \inv{n-2} - \inv{2 n} \\
-		\iff
-			\frac{2}{n(n-2)}
-			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
-			\leq \inv{2n} +\frac{2}{n(n-2)}
-	\end{align}
-	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
-	\end{proof}
-	Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
-	\begin{align}
-		\expect{\bregmanconj( \mu_*; \mu_n) }
-		\in \inv{2n} + \frac{11}{6 n^2} + \frac{4}{n^2(n-2)} + O(n^{-4}) \; .
-	\end{align}
-
-\subsection{Proof of MLE Loose Upper Bound}
-If we apply \eqref{eq:log_bound} to the MLE, we get the upper bound
-\begin{align}
-	\expect{\bregmanconj(\mu^*; \MAPm^\text{MLE})}
-	&\leq  \half \left( \expect{\frac{n}{\chi^2(n)}} + \expect{\frac{\chi^2(n)}{n}} - 2 \right ) \\
-	&= \half \left( \frac{n}{n-2} + 1 - 2 \right)  
-	= \frac{1}{n-2}
-\end{align}
-which is exactly  an additive term $1/2n$ larger than \eqref{eq:MLE_rate}.
-This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.
-
-\subsection{Proof of Expected MAP Natural Parameter}
-\begin{lemma}[Expected MAP natural parameter]
-	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
-	\begin{align}
-		\frac{\mu^*}{\mu_n}
-		\leq \expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\natp^*}} 
-		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n+n_0}} \; .
-	\end{align}
-\end{lemma}
-
-\begin{proof}
-	To start, let us plug in the definition of $\MAPm$
-	\begin{align}
-		\expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\natp^*}} 
-		&= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  \\
-		&= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
-		&= (n_0 +n) \expect{\inv{a + \chi^2(n)}}
-	\end{align}
-	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and 
-\begin{important}
-	\begin{align}
-			a=n_0 \frac{\mu_0}{\mu^*} \; .
-	\end{align}
-\end{important}
-	We reformulate this expectation with the trick
-	\begin{align}
-		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
-		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
-		\implies &\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,
-	\end{align} 
-	to get two integrals instead of one
-	\begin{align}
-		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
-		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
-		&= \int_0^\infty dx\  x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt\  e^{-\half[(a+x)t]}  \\
-		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
-	\end{align}
-	where we used Fubini to switch integrals, without further justification.
-	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
-	\begin{align}
-		\expect{\inv{a + \chi^2(n)}}  
-		&= \frac{e^{\half[a]} }{2^{\half[n]+1}\Gamma(\half[n]) }\int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty 2\frac{dy}{t} \  (2\frac{y}{t})^{\frac{n}{2}-1} e^{-y} \\
-		&= \frac{e^{\half[a]} }{2\Gamma(\half[n]) } 
-		\underbrace{\int_1^\infty  \frac{e^{-\half[a t]}}{t^{\frac{n}{2}}}  dt}_{E_{\half[n]}(\half[a])  } \ 
-		\underbrace{\int_0^\infty  y^{\frac{n}{2}-1} e^{-y} dy}_{\Gamma(\half[n])}
-	\end{align}
-	so we finally get  the formula valid for all $n\geq 1$
-	\begin{align}
-		\boxed{\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])}
-		\label{eq:hard_expectation}
-	\end{align}
-	where the \href{https://dlmf.nist.gov/8.19}{generalized exponential integral function} is defined as 
-	\begin{align}
-		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
-	\end{align}
-	Now our goal is to bound this function with simpler functions.
-	Fortunately, mathematicians have been working on these integrals for decades.
-	When $n=2$, eg $k=1$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8\#E1}{verifies the bound}
-	\begin{align}
-		\half \log(1 + \frac{2}{x}) 
-		\leq e^x E_1(x) 
-		\leq \log(1+\inv{x}) \leq \inv{x} \; .
-	\end{align}
-	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19\#E21}{the general bound}
-	\begin{align}
-		\inv{x+k}
-		\leq e^x E_k(x) 
-		\leq \inv{x + k - 1}
-	\end{align}
-	so that 
-	\begin{align}
-		\inv{a+n}
-		&\leq \expect{\inv{a + \chi^2(n)}}
-		\leq \inv{a + n - 2} \\
-		\iff \frac{n +n_0}{a + n}	
-		&\leq  \expect{\frac{\mu^*}{\MAPm}} 
-		\leq \frac{n +n_0}{a + n - 2}	\\
-		\iff \frac{\mu^*}{\mu_n}
-		&\leq \expect{\frac{\mu^*}{\MAPm}} 
-		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n + n_0}}  
-	\end{align}
-	where we used $\frac{a + n}{n +n_0} = \frac{\mu_n}{\mu^*}$.
-\end{proof}
-
-We are left with a special case when  $n=1$, eg $k=\half$.
-Then the integral can be written with the complementary error function, or \href{https://dlmf.nist.gov/7.8#E1}{Mill's ratio} $M(x) = e^{x^2} \int_x^\infty e^{-t^2} dt$.
-Indeed, taking $t= x \sqrt{u}$
-\begin{align}
-	\int_x^\infty e^{-t^2} dt = x \int_1^\infty \frac{e^{-x^2 u}}{\sqrt{u}} du = x E_{\half}(x^2) \\
-	\implies M(x) = x e^{x^2} E_{\half}(x^2)
-\end{align}
-so that 
-\begin{align}
-		\half e^{\half[a]} E_{\half[1]}(\half[a]) 
-		= \half \sqrt{\frac{2}{a}} M(\sqrt{\half[a]})
-		= \inv{a} \sqrt{\half[a]} M(\sqrt{\half[a]}) \; .
-\end{align}
-and \href{https://dlmf.nist.gov/7.8#E5}{we known that} $x M(x) < \half$ so we get 
-\begin{align}
-	\expect{\inv{a + X^2}} < \inv{2 a}
-\end{align}
-FUCK That's  trivial as  it's true for 1/a.
-
-\subsection{Proof of MAP Bound}
-\begin{theorem}[MAP Upper Bound]
- The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is 
-\begin{equation}
-	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
-	\leq \begin{cases}
-		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
-		\bregmanconj( \mu_*; \mu_0) +\inv{a} \ \text{if}\ n=1,\\
-		\frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} if n\geq 2 \ \text{and}\ a=n_0 \frac{\mu_0}{\mu^*} >0
-	\end{cases}
-\end{equation}
-\end{theorem}
-
-$\frac{1+n_0}{4} e^{\half[a]} \sqrt{\frac{2 \pi}{a}} \mathrm{erfc}(\sqrt{\half[a]})  - \half  + \frac{a - n_0}{2(1+n_0)} $
-
-\begin{proof}
-		Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound}. We get
-\begin{align}
-	2\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \expect{\frac{\mu^*}{\MAPm}}  + \expect{\frac{\MAPm}{\mu^*}} - 1 \\
-	&\leq \frac{\mu^*}{\mu_n - \frac{2 \mu^*}{n + n_0}}   + \frac{\mu_n}{\mu^*} - 1 \\
-	&= \frac{n_0 + n}{a +n - 2} + \frac{a+n}{n_0 + n} - 2 \\
-	&= \frac{n_0 - a + 2}{a +n - 2} + \frac{a - n_0}{n_0 + n} \\
-	&= \frac{2}{a+n-2} + \frac{(n_0 - a)^2}{(a+n-2)(n_0 + n)} \\
-	\implies \expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a)^2}{2(a+n-2)(n_0 + n)} \\
-	&\leq \frac{1}{a+n-2} + \frac{\half(1 - \frac{\mu_0}{\mu^*})^2}{(\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} 
-\end{align}
-where recall that $a=n_0\frac{\mu_0}{\mu^*}$. 
-This inequality highlights a clear variance-bias decomposition, to compare with the MLE.
-Unfortunately it is not tight asymptotically as the inequality \eqref{eq:log_bound} is not quadratically tight around $1$.
-We are basically losing a factor 2 compared to $1/2n$.
-by~\eqref{eq:log_bound}.
-\end{proof}
-
-\section{More on the Bias-Variance Decomposition}
-
-\subsection{Bias Term}
-The pure bias term measures the divergence between the optimum and the expected MAP estimate $\bregmanconj(\mu^*; \mu_n)$.
-
-\paragraph{Asymptote.}
-Using the quadratic approximation of the Bregman divergence near $\mu^*$, we get a $O(n^{-2})$ rate for the pure bias term
-\begin{align}
-	\bregmanconj(\mu^*; \mu_n) 
-	= \frac{\norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}}}{2(1 + \frac{n}{n_0})^{-2} } + O(n^{-3}) \; .
-\end{align}
-
-\begin{example}
-We can use the classical lower bound on the logarithm to upper bound the pure bias term 
-	\begin{align}
-		\bregmanconj(\mu^*; \mu_n)
-		 &= \frac{\mu^*}{\mu_n} -1  - \log \frac{\mu^*}{\mu_n}  \\
-		 &\leq \frac{(\mu^*- \mu_n)^2}{2 \mu^* \mu_n} 
-		 = \half ( \frac{\mu^*}{\mu_n} + \frac{\mu_n}{\mu^*	}) - 1 \\		 
-		 & =  \half (\frac{n_0 + n}{a + n} -1 + \frac{a +n}{n_0+n} -1) \\
-		 & =  \half (\frac{n_0 - a}{a+n} + \frac{a - n_0}{n_0 + n}) \\
-		 & = \half[n_0-a] (\inv{a+n} - \inv{n_0+n}) \\
-		 &= \half[n_0-a] \frac{n_0 + n - a- n}{(a+n)(n_0+n)} \\
-		 &= \frac{(n_0-a)^2}{2(a+n)(n_0+n)}
-	\end{align}
-	which is a $O(n^{-2})$ term, worth $0$ when $n_0=a$, eg $\mu_0=\mu^*$, exactly as we expected.
-\end{example}
-
-\subsection{Variance Term}
-
-\paragraph{Asymptote.}
-Using a linear approximation of the mirror map, we get a $O(n^{-1})$ rate for the covariance
-\begin{align}
-	&\nabla \conj(\mu)  
-	= \nabla \conj(\mu^*) + \nabla^2\conj(\mu^*)(\mu - \mu^*) + O(\norm{\mu - \mu^*}^2) \\
-	\implies &\hat \natp_n 
-	= \natp^* + \Sigma^{-1} (\hat \mu_n - \mu^*)  + O(\E[\norm{\hat \mu_n - \mu^*}^2]) \\
-	\implies &\Cov(\hat \natp_n ; T(X) ) 
-	= \inv{n+n_0} \E[ \norm{T(X) - \mu^*}^2_{\Sigma^{-1}}]   + O(n^{-2}) \\
-	\implies &\Cov(\hat \natp_n ; T(X) ) 
-	 = \frac{d}{n+n_0} + O(n^{-2}) \; .
-\end{align}
-Note that we lost a factor $\half$ compared to the straightforward asymptotic analysis~\eqref{eq:MAP_asymptote}, probably because of the Jensen upper bound.
-
-\begin{example}
-	For the gaussian variance MAP, the variance term is
-		\begin{align}
-			\Cov(\hat \natp_n ; \hat \mu_n )  
-			= \E[\MAPt \MAPm] - \E[\MAPt] \E[\MAPm] 
-			=  - \half - \E[\MAPt] \mu_n
-	\end{align}
-	where we used the fact that $\natp \mu = -\half$ for all valid pairs $(\natp, \mu)$. 
-	Plugging-in \eqref{eq:theta_expectation_bound} yield
-	\begin{align}
-		-\half + \frac{\mu_n}{2 \mu_n}
-		&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
-		\leq -\half + \half \frac{a + n}{a + n - 2} \\
-		\iff 0&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
-		\leq \half \frac{a+n - (a + n -2)}{a + n - 2} \\
-		\iff 0&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
-		\leq \inv{a +n - 2}
-	\end{align}
-	which is exactly the kind of upper bounds we were looking for !
-\end{example}
-
-\subsection{Mixed Bias-Variance Term}
-The mixed bias variance term is $ \frac{n_0}{n+n_0} \langle  {\mu_* - \mu_0} ; {\natp_n - \E[\hat \natp_n]} \rangle $.
-Note that it might be negative !
-Let's focus on the right factor
-\begin{align}
-\theta_n - \expect{\hat \theta_n} = \nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]
-\end{align}
-is sort of a commuting bracket between $\nabla \conj$ and the expectation over $\hat \mu_n$. It will be large if $\nabla \conj$ is highly non linear over the distribution of $\hat \mu_n$.
-
-\paragraph{Asymptote.}
-A linear approximation of $\nabla \conj$ shows that the mixed term is in $O(n^{-3})$, since the linear function commutes with the expectation.
-We need to use a quadratic approximation of $\nabla \conj$ to get a non-trivial approximation
-\begin{align}
-	&\nabla \conj(\mu)  
-	= \nabla \conj(\mu^*) + \nabla^2\conj(\mu^*)(\mu - \mu^*) +  \half \nabla^3 \conj(\mu^*) [ (\mu - \mu^*)  ; (\mu - \mu^*) ]
-	+ O(\norm{\mu - \mu^*}^3) \\
-	&\implies 
-	\nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]
-	= \half \nabla^3 \conj (\mu^*) \Cov(\hat \mu_n) 
-	+ O(\E[\norm{\hat \mu_n - \mu^*}^3]) \\
-	&\implies 
-	\nabla \conj(\E[\hat \mu_n]) - \E[\nabla \conj(\hat \mu_n)]
-	= \frac{\nabla^3 \conj (\mu^*) \Sigma}{2(n+n_0)^2} 
-	+ O(n^{-3}) \\
-\end{align}
-where the axis of the tensor product between third derivative and vector / matrix are all kept implicit for simplicity.
-Consequently the bias-variance term is 
-\begin{align}
-	\frac{n_0}{n+n_0} \langle  \bias{\mu^* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle
-	 = \frac{n_0}{(n+n_0)^3} (\mu^* - \mu_0)^\top \nabla^3 \conj (\mu^*) \Sigma + O(n^{-4}) = O(n^{-3}) \; .
-\end{align}
-
-\begin{example}
-	As we have the tight bounds \eqref{eq:theta_expectation_bound} for the expectation of the natural parameter of the MAP, we can bound above and below this mixed term
-	\begin{align}
-		0
-		\leq \natp_n - \E[\MAPt]
-		\leq \frac{n_0 +n}{2 \mu^*} \left ( \inv{a+n-2} - \inv{a+n} \right)
-		 = \frac{n_0+n}{\mu^*(a+n)(a+n-2)}	 \in O(n^{-1}) \; .
-	\end{align} 
-	Remark how this asymptotic class differs from the $O(n^{-2})$ I expected. Is this bound less tight than I thought, or are my calculus wrong ? 
-	Also remark that $\natp_n$ is always greater than $\E[\MAPt]$, which is a good sanity-check because $\nabla \conj(\mu) = - \inv{2\mu}$ is a concave function.
-
-	The rate on the mixed term depends on the sign of $\mu^* - \mu_0 $
-	\begin{align}
-		\frac{n_0}{n+n_0} \langle  \mu^* - \mu_0 ; \natp_n - \E[\hat \natp_n] \rangle
-		\leq \begin{cases}
-			0 \ \text{ if } \ \mu^*\leq \mu_0, \\
-			\frac{n_0 - a}{(a+n)(a+n-2)} \ \text{ otherwise.}
-		\end{cases}
-	\end{align}
-\end{example}
-
-\subsection{Putting it all together}
-\begin{example}
-\begin{align}
-	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	\leq \bias{\frac{(n_0-a)^2}{2(a+n)(n_0+n)}}
-	+ \frac{\max(n_0 - a , 0)}{(a+n)(a+n-2)}
-	+ \variance{\frac{1}{a+n-2}}
-\end{align}
-\end{example}
-
-\subsection{More Illustrations of the Bias-Variance Decomposition}
-See Figure~\ref{app:fig:bias-variance-numerical}.
-
-\begin{figure}[p]
-	\centering
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=1.pdf}
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=2.pdf}
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=10.pdf}
-	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
-	\label{app:fig:bias-variance-numerical}
-\end{figure}
-
-\clearpage
-
-\section{Tentative with self-concordant function}
-
-We want to have a rate of convergence on
-\[
-    \mathbb{E}_{X_i} \left[ \mathcal{B}_{A^*}(\mu_*,\mu_n) \right],
-\]
-where
-\[
-    \mu_* = \mathbb{E}_{X} T(X), \quad \mu_n =  \frac{\sum_{i=1}^n T(X_i) + n_0 \mu_0}{n_0+n}.
-\]
-We first assume that $A^*$ is self-concordant.
-
-\subsection{Notations}
-\begin{itemize}
-    \item $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$
-    \item $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$
-    \item $\lambda(x) = \|(A^*)'(x)\|_x^*$ (Newton decrements)
-    \item $\omega = t-\ln (1+t)$
-    \item $\omega^* = -t-\ln(1-t)$
-\end{itemize}
-Note that $\omega^*$ is convex and monotonne.
-
-\begin{proposition}
-    (Conversion of norms) We have
-    \[
-        \frac{\|y-x\|_x}{1+\|y-x\|_x} \leq \|y-x\|_y \leq \frac{\|y-x\|_x}{1-\|y-x\|_x}
-    \]
-\end{proposition}
-
-\begin{proposition}
-    (Bounded Hessian change) We have
-    \[
-        (1-\|y-x\|_x)^2 \nabla^2 A^*(x) \leq \nabla^2 A^*(y) \leq \frac{1}{(1-\|y-x\|_x)^2} \nabla^2 A^*(x)
-    \]
-\end{proposition}
-
-\begin{proposition}
-    (Function bound) We have, if $\|y-x\|_x < 1$,
-    \[
-        A^*(y) \leq A^*(x) + \nabla A^*(x)(y-x) + \omega^*(\|y-x\|_x)
-    \]
-\end{proposition}
-
-The last proposition implies that, when $\|y-x\|_x < 1$,
-\[
-    \mathcal{B}_{A^*}(y,x) \leq \omega^*(\|y-x\|_x).
-\]
-Therefore,
-\[
-    \mathcal{B}_{A^*}(\mu^*,\mu_n) \leq \omega^*(\|\mu^*-\mu_n\|_{\mu_n}) .
-\]
-then, if $\|\mu^*-\mu_n\|_{\mu^*} < 1$, we have
-\[
-    \mathcal{B}_{A^*}(\mu^*,\mu_n) \leq \omega^*\left(\frac{\|\mu^*-\mu_n\|_{\mu^*}}{1-\|\mu^*-\mu_n\|_{\mu^*}}\right) .
-\]
-Assume $\|\mu^*-\mu_n\|_{\mu^*} < c$,
-\[
-    \mathcal{B}_{A^*}(\mu^*,\mu_n) \leq \omega^*\left(\frac{\|\mu^*-\mu_n\|_{\mu^*}}{1-c}\right) .
-\]
-
-\subsection{not good}
-However,
-\begin{align*}
-    \|\mu^*-\mu_n\|^2_{\mu_n} & = \left\|\frac{n+n_0-1}{n+n_0}(\mu^*-\mu_{n-1}) + \frac{1}{n+n_0}(\mu^*-T_n)\right\|^2_{\mu_n},\\
-    & = \left\|\frac{n+n_0-1}{n+n_0}(\mu^*-\mu_{n-1})\right\|_{\mu_n}^2 +  \left\|\frac{1}{n+n_0}(\mu^*-T_n)\right\|^2_{\mu_n} + \frac{2(n+n_0+1)}{(n+n_0)^2}\langle \nabla^2 [A(\mu_n)](\mu^*-\mu_{n-1}),\,(\mu^*-T_n) \rangle,\\
-\end{align*}
-and
-\begin{align*}
-    \|\mu^*-\mu_n\|_{\mu_n} & \leq  \frac{n+n_0-1}{n+n_0}\left\|\mu^*-\mu_{n-1}\right\|_{\mu_n} + \frac{1}{n+n_0}\left\|\mu^*-T_n\right\|_{\mu_n} .
-\end{align*}
-Moreover, by the bounded Hessian change,
-\[
-    \left\|\mu^*-\mu_{n-1}\right\|_{\mu_n} \leq \frac{1}{1-\left\|\mu_n-\mu_{n-1}\right\|_{\mu_{n-1}}}\left\|\mu^*-\mu_{n-1}\right\|_{\mu_{n-1}}
-\]
+\bibliography{../references.bib}
 

### 2021-10-02 aistats/paper.tex
Some notes on stuff to cite and arguments we could make

@@
 \newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
+\newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
 \newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}
@@
 \newcommand{\lr}{\gamma} % learning rate, or step-size
+\newcommand{\lin}[1]{\left\langle#1\right\rangle}
 
@@
 \section{Introduction and Background}
+
+\fdk{Consider the optimization of 
+\begin{equation}
+	f(\theta) = \expect[x \sim p(x)]{\frac{1}{2}\norm{\theta}^2 - \lin{x, \theta}},
+\end{equation}
+given samples from $x$. We know how to do that one, 
+and the worst-case scenario happens when $p(x)$ is a Gaussian.
+Now how about the slightly more general case where
+\begin{equation}
+	f(\theta) = \expect[x \sim p(x)]{A(\theta) - \lin{x, \theta}}
+\end{equation}
+and we have full knowledge of $A$, a strictly convex twice differentiable function. 
+How hard can that be?
+Turns out, if $p$ is the maximum entropy distribution with 
+$\expect[x \sim p(x)]{x} = \mu$, 
+then it's a bit hard.}
+
 \RLP{Goal = disseminate related work to make it look nice.}
@@
 
+\clearpage
+\section{Stuff to cite and why }
+
+\begin{itemize}
+	\item Stochastic and online mirror descent: 
+	We obviously have rates in this setting, but they all assume 
+	that the reference function is strongly-convex relative to a norm. 
+	\citep[e.g.][Thm 4.2]{bubeck2015convex}
+	\item What is mirror descent called in the literature?
+		\begin{itemize}
+			\item Mirror Descent \citep{nemirovski1983problem,beck2003mirror} 
+			\item NoLips \citep{bauschke2017descent}
+		\end{itemize}
+	\item Interest in mirror descent boosted by the relatively smooth view 
+	of \citet{bauschke2017descent,lu2018relatively}	
+	\item 
+	The oroblems on how to deal with Bregman divergences abound in optimization, 
+	beyond stochasticity. 
+	For example, we haven't yet figured out the analog of Nesterov-type acceleration 
+	on relatively smooth and strongly convex problems, 
+	to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
+	or just from $1/T$ to $1/T^2$ 
+	in the (non-strongly) convex case.
+	\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
+	The tools developped to make progress on one problem might help make progress on the other. 
+	\item 
+	Stochastic mirror descent in the relatively smooth setting;
+	\citet{hanzely2018fastest},
+	\citet{dragomir2021fast}
+	and the zoo of assumptions
+	\item Something about the statistical asymptotic behavior for the MLE? 
+	\begin{equation}
+		D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
+		\approx \frac{1}{n}?
+	\end{equation}
+	\item Something about how the Bregman divergence is often not considered in the statistics lit 
+	because it can be infinite and is non-symmetric and not nice to work with, 
+	and so folks focus on $\expect{\norm{\theta_n - \theta_*}^2}$ even if it's somewhat meaningless 
+	in the non-asymptotic regime, 
+	and the best non-asymptotic results 
+	still rely on Taylor expansions after $n > $ something big?
+	\item Something about how we don't even know how to deal with decreasing step-sizes 
+	of the order of $1/t$ 
+	\item 
+	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions 
+	that obtain the $1/t$ rate in the smooth, convex setting instead of $1/\sqrt{t}$, 
+	which is the ``expected'' rate for that setting, 
+	and how this approach above is a generalization beyond the Euclidean noise model
+
+\end{itemize}
+
 \end{document}

### 2021-10-03 aistats/paper.tex
Complement the intro

@@
 \usepackage{graphicx}
-\graphicspath{{../figs/}}
+\graphicspath{{../figs/}{../}}
 %\usepackage{subfigure}
@@
 \begin{abstract}
-We raise the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate, or a conjugate maximum a posteriori for the exponential family. 
-Surprisingly, we found no solution to this problem in the literature -- we are not able to tell how many samples we need to fit a gaussian within a few bits of the true distribution.
+We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for the exponential family. 
+Surprisingly, we found no solution to this problem in the literature -- e.g. after seeing 5 samples, we do not know how many bits away (in expectation) our model is from the true distribution.
 After displaying some properties and special cases of this problem, 
-we show it is a special case of several optimization algorithms, but it falls out of their scopes, thus highlighting range of progress in the analysis of these algorithms.
+we show it is an application of several optimization algorithms.
+Yet it falls out of scope for current analysis, thus highlighting areas for progress.
 \end{abstract}
-
-\section{Plan}
-
-\begin{enumerate}
-	\item intro to exponential family, and density estimation
-	\item the thing we want to bound
-	\item Examples : gaussian mean and gaussian variance (+other examples, just mentioned)
-	\item Insight : Strongly convex case. (+ self-concordance that is not verified either)
-	\item insight : bias-variance decomposition
-	\item Optimization perspective : SBPP or SBG. But no analysis hold, revealing a flaw of all these techniques.
-	\item Discussion : we believe finding a convergence rate would bring new tools useful to deal with common objects such as barrier losses.
-\end{enumerate}
-
-Open questions
-\begin{itemize}
-	\item does a base measure change anything ?
-	\item is there multiple conjugate priors ?
-	\item misspecified case. Do we have a   formula ?
-	\item make sure the gaussian entropy is not SC.
-\end{itemize}
 
@@
 
+Call both to stats and optimization crowds by introducing the parallel quite early.
+
 \RLP{Goal = disseminate related work to make it look nice.}
@@
 They are literally the linear model of probabilities.
-The exponential family for data $X\in\cX$  with sufficient statistic $T$ and natural parameter $\natp$ is the model 
+The exponential family for data $X$ with natural parameter $\natp$  is entirely specified by the support set of the data   $\cX$ and a sufficient statistic $T: \cX \rightarrow \real^d$
 \begin{equation}
@@
 where the integral stands for a sum if $x$ has discrete support. \RLP{Use base measure instead.}
-Note that an exponential family is entirely specified by its support set $\cX$ and its sufficient statistic $T$.
-This simple model encompasses both categorical distributions $\cX = \{1, \dots, k\}$ with $T(X)$ being the one-hot encoding, and multivariate normal distributions $\cX=\real, T(X)=(X, X^2)$.
+For instance, this simple model encompasses both categorical distributions $\cX = \{1, \dots, k\}$ with $T(X)$ being the one-hot encoding, and multivariate normal distributions $\cX=\real, T(X)=(X, X^2)$.
 
@@
 If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
+We will write indifferently $\mu$ or  $\natp$ depending on the context, being aware that both represent the same distribution.
 The second identity entails that $\logpart$ is strictly-convex. 
@@
 \end{align}
-It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
+which matches the common notion of \textit{entropy} in information theory.
 If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
@@
 	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$.
-Figure reproduced from \citet{kunstner2020homeomorphic}.}
+Figure reproduced from \citet{kunstner2020homeomorphic}.
+		\RLP{Show the true gaussian bijection instead?}
+	}
 	\label{fig:duality}
 \end{figure}
+
+\paragraph{Conjugate Prior}
+One conjugate prior for $p(X|\natp)$ is
+\begin{align}
+    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
+\end{align}
+where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
+\begin{align}
+    \bregman (\natp ; \natp_0)
+    & = \logpart(\natp) - \logpart(\natp_0) 
+    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
+\end{align}
+with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
+Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
+We can re-write this prior as 
+\begin{align}
+    p(\natp) \propto 
+    \exp( -n_0 \logpart (\natp) 
+    + \langle n_0 \mu_0, \natp \rangle ) \; ,
+\end{align}
+which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
+The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.
+
+\paragraph{Maximum A Posteriori (MAP).}
+The negative log-likelihood of the prior is
+\begin{align*}
+    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
+\end{align*}
+Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
+\begin{multline}
+    -\log p(\cD|\natp)p(\natp) 
+    = (n_0+n) \logpart (\natp) \\
+    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
+\end{multline}
+Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
+\begin{align}
+    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
+\end{align}
+such that the MAP is
+\begin{align}
+    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
+    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
+\end{align}
+When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
+\begin{align}
+	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
+\end{align}
+The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
 
@@
 \end{multline}
-For the exponential family, the KL is also the Bregman divergence induced by the log-partition function (with switched arguments)
-\begin{align}
-	\KL( p(.|\natp^*) ; p(.|\natp)) 
-	= \bregman (\natp ; \natp^*)  \; .
-\end{align}
-There is a general relationship between Bregman divergences and convex conjugates (notice the argument switching)
-\begin{align}
-	\bregman (\natp ; \natp^*)
-    %= \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
-    = \bregmanconj ( \meanp^* ; \meanp)
-\end{align}
-so in the end the suboptimality is a divergence, which can either be seen as a KL between distributions, as a divergence between natural parameters, or as a divergence between mean parameters
+For the exponential family, the KL is equal to the Bregman divergence induced by the log-partition function, or alternatively by the entropy, being careful with the order of the arguments 
 \begin{align}
@@
     = \bregman (\natp ; \natp^*)
-    = \bregmanconj ( \meanp^* ; \meanp) \; .
+    = \bregmanconj ( \meanp^* ; \meanp)
 }
 \end{align}
-
-The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
+Our question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities
 \begin{align}
 	\label{eq:bregmanMLE}
-	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
+	\expect[\cD]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \\
 	\label{eq:bregmanMAP}
-	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
+	\expect[\cD]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} 
 \end{align}
-where the outer expectation is on the dataset $X_1, \dots, X_n$?
+where the outer expectation is on the dataset $\cD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
 
 \paragraph{Remark.}
-What we are looking for is really akin to concentration inequality, expressed with a Bregman divergence instead of a norm.
+This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
 A key difference though, is that the random variable $T(X)$ is connected to the metric $\logpart$. 
@@
 
+	\begin{theorem}[MLE Tight Bound]
+	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
+	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
+	\begin{align}
+		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
+			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
+			\label{eq:MLE_rate}
+	\end{align}
+\end{theorem}
+
+This upper bound is asymptotically tight.
+We illustrate its behavior against empirical data in Figure~\ref{fig:curves}.
+ There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}.
+The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension.
+We do not perform these calculations for now as we want to focus on the simplest possible examples.
+
+As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
+To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
+Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
+We now present a lemma on its expected inverse.
+\begin{lemma}[Expected MAP Natural Parameter]
+	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+	\begin{equation}
+		\frac{\mu^*}{\mu_n}
+		\leq \expect{\frac{\mu^*}{\MAPm}} 
+		= \expect{\frac{\MAPt}{\natp^*}} 
+		\leq \frac{n_0 +n}{a+ (n-2)_+}
+		\label{eq:theta_expectation_bound}
+	\end{equation}
+	where $(x)_+ = \max(0,x)$.
+\end{lemma}
+Note that the lower bound is a simple consequence of the convexity of the inverse function. 
+Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP.
+
+\begin{theorem}[MAP Bound]
+ Let us  define
+ \begin{align}
+	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
+ \end{align}
+The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
+ \begin{equation}
+	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
+	\leq \begin{cases}
+		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
+		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
+	\end{cases}
+	\label{eq:MAP_rate}
+\end{equation}
+\end{theorem}
+
+This inequality highlights a clear variance-bias decomposition.
+In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
+For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
+Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$.
+We are basically losing a factor 2 compared to $1/2n$.
+
+Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
+We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
+
+\begin{corollary}[MLE Loose Bound] When $n\geq 2$, the MLE expected loss is upper bounded as
+\begin{equation}
+	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
+	\leq \inv{n-2} \; .
+	\label{eq:MLE_loose_rate}
+\end{equation}
+\end{corollary}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=.7\textwidth]{figs/MAP_rates/MAP_rate_n0=5.pdf}
+	\caption{
+	Convergence rate of MAP~\eqref{eq:MAP_rate} 
+	\label{fig:MAP_rate} against the ratio between initialization and optimum $ \frac{\mu_0}{\mu^*}$. 
+	We also report the tight convergence rate of MLE~\eqref{eq:MLE_rate} (dash-dotted horizontal lines), and the loose upper bound~\eqref{eq:MLE_loose_rate} (horizontal lines). 
+	We report these curves for various values of $n$, and $n_0=5$. 
+	We observe that the rate of MAP is significantly better (lower) than the rate of MLE, only when we guessed $\mu_0$ right, within an order of magnitude from $\mu^*$, and for a small number of sample ($n<10$).
+	This observation hints towards the optimality of MLE for large $n$. 
+	It makes me feel very interested in convergence rates of MAP in the overparametrized regime $n<\dim(T(X))$.
+	}
+\end{figure}
+
 \bibliographystyle{apalike}
@@
 \clearpage
+\section{Plan}
+
+\begin{enumerate}
+	\item intro to exponential family, and density estimation
+	\item the thing we want to bound
+	\item Examples : gaussian mean and gaussian variance (+other examples, just mentioned)
+	\item Insight : Strongly convex case. (+ self-concordance that is not verified either)
+	\item insight : bias-variance decomposition
+	\item Optimization perspective : SBPP or SBG. But no analysis hold, revealing a flaw of all these techniques.
+	\item Discussion : we believe finding a convergence rate would bring new tools useful to deal with common objects such as barrier losses.
+\end{enumerate}
+
+Open questions
+\begin{itemize}
+	\item does a base measure change anything ?
+	\item is there multiple conjugate priors ?
+	\item misspecified case. Do we have a   formula ?
+\end{itemize}
+
 \section{Stuff to cite and why }
@@
 	\item 
-	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions 
+	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions https://www.di.ens.fr/~fbach/bach_ejs_self_concordance.pdf
 	that obtain the $1/t$ rate in the smooth, convex setting instead of $1/\sqrt{t}$, 
@@
 	and how this approach above is a generalization beyond the Euclidean noise model 
-
+	http://proceedings.mlr.press/v99/marteau-ferey19a/marteau-ferey19a.pdf
+	http://proceedings.mlr.press/v99/ostrovskii19a/ostrovskii19a.pdf
 \end{itemize}

### 2021-10-04 aistats/paper.tex
attempt at a motivation

@@
 \newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
+\newcommand{\cond}{\,\vert\,}
 \newcommand{\logpart}{A}
@@
 
-\section{Introduction and Background}
+\section{Motivation}
 
-\fdk{Consider the optimization of 
+We kind of know how to do optimization 
+on smooth, strongly convex functions 
+when there is some bound on the stochasticity,
+for example in the form of a bounded variance, 
+say for the problem 
 \begin{equation}
-	f(\theta) = \expect[x \sim p(x)]{\frac{1}{2}\norm{\theta}^2 - \lin{x, \theta}},
+\min_\theta f(\theta)
 \end{equation}
-given samples from $x$. We know how to do that one, 
-and the worst-case scenario happens when $p(x)$ is a Gaussian.
-Now how about the slightly more general case where
+with access to stochastic gradients 
+$g(\theta)$ such that 
 \begin{equation}
-	f(\theta) = \expect[x \sim p(x)]{A(\theta) - \lin{x, \theta}}
+	\expect{\norm{g(\theta) - \nabla f(\theta)}^2} \leq \sigma^2.
 \end{equation}
-and we have full knowledge of $A$, a strictly convex twice differentiable function. 
-How hard can that be?
-Turns out, if $p$ is the maximum entropy distribution with 
-$\expect[x \sim p(x)]{x} = \mu$, 
-then it's a bit hard.}
+The best convergence rate, of stochastic gradient methods 
+is obtained by SGD with decreasing step-sizes and averaging, 
+yields a convergence rate of $1/t$ and is asymptotically tight;
+SGD achieves this rate on the following problem, 
+\begin{equation}
+	f(\theta) = \frac{1}{2}\norm{\theta -\mu}^2, \quad
+	g(\theta) \sim \sim \cN(\theta - \mu, \sigma^2).
+\end{equation}
+This problem is estimating the mean of a Gaussian, 
+and it is not improvable by information theoretic arguments.
 
-Call both to stats and optimization crowds by introducing the parallel quite early.
+But our understanding of stochastic optimization lacks in other domains. 
+A generalization of the above problem, 
+at the intersection of optimization and estimation, 
+is the problem of minimizing 
+\begin{equation}
+	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{A(\theta) - \lin{t(x), \theta}},
+\end{equation}
+where $A$ is a convex, Legendre function and 
+$p(x\cond\theta)$ is the exponential family distribution 
+with $A$ as its log-partition function. 
+The Gaussian example above is a special case 
+with $A(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$
+and $p(x\cond\theta) = \cN{\mu, \sigma^2}$.
+
+We obviously have algorithms for this setting, 
+as this is an maximum likelihood problem.
+Giving the (online) maximum likelihood after seeing $t$ samples
+which corresponds as online mirror descent with a decreasing step-size schedule (more on this later).
+But our convergence rates are weak.
+There are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
+
+We do not have a solution. 
+The goal of this paper, and the reason it is submitted at AISTATS, 
+is to bring this issue to the attention of 
+the community working at the intersection of machine learning, optimization and statistics community,
+in the hope that we can make progress on this question;
+Can we build algorithms for stochastic estimation 
+when the objective function is not (close to) quadratic
+and the noise is not (sub-)Gaussian?
+
+We give motivating examples where solution of this problem would enable progress in other areas,
+principally motivated by stochastic optimization, 
+show the connections between the optimization and statistical estimation sides of the problem,
+and discuss where our current theory fails to fully capture the problem.
+
+We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate---this is part of the point of this paper---and that the statistical theory results for the convergence of the MLE to the solution
+are well known. 
+Our point is more subtle, in that for the purposes of optimization, 
+to mimic the results that are known for SGD in Euclidean settings, 
+we would like a result of the form 
+\begin{equation*}
+	\text{optimality gap after $t$ steps}
+	\leq 
+	\frac{\text{``bias''}}{t^2}
+	+ \frac{\text{``variance''}}{t},
+\end{equation*}
+which is the bit that is lacking.
+
+Then; 
+- description of the setting
+- statistical solutions (asymptotic, l1 or l2 error)
+- optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
+- potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
+- conclusion?
+
+\section{Technical background}
 
@@
 	\item 
-	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions https://www.di.ens.fr/~fbach/bach_ejs_self_concordance.pdf
+	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions 
+	\url{https://www.di.ens.fr/~fbach/bach_ejs_self_concordance.pdf}
 	that obtain the $1/t$ rate in the smooth, convex setting instead of $1/\sqrt{t}$, 
@@
 	and how this approach above is a generalization beyond the Euclidean noise model 
-	http://proceedings.mlr.press/v99/marteau-ferey19a/marteau-ferey19a.pdf
-	http://proceedings.mlr.press/v99/ostrovskii19a/ostrovskii19a.pdf
+	\url{http://proceedings.mlr.press/v99/marteau-ferey19a/marteau-ferey19a.pdf}
+	\url{http://proceedings.mlr.press/v99/ostrovskii19a/ostrovskii19a.pdf}
 \end{itemize}

### 2021-10-04 main.tex
Add multivariate gaussian variance

@@
 
+\usepackage{tikz}           % define for loops
 % frames with colored background
@@
 \paragraph{Conjugate Prior}
-\TODO{Under-parametrized conjugate prior ?}
 One conjugate prior for $p(X|\natp)$ is
@@
  There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}.
-The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension.
-We do not perform these calculations for now as we want to focus on the simplest possible examples.
+The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and we report a similar upper bound in the appendix when $n>d+1$.
 
@@
 
-\TODO{add equations for other families, eg gamma or Dirichlet, or Categorical..}
-
 \section{Euclidean Behaviour}
@@
 \begin{align}
-	\E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{n} \Tr(\mM \mSigma)
+	\E \half \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}_\mM^2 = \inv{2 n} \Tr(\mM \mSigma)
 \end{align}
@@
 \begin{align}
-	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
-	&= \frac{n d}{(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
-	&= \frac{d}{n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
+	\E \half \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
+	&= \frac{n d}{2(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \half \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
+	&= \frac{d}{2n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
 \end{align}
@@
 	\item The first one is the variance definition which always end up being infinite when asked to be uniform on every possible trajectories, if the log-partition  has some barrier structure.
+Except for \citet{dorazio2021stochastic} which holds in many cases.
 	\item The second one is the step-size scheduling.
-$O(\inv{t})$ is decreasing too fast for most proof techniques.
+$O(\inv{t})$ is decreasing too fast for most proof techniques, except \citet{dragomir2021fast}.
 Even in the Euclidean setting where the variance does not matter, it results in a disappointing rate $O(\inv{\log t})$.
@@
 \end{align}
-Now let us discover the entropy
+Now let us recover the entropy
 \begin{align}
@@
 In this case, the Bregman divergences induced by the log-partition or the entropy are ugly, so we will not focus on them.
+Note that this entropy is not self-concordant. Taking derivatives in the direction of $\mu_1$, and setting $\mu = (4, 16)$,  the self-concordance inequality is not verified.
 
@@
 This highlights the factor $2$ difference between the log-bound and an exact computation of the expectation of the log term, and a tight upper bound on the digamma function.
+
+\subsection{Multivariate MLE}
+In higher dimensions $d$
+ \begin{align}
+ 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
+ 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
+ \end{align}
+where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
+Using the same upper bound as in the univariate case
+\begin{align}
+	\psi_d(\half[n]) - d \log(\half[n]) 
+	\leq \sum_{i=0}^{d-1} \log(1 - \frac{i}{n}) - \frac{1}{n-i} \; .
+\end{align}
+We can bound the harmonic sum with typical bounds
+\begin{align}
+	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
+	&= H_{n-d} - H_n \\
+	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
+	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
+\end{align}
+which yields 
+\begin{align}
+	\psi_d(\half[n]) - d \log(\half[n]) 
+	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
+	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}  \; .
+\end{align}
+From there we have two options. 
+Either we use a concave tangent bound on the logarithms to get
+\begin{align}
+	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
+	\leq \sum_{i=0}^{d} - \frac{i}{n}
+	 = -\frac{d(d+1)}{2n}
+\end{align}
+which yields the final rate
+\begin{align}
+ 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
+ 	\leq \frac{d(d+1)}{4n} + \frac{d(d+1)^2 + \half[d+1]}{2 n (n - d -1)} \;.
+\end{align}
+Note that this rate is not tight for $d>1$, but it is tight for $d=1$, and we almost recover the previous rate.
+
+Another option, potentially tighter than the tangent bound, is to bound the sum of logarithms with an integral, using the monotonicity of the logarithm $\log(1 - \frac{i}{n}) \leq \log(1- \frac{x}{n}) , \forall x \in [i - 1, i]$
+\begin{align}
+	\sum_{i=1}^{d} \log(1 - \frac{i}{n})
+	&\leq \int_{0}^{d} \log(1- \frac{x}{n}) dx \\
+	& = n \int_{1 - \frac{d}{n}}^{1} \log(y) dy \\
+	& = n [ y (\log(y) - 1) ]_{1 - \frac{d}{n}}^{1} \\
+	& =  - n - (n -d) (\log ( 1 - \frac{d}{n}) - 1) \\
+	& = - (n -d) \log ( 1 - \frac{d}{n}) - d \; .
+\end{align}
+Unfortunately, it is hard to get a tractable and non-vacuous bound for this guy.
+At $n=d$ it is worth $-d$, and then it grows up to $0$.
+and we can bound it using some exponents of $1 - \frac{d}{n}$, but that is not really helpful to get an easy formula.
+Actually looking at \href{https://www.desmos.com/calculator/q4sqk73tld}{graphics}, the first upper bound turns out to be tighter than the second one.
 

### 2021-10-04 aistats/paper.tex
Clean up gaussian variance section

@@
 
-\paragraph{Conjugate Prior}
-One conjugate prior for $p(X|\natp)$ is
-\begin{align}
-    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
-\end{align}
-where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
+\paragraph{Bregman Divergences}
+A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\natp$ and $\natp_0$
 \begin{align}
@@
 with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
+It is a non-symmetric measure of distance between parameters.
+It is equal to the divergence of its convex conjugate with switched arguments
+\begin{align}
+	\bregman (\natp ; \natp_0)
+    = \bregmanconj ( \meanp_0 ; \meanp) \; .
+\end{align}
+
+\paragraph{Conjugate Prior}
+One conjugate prior \citep{agarwal2010geometric} for $p(X|\natp)$ is
+\begin{align}
+    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
+\end{align}
+where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is introduced above.
 Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
-We can re-write this prior as 
-\begin{align}
-    p(\natp) \propto 
-    \exp( -n_0 \logpart (\natp) 
-    + \langle n_0 \mu_0, \natp \rangle ) \; ,
-\end{align}
-which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
-The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.
-
-\paragraph{Maximum A Posteriori (MAP).}
 The negative log-likelihood of the prior is
@@
 \end{align*}
-Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
+which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
+Then the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
 \begin{multline}
@@
     = (n_0+n) \logpart (\natp) \\
-    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
+    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst
+    \label{eq:joint_likelihood}
 \end{multline}
-Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
+meaning the posterior given $\cD=(X_1,\dots, X_n)$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.
+
+\paragraph{Maximum A Posteriori (MAP).}
+Minimizing~\eqref{eq:joint_likelihood} over $\natp$ yields the MAP estimate
 \begin{align}
@@
 \end{align}
-such that the MAP is
+verifying the first oder optimality conditions \RLP{$\natp$ unconstrained, living in an open set with barriers}
 \begin{align}
@@
 \end{align}
+These rules are known as moment matching.
 The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
@@
 \subsection{Gaussian Variance}
-The trailing example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
-The density of a centered normal variable is
+A core example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$. 
+The density is	$p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}$.
+Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
+Mean and natural parameters are roughly inverse of each other $\natp = -\inv{2 \mu}$.
+Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
 \begin{align}
-	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}} \; .
+	\logpart (\natp) &= - \half \log(-\natp)  + \half \log(\pi) \\
+	\conj(\mu) &= \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
 \end{align}
-Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
-Mean and natural parameters are roughly inverse of each other
+Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
+It means the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative Gamma distribution.
+It also means $\bregman$ and $\bregmanconj$ have the same shape
 \begin{align}
-	\natp = -\inv{2 \mu}\; .
-\end{align}
-Now we can match the log-likelihood with the exponential family template to get the log-partition function.
-\begin{align}
-	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
-	= x^2 \natp - \logpart(\natp) \\
-	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
-\end{align}
-
-We can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
-\begin{align}
-	\conj(\mu) = \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
-\end{align}
-We can also take gradient and derivative of $\logpart(\natp)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
-\begin{align}
-	\nabla\logpart(\natp) &= \frac{-1}{2 \natp} = \sigma^2 = \mu = \E[X^2] \\
-	\nabla^2\logpart(\natp) &= \frac{1}{2 \natp^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
-\end{align}
-which we confirm thank to wikipedia since $\E[X^4] = 3 \sigma^4$ and thus $\Var(X^2) = \E[X^4] - \E[X^2]^2 = 3 \sigma^4 - \sigma^4 = 2 \sigma^4$.
-
-The logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
-In particular,
-\begin{align}
-	p(\natp) 
-	&\propto 
-    \exp( -n_0 \logpart (\natp) 
-    + \langle n_0 \mu_0, \natp \rangle ) \\
-    &\propto \exp( \half[n_0] \log(-\natp) + n_0\mu_0 \natp ) \\
-	&\propto (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
-\end{align}
-from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\natp \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$.
-After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.
-
-Both the entropy and the log-partition are roughly negative logarithm $z\mapsto - \log(z)$. Which yields the same shape of Bregman divergence, as visible below (all three lines are equal)
-\begin{align}
-	\KL( \sigma_*^2 ; \sigma_n^2 ) 
-	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
 	\bregmanconj( \mu_*; \mu_n) 
@@
 \end{align}
-
 In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
 \begin{align}
-	\phi(z) := \half (z - 1 - \log(z)) \\
-	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
+	\phi(z) := \half (z - 1 - \log(z))
 \end{align}
-as illustrated in Figure~\ref{fig:phi}. We can get a non-transcendental upper bound thanks to the inequality
-\begin{align}
-	1 - \inv{z} \leq \log(z) 
-	\label{eq:log_bound} 
-	\implies \phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
-\end{align}
+illustrated in Figure~\ref{fig:phi}. 
+Below we report upper bounds on the expected value of this divergence for the MLE and the MAP.
 
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{bregmandef.pdf}
 	\includegraphics[width=.4\textwidth]{phi.pdf}
-	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics.}
+	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics, but it admits another upper-bound (in grey).}
 	\label{fig:phi}
@@
 	\centering
-\includegraphics[width=.48\textwidth]{asymptote.pdf}
-\includegraphics[width=.48\textwidth]{fewsamples.pdf}
+\includegraphics[width=.4\textwidth]{fewsamples.pdf}
 	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$.
 Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval.
-	\textbf{Left:} as $n$ increases, the suboptimality matches the $1/2N$ asymptote.
-	\textbf{Right:} the first few samples significantly deviate from this behavior.
-In fact, for $n=1$ and $n=2$, the expected value is infinite, but we have a closed form solution and a simple upper-bound for $n>2$.
+		While the expected value is infinite for $n=1$ or $n=2$, it quickly matches the upper bound, and the $1/2n$ asymptote.
 	}
@@
 This upper bound is asymptotically tight.
-We illustrate its behavior against empirical data in Figure~\ref{fig:curves}.
- There is also a closed form for the multivariate generalization, thanks to the \href{https://en.wikipedia.org/wiki/Inverse-Wishart_distribution}{inverse Wishart distribution} and the \href{https://en.wikipedia.org/wiki/Wishart_distribution#Log-expectation}{expectation of the log-determinant of a Wishart}.
-The expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension.
-We do not perform these calculations for now as we want to focus on the simplest possible examples.
+We illustrate its numerical behavior in Figure~\ref{fig:curves}.
+We get a similar bound for the multivariate generalization :
+the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.
 
-As for the MAP, we did not manage to get an asymptotically tight upper bound. Nevertheless, using inequality~\eqref{eq:log_bound}, we did get an interesting upper bound. 
-To start, let us recall that the MAP of $\cN(0,\mu_*)$ is $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i X_i^2}{n_0 + n}$.
-Its expectation is simply $\mu_n= \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$.
-We now present a lemma on its expected inverse.
-\begin{lemma}[Expected MAP Natural Parameter]
-	Let's define $a=n_0 \frac {\mu_0 }{ \mu^* }$. For any $n\geq 1$, the expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
-	\begin{equation}
-		\frac{\mu^*}{\mu_n}
-		\leq \expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\natp^*}} 
-		\leq \frac{n_0 +n}{a+ (n-2)_+}
-		\label{eq:theta_expectation_bound}
-	\end{equation}
-	where $(x)_+ = \max(0,x)$.
-\end{lemma}
-Note that the lower bound is a simple consequence of the convexity of the inverse function. 
-Using this lemma, along with the log upper bound on the logarithm~\eqref{eq:log_bound}, we get a convergence rate for the MAP.
+For the MAP, we get an upper bound thanks to the inequality $ - \log(z) \leq \inv{z} - 1$ which implies
+\begin{align}
+	\label{eq:log_bound} 
+	\phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
+\end{align}
 
 \begin{theorem}[MAP Bound]
- Let us  define
+ For $n\in \naturalnumbers$, let us  define
  \begin{align}
-	b = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
+	b_n = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
  \end{align}
@@
 	\leq \begin{cases}
-		\inv{2(n_0+1)}  +  b\ \text{if}\ n=1,\\
-		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b \ \text{if}\ n\geq 2
+		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
+		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b_n \ \text{if}\ n\geq 2
 	\end{cases}
@@
 We are basically losing a factor 2 compared to $1/2n$.
-
-Using the same bound on the log~\eqref{eq:log_bound}, we  get a loose bound on the MLE, which is useful for comparison purposes.
-We relate these loose convergence rates of MAP and MLE in Figure~\ref{fig:MAP_rate}.
-
-\begin{corollary}[MLE Loose Bound] When $n\geq 2$, the MLE expected loss is upper bounded as
-\begin{equation}
-	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
-	\leq \inv{n-2} \; .
-	\label{eq:MLE_loose_rate}
-\end{equation}
-\end{corollary}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.7\textwidth]{figs/MAP_rates/MAP_rate_n0=5.pdf}
-	\caption{
-	Convergence rate of MAP~\eqref{eq:MAP_rate} 
-	\label{fig:MAP_rate} against the ratio between initialization and optimum $ \frac{\mu_0}{\mu^*}$. 
-	We also report the tight convergence rate of MLE~\eqref{eq:MLE_rate} (dash-dotted horizontal lines), and the loose upper bound~\eqref{eq:MLE_loose_rate} (horizontal lines). 
-	We report these curves for various values of $n$, and $n_0=5$. 
-	We observe that the rate of MAP is significantly better (lower) than the rate of MLE, only when we guessed $\mu_0$ right, within an order of magnitude from $\mu^*$, and for a small number of sample ($n<10$).
-	This observation hints towards the optimality of MLE for large $n$. 
-	It makes me feel very interested in convergence rates of MAP in the overparametrized regime $n<\dim(T(X))$.
-	}
-\end{figure}
+\RLP{Add numerical experiments with upper bound for gaussian variance MAP.}
 

### 2021-10-05 main.tex
Add new drawings to manuscript

@@
 	\centering
-	\includegraphics{figs/duality}
-	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$.
-Figure reproduced from \citet{kunstner2020homeomorphic}.}
+	\includegraphics[width= .6\textwidth]{figs/duality} \\
+	\includegraphics[width=.7\textwidth]{figs/2d/mirrormap.pdf}
+	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$.Top figure reproduced from \citet{kunstner2020homeomorphic}.
+Bottom figure represents $\cN(\mu, \sigma^2)$.}
 	\label{fig:duality}
@@
 	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
+	\includegraphics[width=\textwidth]{figs/2d/gaussian_convergence.pdf}
+	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$, for 1 time step  and many repetitions with $n=3$ (top) illustrating the effect of $n_0$ with the Thales triangle.
+Or $50$ time-steps at once and 1 repetition (bottom), illustrating the convergence behavior.}
 	\label{fig:bias-variance-numerical}

### 2021-10-05 aistats/paper.tex
Add to aistat examples and insights sections

@@
 %\setlength{\pdfpagewidth}{8.5in}
+
+\let\oldsection\section
+\renewcommand{\section}[1]{\oldsection{\uppercase{#1}}}
 
@@
 where the outer expectation is on the dataset $\cD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
+More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.
 
@@
 
-\section{Examples}
-\subsection{Gaussian Mean}
+\section{Insight - Asymptote}
+As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
+We can get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
+    + O(\norm{\mu - \mu^*}^3)
+\end{align}
+where the norm  $\| x \|_\mM^2 = x^\top \mM x$  is induced by the hessian at the optimum 
+\begin{align}
+    \nabla^2\conj(\mu^*) 
+    = \nabla^2\logpart(\natp^*)^{-1} 
+    = \Cov_{\natp^*}[T(X)]^{-1}  \; .
+\end{align}
+Both the MLE and the MAP have the same asymptotic rate
+\begin{align}
+\label{eq:MAP_asymptote}
+	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
+	= \frac{d}{2n} + O(n^{- \frac{3}{2}})
+\end{align}
+as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data.
+Remark how this rate is also independent of the optimum $\meanp^*$.
 
-\subsection{Gaussian Variance}
+\section{Example - Gaussian Mean}
+
+\paragraph{$\ell^2$-norm Analysis}
+In the Euclidean realm, the average is bounded by variance becomes a bound on the variance of the average
+\begin{align}
+	\Var T(X) = \E \norm{T(X) - \mu^*}^2 = \sigma^2 \\
+	\implies \E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
+\end{align}
+eg the variance of the mean is $n$ times smaller than the variance of the samples.
+Adding a reference mean $\mu_0$ to get the MAP yields
+\begin{align}
+	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
+	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
+	&= O\left(\frac{\sigma^2}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
+\end{align}
+so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !
+
+\section{Insight - Euclide's World}
+As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as the mean of a gaussian.
+\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    &\leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu} \\
+    &\leq 2L \norm{\mu^* - \mu}
+\end{align}
+so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{t}})$ convergence rate from \RLP{Gaussian rate}.
+\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
+\end{align}
+so $\bregmanconj$ is upper bounded by a quadratic, and we get the same rate as a Gaussian.
+
+\section{Example - Gaussian Variance}
 A core example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$. 
@@
 
+\begin{figure*}[t]
+	\centering
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
+	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
+	\label{fig:bias-variance-numerical}
+\end{figure*}
+
+\section{Bias-Variance Decomposition}
+In both examples (add link), the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
+In fact, it is possible to write such a decomposition for any Bregman divergence
+Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
+\begin{align}
+	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	&= {\bregmanconj(\mu^* ; \tilde \mu_n)}
+	+ {\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
+	\label{eq:primal_pivot}
+\end{align}
+
+\paragraph{Illustrations.}
+We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
+In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
+\RLP{Tune figures.}
+
+\begin{figure*}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/biasvariance/linear_ratio=0.8_n0=10}.pdf}
+	\caption{
+	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
+	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
+	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
+	}
+	\label{fig:variance_decomposition}
+\end{figure*}
+
+\begin{figure*}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/gaussians/new_linear_n0=1}.pdf}
+	\caption{
+	\textbf{Full gaussian} $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. \textbf{Left:} training curves (5,95) percentiles, average and asymptote. 
+	\textbf{Right:} bias-variance decomposition.
+	}
+	\label{fig:gaussian_decomposition}
+\end{figure*}
+
 \bibliographystyle{apalike}

### 2021-10-06 aistats/paper.tex
Remove mixed term from biasvariance figures
attempt at clarification for intro

@@
 \end{equation}
-The best convergence rate, of stochastic gradient methods 
-is obtained by SGD with decreasing step-sizes and averaging, 
-yields a convergence rate of $1/t$ and is asymptotically tight;
-SGD achieves this rate on the following problem, 
-\begin{equation}
-	f(\theta) = \frac{1}{2}\norm{\theta -\mu}^2, \quad
-	g(\theta) \sim \sim \cN(\theta - \mu, \sigma^2).
-\end{equation}
-This problem is estimating the mean of a Gaussian, 
-and it is not improvable by information theoretic arguments.
+Estimating the mean of Gaussian noise is a special case of this problem class, using 
+$f(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$ and $g(\theta) \sim \cN(\theta - \mu, \sigma^2)$.
+SGD with a decreasing step-size coincides with the maximum likelihood estimate (MLE), 
+and attains the $1/t$ convergence rate.
 
 But our understanding of stochastic optimization lacks in other domains. 
-A generalization of the above problem, 
-at the intersection of optimization and estimation, 
+A generalization of the above problem, at the intersection of optimization and statistics, 
 is the problem of minimizing 
 \begin{equation}
-	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{A(\theta) - \lin{t(x), \theta}},
+	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta_*)},
 \end{equation}
-where $A$ is a convex, Legendre function and 
-$p(x\cond\theta)$ is the exponential family distribution 
-with $A$ as its log-partition function. 
-The Gaussian example above is a special case 
-with $A(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$
-and $p(x\cond\theta) = \cN{\mu, \sigma^2}$.
+When $p(x\cond\theta)$ is an exponential family distribution,
+$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (up to constants).
+Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
+The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.
 
-We obviously have algorithms for this setting, 
-as this is an maximum likelihood problem.
-Giving the (online) maximum likelihood after seeing $t$ samples
-which corresponds as online mirror descent with a decreasing step-size schedule (more on this later).
-But our convergence rates are weak.
-There are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
+There are algorithms for this setting, for example the MLE;
+giving the (online) maximum likelihood after seeing $t$ samples
+corresponds to online mirror descent with decreasing step-sizes (more on this later).
+But our convergence rates are weak; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
 
-We do not have a solution. 
-The goal of this paper, and the reason it is submitted at AISTATS, 
-is to bring this issue to the attention of 
-the community working at the intersection of machine learning, optimization and statistics community,
+We do not have a solution. The goal of this paper, and the reason it is submitted at AISTATS, 
+is to bring this issue to the attention of the community working at the intersection of machine learning, optimization and statistics,
 in the hope that we can make progress on this question;
@@
 when the objective function is not (close to) quadratic
-and the noise is not (sub-)Gaussian?
+and the noise is not Gaussian?
 
-We give motivating examples where solution of this problem would enable progress in other areas,
-principally motivated by stochastic optimization, 
-show the connections between the optimization and statistical estimation sides of the problem,
+We show the connections between the optimization and statistical estimation view of the problem
 and discuss where our current theory fails to fully capture the problem.
-
-We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate---this is part of the point of this paper---and that the statistical theory results for the convergence of the MLE to the solution
-are well known. 
-Our point is more subtle, in that for the purposes of optimization, 
-to mimic the results that are known for SGD in Euclidean settings, 
+We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate---this is part of the point of this paper---and 
+that the statistical theory results for the convergence of the MLE to the solution are well known. 
+Our point is more subtle, in that for the purposes of optimization, to mimic the results that are known for SGD in Euclidean settings, 
 we would like a result of the form 
@@
 	\centering
-	\includegraphics[width=\textwidth]{{figs/biasvariance/linear_ratio=0.8_n0=10}.pdf}
+	\includegraphics[width=\textwidth]{{figs/biasvariance/new_linear_ratio=0.8_n0=10}.pdf}
 	\caption{

### 2021-10-06 aistats/paper.tex
attempt to dump a few optim papers

@@
 
-Then; 
-- description of the setting
-- statistical solutions (asymptotic, l1 or l2 error)
-- optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
-- potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
-- conclusion?
+% Then; 
+% - description of the setting
+% - statistical solutions (asymptotic, l1 or l2 error)
+% - optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
+% - potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
+% - conclusion?
 
@@
 
+\fdk{
+Might want to note that a perhaps ``typical'' view of parameter estimation 
+works with norms, e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ (for some norm and some power), 
+or more generally using distances rather than divergences as they are symmetric and have nicer properties. 
+But as the MLE is the minimizer of the log-likelihood, 
+using the log-likelihood (and by extension the KL divergence) as a measure of risk is natural, if inconvenient.
+}
+
 For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
@@
 Remark how this rate is also independent of the optimum $\meanp^*$.
+\fdk{meaning "independent ignoring constants"?}
+
+\fdk{
+That should be well known, maybe find a reference?
+% Something about the statistical asymptotic behavior for the MLE? 
+% \begin{equation}
+% 	D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
+% 	\approx \frac{1}{n}?
+% \end{equation}
+}
 
@@
 
+\section{Optimization related work}
+In the context of optimization, why do we care?
+
+There is a long line work trying the convergence rate of stochastic gradient descent.
+Beyond the results of RM51, 
+we now have proofs n smooth, strongly convex problems and bounded variance 
+(as well as other assumptions on the noise, reviewed later)
+a decreasing step-size and an averaging scheme gets 
+a $1/t$ convergence rate, 
+which matches the asymptotic rate of unbiased estimation,
+through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler}\footnote{
+Those papers assume the stochastic gradients are bounded, 
+but small modifications also work for bounded variance instead.
+}
+
+We do have extensions to other notions of bounded variance, 
+for example assuming that the stochastic gradients are gradients of a perturbed smooth function $f_i$ 
+and that the minimum of $f$ and the minima of the $f_i$ 
+are bounded, $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$,
+or that the gradient noise is bounded only at the minimum. 
+For a review, see \citet{gower2019sgd}.
+
+However, for some problems, combining decreasing step-sizes and averaging is not necessary,
+even when the function is not strongly convex.
+This is the case for maximum likelihood estimation and matches the asymptotic rates, 
+but holds more generally, for example for linear and logistic regressions \citep{bach2013non}.
+
+There has been recent advances in optimization that go beyond the Euclidean case, 
+to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
+neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
+\citep{bauschke2017descent,lu2018relatively},
+where instead of measuring strong-convexity and smoothness relative to the Euclidean norm (or any norm), 
+we use a reference function instead \fdk{maybe add a line}
+and apply mirror descent \citep{nemirovski1983problem,beck2003mirror}
+(also called NoLips \citep{bauschke2017descent}, something else?).
+
+We have convergence results for the 
+application of stochastic mirror descent, 
+but they typically rely on assumptions that are too strong. 
+\begin{itemize}
+	\item The reference function is strongly convex [typical setup in online learning]
+	\item The variance is uniformly bounded
+\end{itemize}
+Some advances have been tried
+\begin{itemize}
+	\item The reference function is ``close to'' a norm
+	and looks like a \citet{hanzely2018fastest}
+	\item 
+	Attempt to adapt the ``bounded gradient at the minimum''
+	\citep{dragomir2021fast}
+\end{itemize}
+
+Other to cite;
+\begin{itemize}
+	\item \citet{marteauferey2019}
+\end{itemize}
+
+\fdk{
+A ``simple open problem'';
+assume we have a deterministic problem, so that we know where the minimum is. 
+Can we figure out a (deterministic) path from $\theta_0$ to $\theta_*$ 
+that has optimality decreasing as $1/n^2$, 
+to mirror the bias term of gradient descent?
+}
+
+\fdk{
+Application
+
+The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity. 
+For example, we haven't yet figured out the analog of Nesterov-type acceleration 
+on relatively smooth and strongly convex problems, 
+to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
+or just from $1/T$ to $1/T^2$ 
+in the (non-strongly) convex case.
+\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
+The tools developped to make progress on one problem might help make progress on the other. 
+}
+
 \bibliographystyle{apalike}
@@
 
-\section{Stuff to cite and why }
-
-\begin{itemize}
-	\item Stochastic and online mirror descent: 
-	We obviously have rates in this setting, but they all assume 
-	that the reference function is strongly-convex relative to a norm. 
-	\citep[e.g.][Thm 4.2]{bubeck2015convex}
-	\item What is mirror descent called in the literature?
-		\begin{itemize}
-			\item Mirror Descent \citep{nemirovski1983problem,beck2003mirror} 
-			\item NoLips \citep{bauschke2017descent}
-		\end{itemize}
-	\item Interest in mirror descent boosted by the relatively smooth view 
-	of \citet{bauschke2017descent,lu2018relatively}	
-	\item 
-	The oroblems on how to deal with Bregman divergences abound in optimization, 
-	beyond stochasticity. 
-	For example, we haven't yet figured out the analog of Nesterov-type acceleration 
-	on relatively smooth and strongly convex problems, 
-	to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
-	or just from $1/T$ to $1/T^2$ 
-	in the (non-strongly) convex case.
-	\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
-	The tools developped to make progress on one problem might help make progress on the other. 
-	\item 
-	Stochastic mirror descent in the relatively smooth setting;
-	\citet{hanzely2018fastest},
-	\citet{dragomir2021fast}
-	and the zoo of assumptions
-	\item Something about the statistical asymptotic behavior for the MLE? 
-	\begin{equation}
-		D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
-		\approx \frac{1}{n}?
-	\end{equation}
-	\item Something about how the Bregman divergence is often not considered in the statistics lit 
-	because it can be infinite and is non-symmetric and not nice to work with, 
-	and so folks focus on $\expect{\norm{\theta_n - \theta_*}^2}$ even if it's somewhat meaningless 
-	in the non-asymptotic regime, 
-	and the best non-asymptotic results 
-	still rely on Taylor expansions after $n > $ something big?
-	\item Something about how we don't even know how to deal with decreasing step-sizes 
-	of the order of $1/t$ 
-	\item 
-	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions 
-	\url{https://www.di.ens.fr/~fbach/bach_ejs_self_concordance.pdf}
-	that obtain the $1/t$ rate in the smooth, convex setting instead of $1/\sqrt{t}$, 
-	which is the ``expected'' rate for that setting, 
-	and how this approach above is a generalization beyond the Euclidean noise model 
-	\url{http://proceedings.mlr.press/v99/marteau-ferey19a/marteau-ferey19a.pdf}
-	\url{http://proceedings.mlr.press/v99/ostrovskii19a/ostrovskii19a.pdf}
-\end{itemize}
-
 \end{document}

### 2021-10-06 aistats/paper.tex
Write about kakade paper

--- 2021-10-05 aistats/paper.tex
@@
 
-\clearpage
+\section{Group Meeting}
+\begin{enumerate}
+	\item new plots : mirror map, and gaussian trajectories
+	\item the MLE is biased according to our definition. See plots. Does this definition of bias make sense ? An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean. 
+	\item \citet{kakade2010learning}
+	\begin{enumerate}
+		\item assumption in the spirit of infinite order self-concordance only at the optimum
+		\item After a burn-in phase, the loss behaves quadratically. No precise description of the burn-in length.
+		\item conclusion : it is worth writing down until the end what we are obtaining with self-concordance. \RLP{TODO}
+		\item idea 1 : try to express some of their results with the mean parameter instead.
+		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD. (exploiting the link MAP -> optimization).
+	\end{enumerate}
+\end{enumerate}
+
 \section{Plan}
-
 \begin{enumerate}
@@
 	\item is there multiple conjugate priors ?
-	\item misspecified case. Do we have a   formula ?
+	\item misspecified case. Do we have a   formula ? Partial answer : a sufficient condition is to have $\mu^*$ be accessible to our model, even if the true distribution is not.
 \end{itemize}

### 2021-10-06 aistats.tex
Rename file and add \hypersetup{draft} to avoid fatal error

--- 2021-10-06 aistats/paper.tex
@@
 \documentclass[twoside]{article}
-\usepackage{aistats2022}
+\usepackage{aistats/aistats2022}
 % If your paper is accepted, change the options for the package
@@
 \usepackage{graphicx}
-\graphicspath{{../figs/}{../}}
+\graphicspath{{figs/}}
 %\usepackage{subfigure}
@@
 % my packages
-\usepackage{../math_commands}
+\usepackage{math_commands}
 % some custom math commands
@@
 \newcommand{\MAPt}{\hat \natp_n}
+
+\hypersetup{draft}
 
@@
 which matches the asymptotic rate of unbiased estimation,
-through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler}\footnote{
+through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler} \footnote{
 Those papers assume the stochastic gradients are bounded, 
@@
 
-\fdk{
-Application
-
+ff
+\fdk{Application 
 The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity. 
@@
 to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
-or just from $1/T$ to $1/T^2$ 
-in the (non-strongly) convex case.
-\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
-The tools developped to make progress on one problem might help make progress on the other. 
+or just from $1/T$ to $1/T^2$ in the (non-strongly) convex case.
 }
 
+\citet{dragomir2021optimal}  shows that naïve application of Bregman updates can not achieve acceleration.
+The tools developed to make progress on one problem might help make progress on the other.
+
 \bibliographystyle{apalike}
-\bibliography{../references.bib}
+\bibliography{references.bib}
 

### 2021-10-06 aistats.tex
Revamp the intro - adding contributions

@@
 Estimating the mean of Gaussian noise is a special case of this problem class, using 
-$f(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$ and $g(\theta) \sim \cN(\theta - \mu, \sigma^2)$.
+$f(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$ and $g(\theta) \sim \cN(- \mu, \sigma^2)$.
 SGD with a decreasing step-size coincides with the maximum likelihood estimate (MLE), 
@@
 When $p(x\cond\theta)$ is an exponential family distribution,
-$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (up to constants).
+$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (omitting the reference distribution).
 Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
@@
 
-There are algorithms for this setting, for example the MLE;
-giving the (online) maximum likelihood after seeing $t$ samples
-corresponds to online mirror descent with decreasing step-sizes (more on this later).
-But our convergence rates are weak; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
+There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
+As we will prove, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
+Despite recent advances in this field, our convergence rates are still lacking ; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
 
-We do not have a solution. The goal of this paper, and the reason it is submitted at AISTATS, 
+We do not have a solution. The goal of this paper
 is to bring this issue to the attention of the community working at the intersection of machine learning, optimization and statistics,
@@
 
-We show the connections between the optimization and statistical estimation view of the problem
-and discuss where our current theory fails to fully capture the problem.
-We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate---this is part of the point of this paper---and 
-that the statistical theory results for the convergence of the MLE to the solution are well known. 
-Our point is more subtle, in that for the purposes of optimization, to mimic the results that are known for SGD in Euclidean settings, 
-we would like a result of the form 
-\begin{equation*}
-	\text{optimality gap after $t$ steps}
-	\leq 
-	\frac{\text{``bias''}}{t^2}
-	+ \frac{\text{``variance''}}{t},
-\end{equation*}
-which is the bit that is lacking.
+We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate -- which is optimal in some Cramer-Rao sense -- and for which we have asymptotic and large sample results \TODO{cite}.
+However, none of the theory we found could address for instance the case of a gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions. 
+%kind of redundant problem - question - problem paragraphs, not disentangling stats and optim. To improve
+
+\paragraph{Contributions}
+\begin{itemize}
+	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the gaussian variance estimation problem.
+	\item We show the connections between the optimization and statistical estimation view of the problem and discuss where current theories fail to fully capture the problem.
+\end{itemize}
+
+The statistics and optimization perspective are so tightly linked in this problem, that we believe progress from one field would help the other one. 
+Progress from statistics could bring forward research in non-euclidean optimization, for instance helping in the development of accelerated methods.
+On the other hand, optimization can become  a more versatile and general tool for statisticians. % handwavy and shady af.
 

### 2021-10-06 aistats.tex
Reorganize aistats sections and clean asymptote section

@@
 
-\section{Insight - Asymptote}
+\section{Insights}
+
+\subsection{Asymptote}
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-We can get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
+Asymptotic results typically show that natural parameters $\natp$ asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
+Here the MAP is more simply expressed with mean parameters $\meanp$.
+We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
 \begin{align}
@@
 \end{align}
-where the norm  $\| x \|_\mM^2 = x^\top \mM x$  is induced by the hessian at the optimum 
+where the Mahalanobis norm  $\| x \|_\mM^2 = x^\top \mM x$  is induced by the hessian of the entropy -- the inverse \textit{Fisher information matrix} -- at the optimum, 
 \begin{align}
-    \nabla^2\conj(\mu^*) 
+    \mF^*
+    :=\nabla^2\conj(\mu^*) 
     = \nabla^2\logpart(\natp^*)^{-1} 
@@
 \end{align}
-Both the MLE and the MAP have the same asymptotic rate
+From there, it is possible to show an asymptotic rate for the MLE and the MAP
 \begin{align}
 \label{eq:MAP_asymptote}
-	\E \bregmanconj \left (\E [T(X)] ;  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} \right ) 
-	= \frac{d}{2n} + O(n^{- \frac{3}{2}})
+	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right ) 
+	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
 \end{align}
-as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data.
-Remark how this rate is also independent of the optimum $\meanp^*$.
-\fdk{meaning "independent ignoring constants"?}
+Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
+This asymptote is also independent of the optimum $\meanp^*$.
 
-\fdk{
-That should be well known, maybe find a reference?
-% Something about the statistical asymptotic behavior for the MLE? 
-% \begin{equation}
-% 	D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
-% 	\approx \frac{1}{n}?
-% \end{equation}
+\subsection{Euclide's World}
+As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as the mean of a gaussian.
+\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    &\leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu} \\
+    &\leq 2L \norm{\mu^* - \mu}
+\end{align}
+so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{t}})$ convergence rate from \RLP{Gaussian rate}.
+\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
+\begin{align}
+    \bregmanconj(\mu^* ; \mu) 
+    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
+\end{align}
+so $\bregmanconj$ is upper bounded by a quadratic, and we get the same rate as a Gaussian.
+
+\subsection{Bias-Variance Decomposition}
+In both examples (add link), the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
+In fact, it is possible to write such a decomposition for any Bregman divergence
+Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
+\begin{align}
+	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	&= {\bregmanconj(\mu^* ; \tilde \mu_n)}
+	+ {\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
+	\label{eq:primal_pivot}
+\end{align}
+
+Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point.
+An estimator will be unbiased if its primal expectation is non-zero.
+This is not true for the MLE.
+In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\mu^* ; \tilde \mu_n)} \leq \frac{2}{n(n-2)}$.
+
+\paragraph{Illustrations.}
+We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
+In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
+\RLP{Tune figures.}
+
+\begin{figure*}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/biasvariance/new_linear_ratio=0.8_n0=10}.pdf}
+	\caption{
+	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
+	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
+	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
 	}
+	\label{fig:variance_decomposition}
+\end{figure*}
 
-\section{Example - Gaussian Mean}
+\begin{figure*}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/gaussians/new_linear_n0=1}.pdf}
+	\caption{
+	\textbf{Full gaussian} $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. \textbf{Left:} training curves (5,95) percentiles, average and asymptote. 
+	\textbf{Right:} bias-variance decomposition.
+	}
+	\label{fig:gaussian_decomposition}
+\end{figure*}
 
+\section{Examples}
+
+\subsection{Gaussian Mean}
 \paragraph{$\ell^2$-norm Analysis}
@@
 
-\section{Insight - Euclide's World}
-As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as the mean of a gaussian.
-\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    &\leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu} \\
-    &\leq 2L \norm{\mu^* - \mu}
-\end{align}
-so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{t}})$ convergence rate from \RLP{Gaussian rate}.
-\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
-\begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
-\end{align}
-so $\bregmanconj$ is upper bounded by a quadratic, and we get the same rate as a Gaussian.
-
-\section{Example - Gaussian Variance}
+\subsection{Gaussian Variance}
 A core example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$. 
@@
 
-\section{Bias-Variance Decomposition}
-In both examples (add link), the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
-In fact, it is possible to write such a decomposition for any Bregman divergence
-Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
-As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
-\begin{align}
-	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	&= {\bregmanconj(\mu^* ; \tilde \mu_n)}
-	+ {\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
-	\label{eq:primal_pivot}
-\end{align}
+\section{"Burn In" Analysis}
+Self-concordant line of work, and \citet{kakade2010learning}.
 
-\paragraph{Illustrations.}
-We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
-In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
-\RLP{Tune figures.}
-
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[width=\textwidth]{{figs/biasvariance/new_linear_ratio=0.8_n0=10}.pdf}
-	\caption{
-	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound. 
-	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
-	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
-	}
-	\label{fig:variance_decomposition}
-\end{figure*}
-
-\begin{figure*}[ht]
-	\centering
-	\includegraphics[width=\textwidth]{{figs/gaussians/new_linear_n0=1}.pdf}
-	\caption{
-	\textbf{Full gaussian} $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. \textbf{Left:} training curves (5,95) percentiles, average and asymptote. 
-	\textbf{Right:} bias-variance decomposition.
-	}
-	\label{fig:gaussian_decomposition}
-\end{figure*}
-
-\section{Optimization related work}
+\section{Optimization Perspective}
 In the context of optimization, why do we care?

### 2021-10-06 aistats.tex
Make technical background more precise

@@
 
-We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate -- which is optimal in some Cramer-Rao sense -- and for which we have asymptotic and large sample results \TODO{cite}.
+We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate for which there is a wealth of asymptotic and large sample results \citep{van2000asymptotic,ostrovskii2021finite}.
+%TODO complete
 However, none of the theory we found could address for instance the case of a gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions. 
-%kind of redundant problem - question - problem paragraphs, not disentangling stats and optim. To improve
+%TODO kind of redundant "problem - question - problem" paragraphs, not disentangling stats and optim. To improve
 
@@
 
-\RLP{Goal = disseminate related work to make it look nice.}
-Exponential Families are an elegant and lean way to model a wide variety of data : binary, categorical, natural numbers, positive float, long or short tailed... 
-They are literally the linear model of probabilities.
-The exponential family for data $X$ with natural parameter $\natp$  is entirely specified by the support set of the data   $\cX$ and a sufficient statistic $T: \cX \rightarrow \real^d$
+Exponential Families are the linear model of probabilities.
+Their convexity along with their modeling power make them one of the most widely used statistical models. 
+Here we review some of their properties. 
+We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
+The exponential family for data $X\in \cX$ with natural (or primal) parameter $\natp$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
 \begin{equation}
@@
 \end{equation}
-where $\logpart$ is the log-partition function -- i.e. the normalization factor
+Here $\logpart$ is the log-partition function -- i.e. the normalization factor
 \begin{align}
-    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx 
+    \logpart(\natp) = \log \int e^{\natp^\top T(x)} \nu(dx) \; .
 \end{align}
-where the integral stands for a sum if $x$ has discrete support. \RLP{Use base measure instead.}
-For instance, this simple model encompasses both categorical distributions $\cX = \{1, \dots, k\}$ with $T(X)$ being the one-hot encoding, and multivariate normal distributions $\cX=\real, T(X)=(X, X^2)$.
+When explicit, we will write the random variable $T = T(X)$ for lightness.
+This model is defined on the open convex set $\natp\in \Theta$ such that $\logpart(\natp) < \infty$. 
+This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}, \nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
 
@@
 \begin{align}
-    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
-    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
+    \nabla\logpart(\natp) &=  \expect[p(T|\natp)]{T} =: \meanp \\
+    \nabla^2 \logpart(\natp) &= \Cov_\natp[T] \geq 0
 \end{align}
-where $\meanp$ is called the mean parameter.
-If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
-We will write indifferently $\mu$ or  $\natp$ depending on the context, being aware that both represent the same distribution.
-The second identity entails that $\logpart$ is strictly-convex. 
-At this point it is useful to introduce the \href{https://en.wikipedia.org/wiki/Convex_conjugate}{convex conjugate} (aka Fenchel-Legendre transform) of the logpartition function
+where $\meanp$ is called the mean (or dual) parameter.
+$\meanp$ lives in the open convex set $\cM$ within the convex hull of $T(\cX)$.
+%TODO why is it convex ? 
+%It is open because it is the reciprocal image of an open set by the continuous mirror map
+From now on, we assume that the sufficient statistic $T$ is minimal 
+-- e.g. its covariance is definite positive for all $\natp$ -- 
+then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
+We will write interchangeably $\mu$ or  $\natp$ depending on the context, being aware that both represent the same distribution.
+At this point it is useful to introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
 \begin{align}
@@
 which matches the common notion of \textit{entropy} in information theory.
-If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
-For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
+The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
+
 \begin{figure}[ht]
@@
 	\includegraphics{duality}
-	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$.
-Figure reproduced from \citet{kunstner2020homeomorphic}.
-		\RLP{Show the true gaussian bijection instead?}
+	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\natp, \meanp$.
+Figure copied from \citet{kunstner2020homeomorphic}.
+% TODO make an original one.
 	}
@@
 which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
-Then the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
+Then the joint log-likelihood of $\cD =(X_1,\dots,X_n)$ and $\natp$ is
 \begin{multline}
@@
     = (n_0+n) \logpart (\natp) \\
-    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst
+    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T_i \right) + \cst
     \label{eq:joint_likelihood}
 \end{multline}
-meaning the posterior given $\cD=(X_1,\dots, X_n)$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.
+where $T_i=T(X_i)$.
+This means the posterior given $\cD$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T_i , -(n_0 + n))$.
 
@@
 \begin{align}
-    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
+    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0) \; .
 \end{align}
-verifying the first oder optimality conditions \RLP{$\natp$ unconstrained, living in an open set with barriers}
+If they can be satisfied, the MAP verifies the first oder optimality conditions
 \begin{align}
     \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
-    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
+    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
 \end{align}
@@
 \begin{align}
-	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T(X_i)}{n}
+	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
 \end{align}
@@
 \subsection{Euclide's World}
-As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as the mean of a gaussian.
+As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
 \paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
@@
 \end{align}
-so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{t}})$ convergence rate from \RLP{Gaussian rate}.
+so $\bregmanconj$ is $2L$-Lipschitz.
 \paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then

### 2021-10-06 aistats.tex
Clean up open problem

@@
 which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
-Then the joint log-likelihood of $\cD =(X_1,\dots,X_n)$ and $\natp$ is
+Then the joint log-likelihood of $\mD =(X_1,\dots,X_n)$ and $\natp$ is
 \begin{multline}
-    -\log p(\cD|\natp)p(\natp) 
+    -\log p(\mD|\natp)p(\natp) 
     = (n_0+n) \logpart (\natp) \\
@@
 where $T_i=T(X_i)$.
-This means the posterior given $\cD$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T_i , -(n_0 + n))$.
+This means the posterior given $\mD$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T_i , -(n_0 + n))$.
 
@@
 \begin{align}
-    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0) \; .
+    \hat \natp = \argmin_\natp -\log p(\mD|\natp) + n_0 \bregman(\natp ; \natp_0) \; .
 \end{align}
@@
 These rules are known as moment matching.
-The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
+The MLE and MAP estimates are statistics of the dataset $\mD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
 
@@
 
-\fdk{
-Might want to note that a perhaps ``typical'' view of parameter estimation 
-works with norms, e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ (for some norm and some power), 
-or more generally using distances rather than divergences as they are symmetric and have nicer properties. 
-But as the MLE is the minimizer of the log-likelihood, 
-using the log-likelihood (and by extension the KL divergence) as a measure of risk is natural, if inconvenient.
-}
-
-For a well-specified model, the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
+Let $\natp^* := \argmin \expect[X\sim\cD]{-\log p(X|\natp)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
+Then the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
 \begin{multline}
-    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
+    \expect[X\sim \cD]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
 	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
 \end{multline}
+Note that $\cD$ does not have to be part of the exponential family.
 For the exponential family, the KL is equal to the Bregman divergence induced by the log-partition function, or alternatively by the entropy, being careful with the order of the arguments 
@@
 \end{align}
-Our question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities
+
+Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \TODO{cite} 
+Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \TODO{cite} 
+A contrario the KL is not smooth, and it can be infinite in misspecified settings.
+But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.
+
+Our question is: how does the KL behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities
 \begin{align}
 	\label{eq:bregmanMLE}
-	\expect[\cD]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \\
+	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
 	\label{eq:bregmanMAP}
-	\expect[\cD]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} 
+	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \mu_0 + \smallsum_i T_i}{n_0+n} \right )} 
 \end{align}
-where the outer expectation is on the dataset $\cD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
+where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
 More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.

### 2021-10-06 aistats.tex

@@
 The statistics and optimization perspective are so tightly linked in this problem, that we believe progress from one field would help the other one. 
-Progress from statistics could bring forward research in non-euclidean optimization, for instance helping in the development of accelerated methods.
+Progress from statistics could bring forward research in non-euclidean optimization, or the optimization of barrier losses, for instance helping in the development of accelerated methods.
 On the other hand, optimization can become  a more versatile and general tool for statisticians. % handwavy and shady af.
@@
 
-Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point.
-An estimator will be unbiased if its primal expectation is non-zero.
-This is not true for the MLE.
-In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\mu^* ; \tilde \mu_n)} \leq \frac{2}{n(n-2)}$.
+Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. 
+This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\mu^* ; \tilde \mu_n)} \leq \frac{2}{n(n-2)}$.
+A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.
 
@@
 We are basically losing a factor 2 compared to $1/2n$.
-\RLP{Add numerical experiments with upper bound for gaussian variance MAP.}
 
@@
 
-\section{"Burn In" Analysis}
-Self-concordant line of work, and \citet{kakade2010learning}.
+\section{"Burn In" Statistical Analysis}
+Self-concordant line of work : 
+\citet{kakade2010learning},
+\cite{anastasiou2017bounds},
+\citet{ostrovskii2021finite}.
 
@@
 
-There is a long line work trying the convergence rate of stochastic gradient descent.
-Beyond the results of RM51, 
-we now have proofs n smooth, strongly convex problems and bounded variance 
+There is a long line work bounding the convergence rate of stochastic gradient descent.
+Beyond the results of \citet{robbins1951stochastic}, 
+we now have proofs on smooth, strongly convex problems and bounded variance 
 (as well as other assumptions on the noise, reviewed later)
@@
 This is the case for maximum likelihood estimation and matches the asymptotic rates, 
-but holds more generally, for example for linear and logistic regressions \citep{bach2013non}.
+but holds more generally, for example for linear and logistic regressions \citep{bach2013nonstronglyconvex,moulines2011non}.
 
@@
 neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
-\citep{bauschke2017descent,lu2018relatively},
-where instead of measuring strong-convexity and smoothness relative to the Euclidean norm (or any norm), 
+\citep{birnbaum2011distributed,bauschke2017descent,lu2018relatively}.
+Instead of measuring strong-convexity and smoothness relative to the Euclidean norm (or any norm), 
 we use a reference function instead \fdk{maybe add a line}
@@
 	and looks like a \citet{hanzely2018fastest}
-	\item 
-	Attempt to adapt the ``bounded gradient at the minimum''
+	\item Attempt to adapt the ``bounded gradient at the minimum''
 	\citep{dragomir2021fast}
+	\item Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}.
 \end{itemize}
 
-Other to cite;
-\begin{itemize}
-	\item \citet{marteauferey2019}
-\end{itemize}
+Other work 
+\citet{marteauferey2019beyond}.
 
@@
 
-ff
 \fdk{Application 
@@
 or just from $1/T$ to $1/T^2$ in the (non-strongly) convex case.
-}
-
 \citet{dragomir2021optimal}  shows that naïve application of Bregman updates can not achieve acceleration.
 The tools developed to make progress on one problem might help make progress on the other. 
+}
 
@@
 		\item After a burn-in phase, the loss behaves quadratically. No precise description of the burn-in length.
-		\item conclusion : it is worth writing down until the end what we are obtaining with self-concordance. \RLP{TODO}
+		\item \TODO{it is worth writing down until the end what we are obtaining with self-concordance.}
 		\item idea 1 : try to express some of their results with the mean parameter instead.
@@
 	\end{enumerate}
+	\item positioning between optimization and statistics.
 \end{enumerate}
 
-\section{Plan}
-\begin{enumerate}
-	\item intro to exponential family, and density estimation
-	\item the thing we want to bound
-	\item Examples : gaussian mean and gaussian variance (+other examples, just mentioned)
-	\item Insight : Strongly convex case. (+ self-concordance that is not verified either)
-	\item insight : bias-variance decomposition
-	\item Optimization perspective : SBPP or SBG. But no analysis hold, revealing a flaw of all these techniques.
-	\item Discussion : we believe finding a convergence rate would bring new tools useful to deal with common objects such as barrier losses.
-\end{enumerate}
-
-Open questions
-\begin{itemize}
-	\item does a base measure change anything ?
-	\item is there multiple conjugate priors ?
-	\item misspecified case. Do we have a   formula ? Partial answer : a sufficient condition is to have $\mu^*$ be accessible to our model, even if the true distribution is not.
-\end{itemize}
-
 \end{document}

### 2021-10-06 main.tex
Update figures

@@
 
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{{figs/gaussians/new_linear_n0=0}.pdf}
+	\includegraphics[width=\textwidth]{{figs/2d/gaussian_convergence_n0=0}.pdf}
+	\caption{
+	The MLE is biased in primal space.
+	}
+	\label{fig:MLE_bias}
+\end{figure}
+
 \subsection{About Variance and Covariance}
@@
 
-In our case, the bias variance decomposition tells us that we are interested in the covariance between primal and dual parameters, rather than the covariance of only of these -- $\Cov(\MAPt;\MAPm)$ rather that $\Var(\MAPm)$ or $\Var(\MAPt)$.
+In our case, the bias variance decomposition tells us that we are interested in the covariance between primal and dual parameters, rather than any of their individual covariance -- $\Cov(\MAPt;\MAPm)$ rather that $\Var(\MAPm)$ or $\Var(\MAPt)$.
 

### 2021-10-07 aistats.tex
Rephrase intro

@@
 
-\aistatstitle{Looking for Convergence Rates\\ for the MAP of the Exponential Family}
+\aistatstitle{Convergence Rates for the MAP of an Exponential Family, and Stochastic Mirror Descent -- an Open Problem}
 
@@
 \begin{abstract}
-We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for the exponential family. 
-Surprisingly, we found no solution to this problem in the literature -- e.g. after seeing 5 samples, we do not know how many bits away (in expectation) our model is from the true distribution.
-After displaying some properties and special cases of this problem, 
-we show it is an application of several optimization algorithms.
-Yet it falls out of scope for current analysis, thus highlighting areas for progress.
+We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family. 
+Surprisingly, we found no general solution to this problem in the literature. 
+In particular, current theories do not hold for a gaussian, or in the very interesting few samples regime.
+After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD), yet it falls out of scope for modern analysis.
+We believe solving this very fundamental problem may bring progress both to the statistics and optimization communities.
 \end{abstract}
+
+%In particular, no rates hold in the few samples regime  -- e.g. after seeing 5 samples from a gaussian, we do not know how many bits away from the true distribution we should expect our model to be.
+%No rates hold either when the loss is ill-behaved : infinite domain, not smooth, not self-concordant
 
@@
 As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
-\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
+\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$, which kinda adresses \citet{bubeck2015entropic}), then
 \begin{align}
@@
 	\item positioning between optimization and statistics.
+Start with stat.
+Convince the reader that this is an essential problem, with asymptotic rate.
+but finite is elusive.
+Inspired by recent work (lepriol , bach2013) we want to use optimization to give a rate, we try.
+but we reveal a hole in current analysis.
+Finite rate would allow to see the importance of the prior, and also help us understand behavior for small numbers.
 \end{enumerate}

### 2021-10-07 aistats.tex
Add technicity to background

@@
 After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD), yet it falls out of scope for modern analysis.
-We believe solving this very fundamental problem may bring progress both to the statistics and optimization communities.
+We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
 \end{abstract}
@@
 
-Exponential Families are the linear model of probabilities.
-Their convexity along with their modeling power make them one of the most widely used statistical models. 
+Exponential Families are to statistics what linear models are to regression.
+Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models. 
 Here we review some of their properties. 
@@
 \end{equation}
-Here $\logpart$ is the log-partition function -- i.e. the normalization factor
+Here $\logpart$ is the log-partition function -- i.e. the normalization term
 \begin{align}
@@
 When explicit, we will write the random variable $T = T(X)$ for lightness.
-This model is defined on the open convex set $\natp\in \Theta$ such that $\logpart(\natp) < \infty$. 
+For convenience, we focus on steep exponential families with $\logpart$ a convex function of Legendre type.
+In particular, the set $\natp \in \Theta$ such that $\logpart(\natp) < \infty$ is open and convex.
 This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}, \nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
+
+% TODO sort out relationships between : steep family, minimal statistics and Legendre type. What implies what ?
 
@@
 where $\meanp$ is called the mean (or dual) parameter.
-$\meanp$ lives in the open convex set $\cM$ within the convex hull of $T(\cX)$.
-%TODO why is it convex ? 
-%It is open because it is the reciprocal image of an open set by the continuous mirror map
 From now on, we assume that the sufficient statistic $T$ is minimal 
@@
 then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
+$\meanp$ lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
 We will write interchangeably $\mu$ or  $\natp$ depending on the context, being aware that both represent the same distribution.
@@
 
-Our question is: how does the KL behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities
+Our question is: how does the KL behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities ?
 \begin{align}
@@
 \end{align}
-where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
+where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
 More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.
@@
 
+\clearpage
 \section{Group Meeting}
 \begin{enumerate}
-	\item new plots : mirror map, and gaussian trajectories
-	\item the MLE is biased according to our definition. See plots. Does this definition of bias make sense ? An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean. 
-	\item \citet{kakade2010learning}
+	\item new plots : mirror map, and gaussian trajectories. Add trajectories to paper, with background level set, and new legend : primal/dual expectation
+	\item the MLE is biased according to our definition.
+See plots.
+Does this definition of bias make sense ?
+An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean.
+Remark : the MLE is biased for standard parameters.
+	\item exciting pap er\citet{kakade2010learning}
 	\begin{enumerate}
@@
 		\item idea 1 : try to express some of their results with the mean parameter instead.
-		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD. (exploiting the link MAP -> optimization).
+		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD.
+(exploiting the link MAP -> optimization).
+How do we prove the linear convergence of Newton to the quadratic regime ?
+That is the problem at stake here.
 	\end{enumerate}
-	\item positioning between optimization and statistics.
-Start with stat.
+	\item Intro : positioning between optimization and statistics.
+Start with stats.
 Convince the reader that this is an essential problem, with asymptotic rate.
 but finite is elusive.
-Inspired by recent work (lepriol , bach2013) we want to use optimization to give a rate, we try.
-but we reveal a hole in current analysis.
-Finite rate would allow to see the importance of the prior, and also help us understand behavior for small numbers.
+Inspired by recent work (lepriol2021 , bach2013) we want to use optimization to give a rate.
+We fail, but we reveal a hole in current analysis.
+Finite rate would allow us to quantify the importance of the prior, and also help us understand behavior for small numbers.
+	\item Intro : 1 clear question and sharper contributions, with references to sections, as a memo for the reviewer writing his review.
+"investigate bias-variance..." "we show all these analysis suffer from this problem."
+	\item Organization : add optim parallel in insigths section, and then sprinkle references until the optim section.
+	\item Optim : find members of the exponential family where the rate of Dragomir et al. holds.
+	\item Technical Background : be precise and add reference for the steep exponential family of Legendre type which guarantees that our sets are open, and have a solution in the middle.
 \end{enumerate}

### 2021-10-08 aistats.tex
Fix hyperref warning
add some descriptions in optim section

@@
 \let\oldsection\section
-\renewcommand{\section}[1]{\oldsection{\uppercase{#1}}}
+\renewcommand{\section}[1]{\oldsection{\texorpdfstring{\uppercase{#1}}{#1}}}
 
@@
 
+\usepackage[capitalise]{cleveref}
 \newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
@@
 
-\hypersetup{draft}
+% \hypersetup{draft}
 
@@
 Surprisingly, we found no general solution to this problem in the literature. 
-In particular, current theories do not hold for a gaussian, or in the very interesting few samples regime.
-After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD), yet it falls out of scope for modern analysis.
+In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
+After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD), 
+yet it falls out of scope for modern analysis.
 We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
@@
 There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
-As we will prove, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
-Despite recent advances in this field, our convergence rates are still lacking ; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
+As we will prove,\footnote{\fdk{I don't know any prior reference for this exact statement, but I'm not sure it's safe to claim this as a ``proof'' as if it was a new statement}} the MAP can be interpreted as online mirror descent with decreasing step-sizes.
+Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
 
@@
 %TODO complete
-However, none of the theory we found could address for instance the case of a gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions. 
+However, none of the theory we found could address for instance the case of a Gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions. 
 %TODO kind of redundant "problem - question - problem" paragraphs, not disentangling stats and optim. To improve
@@
 \begin{itemize}
-	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the gaussian variance estimation problem.
+	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the Gaussian variance estimation problem.
 	\item We show the connections between the optimization and statistical estimation view of the problem and discuss where current theories fail to fully capture the problem.
@@
 As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
-\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$, which kinda adresses \citet{bubeck2015entropic}), then
+
+{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$, which kinda adresses \citet{bubeck2015entropic}), then
 \begin{align}
@@
 so $\bregmanconj$ is $2L$-Lipschitz.
-\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
+
+{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex), then
 \begin{align}
@@
 \begin{align}
-	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
+	&\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
+	\\
 	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
@@
 Self-concordant line of work : 
+\citet{bubeck2015entropic}
 \citet{kakade2010learning},
-\cite{anastasiou2017bounds},
+\citet{anastasiou2017bounds},
+\citet{marteauferey2019beyond},
 \citet{ostrovskii2021finite}.
@@
 \section{Optimization Perspective}
+\fdk{Some sections are obv. too long, and some info irrelevant.}
+
 In the context of optimization, why do we care?
 
-There is a long line work bounding the convergence rate of stochastic gradient descent.
-Beyond the results of \citet{robbins1951stochastic}, 
-we now have proofs on smooth, strongly convex problems and bounded variance 
-(as well as other assumptions on the noise, reviewed later)
-a decreasing step-size and an averaging scheme gets 
-a $1/t$ convergence rate, 
-which matches the asymptotic rate of unbiased estimation,
-through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler} \footnote{
-Those papers assume the stochastic gradients are bounded, 
-but small modifications also work for bounded variance instead.
+\subsection{Stochastic mirror descent}
+
+\fdk{
+- Would mirror descent and the equivalence with MLE/MAP be introduced earlier?
+- Mirror descent \citep{nemirovski1983problem,beck2003mirror}
+- (also called NoLips \citep{bauschke2017descent}, 
+relative gradient descent \citep{hanzely2018fastest},
+Bregman proximal gradient, ...)
 }
 
-We do have extensions to other notions of bounded variance, 
-for example assuming that the stochastic gradients are gradients of a perturbed smooth function $f_i$ 
-and that the minimum of $f$ and the minima of the $f_i$ 
-are bounded, $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$,
-or that the gradient noise is bounded only at the minimum. 
-For a review, see \citet{gower2019sgd}.
+{\bf In online learning, } \fdk{bounded gradient assumption}
 
-However, for some problems, combining decreasing step-sizes and averaging is not necessary,
-even when the function is not strongly convex.
-This is the case for maximum likelihood estimation and matches the asymptotic rates, 
-but holds more generally, for example for linear and logistic regressions \citep{bach2013nonstronglyconvex,moulines2011non}.
-
-There has been recent advances in optimization that go beyond the Euclidean case, 
+{\bf In smooth optimization, } there has been recent advances in optimization that go beyond the Euclidean case, 
 to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
 neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
-\citep{birnbaum2011distributed,bauschke2017descent,lu2018relatively}.
-Instead of measuring strong-convexity and smoothness relative to the Euclidean norm (or any norm), 
-we use a reference function instead \fdk{maybe add a line}
-and apply mirror descent \citep{nemirovski1983problem,beck2003mirror}
-(also called NoLips \citep{bauschke2017descent}, something else?).
+\citep{bauschke2017descent,lu2018relatively}.
+Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm), 
+the $\mu$-strong convexity and $L$-smoothness of a function $f$ 
+relative to a reference function $A$ is defined 
+as 
+\aligns{
+	\mu \cB_{A}(x, y)
+	\leq 
+	\cB_f(x,y)
+	\leq 
+	L \cB_A(x,y).
+}
+Those conditions ensure the linear convergence rate of mirror descent with $A$ as the reference function.
+Maximum likelihood in exponential families
+that fits the theory developped for mirror descent well. 
+Indeed, 
+\aligns{
+	f(\theta) = A(\theta) - \expect[x\sim p(x\cond\theta)]{\lin{x, \theta}}
+}
+is $1$-smooth and $1$-strongly convex relative to $A$,
+and the the optimality gap is the divergence
+\aligns{
+	f(\theta) - f(\theta_*) = 
+	\cB_A(\theta, \theta_*)
+	= \KL(p(\cdot\cond\theta_*);\, p(\cdot\cond\theta)),
+}
+where $\theta_* = \arg\min_\theta f(\theta)$.
+But the known results for stochastic mirror descent do not give convergence rates 
+for all exponential families.
 
-We have convergence results for the 
-application of stochastic mirror descent, 
-but they typically rely on assumptions that are too strong. 
-\begin{itemize}
-	\item The reference function is strongly convex [typical setup in online learning]
-	\item The variance is uniformly bounded
-\end{itemize}
-Some advances have been tried
-\begin{itemize}
-	\item The reference function is ``close to'' a norm
-	and looks like a \citet{hanzely2018fastest}
-	\item Attempt to adapt the ``bounded gradient at the minimum''
-	\citep{dragomir2021fast}
-	\item Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}.
-\end{itemize}
+{\bf Traditional results}
+in stochastic optimization,
+such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
+rely on the assumption that the reference function $A$ 
+is itself strongly convex relative to a norm $\norm{\cdot}$,
+(meaning that $f$ is strongly convex). 
+This apply for the MLE problem of a Gaussian distribution 
+with fixed variance, 
+but not when the covariance is unknown,
+or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
+are not strongly convex.
 
-Other work 
-\citet{marteauferey2019beyond}.
+The work of \citet{bauschke2017descent,lu2018relatively}
+has renewed interest for analyses of mirror descent in the smooth setting, 
+and developping new tools to analyse problems
+without relying on strong-convexity in an underlying norm.
+However, they do not yet address the problem satisfactorily.
 
-\fdk{
-A ``simple open problem'';
-assume we have a deterministic problem, so that we know where the minimum is. 
-Can we figure out a (deterministic) path from $\theta_0$ to $\theta_*$ 
-that has optimality decreasing as $1/n^2$, 
-to mirror the bias term of gradient descent?
+\textbf{An adaptation of the bounded variance,}
+investigated by \citet{hanzely2018fastest}, 
+defines the following bounded quantity.
+Letting $\theta^+(v)$ be a step from $\theta$ with vector $v$, 
+$\theta^+(v) = \nabla A^*(\nabla A(\theta) - v)$,
+\alignn{\label{eq:assumption-hanzely}
+	\expect[\tilde g]{
+	\lin{
+		\nabla f(\theta) - \tilde g, 
+		\theta^+\!\!\paren{\eta \tilde g } - \theta^+\!\!\paren{\eta \nabla f(\theta)}
+	}
+	}
+	\leq 
+	\eta C,
+}
+for some step-size $\eta$ and constant $C$. Let us unpack this quantity;
+
+{\bf For a Gaussian} with known variance, 
+with $A(\theta) = \frac{1}{2}\norm{\theta}^2$, 
+this definition recovers the variance of the stochastic gradient
+$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta_t) - \tilde g_t}{}^2}\leq C$.
+More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
+the left-hand side of \cref{eq:assumption-hanzely} is bounded by 
+\aligns{
+	\frac{1}{\mu} \expect{\norm{\nabla f(\theta_t) - \tilde g_t}{}^2_*},
+}
+and a uniform bound $C$ exist if the variance of the stochastic gradient is bounded.
+
+{\bf In divergence,} this condition states that the (symmetrized) divergence between 
+a deterministic and stochastic update is bounded in expectation,
+\aligns{
+	&
+	\mathbb{E}_{\tilde g_t}\bigg[
+	\cB_{A^*}\paren{\mu - \eta g, \mu - \eta \nabla f(\theta)}
+	\\
+	&\hphantom{\mathbb{E}_{\tilde g}[}
+	+ \cB_{A^*}\paren{\mu - \eta \nabla f(\theta), \mu - \eta g}
+	\bigg]
+	\leq 
+	\eta C.
 }
 
-\fdk{Application 
-The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity. 
-For example, we haven't yet figured out the analog of Nesterov-type acceleration 
-on relatively smooth and strongly convex problems, 
-to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
-or just from $1/T$ to $1/T^2$ in the (non-strongly) convex case.
-\citet{dragomir2021optimal}  shows that naïve application of Bregman updates can not achieve acceleration.
-The tools developed to make progress on one problem might help make progress on the other. 
+{\bf For a maximum likelihood problem, }
+where $\tilde g = \mu - x$ 
+where $x \sim p(x\cond\theta_*)$,
+and $\nabla f(\theta) = \mu - \mu_*$,
+\aligns{
+	&
+	\mathbb{E}_{x}\bigg[
+	\cB_{A^*}\paren{(1-\eta) \mu + \eta x, (1-\eta)\mu + \eta \mu_*}
+	\\
+	&\hphantom{\mathbb{E}_{\tilde g_t}[}
+	\cB_{A^*}\paren{(1-\eta)\mu + \eta \mu_*, (1-\eta) \mu + \eta x}
+	\bigg]
+	\leq 
+	\eta C.
 }
+
+\fdk{Example where this fails?}
+
+\textbf{Bounded variance at the minimum, }
+\fdk{\citep{dragomir2021fast}}
+
+\textbf{Bounded optimality gap,}
+\fdk{
+Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}}
 
@@
 
+\section{Blurbs/ideas?}
+
+\subsection{SGD blurb}
+\fdk{
+%
+There is a long line work bounding the convergence rate of stochastic gradient descent.
+Beyond the results of \citet{robbins1951stochastic}, 
+we now have proofs on smooth, strongly convex problems and bounded variance 
+(as well as other assumptions on the noise, reviewed later)
+a decreasing step-size and an averaging scheme gets 
+a $1/t$ convergence rate, 
+which matches the asymptotic rate of unbiased estimation,
+through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler} 
+\\
+Those papers assume the stochastic gradients are bounded, 
+but small modifications also work for bounded variance instead.
+\\
+We do have extensions to other notions of bounded variance, 
+for example assuming that the stochastic gradients are gradients of a perturbed smooth function $f_i$ 
+and that the minimum of $f$ and the minima of the $f_i$ 
+are bounded, $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$,
+or that the gradient noise is bounded only at the minimum. 
+For a review, see \citet{gower2019sgd}.
+\\
+However, for some problems, combining decreasing step-sizes and averaging is not necessary,
+even when the function is not strongly convex.
+This is the case for maximum likelihood estimation and matches the asymptotic rates, 
+but holds more generally, for example for linear and logistic regressions \citep{bach2013nonstronglyconvex,moulines2011non}.
+}
+
+\subsection{Poisson likelihood?}
+\fdk{
+\citet{bauschke2017descent} and \citet{hanzely2018fastest} both use the example of Poisson inverse problems/Poisson regression
+as examples. 
+The simpler case of the MLE of a Poisson distribution is also unsolved, though. 
+In this case, $h(x) = 1/x!$, $x \in \mathbb{N}$, $A(\theta) = e^\theta$ over $\theta \in \mathbb{R}$,
+$A^*(\mu) = \mu \log \mu - \mu$ over $\mu \in \mathbb{R}_+$.
+It is neither strongly-convex nor self-concordant (at least according to the standard definition, 
+although it satisfies generalized notions of self-concordance as $A'''(\theta) = A''(\theta)$).
+}
+
+\subsection{``simple'' open problem?}
+\fdk{
+A ``simple open problem'';
+assume we have a deterministic problem, so that we know where the minimum is. 
+Can we figure out a (deterministic) path from $\theta_0$ to $\theta_*$ 
+that has optimality decreasing as $1/n^2$, 
+to mirror the bias term of gradient descent?
+}
+
+\subsection{Connection with acceleration?}
+\fdk{
+The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity. 
+For example, we haven't yet figured out the analog of Nesterov-type acceleration 
+on relatively smooth and strongly convex problems, 
+to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
+or just from $1/T$ to $1/T^2$ in the (non-strongly) convex case.
+\citet{dragomir2021optimal}  shows that naïve application of Bregman updates can not achieve acceleration.
+The tools developed to make progress on one problem might help make progress on the other. 
+}
+
 \end{document}

### 2021-10-08 aistats.tex
Add references for technical background

--- 2021-10-07 aistats.tex
@@
 
+\subsection{MAP as Stochastic Mirror Descent}
+\TODO{s
+}
+
 \subsection{Asymptote}
@@
 
-\subsection{Euclide's World}
-As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
+\subsection{Euclidean Case}
+As another reference point, let us consider the case $\logpart = \half \norm{\natp}_2^2$, which corresponds to finding the mean of a Gaussian with variance $1$ ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
+
+If we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
 \paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$, which kinda adresses \citet{bubeck2015entropic}), then
@@
 	\item Technical Background : be precise and add reference for the steep exponential family of Legendre type which guarantees that our sets are open, and have a solution in the middle.
+References: \begin{itemize}
+		\item \href{https://www.jstor.org/stable/4616462?seq=1#metadata_info_tab_contents}{Existence of Maximum Likelihood Estimates for Multi-Dimensional Exponential Families}, 
+		\item The singly truncated normal distribution: A non-steep exponential family, 
+		\item Statistical exponential families: A digest with flash cards,
+		\item Information and Exponential Families In Statistical Theory, 
+		\item Fundamentals of Statistical Exponential Families with Applications in Statistical Decision Theory.
+	\end{itemize}
+
 \end{enumerate}

### 2021-10-08 aistats.tex

--- 2021-10-08 aistats.tex
@@
 
-\subsection{Euclide's World}
-As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
+\subsection{Quadratic Regime}
+As another reference point, let us consider the case $\logpart = \half \norm{\natp}_2^2$, which corresponds to finding the mean of a Gaussian with variance $1$ ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
+If we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
 
@@
 	\item Technical Background : be precise and add reference for the steep exponential family of Legendre type which guarantees that our sets are open, and have a solution in the middle.
+	References: \begin{itemize}
+		\item \href{https://www.jstor.org/stable/4616462?seq=1#metadata_info_tab_contents}{Existence of Maximum Likelihood Estimates for Multi-Dimensional Exponential Families}, 
+		\item The singly truncated normal distribution: A non-steep exponential family, 
+		\item Statistical exponential families: A digest with flash cards,
+		\item Information and Exponential Families In Statistical Theory, 
+		\item Fundamentals of Statistical Exponential Families with Applications in Statistical Decision Theory.
+	\end{itemize}  
 \end{enumerate}

### 2021-10-08 aistats.tex
Complete quadratic comparison

@@
 \newcommand{\bregmanconj}{\cB_{\logpart^*}}
+\newcommand{\nat}{\theta}
 \newcommand{\natp}{\theta}
-\newcommand{\meanp}{\mu}
+\newcommand{\m}{m}
+\newcommand{\meanp}{\m}
 \newcommand{\decrement}{D}
@@
 
-\newcommand{\MAPm}{\hat \mu_n}
+\newcommand{\MAPm}{\hat \m_n}
 \newcommand{\MAPt}{\hat \natp_n}
@@
 
-\subsection{Quadratic Regime}
-As another reference point, let us consider the case $\logpart = \half \norm{\natp}_2^2$, which corresponds to finding the mean of a Gaussian with variance $1$ ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
-If we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
+\subsection{Quadratic Case}
+As another reference point, let us consider the case $\logpart(\natp) = \half \norm{\natp}_2^2$.
+For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
+Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances
+\begin{align}
+	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
+\end{align}
+Thanks to the independence of samples we can break down the MLE into individual point's contributions
+\begin{align}
+	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2} 
+	=\frac{\Var(T)}{2n} 
+\end{align}
+Adding a reference mean $\mu_0$ to get the MAP yields
+\begin{align}
+	\expect{\bregmanconj(\meanp^*; \MAPm)}
+	&= \frac{n \Var(T) +  n_0^2 \norm{\mu^* -  \mu_0}^2}{(n+n_0)^2}
+	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
+\end{align}
+so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. We would like to get a similar result for arbitrary exponential families.
+If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.
 
-{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$, which kinda adresses \citet{bubeck2015entropic}), then
+{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
 \begin{align}
@@
 \end{align}
-so $\bregmanconj$ is $2L$-Lipschitz.
+so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
 
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get the same rate as a Gaussian.
+
+\subsection{Locally Quadratic Case}
+All Bregman divergences are locally quadratic. 
+Under some assumptions, we can quantify  when this quadratic behavior  kicks in.
+For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}. 
+This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant 
+-- e.g. exponential distribution, 
+Pareto distribution with known minimum value, 
+Weibull distribution with known shape k, 
+Laplace distribution
+with known mean
+-- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
+\begin{align}
+	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
+\end{align}
+holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$. 
+Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\natp^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
+
+\TODO{Reference \citet{ostrovskii2021finite}.}
+
+All these works give \textit{large sample} results -- results that hold only for $n\geq N$ for some constant $N$ -- but none of them  apply to small $n$.
 
@@
 \section{Examples}
-
-\subsection{Gaussian Mean}
-\paragraph{$\ell^2$-norm Analysis}
-In the Euclidean realm, the average is bounded by variance becomes a bound on the variance of the average
-\begin{align}
-	\Var T(X) = \E \norm{T(X) - \mu^*}^2 = \sigma^2 \\
-	\implies \E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
-\end{align}
-eg the variance of the mean is $n$ times smaller than the variance of the samples.
-Adding a reference mean $\mu_0$ to get the MAP yields
-\begin{align}
-	&\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
-	\\
-	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
-	&= O\left(\frac{\sigma^2}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
-\end{align}
-so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !
 

### 2021-10-08 aistats.tex
Replace mu by m for mean parameter

@@
 \newcommand{\nat}{\theta}
-\newcommand{\natp}{\theta}
 \newcommand{\m}{m}
@@
 \newcommand{\MAPm}{\hat \m_n}
-\newcommand{\MAPt}{\hat \natp_n}
+\newcommand{\MAPt}{\hat \nat_n}
 
@@
 We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
-The exponential family for data $X\in \cX$ with natural (or primal) parameter $\natp$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
+The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
 \begin{equation}
-	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
+	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
 \end{equation}
@@
 \begin{align}
-    \logpart(\natp) = \log \int e^{\natp^\top T(x)} \nu(dx) \; .
+    \logpart(\nat) = \log \int e^{\nat^\top T(x)} \nu(dx) \; .
 \end{align}
@@
 For convenience, we focus on steep exponential families with $\logpart$ a convex function of Legendre type.
-In particular, the set $\natp \in \Theta$ such that $\logpart(\natp) < \infty$ is open and convex.
+In particular, the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex.
 This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}, \nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
@@
 \begin{align}
-    \nabla\logpart(\natp) &=  \expect[p(T|\natp)]{T} =: \meanp \\
-    \nabla^2 \logpart(\natp) &= \Cov_\natp[T] \geq 0
+    \nabla\logpart(\nat) &=  \expect[p(T|\nat)]{T} =: \meanp \\
+    \nabla^2 \logpart(\nat) &= \Cov_\nat[T] \geq 0
 \end{align}
@@
 From now on, we assume that the sufficient statistic $T$ is minimal 
--- e.g. its covariance is definite positive for all $\natp$ -- 
-then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
+-- e.g. its covariance is definite positive for all $\nat$ -- 
+then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\nat$ and mean parameters $\m$.
 $\meanp$ lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
-We will write interchangeably $\mu$ or  $\natp$ depending on the context, being aware that both represent the same distribution.
+We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both represent the same distribution.
 At this point it is useful to introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
 \begin{align}
-	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
+	\conj(\m) = \langle \m, \nat \rangle - \logpart(\nat) \; .
 \end{align}
 which matches the common notion of \textit{entropy} in information theory.
-The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
+The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).
 
@@
 	\includegraphics{duality}
-	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\natp, \meanp$.
+	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$.
 Figure copied from \citet{kunstner2020homeomorphic}.
@@
 \paragraph{Bregman Divergences}
-A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\natp$ and $\natp_0$
+A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\nat$ and $\nat_0$
 \begin{align}
-    \bregman (\natp ; \natp_0)
-    & = \logpart(\natp) - \logpart(\natp_0) 
-    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
+    \bregman (\nat ; \nat_0)
+    & = \logpart(\nat) - \logpart(\nat_0) 
+    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle
 \end{align}
-with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
+with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. 
 It is a non-symmetric measure of distance between parameters.
@@
 \begin{align}
-	\bregman (\natp ; \natp_0)
+	\bregman (\nat ; \nat_0)
     = \bregmanconj ( \meanp_0 ; \meanp) \; .
@@
 \paragraph{Conjugate Prior}
-One conjugate prior \citep{agarwal2010geometric} for $p(X|\natp)$ is
+One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
 \begin{align}
-    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
+    p(\nat) &\propto \exp( - n_0 \bregman(\nat ; \nat_0) )
 \end{align}
-where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is introduced above.
-Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
+where $n_0$ and $\nat_0$ are (hyper)parameters of the prior, and $\bregman(\nat ; \nat_0)$ is introduced above.
+Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\nat_0$.
 The negative log-likelihood of the prior is
 \begin{align*}
-    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
+    -\log p(\nat) = n_0 (\logpart(\nat)  - \nat^\top \meanp_0 ) + \cst
 \end{align*}
-which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
-Then the joint log-likelihood of $\mD =(X_1,\dots,X_n)$ and $\natp$ is
+which is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
+Then the joint log-likelihood of $\mD =(X_1,\dots,X_n)$ and $\nat$ is
 \begin{multline}
-    -\log p(\mD|\natp)p(\natp) 
-    = (n_0+n) \logpart (\natp) \\
+    -\log p(\mD|\nat)p(\nat) 
+    = (n_0+n) \logpart (\nat) \\
     - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T_i \right) + \cst
@@
 where $T_i=T(X_i)$.
-This means the posterior given $\mD$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T_i , -(n_0 + n))$.
+This means the posterior given $\mD$ is part of the same family, with natural parameters $(n_0 \m_0 + \sum_i T_i , -(n_0 + n))$.
 
 \paragraph{Maximum A Posteriori (MAP).}
-Minimizing~\eqref{eq:joint_likelihood} over $\natp$ yields the MAP estimate
+Minimizing~\eqref{eq:joint_likelihood} over $\nat$ yields the MAP estimate
 \begin{align}
-    \hat \natp = \argmin_\natp -\log p(\mD|\natp) + n_0 \bregman(\natp ; \natp_0) \; .
+    \hat \nat = \argmin_\nat -\log p(\mD|\nat) + n_0 \bregman(\nat ; \nat_0) \; .
 \end{align}
@@
 \begin{align}
-    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
+    \nabla \logpart(\hat \nat_\text{MAP}) = \hat \meanp_\text{MAP}
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
@@
 \begin{align}
-	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
+	\hat \m_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
 \end{align}
 These rules are known as moment matching.
-The MLE and MAP estimates are statistics of the dataset $\mD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
+The MLE and MAP estimates are statistics of the dataset $\mD$. Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.
 
@@
 
-Let $\natp^* := \argmin \expect[X\sim\cD]{-\log p(X|\natp)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
+Let $\nat^* := \argmin \expect[X\sim\cD]{-\log p(X|\nat)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
 Then the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
 \begin{multline}
-    \expect[X\sim \cD]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
-	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
+    \expect[X\sim \cD]{-\log p(X|\nat) + \log p(X|\nat^*) } \\
+	= \KL( p(.|\nat^*) ; p(.|\nat)) \; .
 \end{multline}
@@
 \boxed{
-	\KL( p(.|\natp^*) ; p(.|\natp))
-    = \bregman (\natp ; \natp^*)
+	\KL( p(.|\nat^*) ; p(.|\nat))
+    = \bregman (\nat ; \nat^*)
     = \bregmanconj ( \meanp^* ; \meanp)
@@
 
-Our question is: how does the KL behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities ?
+Our question is: how does the KL behave when $\nat$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities ?
 \begin{align}
@@
 	\label{eq:bregmanMAP}
-	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \mu_0 + \smallsum_i T_i}{n_0+n} \right )} 
+	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )} 
 \end{align}
@@
 Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable. 
-For instance, if we plug in $\conj(\mu)= -\log(\mu)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
+For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
 
@@
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-Asymptotic results typically show that natural parameters $\natp$ asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
+Asymptotic results typically show that natural parameters $\nat$ asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
 Here the MAP is more simply expressed with mean parameters $\meanp$.
@@
 \begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
-    + O(\norm{\mu - \mu^*}^3)
+    \bregmanconj(\m^* ; \m) 
+    &=\half  \norm{\m^* - \m}^2_{\nabla^2\conj(\m^*)}
+    + O(\norm{\m - \m^*}^3)
 \end{align}
@@
     \mF^*
-    :=\nabla^2\conj(\mu^*) 
-    = \nabla^2\logpart(\natp^*)^{-1} 
-    = \Cov_{\natp^*}[T(X)]^{-1}  \; .
+    :=\nabla^2\conj(\m^*) 
+    = \nabla^2\logpart(\nat^*)^{-1} 
+    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
 \end{align}
@@
 \subsection{Quadratic Case}
-As another reference point, let us consider the case $\logpart(\natp) = \half \norm{\natp}_2^2$.
+As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
 For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
@@
 \end{align}
-Adding a reference mean $\mu_0$ to get the MAP yields
+Adding a reference mean $\m_0$ to get the MAP yields
 \begin{align}
 	\expect{\bregmanconj(\meanp^*; \MAPm)}
-	&= \frac{n \Var(T) +  n_0^2 \norm{\mu^* -  \mu_0}^2}{(n+n_0)^2}
-	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
+	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}
+	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
 \end{align}
@@
 \begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    &\leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu} \\
-    &\leq 2L \norm{\mu^* - \mu}
+    \bregmanconj(\m^* ; \m) 
+    &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
+    &\leq 2L \norm{\m^* - \m}
 \end{align}
@@
 \begin{align}
-    \bregmanconj(\mu^* ; \mu) 
-    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
+    \bregmanconj(\m^* ; \m) 
+    \leq \frac{L}{2} \norm{\m^* - \m}^2
 \end{align}
@@
 holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$. 
-Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\natp^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
+Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
 
@@
 In fact, it is possible to write such a decomposition for any Bregman divergence
-Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
 As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
 \begin{align}
-	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
-	&= {\bregmanconj(\mu^* ; \tilde \mu_n)}
-	+ {\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
+	\expect{\bregmanconj(\m^* ; \hat \m_n)} 
+	&= {\bregmanconj(\m^* ; \tilde \m_n)}
+	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
 	\label{eq:primal_pivot}
@@
 Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. 
-This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\mu^* ; \tilde \mu_n)} \leq \frac{2}{n(n-2)}$.
+This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
 A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.
@@
 We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
-In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
+In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
 \RLP{Tune figures.}
@@
 The density is	$p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}$.
-Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
-Mean and natural parameters are roughly inverse of each other $\natp = -\inv{2 \mu}$.
+Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$. 
+Mean and natural parameters are roughly inverse of each other $\nat = -\inv{2 \m}$.
 Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
 \begin{align}
-	\logpart (\natp) &= - \half \log(-\natp)  + \half \log(\pi) \\
-	\conj(\mu) &= \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
+	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi) \\
+	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
 \end{align}
 Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
-It means the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative Gamma distribution.
+It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, eg a negative Gamma distribution.
 It also means $\bregman$ and $\bregmanconj$ have the same shape
 \begin{align}
-	\bregmanconj( \mu_*; \mu_n) 
-	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
-	\bregman( \natp_n; \natp_* )
-	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
+	\bregmanconj( \m_*; \m_n) 
+	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) \\
+	\bregman( \nat_n; \nat_* )
+	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
 \end{align}
-In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
+In other words, this divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function $\phi$
 \begin{align}
@@
 \begin{theorem}[MLE Tight Bound]
-	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
+	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
 	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
 	\begin{align}
-		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
+		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
 			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
@@
  \begin{align}
-	b_n = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
+	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
  \end{align}
-The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
+The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
  \begin{equation}
-	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
+	\expect{\bregmanconj( \m_*; \hat \m_n^\text{MAP})}
 	\leq \begin{cases}
 		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
-		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b_n \ \text{if}\ n\geq 2
+		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
 	\end{cases}
@@
 This inequality highlights a clear variance-bias decomposition.
-In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
-For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
+In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
+For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
 Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$.
@@
 In this case, $h(x) = 1/x!$, $x \in \mathbb{N}$, $A(\theta) = e^\theta$ over $\theta \in \mathbb{R}$,
-$A^*(\mu) = \mu \log \mu - \mu$ over $\mu \in \mathbb{R}_+$.
+$A^*(\m) = \m \log \m - \m$ over $\m \in \mathbb{R}_+$.
 It is neither strongly-convex nor self-concordant (at least according to the standard definition, 

### 2021-10-08 aistats.tex
Cast MAP as SMD

@@
 \newcommand{\MAPt}{\hat \nat_n}
+\DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}
 
@@
 There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
-As we will prove,\footnote{\fdk{I don't know any prior reference for this exact statement, but I'm not sure it's safe to claim this as a ``proof'' as if it was a new statement}} the MAP can be interpreted as online mirror descent with decreasing step-sizes.
+As we show, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
 Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
@@
 \begin{align}
-    p(\nat) &\propto \exp( - n_0 \bregman(\nat ; \nat_0) )
+    p(\nat) 
+    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
+    &\propto \exp(n_0 \m_0^\top \nat - n_0 \logpart(\nat))
 \end{align}
-where $n_0$ and $\nat_0$ are (hyper)parameters of the prior, and $\bregman(\nat ; \nat_0)$ is introduced above.
-Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\nat_0$.
-The negative log-likelihood of the prior is
-\begin{align*}
-    -\log p(\nat) = n_0 (\logpart(\nat)  - \nat^\top \meanp_0 ) + \cst
-\end{align*}
-which is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
-Then the joint log-likelihood of $\mD =(X_1,\dots,X_n)$ and $\nat$ is
-\begin{multline}
-    -\log p(\mD|\nat)p(\nat) 
-    = (n_0+n) \logpart (\nat) \\
-    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T_i \right) + \cst
-    \label{eq:joint_likelihood}
-\end{multline}
-where $T_i=T(X_i)$.
-This means the posterior given $\mD$ is part of the same family, with natural parameters $(n_0 \m_0 + \sum_i T_i , -(n_0 + n))$.
+where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
+This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
+Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.
 
 \paragraph{Maximum A Posteriori (MAP).}
-Minimizing~\eqref{eq:joint_likelihood} over $\nat$ yields the MAP estimate
+Using Bayes rule, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
 \begin{align}
-    \hat \nat = \argmin_\nat -\log p(\mD|\nat) + n_0 \bregman(\nat ; \nat_0) \; .
+	p(\nat | \mD_n)
+	%\propto p(\mD|\nat)p(\nat) 
+    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt))
+    \label{eq:joint_likelihood}
 \end{align}
-If they can be satisfied, the MAP verifies the first oder optimality conditions
+reaching its maximum in $\MAPt$, the MAP estimate verifying the first order optimality conditions
 \begin{align}
-    \nabla \logpart(\hat \nat_\text{MAP}) = \hat \meanp_\text{MAP}
+    \nabla \logpart(\MAPt) = \MAPm
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
 \end{align}
+where $T_i=T(X_i)$.
 When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
 \begin{align}
-	\hat \m_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
+	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
 \end{align}
-These rules are known as moment matching.
-The MLE and MAP estimates are statistics of the dataset $\mD$. Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.
+This last equation is known as moment matching.
+The MLE and MAP estimates are statistics of the dataset $\mD_n$. 
+Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.
 
@@
 This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
-A key difference though, is that the random variable $T(X)$ is connected to the metric $\logpart$. 
+However there is a connection between the random variable $T(X)$ and the metric $\logpart$. 
 Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable. 
@@
 \subsection{MAP as Stochastic Mirror Descent}
-\TODO{s
-}
+The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
+\begin{align*}
+	\hat \nat_{n}
+	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
+    %\label{eq:SBPP}
+\end{align*}
+which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
+\begin{align}
+		\hat \nat_{n}
+	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
+    \label{eq:SMD}
+\end{align}
+where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$. 
+We recognize the update of Stochastic Mirror Descent -- a.k.a. Stochastic Bregman Gradient -- with step-size $\lr_n = \inv{n_0 + n}$.
 
@@
 -- e.g. exponential distribution, 
+Gaussian and Laplace distributions with known mean,
 Pareto distribution with known minimum value, 
-Weibull distribution with known shape k, 
-Laplace distribution
-with known mean
+Weibull distribution with known shape k
 -- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
@@
 
-\TODO{Reference \citet{ostrovskii2021finite}.}
+\TODO{Reference \citet{ostrovskii2021finite}.} ( as well as 
+\citet{anastasiou2017bounds},
+\citet{marteauferey2019beyond})
 
@@
 \end{figure*}
-
-\section{"Burn In" Statistical Analysis}
-Self-concordant line of work : 
-\citet{bubeck2015entropic}
-\citet{kakade2010learning},
-\citet{anastasiou2017bounds},
-\citet{marteauferey2019beyond},
-\citet{ostrovskii2021finite}.
 

### 2021-10-08 aistats.tex
Pass over beginning of optim

@@
 
-% \hypersetup{draft}
+\hypersetup{draft}
 
@@
 \begin{equation}
-	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta_*)},
+	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta)},
 \end{equation}
@@
 \begin{align}
-	p(\nat | \mD_n)
+	p(\nat \cond \mD_n)
 	%\propto p(\mD|\nat)p(\nat) 
@@
 \subsection{MAP as Stochastic Mirror Descent}
+\label{ssec:MAP=SMD}
 The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
@@
 where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$. 
-We recognize the update of Stochastic Mirror Descent -- a.k.a. Stochastic Bregman Gradient -- with step-size $\lr_n = \inv{n_0 + n}$.
+We recognize the update of Stochastic Mirror Descent (a.k.a.
+Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
+This parallel means it would be possible to obtain a finite convergence rate for the  MAP from an analysis of these algorithms. 
+We give a special focus on this approach in Section~\ref{sec:optimization}.
 
@@
 	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
+	\label{eq:MAP_quadratic}
 \end{align}
@@
 \end{align}
-so $\bregmanconj$ is upper bounded by a quadratic, and we get the same rate as a Gaussian.
+so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
 
@@
 
-Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. 
+\textbf{Remark:} In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point.
+An estimator will be unbiased if its primal expectation is non-zero.
 This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
@@
 \section{Optimization Perspective}
-\fdk{Some sections are obv. too long, and some info irrelevant.}
+\label{sec:optimization}
+As we saw in Section~\ref{ssec:MAP=SMD}, MAP can be interpreted as Stochastic Mirror Descent (SMD). 
+There are two consequences of this observation: 1.
+we may obtain a convergence rate for MAP from an optimization analysis, and 2.
+any insights gained from the MAP may inform further design and analysis of SMD.
 
-In the context of optimization, why do we care?
+\subsection{Relative Smoothness}
+Mirror Descent (MD) \footnote{Mirror Descent is going by many names : Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.} \citep{nemirovski1983problem, beck2003mirror} was first introduced to optimize non-smooth functions, typically under bounded gradient assumption on the objective $f$ or strong convexity assumption on the potential $\logpart$\citep{bubeck}\TODO{bubeck online learning 2011, or monograph?}.
 
-\subsection{Stochastic mirror descent}
-
-\fdk{
-- Would mirror descent and the equivalence with MLE/MAP be introduced earlier?
-- Mirror descent \citep{nemirovski1983problem,beck2003mirror}
-- (also called NoLips \citep{bauschke2017descent}, 
-relative gradient descent \citep{hanzely2018fastest},
-Bregman proximal gradient, ...)
-}
-
-{\bf In online learning, } \fdk{bounded gradient assumption}
-
-{\bf In smooth optimization, } there has been recent advances in optimization that go beyond the Euclidean case, 
+More recently, advances in optimization that go beyond the Euclidean case, 
 to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
 neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
-\citep{bauschke2017descent,lu2018relatively}.
+\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
 Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm), 
 the $\mu$-strong convexity and $L$-smoothness of a function $f$ 
-relative to a reference function $A$ is defined 
-as 
+relative to a reference function $A$ is defined  as 
 \aligns{
@@
 Those conditions ensure the linear convergence rate of mirror descent with $A$ as the reference function.
-Maximum likelihood in exponential families
-that fits the theory developped for mirror descent well. 
-Indeed, 
+ Exponential families log-likelihood perfectly fits into this theory.
 \aligns{
@@
 }
-is $1$-smooth and $1$-strongly convex relative to $A$,
-and the the optimality gap is the divergence
-\aligns{
-	f(\theta) - f(\theta_*) = 
-	\cB_A(\theta, \theta_*)
-	= \KL(p(\cdot\cond\theta_*);\, p(\cdot\cond\theta)),
-}
-where $\theta_* = \arg\min_\theta f(\theta)$.
-But the known results for stochastic mirror descent do not give convergence rates 
+is $1$-smooth and $1$-strongly convex relative to $A$.
+However, the MAP is stochastic, and known results for stochastic mirror descent do not give convergence rates 
 for all exponential families.
@@
 are not strongly convex.
-
-The work of \citet{bauschke2017descent,lu2018relatively}
-has renewed interest for analyses of mirror descent in the smooth setting, 
-and developping new tools to analyse problems
+Relative smoothness has renewed interest for analyses of mirror descent
 without relying on strong-convexity in an underlying norm.
 However, they do not yet address the problem satisfactorily.
+
+\TODO{table of results of all 3 papers.}
 

### 2021-10-08 aistats.tex
some comments

@@
 \newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
+\usepackage{suffix}
+\WithSuffix\newcommand\expect*[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator
+
 \newcommand{\cond}{\,\vert\,}
@@
 -- e.g. exponential distribution, 
-Gaussian and Laplace distributions with known mean,
+Gaussian\fdk{with known variance?} and Laplace distributions with known mean,
 Pareto distribution with known minimum value, 
-Weibull distribution with known shape k
+Weibull distribution with known shape $k$
 -- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
@@
 \end{align}
-holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$. 
+holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\fdk{where does this $0.21$ come from?}
 Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
@@
 In fact, it is possible to write such a decomposition for any Bregman divergence
-Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+Let $\tilde \theta_n := \expect*{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
 As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like

### 2021-10-08 aistats.tex
notation questions

@@
 
-\hypersetup{draft}
+%\hypersetup{draft}
 
@@
  \begin{equation}
-	\expect{\bregmanconj( \m_*; \hat \m_n^\text{MAP})}
+	\expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
 	\leq \begin{cases}
@@
 
+\newpage
+\section{Notation convention}
+We use 
+\begin{itemize}
+	\item $\theta, \nat$ for natural parameters
+	\item $\m$ for mean parameters
+	\item $X_1,\ldots,X_n$ for data
+	\item $\cX$ for the data space 
+	\item $T$ for the sufficient statistics
+	\item $\nu$ for the base measure
+	\item $A, A^*$ for the log-partition and its dual
+	\item $\cB_A$ for the Bregman divergence induced by $A$
+	\item $n$ for the number of samples
+	\item $(\theta^*,m^*)$ for the optimum
+	\item $\gamma$ for the step-size
+	\item $\mu,\sigma^2$ for the parameters of a Gaussian
+	\item $\hat\theta_n$ for estimates
+	\item $\tilde\theta_n$ for $\expect{\hat\theta_n}$, which is biased
+	\item $(L, \mu)$ for strong-convexity and smoothness 
+\end{itemize}
+
+\fdk{
+	Do we use $\alpha, \beta$ anywhere? 
+	Might use those for smoothness/strong-convexity rather than $\mu, L$.
+} 
+\fdk{
+	The superscript ${}^*$ denotes the convex conjugate when applied on a function 
+	and the optimum when applied to a parameter. 
+	How about we use $\theta_n, \theta_*$ for parameter at time $n$ and at the optimum 
+	and $f^*$ for the convex conjugate?
+}
+\fdk{
+	On $X_i$ for data; I'm more used to seeing $x_i$ for data, which is the realization of the random variable $X_i$. 
+	We don't talk about random variables and don't need the distinction? 
+}
+
 \end{document}

### 2021-10-09 aistats.tex
add table

@@
 When explicit, we will write the random variable $T = T(X)$ for lightness.
-For convenience, we focus on steep exponential families with $\logpart$ a convex function of Legendre type.
-In particular, the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex.
-This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}, \nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
+For convenience, we focus on steep, regular exponential families 
+where $\logpart$ is strictly convex function of Legendre type,
+and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
+(See \citet{barndoffnielsen2014information}).
+This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
 
@@
 
-\subsection{Relative Smoothness}
 Mirror Descent (MD) \footnote{Mirror Descent is going by many names : Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.} \citep{nemirovski1983problem, beck2003mirror} was first introduced to optimize non-smooth functions, typically under bounded gradient assumption on the objective $f$ or strong convexity assumption on the potential $\logpart$\citep{bubeck}\TODO{bubeck online learning 2011, or monograph?}.
+
+\fdk{Online learning;
+\citet{azoury2001relative,freund1996predicting,dasgupta2007online}
+}
 
@@
 are not strongly convex.
+
 Relative smoothness has renewed interest for analyses of mirror descent
 without relying on strong-convexity in an underlying norm.
+Recent works have made attempts to generalize the assumptions 
+that leads to the convergence of mirror descent 
+with strongly-convex reference functions, 
+such as a uniform bound on the variance of the stochastic gradients $\tilde g(\theta)$,
+$\expect*{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq \sigma^2$, 
+or only requiring this bounded variance at the minimum $\theta_*$.
 However, they do not yet address the problem satisfactorily.
+We present those generalization here 
+and they are summarized in the context of maximum likelihood estimation 
+in \cref{tbl:assumptions}.
 
-\TODO{table of results of all 3 papers.}
+\begin{table*}[t]\begingroup
+\renewcommand*{\arraystretch}{1.25}%
+\centering
+\begin{tabular}{p{.35\textwidth}p{.6\textwidth}}
+\toprule
+Assumption &  \\
+\midrule
+Strongly convex and bounded variance\newline
+\citep[e.g.,][]{??}
+&
+$
+\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
+\quad 
+\expect[x]{\norm{\m - x}}^2\leq\sigma^2
+$
+\\
+Uniformly bounded ``variance''\newline
+\citep{hanzely2018fastest}
+& 
+	$
+	\expect[x]{
+	\begin{aligned}
+	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}\\
+	+\, \cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
+	\end{aligned}
+	}
+	\leq 
+	\eta C
+	$
+\\
+Bounded ``variance'' at $\theta_*$\newline
+\citep{dragomir2021fast}
+&
+	$\expect[x]{
+		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
+	} \leq 2 \eta^2 C$
+\\
+Bounded optimality gap\newline 
+\citep{dorazio2021stochastic} 
+&
+? 
+\\
+\bottomrule
+\end{tabular}
+\caption{Caption}
+\label{tbl:assumptions}
+\endgroup\end{table*}
 
-\textbf{An adaptation of the bounded variance,}
+\subsection{Bounded variance}
+An adaptation of the bounded variance,
 investigated by \citet{hanzely2018fastest}, 
@@
 	\lin{
-		\nabla f(\theta) - \tilde g, 
-		\theta^+\!\!\paren{\eta \tilde g } - \theta^+\!\!\paren{\eta \nabla f(\theta)}
+		\nabla f(\theta) - \tilde g(\theta), 
+		\theta^+\!\!\paren{\eta \tilde g(\theta)} - \theta^+\!\!\paren{\eta \nabla f(\theta)}
 	}
@@
 this definition recovers the variance of the stochastic gradient
-$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta_t) - \tilde g_t}{}^2}\leq C$.
+$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
 More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
@@
 \aligns{
-	\frac{1}{\mu} \expect{\norm{\nabla f(\theta_t) - \tilde g_t}{}^2_*},
+	\frac{1}{\mu} \expect{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*},
 }
@@
 a deterministic and stochastic update is bounded in expectation,
+\alignn{\label{eq:hanzely-in-divergence}
+	\begin{aligned}
+	&
+	\mathbb{E}_{\tilde g}\bigg[
+	\cB_{A^*}\paren{\m - \eta \tilde g(\theta), \m - \eta \nabla f(\theta)}
+	\\
+	&\hphantom{\mathbb{E}_{\tilde g}[}
+	+ \cB_{A^*}\paren{\m - \eta \nabla f(\theta), \m - \eta \tilde g(\theta)}
+	\bigg]
+	\leq 
+	\eta C.
+	\end{aligned}
+}
+
+{\bf For a maximum likelihood problem, }
+where $\tilde g(\theta) = \m - x$ 
+where $x \sim p(x\cond\theta_*)$,
+and $\nabla f(\theta) = \m - \m_*$,
 \aligns{
 	&
-	\mathbb{E}_{\tilde g_t}\bigg[
-	\cB_{A^*}\paren{\mu - \eta g, \mu - \eta \nabla f(\theta)}
+	\mathbb{E}_{x}\bigg[
+	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}
 	\\
-	&\hphantom{\mathbb{E}_{\tilde g}[}
-	+ \cB_{A^*}\paren{\mu - \eta \nabla f(\theta), \mu - \eta g}
+	&\hphantom{\mathbb{E}_{\tilde g_t}[}
+	\cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
 	\bigg]
@@
 }
+\fdk{How do we show that this fails?}
 
+\subsection{Bounded variance at the minimum}
+
+\citep{dragomir2021fast} is similar to \cref{eq:hanzely-in-divergence}, 
+where we take only the first part of it 
+and care only about gradients at the minimum, replacing 
+$\tilde g(\theta)$ and  $\nabla f(\theta)$ with $\tilde g(\theta_*)$ and $\nabla f(\theta_*) = 0$
+\aligns{
+	\expect[\tilde g]{
+		\cB_{A^*}(\m - 2\eta \tilde g(\theta_*), \m)
+	} \leq 2 \eta^2 C.
+}
 {\bf For a maximum likelihood problem, }
-where $\tilde g = \mu - x$ 
-where $x \sim p(x\cond\theta_*)$,
-and $\nabla f(\theta) = \mu - \mu_*$,
 \aligns{
-	&
-	\mathbb{E}_{x}\bigg[
-	\cB_{A^*}\paren{(1-\eta) \mu + \eta x, (1-\eta)\mu + \eta \mu_*}
-	\\
-	&\hphantom{\mathbb{E}_{\tilde g_t}[}
-	\cB_{A^*}\paren{(1-\eta)\mu + \eta \mu_*, (1-\eta) \mu + \eta x}
-	\bigg]
-	\leq 
-	\eta C.
+	\expect[x]{
+		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
+	} \leq 2 \eta^2 C.
 }
-
-\fdk{Example where this fails?}
-
-\textbf{Bounded variance at the minimum, }
-\fdk{\citep{dragomir2021fast}}
+\fdk{How do we show that this fails?}
 
@@
 \fdk{
-Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}}
+Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
+\\
+Do they do something with that? It seems to be fixed step-size only.
+}
 
@@
 }
+
 \fdk{
@@
 
+\fdk{
+	Need to check 
+}
+
 \end{document}

### 2021-10-09 main.tex
Close self-concordance discussion

@@
     &= \argmin_\natp - \log p(X_n|\natp) - \log p(\natp | \cD_{n-1}) \\
-    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
+    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0 - 1) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
     \label{eq:SBPP}
 \end{align} 
-Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0}$.
+Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0 -1}$.
 As a reminder this algorithm is defined with the update
@@
     \item $\lambda(x) = \|(A^*)'(x)\|_x^*$ (Newton decrements)
-    \item $\omega = t-\ln (1+t)$
-    \item $\omega^* = -t-\ln(1-t)$
+    \item $\omega(t) = t-\ln (1+t)$
+    \item $\omega^*(t) = -t-\ln(1-t)$
 \end{itemize}
-Note that $\omega^*$ is convex and monotonne.
+Note that $\omega^*$ is convex and monotone, and $\omega^*(t) \leq t^2, \forall t\leq 0.68$.
 
@@
 \]
-Assume $\|\mu^*-\mu_n\|_{\mu^*} < c$,
+Assume $\|\mu^*-\mu_n\|_{\mu^*} < 0.21$, then
 \[
-    \mathcal{B}_{A^*}(\mu^*,\mu_n) \leq \omega^*\left(\frac{\|\mu^*-\mu_n\|_{\mu^*}}{1-c}\right) .
+    \bregmanconj(\mu^*,\mu_n) \leq \|\mu^*-\mu_n\|_{\mu^*}^2
 \]
+so we have a local quadratic behavior. To put into euclide's world. Problem is it requires global self-concordance.
 

### 2021-10-09 aistats.tex
Updates from Overleaf

--- 2021-10-08 aistats.tex
@@
 
-\usepackage[capitalise]{cleveref}
 \newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
@@
 \newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
-\usepackage{suffix}
-\WithSuffix\newcommand\expect*[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator
-
 \newcommand{\cond}{\,\vert\,}
@@
 \newcommand{\bregmanconj}{\cB_{\logpart^*}}
-\newcommand{\nat}{\theta}
-\newcommand{\m}{m}
-\newcommand{\meanp}{\m}
+\newcommand{\natp}{\theta}
+\newcommand{\meanp}{\mu}
 \newcommand{\decrement}{D}
@@
 
-\newcommand{\MAPm}{\hat \m_n}
-\newcommand{\MAPt}{\hat \nat_n}
-\DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}
+\newcommand{\MAPm}{\hat \mu_n}
+\newcommand{\MAPt}{\hat \natp_n}
 
-\hypersetup{draft}
+%\hypersetup{draft}
 
@@
 
-\aistatstitle{Convergence Rates for the MAP of an Exponential Family, and Stochastic Mirror Descent -- an Open Problem}
+\aistatstitle{Looking for Convergence Rates\\ for the MAP of the Exponential Family}
 
@@
 \begin{abstract}
-We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family. 
-Surprisingly, we found no general solution to this problem in the literature. 
-In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
-After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD), 
-yet it falls out of scope for modern analysis.
-We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
+We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for the exponential family. 
+Surprisingly, we found no solution to this problem in the literature -- e.g. after seeing 5 samples, we do not know how many bits away (in expectation) our model is from the true distribution.
+After displaying some properties and special cases of this problem, 
+we show it is an application of several optimization algorithms.
+Yet it falls out of scope for current analysis, thus highlighting areas for progress.
 \end{abstract}
-
-%In particular, no rates hold in the few samples regime  -- e.g. after seeing 5 samples from a gaussian, we do not know how many bits away from the true distribution we should expect our model to be.
-%No rates hold either when the loss is ill-behaved : infinite domain, not smooth, not self-concordant
 
@@
 \begin{equation}
-	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta)},
+	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta_*)},
 \end{equation}
@@
 There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
-As we show, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
-Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
+As we will prove, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
+Despite recent advances in this field, our convergence rates are still lacking ; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
 
@@
 %TODO complete
-However, none of the theory we found could address for instance the case of a Gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions. 
+However, none of the theory we found could address for instance the case of a gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions. 
 %TODO kind of redundant "problem - question - problem" paragraphs, not disentangling stats and optim. To improve
@@
 \begin{itemize}
-	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the Gaussian variance estimation problem.
+	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the gaussian variance estimation problem.
 	\item We show the connections between the optimization and statistical estimation view of the problem and discuss where current theories fail to fully capture the problem.
@@
 
-Exponential Families are to statistics what linear models are to regression.
-Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models. 
+Exponential Families are the linear model of probabilities.
+Their convexity along with their modeling power make them one of the most widely used statistical models. 
 Here we review some of their properties. 
 We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
-The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
+The exponential family for data $X\in \cX$ with natural (or primal) parameter $\natp$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
 \begin{equation}
-	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
+	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
 \end{equation}
-Here $\logpart$ is the log-partition function -- i.e. the normalization term
+Here $\logpart$ is the log-partition function -- i.e. the normalization factor
 \begin{align}
-    \logpart(\nat) = \log \int e^{\nat^\top T(x)} \nu(dx) \; .
+    \logpart(\natp) = \log \int e^{\natp^\top T(x)} \nu(dx) \; .
 \end{align}
 When explicit, we will write the random variable $T = T(X)$ for lightness.
-For convenience, we focus on steep exponential families with $\logpart$ a convex function of Legendre type.
-In particular, the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex.
+This model is defined on the open convex set $\natp\in \Theta$ such that $\logpart(\natp) < \infty$. 
 This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}, \nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
-
-% TODO sort out relationships between : steep family, minimal statistics and Legendre type. What implies what ?
 
@@
 \begin{align}
-    \nabla\logpart(\nat) &=  \expect[p(T|\nat)]{T} =: \meanp \\
-    \nabla^2 \logpart(\nat) &= \Cov_\nat[T] \geq 0
+    \nabla\logpart(\natp) &=  \expect[p(T|\natp)]{T} =: \meanp \\
+    \nabla^2 \logpart(\natp) &= \Cov_\natp[T] \geq 0
 \end{align}
 where $\meanp$ is called the mean (or dual) parameter.
+$\meanp$ lives in the open convex set $\cM$ within the convex hull of $T(\cX)$.
+%TODO why is it convex ? 
+%It is open because it is the reciprocal image of an open set by the continuous mirror map
 From now on, we assume that the sufficient statistic $T$ is minimal 
--- e.g. its covariance is definite positive for all $\nat$ -- 
-then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\nat$ and mean parameters $\m$.
-$\meanp$ lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
-We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both represent the same distribution.
+-- e.g. its covariance is definite positive for all $\natp$ -- 
+then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
+We will write interchangeably $\mu$ or  $\natp$ depending on the context, being aware that both represent the same distribution.
 At this point it is useful to introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
 \begin{align}
-	\conj(\m) = \langle \m, \nat \rangle - \logpart(\nat) \; .
+	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
 \end{align}
 which matches the common notion of \textit{entropy} in information theory.
-The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).
+The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
 
@@
 	\includegraphics{duality}
-	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$.
+	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\natp, \meanp$.
 Figure copied from \citet{kunstner2020homeomorphic}.
@@
 \paragraph{Bregman Divergences}
-A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\nat$ and $\nat_0$
+A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\natp$ and $\natp_0$
 \begin{align}
-    \bregman (\nat ; \nat_0)
-    & = \logpart(\nat) - \logpart(\nat_0) 
-    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle
+    \bregman (\natp ; \natp_0)
+    & = \logpart(\natp) - \logpart(\natp_0) 
+    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
 \end{align}
-with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. 
+with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
 It is a non-symmetric measure of distance between parameters.
@@
 \begin{align}
-	\bregman (\nat ; \nat_0)
+	\bregman (\natp ; \natp_0)
     = \bregmanconj ( \meanp_0 ; \meanp) \; .
@@
 \paragraph{Conjugate Prior}
-One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
+One conjugate prior \citep{agarwal2010geometric} for $p(X|\natp)$ is
 \begin{align}
-    p(\nat) 
-    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
-    &\propto \exp(n_0 \m_0^\top \nat - n_0 \logpart(\nat))
+    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
 \end{align}
-where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
-This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
-Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.
+where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is introduced above.
+Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
+The negative log-likelihood of the prior is
+\begin{align*}
+    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
+\end{align*}
+which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
+Then the joint log-likelihood of $\mD =(X_1,\dots,X_n)$ and $\natp$ is
+\begin{multline}
+    -\log p(\mD|\natp)p(\natp) 
+    = (n_0+n) \logpart (\natp) \\
+    - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T_i \right) + \cst
+    \label{eq:joint_likelihood}
+\end{multline}
+where $T_i=T(X_i)$.
+This means the posterior given $\mD$ is part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T_i , -(n_0 + n))$.
 
 \paragraph{Maximum A Posteriori (MAP).}
-Using Bayes rule, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
+Minimizing~\eqref{eq:joint_likelihood} over $\natp$ yields the MAP estimate
 \begin{align}
-	p(\nat \cond \mD_n)
-	%\propto p(\mD|\nat)p(\nat) 
-    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt))
-    \label{eq:joint_likelihood}
+    \hat \natp = \argmin_\natp -\log p(\mD|\natp) + n_0 \bregman(\natp ; \natp_0) \; .
 \end{align}
-reaching its maximum in $\MAPt$, the MAP estimate verifying the first order optimality conditions
+If they can be satisfied, the MAP verifies the first oder optimality conditions
 \begin{align}
-    \nabla \logpart(\MAPt) = \MAPm
+    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
 \end{align}
-where $T_i=T(X_i)$.
 When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
 \begin{align}
-	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
+	\hat \mu_\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
 \end{align}
-This last equation is known as moment matching.
-The MLE and MAP estimates are statistics of the dataset $\mD_n$. 
-Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.
+These rules are known as moment matching.
+The MLE and MAP estimates are statistics of the dataset $\mD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
 
@@
 
-Let $\nat^* := \argmin \expect[X\sim\cD]{-\log p(X|\nat)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
+Let $\natp^* := \argmin \expect[X\sim\cD]{-\log p(X|\natp)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
 Then the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
 \begin{multline}
-    \expect[X\sim \cD]{-\log p(X|\nat) + \log p(X|\nat^*) } \\
-	= \KL( p(.|\nat^*) ; p(.|\nat)) \; .
+    \expect[X\sim \cD]{-\log p(X|\natp) + \log p(X|\natp^*) } \\
+	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
 \end{multline}
@@
 \boxed{
-	\KL( p(.|\nat^*) ; p(.|\nat))
-    = \bregman (\nat ; \nat^*)
+	\KL( p(.|\natp^*) ; p(.|\natp))
+    = \bregman (\natp ; \natp^*)
     = \bregmanconj ( \meanp^* ; \meanp)
@@
 
-Our question is: how does the KL behave when $\nat$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities ?
+Our question is: how does the KL behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities
 \begin{align}
@@
 	\label{eq:bregmanMAP}
-	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )} 
+	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \mu_0 + \smallsum_i T_i}{n_0+n} \right )} 
 \end{align}
-where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
+where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d as $X_i\sim \natp^*$?
 More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.
@@
 This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
-However there is a connection between the random variable $T(X)$ and the metric $\logpart$. 
+A key difference though, is that the random variable $T(X)$ is connected to the metric $\logpart$. 
 Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable. 
-For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
+For instance, if we plug in $\conj(\mu)= -\log(\mu)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
 
@@
 
-\subsection{MAP as Stochastic Mirror Descent}
-\label{ssec:MAP=SMD}
-The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
-\begin{align*}
-	\hat \nat_{n}
-	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
-    %\label{eq:SBPP}
-\end{align*}
-which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
-\begin{align}
-		\hat \nat_{n}
-	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
-    \label{eq:SMD}
-\end{align}
-where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$. 
-We recognize the update of Stochastic Mirror Descent (a.k.a.
-Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
-This parallel means it would be possible to obtain a finite convergence rate for the  MAP from an analysis of these algorithms. 
-We give a special focus on this approach in Section~\ref{sec:optimization}.
-
 \subsection{Asymptote}
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-Asymptotic results typically show that natural parameters $\nat$ asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
+Asymptotic results typically show that natural parameters $\natp$ asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
 Here the MAP is more simply expressed with mean parameters $\meanp$.
@@
 \begin{align}
-    \bregmanconj(\m^* ; \m) 
-    &=\half  \norm{\m^* - \m}^2_{\nabla^2\conj(\m^*)}
-    + O(\norm{\m - \m^*}^3)
+    \bregmanconj(\mu^* ; \mu) 
+    &=\half  \norm{\mu^* - \mu}^2_{\nabla^2\conj(\mu^*)}
+    + O(\norm{\mu - \mu^*}^3)
 \end{align}
@@
     \mF^*
-    :=\nabla^2\conj(\m^*) 
-    = \nabla^2\logpart(\nat^*)^{-1} 
-    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
+    :=\nabla^2\conj(\mu^*) 
+    = \nabla^2\logpart(\natp^*)^{-1} 
+    = \Cov_{\natp^*}[T(X)]^{-1}  \; .
 \end{align}
@@
 
-\subsection{Quadratic Case}
-As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
-For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
-Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances
+\subsection{Euclide's World}
+As another reference point, if we make restrictive assumptions on the log-partition function, we can ensure our objective will behave at least as nicely as a quadratic.
+\paragraph{If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
 \begin{align}
-	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
+    \bregmanconj(\mu^* ; \mu) 
+    &\leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu} \\
+    &\leq 2L \norm{\mu^* - \mu}
 \end{align}
-Thanks to the independence of samples we can break down the MLE into individual point's contributions
+so $\bregmanconj$ is $2L$-Lipschitz.
+\paragraph{If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $L^{-1}$-strongly convex), then
 \begin{align}
-	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2} 
-	=\frac{\Var(T)}{2n} 
+    \bregmanconj(\mu^* ; \mu) 
+    \leq \frac{L}{2} \norm{\mu^* - \mu}^2
 \end{align}
-Adding a reference mean $\m_0$ to get the MAP yields
-\begin{align}
-	\expect{\bregmanconj(\meanp^*; \MAPm)}
-	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}
-	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
-	\label{eq:MAP_quadratic}
-\end{align}
-so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. We would like to get a similar result for arbitrary exponential families.
-If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.
-
-{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
-\begin{align}
-    \bregmanconj(\m^* ; \m) 
-    &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
-    &\leq 2L \norm{\m^* - \m}
-\end{align}
-so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
-
-{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex), then
-\begin{align}
-    \bregmanconj(\m^* ; \m) 
-    \leq \frac{L}{2} \norm{\m^* - \m}^2
-\end{align}
-so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
-
-\subsection{Locally Quadratic Case}
-All Bregman divergences are locally quadratic. 
-Under some assumptions, we can quantify  when this quadratic behavior  kicks in.
-For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}. 
-This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant 
--- e.g. exponential distribution, 
-Gaussian\fdk{with known variance?} and Laplace distributions with known mean,
-Pareto distribution with known minimum value, 
-Weibull distribution with known shape $k$
--- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
-\begin{align}
-	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
-\end{align}
-holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\fdk{where does this $0.21$ come from?}
-Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
-
-\TODO{Reference \citet{ostrovskii2021finite}.} ( as well as 
-\citet{anastasiou2017bounds},
-\citet{marteauferey2019beyond})
-
-All these works give \textit{large sample} results -- results that hold only for $n\geq N$ for some constant $N$ -- but none of them  apply to small $n$.
+so $\bregmanconj$ is upper bounded by a quadratic, and we get the same rate as a Gaussian.
 
@@
 In fact, it is possible to write such a decomposition for any Bregman divergence
-Let $\tilde \theta_n := \expect*{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \mu_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
 As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
 \begin{align}
-	\expect{\bregmanconj(\m^* ; \hat \m_n)} 
-	&= {\bregmanconj(\m^* ; \tilde \m_n)}
-	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
+	\expect{\bregmanconj(\mu^* ; \hat \mu_n)} 
+	&= {\bregmanconj(\mu^* ; \tilde \mu_n)}
+	+ {\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}}
 	\label{eq:primal_pivot}
@@
 
-\textbf{Remark:} In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point.
-An estimator will be unbiased if its primal expectation is non-zero.
-This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
+Remark : In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point. An estimator will be unbiased if its primal expectation is non-zero. 
+This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\mu^* ; \tilde \mu_n)} \leq \frac{2}{n(n-2)}$.
 A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.
@@
 We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
-In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
+In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \mu_n^\text{MLE},\hat \mu_n^\text{MAP},\mu_n,\tilde \mu_n, \mu^*$ and $\mu_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
 \RLP{Tune figures.}
@@
 
+\subsection{Gaussian Mean}
+\paragraph{$\ell^2$-norm Analysis}
+In the Euclidean realm, the average is bounded by variance becomes a bound on the variance of the average
+\begin{align}
+	\Var T(X) = \E \norm{T(X) - \mu^*}^2 = \sigma^2 \\
+	\implies \E \norm{\mu^* -  \inv{n}  \smallsum_i T(x_i)}^2 = \frac{\sigma^2}{n} 
+\end{align}
+eg the variance of the mean is $n$ times smaller than the variance of the samples.
+Adding a reference mean $\mu_0$ to get the MAP yields
+\begin{align}
+	\E \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2 
+	&= \frac{n}{(n+n_0)^2} \sigma^2 +  \frac{n_0^2}{(n+n_0)^2} \norm{\mu^* -  \mu_0}^2 \\
+	&= O\left(\frac{\sigma^2}{n} \right) + O\left(\frac{\norm{\mu^* -  \mu_0}^2}{n^2} \right)
+\end{align}
+so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. Let's see if we can get similar estimates for arbitrary exponential families !
+
 \subsection{Gaussian Variance}
@@
 The density is	$p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}$.
-Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$. 
-Mean and natural parameters are roughly inverse of each other $\nat = -\inv{2 \m}$.
+Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
+Mean and natural parameters are roughly inverse of each other $\natp = -\inv{2 \mu}$.
 Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
 \begin{align}
-	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi) \\
-	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
+	\logpart (\natp) &= - \half \log(-\natp)  + \half \log(\pi) \\
+	\conj(\mu) &= \half\left( -\log(\mu) + \log\frac{\pi}{2} - 1 \right) \; .
 \end{align}
 Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
-It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, eg a negative Gamma distribution.
+It means the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative Gamma distribution.
 It also means $\bregman$ and $\bregmanconj$ have the same shape
 \begin{align}
-	\bregmanconj( \m_*; \m_n) 
-	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) \\
-	\bregman( \nat_n; \nat_* )
-	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
+	\bregmanconj( \mu_*; \mu_n) 
+	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
+	\bregman( \natp_n; \natp_* )
+	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
 \end{align}
-In other words, this divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function $\phi$
+In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
 \begin{align}
@@
 \begin{theorem}[MLE Tight Bound]
-	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
+	The MLE of $\cN(0,\mu_*)$ is $\hat \mu_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
 	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
 	\begin{align}
-		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
+		 \expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE}) }
 			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
@@
  \begin{align}
-	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
+	b_n = \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
  \end{align}
-The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
+The expected suboptimality of the MAP of $\cN(0,\mu^*)$ with prior hyper-parameters $(n_0,\mu_0)$ is
  \begin{equation}
-	\expect{\bregmanconj( \m_*; \hat \m_n^\text{MAP})}
+	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MAP})}
 	\leq \begin{cases}
 		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
-		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
+		\frac{1}{n_0 \frac{\mu_0}{\mu^*} +n-2} + b_n \ \text{if}\ n\geq 2
 	\end{cases}
@@
 This inequality highlights a clear variance-bias decomposition.
-In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
-For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
+In particular, there is no bias term when $\frac{\mu_0}{\mu^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
+For instance, when $n_0=1$, it encourages us to set $\mu_0 = 2 \mu^*$.
 Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$.
@@
 
+\section{"Burn In" Statistical Analysis}
+Self-concordant line of work : 
+\citet{kakade2010learning},
+\cite{anastasiou2017bounds},
+\citet{ostrovskii2021finite}.
+
 \section{Optimization Perspective}
-\label{sec:optimization}
-As we saw in Section~\ref{ssec:MAP=SMD}, MAP can be interpreted as Stochastic Mirror Descent (SMD). 
-There are two consequences of this observation: 1.
-we may obtain a convergence rate for MAP from an optimization analysis, and 2.
-any insights gained from the MAP may inform further design and analysis of SMD.
+In the context of optimization, why do we care?
 
-\subsection{Relative Smoothness}
-Mirror Descent (MD) \footnote{Mirror Descent is going by many names : Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.} \citep{nemirovski1983problem, beck2003mirror} was first introduced to optimize non-smooth functions, typically under bounded gradient assumption on the objective $f$ or strong convexity assumption on the potential $\logpart$\citep{bubeck}\TODO{bubeck online learning 2011, or monograph?}.
-
-More recently, advances in optimization that go beyond the Euclidean case, 
-to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
-neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
-\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
-Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm), 
-the $\mu$-strong convexity and $L$-smoothness of a function $f$ 
-relative to a reference function $A$ is defined  as 
-\aligns{
-	\mu \cB_{A}(x, y)
-	\leq 
-	\cB_f(x,y)
-	\leq 
-	L \cB_A(x,y).
-}
-Those conditions ensure the linear convergence rate of mirror descent with $A$ as the reference function.
- Exponential families log-likelihood perfectly fits into this theory.
-\aligns{
-	f(\theta) = A(\theta) - \expect[x\sim p(x\cond\theta)]{\lin{x, \theta}}
-}
-is $1$-smooth and $1$-strongly convex relative to $A$.
-However, the MAP is stochastic, and known results for stochastic mirror descent do not give convergence rates 
-for all exponential families.
-
-{\bf Traditional results}
-in stochastic optimization,
-such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
-rely on the assumption that the reference function $A$ 
-is itself strongly convex relative to a norm $\norm{\cdot}$,
-(meaning that $f$ is strongly convex). 
-This apply for the MLE problem of a Gaussian distribution 
-with fixed variance, 
-but not when the covariance is unknown,
-or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
-are not strongly convex.
-Relative smoothness has renewed interest for analyses of mirror descent
-without relying on strong-convexity in an underlying norm.
-However, they do not yet address the problem satisfactorily.
-
-\TODO{table of results of all 3 papers.}
-
-\textbf{An adaptation of the bounded variance,}
-investigated by \citet{hanzely2018fastest}, 
-defines the following bounded quantity.
-Letting $\theta^+(v)$ be a step from $\theta$ with vector $v$, 
-$\theta^+(v) = \nabla A^*(\nabla A(\theta) - v)$,
-\alignn{\label{eq:assumption-hanzely}
-	\expect[\tilde g]{
-	\lin{
-		\nabla f(\theta) - \tilde g, 
-		\theta^+\!\!\paren{\eta \tilde g } - \theta^+\!\!\paren{\eta \nabla f(\theta)}
-	}
-	}
-	\leq 
-	\eta C,
-}
-for some step-size $\eta$ and constant $C$. Let us unpack this quantity;
-
-{\bf For a Gaussian} with known variance, 
-with $A(\theta) = \frac{1}{2}\norm{\theta}^2$, 
-this definition recovers the variance of the stochastic gradient
-$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta_t) - \tilde g_t}{}^2}\leq C$.
-More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
-the left-hand side of \cref{eq:assumption-hanzely} is bounded by 
-\aligns{
-	\frac{1}{\mu} \expect{\norm{\nabla f(\theta_t) - \tilde g_t}{}^2_*},
-}
-and a uniform bound $C$ exist if the variance of the stochastic gradient is bounded.
-
-{\bf In divergence,} this condition states that the (symmetrized) divergence between 
-a deterministic and stochastic update is bounded in expectation,
-\aligns{
-	&
-	\mathbb{E}_{\tilde g_t}\bigg[
-	\cB_{A^*}\paren{\mu - \eta g, \mu - \eta \nabla f(\theta)}
-	\\
-	&\hphantom{\mathbb{E}_{\tilde g}[}
-	+ \cB_{A^*}\paren{\mu - \eta \nabla f(\theta), \mu - \eta g}
-	\bigg]
-	\leq 
-	\eta C.
-}
-
-{\bf For a maximum likelihood problem, }
-where $\tilde g = \mu - x$ 
-where $x \sim p(x\cond\theta_*)$,
-and $\nabla f(\theta) = \mu - \mu_*$,
-\aligns{
-	&
-	\mathbb{E}_{x}\bigg[
-	\cB_{A^*}\paren{(1-\eta) \mu + \eta x, (1-\eta)\mu + \eta \mu_*}
-	\\
-	&\hphantom{\mathbb{E}_{\tilde g_t}[}
-	\cB_{A^*}\paren{(1-\eta)\mu + \eta \mu_*, (1-\eta) \mu + \eta x}
-	\bigg]
-	\leq 
-	\eta C.
-}
-
-\fdk{Example where this fails?}
-
-\textbf{Bounded variance at the minimum, }
-\fdk{\citep{dragomir2021fast}}
-
-\textbf{Bounded optimality gap,}
-\fdk{
-Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}}
-
-\bibliographystyle{apalike}
-\bibliography{references.bib}
-
-\clearpage
-\section{Group Meeting}
-\begin{enumerate}
-	\item new plots : mirror map, and gaussian trajectories. Add trajectories to paper, with background level set, and new legend : primal/dual expectation
-	\item the MLE is biased according to our definition.
-See plots.
-Does this definition of bias make sense ?
-An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean.
-Remark : the MLE is biased for standard parameters.
-	\item exciting pap er\citet{kakade2010learning}
-	\begin{enumerate}
-		\item assumption in the spirit of infinite order self-concordance only at the optimum
-		\item After a burn-in phase, the loss behaves quadratically. No precise description of the burn-in length.
-		\item \TODO{it is worth writing down until the end what we are obtaining with self-concordance.}
-		\item idea 1 : try to express some of their results with the mean parameter instead.
-		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD.
-(exploiting the link MAP -> optimization).
-How do we prove the linear convergence of Newton to the quadratic regime ?
-That is the problem at stake here.
-	\end{enumerate}
-	\item Intro : positioning between optimization and statistics.
-Start with stats.
-Convince the reader that this is an essential problem, with asymptotic rate.
-but finite is elusive.
-Inspired by recent work (lepriol2021 , bach2013) we want to use optimization to give a rate.
-We fail, but we reveal a hole in current analysis.
-Finite rate would allow us to quantify the importance of the prior, and also help us understand behavior for small numbers.
-	\item Intro : 1 clear question and sharper contributions, with references to sections, as a memo for the reviewer writing his review.
-"investigate bias-variance..." "we show all these analysis suffer from this problem."
-	\item Organization : add optim parallel in insigths section, and then sprinkle references until the optim section.
-	\item Optim : find members of the exponential family where the rate of Dragomir et al. holds.
-	\item Technical Background : be precise and add reference for the steep exponential family of Legendre type which guarantees that our sets are open, and have a solution in the middle.
-	References: \begin{itemize}
-		\item \href{https://www.jstor.org/stable/4616462?seq=1#metadata_info_tab_contents}{Existence of Maximum Likelihood Estimates for Multi-Dimensional Exponential Families}, 
-		\item The singly truncated normal distribution: A non-steep exponential family, 
-		\item Statistical exponential families: A digest with flash cards,
-		\item Information and Exponential Families In Statistical Theory, 
-		\item Fundamentals of Statistical Exponential Families with Applications in Statistical Decision Theory.
-	\end{itemize}  
-\end{enumerate}
-
-\section{Blurbs/ideas?}
-
-\subsection{SGD blurb}
-\fdk{
-%
 There is a long line work bounding the convergence rate of stochastic gradient descent.
@@
 which matches the asymptotic rate of unbiased estimation,
-through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler} 
-\\
+through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler} \footnote{
 Those papers assume the stochastic gradients are bounded, 
 but small modifications also work for bounded variance instead.
-\\
+}
+
 We do have extensions to other notions of bounded variance, 
@@
 For a review, see \citet{gower2019sgd}.
-\\
+
 However, for some problems, combining decreasing step-sizes and averaging is not necessary,
@@
 but holds more generally, for example for linear and logistic regressions \citep{bach2013nonstronglyconvex,moulines2011non}.
-}
 
-\subsection{Poisson likelihood?}
-\fdk{
-\citet{bauschke2017descent} and \citet{hanzely2018fastest} both use the example of Poisson inverse problems/Poisson regression
-as examples. 
-The simpler case of the MLE of a Poisson distribution is also unsolved, though. 
-In this case, $h(x) = 1/x!$, $x \in \mathbb{N}$, $A(\theta) = e^\theta$ over $\theta \in \mathbb{R}$,
-$A^*(\m) = \m \log \m - \m$ over $\m \in \mathbb{R}_+$.
-It is neither strongly-convex nor self-concordant (at least according to the standard definition, 
-although it satisfies generalized notions of self-concordance as $A'''(\theta) = A''(\theta)$).
-}
+There has been recent advances in optimization that go beyond the Euclidean case, 
+to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
+neither strongly convex in the Euclidean norm, using the concept of relative smoothness 
+\citep{birnbaum2011distributed,bauschke2017descent,lu2018relatively}.
+Instead of measuring strong-convexity and smoothness relative to the Euclidean norm (or any norm), 
+we use a reference function instead \fdk{maybe add a line}
+and apply mirror descent \citep{nemirovski1983problem,beck2003mirror}
+(also called NoLips \citep{bauschke2017descent}, something else?).
 
-\subsection{``simple'' open problem?}
+We have convergence results for the 
+application of stochastic mirror descent, 
+but they typically rely on assumptions that are too strong. 
+\begin{itemize}
+	\item The reference function is strongly convex [typical setup in online learning]
+	\item The variance is uniformly bounded
+\end{itemize}
+Some advances have been tried
+\begin{itemize}
+	\item The reference function is ``close to'' a norm
+	and looks like a \citet{hanzely2018fastest}
+	\item Attempt to adapt the ``bounded gradient at the minimum''
+	\citep{dragomir2021fast}
+	\item Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}.
+\end{itemize}
+
+Other work 
+\citet{marteauferey2019beyond}.
+
 \fdk{
@@
 
-\subsection{Connection with acceleration?}
-\fdk{
+\fdk{Application 
 The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity. 
@@
 
+\bibliographystyle{apalike}
+\bibliography{references.bib}
+
+\section{Group Meeting}
+\begin{enumerate}
+	\item new plots : mirror map, and gaussian trajectories
+	\item the MLE is biased according to our definition. See plots. Does this definition of bias make sense ? An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean. 
+	\item \citet{kakade2010learning}
+	\begin{enumerate}
+		\item assumption in the spirit of infinite order self-concordance only at the optimum
+		\item After a burn-in phase, the loss behaves quadratically. No precise description of the burn-in length.
+		\item \TODO{it is worth writing down until the end what we are obtaining with self-concordance.}
+		\item idea 1 : try to express some of their results with the mean parameter instead.
+		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD. (exploiting the link MAP -> optimization).
+	\end{enumerate}
+	\item positioning between optimization and statistics.
+\end{enumerate}
+
 \end{document}

### 2021-10-10 aistats.tex
remove packages that are not compatible with texlive 2015

--- 2021-10-09 aistats.tex
@@
 \newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
-\usepackage{suffix}
-\WithSuffix\newcommand\expect*[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator
 
@@
 \end{align}
-holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\fdk{where does this $0.21$ come from?}
+holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\footnote{The value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
 Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
@@
 In fact, it is possible to write such a decomposition for any Bregman divergence
-Let $\tilde \theta_n := \expect*{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
 As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
@@
 such as a uniform bound on the variance of the stochastic gradients $\tilde g(\theta)$,
-$\expect*{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq \sigma^2$, 
+$\expect{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq \sigma^2$, 
 or only requiring this bounded variance at the minimum $\theta_*$.
@@
 in \cref{tbl:assumptions}.
+
+\subsection{Strongly-convex reference function}
 
@@
 }
+\fdk{Could work with decreasing step-sizes}
+
 \fdk{How do we show that this fails?}

### 2021-10-10 aistats.tex
m

@@
 \usepackage{subcaption}
+\usepackage{placeins}
 \usepackage{hyperref}       % hyperlinks
@@
 \newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
+\newcommand*{\expecti}[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator
 
@@
 
+\newpage~\null\\
+\FloatBarrier
+
 \section{Optimization Perspective}
@@
 
-Mirror Descent (MD) \footnote{Mirror Descent is going by many names : Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.} \citep{nemirovski1983problem, beck2003mirror} was first introduced to optimize non-smooth functions, typically under bounded gradient assumption on the objective $f$ or strong convexity assumption on the potential $\logpart$\citep{bubeck}\TODO{bubeck online learning 2011, or monograph?}.
+Mirror Descent (MD)\footnote{%
+Also known as is going by many names: 
+Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.
+} 
+\citep{nemirovski1983problem} 
+was first introduced to optimize non-smooth functions \fdk{(Is typically encountered in non-smooth, online optimization?)},
+typically under bounded gradient assumption on the objective $f$ 
+or strong convexity assumption on the potential $\logpart$
 
-\fdk{Online learning;
-\citet{azoury2001relative,freund1996predicting,dasgupta2007online}
+\fdk{
+{\bf Online learning paragraph?}
+\citet{azoury2001relative,freund1996predicting,dasgupta2007online},
+Bubeck's paragraph
 }
@@
 }
-Those conditions ensure the linear convergence rate of mirror descent with $A$ as the reference function.
+Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
 Exponential families log-likelihood perfectly fits into this theory.
@@
 
-{\bf Traditional results}
+\subsection{Strongly-convex reference function}
+
+{Traditional results}
 in stochastic optimization,
@@
 with fixed variance, 
-but not when the covariance is unknown,
+but not in a general case. 
+when the covariance is unknown,
 or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
@@
 
-Relative smoothness has renewed interest for analyses of mirror descent
+The introduction of relative smoothness has renewed interest for analyses of mirror descent
 without relying on strong-convexity in an underlying norm.
 Recent works have made attempts to generalize the assumptions 
-that leads to the convergence of mirror descent 
-with strongly-convex reference functions, 
-such as a uniform bound on the variance of the stochastic gradients $\tilde g(\theta)$,
-$\expect{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq \sigma^2$, 
-or only requiring this bounded variance at the minimum $\theta_*$.
+that are used to control the convergence of stochastic gradient descent 
+to this setting. 
+Those include 
+bounded gradients $(\expecti{\norm{\tilde g(\theta)}^2} \leq C$, 
+bounded variance $(\expecti{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq C$
+uniformly or only at the minimum $\theta_*$,
+or bounded disparities between the minimum of $f$ at $\theta_*$ and the minima of individual functions $f_i$, 
+if $f(\theta) = \expect[i]{f_i(\theta)}$ $(\expect{f(\theta_*) - \min_\theta f_i(\theta)} < C)$.
+(\fdk{For a review of those assumptions, see that paper by Richtarik's students})
+
 However, they do not yet address the problem satisfactorily.
@@
 in \cref{tbl:assumptions}.
-
-\subsection{Strongly-convex reference function}
 
@@
 \alignn{\label{eq:assumption-hanzely}
-	\expect[\tilde g]{
+	\expect[\tilde g\!]{
 	\lin{
-		\nabla f(\theta) - \tilde g(\theta), 
-		\theta^+\!\!\paren{\eta \tilde g(\theta)} - \theta^+\!\!\paren{\eta \nabla f(\theta)}
+		\nabla\! f(\theta) 
+		- \tilde g(\theta), 
+		\theta^+\!\!\paren{\eta \tilde g(\theta)} 
+		- \theta^+\!\!\paren{\eta \nabla\! f(\theta)}
 	}
@@
 }
-for some step-size $\eta$ and constant $C$. Let us unpack this quantity;
+for some step-size $\eta$. Let us unpack this quantity;
 
-{\bf For a Gaussian} with known variance, 
-with $A(\theta) = \frac{1}{2}\norm{\theta}^2$, 
+{\bf For a Gaussian} with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$), 
 this definition recovers the variance of the stochastic gradient
@@
 More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
-the left-hand side of \cref{eq:assumption-hanzely} is bounded by 
-\aligns{
-	\frac{1}{\mu} \expect{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*},
-}
-and a uniform bound $C$ exist if the variance of the stochastic gradient is bounded.
+the LHS of \cref{eq:assumption-hanzely} is bounded by 
+$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
+and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:assumption-hanzely}
+to hold.
 
@@
 	\cB_{A^*}\paren{\m - \eta \tilde g(\theta), \m - \eta \nabla f(\theta)}
-	\\
+	\\[-1em]
 	&\hphantom{\mathbb{E}_{\tilde g}[}
@@
 }
-
 {\bf For a maximum likelihood problem, }
@@
 }
+\aligns{
+	A(\theta_*) - \lin{x, \theta_*} - \expecti{ \min_\theta A(\theta) - \lin{x, \theta} }
+}
 
+\subsection{Issues with current analyses}
+
+\fdk{They don't hold for problems we care about}
+
+\fdk{Existing techniques only deal with fixed step-sizes and show convergence to a ball}
+
+\newpage
 \bibliographystyle{apalike}

### 2021-10-11 aistats.tex
remove fancy maths commands
They made my local compilation fail because I am using pdflatex 2015

--- 2021-10-10 aistats.tex
@@
 \newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
+\newcommand*{\expecta}[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator
 
@@
 \end{align}
-holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\footnote{The value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
+holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\fdk{where does this $0.21$ come from?}
 Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
@@
 In fact, it is possible to write such a decomposition for any Bregman divergence
-Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+Let $\tilde \theta_n := \expecta{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
 As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
@@
 such as a uniform bound on the variance of the stochastic gradients $\tilde g(\theta)$,
-$\expect{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq \sigma^2$, 
+$\expecta{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq \sigma^2$, 
 or only requiring this bounded variance at the minimum $\theta_*$.
@@
 }
-\fdk{Could work with decreasing step-sizes}
-
 \fdk{How do we show that this fails?}

### 2021-10-11 aistats.tex

--- 2021-10-10 aistats.tex
@@
 In fact, it is possible to write such a decomposition for any Bregman divergence
-Let $\tilde \theta_n := \expect{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
 As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like

### 2021-10-11 aistats.tex
Correct problem posing
The order in which I brought up the concepts was not so logical.

@@
 \usepackage[capitalise]{cleveref}
-\newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
+\newcommand{\rlp}[1]{\textcolor{red}{RLP:#1}}
 \newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
 \newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}
+\newcommand{\tocite}{\textcolor{purple}{(add citation)}}
 
@@
 
+Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
+Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
+A contrario the KL is not smooth, and it can be infinite in misspecified settings.
+But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.
+
 There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
@@
 \begin{equation}
-	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
+	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
 \end{equation}
+where $\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
 Here $\logpart$ is the log-partition function -- i.e. the normalization term
 \begin{align}
-    \logpart(\nat) = \log \int e^{\nat^\top T(x)} \nu(dx) \; .
+    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
 \end{align}
@@
     &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
-    &\propto \exp(n_0 \m_0^\top \nat - n_0 \logpart(\nat))
+    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat))
 \end{align}
@@
 \section{Open Problem}
-
-Let $\nat^* := \argmin \expect[X\sim\cD]{-\log p(X|\nat)}$, where $\cD$ is any distribution such that $\E[T] = \meanp^* \in \cM$.
-Then the suboptimality on the population log-likelihood is exactly the KL between our current model and the true distribution
-\begin{multline}
-    \expect[X\sim \cD]{-\log p(X|\nat) + \log p(X|\nat^*) } \\
-	= \KL( p(.|\nat^*) ; p(.|\nat)) \; .
-\end{multline}
+Given a distribution $\cD$, our objective is the expected log-likelihood 
+\begin{align}
+	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
+	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
+\end{align}
+Up to a constant, it is also the KL  $f(\nat) = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$.
 Note that $\cD$ does not have to be part of the exponential family.
-For the exponential family, the KL is equal to the Bregman divergence induced by the log-partition function, or alternatively by the entropy, being careful with the order of the arguments
+Let $\nat^* := \argmin f(\nat)$.
+Assuming $\nat^*$ exists within $\Theta$ means $\cD$ is such that $\E[T] = \meanp^* \in \cM$.
+Then the suboptimality on the population log-likelihood can be written with Bregman divergences
+\footnote{
+Since for exponential families, $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, this proves on the way the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
+}
 \begin{align}
-\boxed{
-	\KL( p(.|\nat^*) ; p(.|\nat))
-    = \bregman (\nat ; \nat^*)
-    = \bregmanconj ( \meanp^* ; \meanp)
-}
+	 f(\nat) - f(\nat^*)
+	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
+	 = \bregman(\nat ; \nat^*) 
+	 = \bregmanconj(\m^* ; \m) \; .
 \end{align}
-
-Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \TODO{cite}
-Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \TODO{cite}
-A contrario the KL is not smooth, and it can be infinite in misspecified settings.
-But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.
-
-Our question is: how does the KL behave when $\nat$ is the maximum-likelihood or the MAP estimate ? Can we get upper-bounds on the following quantities ?
+Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
 \begin{align}
@@
 where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
-More explicitly, can we write an upper bound that does not involve an expectation over the dataset -- e.g. where the stochasticity is removed.
+More explicitly, can we write an upper bound that does not involve this expectation over the dataset.
 
-\paragraph{Remark.}
+In general, for arbitrary distributions $\cD$, it is a hard problem, especially since the KL is known for being brittle to misspecification \tocite, so let us focus on the well-specified case first $\cD = p(\cdot \cond \nat^*)$.
+Surely this problem has been solved, hasn't it ?
+In some special cases yes, but it seems the general answer is no. 
+We are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n$.
+
+{\bf Remark.}
 This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
@@
 
-\paragraph{Remark 2.}
+{\bf Remark 2.}
 The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$.
@@
 -- e.g. exponential distribution,
-Gaussian\fdk{with known variance?} and Laplace distributions with known mean,
-Pareto distribution with known minimum value,
-Weibull distribution with known shape $k$
+Gaussian with known mean,
+Laplace with known mean,
+Pareto with known minimum value,
+Weibull with known shape $k$
 -- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
@@
 In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
-\RLP{Tune figures.}
+\rlp{Tune figures.}
 
@@
 	\item $\nu$ for the base measure
-	\item $A, A^*$ for the log-partition and its dual
+	\item $A, A^*, \logpart, \conj$ for the log-partition and its dual
 	\item $\cB_A$ for the Bregman divergence induced by $A$
@@
 	\item $(\theta^*,m^*)$ for the optimum
-	\item $\gamma$ for the step-size
+	\item $\gamma, \lr$ for the step-size
 	\item $\mu,\sigma^2$ for the parameters of a Gaussian

### 2021-10-11 aistats.tex
Clarify that MLE is not optim

@@
 \end{align}
-Up to a constant, it is also the KL  $f(\nat) = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$.
 Note that $\cD$ does not have to be part of the exponential family.
@@
 \footnote{
-Since for exponential families, $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, this proves on the way the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
+Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
+and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
 }
@@
 \label{ssec:MAP=SMD}
-The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
+By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
+Plugging in~\eqref{eq:joint_likelihood}, we see that the MAP verifies
 \begin{align*}
@@
 Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
-This parallel means it would be possible to obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
-We give a special focus on this approach in Section~\ref{sec:optimization}.
+This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
+In particular, it could potentially apply to a wide range of $\cD$.
+Note that this optimization perspective does not hold for the MLE %($n_0=0$)
+as the initial prior uniform prior may correspond to an initialization outside of $\Theta$.
+We focus more on this approach in Section~\ref{sec:optimization}.
 

### 2021-10-11 aistats.tex
Update Optim section
Move MAP=SMD there for simplicity.
Compress the story by immediately introducing SMD, and skipping MD.

@@
 \usepackage[capitalise]{cleveref}
-\newcommand{\rlp}[1]{\textcolor{red}{RLP:#1}}
-\newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
-\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}
+\newcommand{\rlp}[1]{\textcolor{BrickRed}{(RLP:#1)}}
+\newcommand{\fdk}[1]{\textcolor{Periwinkle}{(fdk:#1)}}
+\newcommand{\TODO}[1]{\textcolor{cyan}{(TODO #1)}}
 \newcommand{\tocite}{\textcolor{purple}{(add citation)}}
@@
 
+\paragraph{Notation}
+$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
+
 \section{Technical background}
@@
 \begin{equation}
-	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
+	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; .
+	 \label{eq:def_expfamily}
 \end{equation}
-where $\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
 Here $\logpart$ is the log-partition function -- i.e. the normalization term
@@
     &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat))
+    \label{eq:def_prior}
 \end{align}
@@
 \paragraph{Maximum A Posteriori (MAP).}
-Using Bayes rule, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
+Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
 \begin{align}
@@
 	%\propto p(\mD|\nat)p(\nat)
-    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt))
+    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
     \label{eq:joint_likelihood}
 \end{align}
-reaching its maximum in $\MAPt$, the MAP estimate verifying the first order optimality conditions
+reaching its maximum in $\MAPt^\text{MAP}$ such that
 \begin{align}
-    \nabla \logpart(\MAPt) = \MAPm
+    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
@@
 This last equation is known as moment matching.
+In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
 The MLE and MAP estimates are statistics of the dataset $\mD_n$.
@@
 \section{Insights}
-
-\subsection{MAP as Stochastic Mirror Descent}
-\label{ssec:MAP=SMD}
-By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
-Plugging in~\eqref{eq:joint_likelihood}, we see that the MAP verifies
-\begin{align*}
-	\hat \nat_{n}
-	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
-    %\label{eq:SBPP}
-\end{align*}
-which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
-\begin{align}
-		\hat \nat_{n}
-	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
-    \label{eq:SMD}
-\end{align}
-where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$.
-We recognize the update of Stochastic Mirror Descent (a.k.a.
-Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
-This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
-In particular, it could potentially apply to a wide range of $\cD$.
-Note that this optimization perspective does not hold for the MLE %($n_0=0$)
-as the initial prior uniform prior may correspond to an initialization outside of $\Theta$.
-We focus more on this approach in Section~\ref{sec:optimization}.
 
@@
 \label{sec:optimization}
-As we saw in Section~\ref{ssec:MAP=SMD}, MAP can be interpreted as Stochastic Mirror Descent (SMD).
-There are two consequences of this observation: 1.
-we may obtain a convergence rate for MAP from an optimization analysis, and 2.
-any insights gained from the MAP may inform further design and analysis of SMD.
 
-Mirror Descent (MD)\footnote{%
-Also known as is going by many names:
-Bregman (proximal) Gradient, Relative Gradient Descent, NoLips.
+We now show how MAP can be interpreted as stochastic mirror descent (SMD).
+This means that 1. we may obtain a convergence rate for MAP from an optimization analysis, and 2. any insights gained from the MAP may inform further design and analysis of SMD.
+We then review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
+\rlp{mention that examples have so far been lacking for relative SMD.}
+
+\subsection{MAP as Stochastic Mirror Descent}
+\label{ssec:MAP=SMD}
+By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
+Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
+\begin{align*}
+	\hat \nat_{n}
+	= \argmin_\nat f_n(\nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
+    %\label{eq:SBPP}
+\end{align*}
+which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$.
+%Maybe skip this SBPP story altogether
+Expanding $f_n(\nat) = \logpart(\nat) - T_n$ and  playing with the Bregman we get
+\begin{align*}
+		\hat \nat_{n}
+	= \argmin_\nat \langle g_{n}(\hat \nat_{n-1}), \nat \rangle + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
+    %\label{eq:SMD}
+\end{align*}
+where $g_{n}(\nat) := \nabla f_n(\nat) = \hat \m_{n-1} - T_n$.
+We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$.
+From the first order optimality condition, we see this is a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
+This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
+In particular, it could potentially apply to a wide range of distribution $\cD$ on $X$.
+Note that this optimization perspective does not hold for the MLE %($n_0=0$)
+as its implicit uniform prior may correspond to an initialization outside of $\Theta$.
+
+\subsection{Relative Smoothness}
+Mirror descent (MD)\footnote{
+Also known as
+Bregman (proximal) gradient, relative gradient descent or NoLips.
 }
-\citep{nemirovski1983problem}
-was first introduced to optimize non-smooth functions \fdk{(Is typically encountered in non-smooth, online optimization?)},
-typically under bounded gradient assumption on the objective $f$
-or strong convexity assumption on the potential $\logpart$
+\citep{nemirovski1983problem,beck2003mirror}
+%Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
+and SMD 
+\citet{nemirovski2009robust,ghadimi2012optimal}  \TODO{verify what these citations are talking about} 
+are typically encountered in non-smooth (online) optimization,
+under bounded (or Lipschitz) gradient assumption on the objective $f$
+and strong convexity assumption on the potential $\logpart$
+\citep[Th.4.2 \& Th.6.3]{bubeck2015convex}.
+In our case, none of these assumptions may hold. 
+For instance $\logpart = -\log$ is neither smooth nor strongly convex.
 
@@
 \citet{azoury2001relative,freund1996predicting,dasgupta2007online},
-Bubeck's paragraph
+Bubeck's 2011.
 }
 
-More recently, advances in optimization that go beyond the Euclidean case,
-to optimize a function $f$ when it is not smooth, meaning its gradient is not $L$-Lipschitz,
-neither strongly convex in the Euclidean norm, using the concept of relative smoothness
-\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
-Instead of measuring strong-convexity and smoothness in the Euclidean norm (or any norm),
-the $\mu$-strong convexity and $L$-smoothness of a function $f$
-relative to a reference function $A$ is defined  as
+Recently, these assumptions have been relaxed to allow for non-smooth functions $f$
+\footnote{$f$ is $L$-smooth if its gradient is $L$-Lipschitz}.
+and non-strongly convex potential $\logpart$, first in the deterministic setting
+\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
+These assumptions have been replaced with
+the $\mu$-strong convexity and $L$-smoothness of $f$
+\emph{relative} to $\logpart$, defined as
 \aligns{
@@
 	\leq
-	L \cB_A(x,y).
+	L \cB_A(x,y) \; .
 }
+When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent. 
 Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
-Exponential families log-likelihood perfectly fits into this theory.
+Exponential families log-likelihood perfectly fit into this framework, as
 \aligns{
-	f(\theta) = A(\theta) - \expect[x\sim p(x\cond\theta)]{\lin{x, \theta}}
+	f(\theta) = A(\theta) - \expect{\lin{X, \theta}}
 }
 is $1$-smooth and $1$-strongly convex relative to $A$.
-However, the MAP is stochastic, and known results for stochastic mirror descent do not give convergence rates
-for all exponential families.
 
-\subsection{Strongly-convex reference function}
-
-{Traditional results}
-in stochastic optimization,
-such as the work of \citet{nemirovski2009robust,ghadimi2012optimal},
-rely on the assumption that the reference function $A$
-is itself strongly convex relative to a norm $\norm{\cdot}$,
-(meaning that $f$ is strongly convex).
-This apply for the MLE problem of a Gaussian distribution
-with fixed variance,
-but not in a general case.
-when the covariance is unknown,
-or for a Poisson distribution, as $-\log(\theta)$ and $\exp(\theta)$
-are not strongly convex.
-
-The introduction of relative smoothness has renewed interest for analyses of mirror descent
-without relying on strong-convexity in an underlying norm.
-Recent works have made attempts to generalize the assumptions
-that are used to control the convergence of stochastic gradient descent
-to this setting.
-Those include
-bounded gradients $(\expecti{\norm{\tilde g(\theta)}^2} \leq C$,
-bounded variance $(\expecti{\norm{\nabla f(\theta) - \tilde g(\theta)}^2} \leq C$
-uniformly or only at the minimum $\theta_*$,
-or bounded disparities between the minimum of $f$ at $\theta_*$ and the minima of individual functions $f_i$,
-if $f(\theta) = \expect[i]{f_i(\theta)}$ $(\expect{f(\theta_*) - \min_\theta f_i(\theta)} < C)$.
-(\fdk{For a review of those assumptions, see that paper by Richtarik's students})
-
-However, they do not yet address the problem satisfactorily.
-We present those generalization here
-and they are summarized in the context of maximum likelihood estimation
-in \cref{tbl:assumptions}.
+\subsection{Bounding the Randomness}
+Smoothness is not the only issue.
+To analyze stochastic algorithms, one also need to quantify the randomness. 
+We now review these modern attempts at analyzing SMD, and illustrate their limits  in the context of the MAP.
+See \cref{tbl:assumptions} for a summary.
 
@@
 
-\subsection{Bounded variance}
-An adaptation of the bounded variance,
-investigated by \citet{hanzely2018fastest},
-defines the following bounded quantity.
-Letting $\theta^+(v)$ be a step from $\theta$ with vector $v$,
-$\theta^+(v) = \nabla A^*(\nabla A(\theta) - v)$,
-\alignn{\label{eq:assumption-hanzely}
-	\expect[\tilde g\!]{
-	\lin{
-		\nabla\! f(\theta)
-		- \tilde g(\theta),
-		\theta^+\!\!\paren{\eta \tilde g(\theta)}
-		- \theta^+\!\!\paren{\eta \nabla\! f(\theta)}
+\subsubsection{Bounded variance}
+\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step
+\alignn{
+\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C, \forall n \geq 1
+\label{eq:assumption-hanzely}
 }
-	}
-	\leq
-	\eta C,
-}
-for some step-size $\eta$. Let us unpack this quantity;
-
-{\bf For a Gaussian} with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
+for some constant $C$.
+For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
 this definition recovers the variance of the stochastic gradient
@@
 
-{\bf In divergence,} this condition states that the (symmetrized) divergence between
-a deterministic and stochastic update is bounded in expectation,
-\alignn{\label{eq:hanzely-in-divergence}
-	\begin{aligned}
-	&
-	\mathbb{E}_{\tilde g}\bigg[
-	\cB_{A^*}\paren{\m - \eta \tilde g(\theta), \m - \eta \nabla f(\theta)}
-	\\[-1em]
-	&\hphantom{\mathbb{E}_{\tilde g}[}
-	+ \cB_{A^*}\paren{\m - \eta \nabla f(\theta), \m - \eta \tilde g(\theta)}
-	\bigg]
-	\leq
-	\eta C.
-	\end{aligned}
+Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest} , which is precisely what we wish to obtain.
+
+In general however, this assumption hardly holds.
+For the MAP, this translates into
+\alignn{
+\expect[T_n]{\lin{T_n - \mu^* , \MAPt}} \leq \lr_n C, \forall n \geq 1
 }
-{\bf For a maximum likelihood problem, }
-where $\tilde g(\theta) = \m - x$
-where $x \sim p(x\cond\theta_*)$,
-and $\nabla f(\theta) = \m - \m_*$,
-\aligns{
-	&
-	\mathbb{E}_{x}\bigg[
-	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}
-	\\
-	&\hphantom{\mathbb{E}_{\tilde g_t}[}
-	\cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
-	\bigg]
-	\leq
-	\eta C.
-}
-\fdk{Could work with decreasing step-sizes}
+which fails to hold on the gaussian variance example for instance. \TODO{short proof or formalize.}
 
-\fdk{How do we show that this fails?}
-
-\subsection{Bounded variance at the minimum}
+\subsubsection{Bounded variance at the minimum}
 
@@
 
-\textbf{Bounded optimality gap,}
+\subsubsection{Bounded optimality gap,}
 \fdk{

### 2021-10-12 aistats.tex
Reference a review of SGD assumptions
Write about Dragomir's assumption

@@
 
-%\hypersetup{draft}
+\hypersetup{draft}
 
@@
 To analyze stochastic algorithms, one also need to quantify the randomness.
-We now review these modern attempts at analyzing SMD, and illustrate their limits  in the context of the MAP.
+Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
+We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
 See \cref{tbl:assumptions} for a summary.
+
+Caveat : all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.
 
@@
 	\leq
-	\eta C
+	\eta^2 C
 	$
@@
 
-\subsubsection{Bounded variance}
-\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step
+\subsubsection{Bounded Variance}
+\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step: $\forall n \geq 1, \forall \hat \nat_{n-1},$
 \alignn{
-\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C, \forall n \geq 1
+\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
 \label{eq:assumption-hanzely}
@@
 In general however, this assumption hardly holds.
-For the MAP, this translates into
+For the MAP, this translates into $\forall n \geq 1,$
 \alignn{
-\expect[T_n]{\lin{T_n - \mu^* , \MAPt}} \leq \lr_n C, \forall n \geq 1
+\expect[T_n]{\lin{T_n - \mu^* , \MAPt}} \leq \lr_n C
 }
@@
 
-\subsubsection{Bounded variance at the minimum}
+\subsubsection{Bounded Variance at the Minimum}
 
-\citep{dragomir2021fast} is similar to \cref{eq:hanzely-in-divergence},
-where we take only the first part of it
-and care only about gradients at the minimum, replacing
-$\tilde g(\theta)$ and  $\nabla f(\theta)$ with $\tilde g(\theta_*)$ and $\nabla f(\theta_*) = 0$
-\aligns{
+\citet{dragomir2021fast} introduce a much weaker assumption $\forall n \geq 1, \forall \MAPm,$
+\alignn{
 	\expect[\tilde g]{
-		\cB_{A^*}(\m - 2\eta \tilde g(\theta_*), \m)
-	} \leq 2 \eta^2 C.
+		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
+	} \leq 2 \lr^2 C \; .
+	\label{eq:assumption-dragomir}
 }
-{\bf For a maximum likelihood problem, }
-\aligns{
-	\expect[x]{
-		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
-	} \leq 2 \eta^2 C.
-}
-\fdk{How do we show that this fails?}
+In words, this is bounding the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
+% As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality. 
+Under this assumption, they prove convergence up to a variance ball with a constant step-size.
+Their descent lemma \citep[Eq.(12)]{dragomir2021fast} is the exact analog  of a modern analysis of SGD \citep[Th.3.2]{gower2019sgd}, meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.
 
-\subsubsection{Bounded optimality gap,}
+Seems like a perfect fit then ?
+Well not really, this variance assumption does not hold for exactly the same reason as the one of Hanzely.
+Having something valid for every possible iterates is way too strong.
+
+\subsubsection{Bounded Optimality Gap}
 \fdk{
@@
 \fdk{Existing techniques only deal with fixed step-sizes and show convergence to a ball}
+
+If assumptions~\eqref{eq:assumption-hanzely} and~\eqref{eq:assumption-dragomir} do not hold for every iterate, then let's assume they hold on average.
+Does the proof remain valid ?
+Yes I think so \TODO{check this}.
+Does it become impossible to prove / as hard as the initial problem ?
+Yes.
+\TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
 

### 2021-10-12 aistats.tex
Discuss D'Orazio's paper.

@@
 \midrule
-Strongly convex and bounded variance\newline
-\citep[e.g.,][]{??}
-&
-$
-\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
-\quad
-\expect[x]{\norm{\m - x}}^2\leq\sigma^2
-$
-\\
+%Strongly convex and bounded variance\newline
+%\citep[e.g.,][]{??}
+%&
+%$
+%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
+%\quad
+%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
+%$
+%\\
 Uniformly bounded ``variance''\newline
@@
 
-Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest} , which is precisely what we wish to obtain.
+Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
 
@@
 \subsubsection{Bounded Optimality Gap}
-\fdk{
-Adaptation of the $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$ hypothesis \citep{dorazio2021stochastic}?
-\\
-Do they do something with that? It seems to be fixed step-size only.
+Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explores the hypothesis
+\aligns{
+\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C 
 }
-\aligns{
-	A(\theta_*) - \lin{x, \theta_*} - \expecti{ \min_\theta A(\theta) - \lin{x, \theta} }
-}
+where $f_X$ is a stochastic estimate of $f = \expecti{f_X}$. In our case $f_X(\nat) = - \log p(X\cond \nat)$.
+In words, this is lower bounding the expectation of the minimum of these stochastic estimates.
+For probabilistic models, such a bound is finite as soon as the model cannot give infinite density to any data point $x$.
+This holds for instance for discrete distributions 
+-- $f_x\geq 0$ because the probability mass is upper bounded by $1$ -- 
+but it rules out  many families.
+For normal distributions set $\mu=x$ and $\sigma^2 \rightarrow 0$ to get $p(x\cond \nat) \rightarrow +\infty$. 
+For gamma, set $\alpha = \beta x$ and $beta \rightarrow +\infty$.
+For beta, set $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
+ Other examples include inverse Gaussians, log-normal, gamma, inverse gamma.
+
+ It is possible to overcome this limitation though, by treating batches of samples as 1 sample, and averaging sufficient statistics -- e.g. $Y = \{X_1, \dots, X_k\}$ and $T(Y) = \inv{k}\sum_i T(X_i)$.
+ For instance a multivariate normals of dimension $d$ cannot attribute infinite density to $d+1$ samples that are not in an affine subspace.
+
+So it seems this approach can yield interesting convergence rates for the MAP, and they do prove linear convergence of the last iterate up a to a variance ball with constant step-size, but the size of this ball $\frac{C}{\alpha}$ \rlp{strong convexity notation} is independent of  the step-size.
+Their descent lemma does not lead to convergence with a decreasing $O(1/t)$ step-size.
 

### 2021-10-12 aistats.tex
Format optimization summary table

@@
 \usepackage[dvipsnames]{xcolor}
+
+\usepackage{array}
 
@@
 
-\begin{table*}[t]\begingroup
+\begin{table}[t]
+\begingroup
 \renewcommand*{\arraystretch}{1.25}%
+\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
+\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
 \centering
-\begin{tabular}{p{.35\textwidth}p{.6\textwidth}}
+\begin{tabular}{lccc}
+%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
 \toprule
-Assumption &  \\
+Boundedness & Barrier &  $\lr_n \sim \inv{n}$ & Last iterate \\
 \midrule
@@
 %\\
-Uniformly bounded ``variance''\newline
-\citep{hanzely2018fastest}
-&
-	$
-	\expect[x]{
-	\begin{aligned}
-	\cB_{A^*}\paren{(1-\eta) \m + \eta x, (1-\eta)\m + \eta \m_*}\\
-	+\, \cB_{A^*}\paren{(1-\eta)\m + \eta \m_*, (1-\eta) \m + \eta x}
-	\end{aligned}
-	}
-	\leq
-	\eta^2 C
-	$
+Variance on all $\Theta$ % \newline \citep{hanzely2018fastest}
+& \redxmark & \greencmark  & \redxmark
 \\
-Bounded ``variance'' at $\theta_*$\newline
-\citep{dragomir2021fast}
-&
-	$\expect[x]{
-		\cB_{A^*}(\m - 2\eta (\m_* - x), \m)
-	} \leq 2 \eta^2 C$
+Variance at $\theta_*$ %\newline \citep{dragomir2021fast}
+& \redxmark & \greencmark  & \greencmark
 \\
-Bounded optimality gap\newline
-\citep{dorazio2021stochastic}
-&
-?
+Optimality gap %\newline \citep{dorazio2021stochastic}
+& \greencmark & \redxmark & \greencmark
 \\
@@
 \end{tabular}
-\caption{Caption}
+\caption{
+Summary of results about SMD with relative smoothness and relative strong convexity assumptions. 
+Barrier tells whether the assumption can hold with barrier objectives.
+$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
+}
 \label{tbl:assumptions}
-\endgroup\end{table*}
+\endgroup
+\end{table}
 

### 2021-10-12 aistats.tex
Start formatting figures

@@
 %No rates hold either when the loss is ill-behaved : infinite domain, not smooth, not self-concordant
+
+\begin{figure}[t]
+	\centering
+\includegraphics[width=.4\textwidth]{fewsamples.pdf}
+	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$.
+Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval.
+		While the expected value is infinite for $n=1$ or $n=2$, it quickly matches the upper bound, and the $1/2n$ asymptote.
+	}
+	\label{fig:curves}
+\end{figure}
 
@@
 
-\begin{figure*}[ht]
+\begin{figure}[t]
 	\centering
-	\includegraphics[width=\textwidth]{{figs/biasvariance/new_linear_ratio=0.8_n0=10}.pdf}
+	\includegraphics[width=.4\textwidth]{{figs/gaussian_variance.pdf}}
 	\caption{
@@
 	\label{fig:variance_decomposition}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[ht]
+\begin{figure}[t]
 	\centering
-	\includegraphics[width=\textwidth]{{figs/gaussians/new_linear_n0=1}.pdf}
+	\includegraphics[width=.4\textwidth]{{figs/gaussians/new_linear_n0=1}.pdf}
 	\caption{
@@
 	\label{fig:gaussian_decomposition}
-\end{figure*}
+\end{figure}
 
@@
 	\label{fig:phi}
-\end{figure}
-
-\begin{figure}[ht]
-	\centering
-\includegraphics[width=.4\textwidth]{fewsamples.pdf}
-	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$.
-Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval.
-		While the expected value is infinite for $n=1$ or $n=2$, it quickly matches the upper bound, and the $1/2n$ asymptote.
-	}
-	\label{fig:curves}
 \end{figure}
@@
 For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
+This correlates well with numerical observations.
 Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$.
@@
 \end{figure*}
-
-\newpage~\null\\
-\FloatBarrier
 
@@
 and SMD 
-\citet{nemirovski2009robust,ghadimi2012optimal}  \TODO{verify what these citations are talking about} 
+\citep{nemirovski2009robust,ghadimi2012optimal} 
 are typically encountered in non-smooth (online) optimization,
@@
 \toprule
-Boundedness & Barrier &  $\lr_n \sim \inv{n}$ & Last iterate \\
+Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
 \midrule

### 2021-10-12 aistats.tex
Add a conclusion

@@
 %\\
-Variance on all $\Theta$ % \newline \citep{hanzely2018fastest}
+Variance $\forall \nat\in\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
 & \redxmark & \greencmark  & \redxmark
 \\
-Variance at $\theta_*$ %\newline \citep{dragomir2021fast}
+Variance at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
 & \redxmark & \greencmark  & \greencmark
 \\
-Optimality gap %\newline \citep{dorazio2021stochastic}
+Optimality gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
 & \greencmark & \redxmark & \greencmark
@@
 \Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
-\label{eq:assumption-hanzely}
+\label{eq:hanzely}
 }
@@
 More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
-the LHS of \cref{eq:assumption-hanzely} is bounded by
+the LHS of \cref{eq:hanzely} is bounded by
 $\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
-and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:assumption-hanzely}
+and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:hanzely}
 to hold.
-
+Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
+\aligns{
+\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) = \E[\cS(\MAPm - \lr g_n  )]
+}
+thus showing positivity.
 Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
@@
 	} \leq 2 \lr^2 C \; .
-	\label{eq:assumption-dragomir}
+	\label{eq:dragomir}
 }
@@
 Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explores the hypothesis
-\aligns{
+\alignn{
 \min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C 
+\label{eq:dorazio}
 }
@@
 
-If assumptions~\eqref{eq:assumption-hanzely} and~\eqref{eq:assumption-dragomir} do not hold for every iterate, then let's assume they hold on average.
+If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let's assume they hold on average.
 Does the proof remain valid ?
@@
 \TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
+
+\section{Conclusion}
+Exponential family MLE and MAP are used everywhere -- everytime someone computes a mean and variance to fit some bell-shaped data. 
+MLE and MAP are both maximizing the (regularized) likelihood of the data, whose suboptimality is the Kullback-Leibler (KL) divergence.
+The KL is a very natural measure of discrepancy between a model and the true disribution.
+Yet we do not have convergence rates on the expected KL, even in the well-specified setting.
+Such a rate could inform us on how many samples we may need, or on the influence of the prior.
+There are many results close from such a rate, in particular large sample results \citep{kakade2010learning, ostrovskii2021finite}, which describe how many samples are needed to ensure a locally quadratic regime.
+Another approach to obtain rates is stochastic optimization \citep{bach2013nonstronglyconvex}. 
+Here we observed that MAP perfectly fits the framework of stochastic mirror descent with relative smoothness assumptions.
+Yet none of the current analysis of SMD hold for the MAP, even on a simple family such as $\cN(0,\sigma^2)$, thus revealing area for progress in non-Euclidean optimization.
+In writing this paper, we hope to attract attention to this fundamental problem, leading to progress in both optimization and statistics.
 

### 2021-10-13 aistats.tex
Update aistats.tex
Half pass on the paper - mostly rephrased things and added a few comments (see: D.S. or textcolor red)

@@
 say for the problem
-\begin{equation}
+\begin{equation} %We might want to add a feasible set for theta?
 \min_\theta f(\theta)
@@
 Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
+
+Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
+\[
+R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
+\]
+where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
+Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.
+Unfortunately however, their rate are worse than the offline MAP or MLE algorithms.
+For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$.
+From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
+Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
+This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
 
@@
 
-Exponential Families are to statistics what linear models are to regression.
-Many classical model are exponential families, and their convexity and modeling flexibility make them one of the most widely used statistical models.
-Here we review some of their properties.
+Exponential families are to statistics what linear models are to regression.
+Many classical model are exponential families, for instance the Gaussian or Bernoulli distributions.
+Their convexity or modeling flexibility make them one of the most widely used statistical models.
+In this section, we review some of their properties.
 We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
-The exponential family for data $X\in \cX$ with natural (or primal) parameter $\nat$  is specified by a sufficient statistic $T: \cX \rightarrow \real^d$ and a density with respect to a base measure $\nu(dx)$
+
+The exponential family for data $X\in \cX$ reads
 \begin{equation}
-	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; .
+	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
 	 \label{eq:def_expfamily}
 \end{equation}
-Here $\logpart$ is the log-partition function -- i.e. the normalization term
+where  $\nat$ is called natural (or primal) parameter is specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) $\nu(dx)$, the density with respect to a base measure.
+The function $\logpart$ is called the log-partition function that acts as a normalization term, since
 \begin{align}
@@
 \end{align}
-When explicit, we will write the random variable $T = T(X)$ for lightness.
-For convenience, we focus on steep, regular exponential families
+
+This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
+
+For convenience, in this paper we focus on steep, regular exponential families
 where $\logpart$ is strictly convex function of Legendre type,
 and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
-(See \citet{barndoffnielsen2014information}).
-This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
+(See \citet{barndoffnielsen2014information}). Also, when explicit, we write the random variable $T = T(X)$.
 
@@
 \paragraph{Duality}
-The logpartition function $\logpart$ verifies the two following identities
+The logpartition function $\logpart$ verifies the two following identities:
 \begin{align}
@@
 \end{align}
-where $\meanp$ is called the mean (or dual) parameter.
-From now on, we assume that the sufficient statistic $T$ is minimal
--- e.g. its covariance is definite positive for all $\nat$ --
-then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\nat$ and mean parameters $\m$.
-$\meanp$ lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
+where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
 We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both represent the same distribution.
-At this point it is useful to introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
+
+We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
 \begin{align}
-	\conj(\m) = \langle \m, \nat \rangle - \logpart(\nat) \; .
+	\conj(\m) = \min_{\nat\in\Theta}\langle \m, \nat \rangle - \logpart(\nat) \; ,
 \end{align}
 which matches the common notion of \textit{entropy} in information theory.
-The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).
+In this paper, we assume that the sufficient statistic $T$ is minimal, e.g., its covariance is definite positive for all $\nat$.
+In such case, the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$, and moreover, $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}).
+More concisely,
+\[
+	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
+\]
 
-\begin{figure}[ht]
+% The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).
+
+\begin{figure}[t] % @remi In the figure you use mu instead of m??
 	\centering
@@
 
-\paragraph{Bregman Divergences}
-A very useful object to keep in mind is the Bregman divergence induced by $\logpart$ between two parameters $\nat$ and $\nat_0$
+\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps)
+We now introduce the Bregman divergence induced by $\logpart$, that measure the distance between two parameters $\nat$ and $\nat_0$,
 \begin{align}
@@
     & = \logpart(\nat) - \logpart(\nat_0)
-    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle
+    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
 \end{align}
 with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
-It is a non-symmetric measure of distance between parameters.
-It is equal to the divergence of its convex conjugate with switched arguments
+In general, bregman distances are non-symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
+Finally, it is equal to the divergence of its convex conjugate with switched arguments,
 \begin{align}
@@
     &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
-    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat))
+    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
     \label{eq:def_prior}
@@
 This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
-Intuitively, $n_0$ is a number of fictive data points observed from a distribution with parameter $\nat_0$.
+Intuitively, $n_0$ is the number of fictive data points observed from a distribution with parameter mean $\nat_0$.
 
 \paragraph{Maximum A Posteriori (MAP).} 
+\textcolor{red}{\textbf{D.S.: This section is very unclear. }}
 Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
@@
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-Asymptotic results typically show that natural parameters $\nat$ asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
+Asymptotic results typically show that natural parameters $\nat$ is asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
 Here the MAP is more simply expressed with mean parameters $\meanp$.
@@
     \bregmanconj(\m^* ; \m)
-    &=\half  \norm{\m^* - \m}^2_{\nabla^2\conj(\m^*)}
-    + O(\norm{\m - \m^*}^3)
+    &= \frac{\norm{\m^* - \m}^2_{\mF^*}}{2}
+    + O(\norm{\m - \m^*}^3),
 \end{align}
-where the Mahalanobis norm  $\| x \|_\mM^2 = x^\top \mM x$  is induced by the hessian of the entropy -- the inverse \textit{Fisher information matrix} -- at the optimum,
+where the Mahalanobis norm  $\| x \|_{\mF^*}^2 = x^\top \mF^* x$  is induced by $\mF^*  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
+It happens that  $F^*$ is also the inverse \textit{Fisher information matrix}, since
 \begin{align}
@@
 \end{align}
-From there, it is possible to show an asymptotic rate for the MLE and the MAP
+From there, it is possible to show an asymptotic rate for the MLE and the MAP \textcolor{red}{D.S.: How?}
 \begin{align}
@@
 As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
-For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean ($\cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).
-Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances
+For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean 
+\[ \cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).\]
+Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
 \begin{align}
@@
 \end{align}
-Thanks to the independence of samples we can break down the MLE into individual point's contributions
+Thanks to the independence of samples, we can break down the MLE into individual point's contributions
 \begin{align}
-	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
+	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}.
 	=\frac{\Var(T)}{2n}
 \end{align}
-Adding a reference mean $\m_0$ to get the MAP yields
+Adding a reference mean $\m_0$ to get the MAP yields to
 \begin{align}
 	\expect{\bregmanconj(\meanp^*; \MAPm)}
-	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}
+	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}.
 	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
@@
 \end{align}
-so we have a variance term in $O(n^{-1})$ and a bias term decreasing as $O(n^{-2})$. We would like to get a similar result for arbitrary exponential families.
+In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
 If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.
@@
 \subsection{Locally Quadratic Case}
-All Bregman divergences are locally quadratic.
-Under some assumptions, we can quantify  when this quadratic behavior kicks in.
-For instance {\bf when $\conj$ is self-concordant} \citep[Ch.4.1]{nesterov2003introductory}.
-This is true when $T$ is 1-dimensional and $\logpart$ is self-concordant
--- e.g. exponential distribution,
+\textcolor{red}{Check this! In our case, since we assume $A$ and $A^*$ to be strictly convex,}
+all Bregman divergences are locally quadratic.
+Under some assumptions, like self-concordance \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
+For instance, when $T$ is 1-dimensional, we have that $\logpart$ in the following settings:
+exponential distribution,
 Gaussian with known mean,
@@
 Pareto with known minimum value,
-Weibull with known shape $k$
+Weibull with known shape $k$. This also holds when
 -- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
@@
 
-\subsection{Bias-Variance Decomposition}
+\subsection{Bias-Variance Decomposition} % DS: Arrived here
 In both examples (add link), the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.

### 2021-10-13 aistats.tex
Remove excess $ sign
Update abstract and create related work section

@@
 \begin{abstract}
-We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family.
-Surprisingly, we found no general solution to this problem in the literature.
-In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
-After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD),
-yet it falls out of scope for modern analysis.
+We consider the problem of upper bounding the expected log-likelihood sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family, in a non-asymptotic way.
+Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime. 
+After exhibiting various facets of the problem, we show the MAP can be interpreted as running stochastic mirror descent (SMD) on the log-likelihood, 
+yet modern convergence results do not apply for standard examples of exponential family, highlighting holes in the convergence literature. 
 We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
@@
 
-Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
-Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
-A contrario the KL is not smooth, and it can be infinite in misspecified settings.
-But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.
-
 There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
@@
 Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
-
-Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
-\[
-R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
-\]
-where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
-Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.
-Unfortunately however, their rate are worse than the offline MAP or MLE algorithms.
-For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$.
-From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
-Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
-This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
 
@@
 
+\subsection{Related Work}
+
+\paragraph{Density Estimation}
+Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
+Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
+A contrario the KL is not smooth, and it can be infinite in misspecified settings.
+But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.
+
+\paragraph{Online Learning}
+Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
+\[
+R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
+\]
+where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
+Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.
+Unfortunately however, their rate are worse than the offline MAP or MLE algorithms.
+For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$.
+From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
+Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
+This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
+
 % Then;
@@
 
-\paragraph{Notation}
+\subsection{Notation}
+$X, T(X)$ and $T$ are random variables, while $x$ is a sample. 
+$n$ is the number of samples and $d= \dim(T)$. 
 $\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
@@
 For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean 
-\[ \cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).\]
+\[ \cX=\real,\quad \nu(dx) = \exp(\half[-x^2])dx,\quad T(X)=X.\]
 Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
@@
 	=\frac{\Var(T)}{2n}
+	=\frac{1}{2n}
 \end{align}
@@
 
-\TODO{Reference \citet{ostrovskii2021finite}.} ( as well as
-\citet{anastasiou2017bounds},
+All these works give \textit{large sample} results -- results that hold only for $n\geq N$ for some constant $N$ -- but none of them  apply to small $n$.
+
+\TODO{Incorporate this content} ( about \citet{ostrovskii2021finite} as well as
+\citet{anastasiou2017bounds} and
 \citet{marteauferey2019beyond})
-
-All these works give \textit{large sample} results -- results that hold only for $n\geq N$ for some constant $N$ -- but none of them  apply to small $n$.
+\fdk{
+We know that if the problem is strongly convex (in $\norm{\cdot}_2$), averaged SGD achieves $O(1/n)$ rate 
+%[Robust stochastic approximation approach to stochastic programming].
+(The paper uses a bounded stochastic gradient assumption, but can be relaxed to bounded variance instead.) However, strong-convexity in the Euclidean norm is not strictly necessary to find an algorithm with convergence rate of $O(1/n)$, as shown by [Non-strongly-convex smooth stochastic approximation with convergence rate O(1/n)] for linear and logistic regression.
+Their proof techniques based on self-concordance have been extended to other setting, [Finite-sample analysis of M-estimators using self-concordance, Learning Exponential Families in High-Dimensions: Strong Convexity and Sparsity].
+%The same tools are used in statistics, and evething essentially boils down to controlling how different we are from a Gaussian rv, that we're going to get by the CLT, like [Bounds for the normal approximation of the maximum likelihood estimator, Anastasiou].
+%The paper [Finite-sample analysis of 𝑀-estimators using self-concordance*] has a good background section on the work of local asymptotic normality (1.1 Classical asymptotic theory) that might be useful.
+The first 2-3 pages of [PARAMETRIC ESTIMATION.
+FINITE SAMPLE THEORY] are also nice
+}
 

### 2021-10-13 aistats.tex
Redo the intro

@@
 \newcommand{\nat}{\theta}
-\newcommand{\m}{m}
+\newcommand{\m}{\mu}
 \newcommand{\meanp}{\m}
@@
 
-We kind of know how to do optimization
-on smooth, strongly convex functions
-when there is some bound on the stochasticity,
-for example in the form of a bounded variance,
-say for the problem
-\begin{equation} %We might want to add a feasible set for theta?
-\min_\theta f(\theta)
-\end{equation}
-with access to stochastic gradients
-$g(\theta)$ such that
-\begin{equation}
-	\expect{\norm{g(\theta) - \nabla f(\theta)}^2} \leq \sigma^2.
-\end{equation}
-Estimating the mean of Gaussian noise is a special case of this problem class, using
-$f(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$ and $g(\theta) \sim \cN(- \mu, \sigma^2)$.
-SGD with a decreasing step-size coincides with the maximum likelihood estimate (MLE),
-and attains the $1/t$ convergence rate.
+Exponential families are among the most widely used statistical models.
+Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
+They are flexible both enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
+In particular, they are appreciated for their convex log-likelihood
+\alignn{
+f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat}
+}
+where $\logpart$ is a convex function.
+This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} 
+or variants of principal component analysis \citep{collins2001generalization} among other applications.
+Another interest of exponential families is their closed form maximum-likelihood estimate (MLE) obtained via moment matching
+\begin{align}
+	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} 
+\end{align}
+Given a conjugate prior, a similar formula holds for the maximum a posteriori (MAP).
 
-But our understanding of stochastic optimization lacks in other domains.
-A generalization of the above problem, at the intersection of optimization and statistics,
-is the problem of minimizing
-\begin{equation}
-	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta)},
-\end{equation}
-When $p(x\cond\theta)$ is an exponential family distribution,
-$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (omitting the reference distribution).
-Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
-The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.
+How can we assess the quality of the MLE, or any other estimator?
+Let us distinguish two lines of work answering this question in statistic.
+(1.) Parameter estimation has focused on the quality of the resulting parameters, using norms or distances between parameter themselves, and yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
+(2.) Density estimation has focused on measuring the discrepancy between the model and the true distribution.
+For this purpose, the Kullback-Leibler (KL) divergence arises naturally from information theory,
+but its lack of robustness to misspecification has led statisticians to study other distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}.
+As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research. 
+In this work, we are addressing the following question.
 
-There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
-As we show, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
-Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
+\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP ?}
 
-We do not have a solution. The goal of this paper
-is to bring this issue to the attention of the community working at the intersection of machine learning, optimization and statistics,
-in the hope that we can make progress on this question;
-Can we build algorithms for stochastic estimation
-when the objective function is not (close to) quadratic
-and the noise is not Gaussian?
+If $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
+or close to quadratic, there are simple upper bounds. 
+But without further assumptions, this problem is difficult.
 
-We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate for which there is a wealth of asymptotic and large sample results \citep{van2000asymptotic,ostrovskii2021finite}.
-%TODO complete
-However, none of the theory we found could address for instance the case of a Gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions.
-%TODO kind of redundant "problem - question - problem" paragraphs, not disentangling stats and optim. To improve
+This question has connections with optimization.
+For one, the MLE is minimizing $f$ (or a regularized version for MAP), and the KL is the sub-optimality
+\alignn{
+f(\nat) - f(\nat^*) = \KL\paren{p(\cdot \cond \nat^*) || p(\cdot \cond \nat) }
+}
+where $\nat^* \in \argmin f(\nat)$.
+For two, in exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
+A convergence rate for this algorithm would translate into the upper bound we seek, similarly to other works on stochastic approximation \citep{bach2013nonstronglyconvex}. %TODO check this
+From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g.
+not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
 
 \paragraph{Contributions}
+After formalizing the problem along with its asymptotic properties, we make the following contributions
 \begin{itemize}
-	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the Gaussian variance estimation problem.
-	\item We show the connections between the optimization and statistical estimation view of the problem and discuss where current theories fail to fully capture the problem.
+	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation is valid.
+	\item We investigate a generalized bias-variance decomposition~\citep{pfau2013generalized}.
+	\item We provide a full solution in the case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$.
+	\item We establish a connection with optimization, and use this problem to highlight areas for progress in the stochastic optimization of barrier objectives.
 \end{itemize}
 
-The statistics and optimization perspective are so tightly linked in this problem, that we believe progress from one field would help the other one.
-Progress from statistics could bring forward research in non-euclidean optimization, or the optimization of barrier losses, for instance helping in the development of accelerated methods.
-On the other hand, optimization can become  a more versatile and general tool for statisticians. % handwavy and shady af.
+A solution to this problem would allow the field of optimization tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression).
+It could also help develop accelerated optimization methods.
 
-\subsection{Related Work}
-
-\paragraph{Density Estimation}
-Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
-Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
-A contrario the KL is not smooth, and it can be infinite in misspecified settings.
-But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.
-
-\paragraph{Online Learning}
-Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
-\[
-R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
-\]
-where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
-Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.
-Unfortunately however, their rate are worse than the offline MAP or MLE algorithms.
-For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$.
-From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
-Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
-This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
-
-% Then;
-% - description of the setting
-% - statistical solutions (asymptotic, l1 or l2 error)
-% - optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
-% - potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
-% - conclusion?
-
-\subsection{Notation}
+\paragraph{Notation}
 $X, T(X)$ and $T$ are random variables, while $x$ is a sample. 
@@
 \section{Technical background}
-
-Exponential families are to statistics what linear models are to regression.
-Many classical model are exponential families, for instance the Gaussian or Bernoulli distributions.
-Their convexity or modeling flexibility make them one of the most widely used statistical models.
-In this section, we review some of their properties.
-We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
+In this section, we review the formalism of exponential families, their duality, a conjugate prior and the corresponding MAP.
+We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a more exhaustive introduction.
 
@@
 where $T_i=T(X_i)$.
-When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
-\begin{align}
-	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
-\end{align}
-This last equation is known as moment matching.
+When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE).
 In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
@@
 all Bregman divergences are locally quadratic.
-Under some assumptions, like self-concordance \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
-For instance, when $T$ is 1-dimensional, we have that $\logpart$ in the following settings:
-exponential distribution,
-Gaussian with known mean,
+Under some assumptions, like self-concordance of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
+\begin{proposition}
+Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF^* = \nabla \conj(\m^*)$.
+Then%\footnote{$0.21$ is the value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
+\aligns{
+	\norm{\m^*-\m}_{\mF^*} < 0.21 
+	\implies
+	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
+}
+\end{proposition}
+We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
+This is true for
+exponential distributions,
+Gaussians with known mean,
 Laplace with known mean,
 Pareto with known minimum value,
-Weibull with known shape $k$. This also holds when
--- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
-\begin{align}
-	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
-\end{align}
-holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\footnote{The value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
-Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
+or Weibull with known shape $k$. 
+The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.
+
+\citet{kakade2010learning} assumes a bound similar to self concordance, but only at $\nat^*$ and on all higher order moments.
+They quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply yield a norm on $\m$.
 
@@
 \begingroup
-\renewcommand*{\arraystretch}{1.25}%
+%\renewcommand*{\arraystretch}{1.25}%
 \newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
@@
 %\\
-Variance $\forall \nat\in\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
+Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
 & \redxmark & \greencmark  & \redxmark
 \\
-Variance at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
+Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
 & \redxmark & \greencmark  & \greencmark
 \\
-Optimality gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
+Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
 & \greencmark & \redxmark & \greencmark
@@
 \TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
+
+\subsection{Related Work}
+
+\paragraph{Density Estimation}
+While statistics has a plethora of asymptotic results \citep{vdv1998asymptotic}, there is less work in finite sample bounds.
+Since the KL explode to $\infty$ if the model is unable to give weight to some samples, a lot of this work has been devoted to distances, such as $L^2$ \citep{tsybakov2009introduction}, $L^1$ \citep{devroye2001combinatorial} or more recently Hellinger distance \citep{baraud2017new}.
+Contrary to the KL, these distance are symmetric and behave smoothly.
+
+\paragraph{Online Learning}
+Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
+\[
+R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
+\]
+where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
+Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.
+Unfortunately however, their rate are worse than the offline MAP or MLE algorithms.
+For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$.
+From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
+Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
+This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
 

### 2021-10-13 aistats.tex
Another pass on the intro

@@
 
-\aistatstitle{Convergence Rates for the MAP of an Exponential Family, and Stochastic Mirror Descent -- an Open Problem}
+\aistatstitle{Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem}
 
@@
 Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
-They are flexible both enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
+They are both flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
 In particular, they are appreciated for their convex log-likelihood
@@
 }
-where $\logpart$ is a convex function.
+where $\logpart$ is a convex function and $\nat$ is a parameter vector.
 This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} 
@@
 How can we assess the quality of the MLE, or any other estimator?
-Let us distinguish two lines of work answering this question in statistic.
-(1.) Parameter estimation has focused on the quality of the resulting parameters, using norms or distances between parameter themselves, and yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
-(2.) Density estimation has focused on measuring the discrepancy between the model and the true distribution.
+Let us distinguish two lines of work addressing this question in statistic.
+(1.) \emph{Parameter estimation} has focused on the quality of the resulting parameters, using norms or distances between parameter themselves, and yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
+(2.) \emph{Density estimation} has focused on measuring the discrepancy between the model and the true distribution.
 For this purpose, the Kullback-Leibler (KL) divergence arises naturally from information theory,
@@
 For two, in exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
-A convergence rate for this algorithm would translate into the upper bound we seek, similarly to other works on stochastic approximation \citep{bach2013nonstronglyconvex}. %TODO check this
+A convergence rate for this algorithm would translate into the upper bound we seek.
+%similarly to other works on stochastic approximation \citep{bach2013nonstronglyconvex}. %TODO check this
 From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g.
 not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
+In fact we can see this problem as a very special case to study SMD, revealing the limits of current analysis.
 
@@
 \begin{itemize}
+	\itemsep0em 
 	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation is valid.
-	\item We investigate a generalized bias-variance decomposition~\citep{pfau2013generalized}.
+	\item We investigate a generalized bias-variance decomposition~\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem}
 	\item We provide a full solution in the case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$.
@@
 where  $\nat$ is called natural (or primal) parameter is specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) $\nu(dx)$, the density with respect to a base measure.
+Since the exponential is positive, $p$ has the same support as $\nu$.
 The function $\logpart$ is called the log-partition function that acts as a normalization term, since
@@
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-Asymptotic results typically show that natural parameters $\nat$ is asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
-Here the MAP is more simply expressed with mean parameters $\meanp$.
+Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
+With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
 We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
@@
     \bregmanconj(\m^* ; \m)
-    &= \frac{\norm{\m^* - \m}^2_{\mF^*}}{2}
+    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
     + O(\norm{\m - \m^*}^3),
 \end{align}
-where the Mahalanobis norm  $\| x \|_{\mF^*}^2 = x^\top \mF^* x$  is induced by $\mF^*  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
-It happens that  $F^*$ is also the inverse \textit{Fisher information matrix}, since
+where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
+It happens that  $\mF$ is also the inverse \textit{Fisher information matrix}, since
 \begin{align}
-    \mF^*
+    \mF
     :=\nabla^2\conj(\m^*)
@@
 This asymptote is also independent of the optimum $\meanp^*$.
+We point the curious reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive review.
 
 \subsection{Quadratic Case}
+\rlp{make it multivariate, with a reference to \citep{bach2013nonstronglyconvex}?}
 As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
@@
 so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
+\TODO{connection with optim non-smooth rate.}
 
-{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex), then
+{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
 \begin{align}
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
+\TODO{connection with optim smooth rate.}
 
@@
 \begin{proposition}
-Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF^* = \nabla \conj(\m^*)$.
+Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
 Then%\footnote{$0.21$ is the value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
 \aligns{
-	\norm{\m^*-\m}_{\mF^*} < 0.21 
+	\norm{\m^*-\m}_{\mF} < 0.21 
 	\implies
-	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
+	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
 }
@@
 \fdk{
-We know that if the problem is strongly convex (in $\norm{\cdot}_2$), averaged SGD achieves $O(1/n)$ rate 
-%[Robust stochastic approximation approach to stochastic programming].
+We know that if the problem is strongly convex (in $\norm{\cdot}_2$), averaged SGD achieves $O(1/n)$ rate.
+[Robust stochastic approximation approach to stochastic programming].
 (The paper uses a bounded stochastic gradient assumption, but can be relaxed to bounded variance instead.) However, strong-convexity in the Euclidean norm is not strictly necessary to find an algorithm with convergence rate of $O(1/n)$, as shown by [Non-strongly-convex smooth stochastic approximation with convergence rate O(1/n)] for linear and logistic regression.
 Their proof techniques based on self-concordance have been extended to other setting, [Finite-sample analysis of M-estimators using self-concordance, Learning Exponential Families in High-Dimensions: Strong Convexity and Sparsity].
-%The same tools are used in statistics, and evething essentially boils down to controlling how different we are from a Gaussian rv, that we're going to get by the CLT, like [Bounds for the normal approximation of the maximum likelihood estimator, Anastasiou].
-%The paper [Finite-sample analysis of 𝑀-estimators using self-concordance*] has a good background section on the work of local asymptotic normality (1.1 Classical asymptotic theory) that might be useful.
-The first 2-3 pages of [PARAMETRIC ESTIMATION.
-FINITE SAMPLE THEORY] are also nice
+The same tools are used in statistics, and evething essentially boils down to controlling how different we are from a Gaussian rv, that we're going to get by the CLT, like [Bounds for the normal approximation of the maximum likelihood estimator, Anastasiou].
 }
@@
 
-\fdk{
-{\bf Online learning paragraph?}
-\citet{azoury2001relative,freund1996predicting,dasgupta2007online},
-Bubeck's 2011.
-}
-
 Recently, these assumptions have been relaxed to allow for non-smooth functions $f$
@@
 \centering
+\caption{Comparison of Analysis of SMD, \\
+under relative smoothness and relative strong convexity assumptions. 
+Barrier tells whether the assumption can hold with barrier objectives.
+$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
+}
 \begin{tabular}{lccc}
@@
 \end{tabular}
-\caption{
-Summary of results about SMD with relative smoothness and relative strong convexity assumptions. 
-Barrier tells whether the assumption can hold with barrier objectives.
-$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
-}
 \label{tbl:assumptions}

### 2021-10-13 aistats.tex
Add ref to large deviations

@@
 	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} 
+	\label{eq:defMLE}
 \end{align}
@@
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
+    \label{eq:defMAP}
 \end{align}
@@
 \centering
-\caption{Comparison of Analysis of SMD, \\
+\caption{Summary of results for SMD
 under relative smoothness and relative strong convexity assumptions. 
@@
 
-\paragraph{Density Estimation}
-While statistics has a plethora of asymptotic results \citep{vdv1998asymptotic}, there is less work in finite sample bounds.
+\paragraph{Parameter Estimation.}
+Statistics has a plethora of asymptotic results \citep{vdv1998asymptotic}.
+In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes the decay of the tail.
+
+\paragraph{Density Estimation.}
 Since the KL explode to $\infty$ if the model is unable to give weight to some samples, a lot of this work has been devoted to distances, such as $L^2$ \citep{tsybakov2009introduction}, $L^1$ \citep{devroye2001combinatorial} or more recently Hellinger distance \citep{baraud2017new}.

### 2021-10-13 aistats.tex
Start rephrasing the open problems

@@
 f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat}
+\label{eq:defNLL}
 }
@@
 f(\nat) - f(\nat^*) = \KL\paren{p(\cdot \cond \nat^*) || p(\cdot \cond \nat) }
+\label{eq:suboptimalityKL}
 }
@@
 
-The exponential family for data $X\in \cX$ reads
+The density of an exponential family for data $x \in \cX$ reads
 \begin{equation}
-	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
+	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
 	 \label{eq:def_expfamily}
 \end{equation}
-where  $\nat$ is called natural (or primal) parameter is specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) $\nu(dx)$, the density with respect to a base measure.
+where  $\nat$ is called natural (or primal) parameter.
+It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) a base measure$\nu$ on $\cX$.
 Since the exponential is positive, $p$ has the same support as $\nu$.
@@
 
-For convenience, in this paper we focus on steep, regular exponential families
-where $\logpart$ is strictly convex function of Legendre type,
-and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
-(See \citet{barndoffnielsen2014information}). Also, when explicit, we write the random variable $T = T(X)$.
+For convenience, in this paper we focus on steep, regular exponential families with minimal statistic $T$.
+Then $\logpart$ is a strictly convex function of Legendre type,
+and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex
+\citep{barndoffnielsen2014information}. 
+Also, when explicit, we write the random variable $T = T(X)$.
 
@@
 \paragraph{Duality}
-The logpartition function $\logpart$ verifies the two following identities:
+The log-partition function $\logpart$ verifies the two following identities:
 \begin{align}
-    \nabla\logpart(\nat) &=  \expect[p(T|\nat)]{T} =: \meanp \\
-    \nabla^2 \logpart(\nat) &= \Cov_\nat[T] \geq 0
+    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp \\
+    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0
 \end{align}
 where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
-We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both represent the same distribution.
+Given that $\logpart$ is strictly convex, its Hessian is positive definite and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
+We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both are linked and represent the same distribution.
 
-We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
+We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the log-partition function
 \begin{align}
-	\conj(\m) = \min_{\nat\in\Theta}\langle \m, \nat \rangle - \logpart(\nat) \; ,
+	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
+	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
 \end{align}
 which matches the common notion of \textit{entropy} in information theory.
-In this paper, we assume that the sufficient statistic $T$ is minimal, e.g., its covariance is definite positive for all $\nat$.
-In such case, the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$, and moreover, $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}).
-More concisely,
-\[
+By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}). More concisely,
+\aligns{
 	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
-\]
+}
 
-% The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).
-
-\begin{figure}[t] % @remi In the figure you use mu instead of m??
+\begin{figure}[t] 
 	\centering
@@
 
-\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps)
-We now introduce the Bregman divergence induced by $\logpart$, that measure the distance between two parameters $\nat$ and $\nat_0$,
+\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
+We now introduce the Bregman divergence induced by $\logpart$, that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
 \begin{align}
@@
     = \bregmanconj ( \meanp_0 ; \meanp) \; .
+    \label{eq:bregman_switch}
 \end{align}
@@
 This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
-Intuitively, $n_0$ is the number of fictive data points observed from a distribution with parameter mean $\nat_0$.
+Intuitively, $n_0$ is the number of fictive data points observed from a distribution with natural parameter $\nat_0$.
 
 \paragraph{Maximum A Posteriori (MAP).} 
-\textcolor{red}{\textbf{D.S.: This section is very unclear. }}
-Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
+Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD|\nat)p(\nat)$.
+Plugging in~\eqref{eq:def_expfamily} and ~\eqref{eq:def_prior} yields
 \begin{align}
 	p(\nat \cond \mD_n)
-	%\propto p(\mD|\nat)p(\nat)
     \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
@@
 \end{align}
-reaching its maximum in $\MAPt^\text{MAP}$ such that
+which reaches its maximum in $\MAPt^\text{MAP}$ such that
 \begin{align}
@@
 where $T_i=T(X_i)$.
-When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE).
+When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)~ \eqref{eq:defMLE}.
 In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
-The MLE and MAP estimates are statistics of the dataset $\mD_n$.
-Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.
+%The MLE and MAP estimates are statistics of the dataset $\mD_n$.
 
 \section{Open Problem}
+Let us assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
+We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
+As we will see, this is already an interesting setting.
+We wish to estimate how well the MLE or a MAP model the true distribution.
+
+A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$.
+In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
+Using~\eqref{eq:defNLL} and~\eqref{eq:bregman_switch}, we can also write the KLwith Bregman divergences
+\alignn{
+	\KL(p_{\nat^*} || p_\nat)
+	 = \bregman(\nat ; \nat^*) 
+	 = \bregmanconj(\m^* ; \m) \; .
+}
+
+In the words of statistical decision theory would refer to these quantities as \emph{frequentist risk} of the estimators, when the statistical loss is given by the divergence the KL.
+
 Given a distribution $\cD$, our objective is the expected log-likelihood 
@@
 Note that $\cD$ does not have to be part of the exponential family.
-Let $\nat^* := \argmin f(\nat)$.
-Assuming $\nat^*$ exists within $\Theta$ means $\cD$ is such that $\E[T] = \meanp^* \in \cM$.
+If $\cD$ is such that $\E[T] = \meanp^* \in \cM$, then $f(\nat)$ reaches its minimum $\nat^*\in\Theta$.
 Then the suboptimality on the population log-likelihood can be written with Bregman divergences
-\footnote{
-Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
-and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
-}
+%\footnote{
+%Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
+%and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
+%}
 \begin{align}
@@
 {\bf Remark.}
-This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
-However there is a connection between the random variable $T(X)$ and the metric $\logpart$.
-Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
-For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
-
-{\bf Remark 2.}
-The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$.
-Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.
+While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, for instance in Gaussian covariance estimation when $n \leq d + 1$.
+Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality.
 

### 2021-10-13 aistats.tex
Add optimization paragraph in technical background

@@
     - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
+    \label{eq:defBregman}
 \end{align}
@@
 \begin{align}
-	\bregman (\nat ; \nat_0)
-    = \bregmanconj ( \meanp_0 ; \meanp) \; .
+	\bregman (\nat ; \nat^*)
+    = \bregmanconj ( \meanp^* ; \meanp) \; .
     \label{eq:bregman_switch}
@@
 Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD|\nat)p(\nat)$.
-Plugging in~\eqref{eq:def_expfamily} and ~\eqref{eq:def_prior} yields
+Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
 \begin{align}
@@
 In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
-%The MLE and MAP estimates are statistics of the dataset $\mD_n$.
+
+\paragraph{Optimization}
+We can rewrite~\eqref{eq:defMAP} as
+\alignn{
+\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
+\label{eq:mean_update}
+}
+where $\lr_n := \inv{n_0 + n}$. 
+Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
+If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
+\alignn{
+\nabla\conj(\hat \nat_{n}) 
+= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
+}
+which we recognize as the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ and step-size schedule $\lr_n$.
 
@@
 
-A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$.
+A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
 In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
-Using~\eqref{eq:defNLL} and~\eqref{eq:bregman_switch}, we can also write the KLwith Bregman divergences
+Using~\eqref{eq:defNLL} and~\eqref{eq:bregman_switch}, we can also write the KL with Bregman divergences
 \alignn{
@@
 
-In the words of statistical decision theory would refer to these quantities as \emph{frequentist risk} of the estimators, when the statistical loss is given by the divergence the KL.
-
-Given a distribution $\cD$, our objective is the expected log-likelihood 
-\begin{align}
-	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
-	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
-\end{align}
-Note that $\cD$ does not have to be part of the exponential family.
-If $\cD$ is such that $\E[T] = \meanp^* \in \cM$, then $f(\nat)$ reaches its minimum $\nat^*\in\Theta$.
-Then the suboptimality on the population log-likelihood can be written with Bregman divergences
-%\footnote{
-%Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
-%and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
-%}
-\begin{align}
-	 f(\nat) - f(\nat^*)
-	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
-	 = \bregman(\nat ; \nat^*) 
-	 = \bregmanconj(\m^* ; \m) \; .
-\end{align}
-Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
+How does this quantity behave when $\nat$ is the MLE or a MAP ? 
+For instance can we get upper bounds on the expected KL ?
 \begin{align}
@@
 \end{align}
-where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
-More explicitly, can we write an upper bound that does not involve this expectation over the dataset.
+where the outer expectation is on the dataset $\mD$.
+More explicitly, can we write an upper bound that does not involve this expectation over the dataset?
+Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators, when the statistical loss is given by the KL divergence ?
 
-In general, for arbitrary distributions $\cD$, it is a hard problem, especially since the KL is known for being brittle to misspecification \tocite, so let us focus on the well-specified case first $\cD = p(\cdot \cond \nat^*)$.
-Surely this problem has been solved, hasn't it ?
-In some special cases yes, but it seems the general answer is no. 
+Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
+%TODO. what we did and are going to present next.
 We are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n$.
@@
 {\bf Remark.}
-While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, for instance in Gaussian covariance estimation when $n \leq d + 1$.
+While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, 
+for instance in Gaussian covariance estimation when $n \leq d + 1$.
+Even worse with categorical variables, the MLE is always ill-defined. %TODO
 Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality.

### 2021-10-13 aistats.tex
fixed most of the warnings

@@
 \documentclass[twoside]{article}
+
+\usepackage{silence}
+% Disable all warnings issued by latex starting with "You have..."
+% This is a hack, there is no clean solution for that unless we install locally the package
+\WarningFilter{latex}{You have requested package}
+
+\pdfsuppresswarningpagegroup=1 %removes the warning
+% This is a hack, the other way to remove is to convert all pdf into postscript or eps.
+
 \usepackage{aistats/aistats2022}
@@
 \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{lmodern}
 \usepackage{hyperref}       % hyperlinks
@@
 	pdfauthor={Anonymous},
+	% draft,
+	final,
 }
@@
 \DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}
-
-\hypersetup{draft}
 
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{{figs/gaussian_variance.pdf}}
+	\includegraphics[width=.4\textwidth]{figs/gaussian_variance.pdf}
 	\caption{
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{{figs/gaussians/new_linear_n0=1}.pdf}
+	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
 	\caption{
@@
 From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
-Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
+Also, \citet{dasgupta2007online} prove the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
 This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
@@
 \bibliographystyle{apalike}
-\bibliography{references.bib}
+\bibliography{references} % Removed the .bib as this mess up with the compiler
 

### 2021-10-13 aistats.tex
Add a few refs about quadratics

--- 2021-10-13 aistats.tex
@@
 \begin{align}
-	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}.
+	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
 	=\frac{\Var(T)}{2n}
@@
 	\expect{\bregmanconj(\meanp^*; \MAPm)}
-	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}.
+	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
 	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
+Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
 \TODO{connection with optim smooth rate.}
@@
 \end{proposition}
+Proof in the appendix.
 We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
@@
 
-\citet{kakade2010learning} assumes a bound similar to self concordance, but only at $\nat^*$ and on all higher order moments.
-They quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply yield a norm on $\m$.
+With some more work, it is possible to get a wealth of results under this assumption.
+For instance \citet{ostrovskii2021finite} characterizes the number of samples needed to reach this quadratic behavior for arbitrary parametric models, under the assumptions of self-concordance, or pseudo self-concordance \citep{bach2010self}.
+Or in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher order moments of $\logpart$ in $\nat^*$, to quantify when $\logpart$ can be sandwiched between quadratics in $\nat$, but this does not directly translate into quadratics in $\m$.
 
@@
 
-\TODO{Incorporate this content} ( about \citet{ostrovskii2021finite} as well as
 \citet{anastasiou2017bounds} and
@@
 For normal distributions set $\mu=x$ and $\sigma^2 \rightarrow 0$ to get $p(x\cond \nat) \rightarrow +\infty$. 
-For gamma, set $\alpha = \beta x$ and $beta \rightarrow +\infty$.
+For gamma, set $\alpha = \beta x$ and $\beta \rightarrow +\infty$.
 For beta, set $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.

### 2021-10-13 aistats.tex
Clean the end of Open Problem

@@
 which we recognize as the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ and step-size schedule $\lr_n$.
+\emph{MAP is an instantiation of SMD.}
 
@@
 Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
-%TODO. what we did and are going to present next.
-We are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n$.
+There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights}.
+There are also results for special cases, such as $\cN(0, \sigma^2)$ (See \S\ref{sec:example}). 
+But we are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n \lesssim d$.
 
-{\bf Remark.}
+{\bf Unboundedness of~\eqref{eq:bregmanMLE}.}
 While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, 
 for instance in Gaussian covariance estimation when $n \leq d + 1$.
-Even worse with categorical variables, the MLE is always ill-defined. %TODO
-Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality.
+Even worse, with categorical variables, there is always a non-zero probability to never sample one of the categories.
+In such the MLE will give zero weight to this category and $KL(p_{\nat^*} || p_\text{MLE} ) = +\infty$.
+This means that the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
+Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality but this is a difficult endeavor.
+These examples are rather making a case for regularized estimators such as MAP, for which we may find interesting upper bounds.
 
 \section{Insights}
+\label{sec:insights}
 
@@
 \section{Examples}
+\label{sec:example}
 

### 2021-10-13 aistats.tex
finished the pass on section 5

--- 2021-10-13 aistats.tex
@@
 \begin{align}
-	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
+	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}.
 	=\frac{\Var(T)}{2n}
@@
 	\expect{\bregmanconj(\meanp^*; \MAPm)}
-	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
+	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{(n+n_0)^2}.
 	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
-Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
 \TODO{connection with optim smooth rate.}
@@
 \end{proposition}
-Proof in the appendix.
 We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
@@
 
-With some more work, it is possible to get a wealth of results under this assumption.
-For instance \citet{ostrovskii2021finite} characterizes the number of samples needed to reach this quadratic behavior for arbitrary parametric models, under the assumptions of self-concordance, or pseudo self-concordance \citep{bach2010self}.
-Or in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher order moments of $\logpart$ in $\nat^*$, to quantify when $\logpart$ can be sandwiched between quadratics in $\nat$, but this does not directly translate into quadratics in $\m$.
+\citet{kakade2010learning} assumes a bound similar to self concordance, but only at $\nat^*$ and on all higher order moments.
+They quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply yield a norm on $\m$.
 
@@
 
+\TODO{Incorporate this content} ( about \citet{ostrovskii2021finite} as well as
 \citet{anastasiou2017bounds} and
@@
 
-\subsection{Bias-Variance Decomposition} % DS: Arrived here
-In both examples (add link), the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
-In fact, it is possible to write such a decomposition for any Bregman divergence
-Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
-As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
-\begin{align}
-	\expect{\bregmanconj(\m^* ; \hat \m_n)}
-	&= {\bregmanconj(\m^* ; \tilde \m_n)}
+\subsection{Bias-Variance Decomposition} 
+In both examples \textcolor{red}{(add link)}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
+In fact, such a decomposition exists for any Bregman divergence.
+Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be its corresponding mean parameter.
+As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes into
+\begin{equation}
+	\expect{\bregmanconj(\m^* ; \hat \m_n)} = {\bregmanconj(\m^* ; \tilde \m_n)}
 	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
 	\label{eq:primal_pivot}
-\end{align}
+\end{equation}
 
@@
 An estimator will be unbiased if its primal expectation is non-zero.
-This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
+However, this is not true for the the gaussian variance MLE, where the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
 A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.
@@
 \paragraph{Illustrations.}
-We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
-In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
-\rlp{Tune figures.}
+We plot the bias-variance decompositions for $\cN(0,\sigma^2)$ (Figure~\ref{fig:variance_decomposition}) and  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
+Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:bias-variance-numerical} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) .
+\rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}
 
@@
 \subsection{Gaussian Variance}
-A core example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
-The density is	$p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}$.
+A core example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
+Its density reads 
+\[
+	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
+\]
 Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
-Mean and natural parameters are roughly inverse of each other $\nat = -\inv{2 \m}$.
-Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
+Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
+\textcolor{red}{The explanation here is not clear enough, or too concise} Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
 \begin{align}
-	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi) \\
+	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
 	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
@@
 Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
-It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, eg a negative Gamma distribution.
+It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
 It also means $\bregman$ and $\bregmanconj$ have the same shape
@@
 	\bregmanconj( \m_*; \m_n)
-	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) \\
+	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
 	\bregman( \nat_n; \nat_* )
@@
 \end{align}
-In other words, this divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function $\phi$
+Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
 \begin{align}
@@
 illustrated in Figure~\ref{fig:phi}.
-Below we report upper bounds on the expected value of this divergence for the MLE and the MAP.
 
@@
 \end{figure}
+
+Theorem \ref{eq:MLE_rate} reports upper bounds on the expected value of this divergence for the MLE and the MAP.
 
@@
 
-For the MAP, we get an upper bound thanks to the inequality $ - \log(z) \leq \inv{z} - 1$ which implies
+For the MAP, we get an upper bound thanks to the inequality $ - \log(z) \leq \inv{z} - 1$, which implies
 \begin{align}
@@
 \begin{theorem}[MAP Bound]
- For $n\in \naturalnumbers$, let us  define
+For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
  \begin{align}
-	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
- \end{align}
-The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
- \begin{equation}
-	\expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
+	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
 	\leq \begin{cases}
@@
 	\end{cases}
-	\label{eq:MAP_rate}
-\end{equation}
+	\label{eq:MAP_rate}\\
+	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
+\end{align}
 \end{theorem}
@@
 This correlates well with numerical observations.
-Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$.
-We are basically losing a factor 2 compared to $1/2n$.
+Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$, therefore losing a factor 2 compared to $1/2n$.
 
@@
 We now show how MAP can be interpreted as stochastic mirror descent (SMD).
-This means that 1. we may obtain a convergence rate for MAP from an optimization analysis, and 2. any insights gained from the MAP may inform further design and analysis of SMD.
+This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
 We then review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
@@
 For normal distributions set $\mu=x$ and $\sigma^2 \rightarrow 0$ to get $p(x\cond \nat) \rightarrow +\infty$. 
-For gamma, set $\alpha = \beta x$ and $\beta \rightarrow +\infty$.
+For gamma, set $\alpha = \beta x$ and $beta \rightarrow +\infty$.
 For beta, set $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.

### 2021-10-13 aistats.tex
Fix Asymptote section with Simon's feedback

--- 2021-10-13 aistats.tex
@@
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
+We now review classical asymptotic results. 
+We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
 Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
@@
 We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
-\begin{align}
+\alignn{
     \bregmanconj(\m^* ; \m)
@@
     + O(\norm{\m - \m^*}^3),
-\end{align}
+    \label{eq:bregman_asymptote}
+}
 where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
-It happens that  $\mF$ is also the inverse \textit{Fisher information matrix}, since
-\begin{align}
+It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
+\aligns{
     \mF
@@
     = \Cov_{\nat^*}[T(X)]^{-1}  \; .
-\end{align}
-From there, it is possible to show an asymptotic rate for the MLE and the MAP \textcolor{red}{D.S.: How?}
+}
+Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregman_asymptote}, one can show that MLE and MAP verify 
 \begin{align}
@@
 \end{align}
+See proof in the Appendix.
 Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
 This asymptote is also independent of the optimum $\meanp^*$.
-We point the curious reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive review.
 
 \subsection{Quadratic Case}
-\rlp{make it multivariate, with a reference to \citep{bach2013nonstronglyconvex}?}
-As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
-For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean 
-\[ \cX=\real,\quad \nu(dx) = \exp(\half[-x^2])dx,\quad T(X)=X.\]
+As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
+For instance, this is the log-partition of a Gaussian with variance $I$ and unknown mean 
+\[ \cX=\real^d,\quad \nu(dx) = \exp(\half[-\|x\|^2])dx,\quad T(x)=x.\]
 Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
@@
 so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
+Note that none of the standard exponential families verify this assumption.
 \TODO{connection with optim non-smooth rate.}

### 2021-10-13 aistats.tex
Small fixes in Example section

@@
     + O(\norm{\m - \m^*}^3),
-    \label{eq:bregman_asymptote}
+    \label{eq:bregmanTaylor}
 }
@@
 }
-Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregman_asymptote}, one can show that MLE and MAP verify 
+Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show that MLE and MAP verify 
 \begin{align}
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
+Besides the Gaussian with known mean, no standard exponential family verify this assumption either.
 Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
@@
 \subsection{Locally Quadratic Case}
-\textcolor{red}{Check this! In our case, since we assume $A$ and $A^*$ to be strictly convex,}
-all Bregman divergences are locally quadratic.
+FromTaylor expansion~~\eqref{eq:bregmanTaylor}, 
+we know that all Bregman divergences are locally quadratic.
 Under some assumptions, like self-concordance of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
 \begin{proposition}
-Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
-Then%\footnote{$0.21$ is the value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
+Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
 \aligns{
@@
 \end{proposition}
-Proof in the appendix.
+Proof in the appendix
+\footnote{In case the reader is intrigued, $0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
 We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
@@
 
-\section{Examples}
-\label{sec:example}
-
-\subsection{Gaussian Variance}
-A core example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
+\section{Gaussian Variance Example}
+A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
 Its density reads 
@@
 
-Theorem \ref{eq:MLE_rate} reports upper bounds on the expected value of this divergence for the MLE and the MAP.
-
+In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
 \begin{theorem}[MLE Tight Bound]
+\label{thm:varianceMLE}
 	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
@@
 \end{theorem}
-
 This upper bound is asymptotically tight.
@@
 We get a similar bound for the multivariate generalization :
-the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.
+the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (see appendix).
 
-For the MAP, we get an upper bound thanks to the inequality $ - \log(z) \leq \inv{z} - 1$, which implies
+For MAP, we get an upper bound thanks to
 \begin{align}
 	\label{eq:log_bound}
-	\phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
+	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
 \end{align}
@@
 \begin{theorem}[MAP Bound]
+\label{thm:varianceMAP}
 For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
@@
 \end{theorem}
-
+\rlp{upper bound and compare the symmetrized bregman instead?}
 This inequality highlights a clear variance-bias decomposition.
@@
 For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
-This correlates well with numerical observations.
-Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$, therefore losing a factor 2 compared to $1/2n$.
+This correlates well with numerical observations (see appendix).
+Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
+
+These results are instantiation the bias-variance decomposition and giving us a flavor of what we would like as a general result for exponential families : a finit sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
+In the next section, we investigate what this problem is brining to optimization.
 

### 2021-10-13 aistats.tex
Remove MAP=SMD subsection

@@
 }
-which we recognize as the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ and step-size schedule $\lr_n$.
+which we recognize as the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ and step-size schedule $\lr_n$ initialized at $\nat_0$.
 \emph{MAP is an instantiation of SMD.}
+Note that this analogy does not hold for the MLE %($n_0=0$)
+as it does not have any initialization to start from.
 
@@
 \label{sec:optimization}
-
-We now show how MAP can be interpreted as stochastic mirror descent (SMD).
+As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
 This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
-We then review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
+We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
 \rlp{mention that examples have so far been lacking for relative SMD.}
-
-\subsection{MAP as Stochastic Mirror Descent}
-\label{ssec:MAP=SMD}
-By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
-Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
-\begin{align*}
-	\hat \nat_{n}
-	= \argmin_\nat f_n(\nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
-    %\label{eq:SBPP}
-\end{align*}
-which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$.
-%Maybe skip this SBPP story altogether
-Expanding $f_n(\nat) = \logpart(\nat) - T_n$ and  playing with the Bregman we get
-\begin{align*}
-		\hat \nat_{n}
-	= \argmin_\nat \langle g_{n}(\hat \nat_{n-1}), \nat \rangle + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
-    %\label{eq:SMD}
-\end{align*}
-where $g_{n}(\nat) := \nabla f_n(\nat) = \hat \m_{n-1} - T_n$.
-We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$.
-From the first order optimality condition, we see this is a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
-This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
-In particular, it could potentially apply to a wide range of distribution $\cD$ on $X$.
-Note that this optimization perspective does not hold for the MLE %($n_0=0$)
-as its implicit uniform prior may correspond to an initialization outside of $\Theta$.
 
@@
 and strong convexity assumption on the potential $\logpart$
-\citep[Th.4.2 \& Th.6.3]{bubeck2015convex}.
+\citep[Th.4.2(MD) \& Th.6.3(SMD)]{bubeck2015convex}.
 In our case, none of these assumptions may hold. 

### 2021-10-13 aistats.tex
pass on optimization section

--- 2021-10-13 aistats.tex
@@
 which we recognize as the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ and step-size schedule $\lr_n$.
-\emph{MAP is an instantiation of SMD.}
 
@@
 Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
-There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights}.
-There are also results for special cases, such as $\cN(0, \sigma^2)$ (See \S\ref{sec:example}). 
-But we are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n \lesssim d$.
+%TODO. what we did and are going to present next.
+We are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n$.
 
-{\bf Unboundedness of~\eqref{eq:bregmanMLE}.}
+{\bf Remark.}
 While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, 
 for instance in Gaussian covariance estimation when $n \leq d + 1$.
-Even worse, with categorical variables, there is always a non-zero probability to never sample one of the categories.
-In such the MLE will give zero weight to this category and $KL(p_{\nat^*} || p_\text{MLE} ) = +\infty$.
-This means that the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
-Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality but this is a difficult endeavor.
-These examples are rather making a case for regularized estimators such as MAP, for which we may find interesting upper bounds.
+Even worse with categorical variables, the MLE is always ill-defined. %TODO
+Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality.
 
 \section{Insights}
-\label{sec:insights}
 
@@
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-We now review classical asymptotic results. 
-We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
 Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
@@
 We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
-\alignn{
+\begin{align}
     \bregmanconj(\m^* ; \m)
@@
     + O(\norm{\m - \m^*}^3),
-    \label{eq:bregman_asymptote}
-}
+\end{align}
 where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
-It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
-\aligns{
+It happens that  $\mF$ is also the inverse \textit{Fisher information matrix}, since
+\begin{align}
     \mF
@@
     = \Cov_{\nat^*}[T(X)]^{-1}  \; .
-}
-Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregman_asymptote}, one can show that MLE and MAP verify 
+\end{align}
+From there, it is possible to show an asymptotic rate for the MLE and the MAP \textcolor{red}{D.S.: How?}
 \begin{align}
@@
 \end{align}
-See proof in the Appendix.
 Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
 This asymptote is also independent of the optimum $\meanp^*$.
+We point the curious reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive review.
 
 \subsection{Quadratic Case}
-As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
-For instance, this is the log-partition of a Gaussian with variance $I$ and unknown mean 
-\[ \cX=\real^d,\quad \nu(dx) = \exp(\half[-\|x\|^2])dx,\quad T(x)=x.\]
+\rlp{make it multivariate, with a reference to \citep{bach2013nonstronglyconvex}?}
+As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
+For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean 
+\[ \cX=\real,\quad \nu(dx) = \exp(\half[-x^2])dx,\quad T(X)=X.\]
 Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
@@
 so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
-Note that none of the standard exponential families verify this assumption.
 \TODO{connection with optim non-smooth rate.}
@@
 \section{Examples}
-\label{sec:example}
 
@@
 
-We now show how MAP can be interpreted as stochastic mirror descent (SMD).
-This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
-We then review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
+This section show how MAP can be interpreted as stochastic mirror descent (SMD).
+Thanks to this strong link, we show that \textbf{1)} we may obtain a convergence rate for MAP from the rate od SMD, and \textbf{2)} any insights gained from the MAP may inform on further design and analysis of SMD.
+Then, we review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three different analyses of SMD with the MAP.
 \rlp{mention that examples have so far been lacking for relative SMD.}
@@
 \label{ssec:MAP=SMD}
-By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
+By the Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
 Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
@@
 %Maybe skip this SBPP story altogether
-Expanding $f_n(\nat) = \logpart(\nat) - T_n$ and  playing with the Bregman we get
+If we expand $f_n(\nat) = \logpart(\nat) - T_n$,  we get (after some manipulation with the Bregman divergence)
 \begin{align*}
@@
 \subsection{Relative Smoothness}
-Mirror descent (MD)\footnote{
-Also known as
-Bregman (proximal) gradient, relative gradient descent or NoLips.
-}
-\citep{nemirovski1983problem,beck2003mirror}
+Mirror descent (MD) \citep{nemirovski1983problem,beck2003mirror}, also known as
+Bregman (proximal) gradient, relative gradient descent or NoLips,
 %Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
@@
 
-Recently, these assumptions have been relaxed to allow for non-smooth functions $f$
-\footnote{$f$ is $L$-smooth if its gradient is $L$-Lipschitz}.
-and non-strongly convex potential $\logpart$, first in the deterministic setting
-\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
-These assumptions have been replaced with
-the $\mu$-strong convexity and $L$-smoothness of $f$
+Recently, these assumptions have been relaxed to the $\mu$-strong convexity and $L$-smoothness of $f$
 \emph{relative} to $\logpart$, defined as
@@
 }
+This extends the theory to a family of non-smooth functions $f$ and non-strongly convex potential $\logpart$, first in the deterministic setting
+\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
 When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
 Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
-Exponential families log-likelihood perfectly fit into this framework, as
+
+\textcolor{red}{Sure of this formulation?} Exponential families log-likelihood perfectly fit into this framework, as
 \aligns{
@@
 is $1$-smooth and $1$-strongly convex relative to $A$.
+However, the (relative) smoothness is not biggest issue in our convergence analysis, but the bound on the variance, as explained in the next section.
 
 \subsection{Bounding the Randomness}
-Smoothness is not the only issue.
 To analyze stochastic algorithms, one also need to quantify the randomness.
@@
 this definition recovers the variance of the stochastic gradient
-$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
-More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
+\[
+	\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C.
+\]
+More generally, if $A$ is $\mu$-strongly convex w.r.t. $\norm{\cdot}$,
 the LHS of \cref{eq:hanzely} is bounded by
@@
 to hold.
-Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
+\textcolor{red}{check formulation} Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
 \aligns{
@@
 thus showing positivity.
-Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
+Under this assumption, \textcolor{red}{who is they?}they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
 
@@
 }
-In words, this is bounding the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
+In words, this bounds the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
 % As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality. 
@@
 
-Seems like a perfect fit then ?
-Well not really, this variance assumption does not hold for exactly the same reason as the one of Hanzely.
+\textcolor{red}{Too informal, plz rephrase} Seems like a perfect fit then ?
+Well not really, this variance assumption does not hold for exactly the same reason as the one of \textcolor{red}{use cite command} Hanzely.
 Having something valid for every possible iterates is way too strong.
@@
 \subsubsection{Bounded Optimality Gap}
-Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explores the hypothesis
+Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explore the hypothesis
 \alignn{
-\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C 
+\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C,
 \label{eq:dorazio}
@@
 where $f_X$ is a stochastic estimate of $f = \expecti{f_X}$. In our case $f_X(\nat) = - \log p(X\cond \nat)$.
-In words, this is lower bounding the expectation of the minimum of these stochastic estimates.
+In words, this lower bounds the expectation of the minimum of the stochastic estimates.
 For probabilistic models, such a bound is finite as soon as the model cannot give infinite density to any data point $x$.
-This holds for instance for discrete distributions 
--- $f_x\geq 0$ because the probability mass is upper bounded by $1$ -- 
+This holds for instance for discrete distributions because the probability mass is upper bounded by $1$,
 but it rules out  many families.
-For normal distributions set $\mu=x$ and $\sigma^2 \rightarrow 0$ to get $p(x\cond \nat) \rightarrow +\infty$. 
-For gamma, set $\alpha = \beta x$ and $\beta \rightarrow +\infty$.
-For beta, set $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
+In the case of normal distributions, setting $\mu=x$ and $\sigma^2 \rightarrow 0$ gets $p(x\cond \nat) \rightarrow +\infty$.
+\textcolor{red}{Check this, I changed the formulation but i am not sure if this remains correct or not.} We have a similar behavior for gamma distribution with $\alpha = \beta x$ and $\beta \rightarrow +\infty$, or with the beta distribution with $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
  Other examples include inverse Gaussians, log-normal, gamma, inverse gamma.
 
- It is possible to overcome this limitation though, by treating batches of samples as 1 sample, and averaging sufficient statistics -- e.g. $Y = \{X_1, \dots, X_k\}$ and $T(Y) = \inv{k}\sum_i T(X_i)$.
- For instance a multivariate normals of dimension $d$ cannot attribute infinite density to $d+1$ samples that are not in an affine subspace.
+It is possible to overcome this limitation by treating batches of samples as single samples by averaging sufficient statistics, e.g.,. $Y = \{X_1, \dots, X_k\}$ and $T(Y) = \inv{k}\sum_i T(X_i)$.
+For instance, a multivariate normals of dimension $d$ cannot attribute infinite density to $d+1$ samples that are not in an affine subspace.
 
-So it seems this approach can yield interesting convergence rates for the MAP, and they do prove linear convergence of the last iterate up a to a variance ball with constant step-size, but the size of this ball $\frac{C}{\alpha}$ \rlp{strong convexity notation} is independent of  the step-size.
-Their descent lemma does not lead to convergence with a decreasing $O(1/t)$ step-size.
+In conclusion, it seems these approaches can yield to interesting convergence rates for the MAP, and they do prove linear convergence of the last iterate up a to a variance ball with constant step-size, but the size of this ball $\frac{C}{\alpha}$ \rlp{strong convexity notation} is independent of the step-size.
+\textcolor{red}{Who is their?} Their descent lemma does not lead to convergence with a decreasing $O(1/t)$ step-size.
 

### 2021-10-13 aistats.tex

--- 2021-10-13 aistats.tex
@@
 \label{sec:optimization}
+
+%This section show how MAP can be interpreted as stochastic mirror descent (SMD).
+Thanks to this strong link, we show that \textbf{1)} we may obtain a convergence rate for MAP from the rate od SMD, and \textbf{2)} any insights gained from the MAP may inform on further design and analysis of SMD.
+%Then, we review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three different analyses of SMD with the MAP.
+%\rlp{mention that examples have so far been lacking for relative SMD.}
+%
+%\subsection{MAP as Stochastic Mirror Descent}
+%\label{ssec:MAP=SMD}
+%By the Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
+%Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
+%\begin{align*}
+%	\hat \nat_{n}
+%	= \argmin_\nat f_n(\nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
+%    %\label{eq:SBPP}
+%\end{align*}
+%which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$.
+%%Maybe skip this SBPP story altogether
+%If we expand $f_n(\nat) = \logpart(\nat) - T_n$,  we get (after some manipulation with the Bregman divergence)
+%\begin{align*}
+%		\hat \nat_{n}
+%	= \argmin_\nat \langle g_{n}(\hat \nat_{n-1}), \nat \rangle + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
+%    %\label{eq:SMD}
+%\end{align*}
+%where $g_{n}(\nat) := \nabla f_n(\nat) = \hat \m_{n-1} - T_n$.
+%We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$.
+%From the first order optimality condition, we see this is a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
+%This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
+%In particular, it could potentially apply to a wide range of distribution $\cD$ on $X$.
+%Note that this optimization perspective does not hold for the MLE %($n_0=0$)
+%as its implicit uniform prior may correspond to an initialization outside of $\Theta$.
+
 As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).

### 2021-10-13 aistats.tex
Add self-concordance sample complexity

--- 2021-10-13 aistats.tex
@@
 \footnote{In case the reader is intrigued, $0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
+To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
+For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
+For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$ (see appendix).
+This means that on average, we need much more samples than the dimension to reach the quadratic regime.
+
 We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.

### 2021-10-14 aistats.tex
Correct all warnings + pass on the intro

@@
 %\usepackage[accepted]{aistats2022}
+
+\setlength{\footskip}{3.31pt}
 
@@
 Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
-They are both flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
+They are flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
 In particular, they are appreciated for their convex log-likelihood
 \alignn{
-f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat}
+f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
 \label{eq:defNLL}
 }
-where $\logpart$ is a convex function and $\nat$ is a parameter vector.
+where $\logpart$ is a convex function and $\nat$ is a vector\textcolor{red}{DS: double check this}, called \textit{natural parameter}.
 This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} 
-or variants of principal component analysis \citep{collins2001generalization} among other applications.
-Another interest of exponential families is their closed form maximum-likelihood estimate (MLE) obtained via moment matching
-\begin{align}
-	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} 
+or variants of principal component analysis \citep{collins2001generalization}, among other applications.
+\textcolor{red}{First, i think its better to have a new paragraph here.
+Then you need to introduce the problematic (like: we need to estimate the parameter theta, and its easy with exponential families becase blabla) }Another interest of exponential families is their closed form maximum-likelihood estimate (MLE) obtained via moment matching:
+\begin{align}
+	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} .
 	\label{eq:defMLE}
@@
 Given a conjugate prior, a similar formula holds for the maximum a posteriori (MAP).
-
-How can we assess the quality of the MLE, or any other estimator?
+\textcolor{red}{DS: I know that for conciseness we do not write the MAP estimate here, but it's super central and I think its important to write the equation.
+WDYT?}
+
+\textcolor{red}{This comes a bit from nowhere.
+Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research.
+We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their ascociated distribution.
+In the first case, the analysis use norms blabla.
+In the second, we use KL"} How can we assess the quality of the MLE, or any other estimator?
 Let us distinguish two lines of work addressing this question in statistic.
@@
 As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
-In this work, we are addressing the following question.
-
-\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP ?}
+
+\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution.
+Hence the natural question:}
+% In this work, we are addressing the following question.
+
+\begin{center}
+	\framebox{
+		\parbox[c][]{0.9\linewidth}{
+			\begin{center}
+			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
+		\end{center}
+		}
+	}
+\end{center}
 
@@
 or close to quadratic, there are simple upper bounds.
+\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
 But without further assumptions, this problem is difficult.
 
+\textcolor{red}{My main problem with the following paragraph here is that its at the same time too detailed for the intro, but also not enough detailed to understand it.}
+
 This question has connections with optimization.
+\textcolor{red}{I would start with "Consider the problem $\min_{\theta\in \Theta} f(\theta)$ (write the equation in display).
+Setting $f=...$, we obtain that KL is the measure of suboptimality, i.e., (equation (3)).
+Moreover, we have that MAP solve a regularized version of f, (write equation) and MLE is the special case where (regularisation = 0)"}
 For one, the MLE is minimizing $f$ (or a regularized version for MAP), and the KL is the sub-optimality
@@
 \paragraph{Contributions}
-After formalizing the problem along with its asymptotic properties, we make the following contributions
+After formalizing the problem \textcolor{red}{restate the problem} along with its asymptotic properties, we make the following contributions.
+\textcolor{red}{Add ref to the main text in the contribution.}
 \begin{itemize}
 	\itemsep0em 
-	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation is valid.
-	\item We investigate a generalized bias-variance decomposition~\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem}
-	\item We provide a full solution in the case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$.
-	\item We establish a connection with optimization, and use this problem to highlight areas for progress in the stochastic optimization of barrier objectives.
+	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation is valid. \textcolor{red}{Approx. of what?}
+	\item We investigate a generalized bias-variance decomposition~\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem} \textcolor{red}{Decomposition of what?}
+	\item We provide a full solution \textcolor{red}{full solution of what?} in the case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$.
+	\item We establish a connection \textcolor{red}{Connection between something (you need to precise what) and optimization} with optimization, and use this problem to highlight areas for progress in the stochastic optimization of barrier objectives.
 \end{itemize}
 
-A solution to this problem would allow the field of optimization tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression).
+\textcolor{red}{I would emphasize more that you are talking about the impact of the work rather than contribution.
+Also I think its better to write this before the contribution, to motivate better why we did this work.
+}A solution to this problem would allow the field of optimization tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression).
 It could also help develop accelerated optimization methods.
@@
 \paragraph{Notation}
-$X, T(X)$ and $T$ are random variables, while $x$ is a sample. 
+$X,\,T(X)$ and $T$ \textcolor{red}{Don't we have T = T(X), but written tuch that it's simpler?
+I think its better to say it that we write T instead of T(X) for simplicity, rather than saying its a random variable.} are random variables, while $x$ is a sample.
 $n$ is the number of samples and $d= \dim(T)$. 
@@
 For instance, this is the log-partition of a Gaussian with variance $I$ and unknown mean 
-\[ \cX=\real^d,\quad \nu(dx) = \exp(\half[-\|x\|^2])dx,\quad T(x)=x.\]
+\[ 
+	\cX=\real^d,\quad \nu(dx) = \exp({\textstyle \half[-\|x\|^2]})dx,\quad T(x)=x.
+\]
 Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
@@
 
-\section{Gaussian Variance Example}
+\section{Gaussian Variance Example}\label{sec:example}
 A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.

### 2021-10-14 aistats.tex
small things

@@
 
-The density of an exponential family for data $x \in \cX$ reads
+The density of an exponential family for a sample $x$ is
 \begin{equation}
@@
 where  $\nat$ is called natural (or primal) parameter.
-It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) a base measure$\nu$ on $\cX$.
+It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, 
+and 2) a base measure $\nu$ on $\cX$.
 Since the exponential is positive, $p$ has the same support as $\nu$.
-The function $\logpart$ is called the log-partition function that acts as a normalization term, since
-\begin{align}
-    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
-\end{align}
-
+The log-partition function $\logpart$ acts as a normalization term, since
+\begin{align}
+    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \;.
+\end{align}
 This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
 
-For convenience, in this paper we focus on steep, regular exponential families with minimal statistic $T$.
+For convenience, we focus on steep, regular exponential families with minimal statistic $T$.
 Then $\logpart$ is a strictly convex function of Legendre type,
@@
 \citep{barndoffnielsen2014information}. 
-Also, when explicit, we write the random variable $T = T(X)$.
+When explicit, we write the random variable $T = T(X)$.
 
@@
 
-We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the log-partition function
+We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
 \begin{align}
@@
 \end{align}
-which matches the common notion of \textit{entropy} in information theory.
-By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}). More concisely,
+which is the common notion of \textit{entropy} in information theory.
+By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$ 
+(see Fig.~\ref{fig:duality}), giving
 \aligns{
@@
 }
-
 \begin{figure}[t] 
@@
 	\label{fig:duality}
-\end{figure}
-
-\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
-We now introduce the Bregman divergence induced by $\logpart$, that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
+\end{figure}%
+%
+%\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
+% We now introduce the Bregman divergence induced by $\logpart$, 
+% that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
+The Bregman divergence induced by $\logpart$, 
+measures the discrepancy between two parameters $\nat$ and $\nat_0$,
 \begin{align}
@@
 with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. 
-In general, bregman distances are non-symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
+In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$. 
 Finally, it is equal to the divergence of its convex conjugate with switched arguments,
 \begin{align}
-	\bregman (\nat ; \nat^*)
-    = \bregmanconj ( \meanp^* ; \meanp) \; .
+	\bregman (\nat ; \nat_0)
+    = \bregmanconj ( \meanp_0 ; \meanp) \; .
     \label{eq:bregman_switch}
@@
 where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
-This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
+This is an exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and natural parameter $(n_0 \m_0, -n_0)$.
 Intuitively, $n_0$ is the number of fictive data points observed from a distribution with natural parameter $\nat_0$.
@@
 \end{align}
-which reaches its maximum in $\MAPt^\text{MAP}$ such that
+which reaches its maximum attempts $\MAPt^\text{MAP}$ such that
 \begin{align}
@@
 where $T_i=T(X_i)$.
-When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)~ \eqref{eq:defMLE}.
-In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
+When $n_0=0$ -- we have no samples from the prior -- we recover the MLE~\eqref{eq:defMLE}.
+We write $\MAPt$ for the MAP, and view the MLE as a special case.
 
@@
 }
-which we recognize as the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ and step-size schedule $\lr_n$ initialized at $\nat_0$.
+which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ 
+and step-size schedule $\lr_n$, initialized at $\nat_0$.
 \emph{MAP is an instantiation of SMD.}
@@
 as it does not have any initialization to start from.
+\fdk{It does, you just forget the initial condition after the first step with step-size 1?}
 
 \section{Open Problem}
-Let us assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
-We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
-As we will see, this is already an interesting setting.
-We wish to estimate how well the MLE or a MAP model the true distribution.
+% Assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
+% We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
+% As we will see, this is already an interesting setting.
+% We wish to estimate how well the MLE or a MAP model the true distribution.
+
+Assume we observe a dataset $\mD$ drawn i.i.d. from an exponential family distribution 
+with parameters $\nat^*$, $p(\cdot \cond\nat^*)$.
+We wish to quantify how well the MLE or a MAP models the true distribution.
 
@@
 }
-
-How does this quantity behave when $\nat$ is the MLE or a MAP ? 
-For instance can we get upper bounds on the expected KL ?
+How does this quantity behave when $\nat$ is the MLE or MAP?
+Can we get upper bounds on the expected KL?
 \begin{align}
 	\label{eq:bregmanMLE}
-	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
+	\text{MLE: } &\quad \expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
 	\label{eq:bregmanMAP}
-	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
-\end{align}
-where the outer expectation is on the dataset $\mD$.
-More explicitly, can we write an upper bound that does not involve this expectation over the dataset?
-Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators, when the statistical loss is given by the KL divergence ?
+	\text{MAP: } &\quad \expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
+\end{align}
+where the expectation is on the data $\mD$.
+More explicitly, is there an upper bound that does not involve this expectation over the dataset?
+Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators 
+when the loss is the KL divergence?
 
 Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
-There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights}.
-There are also results for special cases, such as $\cN(0, \sigma^2)$ (See \S\ref{sec:example}). 
-But we are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n \lesssim d$.
+There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights},
+or results for special cases such as $\cN(0, \sigma^2)$ (see \S\ref{sec:example}). 
+But we are yet to find a solution encompassing a broad range of exponential families, 
+and applicable to small sample sizes $n \lesssim d$.
 
@@
 While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite, 
-for instance in Gaussian covariance estimation when $n \leq d + 1$.
-Even worse, with categorical variables, there is always a non-zero probability to never sample one of the categories.
-In such the MLE will give zero weight to this category and $KL(p_{\nat^*} || p_\text{MLE} ) = +\infty$.
-This means that the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
-Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality but this is a difficult endeavor.
-These examples are rather making a case for regularized estimators such as MAP, for which we may find interesting upper bounds.
+for instance when estimating the covariance of a Gaussian when $n \leq d + 1$.
+Even worse, with categorical variables, there is a non-zero probability to never sample one of the categories. 
+In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$,
+and the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
+Instead of taking the expectation, one might want to bound the risk in high probability, 
+without resorting to Markov inequality but this is a difficult endeavor.
+These examples make a case for regularized estimators such as MAP, 
+for which we may find upper bounds.
 
@@
 As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
-For instance, this is the log-partition of a Gaussian with variance $I$ and unknown mean 
+For instance, this is the log-partition of a Gaussian with known variance $I$,
 \[ 
-	\cX=\real^d,\quad \nu(dx) = \exp({\textstyle \half[-\|x\|^2]})dx,\quad T(x)=x.
+	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
 \]

### 2021-10-14 aistats.tex
Start incorporating Damien's feedback

--- 2021-10-14 aistats.tex
@@
 \alignn{
-f(\nat) = \E[-\log p(X \cond \nat)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
+f(\nat) = \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
 \label{eq:defNLL}
 }
-where $\logpart$ is a convex function and $\nat$ is a vector\textcolor{red}{DS: double check this}, called \textit{natural parameter}.
+where $\logpart$ is a convex function and $\nat\in|Theta$ is a vector called \textit{natural parameter}.
 This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} 
 or variants of principal component analysis \citep{collins2001generalization}, among other applications.
-\textcolor{red}{First, i think its better to have a new paragraph here.
-Then you need to introduce the problematic (like: we need to estimate the parameter theta, and its easy with exponential families becase blabla) }Another interest of exponential families is their closed form maximum-likelihood estimate (MLE) obtained via moment matching:
-\begin{align}
-	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} .
+
+In general, we want to estimate $\nat$ from some data $\mD = (x_1, \dots, x_n)$.
+For this purpose, not only is $f$ convex, but it even enjoys a closed form maximum-likelihood estimate (MLE)
+\begin{align}
+	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
 	\label{eq:defMLE}
 \end{align}
-Given a conjugate prior, a similar formula holds for the maximum a posteriori (MAP).
-\textcolor{red}{DS: I know that for conciseness we do not write the MAP estimate here, but it's super central and I think its important to write the equation.
-WDYT?}
-
-\textcolor{red}{This comes a bit from nowhere.
+This rule is known as moment matching.
+Given a conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP). 
+%\textcolor{red}{DS: I know that for conciseness we do not write the MAP estimate here, but it's super central and I think its important to write the equation. WDYT?}
+% I think it's okay to have it only later, and refernece it here.
+
+%\textcolor{red}{This comes a bit from nowhere.
 Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research.
-We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their ascociated distribution.
+We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution.
 In the first case, the analysis use norms blabla.
-In the second, we use KL"} How can we assess the quality of the MLE, or any other estimator?
-Let us distinguish two lines of work addressing this question in statistic.
-(1.) \emph{Parameter estimation} has focused on the quality of the resulting parameters, using norms or distances between parameter themselves, and yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
-(2.) \emph{Density estimation} has focused on measuring the discrepancy between the model and the true distribution.
+In the second, we use KL"}
+Once we have an estimator $\hat \nat$, we wish to assess its quality:
+how close is it from some optimal $\nat^*$.
+This is necessary, for instance to choose between 2 estimators (model selection).
+We distinguish two ways of doing that.
+{\bf 1)} distance in parameter space.
+This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
+{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
 For this purpose, the Kullback-Leibler (KL) divergence arises naturally from information theory,
-but its lack of robustness to misspecification has led statisticians to study other distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}.
+but its lack of robustness to misspecification\footnote{
+$\KL(p||q) = +\infty$ if the model $q(x)=0$ for some $x$ from the true distribution$p(x)>0$.
+}
+has led statisticians to study other distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction}.
 As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research. 
-
-\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution.
-Hence the natural question:}
-% In this work, we are addressing the following question.
+Hence the natural question:
+
+%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution.
+}
+% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
 
@@
 
-If $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
-or close to quadratic, there are simple upper bounds.
-\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
-But without further assumptions, this problem is difficult.
-
-\textcolor{red}{My main problem with the following paragraph here is that its at the same time too detailed for the intro, but also not enough detailed to understand it.}
-
-This question has connections with optimization.
-\textcolor{red}{I would start with "Consider the problem $\min_{\theta\in \Theta} f(\theta)$ (write the equation in display).
-Setting $f=...$, we obtain that KL is the measure of suboptimality, i.e., (equation (3)).
-Moreover, we have that MAP solve a regularized version of f, (write equation) and MLE is the special case where (regularisation = 0)"}
-For one, the MLE is minimizing $f$ (or a regularized version for MAP), and the KL is the sub-optimality
+We know asymptotic answers. 
+We also know simple upper bounds if $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
+or close to quadratic (see \S\ref{ssec:quadratic}).
+But in general, this problem is difficult. In this paper, we show some partial solutions, and give some idea on how to solve it.
+%\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
+%RLP poroblem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.
+
+%\textcolor{red}{My main problem with the following paragraph here is that its at the same time too detailed for the intro, but also not enough detailed to understand it.}
+%RLP right, not sure about how to fix this. It is hard to decide on the right amount of space for optimization versus statistics.
+
+One lens on this question is stochastic optimization.
+Consider the problem
 \alignn{
-f(\nat) - f(\nat^*) = \KL\paren{p(\cdot \cond \nat^*) || p(\cdot \cond \nat) }
+	\min_{\theta\in \Theta} f(\theta)
+	\label{eq:optimization_problem}
+}
+solved by $\nat^*\in \Theta$.
+Setting $f$ as in~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
+\alignn{
+	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
 	\label{eq:suboptimalityKL}
 }
-where $\nat^* \in \argmin f(\nat)$.
-For two, in exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
+Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
+In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
 A convergence rate for this algorithm would translate into the upper bound we seek.
-%similarly to other works on stochastic approximation \citep{bach2013nonstronglyconvex}. %TODO check this
-From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g.
+%From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g.
 not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
-In fact we can see this problem as a very special case to study SMD, revealing the limits of current analysis.
+Conversely we can see this problem as a very special case to study SMD, revealing the limits of current analysis.
 
@@
 Also, when explicit, we write the random variable $T = T(X)$.
-
 % TODO sort out relationships between : steep family, minimal statistics and Legendre type. What implies what ?
@@
 \subsection{Quadratic Case}
+\label{ssec:quadratic}
 As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
@@
 	\label{eq:MAP_rate}\\
-	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
+	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
 \end{align}

### 2021-10-14 aistats.tex

@@
 where  $\nat$ is called natural (or primal) parameter.
-It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) a base measure$\nu$ on $\cX$.
+It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic,
+and 2) a base measure $\nu$ on $\cX$.
 Since the exponential is positive, $p$ has the same support as $\nu$.
-The function $\logpart$ is called the log-partition function that acts as a normalization term, since
-\begin{align}
-    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
-\end{align}
-
+The log-partition function $\logpart$ acts as a normalization term, since
+\begin{align}
+    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \;.
+\end{align}
 This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
 
-For convenience, in this paper we focus on steep, regular exponential families with minimal statistic $T$.
+For convenience, we focus on steep, regular exponential families with minimal statistic $T$.
 Then $\logpart$ is a strictly convex function of Legendre type,
@@
 \citep{barndoffnielsen2014information}.
-Also, when explicit, we write the random variable $T = T(X)$.
+When explicit, we write the random variable $T = T(X)$.
+
 % TODO sort out relationships between : steep family, minimal statistics and Legendre type. What implies what ?
@@
 \end{align}
-which matches the common notion of \textit{entropy} in information theory.
-By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}). More concisely,
+which is the common notion of \textit{entropy} in information theory.
+By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$
+(see Fig.~\ref{fig:duality}), giving
 \aligns{
@@
 }
-
 \begin{figure}[t]
@@
 	\label{fig:duality}
-\end{figure}
-
-\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
-We now introduce the Bregman divergence induced by $\logpart$, that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
+\end{figure}%
+%
+%\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
+% We now introduce the Bregman divergence induced by $\logpart$,
+% that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
+The Bregman divergence induced by $\logpart$,
+measures the discrepancy between two parameters $\nat$ and $\nat_0$,
 \begin{align}
@@
 with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
-In general, bregman distances are non-symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
+In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
 Finally, it is equal to the divergence of its convex conjugate with switched arguments,
 \begin{align}
-	\bregman (\nat ; \nat^*)
-    = \bregmanconj ( \meanp^* ; \meanp) \; .
+	\bregman (\nat ; \nat_0)
+    = \bregmanconj ( \meanp_0 ; \meanp) \; .
     \label{eq:bregman_switch}
@@
 }
-which we recognize as the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$ and step-size schedule $\lr_n$ initialized at $\nat_0$.
+which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$
+and step-size schedule $\lr_n$, initialized at $\nat_0$.
 \emph{MAP is an instantiation of SMD.}
@@
 as it does not have any initialization to start from.
+\fdk{It does, you just forget the initial condition after the first step with step-size 1?}
 
 \section{Open Problem}
-Let us assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
-We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
-As we will see, this is already an interesting setting.
-We wish to estimate how well the MLE or a MAP model the true distribution.
+% Assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
+% We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
+% As we will see, this is already an interesting setting.
+% We wish to estimate how well the MLE or a MAP model the true distribution.
+
+Assume we observe a dataset $\mD$ drawn i.i.d. from an exponential family distribution
+with parameters $\nat^*$, $p(\cdot \cond\nat^*)$.
+We wish to quantify how well the MLE or a MAP models the true distribution.
 
@@
 }
-
-How does this quantity behave when $\nat$ is the MLE or a MAP ? 
-For instance can we get upper bounds on the expected KL ?
+How does this quantity behave when $\nat$ is the MLE or MAP?
+Can we get upper bounds on the expected KL?
 \begin{align}
 	\label{eq:bregmanMLE}
-	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
+	\text{MLE: } &\quad \expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
 	\label{eq:bregmanMAP}
-	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
-\end{align}
-where the outer expectation is on the dataset $\mD$.
-More explicitly, can we write an upper bound that does not involve this expectation over the dataset?
-Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators, when the statistical loss is given by the KL divergence ?
+	\text{MAP: } &\quad \expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
+\end{align}
+where the expectation is on the data $\mD$.
+More explicitly, is there an upper bound that does not involve this expectation over the dataset?
+Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators
+when the loss is the KL divergence?
 
 Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
-There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights}.
-There are also results for special cases, such as $\cN(0, \sigma^2)$ (See \S\ref{sec:example}). 
-But we are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n \lesssim d$.
+There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights},
+or results for special cases such as $\cN(0, \sigma^2)$ (see \S\ref{sec:example}).
+But we are yet to find a solution encompassing a broad range of exponential families,
+and applicable to small sample sizes $n \lesssim d$.
 
@@
 While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite,
-for instance in Gaussian covariance estimation when $n \leq d + 1$.
-Even worse, with categorical variables, there is always a non-zero probability to never sample one of the categories.
-In such the MLE will give zero weight to this category and $KL(p_{\nat^*} || p_\text{MLE} ) = +\infty$.
-This means that the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
-Instead of taking the expectation, one might want to bound the risk in high probability, without resorting to Markov inequality but this is a difficult endeavor.
-These examples are rather making a case for regularized estimators such as MAP, for which we may find interesting upper bounds.
+for instance when estimating the covariance of a Gaussian when $n \leq d + 1$.
+Even worse, with categorical variables, there is a non-zero probability to never sample one of the categories.
+In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$,
+and the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
+Instead of taking the expectation, one might want to bound the risk in high probability,
+without resorting to Markov inequality but this is a difficult endeavor.
+These examples make a case for regularized estimators such as MAP,
+for which we may find upper bounds.
 
@@
 As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
-For instance, this is the log-partition of a Gaussian with variance $I$ and unknown mean 
+For instance, this is the log-partition of a Gaussian with known variance $I$,
 \[
-	\cX=\real^d,\quad \nu(dx) = \exp({\textstyle \half[-\|x\|^2]})dx,\quad T(x)=x.
+	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
 \]

### 2021-10-14 aistats.tex
Pass over contributions

@@
 \section{Motivation}
-
-Exponential families are among the most widely used statistical models.
+\label{sec:motivation}
+
+{\bf Exponential families} are among the most widely used statistical models.
 Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
@@
 
+{\bf Estimators.}
 In general, we want to estimate $\nat$ from some data $\mD = (x_1, \dots, x_n)$.
@@
 In the second, we use KL"}
+{\bf Metrics.}
 Once we have an estimator $\hat \nat$, we wish to assess its quality:
@@
 We distinguish two ways of doing that.
-{\bf 1)} distance in parameter space.
+{\bf 1)} distance in parameter space $d(\nat,\nat^*)$.
 This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
 {\bf 2)} distance between distributions, as studied in \emph{density estimation}.
-For this purpose, the Kullback-Leibler (KL) divergence arises naturally from information theory,
+For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
 but its lack of robustness to misspecification\footnote{
@@
 
-\begin{center}
-	\framebox{
-		\parbox[c][]{0.9\linewidth}{
-			\begin{center}
-			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
-		\end{center}
-		}
-	}
-\end{center}
+%\begin{center}
+%	\framebox{
+%		\parbox[c][]{0.9\linewidth}{
+%			\begin{center}
+%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
+%		\end{center}
+%		}
+%	}
+%\end{center}
+
+\begin{equation}
+\boxed{
+\begin{aligned}
+	\textit{Can we upper}&\textit{ bound  the expected value of} \\
+	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} })?
+\end{aligned}
+}
+\tag{$\star$}
+\label{problem}
+\end{equation}
+% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.
 
@@
 
-One lens on this question is stochastic optimization.
+{\bf Stochastic optimization} offers an interesting perspective on this question.
 Consider the problem
@@
 solved by $\nat^*\in \Theta$.
-Setting $f$ as in~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
+Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
 \alignn{
@@
 
-\paragraph{Contributions}
-After formalizing the problem \textcolor{red}{restate the problem} along with its asymptotic properties, we make the following contributions.
-\textcolor{red}{Add ref to the main text in the contribution.}
-\begin{itemize}
-	\itemsep0em
-	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation is valid. \textcolor{red}{Approx. of what?}
-	\item We investigate a generalized bias-variance decomposition~\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem} \textcolor{red}{Decomposition of what?}
-	\item We provide a full solution \textcolor{red}{full solution of what?} in the case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$.
-	\item We establish a connection \textcolor{red}{Connection between something (you need to precise what) and optimization} with optimization, and use this problem to highlight areas for progress in the stochastic optimization of barrier objectives.
-\end{itemize}
-
 \textcolor{red}{I would emphasize more that you are talking about the impact of the work rather than contribution.
@@
 It could also help develop accelerated optimization methods.
+
+\paragraph{Contributions}
+After formalizing the problem of upper bounding the KL of MAP estimators (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
+\begin{itemize}
+	\itemsep0em
+	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
+	\item We investigate a generalized bias-variance decomposition of the KL\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem} (\S\ref{ssec:bias-variance})
+	\item We answer~\eqref{problem} with an upper bound in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
+	\item We establish a connection between~\eqref{problem} and optimization, and use it to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
+\end{itemize}
 
@@
 \section{Technical background}
+\label{sec:background}
 In this section, we review the formalism of exponential families, their duality, a conjugate prior and the corresponding MAP.
@@
 \end{align}
-which reaches its maximum attempts $\MAPt^\text{MAP}$ such that
+which reaches its maximum at $\MAPt^\text{MAP}$ such that
 \begin{align}
@@
 \section{Open Problem}
+\label{sec:problem}
 % Assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
@@
 	\label{eq:bregmanMLE}
-	\text{MLE: } &\quad \expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
+	\text{MLE: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
 	\label{eq:bregmanMAP}
-	\text{MAP: } &\quad \expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
+	\text{MAP: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
 \end{align}
@@
 \subsection{Asymptote}
+\label{ssec:asymptote}
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
@@
 \subsection{Locally Quadratic Case}
+\label{ssec:local-quadratic}
 FromTaylor expansion~~\eqref{eq:bregmanTaylor},
@@
 \subsection{Bias-Variance Decomposition}
+\label{ssec:bias-variance}
 In both examples \textcolor{red}{(add link)}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
@@
 
-\section{Gaussian Variance Example}\label{sec:example}
+\section{Examples}\label{sec:example}
+\subsection{Gaussian Variance}\label{ssec:gaussian-variance}
+
 A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
@@
 \end{figure*}
+
+\subsection{Gaussian}
+\label{ssec:gaussian}
+The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
 

### 2021-10-14 aistats.tex
Write about impact

@@
 Hence the natural question:
-
 %\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution.
@@
 % RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
-
 %\begin{center}
@@
 	\textit{Can we upper}&\textit{ bound  the expected value of} \\
-	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} })?
+	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \ ?
 \end{aligned}
@@
 or close to quadratic (see \S\ref{ssec:quadratic}).
-But in general, this problem is difficult. In this paper, we show some partial solutions, and give some idea on how to solve it.
+But in general, this problem is difficult. In this paper, we show some partial solutions, and give some ideas on how to solve it.
 %\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
@@
 
-\textcolor{red}{I would emphasize more that you are talking about the impact of the work rather than contribution.
-Also I think its better to write this before the contribution, to motivate better why we did this work.
-}A solution to this problem would allow the field of optimization tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression).
-It could also help develop accelerated optimization methods.
-
-\paragraph{Contributions}
-After formalizing the problem of upper bounding the KL of MAP estimators (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
+{\bf Impact.}
+While a solution to the fundamental problem~\eqref{problem} would be of independent interest -- e.g.
+assessing the importance of the prior in MAP with few samples-- we believe it would have repercussions to other fields.
+For instance, it could enable stochastic optimization to tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression) \citep{bach2013nonstronglyconvex}.
+%Or it could help developing accelerated (stochastic?) algorithms for barrier objectives  overcoming the limitations of non-Euclidean geometry \citep{dragomir2021optimal}.
+\rlp{Fred I need your feeedback on that one.}
+
+% outcomes
+% quantifying the importance of the prior, and few sample behavior, for instance in Gaussians
+% stochastic algorithms for barrier objectives
+% such as a lot of GLM
+% accelerated methods ?
+
+{\bf Contributions.}
+After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
 \begin{itemize}
@@
 
-\paragraph{Notation}
-$X,\,T(X)$ and $T$ \textcolor{red}{Don't we have T = T(X), but written tuch that it's simpler?
-I think its better to say it that we write T instead of T(X) for simplicity, rather than saying its a random variable.} are random variables, while $x$ is a sample.
+{\bf Notation.}
+$X$ and $T=T(X)$ are random variables, while $x$ is a sample.
 $n$ is the number of samples and $d= \dim(T)$.

### 2021-10-14 aistats.tex
Re-write impact

@@
 For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
-but its lack of robustness to misspecification\footnote{
-$\KL(p||q) = +\infty$ if the model $q(x)=0$ for some $x$ from the true distribution$p(x)>0$.
+but its lack of robustness to misspecification\footnote{For $p$ and $q$ continuous densities of the ground truth and a model,
+$\KL(p||q) = +\infty$ if $q(x)=0$ for some $x$ such that $p(x)>0$.
 }
@@
 As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
-Hence the natural question:
+Hence the fundamental question:
 %\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution.
@@
 
-{\bf Impact.}
-While a solution to the fundamental problem~\eqref{problem} would be of independent interest -- e.g.
-assessing the importance of the prior in MAP with few samples-- we believe it would have repercussions to other fields.
-For instance, it could enable stochastic optimization to tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression) \citep{bach2013nonstronglyconvex}.
+{\bf Expected Ouctomes.}
+A solution to~\eqref{problem} would quantify the importance of the prior in MAP, in particular in the few sample regime, for instance for Gaussians $\cN(m,\sigma^2)$. 
+More generally it could enable stochastic optimization to tackle a wide class of barrier objectives\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}.
+A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
+It could also help in assessing the impact of alternative forms of regularization (prior) for these models.
+
 %Or it could help developing accelerated (stochastic?) algorithms for barrier objectives  overcoming the limitations of non-Euclidean geometry \citep{dragomir2021optimal}.
-\rlp{Fred I need your feeedback on that one.}
-
-% outcomes
-% quantifying the importance of the prior, and few sample behavior, for instance in Gaussians
-% stochastic algorithms for barrier objectives
-% such as a lot of GLM
-% accelerated methods ?
+\rlp{Fred I need your feedback on that one.}
 
@@
 The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.
+This is true for categorical or Dirichlet distributions.
 

### 2021-10-14 aistats.tex
Updates from Overleaf

--- 2021-10-14 aistats.tex
@@
 \documentclass[twoside]{article}
-
-\usepackage{silence}
-% Disable all warnings issued by latex starting with "You have..."
-% This is a hack, there is no clean solution for that unless we install locally the package
-\WarningFilter{latex}{You have requested package}
-
-\pdfsuppresswarningpagegroup=1 %removes the warning
-% This is a hack, the other way to remove is to convert all pdf into postscript or eps.
-
 \usepackage{aistats/aistats2022}
@@
 
-\setlength{\footskip}{3.31pt}
-
 % If you set papersize explicitly, activate the following three lines:
@@
 \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{lmodern}
 \usepackage{hyperref}       % hyperlinks
@@
 	pdfauthor={Anonymous},
-	% draft,
-	final,
 }
@@
 \newcommand{\nat}{\theta}
-\newcommand{\m}{\mu}
+\newcommand{\m}{m}
 \newcommand{\meanp}{\m}
@@
 \DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}
+
+%\hypersetup{draft}
 
@@
 
-\aistatstitle{Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem}
+\aistatstitle{Convergence Rates for the MAP of an Exponential Family, and Stochastic Mirror Descent -- an Open Problem}
 
@@
 \begin{abstract}
-We consider the problem of upper bounding the expected log-likelihood sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family, in a non-asymptotic way.
-Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
-After exhibiting various facets of the problem, we show the MAP can be interpreted as running stochastic mirror descent (SMD) on the log-likelihood,
-yet modern convergence results do not apply for standard examples of exponential family, highlighting holes in the convergence literature.
+We consider the problem of upper bounding the expected sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family.
+Surprisingly, we found no general solution to this problem in the literature.
+In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
+After exhibiting various facets of the problem, we show the MAP can be interpreted as the result of Stochastic Mirror Descent (SMD),
+yet it falls out of scope for modern analysis.
 We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
@@
 \section{Motivation}
-\label{sec:motivation}
-
-{\bf Exponential families} are among the most widely used statistical models.
-Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
-They are flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
-In particular, they are appreciated for their convex log-likelihood
-\alignn{
-f(\nat) = \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
-\label{eq:defNLL}
-}
-where $\logpart$ is a convex function and $\nat\in|Theta$ is a vector called \textit{natural parameter}.
-This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized}
-or variants of principal component analysis \citep{collins2001generalization}, among other applications.
-
-{\bf Estimators.}
-In general, we want to estimate $\nat$ from some data $\mD = (x_1, \dots, x_n)$.
-For this purpose, not only is $f$ convex, but it even enjoys a closed form maximum-likelihood estimate (MLE)
-\begin{align}
-	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
-	\label{eq:defMLE}
-\end{align}
-This rule is known as moment matching.
-Given a conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
-%\textcolor{red}{DS: I know that for conciseness we do not write the MAP estimate here, but it's super central and I think its important to write the equation. WDYT?}
-% I think it's okay to have it only later, and refernece it here.
-
-%\textcolor{red}{This comes a bit from nowhere.
-Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research.
-We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution.
-In the first case, the analysis use norms blabla.
-In the second, we use KL"}
-{\bf Metrics.}
-Once we have an estimator $\hat \nat$, we wish to assess its quality:
-how close is it from some optimal $\nat^*$.
-This is necessary, for instance to choose between 2 estimators (model selection).
-We distinguish two ways of doing that.
-{\bf 1)} distance in parameter space $d(\nat,\nat^*)$.
-This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
-{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
-For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
-but its lack of robustness to misspecification\footnote{
-$\KL(p||q) = +\infty$ if the model $q(x)=0$ for some $x$ from the true distribution$p(x)>0$.
-}
-has led statisticians to study other distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction}.
-As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
-Hence the natural question:
-
-%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution.
-}
-% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
-
-%\begin{center}
-%	\framebox{
-%		\parbox[c][]{0.9\linewidth}{
-%			\begin{center}
-%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
-%		\end{center}
-%		}
-%	}
-%\end{center}
-
+
+We kind of know how to do optimization
+on smooth, strongly convex functions
+when there is some bound on the stochasticity,
+for example in the form of a bounded variance,
+say for the problem
+\begin{equation} %We might want to add a feasible set for theta?
+\min_\theta f(\theta)
+\end{equation}
+with access to stochastic gradients
+$g(\theta)$ such that
 \begin{equation}
-\boxed{
-\begin{aligned}
-	\textit{Can we upper}&\textit{ bound  the expected value of} \\
-	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} })?
-\end{aligned}
-}
-\tag{$\star$}
-\label{problem}
+	\expect{\norm{g(\theta) - \nabla f(\theta)}^2} \leq \sigma^2.
 \end{equation}
-% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.
-
-We know asymptotic answers.
-We also know simple upper bounds if $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
-or close to quadratic (see \S\ref{ssec:quadratic}).
-But in general, this problem is difficult. In this paper, we show some partial solutions, and give some idea on how to solve it.
-%\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
-%RLP poroblem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.
-
-%\textcolor{red}{My main problem with the following paragraph here is that its at the same time too detailed for the intro, but also not enough detailed to understand it.}
-%RLP right, not sure about how to fix this. It is hard to decide on the right amount of space for optimization versus statistics.
-
-{\bf Stochastic optimization} offers an interesting perspective on this question.
-Consider the problem
-\alignn{
-	\min_{\theta\in \Theta} f(\theta)
-	\label{eq:optimization_problem}
-}
-solved by $\nat^*\in \Theta$.
-Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
-\alignn{
-	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
-	\label{eq:suboptimalityKL}
-}
-Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
-In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
-A convergence rate for this algorithm would translate into the upper bound we seek.
-%From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g.
-not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
-Conversely we can see this problem as a very special case to study SMD, revealing the limits of current analysis.
-
-\textcolor{red}{I would emphasize more that you are talking about the impact of the work rather than contribution.
-Also I think its better to write this before the contribution, to motivate better why we did this work.
-}A solution to this problem would allow the field of optimization tackle a wider class of generalized linear models than fixed variance gaussians (linear regression) and Bernoulli (logistic regression).
-It could also help develop accelerated optimization methods.
+Estimating the mean of Gaussian noise is a special case of this problem class, using
+$f(\theta) = \frac{1}{2}\norm{\theta - \mu}^2$ and $g(\theta) \sim \cN(- \mu, \sigma^2)$.
+SGD with a decreasing step-size coincides with the maximum likelihood estimate (MLE),
+and attains the $1/t$ convergence rate.
+
+But our understanding of stochastic optimization lacks in other domains.
+A generalization of the above problem, at the intersection of optimization and statistics,
+is the problem of minimizing
+\begin{equation}
+	f(\theta) = \expect[x \sim p(x\cond\theta_*)]{-\log p(x\cond\theta)},
+\end{equation}
+When $p(x\cond\theta)$ is an exponential family distribution,
+$-\log p(x\cond\theta) = A(\theta) - \lin{t(x), \theta}$ (omitting the reference distribution).
+Here, $A$ is the log-partition function, a Legendre function that we assume can be minimized in closed form.
+The Gaussian example is the special case $A(\theta) = \frac{1}{2}\norm{\theta}^2$.
+
+Typical works on parameter estimation bound the deviation with norms -- e.g. bounding $\expect[p(x\cond\theta_*)]{\norm{\theta-\theta_*}^p}$ for some norm and some power. \tocite
+Or more generally, they  use distances rather than divergences as they are symmetric and usually behave smoothly. \tocite 
+A contrario the KL is not smooth, and it can be infinite in misspecified settings.
+But as the MLE is the minimizer of the log-likelihood, we wish to bound the sub-optimality.
+
+There are algorithms for this setting, for example the MLE -- the (online) maximum likelihood after seeing $t$ samples -- or adding regularization, the MAP -- the maximum a posteriori.
+As we show, the MAP can be interpreted as online mirror descent with decreasing step-sizes.
+Despite recent advances in this field, our convergence rates are still lacking; they are either asymptotic, cover only special cases or require having seen many ($n \gg d$) samples.
+
+Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
+\[
+R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
+\]
+where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
+Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.
+Unfortunately however, their rate are worse than the offline MAP or MLE algorithms.
+For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$.
+From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
+Also, \citet{dasgupta2007online} proves the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
+This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
+
+We do not have a solution. The goal of this paper
+is to bring this issue to the attention of the community working at the intersection of machine learning, optimization and statistics,
+in the hope that we can make progress on this question;
+Can we build algorithms for stochastic estimation
+when the objective function is not (close to) quadratic
+and the noise is not Gaussian?
+
+We are aware that \emph{a} solution to this problem would be the maximum likelihood estimate for which there is a wealth of asymptotic and large sample results \citep{van2000asymptotic,ostrovskii2021finite}.
+%TODO complete
+However, none of the theory we found could address for instance the case of a Gaussian with unknown mean and variance $\cN(\mu,\sigma^2)$, hence our contributions.
+%TODO kind of redundant "problem - question - problem" paragraphs, not disentangling stats and optim. To improve
 
 \paragraph{Contributions}
-After formalizing the problem of upper bounding the KL of MAP estimators (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
 \begin{itemize}
-	\itemsep0em
-	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
-	\item We investigate a generalized bias-variance decomposition of the KL\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem} (\S\ref{ssec:bias-variance})
-	\item We answer~\eqref{problem} with an upper bound in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
-	\item We establish a connection between~\eqref{problem} and optimization, and use it to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
+	\item We provide insights on different facets of this problem : quadratic approximations ; bias variance decomposition; and a detailed analysis of the Gaussian variance estimation problem.
+	\item We show the connections between the optimization and statistical estimation view of the problem and discuss where current theories fail to fully capture the problem.
 \end{itemize}
 
+The statistics and optimization perspective are so tightly linked in this problem, that we believe progress from one field would help the other one.
+Progress from statistics could bring forward research in non-euclidean optimization, or the optimization of barrier losses, for instance helping in the development of accelerated methods.
+On the other hand, optimization can become  a more versatile and general tool for statisticians. % handwavy and shady af.
+
+% Then;
+% - description of the setting
+% - statistical solutions (asymptotic, l1 or l2 error)
+% - optimization solutions (online+bounded gradient, self-concordance, more recent relative smoothness results)
+% - potential application of ideas that would solve this, beyond this problem (acceleration, causality?)
+% - conclusion?
+
 \paragraph{Notation}
-$X,\,T(X)$ and $T$ \textcolor{red}{Don't we have T = T(X), but written tuch that it's simpler?
-I think its better to say it that we write T instead of T(X) for simplicity, rather than saying its a random variable.} are random variables, while $x$ is a sample.
-$n$ is the number of samples and $d= \dim(T)$.
 $\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
@@
 \section{Technical background}
-\label{sec:background}
-In this section, we review the formalism of exponential families, their duality, a conjugate prior and the corresponding MAP.
-We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a more exhaustive introduction.
-
-The density of an exponential family for a sample $x$ is
+
+Exponential families are to statistics what linear models are to regression.
+Many classical model are exponential families, for instance the Gaussian or Bernoulli distributions.
+Their convexity or modeling flexibility make them one of the most widely used statistical models.
+In this section, we review some of their properties.
+We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a full review.
+
+The exponential family for data $X\in \cX$ reads
 \begin{equation}
-	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
+	 p(X|\nat) = \exp( \langle \nat, T(X) \rangle - \logpart(\nat)) \; ,
 	 \label{eq:def_expfamily}
 \end{equation}
-where  $\nat$ is called natural (or primal) parameter.
-It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic,
-and 2) a base measure $\nu$ on $\cX$.
-Since the exponential is positive, $p$ has the same support as $\nu$.
-The log-partition function $\logpart$ acts as a normalization term, since
-\begin{align}
-    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \;.
-\end{align}
+where  $\nat$ is called natural (or primal) parameter is specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic, and 2) $\nu(dx)$, the density with respect to a base measure.
+The function $\logpart$ is called the log-partition function that acts as a normalization term, since
+\begin{align}
+    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \; .
+\end{align}
+
 This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
 
-For convenience, we focus on steep, regular exponential families with minimal statistic $T$.
-Then $\logpart$ is a strictly convex function of Legendre type,
-and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex
-\citep{barndoffnielsen2014information}.
-When explicit, we write the random variable $T = T(X)$.
+For convenience, in this paper we focus on steep, regular exponential families
+where $\logpart$ is strictly convex function of Legendre type,
+and the set $\nat \in \Theta$ such that $\logpart(\nat) < \infty$ is open and convex
+(See \citet{barndoffnielsen2014information}). Also, when explicit, we write the random variable $T = T(X)$.
 
@@
 \paragraph{Duality}
-The log-partition function $\logpart$ verifies the two following identities:
-\begin{align}
-    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp \\
-    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0
+The logpartition function $\logpart$ verifies the two following identities:
+\begin{align}
+    \nabla\logpart(\nat) &=  \expect[p(T|\nat)]{T} =: \meanp \\
+    \nabla^2 \logpart(\nat) &= \Cov_\nat[T] \geq 0
 \end{align}
 where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
-Given that $\logpart$ is strictly convex, its Hessian is positive definite and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
-We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both are linked and represent the same distribution.
-
-We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
-\begin{align}
-	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
-	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
-\end{align}
-which is the common notion of \textit{entropy} in information theory.
-By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$
-(see Fig.~\ref{fig:duality}), giving
-\aligns{
+We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both represent the same distribution.
+
+We now introduce the convex conjugate (a.k.a. Fenchel-Legendre transform) of the logpartition function
+\begin{align}
+	\conj(\m) = \min_{\nat\in\Theta}\langle \m, \nat \rangle - \logpart(\nat) \; ,
+\end{align}
+which matches the common notion of \textit{entropy} in information theory.
+In this paper, we assume that the sufficient statistic $T$ is minimal, e.g., its covariance is definite positive for all $\nat$.
+In such case, the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$, and moreover, $\nabla\conj=\nabla\logpart^{-1}$ (cf Fig.~\ref{fig:duality}).
+More concisely,
+\[
 	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
-}
-\begin{figure}[t]
+\]
+
+% The entropy is eating mean parameters, and its gradient is mapping back to natural parameters $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).
+
+\begin{figure}[t] % @remi In the figure you use mu instead of m??
 	\centering
@@
 	\label{fig:duality}
-\end{figure}%
-%
-%\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
-% We now introduce the Bregman divergence induced by $\logpart$,
-% that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
-The Bregman divergence induced by $\logpart$,
-measures the discrepancy between two parameters $\nat$ and $\nat_0$,
+\end{figure}
+
+\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps)
+We now introduce the Bregman divergence induced by $\logpart$, that measure the distance between two parameters $\nat$ and $\nat_0$,
 \begin{align}
@@
     - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
-    \label{eq:defBregman}
 \end{align}
 with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
-In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
+In general, bregman distances are non-symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
 Finally, it is equal to the divergence of its convex conjugate with switched arguments,
@@
     = \bregmanconj ( \meanp_0 ; \meanp) \; .
-    \label{eq:bregman_switch}
 \end{align}
@@
 where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
-This is an exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and natural parameter $(n_0 \m_0, -n_0)$.
-Intuitively, $n_0$ is the number of fictive data points observed from a distribution with natural parameter $\nat_0$.
+This is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \m_0, -n_0)$.
+Intuitively, $n_0$ is the number of fictive data points observed from a distribution with parameter mean $\nat_0$.
 
 \paragraph{Maximum A Posteriori (MAP).} 
-Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD|\nat)p(\nat)$.
-Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
+\textcolor{red}{\textbf{D.S.: This section is very unclear. }}
+Using Bayes rule with~\eqref{eq:def_expfamily}~\&~\eqref{eq:def_prior}, the posterior of $\nat$ given a dataset $\mD_n =(X_1,\dots,X_n)$ is
 \begin{align}
 	p(\nat \cond \mD_n)
+	%\propto p(\mD|\nat)p(\nat)
     \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
@@
 \end{align}
-which reaches its maximum at $\MAPt^\text{MAP}$ such that
+reaching its maximum in $\MAPt^\text{MAP}$ such that
 \begin{align}
@@
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
-    \label{eq:defMAP}
 \end{align}
 where $T_i=T(X_i)$.
-When $n_0=0$ -- we have no samples from the prior -- we recover the MLE~\eqref{eq:defMLE}.
-We write $\MAPt$ for the MAP, and view the MLE as a special case.
-
-\paragraph{Optimization}
-We can rewrite~\eqref{eq:defMAP} as
-\alignn{
-\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
-\label{eq:mean_update}
-}
-where $\lr_n := \inv{n_0 + n}$.
-Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
-If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
-\alignn{
-	\nabla\conj(\hat \nat_{n})
-	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
-}
-which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$
-and step-size schedule $\lr_n$, initialized at $\nat_0$.
-\emph{MAP is an instantiation of SMD.}
-Note that this analogy does not hold for the MLE %($n_0=0$)
-as it does not have any initialization to start from.
-\fdk{It does, you just forget the initial condition after the first step with step-size 1?}
+When $n_0=0$ -- eg we observed zero samples from the prior -- we recover the Maximum Likelihood Estimate (MLE)
+\begin{align}
+	\hat \m_n^\text{MLE} = \frac{\sum_{i=1}^n T_i}{n}
+\end{align}
+This last equation is known as moment matching.
+In general, we will write $\MAPt$ to refer to the MAP, and view the MLE as a very special case.
+The MLE and MAP estimates are statistics of the dataset $\mD_n$.
+Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.
 
 \section{Open Problem}
-\label{sec:problem}
-% Assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
-% We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
-% As we will see, this is already an interesting setting.
-% We wish to estimate how well the MLE or a MAP model the true distribution.
-
-Assume we observe a dataset $\mD$ drawn i.i.d. from an exponential family distribution
-with parameters $\nat^*$, $p(\cdot \cond\nat^*)$.
-We wish to quantify how well the MLE or a MAP models the true distribution.
-
-A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
-In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
-Using~\eqref{eq:defNLL} and~\eqref{eq:bregman_switch}, we can also write the KL with Bregman divergences
-\alignn{
-	\KL(p_{\nat^*} || p_\nat)
+Given a distribution $\cD$, our objective is the expected log-likelihood 
+\begin{align}
+	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
+	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
+\end{align}
+Note that $\cD$ does not have to be part of the exponential family.
+Let $\nat^* := \argmin f(\nat)$.
+Assuming $\nat^*$ exists within $\Theta$ means $\cD$ is such that $\E[T] = \meanp^* \in \cM$.
+Then the suboptimality on the population log-likelihood can be written with Bregman divergences
+\footnote{
+Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
+and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
+}
+\begin{align}
+	 f(\nat) - f(\nat^*)
+	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
 	 = \bregman(\nat ; \nat^*) 
 	 = \bregmanconj(\m^* ; \m) \; .
-}
-How does this quantity behave when $\nat$ is the MLE or MAP?
-Can we get upper bounds on the expected KL?
+\end{align}
+Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
 \begin{align}
 	\label{eq:bregmanMLE}
-	\text{MLE: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
+	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
 	\label{eq:bregmanMAP}
-	\text{MAP: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
-\end{align}
-where the expectation is on the data $\mD$.
-More explicitly, is there an upper bound that does not involve this expectation over the dataset?
-Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators
-when the loss is the KL divergence?
-
-Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
-There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights},
-or results for special cases such as $\cN(0, \sigma^2)$ (see \S\ref{sec:example}).
-But we are yet to find a solution encompassing a broad range of exponential families,
-and applicable to small sample sizes $n \lesssim d$.
-
-{\bf Unboundedness of~\eqref{eq:bregmanMLE}.}
-While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite,
-for instance when estimating the covariance of a Gaussian when $n \leq d + 1$.
-Even worse, with categorical variables, there is a non-zero probability to never sample one of the categories.
-In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$,
-and the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
-Instead of taking the expectation, one might want to bound the risk in high probability,
-without resorting to Markov inequality but this is a difficult endeavor.
-These examples make a case for regularized estimators such as MAP,
-for which we may find upper bounds.
+	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
+\end{align}
+where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
+More explicitly, can we write an upper bound that does not involve this expectation over the dataset.
+
+In general, for arbitrary distributions $\cD$, it is a hard problem, especially since the KL is known for being brittle to misspecification \tocite, so let us focus on the well-specified case first $\cD = p(\cdot \cond \nat^*)$.
+Surely this problem has been solved, hasn't it ?
+In some special cases yes, but it seems the general answer is no. 
+We are yet to find a solution encompassing a broad range of exponential families, and applicable to small sample sizes $n$.
+
+{\bf Remark.}
+This could be seen as a concentration inequality, expressed with a Bregman divergence instead of a norm.
+However there is a connection between the random variable $T(X)$ and the metric $\logpart$.
+Indeed expressions~\eqref{eq:bregmanMLE} or~\eqref{eq:bregmanMAP} can be infinite for another choice of random variable.
+For instance, if we plug in $\conj(\m)= -\log(\m)$, which defines a divergence on positive numbers, and $T(X) \sim \cN(0,1)$ which can be negative.
+
+{\bf Remark 2.}
+The expectation of the MLE may be infinite, for instance with $\cN(0,\sigma^2)$ and $n\leq 2$.
+Instead of taking the expectation,  we might want to bound this quantity in high probability, without resorting to Markov inequality, but that is a difficult endeavor.
 
 \section{Insights}
-\label{sec:insights}
 
 \subsection{Asymptote}
-\label{ssec:asymptote}
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-We now review classical asymptotic results.
-We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
-Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
-With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
+Asymptotic results typically show that natural parameters $\nat$ is asymptotically normal ; reaching the Cramer-Rao lower bound \citep[for instance Ch4.2]{van2000asymptotic}.
+Here the MAP is more simply expressed with mean parameters $\meanp$.
 We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
-\alignn{
+\begin{align}
     \bregmanconj(\m^* ; \m)
-    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
+    &= \frac{\norm{\m^* - \m}^2_{\mF^*}}{2}
     + O(\norm{\m - \m^*}^3),
-    \label{eq:bregmanTaylor}
-}
-where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
-It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
-\aligns{
-    \mF
+\end{align}
+where the Mahalanobis norm  $\| x \|_{\mF^*}^2 = x^\top \mF^* x$  is induced by $\mF^*  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
+It happens that  $F^*$ is also the inverse \textit{Fisher information matrix}, since
+\begin{align}
+    \mF^*
     :=\nabla^2\conj(\m^*)
@@
     = \Cov_{\nat^*}[T(X)]^{-1}  \; .
-}
-Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show that MLE and MAP verify
+\end{align}
+From there, it is possible to show an asymptotic rate for the MLE and the MAP \textcolor{red}{D.S.: How?}
 \begin{align}
@@
 \end{align}
-See proof in the Appendix.
 Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
@@
 \subsection{Quadratic Case}
-\label{ssec:quadratic}
-As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
-For instance, this is the log-partition of a Gaussian with known variance $I$,
-\[
-	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
-\]
+As another reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
+For instance, this is the log-partition of a Gaussian with variance $1$ and unknown mean 
+\[ \cX=\real, \nu(dx) = \exp(\half[-x^2])dx, T(X)=X$).\]
 Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
@@
 \begin{align}
-	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
+	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}.
 	=\frac{\Var(T)}{2n}
-	=\frac{1}{2n}
 \end{align}
@@
 so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
-Note that none of the standard exponential families verify this assumption.
-\TODO{connection with optim non-smooth rate.}
-
-{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
+
+{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex), then
 \begin{align}
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
-Besides the Gaussian with known mean, no standard exponential family verify this assumption either.
-Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
-\TODO{connection with optim smooth rate.}
 
 \subsection{Locally Quadratic Case}
-\label{ssec:local-quadratic}
-FromTaylor expansion~~\eqref{eq:bregmanTaylor},
-we know that all Bregman divergences are locally quadratic.
-Under some assumptions, like self-concordance of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
-\begin{proposition}
-Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
-\aligns{
-	\norm{\m^*-\m}_{\mF} < 0.21
-	\implies
-	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
-}
-\end{proposition}
-Proof in the appendix
-\footnote{In case the reader is intrigued, $0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
-To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
-For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
-For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$ (see appendix).
-This means that on average, we need much more samples than the dimension to reach the quadratic regime.
-
-We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
-This is true for
-exponential distributions,
-Gaussians with known mean,
+\textcolor{red}{Check this! In our case, since we assume $A$ and $A^*$ to be strictly convex,}
+all Bregman divergences are locally quadratic.
+Under some assumptions, like self-concordance \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
+For instance, when $T$ is 1-dimensional, we have that $\logpart$ in the following settings:
+exponential distribution,
+Gaussian with known mean,
 Laplace with known mean,
 Pareto with known minimum value,
-or Weibull with known shape $k$.
-The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.
-
-With some more work, it is possible to get a wealth of results under this assumption.
-For instance \citet{ostrovskii2021finite} characterizes the number of samples needed to reach this quadratic behavior for arbitrary parametric models, under the assumptions of self-concordance, or pseudo self-concordance \citep{bach2010self}.
-Or in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher order moments of $\logpart$ in $\nat^*$, to quantify when $\logpart$ can be sandwiched between quadratics in $\nat$, but this does not directly translate into quadratics in $\m$.
+Weibull with known shape $k$. This also holds when
+-- or when $T$ lives in a compact \citep{bubeck2015entropic}. Then the bound
+\begin{align}
+	 \bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF^*}^2
+\end{align}
+holds whenever $\norm{\m^*-\m}_{\mF^*} < 0.21$.\footnote{The value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
+Similarly, \citet{kakade2010learning} assumes a bound on all higher order moments at $\nat^*$ to quantify when $\logpart$ starts behaving quadratically, but their rate does not directly apply to $\conj$ and $\m$.
+
+\TODO{Reference \citet{ostrovskii2021finite}.} ( as well as
+\citet{anastasiou2017bounds},
+\citet{marteauferey2019beyond})
 
@@
 
-\citet{anastasiou2017bounds} and
-\citet{marteauferey2019beyond})
-\fdk{
-We know that if the problem is strongly convex (in $\norm{\cdot}_2$), averaged SGD achieves $O(1/n)$ rate.
-[Robust stochastic approximation approach to stochastic programming].
-(The paper uses a bounded stochastic gradient assumption, but can be relaxed to bounded variance instead.) However, strong-convexity in the Euclidean norm is not strictly necessary to find an algorithm with convergence rate of $O(1/n)$, as shown by [Non-strongly-convex smooth stochastic approximation with convergence rate O(1/n)] for linear and logistic regression.
-Their proof techniques based on self-concordance have been extended to other setting, [Finite-sample analysis of M-estimators using self-concordance, Learning Exponential Families in High-Dimensions: Strong Convexity and Sparsity].
-The same tools are used in statistics, and evething essentially boils down to controlling how different we are from a Gaussian rv, that we're going to get by the CLT, like [Bounds for the normal approximation of the maximum likelihood estimator, Anastasiou].
-}
-
-\subsection{Bias-Variance Decomposition}
-\label{ssec:bias-variance}
-In both examples \textcolor{red}{(add link)}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
-In fact, such a decomposition exists for any Bregman divergence.
-Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be its corresponding mean parameter.
-As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes into
-\begin{equation}
-	\expect{\bregmanconj(\m^* ; \hat \m_n)} = {\bregmanconj(\m^* ; \tilde \m_n)}
+\subsection{Bias-Variance Decomposition} % DS: Arrived here
+In both examples (add link), the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
+In fact, it is possible to write such a decomposition for any Bregman divergence
+Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be the corresponding mean parameter.
+As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes like
+\begin{align}
+	\expect{\bregmanconj(\m^* ; \hat \m_n)}
+	&= {\bregmanconj(\m^* ; \tilde \m_n)}
 	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
 	\label{eq:primal_pivot}
-\end{equation}
+\end{align}
 
@@
 An estimator will be unbiased if its primal expectation is non-zero.
-However, this is not true for the the gaussian variance MLE, where the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
+This is not true for the MLE. In fact for the gaussian variance MLE, the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
 A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.
@@
 \paragraph{Illustrations.}
-We plot the bias-variance decompositions for $\cN(0,\sigma^2)$ (Figure~\ref{fig:variance_decomposition}) and  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
-Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:bias-variance-numerical} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) .
-\rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}
+We show these decompositions for $\cN(0,\sigma^2)$ in Figure~\ref{fig:variance_decomposition} and  $\cN(\mu, \sigma^2)$ in Figure~\ref{fig:gaussian_decomposition}.
+In particular for $\cN(\mu, \sigma^2)$, we illustrate the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters)  in Figure~\ref{fig:bias-variance-numerical}.
+\rlp{Tune figures.}
 
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{figs/gaussian_variance.pdf}
+	\includegraphics[width=.4\textwidth]{{figs/gaussian_variance.pdf}}
 	\caption{
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
+	\includegraphics[width=.4\textwidth]{{figs/gaussians/new_linear_n0=1}.pdf}
 	\caption{
@@
 
-\section{Examples}\label{sec:example}
-\subsection{Gaussian Variance}\label{ssec:gaussian-variance}
-
-A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
-Its density reads
-\[
-	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
-\]
+\section{Examples}
+
+\subsection{Gaussian Variance}
+A core example of this paper is a centered gaussian with unknown variance $\cN(0,\sigma^2)$.
+The density is	$p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}$.
 Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
-Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
-\textcolor{red}{The explanation here is not clear enough, or too concise} Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
-\begin{align}
-	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
+Mean and natural parameters are roughly inverse of each other $\nat = -\inv{2 \m}$.
+Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
+\begin{align}
+	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi) \\
 	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
@@
 Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
-It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
+It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, eg a negative Gamma distribution.
 It also means $\bregman$ and $\bregmanconj$ have the same shape
@@
 	\bregmanconj( \m_*; \m_n)
-	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
+	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) \\
 	\bregman( \nat_n; \nat_* )
@@
 \end{align}
-Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
+In other words, this divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function $\phi$
 \begin{align}
@@
 illustrated in Figure~\ref{fig:phi}.
+Below we report upper bounds on the expected value of this divergence for the MLE and the MAP.
 
@@
 
-In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
 \begin{theorem}[MLE Tight Bound]
-\label{thm:varianceMLE}
 	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
@@
 \end{theorem}
+
 This upper bound is asymptotically tight.
@@
 We get a similar bound for the multivariate generalization :
-the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (see appendix).
-
-For MAP, we get an upper bound thanks to
+the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.
+
+For the MAP, we get an upper bound thanks to the inequality $ - \log(z) \leq \inv{z} - 1$ which implies
 \begin{align}
 	\label{eq:log_bound}
-	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
+	\phi(z) \leq \half (z + \inv{z}) - 1 = \frac{(z-1)^2}{2 z} \; .
 \end{align}
@@
 \begin{theorem}[MAP Bound]
-\label{thm:varianceMAP}
-For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
- \begin{align}
-	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
+ For $n\in \naturalnumbers$, let us  define
+ \begin{align}
+	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; .
+ \end{align}
+The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
+ \begin{equation}
+	\expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
 	\leq \begin{cases}
@@
 	\end{cases}
-	\label{eq:MAP_rate}\\
-	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
-\end{align}
+	\label{eq:MAP_rate}
+\end{equation}
 \end{theorem}
-\rlp{upper bound and compare the symmetrized bregman instead?}
+
 This inequality highlights a clear variance-bias decomposition.
@@
 For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
-This correlates well with numerical observations (see appendix).
-Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
-
-These results are instantiation the bias-variance decomposition and giving us a flavor of what we would like as a general result for exponential families : a finit sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
-In the next section, we investigate what this problem is brining to optimization.
+This correlates well with numerical observations.
+Remark that the variance term is not asymptotically tight as the log-inequality we used \eqref{eq:log_bound} is not quadratically tight around $1$.
+We are basically losing a factor 2 compared to $1/2n$.
 
@@
 
-\subsection{Gaussian}
-\label{ssec:gaussian}
-The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
-
 \section{Optimization Perspective}
@@
 
-%This section show how MAP can be interpreted as stochastic mirror descent (SMD).
-Thanks to this strong link, we show that \textbf{1)} we may obtain a convergence rate for MAP from the rate od SMD, and \textbf{2)} any insights gained from the MAP may inform on further design and analysis of SMD.
-%Then, we review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three different analyses of SMD with the MAP.
-%\rlp{mention that examples have so far been lacking for relative SMD.}
-%
-%\subsection{MAP as Stochastic Mirror Descent}
-%\label{ssec:MAP=SMD}
-%By the Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
-%Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
-%\begin{align*}
-%	\hat \nat_{n}
-%	= \argmin_\nat f_n(\nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
-%    %\label{eq:SBPP}
-%\end{align*}
-%which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$.
-%%Maybe skip this SBPP story altogether
-%If we expand $f_n(\nat) = \logpart(\nat) - T_n$,  we get (after some manipulation with the Bregman divergence)
-%\begin{align*}
-%		\hat \nat_{n}
-%	= \argmin_\nat \langle g_{n}(\hat \nat_{n-1}), \nat \rangle + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
-%    %\label{eq:SMD}
-%\end{align*}
-%where $g_{n}(\nat) := \nabla f_n(\nat) = \hat \m_{n-1} - T_n$.
-%We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$.
-%From the first order optimality condition, we see this is a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
-%This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
-%In particular, it could potentially apply to a wide range of distribution $\cD$ on $X$.
-%Note that this optimization perspective does not hold for the MLE %($n_0=0$)
-%as its implicit uniform prior may correspond to an initialization outside of $\Theta$.
-
-As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
-This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
-We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
+We now show how MAP can be interpreted as stochastic mirror descent (SMD).
+This means that 1. we may obtain a convergence rate for MAP from an optimization analysis, and 2. any insights gained from the MAP may inform further design and analysis of SMD.
+We then review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
 \rlp{mention that examples have so far been lacking for relative SMD.}
 
+\subsection{MAP as Stochastic Mirror Descent}
+\label{ssec:MAP=SMD}
+By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
+Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
+\begin{align*}
+	\hat \nat_{n}
+	= \argmin_\nat f_n(\nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
+    %\label{eq:SBPP}
+\end{align*}
+which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$.
+%Maybe skip this SBPP story altogether
+Expanding $f_n(\nat) = \logpart(\nat) - T_n$ and  playing with the Bregman we get
+\begin{align*}
+		\hat \nat_{n}
+	= \argmin_\nat \langle g_{n}(\hat \nat_{n-1}), \nat \rangle + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
+    %\label{eq:SMD}
+\end{align*}
+where $g_{n}(\nat) := \nabla f_n(\nat) = \hat \m_{n-1} - T_n$.
+We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$.
+From the first order optimality condition, we see this is a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
+This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
+In particular, it could potentially apply to a wide range of distribution $\cD$ on $X$.
+Note that this optimization perspective does not hold for the MLE %($n_0=0$)
+as its implicit uniform prior may correspond to an initialization outside of $\Theta$.
+
 \subsection{Relative Smoothness}
-Mirror descent (MD) \citep{nemirovski1983problem,beck2003mirror}, also known as
-Bregman (proximal) gradient, relative gradient descent or NoLips,
+Mirror descent (MD)\footnote{
+Also known as
+Bregman (proximal) gradient, relative gradient descent or NoLips.
+}
+\citep{nemirovski1983problem,beck2003mirror}
 %Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
@@
 and strong convexity assumption on the potential $\logpart$
-\citep[Th.4.2(MD) \& Th.6.3(SMD)]{bubeck2015convex}.
+\citep[Th.4.2 \& Th.6.3]{bubeck2015convex}.
 In our case, none of these assumptions may hold. 
@@
 
-Recently, these assumptions have been relaxed to the $\mu$-strong convexity and $L$-smoothness of $f$
+\fdk{
+{\bf Online learning paragraph?}
+\citet{azoury2001relative,freund1996predicting,dasgupta2007online},
+Bubeck's 2011.
+}
+
+Recently, these assumptions have been relaxed to allow for non-smooth functions $f$
+\footnote{$f$ is $L$-smooth if its gradient is $L$-Lipschitz}.
+and non-strongly convex potential $\logpart$, first in the deterministic setting
+\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
+These assumptions have been replaced with
+the $\mu$-strong convexity and $L$-smoothness of $f$
 \emph{relative} to $\logpart$, defined as
@@
 }
-This extends the theory to a family of non-smooth functions $f$ and non-strongly convex potential $\logpart$, first in the deterministic setting
-\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
 When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent. 
 Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
-
-\textcolor{red}{Sure of this formulation?} Exponential families log-likelihood perfectly fit into this framework, as
+Exponential families log-likelihood perfectly fit into this framework, as
 \aligns{
@@
 is $1$-smooth and $1$-strongly convex relative to $A$.
-However, the (relative) smoothness is not biggest issue in our convergence analysis, but the bound on the variance, as explained in the next section.
 
 \subsection{Bounding the Randomness}
+Smoothness is not the only issue.
 To analyze stochastic algorithms, one also need to quantify the randomness.
@@
 \begingroup
-%\renewcommand*{\arraystretch}{1.25}%
+\renewcommand*{\arraystretch}{1.25}%
 \newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
@@
 \centering
-\caption{Summary of results for SMD
-under relative smoothness and relative strong convexity assumptions.
-Barrier tells whether the assumption can hold with barrier objectives.
-$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
-}
 \begin{tabular}{lccc}
@@
 %\\
-Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
+Variance $\forall \nat\in\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
 & \redxmark & \greencmark  & \redxmark
 \\
-Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
+Variance at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
 & \redxmark & \greencmark  & \greencmark
 \\
-Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
+Optimality gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
 & \greencmark & \redxmark & \greencmark
@@
 \end{tabular}
+\caption{
+Summary of results about SMD with relative smoothness and relative strong convexity assumptions. 
+Barrier tells whether the assumption can hold with barrier objectives.
+$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
+}
 \label{tbl:assumptions}
@@
 this definition recovers the variance of the stochastic gradient
-\[
-	\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C.
-\]
-More generally, if $A$ is $\mu$-strongly convex w.r.t. $\norm{\cdot}$,
+$\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C$.
+More generally, if $A$ is $\mu$-strongly convex in $\norm{\cdot}$,
 the LHS of \cref{eq:hanzely} is bounded by
@@
 to hold.
-\textcolor{red}{check formulation} Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
+Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
 \aligns{
@@
 thus showing positivity.
-Under this assumption, \textcolor{red}{who is they?}they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
+Under this assumption, they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
 
@@
 }
-In words, this bounds the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
+In words, this is bounding the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
 % As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality. 
@@
 
-\textcolor{red}{Too informal, plz rephrase} Seems like a perfect fit then ?
-Well not really, this variance assumption does not hold for exactly the same reason as the one of \textcolor{red}{use cite command} Hanzely.
+Seems like a perfect fit then ?
+Well not really, this variance assumption does not hold for exactly the same reason as the one of Hanzely.
 Having something valid for every possible iterates is way too strong.
@@
 \subsubsection{Bounded Optimality Gap}
-Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explore the hypothesis
+Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explores the hypothesis
 \alignn{
-\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C,
+\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C 
 \label{eq:dorazio}
@@
 where $f_X$ is a stochastic estimate of $f = \expecti{f_X}$. In our case $f_X(\nat) = - \log p(X\cond \nat)$.
-In words, this lower bounds the expectation of the minimum of the stochastic estimates.
+In words, this is lower bounding the expectation of the minimum of these stochastic estimates.
 For probabilistic models, such a bound is finite as soon as the model cannot give infinite density to any data point $x$.
-This holds for instance for discrete distributions because the probability mass is upper bounded by $1$,
+This holds for instance for discrete distributions 
+-- $f_x\geq 0$ because the probability mass is upper bounded by $1$ -- 
 but it rules out  many families.
-In the case of normal distributions, setting $\mu=x$ and $\sigma^2 \rightarrow 0$ gets $p(x\cond \nat) \rightarrow +\infty$.
-\textcolor{red}{Check this, I changed the formulation but i am not sure if this remains correct or not.} We have a similar behavior for gamma distribution with $\alpha = \beta x$ and $\beta \rightarrow +\infty$, or with the beta distribution with $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
+For normal distributions set $\mu=x$ and $\sigma^2 \rightarrow 0$ to get $p(x\cond \nat) \rightarrow +\infty$. 
+For gamma, set $\alpha = \beta x$ and $beta \rightarrow +\infty$.
+For beta, set $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
  Other examples include inverse Gaussians, log-normal, gamma, inverse gamma.
 
-It is possible to overcome this limitation by treating batches of samples as single samples by averaging sufficient statistics, e.g.,. $Y = \{X_1, \dots, X_k\}$ and $T(Y) = \inv{k}\sum_i T(X_i)$.
-For instance, a multivariate normals of dimension $d$ cannot attribute infinite density to $d+1$ samples that are not in an affine subspace.
-
-In conclusion, it seems these approaches can yield to interesting convergence rates for the MAP, and they do prove linear convergence of the last iterate up a to a variance ball with constant step-size, but the size of this ball $\frac{C}{\alpha}$ \rlp{strong convexity notation} is independent of the step-size.
-\textcolor{red}{Who is their?} Their descent lemma does not lead to convergence with a decreasing $O(1/t)$ step-size.
+ It is possible to overcome this limitation though, by treating batches of samples as 1 sample, and averaging sufficient statistics -- e.g. $Y = \{X_1, \dots, X_k\}$ and $T(Y) = \inv{k}\sum_i T(X_i)$.
+ For instance a multivariate normals of dimension $d$ cannot attribute infinite density to $d+1$ samples that are not in an affine subspace.
+
+So it seems this approach can yield interesting convergence rates for the MAP, and they do prove linear convergence of the last iterate up a to a variance ball with constant step-size, but the size of this ball $\frac{C}{\alpha}$ \rlp{strong convexity notation} is independent of  the step-size.
+Their descent lemma does not lead to convergence with a decreasing $O(1/t)$ step-size.
 
@@
 \TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
-
-\subsection{Related Work}
-
-\paragraph{Parameter Estimation.}
-Statistics has a plethora of asymptotic results \citep{vdv1998asymptotic}.
-In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes the decay of the tail.
-
-\paragraph{Density Estimation.}
-Since the KL explode to $\infty$ if the model is unable to give weight to some samples, a lot of this work has been devoted to distances, such as $L^2$ \citep{tsybakov2009introduction}, $L^1$ \citep{devroye2001combinatorial} or more recently Hellinger distance \citep{baraud2017new}.
-Contrary to the KL, these distance are symmetric and behave smoothly.
-
-\paragraph{Online Learning}
-Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
-\[
-R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
-\]
-where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
-Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.
-Unfortunately however, their rate are worse than the offline MAP or MLE algorithms.
-For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$.
-From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
-Also, \citet{dasgupta2007online} prove the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
-This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
 
@@
 \bibliographystyle{apalike}
-\bibliography{references} % Removed the .bib as this mess up with the compiler
+\bibliography{references.bib}
 

### 2021-10-14 aistats.tex
Address Damien's comments on optim

--- 2021-10-14 aistats.tex
@@
 
+\newcommand{\stgcvx}{\alpha} % strong convexity
+\newcommand{\smooth}{\beta} % smoothness
+
 \begin{document}
@@
 This rule is known as moment matching.
-Given a conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
+Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
 %\textcolor{red}{DS: I know that for conciseness we do not write the MAP estimate here, but it's super central and I think its important to write the equation. WDYT?}
@@
 \fdk{It does, you just forget the initial condition after the first step with step-size 1?}
+\rlp{to discuss.}
 
@@
 
-{\bf If $\conj$ is $L$-smooth} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
+{\bf If $\conj$ is $L$-smooth}\footnote{e.g. $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
 \begin{align}
@@
 Proof in the appendix
-\footnote{In case the reader is intrigued, $0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
+\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
 To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
@@
 
-Recently, these assumptions have been relaxed to the $\mu$-strong convexity and $L$-smoothness of $f$
+Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
 \emph{relative} to $\logpart$, defined as
 \aligns{
-	\mu \cB_{A}(x, y)
+	\stgcvx \cB_{A}(x, y)
 	\leq
@@
 	\leq
-	L \cB_A(x,y) \; .
-}
-This extends the theory to a family of non-smooth functions $f$ and non-strongly convex potential $\logpart$, first in the deterministic setting
-\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
+	\smooth \cB_A(x,y) \; .
+}
 When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
-Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
-
-\textcolor{red}{Sure of this formulation?} Exponential families log-likelihood perfectly fit into this framework, as
+Those conditions ensure the linear convergence of MD with $A$ as the reference function,
+thus extending the theory to a family of non-smooth functions $f$ and non-strongly convex potential $\logpart$
+\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
+
+For exponential families, the log-likelihood perfectly fit into this framework, as
 \aligns{
@@
 }
-is $1$-smooth and $1$-strongly convex relative to $A$.
-However, the (relative) smoothness is not biggest issue in our convergence analysis, but the bound on the variance, as explained in the next section.
+is both $1$-smooth and $1$-strongly convex relative to $A$. 
+So we are left with finding an applicable convergence rate for SMD under relative assumptions.
+As we investigate in the next section, several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
 
@@
 \]
-More generally, if $A$ is $\mu$-strongly convex w.r.t. $\norm{\cdot}$,
+More generally, if $A$ is $\stgcvx$-strongly convex w.r.t. $\norm{\cdot}$,
 the LHS of \cref{eq:hanzely} is bounded by
-$\expecti{\frac{1}{\mu}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$,
-and a bound on the variance of the stochastic gradient is a sufficient condition for \cref{eq:hanzely}
+$\expecti{\frac{1}{\stgcvx}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$.
+Then a bound on the variance of stochastic gradients is sufficient for \cref{eq:hanzely}
 to hold.
-\textcolor{red}{check formulation} Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
-\aligns{
-\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) = \E[\cS(\MAPm - \lr g_n  )]
-}
-thus showing positivity.
-Under this assumption, \textcolor{red}{who is they?}they prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size \citep[Lem.4.8]{hanzely2018fastest}, by applying a polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
+%Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
+%\aligns{
+%\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) = \E[\cS(\MAPm - \lr g_n  )]
+%}
+%thus showing positivity.
+Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size, by applying polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
 
@@
 
-\textcolor{red}{Too informal, plz rephrase} Seems like a perfect fit then ?
-Well not really, this variance assumption does not hold for exactly the same reason as the one of \textcolor{red}{use cite command} Hanzely.
-Having something valid for every possible iterates is way too strong.
+This seems to be a perfect solution to our problem.
+Yet it suffers from the exact same qualm as \citet{hanzely2018fastest}.
+For barrier objectives, \cref{eq:dragomir} cannot hold uniformly on every possible path of iterates, because these paths can get arbitrarily close from boundaries where $\bregmanconj$ becomes infinite.
 
@@
 but it rules out  many families.
-In the case of normal distributions, setting $\mu=x$ and $\sigma^2 \rightarrow 0$ gets $p(x\cond \nat) \rightarrow +\infty$.
-\textcolor{red}{Check this, I changed the formulation but i am not sure if this remains correct or not.} We have a similar behavior for gamma distribution with $\alpha = \beta x$ and $\beta \rightarrow +\infty$, or with the beta distribution with $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
- Other examples include inverse Gaussians, log-normal, gamma, inverse gamma.
+In the case of normal distributions$\cN(m, \sigma^2)$, setting $\mu=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$.
+We have a similar behavior for gamma distribution with $\alpha = \beta x$ and $\beta \rightarrow +\infty$, or with the beta distribution with $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
+ Other counter-examples include inverse Gaussians, log-normal, gamma, inverse gamma.
 
@@
 
-In conclusion, it seems these approaches can yield to interesting convergence rates for the MAP, and they do prove linear convergence of the last iterate up a to a variance ball with constant step-size, but the size of this ball $\frac{C}{\alpha}$ \rlp{strong convexity notation} is independent of the step-size.
-\textcolor{red}{Who is their?} Their descent lemma does not lead to convergence with a decreasing $O(1/t)$ step-size.
+Overall, this approach seems to answer~\eqref{problem}.
+Yet it fails to account for the step-size $\lr_n = \inv{n_0+n}$, as \citet[Thm.1]{dorazio2021stochastic} only proves linear convergence to a variance ball of size $\frac{C}{\stgcvx}$ under constant step-size $\lr \leq \inv{\smooth}$.
 

### 2021-10-14 aistats.tex
Move examples earlier
And rename sections

@@
 
-\section{Motivation}
+\section{Introduction}
 \label{sec:motivation}
@@
 }
-where $\logpart$ is a convex function and $\nat\in|Theta$ is a vector called \textit{natural parameter}.
+where $\logpart$ is a convex function and $\nat\in\Theta$ is a vector called \textit{natural parameter}.
 This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized}
@@
 \paragraph{Maximum A Posteriori (MAP).}
-Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD|\nat)p(\nat)$.
+Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
 Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
@@
 
-\section{Open Problem}
+\section{Problem Formulation}
 \label{sec:problem}
@@
 
-\section{Insights}
+We start with special examples, with a concrete upper bound, to give insights into what we want to achieve.
+
+\section{Examples}\label{sec:example}
+\subsection{Gaussian Variance}\label{ssec:gaussian-variance}
+
+A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
+Its density reads
+\[
+	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
+\]
+Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
+Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
+\textcolor{red}{The explanation here is not clear enough, or too concise} Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
+\begin{align}
+	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
+	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
+\end{align}
+Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
+It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
+It also means $\bregman$ and $\bregmanconj$ have the same shape
+\begin{align}
+	\bregmanconj( \m_*; \m_n)
+	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
+	\bregman( \nat_n; \nat_* )
+	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
+\end{align}
+Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
+\begin{align}
+	\phi(z) := \half (z - 1 - \log(z))
+\end{align}
+illustrated in Figure~\ref{fig:phi}.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=.4\textwidth]{phi.pdf}
+	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics, but it admits another upper-bound (in grey).}
+	\label{fig:phi}
+\end{figure}
+
+In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
+\begin{theorem}[MLE Tight Bound]
+\label{thm:varianceMLE}
+	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
+	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
+	\begin{align}
+		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
+			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
+			\label{eq:MLE_rate}
+	\end{align}
+\end{theorem}
+This upper bound is asymptotically tight.
+We illustrate its numerical behavior in Figure~\ref{fig:curves}.
+We get a similar bound for the multivariate generalization :
+the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (see appendix).
+
+For MAP, we get an upper bound thanks to
+\begin{align}
+	\label{eq:log_bound}
+	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
+\end{align}
+
+\begin{theorem}[MAP Bound]
+\label{thm:varianceMAP}
+For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
+ \begin{align}
+	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
+	\leq \begin{cases}
+		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
+		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
+	\end{cases}
+	\label{eq:MAP_rate}\\
+	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
+\end{align}
+\end{theorem}
+\rlp{upper bound and compare the symmetrized bregman instead?}
+This inequality highlights a clear variance-bias decomposition.
+In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
+For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
+This correlates well with numerical observations (see appendix).
+Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
+
+These results are instantiation the bias-variance decomposition and giving us a flavor of what we would like as a general result for exponential families : a finit sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
+In the next section, we investigate what this problem is brining to optimization.
+
+\begin{figure*}[t]
+	\centering
+	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
+	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
+	\label{fig:bias-variance-numerical}
+\end{figure*}
+
+\subsection{Gaussian}
+\label{ssec:gaussian}
+The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
+
+\section{Partial Solutions}
 \label{sec:insights}
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
-Besides the Gaussian with known mean, no standard exponential family verify this assumption either.
+Besides the Gaussian with known variance, no standard exponential family verify this assumption either.
 Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
@@
 
+\textbf{Conjecture:} the bias decreases like $O(1/n^2)$.
+
 \paragraph{Illustrations.}
@@
 
-\section{Examples}\label{sec:example}
-\subsection{Gaussian Variance}\label{ssec:gaussian-variance}
-
-A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
-Its density reads
-\[
-	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
-\]
-Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
-Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
-\textcolor{red}{The explanation here is not clear enough, or too concise} Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
-\begin{align}
-	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
-	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
-\end{align}
-Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
-It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
-It also means $\bregman$ and $\bregmanconj$ have the same shape
-\begin{align}
-	\bregmanconj( \m_*; \m_n)
-	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
-	\bregman( \nat_n; \nat_* )
-	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
-\end{align}
-Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
-\begin{align}
-	\phi(z) := \half (z - 1 - \log(z))
-\end{align}
-illustrated in Figure~\ref{fig:phi}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.4\textwidth]{phi.pdf}
-	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics, but it admits another upper-bound (in grey).}
-	\label{fig:phi}
-\end{figure}
-
-In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
-\begin{theorem}[MLE Tight Bound]
-\label{thm:varianceMLE}
-	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
-	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
-	\begin{align}
-		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
-			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
-			\label{eq:MLE_rate}
-	\end{align}
-\end{theorem}
-This upper bound is asymptotically tight.
-We illustrate its numerical behavior in Figure~\ref{fig:curves}.
-We get a similar bound for the multivariate generalization :
-the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (see appendix).
-
-For MAP, we get an upper bound thanks to
-\begin{align}
-	\label{eq:log_bound}
-	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
-\end{align}
-
-\begin{theorem}[MAP Bound]
-\label{thm:varianceMAP}
-For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
- \begin{align}
-	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
-	\leq \begin{cases}
-		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
-		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
-	\end{cases}
-	\label{eq:MAP_rate}\\
-	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
-\end{align}
-\end{theorem}
-\rlp{upper bound and compare the symmetrized bregman instead?}
-This inequality highlights a clear variance-bias decomposition.
-In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
-For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
-This correlates well with numerical observations (see appendix).
-Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
-
-These results are instantiation the bias-variance decomposition and giving us a flavor of what we would like as a general result for exponential families : a finit sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
-In the next section, we investigate what this problem is brining to optimization.
-
-\begin{figure*}[t]
-	\centering
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
-	\label{fig:bias-variance-numerical}
-\end{figure*}
-
-\subsection{Gaussian}
-\label{ssec:gaussian}
-The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
-
-\section{Optimization Perspective}
+\section{AN OPTIMIZATION PROBLEM}
 \label{sec:optimization}
@@
 \subsection{Bounding the Randomness}
+\TODO{compress this section}
+
 To analyze stochastic algorithms, one also need to quantify the randomness.
@@
 }
-which fails to hold on the gaussian variance example for instance. \TODO{short proof or formalize.}
+which fails to hold on the gaussian variance example for instance. \TODO{ DAMIEN short proof or formalize.}
 
@@
 
-\subsection{Related Work}
+\section{Related Work}
 

### 2021-10-14 aistats.tex
Move some related work to intro

@@
 {\bf 1)} distance in parameter space $d(\nat,\nat^*)$.
-This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation}.
+This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
+In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
 {\bf 2)} distance between distributions, as studied in \emph{density estimation}.
 For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
-but its lack of robustness to misspecification\footnote{For $p$ and $q$ continuous densities of the ground truth and a model,
-$\KL(p||q) = +\infty$ if $q(x)=0$ for some $x$ such that $p(x)>0$.
-}
-has led statisticians to study other distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction}.
-As we will see, with exponential families, the KL divergence is also a Bregman divergence between parameters, thus drawing a connection between these two lines of research.
+but its lack of robustness to misspecification\footnote{
+For $p$ and $q$ continuous densities,
+$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \&p(x)>0$.
+%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
+}
+has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[e.g.
+\S1.2]{tsybakov2009introduction} , the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
+With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
 Hence the fundamental question:
@@
 
+\begin{center}
 \begin{equation}
@@
 \begin{aligned}
-	\textit{Can we upper}&\textit{ bound  the expected value of} \\
+	\textit{Can we upper}&\textit{ bound  the expected value of } \\
 	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \ ?
@@
 \end{equation}
+\end{center}
 % RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.
@@
 \section{Related Work}
-
-\paragraph{Parameter Estimation.}
-Statistics has a plethora of asymptotic results \citep{vdv1998asymptotic}.
-In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes the decay of the tail.
-
-\paragraph{Density Estimation.}
-Since the KL explode to $\infty$ if the model is unable to give weight to some samples, a lot of this work has been devoted to distances, such as $L^2$ \citep{tsybakov2009introduction}, $L^1$ \citep{devroye2001combinatorial} or more recently Hellinger distance \citep{baraud2017new}.
-Contrary to the KL, these distance are symmetric and behave smoothly.
 

### 2021-10-14 aistats.tex
Reorganize contributions to match main text

@@
 With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
-Hence the fundamental question:
+Hence the fundamental problem:
 %\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution.
@@
 %\end{center}
-
-\begin{center}
 \begin{equation}
@@
 \begin{aligned}
-	\textit{Can we upper}&\textit{ bound  the expected value of } \\
-	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \ ?
+	\textit{Find an upper}&\textit{ bound on the expected value of } \\
+	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
 \end{aligned}
@@
 \end{equation}
-\end{center}
 % RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.
@@
 	\itemsep0em
+	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
 	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
-	\item We investigate a generalized bias-variance decomposition of the KL\citep{pfau2013generalized}. \TODO{revealing what ? the ideal shape of a rate and a sub-problem} (\S\ref{ssec:bias-variance})
-	\item We answer~\eqref{problem} with an upper bound in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
-	\item We establish a connection between~\eqref{problem} and optimization, and use it to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
+	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance})
+	\item We use a connection between~\eqref{problem} and optimization to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
 \end{itemize}

### 2021-10-14 aistats.tex
Clean and add links to appendix

@@
 }
-has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[e.g.
-\S1.2]{tsybakov2009introduction} , the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
+has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction} , the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
 With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
@@
 
-We know asymptotic answers.
-We also know simple upper bounds if $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
-or close to quadratic (see \S\ref{ssec:quadratic}).
-But in general, this problem is difficult. In this paper, we show some partial solutions, and give some ideas on how to solve it.
+We have asymptotic solutions (\S\ref{ssec:asymptote}),
+or when $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
+or close to quadratic (see \S\ref{ssec:quadratic}),
+but a  general solution for finite $n$ remains elusive.
+In this paper, we review partial solutions, and give some ideas on how to solve it.
 %\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
-%RLP poroblem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.
+%RLP problem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.
 
@@
 In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
-A convergence rate for this algorithm would translate into the upper bound we seek.
-%From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g.
-not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
-Conversely we can see this problem as a very special case to study SMD, revealing the limits of current analysis.
+Inspired by recent work~\citep{lepriol2021analysis}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
+Unfortunately, none of the current analysis apply, which frames another open problem for the analysis of SMD.
 
@@
 It could also help in assessing the impact of alternative forms of regularization (prior) for these models.
-
-%Or it could help developing accelerated (stochastic?) algorithms for barrier objectives  overcoming the limitations of non-Euclidean geometry \citep{dragomir2021optimal}.
-\rlp{Fred I need your feedback on that one.}
 
@@
 	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance})
-	\item We use a connection between~\eqref{problem} and optimization to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
+	\item By identifying MAP and stochastic mirror descent (SMD), we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
 \end{itemize}
@@
 
-\paragraph{Optimization}
-We can rewrite~\eqref{eq:defMAP} as
-\alignn{
-\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
-\label{eq:mean_update}
-}
-where $\lr_n := \inv{n_0 + n}$.
-Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
-If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
-\alignn{
-	\nabla\conj(\hat \nat_{n})
-	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
-}
-which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$
-and step-size schedule $\lr_n$, initialized at $\nat_0$.
-\emph{MAP is an instantiation of SMD.}
-Note that this analogy does not hold for the MLE %($n_0=0$)
-as it does not have any initialization to start from.
-\fdk{It does, you just forget the initial condition after the first step with step-size 1?}
-\rlp{to discuss.}
-
 \section{Problem Formulation}
@@
 
+\paragraph{Optimization}
+We can rewrite~\eqref{eq:defMAP} as
+\alignn{
+\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
+\label{eq:mean_update}
+}
+where $\lr_n := \inv{n_0 + n}$.
+Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
+If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
+\alignn{
+	\nabla\conj(\hat \nat_{n})
+	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
+}
+which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$
+and step-size schedule $\lr_n$, initialized at $\nat_0$.
+\emph{MAP is an instantiation of SMD.}
+In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
+
 We start with special examples, with a concrete upper bound, to give insights into what we want to achieve.
@@
 In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
+Proofs are in \cref{app:gaussian-variance}.
 \begin{theorem}[MLE Tight Bound]
@@
 We get a similar bound for the multivariate generalization :
-the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (see appendix).
+the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (c.f. App.\ref{app:gaussian-variance}).
 
@@
 For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
-This correlates well with numerical observations (see appendix).
+This correlates well with numerical observations (cf App.\ref{app:gaussian-variance}).
 Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
@@
 \end{align}
-See proof in the Appendix.
+Detailed derivation is in~\cref{app:asymptote}
 Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
@@
 we know that all Bregman divergences are locally quadratic.
-Under some assumptions, like self-concordance of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
+Under some assumptions, such as self-concordance of $\conj$\footnote{
+In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
+} \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
 \begin{proposition}
@@
 \end{proposition}
-Proof in the appendix
 \footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
@@
 For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
-For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$ (see appendix).
+For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
 This means that on average, we need much more samples than the dimension to reach the quadratic regime.
@@
 \bibliographystyle{apalike}
-\bibliography{references} % Removed the .bib as this mess up with the compiler
+\bibliography{references}
 
 \clearpage
-\section{Group Meeting}
+\appendix
+\onecolumn
+
+\section{PROOFS FOR GAUSSIAN VARIANCE}
+\label{app:gaussian-variance}
+
+\section{ASYMPTOTIC DERIVATION}
+\label{app:asymptote}
+
+\section{SELF-CONCORDANCE}
+\label{app:self-concordant}
+
+\section{TODO}
 \begin{enumerate}
-	\item new plots : mirror map, and gaussian trajectories. Add trajectories to paper, with background level set, and new legend : primal/dual expectation
-	\item the MLE is biased according to our definition.
-See plots.
-Does this definition of bias make sense ?
-An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean.
-Remark : the MLE is biased for standard parameters.
-	\item exciting pap er\citet{kakade2010learning}
-	\begin{enumerate}
-		\item assumption in the spirit of infinite order self-concordance only at the optimum
-		\item After a burn-in phase, the loss behaves quadratically. No precise description of the burn-in length.
-		\item \TODO{it is worth writing down until the end what we are obtaining with self-concordance.}
-		\item idea 1 : try to express some of their results with the mean parameter instead.
-		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD.
-(exploiting the link MAP -> optimization).
-How do we prove the linear convergence of Newton to the quadratic regime ?
-That is the problem at stake here.
-	\end{enumerate}
-	\item Intro : positioning between optimization and statistics.
-Start with stats.
-Convince the reader that this is an essential problem, with asymptotic rate.
-but finite is elusive.
-Inspired by recent work (lepriol2021 , bach2013) we want to use optimization to give a rate.
-We fail, but we reveal a hole in current analysis.
-Finite rate would allow us to quantify the importance of the prior, and also help us understand behavior for small numbers.
-	\item Intro : 1 clear question and sharper contributions, with references to sections, as a memo for the reviewer writing his review.
-"investigate bias-variance..." "we show all these analysis suffer from this problem."
-	\item Organization : add optim parallel in insigths section, and then sprinkle references until the optim section.
-	\item Optim : find members of the exponential family where the rate of Dragomir et al. holds.
+	\item new plots : add 2d trajectories to paper, with background level set, and new legend : primal/dual expectation, and clear explanation !
+	\item Incorporate related work by Frederik.
 	\item Technical Background : be precise and add reference for the steep exponential family of Legendre type which guarantees that our sets are open, and have a solution in the middle.
@@
 	\item $\tilde\theta_n$ for $\expect{\hat\theta_n}$, which is biased
-	\item $(L, \mu)$ for strong-convexity and smoothness
+	\item $(\stgcvx , \smooth)$ for strong-convexity and smoothness
 \end{itemize}
 
-\fdk{
-	Do we use $\alpha, \beta$ anywhere?
-	Might use those for smoothness/strong-convexity rather than $\mu, L$.
-}
-
-\fdk{
-	The superscript ${}^*$ denotes the convex conjugate when applied on a function
-	and the optimum when applied to a parameter.
-	How about we use $\theta_n, \theta_*$ for parameter at time $n$ and at the optimum
-	and $f^*$ for the convex conjugate?
-}
-
-\fdk{
-	On $X_i$ for data; I'm more used to seeing $x_i$ for data, which is the realization of the random variable $X_i$.
-	We don't talk about random variables and don't need the distinction?
-}
-
-\fdk{
-	Need to check
-}
-
 \end{document}

### 2021-10-14 aistats.tex
Remove comments from intro
Lighten gaussian variance section

@@
 Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
-%\textcolor{red}{DS: I know that for conciseness we do not write the MAP estimate here, but it's super central and I think its important to write the equation. WDYT?}
-% I think it's okay to have it only later, and refernece it here.
-
-%\textcolor{red}{This comes a bit from nowhere.
-Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research.
-We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution.
-In the first case, the analysis use norms blabla.
-In the second, we use KL"}
+
 {\bf Metrics.}
@@
 Hence the fundamental problem:
-%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution.
-}
-% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
 %\begin{center}
@@
 In this paper, we review partial solutions, and give some ideas on how to solve it.
-%\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
-%RLP problem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.
-
-%\textcolor{red}{My main problem with the following paragraph here is that its at the same time too detailed for the intro, but also not enough detailed to understand it.}
-%RLP right, not sure about how to fix this. It is hard to decide on the right amount of space for optimization versus statistics.
 
@@
 
-{\bf Unboundedness of~\eqref{eq:bregmanMLE}.}
+{\bf Difficulty with MLE.}
 While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite,
@@
 	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
-	\conj(\m) &= \half\left( -\log(\m) + \log\frac{\pi}{2} - 1 \right) \; .
+	\conj(\m) &= - \half \log(\m) + \half \log\frac{\pi}{2} - \half  \; .
 \end{align}
@@
 \end{align}
-Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
-\begin{align}
-	\phi(z) := \half (z - 1 - \log(z))
-\end{align}
-illustrated in Figure~\ref{fig:phi}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.4\textwidth]{phi.pdf}
-	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics, but it admits another upper-bound (in grey).}
-	\label{fig:phi}
-\end{figure}
-
 In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
-Proofs are in \cref{app:gaussian-variance}.
-\begin{theorem}[MLE Tight Bound]
+All proofs for this section are in \cref{app:gaussian-variance}.
+\begin{theorem}[MLE Bound]
 \label{thm:varianceMLE}
@@
 \end{theorem}
-This upper bound is asymptotically tight.
+This upper bound matches the asymptote~\eqref{eq:asymptote} .
 We illustrate its numerical behavior in Figure~\ref{fig:curves}.
-We get a similar bound for the multivariate generalization :
-the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (c.f. App.\ref{app:gaussian-variance}).
-
-For MAP, we get an upper bound thanks to
-\begin{align}
-	\label{eq:log_bound}
-	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
-\end{align}
+With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.
 
@@
 \end{theorem}
-\rlp{upper bound and compare the symmetrized bregman instead?}
-This inequality highlights a clear variance-bias decomposition.
-In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
-For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
-This correlates well with numerical observations (cf App.\ref{app:gaussian-variance}).
-Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
-
-These results are instantiation the bias-variance decomposition and giving us a flavor of what we would like as a general result for exponential families : a finit sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
-In the next section, we investigate what this problem is brining to optimization.
+Note that this inequality actually holds for the symmetrized Bregman $\cB(a,b) + \cB(b,a)$, which explains why it loses a factor 2 compared to the asymptote~\eqref{eq:asymptote}, but we leave this to the appendix for clarity.
+Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
+In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth, which correlates well with numerical observations (cf App.\ref{app:gaussian-variance}).
+%For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
+%Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
 
@@
 The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
+
+Equipped with the knowledge of these 2 examples, we are ready to discuss the general problem, and some ways to solve with direct expansions of the Bregman.
 
@@
 \begin{align}
-	\label{eq:MAP_asymptote}
+	\label{eq:asymptote}
 	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
@@
 \label{ssec:bias-variance}
-In both examples \textcolor{red}{(add link)}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, reminding us of the bias-variance decomposition.
+In both the quadratic~\eqref{eq:MAP_quadratic} and the Gaussiance variance examples~\eqref{eq:MAP_rate}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, 
+giving us a flavor of what we would like as a general result for exponential families : a finite sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
 In fact, such a decomposition exists for any Bregman divergence.
@@
 
+Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
+\begin{align}
+	\phi(z) := \half (z - 1 - \log(z))
+\end{align}
+illustrated in Figure~\ref{fig:phi}.
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=.4\textwidth]{phi.pdf}
+	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics, but it admits another upper-bound (in grey).}
+	\label{fig:phi}
+\end{figure}
+
+For MAP, we get an upper bound thanks to
+\begin{align}
+	\label{eq:log_bound}
+	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
+\end{align}
+
 \section{ASYMPTOTIC DERIVATION}

### 2021-10-14 aistats.tex
Compressions in tech. background

@@
 In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
-Inspired by recent work~\citep{lepriol2021analysis}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
+Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
 Unfortunately, none of the current analysis apply, which frames another open problem for the analysis of SMD.
@@
 
-For convenience, we focus on steep, regular exponential families with minimal statistic $T$.
+For convenience, we focus on steep, regular exponential families with minimal statistic $T$ \citep{barndoffnielsen2014information}.
 Then $\logpart$ is a strictly convex function of Legendre type,
-and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex
-\citep{barndoffnielsen2014information}.
+and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex.
 When explicit, we write the random variable $T = T(X)$.
@@
 
-\paragraph{Duality}
+{\bf Duality.}
 The log-partition function $\logpart$ verifies the two following identities:
@@
 %
-%\paragraph{Bregman Divergences} %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
-% We now introduce the Bregman divergence induced by $\logpart$,
-% that measures the discrepancy between two parameters $\nat$ and $\nat_0$,
-The Bregman divergence induced by $\logpart$,
+%Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
+{\bf The Bregman Divergence} induced by $\logpart$,
 measures the discrepancy between two parameters $\nat$ and $\nat_0$,
@@
 In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
-Finally, it is equal to the divergence of its convex conjugate with switched arguments,
-\begin{align}
-	\bregman (\nat ; \nat_0)
-    = \bregmanconj ( \meanp_0 ; \meanp) \; .
-    \label{eq:bregman_switch}
-\end{align}
-
-\paragraph{Conjugate Prior}
-One conjugate prior \citep{agarwal2010geometric} for $p(X|\nat)$ is
+%Finally, it is equal to the divergence of its convex conjugate with switched arguments,
+%\begin{align}
+%	\bregman (\nat ; \nat_0)
+%    = \bregmanconj ( \meanp_0 ; \meanp) \; .
+%    \label{eq:bregman_switch}
+%\end{align}
+
+{\bf A Conjugate Prior} for $p(X|\nat)$ is
 \begin{align}
@@
 \end{align}
-where $n_0$ and $\nat_0$ are (hyper)parameters of the prior.
+where $n_0$ and $\nat_0$ are (hyper)parameters of the prior  \citep{agarwal2010geometric}.
 This is an exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and natural parameter $(n_0 \m_0, -n_0)$.
@@
 
-\paragraph{Maximum A Posteriori (MAP).}
+{\bf Maximum A Posteriori (MAP)}
 Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
@@
 In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
-Using~\eqref{eq:defNLL} and~\eqref{eq:bregman_switch}, we can also write the KL with Bregman divergences
+With exponential families, the KL is also a Bregman divergence
 \alignn{
@@
 }
+where the second equality is a general property of Bregman divergences and convex conjugates. (Be careful about the 
 How does this quantity behave when $\nat$ is the MLE or MAP?
@@
 
-{\bf Difficulty with MLE.}
+{\bf A Difficulty with the MLE.}
 While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite,

### 2021-10-14 aistats.tex
Complement optimization in section problems

@@
 \newtheorem{proposition}{Proposition}
+\newtheorem{problem}{Open Problem}
+
 \newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
@@
 We write $\MAPt$ for the MAP, and view the MLE as a special case.
-
-\section{Problem Formulation}
+We are now ready to formalize the main problem of this paper.
+
+\section{PROBLEMS FORMULATION}
 \label{sec:problem}
@@
 }
-where the second equality is a general property of Bregman divergences and convex conjugates. (Be careful about the 
+where the second equality is a general property of Bregman divergences and convex conjugates. (Be careful about the order of arguments.)
 How does this quantity behave when $\nat$ is the MLE or MAP?
-Can we get upper bounds on the expected KL?
+ Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators when the loss is the KL divergence? 
+This is our first problem.
+
+\begin{problem}[Bounding MAP and MLE]
+Upper bound the following quantities:
 \begin{align}
@@
 \end{align}
-where the expectation is on the data $\mD$.
-More explicitly, is there an upper bound that does not involve this expectation over the dataset?
-Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators
-when the loss is the KL divergence?
-
+where the expectation is on the data $\mD = (T_1, \dots, T_n)$.
+\end{problem}
+
+More explicitly, we want an upper bound that does not involve this expectation over the dataset.
 Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
-There is a large body of work about asymptotic or large sample results, which we partially review in \S\ref{sec:insights},
-or results for special cases such as $\cN(0, \sigma^2)$ (see \S\ref{sec:example}).
-But we are yet to find a solution encompassing a broad range of exponential families,
+In \S\ref{sec:example}, we provide results for special cases such as $\cN(0, \sigma^2)$,
+while in \S\ref{sec:insights} we provide realistic conditions to obtain bounds that are valid after seeing a large number of samples.
+But we are yet to find a solution encompassing both a broad range of exponential families,
 and applicable to small sample sizes $n \lesssim d$.
@@
 \paragraph{Optimization}
-We can rewrite~\eqref{eq:defMAP} as
+With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD).}
+More precisely, let us re-write~\eqref{eq:defMAP} as
 \alignn{
@@
 where $\lr_n := \inv{n_0 + n}$.
-Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
-If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
+Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
+If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
 \alignn{
@@
 }
-which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$
+which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
 and step-size schedule $\lr_n$, initialized at $\nat_0$.
-\emph{MAP is an instantiation of SMD.}
 In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.	
-
-We start with special examples, with a concrete upper bound, to give insights into what we want to achieve.
+The observation MAP$\in$SMD brings us to our second problem.
+\begin{problem}[Convergence rate for SMD]
+Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(m, \sigma^2)$.
+\end{problem}
+
+To address these problems, we start by investigating examples, for which we provide solutions to problem 1 and get insights into what is achievable.
 

### 2021-10-14 aistats.tex
Remove frederik's figure

@@
 which is the common notion of \textit{entropy} in information theory.
-By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$
-(see Fig.~\ref{fig:duality}), giving
+By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$, giving
 \aligns{
@@
 }
-\begin{figure}[t]
-	\centering
-	\includegraphics{duality}
-	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$.
+%\begin{figure}[t]
+%	\centering
+%	\includegraphics{duality}
+%	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$.
 Figure copied from \citet{kunstner2020homeomorphic}.
 % TODO make an original one.
-	}
-	\label{fig:duality}
-\end{figure}%
+%	}
+%	\label{fig:duality}
+%\end{figure}%
 %
@@
 
-\begin{problem}[Bounding MAP and MLE]
+\begin{problem}[Upper-bounding MAP and MLE]
 Upper bound the following quantities:
@@
 Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
-\textcolor{red}{The explanation here is not clear enough, or too concise} Now we can match the log-likelihood with the exponential family template to get the log-partition function, and taking the conjugate to find the entropy
+\textcolor{red}{The explanation here is not clear enough, or too concise} 
+Now we can match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
 \begin{align}

### 2021-10-14 aistats.tex
Compress gaussian variance and add gaussian

@@
 A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
-Its density reads
-\[
-	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
-\]
+Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
 Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
@@
 Now we can match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
-\begin{align}
-	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
-	\conj(\m) &= - \half \log(\m) + \half \log\frac{\pi}{2} - \half  \; .
-\end{align}
+\aligns{
+	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
+	\conj(\m) = - \half \log(\m)  \; ,
+}
+up to constants.
 Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
@@
 \label{thm:varianceMAP}
-For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
+The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
  \begin{align}
@@
 Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
-In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth, which correlates well with numerical observations (cf App.\ref{app:gaussian-variance}).
-%For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
-%Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
+In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth
+This correlates well with our numerical observations (cf App.\ref{app:gaussian-variance}).
 
@@
 	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
-	\label{fig:bias-variance-numerical}
+	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.
+	}
+	\label{fig:thales}
 \end{figure*}
@@
 \label{ssec:gaussian}
-The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
+The ubiquitous Gaussian $\cN(m,\sigma^2)$ actually offers a highly non-trivial example for Problem 1.
+Its log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
+With sufficient statistic $T(x)=(x, x^2)$, 
+the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
+and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$. 
+$\cM$ and $\Theta$ are delimited in grey in \cref{fig:thales}.
+Given these parameters, log-partition and entropy are
+\alignn{
+	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
+	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2)
+}
+up to constants.
+
+ Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
 
@@
 We plot the bias-variance decompositions for $\cN(0,\sigma^2)$ (Figure~\ref{fig:variance_decomposition}) and  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
-Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:bias-variance-numerical} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) .
+Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:thales} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) .
 \rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}
@@
 
-Caveat : all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.
+\textbf{Caveat :} all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.
 

### 2021-10-14 aistats.tex
Finish gaussian section

@@
 \label{ssec:gaussian}
-The ubiquitous Gaussian $\cN(m,\sigma^2)$ actually offers a highly non-trivial example for Problem 1.
-Its log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
+Now that we have solved the case of $\cN(0,\sigma^2)$, let's move on to the ubiquitous Gaussian $\cN(m,\sigma^2)$.
+Gaussians actually offers a highly non-trivial example for Problem 1.
+Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
 With sufficient statistic $T(x)=(x, x^2)$, 
@@
 and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$. 
-$\cM$ and $\Theta$ are delimited in grey in \cref{fig:thales}.
-Given these parameters, log-partition and entropy are
+Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
+Given these parameters, log-partition and entropy are, up to constants, 
 \alignn{
@@
 }
-up to constants.
-
- Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
-
-Equipped with the knowledge of these 2 examples, we are ready to discuss the general problem, and some ways to solve with direct expansions of the Bregman.
+
+These functions are neither smooth, nor strongly convex, nor self-concordant (see \cref{app:gaussian}).
+Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.
 
@@
 
+\section{PROOFS FOR GAUSSIAN}
+\label{app:gaussian}
+
 \section{ASYMPTOTIC DERIVATION}

### 2021-10-14 aistats.tex
Update fig 1 legend

@@
 \includegraphics[width=.4\textwidth]{fewsamples.pdf}
-	\caption{Suboptimality of a Gaussian variance MLE against number of samples $n$.
-Bold curve is average over 100 trials,  dark shaded area is 90\% (dark) confidence interval, light shade is min-max interval.
-		While the expected value is infinite for $n=1$ or $n=2$, it quickly matches the upper bound, and the $1/2n$ asymptote.
+	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance MLE (\S\ref{ssec:gaussian-variance}) against number of samples $n$.
+Bold curve is an average over 10000 trials,  dark shaded area is 90\% confidence interval, light shade is min-max interval.
+		The expected value is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}, while the variance remains large.
+		We wish to find similar upper bounds for a variety of exponential families.
 	}

### 2021-10-14 aistats.tex
minor modifs

--- 2021-10-14 aistats.tex
@@
 Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
+The parameter $\theta$ can also be estimated with online-learning techniques, but they are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.
 
 {\bf Metrics.}
-Once we have an estimator $\hat \nat$, we wish to assess its quality:
-how close is it from some optimal $\nat^*$.
+Once we have an estimator $\hat \nat$, we wish to assess its quality, i.e., its closeness to some optimal $\nat^*$.
 This is necessary, for instance to choose between 2 estimators (model selection).
-We distinguish two ways of doing that.
-{\bf 1)} distance in parameter space $d(\nat,\nat^*)$.
+We distinguish here two ways: \textit{disntance in parameter space} and \textit{distance between distributions}.
+{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$.
 This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
 In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
-{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
+{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
 For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
@@
 }
-has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction} , the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
+has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
 With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
@@
 
-We have asymptotic solutions (\S\ref{ssec:asymptote}),
-or when $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
-or close to quadratic (see \S\ref{ssec:quadratic}),
-but a  general solution for finite $n$ remains elusive.
-In this paper, we review partial solutions, and give some ideas on how to solve it.
+We have already asymptotic solutions (\S\ref{ssec:asymptote}) when $\logpart$ is quadratic (e.g. $X$ is Gaussian with known variance)
+or close to quadratic (\S\ref{ssec:quadratic}).
+However, a  general solution for finite $n$ remains elusive.
+In this paper, we review partial solutions, and give ideas on how to solve it.
 
@@
 \alignn{
-	\min_{\theta\in \Theta} f(\theta)
+	\min_{\theta\in \Theta} f(\theta)\,,
 	\label{eq:optimization_problem}
@@
 Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
-Unfortunately, none of the current analysis apply, which frames another open problem for the analysis of SMD.
+Unfortunately, none of the current analyses apply, which frames another open problem for the analysis of SMD.
 
 {\bf Expected Ouctomes.}
-A solution to~\eqref{problem} would quantify the importance of the prior in MAP, in particular in the few sample regime, for instance for Gaussians $\cN(m,\sigma^2)$. 
-More generally it could enable stochastic optimization to tackle a wide class of barrier objectives\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}.
+A solution to~\eqref{problem} would quantify the importance of the prior in MAP, in particular in the few sample regime.%, for instance for Gaussians $\cN(m,\sigma^2)$. 
+Also, it could enable stochastic optimization to tackle a wide class of barrier objectives\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}.
 A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
@@
 	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
-	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance})
-	\item By identifying MAP and stochastic mirror descent (SMD), we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
+	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
+	\item By linking MAP and SMD, we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
 \end{itemize}
@@
 {\bf Notation.}
-$X$ and $T=T(X)$ are random variables, while $x$ is a sample.
-$n$ is the number of samples and $d= \dim(T)$.
+$X$ and $T=T(X)$ are random variables, $x$ is a sample, $n$ is the number of samples and $d= \dim(T)$.
 $\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
@@
 \begin{align}
-    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp \\
-    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0
+    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \\
+    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
 \end{align}
@@
 %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
-{\bf The Bregman Divergence} induced by $\logpart$,
-measures the discrepancy between two parameters $\nat$ and $\nat_0$,
+{\bf The Bregman Divergence} induced by $\logpart$ measures the discrepancy between two parameters $\nat$ and $\nat_0$,
 \begin{align}
@@
 A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
-Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
+Its density reads
+\[
+	p(x) = \inv{\sqrt{2\pi \sigma^2}} e^{-\frac{x^2}{2 \sigma^2}}.
+\]
 Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
@@
 Now we can match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
-\aligns{
-	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
-	\conj(\m) = - \half \log(\m)  \; ,
-}
-up to constants.
+\begin{align}
+	\logpart (\nat) &= - \half \log(-\nat)  + \half \log(\pi), \\
+	\conj(\m) &= - \half \log(\m) + \half \log\frac{\pi}{2} - \half  \; .
+\end{align}
 Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
@@
 \label{thm:varianceMAP}
-The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
+For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
  \begin{align}
@@
 Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
-In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth
-This correlates well with our numerical observations (cf App.\ref{app:gaussian-variance}).
+In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth, which correlates well with numerical observations (cf App.\ref{app:gaussian-variance}).
+%For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$.
+%Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.
 
@@
 	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.
-	}
-	\label{fig:thales}
+	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.}
+	\label{fig:bias-variance-numerical}
 \end{figure*}
@@
 \label{ssec:gaussian}
-The ubiquitous Gaussian $\cN(m,\sigma^2)$ actually offers a highly non-trivial example for Problem 1.
-Its log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
-With sufficient statistic $T(x)=(x, x^2)$, 
-the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
-and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$. 
-$\cM$ and $\Theta$ are delimited in grey in \cref{fig:thales}.
-Given these parameters, log-partition and entropy are
-\alignn{
-	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
-	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2)
-}
-up to constants.
-
- Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
+The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
 
@@
 We plot the bias-variance decompositions for $\cN(0,\sigma^2)$ (Figure~\ref{fig:variance_decomposition}) and  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
-Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:thales} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) .
+Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:bias-variance-numerical} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) .
 \rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}
@@
 
-\textbf{Caveat :} all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.
+Caveat : all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.
 
@@
 
-\section{Related Work}
-
-\paragraph{Online Learning}
-Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
-\[
-R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
-\]
-where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
+%\section{Related Work}
+
+%\paragraph{Online Learning}
+%Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
+%\[
+%R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
+%\]
+%where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
 Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.

### 2021-10-14 aistats.tex

@@
 \label{ssec:gaussian}
-The simple Gaussian $\cN(m,\sigma^2)$ is highly non-trivial. Neither its log-partition nor its entropy are smooth, strongly-convex, or self-concordant.
-
-Equipped with the knowledge of these 2 examples, we are ready to discuss the general problem, and some ways to solve with direct expansions of the Bregman.
+Now that we have solved the case of $\cN(0,\sigma^2)$, let's move on to the ubiquitous Gaussian $\cN(m,\sigma^2)$.
+Gaussians actually offers a highly non-trivial example for Problem 1.
+Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
+With sufficient statistic $T(x)=(x, x^2)$, 
+the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
+and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$. 
+Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
+Given these parameters, log-partition and entropy are, up to constants, 
+\alignn{
+	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
+	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2)
+}
+
+These functions are neither smooth, nor strongly convex, nor self-concordant (see \cref{app:gaussian}).
+Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.
 

### 2021-10-14 aistats.tex
Add transitions in partial solutions

--- 2021-10-14 aistats.tex
@@
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-We now review classical asymptotic results.
-We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
-Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
-With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
-We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
+We now detail a classical asymptotic result.
+Proofs are in \cref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
+
+Statistics typically give results on $\nat$, but the MAP~\eqref{eq:defMAP} is more simply expressed with $\meanp$, so let us focus on $\bregmanconj$.
+Bregman divergences are locally quadratic, as seen via a second order Taylor expansion
 \alignn{
@@
 }
-Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show that MLE and MAP verify
+Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show  that 
 \begin{align}
@@
 \end{align}
-Detailed derivation is in~\cref{app:asymptote}
 Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
-This asymptote is also independent of the optimum $\meanp^*$.
+For well-specified models, this asymptote is also independent of the optimum $\meanp^*$. or $\mF$.
+Let us see another example for which it is trivial to get a rate matching~\eqref{eq:asymptote}.
 
@@
 	=\frac{\Var(T)}{2n}
-	=\frac{1}{2n}
-\end{align}
-Adding a reference mean $\m_0$ to get the MAP yields to
+	=\frac{d}{2n}
+\end{align}
+Adding a reference mean $\m_0$ to get the MAP yields
 \begin{align}
@@
 \end{align}
-In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
+So we see a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$.
+Can we generalize these results to other families ?
 If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.
@@
 \end{align}
-so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
-Note that none of the standard exponential families verify this assumption.
-\TODO{connection with optim non-smooth rate.}
-
-{\bf If $\conj$ is $L$-smooth}\footnote{e.g. $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
+so $\bregmanconj$ is $2L$-Lipschitz, and~\eqref{eq:MAP_quadratic} yields a $O(\inv{\sqrt{n}})$ rate.
+However none of the common exponential families verify this assumption.
+
+{\bf If $\conj$ is $L$-smooth}\footnote{$\conj$ is $L$-smooth iff $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
 \begin{align}
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
-Besides the Gaussian with known variance, no standard exponential family verify this assumption either.
-Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
-\TODO{connection with optim smooth rate.}
+It is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
+The problem is that, besides the Gaussian with known variance, no standard exponential family has a strongly convex log-partition function.
+So let us focus on a more realistic set of assumptions in the next section.
 
@@
 \label{ssec:local-quadratic}
-FromTaylor expansion~~\eqref{eq:bregmanTaylor},
+From Taylor expansion~\eqref{eq:bregmanTaylor},
 we know that all Bregman divergences are locally quadratic.
-Under some assumptions, such as self-concordance of $\conj$\footnote{
+Under some assumptions, such as self-concordance\footnote{
 In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
-} \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
+} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
 \begin{proposition}
-Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
+Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
+Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
 \aligns{
@@
 \end{proposition}
-\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
 To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
@@
 For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
-This means that on average, we need much more samples than the dimension to reach the quadratic regime.
+This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime where a formula like~\eqref{eq:MAP_quadratic} is valid.
 

### 2021-10-14 aistats.tex
pass up to section 4

--- 2021-10-14 aistats.tex
@@
     \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
-    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; .
+    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; ,
     \label{eq:defMAP}
@@
 where $T_i=T(X_i)$.
-When $n_0=0$ -- we have no samples from the prior -- we recover the MLE~\eqref{eq:defMLE}.
+When $n_0=0$ (no samples from the prior), we recover the MLE~\eqref{eq:defMLE}.
 We write $\MAPt$ for the MAP, and view the MLE as a special case.
-We are now ready to formalize the main problem of this paper.
 
@@
 
-Assume we observe a dataset $\mD$ drawn i.i.d. from an exponential family distribution
-with parameters $\nat^*$, $p(\cdot \cond\nat^*)$.
+We are now ready to formalize the main problem of this paper. Assume we observe a dataset $\mD$ drawn i.i.d. from $p(\cdot \cond\nat^*)$, an exponential family distribution
+with parameters $\nat^*$.
 We wish to quantify how well the MLE or a MAP models the true distribution.
@@
 In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
-With exponential families, the KL is also a Bregman divergence
+With exponential families, the KL is also a Bregman divergence:
 \alignn{
@@
 }
-where the second equality is a general property of Bregman divergences and convex conjugates. (Be careful about the order of arguments.)
+The second equality is a general property of Bregman divergences and convex conjugates. % (Be careful about the order of arguments.)
 How does this quantity behave when $\nat$ is the MLE or MAP?
@@
 Even worse, with categorical variables, there is a non-zero probability to never sample one of the categories.
-In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$,
-and the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
+In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$.
+Therefore, the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
 Instead of taking the expectation, one might want to bound the risk in high probability,
@@
 	\nabla\conj(\hat \nat_{n})
-	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
+	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
 }
@@
 \begin{problem}[Convergence rate for SMD]
-Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(m, \sigma^2)$.
+Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(\meanp, \sigma^2)$.
 \end{problem}

### 2021-10-14 aistats.tex
small shortening on 6

--- 2021-10-14 aistats.tex
@@
 As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-We now detail a classical asymptotic result.
-Proofs are in \cref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
-
-Statistics typically give results on $\nat$, but the MAP~\eqref{eq:defMAP} is more simply expressed with $\meanp$, so let us focus on $\bregmanconj$.
-Bregman divergences are locally quadratic, as seen via a second order Taylor expansion
+We now review classical asymptotic results.
+We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
+Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
+With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
+We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
 \alignn{
@@
 }
-Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show  that 
+Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show that MLE and MAP verify
 \begin{align}
@@
 \end{align}
+Detailed derivation is in~\cref{app:asymptote}
 Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
-For well-specified models, this asymptote is also independent of the optimum $\meanp^*$. or $\mF$.
-Let us see another example for which it is trivial to get a rate matching~\eqref{eq:asymptote}.
+This asymptote is also independent of the optimum $\meanp^*$.
 
@@
 	=\frac{\Var(T)}{2n}
-	=\frac{d}{2n}
-\end{align}
-Adding a reference mean $\m_0$ to get the MAP yields
+	=\frac{1}{2n}
+\end{align}
+Adding a reference mean $\m_0$ to get the MAP yields to
 \begin{align}
@@
 \end{align}
-So we see a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$.
-Can we generalize these results to other families ?
+In both cases, we have a variance term in $O(n^{-1})$ and a bias term in $O(n^{-2})$. Ideally, such result should also similarly for arbitrary exponential families.
 If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.
@@
 \end{align}
-so $\bregmanconj$ is $2L$-Lipschitz, and~\eqref{eq:MAP_quadratic} yields a $O(\inv{\sqrt{n}})$ rate.
-However none of the common exponential families verify this assumption.
-
-{\bf If $\conj$ is $L$-smooth}\footnote{$\conj$ is $L$-smooth iff $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
+so $\bregmanconj$ is $2L$-Lipschitz, and we get a $O(\inv{\sqrt{n}})$ rate.
+Note that none of the standard exponential families verify this assumption.
+\TODO{connection with optim non-smooth rate.}
+
+{\bf If $\conj$ is $L$-smooth}\footnote{e.g. $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
 \begin{align}
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
-It is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
-The problem is that, besides the Gaussian with known variance, no standard exponential family has a strongly convex log-partition function.
-So let us focus on a more realistic set of assumptions in the next section.
+Besides the Gaussian with known variance, no standard exponential family verify this assumption either.
+Note that it is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
+\TODO{connection with optim smooth rate.}
 
@@
 \label{ssec:local-quadratic}
-From Taylor expansion~\eqref{eq:bregmanTaylor},
+FromTaylor expansion~~\eqref{eq:bregmanTaylor},
 we know that all Bregman divergences are locally quadratic.
-Under some assumptions, such as self-concordance\footnote{
+Under some assumptions, such as self-concordance of $\conj$\footnote{
 In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
-} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
+} \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
 \begin{proposition}
-Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
-Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
+Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
 \aligns{
@@
 \end{proposition}
+\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
 To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
@@
 For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
-This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime where a formula like~\eqref{eq:MAP_quadratic} is valid.
+This means that on average, we need much more samples than the dimension to reach the quadratic regime.
 
@@
 
-As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
-This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
-We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
+As introduced in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
+This means that we may obtain a convergence rate for MAP from an optimization analysis, 
+and any insights gained from the MAP may inform further design and analysis of SMD.
+We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, 
+before investigating 3 analysis of SMD with the MAP.
 \rlp{mention that examples have so far been lacking for relative SMD.}
@@
 %Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
-and SMD
-\citep{nemirovski2009robust,ghadimi2012optimal}
+and SMD \citep{nemirovski2009robust,ghadimi2012optimal}
 are typically encountered in non-smooth (online) optimization,
@@
 \citep[Th.4.2(MD) \& Th.6.3(SMD)]{bubeck2015convex}.
-In our case, none of these assumptions may hold.
+In our case, these assumptions do not hold.
 For instance $\logpart = -\log$ is neither smooth nor strongly convex.
@@
 Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
-\emph{relative} to $\logpart$, defined as
+\emph{relative} to a reference function $\logpart$, defined as
 \aligns{
@@
 When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
-Those conditions ensure the linear convergence of MD with $A$ as the reference function,
-thus extending the theory to a family of non-smooth functions $f$ and non-strongly convex potential $\logpart$
+Those conditions ensure the linear convergence of MD with mirror map $\nabla A$,
+extending the theory to non-smooth functions $f$ and non-strongly convex potential $\logpart$
 \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
@@
 \aligns{
-	f(\theta) = A(\theta) - \expect{\lin{X, \theta}}
+	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
 }
@@
 
-\textbf{Caveat :} all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.
+\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.
 
@@
 %thus showing positivity.
-Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size, by applying polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
-
-In general however, this assumption hardly holds.
+Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with $O(1/n)$ step-sizes, 
+by applying tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
+But in general, this assumption rarely holds.
 For the MAP, this translates into $\forall n \geq 1,$
@@
 
-\citet{dragomir2021fast} introduce a much weaker assumption $\forall n \geq 1, \forall \MAPm,$
+\citet{dragomir2021fast} use the weaker assumption 
 \alignn{
@@
 		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
-	} \leq 2 \lr^2 C \; .
+	} \leq 2 \lr^2 C \; ,
+	\quad
+	\forall n \geq 1.
 	\label{eq:dragomir}
@@
 Under this assumption, they prove convergence up to a variance ball with a constant step-size.
-Their descent lemma \citep[Eq.(12)]{dragomir2021fast} is the exact analog  of a modern analysis of SGD \citep[Th.3.2]{gower2019sgd}, meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.
-
-This seems to be a perfect solution to our problem.
-Yet it suffers from the exact same qualm as \citet{hanzely2018fastest}.
-For barrier objectives, \cref{eq:dragomir} cannot hold uniformly on every possible path of iterates, because these paths can get arbitrarily close from boundaries where $\bregmanconj$ becomes infinite.
+Their descent lemma \citep[Eq.(12)]{dragomir2021fast} 
+is the analog of a modern analysis of SGD~\citep[Th.3.2]{gower2019sgd}, 
+meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.
+
+This seems to be a solution to our problem, but it suffers from the same issues as \citet{hanzely2018fastest}. 
+For barrier objectives, \cref{eq:dragomir} does not hold uniformly on every path of iterates, 
+because these paths can get arbitrarily close to boundaries where $\bregmanconj$ diverges.
 

### 2021-10-14 aistats.tex
clean up locally quadratic

--- 2021-10-14 aistats.tex
@@
 For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
-This means that on average, we need much more samples than the dimension to reach the quadratic regime.
-
-We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
-This is true for
+This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
+
+The entropy $\conj$ is self-concordant for a lot of common families.
+For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
+as is the case with:
 exponential distributions,
@@
 or Weibull with known shape $k$.
-The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic}.
-This is true for categorical or Dirichlet distributions.
-
-With some more work, it is possible to get a wealth of results under this assumption.
-For instance \citet{ostrovskii2021finite} characterizes the number of samples needed to reach this quadratic behavior for arbitrary parametric models, under the assumptions of self-concordance, or pseudo self-concordance \citep{bach2010self}.
-Or in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher order moments of $\logpart$ in $\nat^*$, to quantify when $\logpart$ can be sandwiched between quadratics in $\nat$, but this does not directly translate into quadratics in $\m$.
-
-All these works give \textit{large sample} results -- results that hold only for $n\geq N$ for some constant $N$ -- but none of them  apply to small $n$.
-
-\citet{anastasiou2017bounds} and
-\citet{marteauferey2019beyond})
-\fdk{
-We know that if the problem is strongly convex (in $\norm{\cdot}_2$), averaged SGD achieves $O(1/n)$ rate.
-[Robust stochastic approximation approach to stochastic programming].
-(The paper uses a bounded stochastic gradient assumption, but can be relaxed to bounded variance instead.) However, strong-convexity in the Euclidean norm is not strictly necessary to find an algorithm with convergence rate of $O(1/n)$, as shown by [Non-strongly-convex smooth stochastic approximation with convergence rate O(1/n)] for linear and logistic regression.
-Their proof techniques based on self-concordance have been extended to other setting, [Finite-sample analysis of M-estimators using self-concordance, Learning Exponential Families in High-Dimensions: Strong Convexity and Sparsity].
-The same tools are used in statistics, and evething essentially boils down to controlling how different we are from a Gaussian rv, that we're going to get by the CLT, like [Bounds for the normal approximation of the maximum likelihood estimator, Anastasiou].
-}
+The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g. categorical and Dirichlet distributions.
+
+There are quite a few pieces of work in this direction. 
+\citet{ostrovskii2021finite} characterizes the number of samples needed to reach the asymptotic quadratic behavior, for any parametric models with a self-concordant log-likelihood.
+\citet{anastasiou2017bounds} obtains a similar flavor of result under other assumptions on the third derivative of $f$.
+More closely, in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher order moments of $\logpart$ in $\nat^*$, to obtain a similar result.
+However these results are expressed with quadratics in $\nat$, not $\m$, and they do not directly translate to convergence rates for the MAP, but they might with some more work.
+
+These results partially answer~\eqref{problem}. 
+Partially, because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
+None of them apply to small $n$.
+That is one caveat. Another one is the gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
+A full solution to~\eqref{problem} would address both of these caveats.
+Informed by the properties we have seen so far, let us investigate a bias-variance decomposition to formulate a conjecture.
+%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.
 
@@
 
-As introduced in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
-This means that we may obtain a convergence rate for MAP from an optimization analysis, 
-and any insights gained from the MAP may inform further design and analysis of SMD.
-We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, 
-before investigating 3 analysis of SMD with the MAP.
+As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
+This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
+We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
 \rlp{mention that examples have so far been lacking for relative SMD.}
@@
 %Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
-and SMD \citep{nemirovski2009robust,ghadimi2012optimal}
+and SMD
+\citep{nemirovski2009robust,ghadimi2012optimal}
 are typically encountered in non-smooth (online) optimization,
@@
 \citep[Th.4.2(MD) \& Th.6.3(SMD)]{bubeck2015convex}.
-In our case, these assumptions do not hold.
+In our case, none of these assumptions may hold.
 For instance $\logpart = -\log$ is neither smooth nor strongly convex.
@@
 Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
-\emph{relative} to a reference function $\logpart$, defined as
+\emph{relative} to $\logpart$, defined as
 \aligns{
@@
 When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
-Those conditions ensure the linear convergence of MD with mirror map $\nabla A$,
-extending the theory to non-smooth functions $f$ and non-strongly convex potential $\logpart$
+Those conditions ensure the linear convergence of MD with $A$ as the reference function,
+thus extending the theory to a family of non-smooth functions $f$ and non-strongly convex potential $\logpart$
 \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
@@
 \aligns{
-	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
+	f(\theta) = A(\theta) - \expect{\lin{X, \theta}}
 }
@@
 
-\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.
+\textbf{Caveat :} all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.
 
@@
 %thus showing positivity.
-Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with $O(1/n)$ step-sizes, 
-by applying tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
-But in general, this assumption rarely holds.
+Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size, by applying polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
+
+In general however, this assumption hardly holds.
 For the MAP, this translates into $\forall n \geq 1,$
@@
 
-\citet{dragomir2021fast} use the weaker assumption 
+\citet{dragomir2021fast} introduce a much weaker assumption $\forall n \geq 1, \forall \MAPm,$
 \alignn{
@@
 		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
-	} \leq 2 \lr^2 C \; ,
-	\quad
-	\forall n \geq 1.
+	} \leq 2 \lr^2 C \; .
 	\label{eq:dragomir}
@@
 Under this assumption, they prove convergence up to a variance ball with a constant step-size.
-Their descent lemma \citep[Eq.(12)]{dragomir2021fast} 
-is the analog of a modern analysis of SGD~\citep[Th.3.2]{gower2019sgd}, 
-meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.
-
-This seems to be a solution to our problem, but it suffers from the same issues as \citet{hanzely2018fastest}. 
-For barrier objectives, \cref{eq:dragomir} does not hold uniformly on every path of iterates, 
-because these paths can get arbitrarily close to boundaries where $\bregmanconj$ diverges.
+Their descent lemma \citep[Eq.(12)]{dragomir2021fast} is the exact analog  of a modern analysis of SGD \citep[Th.3.2]{gower2019sgd}, meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.
+
+This seems to be a perfect solution to our problem.
+Yet it suffers from the exact same qualm as \citet{hanzely2018fastest}.
+For barrier objectives, \cref{eq:dragomir} cannot hold uniformly on every possible path of iterates, because these paths can get arbitrarily close from boundaries where $\bregmanconj$ becomes infinite.
 

### 2021-10-14 aistats.tex

--- 2021-10-14 aistats.tex
@@
 For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
-This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime where a formula like~\eqref{eq:MAP_quadratic} is valid.
-
-We know that $\conj$ is self-concordant when $T$ is 1-dimensional and $\logpart$ is self-concordant.
-This is true for
+This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
+
+The entropy $\conj$ is self-concordant for a lot of common families.
+For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
+as is the case with:
 exponential distributions,

### 2021-10-14 aistats.tex
update section 6

--- 2021-10-14 aistats.tex
@@
 \alignn{
-    \bregmanconj(\m^* ; \m)
-    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
+    \textstyle \bregmanconj(\m^* ; \m)
+    &\textstyle = \frac{1}{2}\norm{\m^* - \m}^2_{\mF}
     + O(\norm{\m - \m^*}^3),
@@
 }
-Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show  that 
+Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, we have
 \begin{align}
@@
 \end{align}
-Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
-For well-specified models, this asymptote is also independent of the optimum $\meanp^*$. or $\mF$.
-Let us see another example for which it is trivial to get a rate matching~\eqref{eq:asymptote}.
+Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible for large $n$.
+For well-specified models, this asymptote is also independent of the optimum $\meanp^*$ or $\mF$.
+Let us see another example for which we get a rate matching~\eqref{eq:asymptote}.
 
@@
 \label{ssec:quadratic}
-As another classical reference point, let us consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
+As another classical reference point, we consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
 For instance, this is the log-partition of a Gaussian with known variance $I$,
@@
 \]
-Then $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances:
+In this case, $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances since
 \begin{align}
@@
 \end{align}
-Thanks to the independence of samples, we can break down the MLE into individual point's contributions
+Thanks to the independence of samples, we can break down the MLE into individual point's contributions:
 \begin{align}
@@
 	=\frac{\Var(T)}{2n}
-	=\frac{d}{2n}
+	=\frac{d}{2n}.
 \end{align}
@@
 \end{align}
-So we see a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$.
-Can we generalize these results to other families ?
-If we make restrictive assumptions on the log-partition function, we can relate $\bregmanconj$ to a norm or a quadratic.
+We see here a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$.
+However, this result does not generalize well to other families, unless we make restrictive assumptions on $\conj$.
+% we can relate $\bregmanconj$ to a norm or a quadratic.
+% Can we generalize these results to other families?
 
@@
     &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
-    &\leq 2L \norm{\m^* - \m}
+    &\leq 2L \norm{\m^* - \m},
 \end{align}
@@
     \bregmanconj(\m^* ; \m)
-    \leq \frac{L}{2} \norm{\m^* - \m}^2
+    \leq \frac{L}{2} \norm{\m^* - \m}^2,
 \end{align}
@@
 It is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
-The problem is that, besides the Gaussian with known variance, no standard exponential family has a strongly convex log-partition function.
-So let us focus on a more realistic set of assumptions in the next section.
+The problem is that, besides the Gaussian with known variance, no standard exponential family has a \textit{global} strongly convex log-partition function.
+The next section focuses on \textit{local} quadratic behavior, which is more realistic.
+% So let us focus on a more realistic set of assumptions in the next section.
 
@@
 Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
-Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
+Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
 \aligns{
@@
 	\implies
-	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
+	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
 }
@@
 In both the quadratic~\eqref{eq:MAP_quadratic} and the Gaussiance variance examples~\eqref{eq:MAP_rate}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$, 
-giving us a flavor of what we would like as a general result for exponential families : a finite sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
+giving us a flavor of what we would like as a general result for exponential families: a finite sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
 In fact, such a decomposition exists for any Bregman divergence.
@@
 However, this is not true for the the gaussian variance MLE, where the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
-A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.
+A contrario, the MLE is unbiased w.r.t. the dual parameter, but the decomposition there is not clean.
 
@@
 
-As introduced in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
-This means that we may obtain a convergence rate for MAP from an optimization analysis, 
-and any insights gained from the MAP may inform further design and analysis of SMD.
-We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, 
-before investigating 3 analysis of SMD with the MAP.
+As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
+This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
+We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three analyses of SMD with the MAP.
 \rlp{mention that examples have so far been lacking for relative SMD.}
@@
 \citep[Th.4.2(MD) \& Th.6.3(SMD)]{bubeck2015convex}.
-In our case, these assumptions do not hold.
+In our case, none of these assumptions may hold.
 For instance $\logpart = -\log$ is neither smooth nor strongly convex.
 
-Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
-\emph{relative} to a reference function $\logpart$, defined as
+Recent works extend these assumptions to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
+\emph{relative} to $\logpart$,
 \aligns{
@@
 
-\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.
+\textbf{Caveat :} all of these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.
 
@@
 %thus showing positivity.
-Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with $O(1/n)$ step-sizes, 
-by applying tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
-But in general, this assumption rarely holds.
+Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with a $O(1/n)$ step-size, by applying polynomial tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
+
+In general however, this assumption hardly holds.
 For the MAP, this translates into $\forall n \geq 1,$
 \alignn{
-\expect[T_n]{\lin{T_n - \mu^* , \MAPt}} \leq \lr_n C
-}
-which fails to hold on the gaussian variance example for instance. \TODO{ DAMIEN short proof or formalize.}
+\expect[T_n]{\lin{T_n - \meanp^* , \MAPt}} \leq \lr_n C
+}
+which fails to hold on the gaussian variance example for instance.
+Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero.
+Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.
+
+%  \TODO{ DAMIEN short proof or formalize.}
 

### 2021-10-14 aistats.tex

--- 2021-10-14 aistats.tex
@@
 
-\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging in the primal.
+\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.
 
@@
 \alignn{
-\expect[T_n]{\lin{T_n - \mu^* , \MAPt}} \leq \lr_n C
-}
-which fails to hold on the gaussian variance example for instance. \TODO{ DAMIEN short proof or formalize.}
+\expect[T_n]{\lin{T_n - \meanp^* , \MAPt}} \leq \lr_n C
+}
+which fails to hold on the gaussian variance example for instance.
+Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero.
+Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.
+
+%  \TODO{ DAMIEN short proof or formalize.}
 

### 2021-10-15 aistats.tex
Start editing bias-variance

--- 2021-10-14 aistats.tex
@@
 Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$. 
-\textcolor{red}{The explanation here is not clear enough, or too concise} 
-Now we can match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
+Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
 \aligns{
@@
 	\bregmanconj( \m_*; \m_n)
-	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right), \\
-	\bregman( \nat_n; \nat_* )
-	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
+	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) \; .
+%	\bregman( \nat_n; \nat_* )
+%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
 \end{align}
 In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
-All proofs for this section are in \cref{app:gaussian-variance}.
+All proofs for this section are in App.\ref{app:gaussian-variance}.
 \begin{theorem}[MLE Bound]
@@
 \end{theorem}
-Note that this inequality actually holds for the symmetrized Bregman $\cB(a,b) + \cB(b,a)$, which explains why it loses a factor 2 compared to the asymptote~\eqref{eq:asymptote}, but we leave this to the appendix for clarity.
+Note that this inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
+This explains the loss of a factor 2 compared to the asymptote~\eqref{eq:asymptote}. 
 Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
@@
 
-These functions are neither smooth, nor strongly convex, nor self-concordant (see \cref{app:gaussian}).
+These functions are neither smooth, nor strongly convex, nor self-concordant (see App.\ref{app:gaussian}).
 Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.
@@
 We now detail a classical asymptotic result.
-Proofs are in \cref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
+Proofs are in App.\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
 
@@
 In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
-} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
+} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.\ref{app:self-concordant}.
 \begin{proposition}
@@
 A full solution to~\eqref{problem} would address both of these caveats.
-Informed by the properties we have seen so far, let us investigate a bias-variance decomposition to formulate a conjecture.
+Informed by the properties we have seen so far, let us formulate a conjecture.
 %If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.
@@
 giving us a flavor of what we would like as a general result for exponential families : a finite sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
-In fact, such a decomposition exists for any Bregman divergence.
-Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be its corresponding mean parameter.
-As described by \citet[Theorem 0.1]{pfau2013generalized}, the  expected Bregman decomposes into
+In fact, such a decomposition exists for any Bregman divergence \citep[Theorem 0.1]{pfau2013generalized}.
+\begin{theorem}[Bregman Bias-Variance Decomposition]
+	Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be its dual representation.
+The  expected Bregman decomposes into
 \begin{equation}
@@
 \end{equation}
-
-\textbf{Remark:} In this decomposition, the primal expectation $\expect{\hat \theta_n}$ is taking the role of pivot / reference point.
-An estimator will be unbiased if its primal expectation is non-zero.
-However, this is not true for the the gaussian variance MLE, where the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$.
-A contrario, the MLE is unbiased wrt the dual parameter, but the decomposition there is not clean.
+\end{theorem}
+
+\textbf{Remark:} In this decomposition, the primal expectation $\expecti{\hat \theta_n}$ is the reference point. 
+An estimator will be unbiased if $\tilde \nat_n = \nat^*$.
+This is not true for the MLE, which is unbiased w.r.t. the dual parameter $\expecti{\hat \m_n}=\m^*$.
+
+We show in App.\ref{app:bias-variance} that the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$ for Gaussian variance MAP, and
+ ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{ \norm{\m^* - \m_0}^2 }{(1 + \frac{n}{n_0})^2 }$ for quadratics.
 
@@
 \paragraph{Illustrations.}
-We plot the bias-variance decompositions for $\cN(0,\sigma^2)$ (Figure~\ref{fig:variance_decomposition}) and  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
+We plot the bias-variance decompositions  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
 Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:thales} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) .
@@
 
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=.4\textwidth]{figs/gaussian_variance.pdf}
-	\caption{
-	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound.
-	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
-	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
-	}
-	\label{fig:variance_decomposition}
-\end{figure}
+%\begin{figure}[t]
+%	\centering
+%	\includegraphics[width=.4\textwidth]{figs/gaussian_variance.pdf}
+%	\caption{
+%	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound.
+%	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
+%	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
+%	}
+%	\label{fig:variance_decomposition}
+%\end{figure}
 
@@
 	\caption{
-	\textbf{Full gaussian} $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. \textbf{Left:} training curves (5,95) percentiles, average and asymptote.
-	\textbf{Right:} bias-variance decomposition.
-	}
+	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$.
+	} %TODO this figure has no labels.t
 	\label{fig:gaussian_decomposition}
@@
 
+\section{BIAS-VARIANCE}
+\label{app:bias-variance}
+
 \section{TODO}

### 2021-10-15 aistats.tex
Conclude bias-variance

@@
 A full solution to~\eqref{problem} would address both of these caveats.
-Informed by the properties we have seen so far, let us formulate a conjecture.
+Informed by the properties we have seen so far, let us investigate a general decomposition of the Bregman that could guide us towards a solution.
 %If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.
@@
 \end{theorem}
+We plot this decomposition for $\cN(\mu, \sigma^2)$ in Fig.\ref{fig:gaussian_decomposition} , and we illustrate the primal mean $\tilde \nat_n$ in  Fig.\ref{fig:thales}.
 
@@
 
-We show in App.\ref{app:bias-variance} that the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$ for Gaussian variance MAP, and
- ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{ \norm{\m^* - \m_0}^2 }{(1 + \frac{n}{n_0})^2 }$ for quadratics.
-
-\textbf{Conjecture:} the bias decreases like $O(1/n^2)$.
-
-\paragraph{Illustrations.}
-We plot the bias-variance decompositions  $\cN(\mu, \sigma^2)$ (Figure~\ref{fig:gaussian_decomposition}).
-Additionally,  for $\cN(\mu, \sigma^2)$, Figure~\ref{fig:thales} plots the characters featured in these decompositions $\hat \m_n^\text{MLE},\hat \m_n^\text{MAP},\m_n,\tilde \m_n, \m^*$ and $\m_0$ (and corresponding primal parameters) .
-\rlp{Tune figures.}\textcolor{red}{Not super clear, what is the "characters featured in these decompositions?"}
-
-%\begin{figure}[t]
-%	\centering
-%	\includegraphics[width=.4\textwidth]{figs/gaussian_variance.pdf}
-%	\caption{
-%	\textbf{Gaussian variance $\cN(0,\sigma^2)$ example. Left:} training curves and analytic upper bound.
-%	\textbf{Center:} bias-mixed-variance decomposition, using the arithmetic mean.
-%	\textbf{Right:} bias-variance decomposition, using the harmonic mean.
-%	}
-%	\label{fig:variance_decomposition}
-%\end{figure}
+We show in App.\ref{app:bias-variance} that the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$ for Gaussian variance MLE, and
+ ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{ \norm{\m^* - \m_0}^2 }{(1 + \frac{n}{n_0})^2 }$ for a quadratic MAP.
+These observations are hinting towards a general $O(1/n^2)$ upper bound for the bias, while the variance may be less dependent on the initialization $\nat_0$.
 
@@
 	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$.
-	} %TODO this figure has no labels.t
+	} 
 	\label{fig:gaussian_decomposition}
 \end{figure}
+
+In this section, we considered direct expansions of~\eqref{eq:bregmanMAP}.
+None of them could really solve~\eqref{problem}. 
+Is it possible to solve this problem through an optimization approach ?
 

### 2021-10-15 aistats.tex
Pass on introduction.

--- 2021-10-15 aistats.tex
@@
 
-{\bf Exponential families} are among the most widely used statistical models.
-Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
+{\bf Exponential families} are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
+Many standard random variables are exponential families: Gaussians, categorical, gamma or Dirichlet for example.
 They are flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
-In particular, they are appreciated for their convex log-likelihood
-\alignn{
-f(\nat) = \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
+They are particularly appreciated for their convex log-likelihood
+\alignn{
+f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
 \label{eq:defNLL}
 }
-where $\logpart$ is a convex function and $\nat\in\Theta$ is a vector called \textit{natural parameter}.
+where $\logpart$ is the convex log-partition function and $\nat\in\Theta$ is the \textit{natural parameter}.
 This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized}
@@
 {\bf Estimators.}
-In general, we want to estimate $\nat$ from some data $\mD = (x_1, \dots, x_n)$.
-For this purpose, not only is $f$ convex, but it even enjoys a closed form maximum-likelihood estimate (MLE)
-\begin{align}
-	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
+In this paper, we consider the problem of estimating $\nat$ from a dataset $\mD = (X_1, \dots, X_n)$ of iid observations from $p_\theta$ in an exponential family.
+In this case, not only is $f$ convex, but it yields a simple condition for the maximum-likelihood estimate (MLE)
+\begin{align}
+ \hat \mu_n^\text{MLE} = \nabla  \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
 	\label{eq:defMLE}
 \end{align}
-This rule is known as moment matching.
+This rule is also known as moment matching.
 Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
-The parameter $\theta$ can also be estimated with online-learning techniques, but they are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.
-
-{\bf Metrics.}
-Once we have an estimator $\hat \nat$, we wish to assess its quality, i.e., its closeness to some optimal $\nat^*$.
-This is necessary, for instance to choose between 2 estimators (model selection).
-We distinguish here two ways: \textit{disntance in parameter space} and \textit{distance between distributions}.
+In this paper, we will focus on analyzing MLE and MAP estimators.\footnote{A related analysis is present in the online-learning literature, but for different online estimators which are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.}
+
+{\bf Statistical decision theory.}
+To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
+We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
 {\bf 1)} Distance in parameter space $d(\nat,\nat^*)$.
@@
 For $p$ and $q$ continuous densities,
-$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \&p(x)>0$.
+$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
 %Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
@@
 With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
-Hence the fundamental problem:
+We thus consider the fundamental problem:
 %\begin{center}
@@
 
-We have already asymptotic solutions (\S\ref{ssec:asymptote}) when $\logpart$ is quadratic (e.g. $X$ is Gaussian with known variance)
+There are already general asymptotic results (\S\ref{ssec:asymptote}), and a finite $n$ result when $\logpart$ is quadratic (e.g. $X$ is Gaussian with known variance)
 or close to quadratic (\S\ref{ssec:quadratic}).
@@
 \alignn{
-	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
+	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} } .
 	\label{eq:suboptimalityKL}
@@
 Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
-In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
-Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
-Unfortunately, none of the current analyses apply, which frames another open problem for the analysis of SMD.
-
-{\bf Expected Ouctomes.}
-A solution to~\eqref{problem} would quantify the importance of the prior in MAP, in particular in the few sample regime.%, for instance for Gaussians $\cN(m,\sigma^2)$.
-Also, it could enable stochastic optimization to tackle a wide class of barrier objectives\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}.
+In particular, with exponential families, MAP is an instantiation of the so-called stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
+Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we consider using existing optimization convergence rates for SMD to get the upper bound we seek.
+Unfortunately, none of the current analyses apply, highlighting open problems for the analysis of SMD.
+
+{\bf Expected Outcomes.}
+A solution to~\eqref{problem} can clarify the importance of the prior in MAP, in particular in the few sample regime. %, for instance for Gaussians $\cN(m,\sigma^2)$.
+Also, it could enable stochastic optimization to tackle a wide class of barrier objectives.\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}
 A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
@@
 	\itemsep0em
-	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
+	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}), illustrating that tight rates are possible even though the current theory does not cover them.
 	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
-	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
+	%SLJ REMOVED per Slack discussion
+	%\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
 	\item By linking MAP and SMD, we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).

### 2021-10-15 aistats.tex
Update figure 3 with legend

--- 2021-10-15 aistats.tex
@@
 	\centering
-	\includegraphics[width=\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\caption{A numerical illustration of the different characters featured in the bias-variance decomposition, for a 1D Gaussian $\cN(\mu, \sigma^2)$.
+	\includegraphics[width=.8\textwidth]{figs/thales/numerical_schema_n=3.pdf}
+	\caption{
+	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MAP (blue) and MLE (orange) as detailed in \S\ref{ssec:gaussian} with $n=3$ . 
+	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
+	But in primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition.
+	The hyperparameter of the prior $\nat_0$ is controlling the location of the brown point, while varying $n_0$ spans the Thales triangle and the red-curve.
+	Large blurry circles in the background are other instances of MAP and MLE revealing the distribution of these estimators.
 	}

### 2021-10-15 aistats.tex
Quick pass on 2 and 3; add appendix F for reviewing SMD.

--- 2021-10-15 aistats.tex
@@
 \end{align}
-This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
+This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding and multivariate normal distributions $\cX=\real^d, \nu$ Lebesgue and $T(X)=(X, X X^\top)$.
 
@@
 
-{\bf Maximum A Posteriori (MAP)}
+{\bf Maximum A Posteriori (MAP).}
 Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
@@
 
-We are now ready to formalize the main problem of this paper. Assume we observe a dataset $\mD$ drawn i.i.d. from $p(\cdot \cond\nat^*)$, an exponential family distribution
+We are now ready to formalize the main problem of this paper. Assume we observe a dataset $\mD_n$ drawn i.i.d. from $p(\cdot \cond\nat^*)$, an exponential family distribution
 with parameters $\nat^*$.
-We wish to quantify how well the MLE or a MAP models the true distribution.
+We wish to quantify how well the MLE or a MAP approximates the true distribution.
 
 A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
-In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
+In the well-specified setting, it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
 With exponential families, the KL is also a Bregman divergence:
@@
 The second equality is a general property of Bregman divergences and convex conjugates. % (Be careful about the order of arguments.)
-How does this quantity behave when $\nat$ is the MLE or MAP?
+How does this quantity behave when $\hat \nat$ is the MLE or MAP?
  Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators when the loss is the KL divergence?
@@
 	\label{eq:bregmanMLE}
-	\text{MLE: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
+	\text{MLE: } \quad &\expect[\mD_n]{\bregmanconj \left (\E_{\nat^*} [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
 	\label{eq:bregmanMAP}
-	\text{MAP: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
-\end{align}
-where the expectation is on the data $\mD = (T_1, \dots, T_n)$.
+	\text{MAP: } \quad &\expect[\mD_n]{\bregmanconj \left (\E_{\nat^*} [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
+\end{align}
+where the expectation is on the data $\mD_n = (T_1, \dots, T_n)$.
 \end{problem}
@@
 
-\paragraph{Optimization}
-With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD).}
+\paragraph{Optimization.}
+With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD)}, reviewed in App.~\ref{app:SMD}.
 More precisely, let us re-write~\eqref{eq:defMAP} as
@@
 In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
-The observation MAP$\in$SMD brings us to our second problem.
+The observation MAP $\in$ SMD brings us to our second problem.
 \begin{problem}[Convergence rate for SMD]
@@
 
+\section{REVIEW OF SMD}
+\label{app:SMD}
+
 \section{TODO}

### 2021-10-15 aistats.tex
Pass on 4

@@
 
-To address these problems, we start by investigating examples, for which we provide solutions to problem 1 and get insights into what is achievable.
-
-\section{Examples}\label{sec:example}
-\subsection{Gaussian Variance}\label{ssec:gaussian-variance}
+To address these problems, we start by investigating simple examples for which we can provide solutions to Problem 1, getting insights into what is achievable.
+
+\section{Illustrating Examples}\label{sec:example}
+\subsection{Gaussian with Unknown Variance}\label{ssec:gaussian-variance}
 
@@
 Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
-It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative Gamma distribution.
+It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative gamma distribution.
 It also means $\bregman$ and $\bregmanconj$ have the same shape
 \begin{align}
-	\bregmanconj( \m_*; \m_n)
-	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) \; .
+	\bregmanconj( \m^*; \m_n)
+	&= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
 %	\bregman( \nat_n; \nat_* )
@@
 \end{align}
-In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
-All proofs for this section are in App.\ref{app:gaussian-variance}.
+In Theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
+All proofs for this section are in App.~\ref{app:gaussian-variance}.
 \begin{theorem}[MLE Bound]
 \label{thm:varianceMLE}
-	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
+	The MLE of $\cN(0,\m^*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
 	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
 	\begin{align}
-		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
+		 \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
 			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
@@
 \end{theorem}
-This upper bound matches the asymptote~\eqref{eq:asymptote} .
+This upper bound matches the asymptotic result~\eqref{eq:asymptote} that we derive in \S\ref{ssec:asymptote}.
 We illustrate its numerical behavior in Figure~\ref{fig:curves}.
@@
  \begin{align}
-	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
+	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
 	\leq \begin{cases}
@@
 Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
-In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth
-This correlates well with our numerical observations (cf App.\ref{app:gaussian-variance}).
+In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth. 
+This correlates well with our numerical observations (cf App.~\ref{app:gaussian-variance}).
 
@@
 
-\subsection{Gaussian}
+\subsection{Full Gaussian (Non-Trivial)}
 \label{ssec:gaussian}
-Now that we have solved the case of $\cN(0,\sigma^2)$, let's move on to the ubiquitous Gaussian $\cN(m,\sigma^2)$.
-Gaussians actually offers a highly non-trivial example for Problem 1.
+Now that we have solved the case of $\cN(0,\sigma^2)$, let us consider the full Gaussian $\cN(m,\sigma^2)$ which actually offers a highly non-trivial example for Problem 1.
 Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
@@
 
-These functions are neither smooth, nor strongly convex, nor self-concordant (see App.\ref{app:gaussian}).
+These functions are neither smooth, nor strongly convex, nor self-concordant (see App.~\ref{app:gaussian}).
 Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.

### 2021-10-15 aistats.tex
Pass on 6.1

--- 2021-10-15 aistats.tex
@@
 \end{align}
-This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding and multivariate normal distributions $\cX=\real^d, \nu$ Lebesgue and $T(X)=(X, X X^\top)$.
+This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.
 
@@
 
-{\bf Maximum A Posteriori (MAP).}
+{\bf Maximum A Posteriori (MAP)}
 Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
@@
 
-We are now ready to formalize the main problem of this paper. Assume we observe a dataset $\mD_n$ drawn i.i.d. from $p(\cdot \cond\nat^*)$, an exponential family distribution
+We are now ready to formalize the main problem of this paper. Assume we observe a dataset $\mD$ drawn i.i.d. from $p(\cdot \cond\nat^*)$, an exponential family distribution
 with parameters $\nat^*$.
-We wish to quantify how well the MLE or a MAP approximates the true distribution.
+We wish to quantify how well the MLE or a MAP models the true distribution.
 
 A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
-In the well-specified setting, it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
+In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
 With exponential families, the KL is also a Bregman divergence:
@@
 The second equality is a general property of Bregman divergences and convex conjugates. % (Be careful about the order of arguments.)
-How does this quantity behave when $\hat \nat$ is the MLE or MAP?
+How does this quantity behave when $\nat$ is the MLE or MAP?
  Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators when the loss is the KL divergence?
@@
 	\label{eq:bregmanMLE}
-	\text{MLE: } \quad &\expect[\mD_n]{\bregmanconj \left (\E_{\nat^*} [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
+	\text{MLE: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
 	\label{eq:bregmanMAP}
-	\text{MAP: } \quad &\expect[\mD_n]{\bregmanconj \left (\E_{\nat^*} [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
-\end{align}
-where the expectation is on the data $\mD_n = (T_1, \dots, T_n)$.
+	\text{MAP: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
+\end{align}
+where the expectation is on the data $\mD = (T_1, \dots, T_n)$.
 \end{problem}
@@
 
-\paragraph{Optimization.}
-With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD)}, reviewed in App.~\ref{app:SMD}.
+\paragraph{Optimization}
+With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD).}
 More precisely, let us re-write~\eqref{eq:defMAP} as
@@
 In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
-The observation MAP $\in$ SMD brings us to our second problem.
+The observation MAP$\in$SMD brings us to our second problem.
 \begin{problem}[Convergence rate for SMD]
@@
 	\caption{
-	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$.
+	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
 	} 
@@
 
-As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
+As we saw in \S\ref{sec:problem}, MAP can be interpreted as stochastic mirror descent (SMD).
 This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
-We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three analyses of SMD with the MAP.
-\rlp{mention that examples have so far been lacking for relative SMD.}
+We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating recent analyses of SMD with the MAP.
 
@@
 When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
-Those conditions ensure the linear convergence of MD with mirror map $\nabla A$,
-extending the theory to non-smooth functions $f$ and non-strongly convex potential $\logpart$
-\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
-
-For exponential families, the log-likelihood perfectly fit into this framework, as
+These relative conditions ensure the linear convergence of MD with mirror map $\nabla A$
+\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively},
+even when $f$ is not smooth and $\logpart$ is not strongly convex.
+
+For exponential families, MAP perfectly fits into this framework, as
 \aligns{
@@
 is both $1$-smooth and $1$-strongly convex relative to $A$.
-So we are left with finding an applicable convergence rate for SMD under relative assumptions.
-As we investigate in the next section, several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
+Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
+Several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, 
+but they have so far been lacking concrete examples.
+In the next section, we challenge their hypothesis thanks with the MAP.
 
 \subsection{Bounding the Randomness}
-\TODO{compress this section}
 
 To analyze stochastic algorithms, one also need to quantify the randomness.
-Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
-We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
-See \cref{tbl:assumptions} for a summary.
-
-\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.
+While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness.
+We review these in the light of  the MAP and provide a summary in \cref{tbl:assumptions}.
 
 \begin{table}[t]
-\begingroup
-%\renewcommand*{\arraystretch}{1.25}%
 \newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
 \newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
-\centering
 \caption{Summary of results for SMD
 under relative smoothness and relative strong convexity assumptions.
-Barrier tells whether the assumption can hold with barrier objectives.
-$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
-}
-\begin{tabular}{lccc}
+Each row correspond to one analysis, and each columns answers one question.
+($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
+($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
+($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^*,\hat\nat_n)$? 
+ ($\hat\nat_n$) is it for the last iterate or an average ?
+None of these analysis check all the boxes needed to address~\eqref{problem}.
+}
+\begin{center}
+\begin{tabular}{lcccc}
 %{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
 \toprule
-Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
+Boundedness & $-\log$ &  $\lr_n \sim \inv{n}$ & $f$ & $\hat\nat_n$ \\
 \midrule
@@
 Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
-& \redxmark & \greencmark  & \redxmark
+& \redxmark & \greencmark & \greencmark  & \redxmark
 \\
 Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
-& \redxmark & \greencmark  & \greencmark
+& \redxmark & \greencmark & \redxmark  & \greencmark
 \\
 Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
-& \greencmark & \redxmark & \greencmark
+& \greencmark & \redxmark & \redxmark & \greencmark
 \\
@@
 \end{tabular}
+\end{center}
 \label{tbl:assumptions}
-\endgroup
 \end{table}
@@
 \label{app:bias-variance}
-
-\section{REVIEW OF SMD}
-\label{app:SMD}
 

### 2021-10-15 aistats.tex
minor changes

@@
 \paragraph{Optimization.}
-With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD)}, reviewed in App.~\ref{app:SMD}.
+With exponential families, MAP can be linked to \emph{stochastic mirror descent (SMD)}, see App.~\ref{app:SMD}.
 More precisely, let us re-write~\eqref{eq:defMAP} as
@@
 up to constants.
-Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
+Both $A$ and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
 It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative gamma distribution.

### 2021-10-15 aistats.tex
grammarly pass + squeeze to get 8 pages (almost done!)

@@
 We consider the problem of upper bounding the expected log-likelihood sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family, in a non-asymptotic way.
-Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian, or in the very interesting few samples regime.
-After exhibiting various facets of the problem, we show the MAP can be interpreted as running stochastic mirror descent (SMD) on the log-likelihood,
-yet modern convergence results do not apply for standard examples of exponential family, highlighting holes in the convergence literature.
+Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian or in the interesting few samples regime.
+After exhibiting various facets of the problem, we show we can interpret the MAP as running stochastic mirror descent (SMD) on the log-likelihood.
+However, modern convergence results do not apply for standard examples of the exponential family, highlighting holes in the convergence literature.
 We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
@@
 	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance MLE (\S\ref{ssec:gaussian-variance}) against number of samples $n$.
-Bold curve is an average over 10000 trials,  dark shaded area is 90\% confidence interval, light shade is min-max interval.
+The bold curve plots the average over 10000 trials,  dark shaded area is 90\% confidence interval, and light shade is a min-max interval.
 		The expected value is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}, while the variance remains large.
@@
 {\bf Exponential families} are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
-Many standard random variables are exponential families: Gaussians, categorical, gamma or Dirichlet for example.
-They are flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
+Many standard random variables are exponential families: Gaussians, categorical, gamma, or Dirichlet, for example.
+They are flexible enough to model a variety of data sources $X$ and easy to describe with some sufficient statistics $T(X) \in \real^d$.
 They are particularly appreciated for their convex log-likelihood
@@
 where $\logpart$ is the convex log-partition function and $\nat\in\Theta$ is the \textit{natural parameter}.
-This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized}
-or variants of principal component analysis \citep{collins2001generalization}, among other applications.
+This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} or variants of principal component analysis \citep{collins2001generalization}, among other applications.
 
@@
 This rule is also known as moment matching.
-Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
-In this paper, we will focus on analyzing MLE and MAP estimators.\footnote{A related analysis is present in the online-learning literature, but for different online estimators which are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.}
+Given a specific conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
+In this paper, we will focus on analyzing MLE and MAP estimators.\footnote{A related analysis is present in the online-learning literature, but for different online estimators, which are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.}
 
@@
 
-There are already general asymptotic results (\S\ref{ssec:asymptote}), and a finite $n$ result when $\logpart$ is quadratic (e.g. $X$ is Gaussian with known variance)
+There are already general asymptotic results (\S\ref{ssec:asymptote}), and a finite $n$ result when $\logpart$ is quadratic (e.g., $X$ is Gaussian with known variance)
 or close to quadratic (\S\ref{ssec:quadratic}).
 However, a  general solution for finite $n$ remains elusive.
-In this paper, we review partial solutions, and give ideas on how to solve it.
+In this paper, we review partial solutions and give ideas on how to solve the problem.
 
@@
 Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
-In particular, with exponential families, MAP is an instantiation of the so-called stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
-Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we consider using existing optimization convergence rates for SMD to get the upper bound we seek.
+In particular, with exponential families, MAP is equivalent to stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
+Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we consider using existing convergence rates for SMD to get the upper bound we seek.
 Unfortunately, none of the current analyses apply, highlighting open problems for the analysis of SMD.
@@
 A solution to~\eqref{problem} can clarify the importance of the prior in MAP, in particular in the few sample regime. %, for instance for Gaussians $\cN(m,\sigma^2)$.
-Also, it could enable stochastic optimization to tackle a wide class of barrier objectives.\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}
-A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
-It could also help in assessing the impact of alternative forms of regularization (prior) for these models.
+Also, it could enable stochastic optimization to tackle a broad class of barrier objectives.\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}
+A good example is the generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
+It could also help assess the impact of alternative forms of regularization (prior) for these models.
 
@@
 	\itemsep0em
-	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}), illustrating that tight rates are possible even though the current theory does not cover them.
+	\item We provide an upper bound on the KL in the particular case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}), illustrating that tight rates are possible even though the current theory does not cover them.
 	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
@@
 	%\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
-	\item By linking MAP and SMD, we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
+	\item By linking MAP and SMD, we show that modern analysis of SMD is yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
 \end{itemize}
@@
 \label{sec:background}
-In this section, we review the formalism of exponential families, their duality, a conjugate prior and the corresponding MAP.
-We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a more exhaustive introduction.
+This section reviews the formalism of exponential families, their duality, a conjugate prior, and the corresponding MAP.
+We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a more detailed introduction.
 
@@
 where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
-Given that $\logpart$ is strictly convex, its Hessian is positive definite and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
-We will write interchangeably $\m$ or  $\nat$ depending on the context, being aware that both are linked and represent the same distribution.
+Given that $\logpart$ is strictly convex, its Hessian is positive definite, and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
+We will write $\m$ or $\nat$ interchangeably depending on the context, being aware that both are linked and represent the same distribution.
 
@@
 When $n_0=0$ (no samples from the prior), we recover the MLE~\eqref{eq:defMLE}.
-We write $\MAPt$ for the MAP, and view the MLE as a special case.
+We write $\MAPt$ for the MAP and view the MLE as a particular case.
 
@@
 More explicitly, we want an upper bound that does not involve this expectation over the dataset.
-Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our own means.
+Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our means.
 In \S\ref{sec:example}, we provide results for special cases such as $\cN(0, \sigma^2)$,
-while in \S\ref{sec:insights} we provide realistic conditions to obtain bounds that are valid after seeing a large number of samples.
-But we are yet to find a solution encompassing both a broad range of exponential families,
+while in \S,\ref{sec:insights} we provide realistic conditions to obtain valid bounds after seeing a large number of samples.
+However, we have yet to find a solution encompassing both a broad range of exponential families
 and applicable to small sample sizes $n \lesssim d$.
@@
 for instance when estimating the covariance of a Gaussian when $n \leq d + 1$.
-Even worse, with categorical variables, there is a non-zero probability to never sample one of the categories.
+Even worse, there is a non-zero probability never to sample one of the categories with categorical variables.
 In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$.
 Therefore, the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
-Instead of taking the expectation, one might want to bound the risk in high probability,
-without resorting to Markov inequality but this is a difficult endeavor.
+Instead of taking the expectation, one might want to bound the risk in high probability
+without resorting to Markov inequality, but this is a difficult endeavor.
 These examples make a case for regularized estimators such as MAP,
@@
 and step-size schedule $\lr_n$, initialized at $\nat_0$.
-In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
+In this view, MLE forgets its (arbitrary) initialization after the first step with step size 1.
 The observation MAP $\in$ SMD brings us to our second problem.
@@
 
-To address these problems, we start by investigating simple examples for which we can provide solutions to Problem 1, getting insights into what is achievable.
+To address these problems, we start by investigating simple examples to provide solutions to Problem 1, getting insights into what is achievable.
 
@@
 
-A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
+A non-trivial yet straightforward example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
 Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
@@
 \end{theorem}
-Note that this inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
+This inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
 This explains the loss of a factor 2 compared to the asymptote~\eqref{eq:asymptote}.
-Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights a clear variance-bias decomposition.
+Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights an explicit variance-bias decomposition.
 In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth. 
@@
 	\caption{
-	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MAP (blue) and MLE (orange) as detailed in \S\ref{ssec:gaussian} with $n=3$ . 
+	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MAP (blue) and MLE (orange) (\S\ref{ssec:gaussian} with $n=3$). 
 	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
-	But in primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition.
-	The hyperparameter of the prior $\nat_0$ is controlling the location of the brown point, while varying $n_0$ spans the Thales triangle and the red-curve.
+	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition.
+	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the Thales triangle and the red curve.
 	Large blurry circles in the background are other instances of MAP and MLE revealing the distribution of these estimators.
@@
 \label{ssec:gaussian}
-Now that we have solved the case of $\cN(0,\sigma^2)$, let us consider the full Gaussian $\cN(m,\sigma^2)$ which actually offers a highly non-trivial example for Problem 1.
+Now that we have solved the case of $\cN(0,\sigma^2)$, consider the full Gaussian $\cN(m,\sigma^2)$, which offers a highly non-trivial example for Problem 1.
 Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
@@
 \alignn{
-	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
-	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2)
+	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
+	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
 }
@@
 These functions are neither smooth, nor strongly convex, nor self-concordant (see App.~\ref{app:gaussian}).
-Let us keep this in mind, as we move on to discussing the general problem, and some ways to solve it via direct expansions of the Bregman divergence.
+We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.
 
@@
 Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible for large $n$.
-For well-specified models, this asymptote is also independent of the optimum $\meanp^*$ or $\mF$.
+This asymptote is independent of the optimum $\meanp^*$ or $\mF$ for well-specified models.
 Let us see another example for which we get a rate matching~\eqref{eq:asymptote}.
@@
 We see here a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$.
-However, this result does not generalize well to other families, unless we make restrictive assumptions on $\conj$.
+However, this result does not generalize well to other families unless we make restrictive assumptions on $\conj$.
 % we can relate $\bregmanconj$ to a norm or a quadratic.
@@
 \end{align}
-so $\bregmanconj$ is $2L$-Lipschitz, and~\eqref{eq:MAP_quadratic} yields a $O(\inv{\sqrt{n}})$ rate.
-However none of the common exponential families verify this assumption.
+so $\bregmanconj$ is Lipschitz, and~\eqref{eq:MAP_quadratic} yields a $O(\inv{\sqrt{n}})$ rate, but no common exponential families verify the assumption.
 
@@
 It is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
-The problem is that, besides the Gaussian with known variance, no standard exponential family has a \textit{global} strongly convex log-partition function.
+Besides the Gaussian with known variance, the problem is that no standard exponential family has a \textit{global} strongly convex log-partition function.
 The next section focuses on \textit{local} quadratic behavior, which is more realistic.
@@
 For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
-This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
+This means that, on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper bound like~\eqref{eq:MAP_quadratic}.
 
 The entropy $\conj$ is self-concordant for a lot of common families.
-For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
+For instance, when $T$ is 1-dimensional, and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
 as is the case with:
@@
 or Weibull with known shape $k$.
-The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g. categorical and Dirichlet distributions.
+The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g., categorical and Dirichlet distributions.
 
 There are quite a few pieces of work in this direction.
-\citet{ostrovskii2021finite} characterizes the number of samples needed to reach the asymptotic quadratic behavior, for any parametric models with a self-concordant log-likelihood.
+\citet{ostrovskii2021finite} characterizes the number of samples needed to reach the asymptotic quadratic behavior for any parametric models with a self-concordant log-likelihood.
 \citet{anastasiou2017bounds} obtains a similar flavor of result under other assumptions on the third derivative of $f$.
-More closely, in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher order moments of $\logpart$ in $\nat^*$, to obtain a similar result.
-However these results are expressed with quadratics in $\nat$, not $\m$, and they do not directly translate to convergence rates for the MAP, but they might with some more work.
+More closely, in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher-order moments of $\logpart$ in $\nat^*$, to obtain a similar result.
+However, these results are expressed with quadratics in $\nat$, not $\m$, and they do not directly translate to convergence rates for the MAP, but they might with some more work.
 
 These results partially answer~\eqref{problem}.
-Partially, because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
+Partially because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
 None of them apply to small $n$.
@@
 \label{ssec:bias-variance}
-In both the quadratic~\eqref{eq:MAP_quadratic} and the Gaussiance variance examples~\eqref{eq:MAP_rate}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$,
-giving us a flavor of what we would like as a general result for exponential families : a finite sample convergence rate, with a variance and a bias terms that reflect the important constants of the problem.
-In fact, such a decomposition exists for any Bregman divergence \citep[Theorem 0.1]{pfau2013generalized}.
+In both the quadratic~\eqref{eq:MAP_quadratic} and the Gaussian variance examples~\eqref{eq:MAP_rate}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$,
+giving us a flavor of what we would like as a general result for exponential families: a finite sample convergence rate, with variance and bias terms that reflect the important constants of the problem.
+Such a decomposition exists for any Bregman divergence \citep[Theorem 0.1]{pfau2013generalized}.
 \begin{theorem}[Bregman Bias-Variance Decomposition]
@@
  ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{ \norm{\m^* - \m_0}^2 }{(1 + \frac{n}{n_0})^2 }$ for a quadratic MAP.
-These observations are hinting towards a general $O(1/n^2)$ upper bound for the bias, while the variance may be less dependent on the initialization $\nat_0$.
+These observations hint towards a general $O(1/n^2)$ upper bound for the bias, while the variance may be less dependent on the initialization $\nat_0$.
 
@@
 None of them could really solve~\eqref{problem}. 
-Is it possible to solve this problem through an optimization approach ?
+Is it possible to solve this problem through an optimization approach?
 
@@
 This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
-We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating recent analyses of SMD with the MAP.
+We first review the assumptions of relative smoothness, helpful to deal with non-smooth functions, before investigating recent analyses of SMD with the MAP.
 
@@
 }
-When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
-These relative conditions ensure the linear convergence of MD with mirror map $\nabla A$
+When $\logpart = \norm{\cdot}^2$, we recover the standard smoothness and gradient descent.
+These conditions ensure the linear convergence of MD with mirror map $\nabla A$
 \citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively},
-even when $f$ is not smooth and $\logpart$ is not strongly convex.
+even when $f$ is not smooth, and $\logpart$ not strongly convex.
 
@@
 }
-is both $1$-smooth and $1$-strongly convex relative to $A$.
+is $1$-smooth and $1$-strongly convex relative to $A$.
 Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
-Several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, 
-but they have so far been lacking concrete examples.
-In the next section, we challenge their hypothesis thanks with the MAP.
+Several pieces of work address this question under various assumptions on stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, 
+but they have so far lacked concrete examples.
+In the next section, we challenge their hypothesis thanks to the MAP.
 
@@
 
-To analyze stochastic algorithms, one also need to quantify the randomness.
+To analyze stochastic algorithms, one also needs to quantify the randomness.
 While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness.
@@
 by applying tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
-But in general, this assumption rarely holds.
+However, in general, this assumption rarely holds.
 For the MAP, this translates into $\forall n \geq 1,$
@@
 }
-which fails to hold on the gaussian variance example for instance.
+which fails to hold on the gaussian variance example, for instance.
 Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero.
@@
 }
-In words, this bounds the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
+This bounds the variance of the gradients at the minimum with a metric depending on the iterate $\MAPm$. This bound should hold for every iterate we may encounter on the stochastic trajectory.
 % As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality.
-Under this assumption, they prove convergence up to a variance ball with a constant step-size.
+Under this assumption, they prove convergence up to a variance ball with a constant step size.
 Their descent lemma \citep[Eq.(12)]{dragomir2021fast}
-is the analog of a modern analysis of SGD~\citep[Th.3.2]{gower2019sgd},
+is the analog of modern analysis of SGD~\citep[Th.3.2]{gower2019sgd},
 meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.
@@
 where $f_X$ is a stochastic estimate of $f = \expecti{f_X}$. In our case $f_X(\nat) = - \log p(X\cond \nat)$.
-In words, this lower bounds the expectation of the minimum of the stochastic estimates.
+In other words, this lower bounds the expectation of the minimum of the stochastic estimates.
 For probabilistic models, such a bound is finite as soon as the model cannot give infinite density to any data point $x$.
-This holds for instance for discrete distributions because the probability mass is upper bounded by $1$,
-but it rules out  many families.
+This holds, for instance, for discrete distributions because the probability mass is upper bounded by $1$; however, it rules out  many families.
 In the case of normal distributions$\cN(m, \sigma^2)$, setting $\mu=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$.
@@
 It is possible to overcome this limitation by treating batches of samples as single samples by averaging sufficient statistics, e.g.,. $Y = \{X_1, \dots, X_k\}$ and $T(Y) = \inv{k}\sum_i T(X_i)$.
-For instance, a multivariate normals of dimension $d$ cannot attribute infinite density to $d+1$ samples that are not in an affine subspace.
-
-Overall, this approach seems to answer~\eqref{problem}.
-Yet it fails to account for the step-size $\lr_n = \inv{n_0+n}$, as \citet[Thm.1]{dorazio2021stochastic} only proves linear convergence to a variance ball of size $\frac{C}{\stgcvx}$ under constant step-size $\lr \leq \inv{\smooth}$.
+For instance, a multivariate normal of dimension $d$ cannot attribute infinite density to $d+1$ samples that are not in an affine subspace.
+
+Overall, this approach seems to answer~\eqref{problem}, but fails to account for the step-size $\lr_n = \inv{n_0+n}$, as \citet[Thm.1]{dorazio2021stochastic} only proves linear convergence to a variance ball of size $\frac{C}{\stgcvx}$ under constant step size.%  $\lr \leq \inv{\smooth}$.
 
@@
 
-\fdk{They don't hold for problems we care about}
+\fdk{They do not hold for problems we care about}
 
@@
 
-If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let's assume they hold on average.
-Does the proof remain valid ?
-Yes I think so \TODO{check this}.
-Does it become impossible to prove / as hard as the initial problem ?
+If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let us assume they hold on average.
+Does the proof remain valid?
+Yes, I think so \TODO{check this}.
+Does it become impossible to prove / as hard as the initial problem?
 Yes.
@@
 \section{Conclusion}
-Exponential family MLE and MAP are used everywhere -- everytime someone computes a mean and variance to fit some bell-shaped data.
+Exponential family MLE and MAP are used everywhere -- every time someone computes a mean and variance to fit some bell-shaped data.
 MLE and MAP are both maximizing the (regularized) likelihood of the data, whose suboptimality is the Kullback-Leibler (KL) divergence.
-The KL is a very natural measure of discrepancy between a model and the true disribution.
-Yet we do not have convergence rates on the expected KL, even in the well-specified setting.
-Such a rate could inform us on how many samples we may need, or on the influence of the prior.
-There are many results close from such a rate, in particular large sample results \citep{kakade2010learning, ostrovskii2021finite}, which describe how many samples are needed to ensure a locally quadratic regime.
+The KL is a very natural measure of discrepancy between a model and the true distribution.
+However, we do not have convergence rates on the expected KL, even in the well-specified setting.
+Such a rate could inform us on how many samples we may need or on the influence of the prior.
+There are many results close to such a rate, particularly large sample results \citep{kakade2010learning, ostrovskii2021finite}, which describe how many samples are needed to ensure a locally quadratic regime.
 Another approach to obtain rates is stochastic optimization \citep{bach2013nonstronglyconvex}.
 Here we observed that MAP perfectly fits the framework of stochastic mirror descent with relative smoothness assumptions.
-Yet none of the current analysis of SMD hold for the MAP, even on a simple family such as $\cN(0,\sigma^2)$, thus revealing area for progress in non-Euclidean optimization.
+Nevertheless, none of the current analyses of SMD hold for the MAP, even on a simple family such as $\cN(0,\sigma^2)$, thus revealing an area for progress in non-Euclidean optimization.
 In writing this paper, we hope to attract attention to this fundamental problem, leading to progress in both optimization and statistics.
@@
 	\includegraphics[width=.4\textwidth]{phi.pdf}
-	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it is poorly approximated by quadratics, but it admits another upper-bound (in grey).}
+	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it admits another upper bound (in grey) despite being poorly approximated by quadratics.}
 	\label{fig:phi}
@@
 \begin{enumerate}
-	\item new plots : add 2d trajectories to paper, with background level set, and new legend : primal/dual expectation, and clear explanation !
+	\item new plots: add 2d trajectories to paper, with background level set, and new legend: primal/dual expectation, and clear explanation!
 	\item Incorporate related work by Frederik.
-	\item Technical Background : be precise and add reference for the steep exponential family of Legendre type which guarantees that our sets are open, and have a solution in the middle.
+	\item Technical Background: be precise and add a reference for the steep exponential family of Legendre type, which guarantees that our sets are open and have a solution in the middle.
 	References: \begin{itemize}
@@
 		\item The singly truncated normal distribution: A non-steep exponential family,
-		\item Statistical exponential families: A digest with flash cards,
+		\item Statistical exponential families: A digest with flashcards,
 		\item Information and Exponential Families In Statistical Theory,
@@
 %
-There is a long line work bounding the convergence rate of stochastic gradient descent.
+There is a long line of work bounding the convergence rate of stochastic gradient descent.
 Beyond the results of \citet{robbins1951stochastic},
@@
 Those papers assume the stochastic gradients are bounded,
-but small modifications also work for bounded variance instead.
+but minor modifications also work for bounded variance instead.
 \\
 We do have extensions to other notions of bounded variance,
-for example assuming that the stochastic gradients are gradients of a perturbed smooth function $f_i$
+for example, assuming that the stochastic gradients are gradients of a perturbed smooth function $f_i$
 and that the minimum of $f$ and the minima of the $f_i$
@@
 This is the case for maximum likelihood estimation and matches the asymptotic rates,
-but holds more generally, for example for linear and logistic regressions \citep{bach2013nonstronglyconvex,moulines2011non}.
+but holds more generally, for example, for linear and logistic regressions \citep{bach2013nonstronglyconvex,moulines2011non}.
 }
@@
 A ``simple open problem'';
-assume we have a deterministic problem, so that we know where the minimum is.
+assume we have a deterministic problem so that we know where the minimum is.
 Can we figure out a (deterministic) path from $\theta_0$ to $\theta_*$
@@
 The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity.
-For example, we haven't yet figured out the analog of Nesterov-type acceleration
-on relatively smooth and strongly convex problems,
+For example, we have not yet figured out the analog of Nesterov-type acceleration
+on relatively smooth and strongly convex problems
 to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$,

### 2021-10-15 aistats.tex
Partial pass on 5.1-5.2-5.3. pushing for damien.

--- 2021-10-15 aistats.tex
@@
 
-\subsection{Asymptote}
+\subsection{Asymptotic Rate}
 \label{ssec:asymptote}
@@
 We now detail a classical asymptotic result.
-Proofs are in App.\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
+Proofs are in App.~\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
 
@@
 }
-where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
+where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the Hessian of the entropy at the optimum.
 It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
@@
 For well-specified models, this asymptote is also independent of the optimum $\meanp^*$ or $\mF$.
-Let us see another example for which we get a rate matching~\eqref{eq:asymptote}.
+Next, we give another example for which we get a rate matching~\eqref{eq:asymptote}.
 
@@
 \begin{align}
-	\expect{\bregmanconj(\meanp^*; \MAPm)}
-	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
+		\!\!\!\!\! \expect{\half \norm{\m^* -   \MAPm^\text{MAP}}_2^2} \!
+	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.\!
 	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
@@
 In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
-} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.\ref{app:self-concordant}.
+} of $\conj$ \citep[Ch.~4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.~\ref{app:self-concordant}.
 \begin{proposition}
 Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
-Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
+Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
 \aligns{
@@
 To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
-For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
-For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
-This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
-
-The entropy $\conj$ is self-concordant for a lot of common families.
-For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
+For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
+For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition \TODO{FIX THIS!} $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
+This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
+
+The entropy $\conj$ is self-concordant for several common families.
+For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.~\ref{app:self-concordant}),
 as is the case with:
@@
 None of them apply to small $n$.
-That is one caveat. Another one is the gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
+That is one caveat. Another one is the Gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
 A full solution to~\eqref{problem} would address both of these caveats.
@@
 	\caption{
-	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
+	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$.
 	} 
@@
 
-As we saw in \S\ref{sec:problem}, MAP can be interpreted as stochastic mirror descent (SMD).
+As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
 This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
-We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating recent analyses of SMD with the MAP.
+We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three analyses of SMD with the MAP.
+\rlp{mention that examples have so far been lacking for relative SMD.}
 
@@
 When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
-These relative conditions ensure the linear convergence of MD with mirror map $\nabla A$
-\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively},
-even when $f$ is not smooth and $\logpart$ is not strongly convex.
-
-For exponential families, MAP perfectly fits into this framework, as
+Those conditions ensure the linear convergence of MD with mirror map $\nabla A$,
+extending the theory to non-smooth functions $f$ and non-strongly convex potential $\logpart$
+\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.
+
+For exponential families, the log-likelihood perfectly fit into this framework, as
 \aligns{
@@
 is both $1$-smooth and $1$-strongly convex relative to $A$.
-Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
-Several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, 
-but they have so far been lacking concrete examples.
-In the next section, we challenge their hypothesis thanks with the MAP.
+So we are left with finding an applicable convergence rate for SMD under relative assumptions.
+As we investigate in the next section, several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
 
 \subsection{Bounding the Randomness}
+\TODO{compress this section}
 
 To analyze stochastic algorithms, one also need to quantify the randomness.
-While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness.
-We review these in the light of  the MAP and provide a summary in \cref{tbl:assumptions}.
+Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
+We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
+See \cref{tbl:assumptions} for a summary.
+
+\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.
 
 \begin{table}[t]
+\begingroup
+%\renewcommand*{\arraystretch}{1.25}%
 \newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
 \newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
+\centering
 \caption{Summary of results for SMD
 under relative smoothness and relative strong convexity assumptions.
-Each row correspond to one analysis, and each columns answers one question.
-($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
-($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
-($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^*,\hat\nat_n)$? 
- ($\hat\nat_n$) is it for the last iterate or an average ?
-None of these analysis check all the boxes needed to address~\eqref{problem}.
-}
-\begin{center}
-\begin{tabular}{lcccc}
+Barrier tells whether the assumption can hold with barrier objectives.
+$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
+}
+\begin{tabular}{lccc}
 %{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
 \toprule
-Boundedness & $-\log$ &  $\lr_n \sim \inv{n}$ & $f$ & $\hat\nat_n$ \\
+Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
 \midrule
@@
 Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
-& \redxmark & \greencmark & \greencmark  & \redxmark
+& \redxmark & \greencmark  & \redxmark
 \\
 Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
-& \redxmark & \greencmark & \redxmark  & \greencmark
+& \redxmark & \greencmark  & \greencmark
 \\
 Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
-& \greencmark & \redxmark & \redxmark & \greencmark
+& \greencmark & \redxmark & \greencmark
 \\
@@
 \end{tabular}
-\end{center}
 \label{tbl:assumptions}
+\endgroup
 \end{table}

### 2021-10-15 aistats.tex

--- 2021-10-15 aistats.tex
@@
 This asymptote is independent of the optimum $\meanp^*$ or $\mF$ for well-specified models.
-Let us see another example for which we get a rate matching~\eqref{eq:asymptote}.
+Next, we give another example for which we get a rate matching~\eqref{eq:asymptote}.
 
@@
 To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
-For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
-For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
-This means that, on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper bound like~\eqref{eq:MAP_quadratic}.
-
-The entropy $\conj$ is self-concordant for a lot of common families.
-For instance, when $T$ is 1-dimensional, and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
+For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
+For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition \TODO{FIX THIS!} $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
+This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
+
+The entropy $\conj$ is self-concordant for several common families.
+For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.~\ref{app:self-concordant}),
 as is the case with:

### 2021-10-15 aistats.tex
Merge Hanzely and Dragomir

--- 2021-10-15 aistats.tex
@@
 
-\subsection{Asymptotic Rate}
+\subsection{Asymptote}
 \label{ssec:asymptote}
@@
 We now detail a classical asymptotic result.
-Proofs are in App.~\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
+Proofs are in App.\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
 
@@
 }
-where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the Hessian of the entropy at the optimum.
+where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum.
 It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
@@
 For well-specified models, this asymptote is also independent of the optimum $\meanp^*$ or $\mF$.
-Next, we give another example for which we get a rate matching~\eqref{eq:asymptote}.
+Let us see another example for which we get a rate matching~\eqref{eq:asymptote}.
 
@@
 \begin{align}
-		\!\!\!\!\! \expect{\half \norm{\m^* -   \MAPm^\text{MAP}}_2^2} \!
-	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.\!
+	\expect{\bregmanconj(\meanp^*; \MAPm)}
+	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.
 	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
@@
 In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
-} of $\conj$ \citep[Ch.~4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.~\ref{app:self-concordant}.
+} of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.\ref{app:self-concordant}.
 \begin{proposition}
 Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
-Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
+Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}
 \aligns{
@@
 To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
-For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
-For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition \TODO{FIX THIS!} $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
-This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
-
-The entropy $\conj$ is self-concordant for several common families.
-For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.~\ref{app:self-concordant}),
+For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
+For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
+This means that on average, we need $5$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
+
+The entropy $\conj$ is self-concordant for a lot of common families.
+For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.\ref{app:self-concordant}),
 as is the case with:
@@
 None of them apply to small $n$.
-That is one caveat. Another one is the Gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
+That is one caveat. Another one is the gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
 A full solution to~\eqref{problem} would address both of these caveats.
@@
 is both $1$-smooth and $1$-strongly convex relative to $A$.
-So we are left with finding an applicable convergence rate for SMD under relative assumptions.
-As we investigate in the next section, several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
+Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
 
 \subsection{Bounding the Randomness}
-\TODO{compress this section}
 
 To analyze stochastic algorithms, one also need to quantify the randomness.
-Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
-We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
-See \cref{tbl:assumptions} for a summary.
-
-\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.
+While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, but they have so far been lacking concrete examples.
+We review these analyses in the light of the MAP and provide a summary in \cref{tbl:assumptions}.
 
 \begin{table}[t]
-\begingroup
-%\renewcommand*{\arraystretch}{1.25}%
 \newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
 \newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
-\centering
 \caption{Summary of results for SMD
 under relative smoothness and relative strong convexity assumptions.
-Barrier tells whether the assumption can hold with barrier objectives.
-$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
-}
-\begin{tabular}{lccc}
+Each row correspond to one analysis, and each columns answers one question.
+($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
+($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
+($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^*,\hat\nat_n)$? 
+ ($\hat\nat_n$) is it for the last iterate or an average ?
+None of these analysis check all the boxes needed to address~\eqref{problem}.
+}
+\begin{center}
+\begin{tabular}{lcccc}
 %{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
 \toprule
-Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
+Boundedness & $-\log$ &  $\lr_n \sim \inv{n}$ & $f$ & $\hat\nat_n$ \\
 \midrule
@@
 \end{tabular}
+\end{center}
 \label{tbl:assumptions}
-\endgroup
 \end{table}
 
-\subsubsection{Bounded Variance}
-\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step: $\forall n \geq 1, \forall \hat \nat_{n-1},$
+%Before we proceed, let $f_X$ be a stochastic estimate of $f = \expecti{f_X}$, with stochastic gradients $g_X(\nat) =\nabla f_X(\nat)$.
+In our case $f_X(\nat) = - \log p(X\cond \nat)$ and $g_X(\nat) = \m - T(X)$.
+
+\subsubsection{Analogs of the Variance}
+let us introduce $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ th symmetrized Bregman induced by $\conj$.
+\citet{hanzely2018fastest} assume that the expectation of $\cS$ between stochastic and deterministic updates verifies
 \alignn{
-	\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
+\expect[g]{\cS_\conj \paren{
+	\hat\m_{n} - \lr g(\hat \nat_{n}), 
+	\hat\m_{n}   -\lr \nabla f(\hat \nat_{n})
+}} \leq \lr^2 C
 \label{eq:hanzely}
 }
-for some constant $C$.
-For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
+for all possible iterates $\MAPt$, relevant step-sizes $\lr$ and for some constant $C$.
+When $A(\theta) = \frac{1}{2}\norm{\theta}^2$ or when $A$ is $\stgcvx$-strongly convex,
 this definition recovers the variance of the stochastic gradient
 \[
-	\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C.
+	\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla f(\theta) - g(\theta)}^2}\leq C.
 \]
-More generally, if $A$ is $\stgcvx$-strongly convex w.r.t. $\norm{\cdot}$,
-the LHS of \cref{eq:hanzely} is bounded by
-$\expecti{\frac{1}{\stgcvx}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$.
-Then a bound on the variance of stochastic gradients is sufficient for \cref{eq:hanzely}
-to hold.
-%Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
-%\aligns{
-%\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) = \E[\cS(\MAPm - \lr g_n  )]
-%}
-%thus showing positivity.
-Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with $O(1/n)$ step-sizes,
-by applying tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
-But in general, this assumption rarely holds.
-For the MAP, this translates into $\forall n \geq 1,$
-\alignn{
-\expect[T_n]{\lin{T_n - \meanp^* , \MAPt}} \leq \lr_n C
-}
-which fails to hold on the gaussian variance example for instance.
-Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero.
-Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.
-
-%  \TODO{ DAMIEN short proof or formalize.}
-
-\subsubsection{Bounded Variance at the Minimum}
-
-\citet{dragomir2021fast} use the weaker assumption
+Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate on function values with $O(1/n)$ step-sizes and tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
+
+\citet{dragomir2021fast} define the weaker assumption
 \alignn{
@@
 		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
-	} \leq 2 \lr^2 C \; ,
-	\quad
-	\forall n \geq 1.
+	} \leq 2 \lr^2 C \; .
 	\label{eq:dragomir}
 }
-In words, this bounds the variance of the gradients at the minimum, but with a metric that depends on the iterate $\MAPm$, and this bound should hold for every iterate we may encounter on the stochastic trajectory.
-% As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality.
-Under this assumption, they prove convergence up to a variance ball with a constant step-size.
-Their descent lemma \citep[Eq.(12)]{dragomir2021fast}
-is the analog of a modern analysis of SGD~\citep[Th.3.2]{gower2019sgd},
-meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.
-
-This seems to be a solution to our problem, but it suffers from the same issues as \citet{hanzely2018fastest}.
-For barrier objectives, \cref{eq:dragomir} does not hold uniformly on every path of iterates,
-because these paths can get arbitrarily close to boundaries where $\bregmanconj$ diverges.
+When $\logpart$ is $\stgcvx$-strongly-convex, we recover the variance of the gradients at the optimum
+\[
+	\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla g(\theta^*)}^2}\leq C.
+\]
+Using their descent lemma \citep[Eq.(12)]{dragomir2021fast} with the $O(1/n)$ step-size used by \citet[Th.3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^*, \hat\nat_n)$.
+
+These two analyses seem promising for~\eqref{problem}, but none of these assumptions hold in front of barrier objectives such as the $-\log$ from \S\ref{ssec:gaussian-variance}.
+Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero.
+Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.
 
@@
 
-\subsection{Issues with current analyses}
-
-\fdk{They don't hold for problems we care about}
-
-\fdk{Existing techniques only deal with fixed step-sizes and show convergence to a ball}
-
 If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let's assume they hold on average.
@@
 \TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
-
-%\section{Related Work}
-
-%\paragraph{Online Learning}
-%Other works focus instead on the online estimation of $\theta$ and minimize the \textit{regret}
-%\[
-%R_T=\sum_{t=1}^T f_t(\hat{\theta}_t)-\min_{\theta\in\Theta}\sum_{t=1}^Tf_t(\theta),
-%\]
-%where $f_t(\theta)\triangleq-\log p(x_t|\theta)$, and the estimator $\hat{\theta}_t$ is computed using only the $t-1$ samples $\{x_i\}$.
-Usually, the \textit{follow-the-leader} strategy works best, i.e., moving the current estimate towards $x_i$ with a certain step size schedule.
-Unfortunately however, their rate are worse than the offline MAP or MLE algorithms.
-For instance, in the specific case where we estimate the mean parameter of a Bernoulli or Gaussian distribution, \citet{azoury2001relative} show that $R_T\leq log(T)$.
-From this bound, we can deduce the rate  to the real mean parameter, being $f(\hat{\theta}_T)-f^*\leq \frac{log(T)}{T}$.
-Also, \citet{dasgupta2007online} prove the asymptotic rate $\lim\sup_{T>1}\frac{R_T}{T}\leq 0$ when we estimate the mean and variance of a Gaussian, under the (strong) assumption that the data are finite, i.e., $\|x_t\|<C$.
-This means the estimator converges asymptotically, but with an unknown rate of convergence, therefore highlighting the difficulty of that setting.
 

### 2021-10-15 aistats.tex

--- 2021-10-15 aistats.tex
@@
 Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
-Several pieces of work address this question under various assumptions on stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, 
-but they have so far lacked concrete examples.
-In the next section, we challenge their hypothesis thanks to the MAP.
 
@@
 
-To analyze stochastic algorithms, one also needs to quantify the randomness.
-While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness.
-We review these in the light of  the MAP and provide a summary in \cref{tbl:assumptions}.
+To analyze stochastic algorithms, one also need to quantify the randomness.
+While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, but they have so far been lacking concrete examples.
+We review these analyses in the light of the MAP and provide a summary in \cref{tbl:assumptions}.
 
@@
 
-\subsubsection{Bounded Variance}
-\citet{hanzely2018fastest} assume the covariance between negative gradient and next iterate is upper bounded at every time step: $\forall n \geq 1, \forall \hat \nat_{n-1},$
+%Before we proceed, let $f_X$ be a stochastic estimate of $f = \expecti{f_X}$, with stochastic gradients $g_X(\nat) =\nabla f_X(\nat)$.
+In our case $f_X(\nat) = - \log p(X\cond \nat)$ and $g_X(\nat) = \m - T(X)$.
+
+\subsubsection{Analogs of the Variance}
+let us introduce $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ th symmetrized Bregman induced by $\conj$.
+\citet{hanzely2018fastest} assume that the expectation of $\cS$ between stochastic and deterministic updates verifies
 \alignn{
-	\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) \leq \lr_n C
+\expect[g]{\cS_\conj \paren{
+	\hat\m_{n} - \lr g(\hat \nat_{n}),
+	\hat\m_{n}   -\lr \nabla f(\hat \nat_{n})
+}} \leq \lr^2 C
 \label{eq:hanzely}
 }
-for some constant $C$.
-For a Gaussian with known variance ($A(\theta) = \frac{1}{2}\norm{\theta}^2$),
+for all possible iterates $\MAPt$, relevant step-sizes $\lr$ and for some constant $C$.
+When $A(\theta) = \frac{1}{2}\norm{\theta}^2$ or when $A$ is $\stgcvx$-strongly convex,
 this definition recovers the variance of the stochastic gradient
 \[
-	\expect[\tilde g_t\!\!]{\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2}\leq C.
+	\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla f(\theta) - g(\theta)}^2}\leq C.
 \]
-More generally, if $A$ is $\stgcvx$-strongly convex w.r.t. $\norm{\cdot}$,
-the LHS of \cref{eq:hanzely} is bounded by
-$\expecti{\frac{1}{\stgcvx}\norm{\nabla f(\theta) - \tilde g(\theta)}{}^2_*}$.
-Then a bound on the variance of stochastic gradients is sufficient for \cref{eq:hanzely}
-to hold.
-%Another perspective on this covariance is as the expectation of the symmetrized Bregman $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ between stochastic and deterministic updates
-%\aligns{
-%\Cov(-g_n(\hat \nat_{n-1}) , \MAPt) = \E[\cS(\MAPm - \lr g_n  )]
-%}
-%thus showing positivity.
-Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate with $O(1/n)$ step-sizes,
-by applying tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
-However, in general, this assumption rarely holds.
-For the MAP, this translates into $\forall n \geq 1,$
-\alignn{
-\expect[T_n]{\lin{T_n - \meanp^* , \MAPt}} \leq \lr_n C
-}
-which fails to hold on the gaussian variance example, for instance.
-Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero.
-Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.
-
-%  \TODO{ DAMIEN short proof or formalize.}
-
-\subsubsection{Bounded Variance at the Minimum}
-
-\citet{dragomir2021fast} use the weaker assumption
+Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate on function values with $O(1/n)$ step-sizes and tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
+
+\citet{dragomir2021fast} define the weaker assumption
 \alignn{
@@
 		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
-	} \leq 2 \lr^2 C \; ,
-	\quad
-	\forall n \geq 1.
+	} \leq 2 \lr^2 C \; .
 	\label{eq:dragomir}
 }
-This bounds the variance of the gradients at the minimum with a metric depending on the iterate $\MAPm$. This bound should hold for every iterate we may encounter on the stochastic trajectory.
-% As observed by Frederick, this is really weaker than Hanzely, because it is only the first half of their assumption. There is only a factor 2 that's preventing a direct inequality.
-Under this assumption, they prove convergence up to a variance ball with a constant step size.
-Their descent lemma \citep[Eq.(12)]{dragomir2021fast}
-is the analog of modern analysis of SGD~\citep[Th.3.2]{gower2019sgd},
-meaning we can plug in a $O(1/n)$ decreasing step-size and obtain a $O(1/n)$ convergence rate.
-
-This seems to be a solution to our problem, but it suffers from the same issues as \citet{hanzely2018fastest}.
-For barrier objectives, \cref{eq:dragomir} does not hold uniformly on every path of iterates,
-because these paths can get arbitrarily close to boundaries where $\bregmanconj$ diverges.
+When $\logpart$ is $\stgcvx$-strongly-convex, we recover the variance of the gradients at the optimum
+\[
+	\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla g(\theta^*)}^2}\leq C.
+\]
+Using their descent lemma \citep[Eq.(12)]{dragomir2021fast} with the $O(1/n)$ step-size used by \citet[Th.3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^*, \hat\nat_n)$.
+
+These two analyses seem promising for~\eqref{problem}, but none of these assumptions hold in front of barrier objectives such as the $-\log$ from \S\ref{ssec:gaussian-variance}.
+Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero.
+Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.
 
@@
 
-\subsection{Issues with current analyses}
-
-\fdk{They do not hold for problems we care about}
-
-\fdk{Existing techniques only deal with fixed step-sizes and show convergence to a ball}
-
 If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let us assume they hold on average.

### 2021-10-15 aistats.tex
Pass on 5.4

--- 2021-10-15 aistats.tex
@@
 	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
-	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition.
+	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:primal_pivot} from~\S\ref{ssec:bias-variance}.
 	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the Thales triangle and the red curve.
-	Large blurry circles in the background are other instances of MAP and MLE revealing the distribution of these estimators.
+	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution.
 	}
@@
 \end{theorem}
-We plot this decomposition for $\cN(\mu, \sigma^2)$ in Fig.\ref{fig:gaussian_decomposition} , and we illustrate the primal mean $\tilde \nat_n$ in  Fig.\ref{fig:thales}.
+We plot this decomposition for $\cN(\mu, \sigma^2)$ in Fig.~\ref{fig:gaussian_decomposition} , and we illustrate the primal mean $\tilde \nat_n$ in  Fig.~\ref{fig:thales}.
 
@@
 
-We show in App.\ref{app:bias-variance} that the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$ for Gaussian variance MLE, and
+We show in App.~\ref{app:bias-variance} that the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$ for Gaussian variance MLE, and
  ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{ \norm{\m^* - \m_0}^2 }{(1 + \frac{n}{n_0})^2 }$ for a quadratic MAP.
@@
 In this section, we considered direct expansions of~\eqref{eq:bregmanMAP}.
-None of them could really solve~\eqref{problem}. 
-Is it possible to solve this problem through an optimization approach?
+None of them could fully solve~\eqref{problem}.
+Next, we investigate whether an optimization approach could solve it.
 

### 2021-10-15 aistats.tex
Polish discussion of variance flaw

--- 2021-10-15 aistats.tex
@@
 and strong convexity assumption on the potential $\logpart$
-\citep[Th.4.2(MD) \& Th.6.3(SMD)]{bubeck2015convex}.
+\citep[Th. 4.2(MD) \& Th. 6.3(SMD)]{bubeck2015convex}.
 In our case, these assumptions do not hold.
@@
 \]
-Using their descent lemma \citep[Eq.(12)]{dragomir2021fast} with the $O(1/n)$ step-size used by \citet[Th.3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^*, \hat\nat_n)$.
+Using their descent lemma \citep[Eq.
+(12)]{dragomir2021fast} with the $O(1/n)$ step-size used by \citet[Th.
+3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^*, \hat\nat_n)$.
 
 These two analyses seem promising for~\eqref{problem}, but none of these assumptions hold in front of barrier objectives such as the $-\log$ from \S\ref{ssec:gaussian-variance}.
-Indeed, we cannot control the samples $x_i$, therefore there is a non-zero probability they make $\nabla\conj(\MAPt) = \hat\meanp_t$ arbitrarily close to zero.
-Since the conjugate of the log partition is $\conj=-0.5\log$, this makes $\MAPt$ unbounded.
+Indeed, they  both assume their bound holds uniformly for every possible iterate $\hat \nat_n$.
+Yet $\cN(0,\sigma^2)$ has a positive mass around $0$.
+This means that $\hat \m_n$ can get arbitrarily close from $0$, where the $-\log$ is unbounded, along with the associated  Bregman divergences~\eqref{eq:hanzely} and~\eqref{eq:dragomir}.
+In general, this uniform bound over $\hat \m_n$ cannot hold for \emph{barrier} objectives -- functions  exploding to infinity in some finite point of space.
+
+However, both of their proofs hold if we add an expectation over $\MAPm$ to their assumption.
+But this is not helpful for us, as verifying the assumption become as hard as the initial problem.
+For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper-bound on the variance term of~\eqref{eq:primal_pivot}.
+Confronted to this difficulty, let us move on to an alternative definition of variance.
 

### 2021-10-15 aistats.tex

--- 2021-10-15 aistats.tex
@@
 In this section, we considered direct expansions of~\eqref{eq:bregmanMAP}.
-None of them could really solve~\eqref{problem}.
-Is it possible to solve this problem through an optimization approach?
+None of them could fully solve~\eqref{problem}.
+Next, we investigate whether an optimization approach could solve it.
 

### 2021-10-15 aistats.tex
Cleanup end of optimization

--- 2021-10-15 aistats.tex
@@
 
-To analyze stochastic algorithms, one also need to quantify the randomness.
+To analyze stochastic algorithms, one also need to quantify the randomness of stochastic gradients $g(\nat)$ .
 While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, but they have so far been lacking concrete examples.
@@
 This holds, for instance, for discrete distributions because the probability mass is upper bounded by $1$; however, it rules out  many families.
-In the case of normal distributions$\cN(m, \sigma^2)$, setting $\mu=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$.
+In the case of normal distributions $\cN(m, \sigma^2)$, setting $m=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$.
 We have a similar behavior for gamma distribution with $\alpha = \beta x$ and $\beta \rightarrow +\infty$, or with the beta distribution with $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
@@
 
-If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let us assume they hold on average.
-Does the proof remain valid?
-Yes, I think so \TODO{check this}.
-Does it become impossible to prove / as hard as the initial problem?
-Yes.
-\TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
+%\rlp{Add discussion here.}
 

### 2021-10-15 aistats.tex
Discuss end of optim

@@
 
-Overall, this approach seems to answer~\eqref{problem}, but fails to account for the step-size $\lr_n = \inv{n_0+n}$, as \citet[Thm.1]{dorazio2021stochastic} only proves linear convergence to a variance ball of size $\frac{C}{\stgcvx}$ under constant step size.%  $\lr \leq \inv{\smooth}$.
-
-%\rlp{Add discussion here.}
+Overall, \eqref{eq:dorazio} can partially handle barrier objectives, but it fails to account for the step-size $\lr_n = \inv{n_0+n}$, as \citet[Thm.1]{dorazio2021stochastic} only proves linear convergence to a variance ball of size $\frac{C}{\stgcvx}$ under constant step-size.
+This is in contrast with~\citet{dragomir2021fast} which can handle decreasing step-sizes but not barrier objectives.
+Proving convergence of stochastic mirror descent on barrier loss remains an open problem.
 

### 2021-10-15 aistats.tex
Fix numbers in self-concordance
Pass on conclusion.

--- 2021-10-15 aistats.tex
@@
 
-To analyze stochastic algorithms, one also need to quantify the randomness of stochastic gradients $g(\nat)$ .
+To analyze stochastic algorithms, one also need to quantify the randomness.
 While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, but they have so far been lacking concrete examples.
@@
 This holds, for instance, for discrete distributions because the probability mass is upper bounded by $1$; however, it rules out  many families.
-In the case of normal distributions $\cN(m, \sigma^2)$, setting $m=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$.
+In the case of normal distributions$\cN(m, \sigma^2)$, setting $\mu=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$.
 We have a similar behavior for gamma distribution with $\alpha = \beta x$ and $\beta \rightarrow +\infty$, or with the beta distribution with $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
@@
 
-%\rlp{Add discussion here.}
+If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let us assume they hold on average.
+Does the proof remain valid?
+Yes, I think so \TODO{check this}.
+Does it become impossible to prove / as hard as the initial problem?
+Yes.
+\TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
 
 \section{Conclusion}
-Exponential family MLE and MAP are used everywhere -- every time someone computes a mean and variance to fit some bell-shaped data.
-MLE and MAP are both maximizing the (regularized) likelihood of the data, whose suboptimality is the Kullback-Leibler (KL) divergence.
-The KL is a very natural measure of discrepancy between a model and the true distribution.
-However, we do not have convergence rates on the expected KL, even in the well-specified setting.
-Such a rate could inform us on how many samples we may need or on the influence of the prior.
-There are many results close to such a rate, particularly large sample results \citep{kakade2010learning, ostrovskii2021finite}, which describe how many samples are needed to ensure a locally quadratic regime.
-Another approach to obtain rates is stochastic optimization \citep{bach2013nonstronglyconvex}.
-Here we observed that MAP perfectly fits the framework of stochastic mirror descent with relative smoothness assumptions.
+Despite the MLE and MAP estimators in the exponential family being classical and known in statistics for decades, we highlighted in this paper open problems to bound their frequentist risk (the expected KL) in a non-asymptotic way.
+We reviewed some partial results, such as a large sample analysis that describes how many samples are needed to ensure a locally quadratic regime~\citep{kakade2010learning, ostrovskii2021finite} for which rates are known.
+We also related this problem to the one of obtaining convergence rates in stochastic optimization, observing that MAP fits the framework of stochastic mirror descent with relative smoothness assumptions.
 Nevertheless, none of the current analyses of SMD hold for the MAP, even on a simple family such as $\cN(0,\sigma^2)$, thus revealing an area for progress in non-Euclidean optimization.

### 2021-10-15 aistats.tex

@@
 For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
-For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition \TODO{FIX THIS!} $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
+For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 25d + 5 \norm{\m^* -  \m_0} - n_0$.
 This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
@@
 
-Overall, this approach seems to answer~\eqref{problem}, but fails to account for the step-size $\lr_n = \inv{n_0+n}$, as \citet[Thm.1]{dorazio2021stochastic} only proves linear convergence to a variance ball of size $\frac{C}{\stgcvx}$ under constant step size.%  $\lr \leq \inv{\smooth}$.
-
-If assumptions~\eqref{eq:hanzely} and~\eqref{eq:dragomir} do not hold for every iterate, then let us assume they hold on average.
-Does the proof remain valid?
-Yes, I think so \TODO{check this}.
-Does it become impossible to prove / as hard as the initial problem?
-Yes.
-\TODO{Does it match exactly the variance from the bias-(mixed)-variance decomposition ?}
+Overall, \eqref{eq:dorazio} can partially handle barrier objectives, but it fails to account for the step-size $\lr_n = \inv{n_0+n}$, as \citet[Thm.1]{dorazio2021stochastic} only proves linear convergence to a variance ball of size $\frac{C}{\stgcvx}$ under constant step-size.
+This is in contrast with~\citet{dragomir2021fast} which can handle decreasing step-sizes but not barrier objectives.
+Proving convergence of stochastic mirror descent on barrier loss remains an open problem.
 

### 2021-10-15 aistats.tex
pass on optimization section

--- 2021-10-15 aistats.tex
@@
 As we saw in \S\ref{sec:problem}, MAP can be interpreted as stochastic mirror descent (SMD).
-This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
+This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform other designs and analyses of SMD.
 We first review the assumptions of relative smoothness, helpful to deal with non-smooth functions, before investigating recent analyses of SMD with the MAP.
@@
 
-To analyze stochastic algorithms, one also need to quantify the randomness of stochastic gradients $g(\nat)$ .
+To analyze stochastic algorithms, one also needs to quantify the randomness of stochastic gradients $g(\nat)$.
 While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, but they have so far been lacking concrete examples.
@@
 \subsubsection{Analogs of the Variance}
-let us introduce $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$ th symmetrized Bregman induced by $\conj$.
+Let us introduce the symmetrized Bregman induced by $\conj$, written $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$.
 \citet{hanzely2018fastest} assume that the expectation of $\cS$ between stochastic and deterministic updates verifies
@@
 However, both of their proofs hold if we add an expectation over $\MAPm$ to their assumption.
-But this is not helpful for us, as verifying the assumption become as hard as the initial problem.
-For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper-bound on the variance term of~\eqref{eq:primal_pivot}.
-Confronted to this difficulty, let us move on to an alternative definition of variance.
+However, this is not helpful, as verifying the assumption becomes as hard as the initial problem.
+For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper bound on the variance term of~\eqref{eq:primal_pivot}.
+Confronted with this difficulty, we investigate an alternative definition of variance.
 

### 2021-10-15 aistats.tex
minor change
Add link to appendix in section optim

@@
 
-However, both of their proofs hold if we add an expectation over $\MAPm$ to their assumption.
+Both of their proofs hold if we add an expectation over $\MAPm$ to their assumption.
 However, this is not helpful, as verifying the assumption becomes as hard as the initial problem.
-For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper bound on the variance term of~\eqref{eq:primal_pivot}.
+For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper bound on the variance term of~\eqref{eq:primal_pivot} (cf App.~\ref{app:bias-variance}).
 Confronted with this difficulty, we investigate an alternative definition of variance.

### 2021-10-15 aistats.tex
clarify categorical issue with Proposition 1 in 5.3.
Now make a submission!

--- 2021-10-15 aistats.tex
@@
 \label{ssec:local-quadratic}
-From Taylor expansion~\eqref{eq:bregmanTaylor},
+From the Taylor expansion~\eqref{eq:bregmanTaylor},
 we know that all Bregman divergences are locally quadratic.
@@
 \begin{proposition}
+	\label{prop:selfConcordant}
 Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
@@
 
-These results partially answer~\eqref{problem}.
-Partially because the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$.
+These results answer~\eqref{problem} only \emph{partially} for the following reasons.
+First, the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$. 
 None of them apply to small $n$.
-That is one caveat. Another one is the Gaussian example we saw in \S\ref{ssec:gaussian}, which is not self-concordant.
-A full solution to~\eqref{problem} would address both of these caveats.
-Informed by the properties we have seen so far, let us investigate a general decomposition of the Bregman that could guide us towards a solution.
+Second, they do not apply to some standard exponential families like the full Gaussian example that we saw in \S\ref{ssec:gaussian} which is \emph{not} self-concordant.
+Third, Proposition~\ref{prop:selfConcordant} does not imply directly a bound on the \emph{expected} Bregman.\footnote{One way to see this is that while the entropy of the categorical distribution is self-concordant, and thus Proposition~\ref{prop:selfConcordant} applies to it, the expected KL of its MLE is still infinite as previously mentioned in \S\ref{sec:problem}.}
+A full solution to~\eqref{problem} would address all of these caveats.
+Informed by the properties that we have seen so far, we next investigate a general decomposition of the Bregman that could guide us towards a solution.
 %If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.
@@
 			%\\
-			Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
+			Variance on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
 			& \redxmark & \greencmark & \greencmark  & \redxmark
 			\\
-			Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
+			Variance at $\theta^*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
 			& \redxmark & \greencmark & \redxmark  & \greencmark
 			\\
-			Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
+			Optimization gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
 			& \greencmark & \redxmark & \redxmark & \greencmark

### 2021-10-15 aistats.tex
Rephrase locally quadratic to better situate the work of ostrovskii

@@
 The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g., categorical and Dirichlet distributions.
-
-There are quite a few pieces of work in this direction.
-\citet{ostrovskii2021finite} characterizes the number of samples needed to reach the asymptotic quadratic behavior for any parametric models with a self-concordant log-likelihood.
+Precisely, categorical variables illustrate that Proposition~\ref{prop:selfConcordant} does not imply directly a bound on the \emph{expected} Bregman.
+The expected KL of the categorical MLE is always infinite, as previously mentioned in \S\ref{sec:problem}.
+
+Overcoming this limitation,
+\citet{ostrovskii2021finite} characterize the number of samples needed to be upper bounded by a quadratic \emph{with high-probability}, for any parametric models with a self-concordant log-likelihood.
 \citet{anastasiou2017bounds} obtains a similar flavor of result under other assumptions on the third derivative of $f$.
-More closely, in the world of exponential families, \citet{kakade2010learning} assumes a local bound on all higher-order moments of $\logpart$ in $\nat^*$, to obtain a similar result.
+More closely, in the world of exponential families, \citet{kakade2010learning} prove a result similar to Proposition~\ref{prop:selfConcordant} from a local bound on all higher-order moments of $\logpart$ in $\nat^*$.
 However, these results are expressed with quadratics in $\nat$, not $\m$, and they do not directly translate to convergence rates for the MAP, but they might with some more work.
 
-These results answer~\eqref{problem} only \emph{partially} for the following reasons.
+More generally, these results answer~\eqref{problem} only \emph{partially} for the following reasons.
 First, the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$. 
@@
 Second, they do not apply to some standard exponential families like the full Gaussian example that we saw in \S\ref{ssec:gaussian} which is \emph{not} self-concordant. 
-Third, Proposition~\ref{prop:selfConcordant} does not imply directly a bound on the \emph{expected} Bregman.\footnote{One way to see this is that while the entropy of the categorical distribution is self-concordant, and thus Proposition~\ref{prop:selfConcordant} applies to it, the expected KL of its MLE is still infinite as previously mentioned in \S\ref{sec:problem}.}
-A full solution to~\eqref{problem} would address all of these caveats.
+A full solution to~\eqref{problem} would address these caveats.
 Informed by the properties that we have seen so far, we next investigate a general decomposition of the Bregman that could guide us towards a solution.
-%If no current approach solves~\eqref{problem}, then let us formulate a conjecture informed by the examples and partial solutions we have s.
 

### 2021-10-15 aistats.tex
Switch to real paragraphs in the intro to exploit the space

@@
 
-{\bf Exponential families} are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
+\paragraph{Models}
+Exponential families are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
 Many standard random variables are exponential families: Gaussians, categorical, gamma, or Dirichlet, for example.
@@
 
-{\bf Estimators.}
+\paragraph{Estimators}
 In this paper, we consider the problem of estimating $\nat$ from a dataset $\mD = (X_1, \dots, X_n)$ of iid observations from $p_\theta$ in an exponential family.
@@
 
-{\bf Statistical decision theory.}
+\paragraph{Statistical decision theory}
 To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
@@
 
-{\bf Stochastic optimization} offers an interesting perspective on this question.
+\paragraph{Optimization} 
+Stochastic optimization offers an interesting perspective on~\eqref{problem}.
 Consider the problem
@@
 
-{\bf Expected Outcomes.}
+\paragraph{Expected Outcomes}
 A solution to~\eqref{problem} can clarify the importance of the prior in MAP, in particular in the few sample regime. %, for instance for Gaussians $\cN(m,\sigma^2)$.
@@
 
-{\bf Contributions.}
+\paragraph{Contributions}
 After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
@@
 
-{\bf Notation.}
+\paragraph{Notation}
 $X$ and $T=T(X)$ are random variables, $x$ is a sample, $n$ is the number of samples and $d= \dim(T)$.

### 2021-10-15 aistats.tex
Minor clarity tweaks

@@
 We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
-\begin{align}
+\aligns{
 	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
 	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
-\end{align}
+}
 which is the common notion of \textit{entropy} in information theory.
@@
     p(\nat)
-    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \\
+    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \nonumber \\
     &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
@@
 Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
-\begin{align}
+\aligns{
 	p(\nat \cond \mD_n)
     \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
-    \label{eq:joint_likelihood}
-\end{align}
+%    \label{eq:joint_likelihood}
+}
 which reaches its maximum at $\MAPt^\text{MAP}$ such that
@@
 Instead of taking the expectation, one might want to bound the risk in high probability
-without resorting to Markov inequality, but this is a difficult endeavor.
+without resorting to Markov inequality, as achieved by~\citep{ostrovskii2021finite},
+ but this is a difficult endeavor.
 These examples make a case for regularized estimators such as MAP,
@@
 We illustrate its numerical behavior in Figure~\ref{fig:curves}.
-With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.
+With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O\paren{\frac{d^2}{n} + \frac{d^3}{n(n-d-1)}}$.
 
@@
 \end{theorem}
+Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights an explicit $O(\frac{v}{n} + \frac{b}{n^2})$ variance-bias decomposition.
 This inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
-This explains the loss of a factor 2 compared to the asymptote~\eqref{eq:asymptote}.
-Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights an explicit variance-bias decomposition.
-In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
+This explains why the variance term is twice larger than the asymptote~\eqref{eq:asymptote}.
+Regarding the bias, it vanishes when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
 This correlates well with our numerical observations (cf App.~\ref{app:gaussian-variance}).
@@
 	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:primal_pivot} from~\S\ref{ssec:bias-variance}.
-	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the Thales triangle and the red curve.
+	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
 	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution.

### 2021-10-15 aistats.tex
Fix self-concordance

@@
 
-These functions are neither smooth, nor strongly convex, nor self-concordant (see App.~\ref{app:gaussian}).
+These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
+\citep[p.177]{nesterov2003introductory}
+, but $\logpart$ is not (see App.~\ref{app:gaussian}).
 We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.
@@
 \label{ssec:asymptote}
-As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-We now detail a classical asymptotic result.
-Proofs are in App.~\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a great review.
+As a reference point for any finite convergence rate, it is interesting to briefly review the classical asymptotic behavior of these quantities as $n \rightarrow +\infty$.
+Proofs are in App.~\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a more comprehensive review.
 
@@
 It is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
-Besides the Gaussian with known variance, the problem is that no standard exponential family has a \textit{global} strongly convex log-partition function.
+Besides the Gaussian with known variance, the problem is that no standard exponential family has a \textit{globally} strongly convex log-partition function.
 The next section focuses on \textit{local} quadratic behavior, which is more realistic.
-% So let us focus on a more realistic set of assumptions in the next section.
 
@@
 The entropy $\conj$ is self-concordant for several common families.
-For instance when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.~\ref{app:self-concordant}),
+For instance Gaussians (\S\ref{ssec:gaussian}), or when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.~\ref{app:self-concordant}),
 as is the case with:
 exponential distributions,
-Gaussians with known mean,
 Laplace with known mean,
@@
 Overcoming this limitation,
-\citet{ostrovskii2021finite} characterize the number of samples needed to be upper bounded by a quadratic \emph{with high-probability}, for any parametric models with a self-concordant log-likelihood.
+\citet{ostrovskii2021finite} characterize the number of samples needed to be upper bounded by a quadratic \emph{with high-probability}, for any parametric models with a self-concordant log-likelihood $f$.
 \citet{anastasiou2017bounds} obtains a similar flavor of result under other assumptions on the third derivative of $f$.
@@
 None of them apply to small $n$.
-Second, they do not apply to some standard exponential families like the full Gaussian example that we saw in \S\ref{ssec:gaussian} which is \emph{not} self-concordant. 
+Second, they do not apply to some standard exponential families like the full Gaussian example that we saw in \S\ref{ssec:gaussian}  for which $f$ is \emph{not} self-concordant. 
 A full solution to~\eqref{problem} would address these caveats.
@@
 for all possible iterates $\MAPt$, relevant step-sizes $\lr$ and for some constant $C$.
-When $A(\theta) = \frac{1}{2}\norm{\theta}^2$ or when $A$ is $\stgcvx$-strongly convex,
+When $A(\theta) = \frac{1}{2}\norm{\theta}^2$,
+%or when $A$ is $\stgcvx$-strongly convex,
 this definition recovers the variance of the stochastic gradient
 \[
-\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla f(\theta) - g(\theta)}^2}\leq C.
+\expect[g\!\!]{\norm{\nabla f(\theta) - g(\theta)}^2}\leq C.
 \]
@@
 }
-When $\logpart$ is $\stgcvx$-strongly-convex, we recover the variance of the gradients at the optimum
+When $A(\theta) = \frac{1}{2}\norm{\theta}^2$,
+% or when $\logpart$ is $\stgcvx$-strongly-convex,
+we recover the variance of the gradients at the optimum
 \[
-\expect[g\!\!]{\inv{\stgcvx}\norm{\nabla g(\theta^*)}^2}\leq C.
+\expect[g\!\!]{\norm{\nabla g(\theta^*)}^2}\leq C.
 \]

### 2021-10-16 main.tex
Compress gaussian variance and add gaussian
Add bound for Gamma(rate) distribution

@@
 
+\section{Gamma with known shape}
+
+We are looking for $\Gamma(\alpha, \beta^*)$, where the shape $\alpha$ is known, but the rate $\beta^*$ is unknown.
+The logpartition is $\logpart(\beta) = - \alpha \log(\beta)$ and the mean parameter is $\mu* = \frac{\alpha}{\beta}$.
+The symmetrized Bregman of the MAP verifies the following upper bound
+\alignn{
+    \forall n\geq \inv{\alpha},\quad
+	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
+	\leq\frac{1}{n_0+n} + 
+	\frac{\alpha\paren{\frac{\mu_0}{\mu^*} - 1 - \inv{\alpha n_0}}^2}{\paren{1+\frac{n}{n_0}}\paren{\frac{\mu_0}{\mu^*} + \frac{n - \inv{\alpha}}{n_0}}}
+}
+where the bias vanish when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.
+
 \section{Euclidean Behaviour}
@@
 	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
-	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-\natp_2) \; .
+	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-2\natp_2) \; .
 \end{align}
@@
 	\expect{\bregmanconj(\mu^*; \MAPm)} 
+	& = \half \left(\frac{n_0 - a + 2}{a +n - 2}  + \frac{a - n_0}{n_0+n} \right )\\
 	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
-	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \frac{n_0 - a + 2}{(a+n-2)(n_0+n)} \\
-	&= \frac{1}{a+n-2} + \frac{(n_0 - a)^2 + 2(n_0-a) \pm 1}{2(a+n-2)(n_0+n)} \\
+	&\leq \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
 	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
@@
 	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
-	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 - a + 1}{a} \\
+	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 +1 - a }{a} \\
 	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
 	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
-	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 \frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
+	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{\frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
 \end{align}

### 2021-10-18 aistats.tex
Add comment
Write section titles full cap

@@
 
-\let\oldsection\section
-\renewcommand{\section}[1]{\oldsection{\texorpdfstring{\uppercase{#1}}{#1}}}
+%\let\oldsection\section
+%\renewcommand{\section}[1]{\oldsection{\texorpdfstring{\uppercase{#1}}{#1}}}
 
@@
 
-\section{Introduction}
+\section{INTRODUCTION}
 \label{sec:motivation}
@@
 
-\section{Technical background}
+\section{TECHNICAL BACKGROUND}
 \label{sec:background}
@@
 
-\section{Illustrating Examples}\label{sec:example}
+\section{ILLUSTRATING EXAMPLES}\label{sec:example}
 \subsection{Gaussian with Unknown Variance}\label{ssec:gaussian-variance}
@@
 
-\section{Partial Solutions}
+\section{PARTIAL SOLUTIONS}
 \label{sec:insights}
@@
 \label{sec:optimization}
-
-%This section show how MAP can be interpreted as stochastic mirror descent (SMD).
-Thanks to this strong link, we show that \textbf{1)} we may obtain a convergence rate for MAP from the rate od SMD, and \textbf{2)} any insights gained from the MAP may inform on further design and analysis of SMD.
-%Then, we review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three different analyses of SMD with the MAP.
-%\rlp{mention that examples have so far been lacking for relative SMD.}
-%
-%\subsection{MAP as Stochastic Mirror Descent}
-%\label{ssec:MAP=SMD}
-%By the Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
-%Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
-%\begin{align*}
-%	\hat \nat_{n}
-%	= \argmin_\nat f_n(\nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
-%    %\label{eq:SBPP}
-%\end{align*}
-%which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$.
-%%Maybe skip this SBPP story altogether
-%If we expand $f_n(\nat) = \logpart(\nat) - T_n$,  we get (after some manipulation with the Bregman divergence)
-%\begin{align*}
-%		\hat \nat_{n}
-%	= \argmin_\nat \langle g_{n}(\hat \nat_{n-1}), \nat \rangle + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
-%    %\label{eq:SMD}
-%\end{align*}
-%where $g_{n}(\nat) := \nabla f_n(\nat) = \hat \m_{n-1} - T_n$.
-%We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$.
-%From the first order optimality condition, we see this is a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
-%This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
-%In particular, it could potentially apply to a wide range of distribution $\cD$ on $X$.
-%Note that this optimization perspective does not hold for the MLE %($n_0=0$)
-%as its implicit uniform prior may correspond to an initialization outside of $\Theta$.
 
@@
 This holds, for instance, for discrete distributions because the probability mass is upper bounded by $1$; however, it rules out many families.
-In the case of normal distributions $\cN(m, \sigma^2)$, setting $m=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$.
+In the case of normal distributions $\cN(m, \sigma^2)$, setting $m=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$,.
+%A similar behavior holds for Gamma, Beta, inverse Gaussians, log-normal, Gamma or inverse Gamma distributions.
 We have a similar behavior for gamma distribution with $\alpha = \beta x$ and $\beta \rightarrow +\infty$, or with the beta distribution with $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
@@
 
-\section{Conclusion}
+\section{CONCLUSION}
 Despite the MLE and MAP estimators in the exponential family being classical and known in statistics for decades, we highlighted in this paper open problems to bound their frequentist risk (the expected KL) in a non-asymptotic way.

### 2021-10-18 main.tex
Replace \natp by \nat in main.tex

@@
 \newcommand{\bregmanconj}{\cB_{\logpart^*}}
-\newcommand{\natp}{\theta}
+\newcommand{\nat}{\theta}
 \newcommand{\meanp}{\mu}
@@
 \newcommand{\MAPm}{\hat \mu_n}
-\newcommand{\MAPt}{\hat \natp_n}
+\newcommand{\MAPt}{\hat \nat_n}
 
@@
 \paragraph{Exponential Family.}
-The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model 
+The exponential family member with sufficient statistic $T$ and natural parameter $\nat$ is the model 
 \begin{equation}
-	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
+	 p(X|\nat) = \exp( \nat^\top T(X) - \logpart(\nat)) \; ,
 \end{equation}
@@
 \begin{align}
-    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
+    \logpart(\nat) = \log \int e^{\nat^\top T(x)} dx \; .
 \end{align}
@@
 \end{align}
-Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\natp = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
+Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\mu=\E[T(X)] = \sigma^2 >0$. 
 Mean and natural parameters are roughly inverse of each other
 \begin{align}
-	\natp = -\inv{2 \mu}\; .
+	\nat = -\inv{2 \mu}\; .
 \end{align}
@@
 	\log p(x) = - \frac{x^2}{2 \sigma^2} - \half \log(2 \pi \sigma^2 ) 
-	= x^2 \natp - \logpart(\natp) \\
-	\implies \logpart (\natp) = - \half \log(-\natp)  + \half \log(\pi) 
+	= x^2 \nat - \logpart(\nat) \\
+	\implies \logpart (\nat) = - \half \log(-\nat)  + \half \log(\pi) 
 \end{align}
@@
 \begin{align}
-    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
-    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
+    \nabla\logpart(\nat) &=  \expect[p(X|\nat)]{T(X)} =: \meanp \\
+    \nabla^2 \logpart(\nat) &= \Cov_\nat[T(X)] > 0
 \end{align}
 where $\meanp$ is called the mean parameter.
-If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\natp$ and mean parameters $\mu$.
+If the sufficient statistic $T$ is minimal, then the log-partition function $\logpart$ is strictly convex and its gradient $\nabla \logpart$ is a bijection between natural parameters $\nat$ and mean parameters $\mu$.
 The second identity entails that $\logpart$ is strictly-convex. 
@@
 \begin{align}
-	\conj(\mu) = \langle \mu, \natp \rangle - \logpart(\natp) \; .
+	\conj(\mu) = \langle \mu, \nat \rangle - \logpart(\nat) \; .
 \end{align}
 It turns out that $\conj$ matches the common notion of \textit{entropy} in information theory, so we will call it entropy.
-If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\natp) = \natp$ (cf Fig.~\ref{fig:duality}).
+If $\logpart$ is strictly convex, then its gradient is strictly monotone, so it is a bijection, and its inverse is the gradient of its dual $\nabla\conj \circ \nabla\logpart(\nat) = \nat$ (cf Fig.~\ref{fig:duality}).
 For a full review of exponential families and their duality, see \citet[Chapter 3]{wainwright2008graphical}.
@@
 	\includegraphics[width=.7\textwidth]{figs/2d/mirrormap.pdf}
-	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\natp, \meanp$.Top figure reproduced from \citet{kunstner2020homeomorphic}.
+	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between the natural and mean parameters $\nat, \meanp$.Top figure reproduced from \citet{kunstner2020homeomorphic}.
 Bottom figure represents $\cN(\mu, \sigma^2)$.}
@@
 \begin{example}
-For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \natp - \logpart(\natp)$ to get the entropy
+For $\cN(0,\sigma^2)$,we can use the formula $\conj(\mu) = \mu \nat - \logpart(\nat)$ to get the entropy
 \begin{align}
@@
 \end{align}
-We can also take gradient and derivative of $\logpart(\natp)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
-\begin{align}
-	\nabla\logpart(\natp) &= \frac{-1}{2 \natp} = \sigma^2 = \mu = \E[X^2] \\
-	\nabla^2\logpart(\natp) &= \frac{1}{2 \natp^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
+We can also take gradient and derivative of $\logpart(\nat)$ to retrieve the mean and covariance of the sufficient statistic $X^2$
+\begin{align}
+	\nabla\logpart(\nat) &= \frac{-1}{2 \nat} = \sigma^2 = \mu = \E[X^2] \\
+	\nabla^2\logpart(\nat) &= \frac{1}{2 \nat^2} = 2 \sigma^4 = 2 \mu^2 = \Var(X^2) 
 \end{align}
@@
 \paragraph{Conjugate Prior}
-One conjugate prior for $p(X|\natp)$ is
-\begin{align}
-    p(\natp) &\propto \exp( - n_0 \bregman(\natp ; \natp_0) )
-\end{align}
-where $n_0$ and $\natp_0$ are (hyper)parameters of the prior, and $\bregman(\natp ; \natp_0)$ is the Bregman divergence induced by $\logpart$ between $\natp$ and $\natp_0$
-\begin{align}
-    \bregman (\natp ; \natp_0)
-    & = \logpart(\natp) - \logpart(\natp_0) 
-    - \langle \nabla \logpart(\natp_0)  , \natp - \natp_0 \rangle
-\end{align}
-with $\nabla \logpart(\natp_0) = \expect[\natp_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\natp_0$. 
-Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\natp_0$.
+One conjugate prior for $p(X|\nat)$ is
+\begin{align}
+    p(\nat) &\propto \exp( - n_0 \bregman(\nat ; \nat_0) )
+\end{align}
+where $n_0$ and $\nat_0$ are (hyper)parameters of the prior, and $\bregman(\nat ; \nat_0)$ is the Bregman divergence induced by $\logpart$ between $\nat$ and $\nat_0$
+\begin{align}
+    \bregman (\nat ; \nat_0)
+    & = \logpart(\nat) - \logpart(\nat_0) 
+    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle
+\end{align}
+with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$. 
+Intuitively, $n_0$ is a number of fictive points observed from a distribution with parameter $\nat_0$.
 We can re-write this prior as 
 \begin{align}
-    p(\natp) \propto 
-    \exp( -n_0 \logpart (\natp) 
-    + \langle n_0 \mu_0, \natp \rangle ) \; ,
-\end{align}
-which is the formula for the exponential family with sufficient statistics $(\natp ,\logpart(\natp))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
+    p(\nat) \propto 
+    \exp( -n_0 \logpart (\nat) 
+    + \langle n_0 \mu_0, \nat \rangle ) \; ,
+\end{align}
+which is the formula for the exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and with natural parameter $(n_0 \mu_0, -n_0)$.
 The posterior given $\cD=(X_1,\dots, X_n)$ is then part of the same family, with natural parameters $(n_0 \mu_0 + \sum_i T(X_i) , -(n_0 + n))$.
@@
 \begin{example}
-In the case of $\cN(0,\sigma^2)$, the logpartition function is $\logpart(\natp) = -\log(-\natp) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\natp, \log(-\natp) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
+In the case of $\cN(0,\sigma^2)$, the logpartition function is $\logpart(\nat) = -\log(-\nat) /2 + \cst$, thus the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, eg a negative \href{https://en.wikipedia.org/wiki/Gamma_distribution}{Gamma distribution}.
 In particular,
 \begin{align}
-	p(\natp) 
+	p(\nat) 
 	&\propto 
-    \exp( -n_0 \logpart (\natp) 
-    + \langle n_0 \mu_0, \natp \rangle ) \\
-    &\propto \exp( \half[n_0] \log(-\natp) + n_0\mu_0 \natp ) \\
-	&\propto (-\natp)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\natp)} / Z
-\end{align}
-from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\natp \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$.
+    \exp( -n_0 \logpart (\nat) 
+    + \langle n_0 \mu_0, \nat \rangle ) \\
+    &\propto \exp( \half[n_0] \log(-\nat) + n_0\mu_0 \nat ) \\
+	&\propto (-\nat)^{1 + \half[n_0] -1 } e^{-n_0 \mu_0 (-\nat)} / Z
+\end{align}
+from which we infer the shape parameter $\alpha=1 + \half[n_0]$ and the rate parameter $\beta = n_0 \mu_0$, eg $\nat \sim \Gamma (1+\half[n_0] , n_0 \mu_0)$.
 After seeing $n$ samples, the posterior is $\Gamma\left(1+\half[n_0+n] , n_0 \mu_0 + \sum_i T(X_i) \right)$.
@@
 \begin{align*}
-    -\log p(\natp) = n_0 (\logpart(\natp)  - \natp^\top \meanp_0 ) + \cst
+    -\log p(\nat) = n_0 (\logpart(\nat)  - \nat^\top \meanp_0 ) + \cst
 \end{align*}
-Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\natp)$ is
-\begin{align}
-    -\log p(\cD|\natp)p(\natp) 
-    = (n_0+n) \logpart (\natp) 
+Thus the joint log-likelihood of $\cD =(X_1,\dots,X_n,\nat)$ is
+\begin{align}
+    -\log p(\cD|\nat)p(\nat) 
+    = (n_0+n) \logpart (\nat) 
     - \theta^\top \left(n_0 \meanp_0 + \sum_{i=1}^n T(X_i) \right) + \cst \; .
 \end{align}
-Minimizing this expression over $\natp$ yields the Maximum A Posteriori estimate
-\begin{align}
-    \hat \natp = \argmin_\natp -\log p(\cD|\natp) + n_0 \bregman(\natp ; \natp_0)
+Minimizing this expression over $\nat$ yields the Maximum A Posteriori estimate
+\begin{align}
+    \hat \nat = \argmin_\nat -\log p(\cD|\nat) + n_0 \bregman(\nat ; \nat_0)
 \end{align}
@@
 \begin{align}
-    \nabla \logpart(\hat \natp_\text{MAP}) = \hat \meanp_\text{MAP}
+    \nabla \logpart(\hat \nat_\text{MAP}) = \hat \meanp_\text{MAP}
     = \frac{n_0 \meanp_0 + \sum_{i=1}^n T(X_i) }{n_0+n} \; .
@@
 \end{align}
-The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\natp^*$ or $\meanp^*$.
+The MLE and MAP estimates are statistics of the dataset $\cD$. Given a random dataset, we wish to bound their deviation from the optimum $\nat^*$ or $\meanp^*$.
 
@@
 \begin{align}
-    \expect[X\sim p(.|\natp^*)]{-\log p(X|\natp) + \log p(X|\natp^*) }
-	= \KL( p(.|\natp^*) ; p(.|\natp)) \; .
+    \expect[X\sim p(.|\nat^*)]{-\log p(X|\nat) + \log p(X|\nat^*) }
+	= \KL( p(.|\nat^*) ; p(.|\nat)) \; .
 \end{align}
@@
 \begin{align}
-	\KL( p(.|\natp^*) ; p(.|\natp)) 
-	= \bregman (\natp ; \natp^*)  \; .
+	\KL( p(.|\nat^*) ; p(.|\nat)) 
+	= \bregman (\nat ; \nat^*)  \; .
 \end{align}
@@
 \begin{align}
-	\bregman (\natp ; \natp^*)
-    = \logpart(\natp) - \langle \natp , \mu^* \rangle + \conj(\mu^*)
+	\bregman (\nat ; \nat^*)
+    = \logpart(\nat) - \langle \nat , \mu^* \rangle + \conj(\mu^*)
     = \bregmanconj ( \meanp^* ; \meanp)
@@
 \boxed{
-	\KL( p(.|\natp^*) ; p(.|\natp))
-    = \bregman (\natp ; \natp^*)
+	\KL( p(.|\nat^*) ; p(.|\nat))
+    = \bregman (\nat ; \nat^*)
     = \bregmanconj ( \meanp^* ; \meanp) \; .
@@
 \begin{important}
-The question is: how does this quantity behave when $\natp$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
+The question is: how does this quantity behave when $\nat$ is the maximum-likelihood or the MAP estimate ? Can we get bounds on the following quantities
 \begin{align}
 	\label{eq:bregmanMLE}
-	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
+	\expect[X_i\sim \nat^*]{\bregmanconj \left (\E [T(X)] ;  \inv{n}  \smallsum_i T(X_i) \right )} \leq \ ? \; , \\
 	\label{eq:bregmanMAP}
-	\expect[X_i\sim \natp^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
+	\expect[X_i\sim \nat^*]{\bregmanconj \left (\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(X_i)}{n_0+n} \right )} \leq \ ? \; ,
 \end{align}
@@
 	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
-	\bregman( \natp_n; \natp_* ) 
-	&=  \half \left ( \frac{ \natp_n}{\natp_*} - 1 - \log  \frac{ \natp_n}{\natp_*} \right) \; .
+	\bregman( \nat_n; \nat_* ) 
+	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
 \end{align}
 \end{example}
-In other words, this divergence measures the discrepancy between the ratio $\frac{ \natp_n}{\natp_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
+In other words, this divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\mu_*}{ \mu_n}  $ and $1$ via the function $\phi$
 \begin{align}
 	\phi(z) := \half (z - 1 - \log(z)) \\
-	\bregman( \natp_n; \natp_* )   = \phi(\frac{ \natp_n}{\natp_*}) =  \phi(\frac{\mu_*}{ \mu_n})
+	\bregman( \nat_n; \nat_* )   = \phi(\frac{ \nat_n}{\nat_*}) =  \phi(\frac{\mu_*}{ \mu_n})
 \end{align}
@@
 		\leq \expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\natp^*}} 
+		= \expect{\frac{\MAPt}{\nat^*}} 
 		\leq \frac{n_0 +n}{a+ (n-2)_+}
@@
     \bregmanconj(\mu^* ; \mu) 
-    \leq L \norm{\mu^* - \mu} + \norm{\natp} \norm{\mu^* - \mu}
+    \leq L \norm{\mu^* - \mu} + \norm{\nat} \norm{\mu^* - \mu}
     \leq 2L \norm{\mu^* - \mu}
@@
     \nabla^2\conj(\mu^*) 
-    = \nabla^2\logpart(\natp^*)^{-1} 
-    = \Cov_{\natp^*}[T(X)]^{-1} 
+    = \nabla^2\logpart(\nat^*)^{-1} 
+    = \Cov_{\nat^*}[T(X)]^{-1} 
     =: \mSigma^{-1}  \; .
@@
 \subsection{Dual Expectation Pivot.}
-Let $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ be the MAP estimate of the mean (dual) parameter, $\hat \natp_n = \nabla \conj(\hat \mu_n)$ the corresponding natural (primal) parameter, and
+Let $\hat \mu_n = \frac{n_0 \mu_0 + \sum_i T(X_i)}{n_0 + n}$ be the MAP estimate of the mean (dual) parameter, $\hat \nat_n = \nabla \conj(\hat \mu_n)$ the corresponding natural (primal) parameter, and
 $\mu_n = \expect{\hat \mu_n} = \frac{n_0 \mu_0 + n \mu^*}{n_0 + n}$ the expected MAP dual estimate
@@
 	-  \expect{\conj(\hat \mu_n)}
-	+ \E[\langle \hat \natp_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] 
+	+ \E[\langle \hat \nat_n ; \hat \mu_n   \pm \mu_n - \mu^*  \rangle ] 
 	\nonumber \\
@@
 	- \conj(\mu_n)  + \E[\bregmanconj(\mu_n, \MAPm)]
-	+ \langle \E[\hat \natp_n] \pm \nabla \conj(\mu_n) ;  \mu_n - \mu^* \rangle
+	+ \langle \E[\hat \nat_n] \pm \nabla \conj(\mu_n) ;  \mu_n - \mu^* \rangle
 	\nonumber\\
 	& = \bias{\bregmanconj(\mu^*; \mu_n)}
-	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \natp_n] - \natp_n} \rangle 
+	+ \frac{n_0}{n+n_0} \langle  \bias{\mu_0 - \mu_*} ; \variance{\E[\hat \nat_n] - \nat_n} \rangle 
 	+ \variance{\E[\bregmanconj(\mu_n, \MAPm)]} \; .
@@
 	+ \variance{\expect{\bregmanconj(\tilde \mu_n ; \MAPm)}} \\
-	\expect{\bregman(\MAPt; \natp^*)} 
-	&= \bias{\bregman(\E[\MAPt]; \natp^*)}
+	\expect{\bregman(\MAPt; \nat^*)} 
+	&= \bias{\bregman(\E[\MAPt]; \nat^*)}
 	+ \variance{\expect{\bregman(\MAPt; \E[\MAPt])}}  \; .
@@
 	\begin{align}
-	\Cov(\hat \natp_n ; \MAPm )  
+	\Cov(\hat \nat_n ; \MAPm )  
 	&=\expect{\bregmanconj(\MAPm ; \mu_n)
@@
 	 = \expect{\langle \MAPt; \MAPm \rangle} - \langle \expect{\MAPt} ; \MAPm \rangle
-	= \Cov(\hat \natp_n ; \hat \mu_n ) 
+	= \Cov(\hat \nat_n ; \hat \mu_n ) 
 \end{align}
@@
 Note that, as a corollary, the covariance between primal and dual MAP
-$\Cov(\hat \natp_n ; \hat \mu_n )$ is positive, and it can be seen as a non-linear variance of $\hat \mu_n$ -- e.g.
+$\Cov(\hat \nat_n ; \hat \mu_n )$ is positive, and it can be seen as a non-linear variance of $\hat \mu_n$ -- e.g.
 we can define a variance operator with a non-linear monotone operator $\nabla\conj$ plugged in the middle $\Var_{\nabla \conj}(\hat \mu_n) := \Cov(\nabla\conj(\hat \mu_n) ; \hat \mu_n ) \geq 0$.
@@
 \begin{align}
-	\expect{\langle \hat \natp_n ; \hat \mu_n - \mu_n \rangle}
-	&= \expect{\langle \hat \natp_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
-	& = \frac{n}{n+n_0} \expect{\langle \hat \natp_n ; T(X_0) - \mu^*\rangle} \\
-	&= \frac{n}{n+n_0} \Cov(\hat \natp_n ; T(X_0) ) 
+	\expect{\langle \hat \nat_n ; \hat \mu_n - \mu_n \rangle}
+	&= \expect{\langle \hat \nat_n ; \frac{\sum_i T(X_i) - \mu^*}{n+n_0} \rangle}  \\
+	& = \frac{n}{n+n_0} \expect{\langle \hat \nat_n ; T(X_0) - \mu^*\rangle} \\
+	&= \frac{n}{n+n_0} \Cov(\hat \nat_n ; T(X_0) ) 
 	\label{eq:covariance_expansion}
@@
 \begin{align}
-	f^* = \min_\natp \E[f(\natp ; X)] := -\langle \E[T(X)] ; \natp \rangle + \logpart(\natp)
+	f^* = \min_\nat \E[f(\nat ; X)] := -\langle \E[T(X)] ; \nat \rangle + \logpart(\nat)
 \end{align}
 where $\logpart$ is a convex function.
-We assume access to stochastic oracles $X \sim p(X | \natp^*)$ -- e.g. we observe $X$ such that $\E[T(X)] = \mu^*$ and we can compute  stochastic function values and gradients 
-\begin{align}
-	f(\natp ; X) &= -\langle T(X) ; \natp \rangle + \logpart(\natp) \\
-	\nabla f(\natp ; X) &= \nabla \logpart(\natp) - T(X)\; .
+We assume access to stochastic oracles $X \sim p(X | \nat^*)$ -- e.g. we observe $X$ such that $\E[T(X)] = \mu^*$ and we can compute  stochastic function values and gradients 
+\begin{align}
+	f(\nat ; X) &= -\langle T(X) ; \nat \rangle + \logpart(\nat) \\
+	\nabla f(\nat ; X) &= \nabla \logpart(\nat) - T(X)\; .
 \end{align}
@@
 \paragraph{Interpretation.}
-As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms with divergence $\bregman$ and initialization $\natp_0$ : 
+As we will see in the two upcoming sections, MAP can be seen as an application of more generic iterative algorithms with divergence $\bregman$ and initialization $\nat_0$ : 
 \begin{itemize}
@@
 \end{itemize}
-In both cases, $f(\natp)$ is convex and 1-smooth  and 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.
-
-\paragraph{Goal.} We want to bound the suboptimality $f(\natp) - f^*$ as a function of the number of samples.
+In both cases, $f(\nat)$ is convex and 1-smooth  and 1-strongly convex relative to the potential $\logpart$. Be cautious that the learning rate is decreasing.
+
+\paragraph{Goal.} We want to bound the suboptimality $f(\nat) - f^*$ as a function of the number of samples.
 
@@
 \begin{align}
-	p(\natp |\cD_n)
-	=p(\natp|X_n,\cD_{n-1})
-	\propto p(X_n|\natp,\cD_{n-1}) p(\natp|\cD_{n-1})
-	=p(X_n|\natp)p(\natp |\cD_{n-1}) \; .
+	p(\nat |\cD_n)
+	=p(\nat|X_n,\cD_{n-1})
+	\propto p(X_n|\nat,\cD_{n-1}) p(\nat|\cD_{n-1})
+	=p(X_n|\nat)p(\nat |\cD_{n-1}) \; .
 \end{align}
@@
 \begin{align}
-	- \log p(\natp | \cD_n) 
-	&= \sum_{i=1}^n -\log p(X_i | \natp) - \log p(\natp) + \cst\\
-	&= \sum_i - \langle T(X_i) ; \natp \rangle  + n \logpart(\natp) + n_0 \bregman(\natp ; \natp_0) +\cst \\
-	&= (n+n_0)\logpart(\natp)  - \langle n_0 \mu_0 + \sum_i T(X_i) ; \natp \rangle + \cst \\
-	&= (n+n_0) \bregman(\natp; \hat \natp_n^\text{MAP}) + \cst \; ,
+	- \log p(\nat | \cD_n) 
+	&= \sum_{i=1}^n -\log p(X_i | \nat) - \log p(\nat) + \cst\\
+	&= \sum_i - \langle T(X_i) ; \nat \rangle  + n \logpart(\nat) + n_0 \bregman(\nat ; \nat_0) +\cst \\
+	&= (n+n_0)\logpart(\nat)  - \langle n_0 \mu_0 + \sum_i T(X_i) ; \nat \rangle + \cst \\
+	&= (n+n_0) \bregman(\nat; \hat \nat_n^\text{MAP}) + \cst \; ,
 \end{align}
@@
 \begin{align}
-    \hat \natp_n^\text{MAP}
-    &= \argmin_\natp - \log p(\natp | \cD_n) \\
-    &= \argmin_\natp - \log p(X_n|\natp) - \log p(\natp | \cD_{n-1}) \\
-    &= \argmin_\natp f(\natp; X_{n}) + (n+n_0 - 1) \bregman(\natp; \hat \natp_{n-1}^\text{MAP})  \; .
+    \hat \nat_n^\text{MAP}
+    &= \argmin_\nat - \log p(\nat | \cD_n) \\
+    &= \argmin_\nat - \log p(X_n|\nat) - \log p(\nat | \cD_{n-1}) \\
+    &= \argmin_\nat f(\nat; X_{n}) + (n+n_0 - 1) \bregman(\nat; \hat \nat_{n-1}^\text{MAP})  \; .
     \label{eq:SBPP}
 \end{align} 
-Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\natp) = \E_X[-\log p(X|\natp)]$ , with step-size $\lr_n = \inv{n+n_0 -1}$.
+Coincidentally, this is exactly the formula of a Stochastic Bregman Proximal Point update on the function $f(\nat) = \E_X[-\log p(X|\nat)]$ , with step-size $\lr_n = \inv{n+n_0 -1}$.
 As a reminder this algorithm is defined with the update
 \begin{align}
-	 \hat \natp_n
-    &= \argmin_\natp \lr_n f(\natp; X_{n}) + \bregman(\natp; \hat \natp_{n-1}) \\
-    \iff \nabla \logpart(\natp_{n-1}) &= \nabla\logpart(\natp_n) + \lr_t \nabla f(\natp_n ; X_t) \; .
-\end{align}
-In words, if it converges to $\natp^*$, it can be seen as a backward mirror ascent starting from $\natp^*$.
+	 \hat \nat_n
+    &= \argmin_\nat \lr_n f(\nat; X_{n}) + \bregman(\nat; \hat \nat_{n-1}) \\
+    \iff \nabla \logpart(\nat_{n-1}) &= \nabla\logpart(\nat_n) + \lr_t \nabla f(\nat_n ; X_t) \; .
+\end{align}
+In words, if it converges to $\nat^*$, it can be seen as a backward mirror ascent starting from $\nat^*$.
 
@@
 \begin{align}
-	 \forall \natp, \lr_t f(\natp) + \cB(\natp ; \natp_t) \geq \lr_t f(\natp_{t+1}) + \cB( \natp; \natp_{t+1}) +\cB(\natp_{t+1} ; \natp_{t}) \; .
+	 \forall \nat, \lr_t f(\nat) + \cB(\nat ; \nat_t) \geq \lr_t f(\nat_{t+1}) + \cB( \nat; \nat_{t+1}) +\cB(\nat_{t+1} ; \nat_{t}) \; .
 \end{align}
@@
 \paragraph{Monotonicity of Function Values.}
-Applying it with $\natp=\natp_t$, we prove that function values are decreasing
-\begin{align}
-	f(\natp_{t+1})  \leq f(\natp_{t})  - \inv{\lr_t} \underbrace{\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) }_{=: \cS(\natp_{t} ; \natp_{t+1})}
+Applying it with $\nat=\nat_t$, we prove that function values are decreasing
+\begin{align}
+	f(\nat_{t+1})  \leq f(\nat_{t})  - \inv{\lr_t} \underbrace{\left (\cB(\nat_{t+1} ; \nat_{t}) + \cB(\nat_{t} ; \nat_{t+1}) \right) }_{=: \cS(\nat_{t} ; \nat_{t+1})}
 \end{align}
@@
 \paragraph{Descent Lemma and Telescopic Sum.}
-Applying the three point lemma with $\natp=\natp^*$ (assuming such a minimizer exists), we obtain a telescopic sum
-\begin{align}
-	\lr_t(f(\natp_{t+1}) - f(\natp^*)) +  \cB(\natp_{t+1} ; \natp_{t}) 
-	&\leq  \cB( \natp^*; \natp_{t})  - \cB( \natp^*; \natp_{t+1}) \\
-	\implies \sum_{t=0}^{T-1} \lr_t (f(\natp_{t+1}) - f(\natp^*)) 
-	+ \sum_{t=0}^{T-1} \cB(\natp_{t+1} ; \natp_{t}) 
-	&\leq \cB( \natp^*; \natp_{0})  - \cB( \natp^*; \natp_{T})  
-	\leq \cB( \natp^*; \natp_{0})  \; ,
-\end{align}
-which proves on the way that the total path length $ \sum_{t=0}^{t-1} \cB(\natp_{t+1} ; \natp_{t}) $ is bounded.
-Using the monotonicity of $f(\natp_t)$, we get
-\begin{align}
-	f(\natp_{T}) - f(\natp^*) \leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t} \; .
+Applying the three point lemma with $\nat=\nat^*$ (assuming such a minimizer exists), we obtain a telescopic sum
+\begin{align}
+	\lr_t(f(\nat_{t+1}) - f(\nat^*)) +  \cB(\nat_{t+1} ; \nat_{t}) 
+	&\leq  \cB( \nat^*; \nat_{t})  - \cB( \nat^*; \nat_{t+1}) \\
+	\implies \sum_{t=0}^{T-1} \lr_t (f(\nat_{t+1}) - f(\nat^*)) 
+	+ \sum_{t=0}^{T-1} \cB(\nat_{t+1} ; \nat_{t}) 
+	&\leq \cB( \nat^*; \nat_{0})  - \cB( \nat^*; \nat_{T})  
+	\leq \cB( \nat^*; \nat_{0})  \; ,
+\end{align}
+which proves on the way that the total path length $ \sum_{t=0}^{t-1} \cB(\nat_{t+1} ; \nat_{t}) $ is bounded.
+Using the monotonicity of $f(\nat_t)$, we get
+\begin{align}
+	f(\nat_{T}) - f(\nat^*) \leq \frac{\cB( \nat^*; \nat_{0}) }{\sum_{t=0}^{T-1} \lr_t} \; .
 \end{align}
@@
 \begin{align}
-\boxed{\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t+1})} \leq \lr_t \sigma^2 \; , }
+\boxed{\expect{f(\nat_{t+1}) - f(\nat_{t+1} ; X_{t+1})} \leq \lr_t \sigma^2 \; , }
 \label{eq:proximal_variance}
@@
 Asymptotically, this assumption is equivalent to the one of~\citet{hanzely2018fastest}. 
-To see this, linearize $f(\natp_{t+1})$ and $f(\natp_{t+1} ; X_{t+1}) $ in $\natp_t$
-\begin{align}
-	\expect{f(\natp_{t+1}) - f(\natp_{t+1} ; X_{t+1})} 
-	&\approx \expect{f(\natp_{t}) - f(\natp_{t} ; X_{t+1})} + \expect{\langle \nabla f(\natp_t) - \nabla f(\natp_t ; X_{t+1}) ; \natp_{t+1} - \natp_t \rangle } \\
-	&= 0 + \expect{\langle \nabla f(\natp_t) - \nabla f(\natp_t ; X_{t+1}) ; \natp_{t+1} \rangle } \\
-	& = \Cov(- \nabla f(\natp_t ; X_{t+1})  ; \natp_{t+1} ) \; .
+To see this, linearize $f(\nat_{t+1})$ and $f(\nat_{t+1} ; X_{t+1}) $ in $\nat_t$
+\begin{align}
+	\expect{f(\nat_{t+1}) - f(\nat_{t+1} ; X_{t+1})} 
+	&\approx \expect{f(\nat_{t}) - f(\nat_{t} ; X_{t+1})} + \expect{\langle \nabla f(\nat_t) - \nabla f(\nat_t ; X_{t+1}) ; \nat_{t+1} - \nat_t \rangle } \\
+	&= 0 + \expect{\langle \nabla f(\nat_t) - \nabla f(\nat_t ; X_{t+1}) ; \nat_{t+1} \rangle } \\
+	& = \Cov(- \nabla f(\nat_t ; X_{t+1})  ; \nat_{t+1} ) \; .
 \end{align}
@@
 \paragraph{Monotonicity of Expected Function Values.}
-Applying the tree point lemma with $\natp=\natp_t$ gives
-\begin{align}
-	f(\natp_{t+1}, X_{t+1})  
-	&\leq f(\natp_{t}, X_{t+1})  - \inv{\lr_t}\left (\cB(\natp_{t+1} ; \natp_{t}) + \cB(\natp_{t} ; \natp_{t+1}) \right) \\
-	&= f(\natp_{t}, X_{t+1})  - \inv{\lr_t} \cS(\natp_{t+1} ; \natp_{t}) \\
-	\implies \expect{f(\natp_{t+1}, X_{t+1})} 
-	&\leq f(\natp_{t}) - \inv{\lr_t}\expect{\cS(\natp_{t+1} ; \natp_{t})}  \; .
+Applying the tree point lemma with $\nat=\nat_t$ gives
+\begin{align}
+	f(\nat_{t+1}, X_{t+1})  
+	&\leq f(\nat_{t}, X_{t+1})  - \inv{\lr_t}\left (\cB(\nat_{t+1} ; \nat_{t}) + \cB(\nat_{t} ; \nat_{t+1}) \right) \\
+	&= f(\nat_{t}, X_{t+1})  - \inv{\lr_t} \cS(\nat_{t+1} ; \nat_{t}) \\
+	\implies \expect{f(\nat_{t+1}, X_{t+1})} 
+	&\leq f(\nat_{t}) - \inv{\lr_t}\expect{\cS(\nat_{t+1} ; \nat_{t})}  \; .
 	\label{eq:SPP_monotone}
 \end{align}
-This inequality bears on $\expect{f(\natp_{t+1}, X_{t+1})}$, which features a correlation between $\natp_{t+1}$ and $X_{t+1}$, and is different from $f(\natp_{t+1})$.
+This inequality bears on $\expect{f(\nat_{t+1}, X_{t+1})}$, which features a correlation between $\nat_{t+1}$ and $X_{t+1}$, and is different from $f(\nat_{t+1})$.
 Combining~\eqref{eq:SPP_monotone} with the variance assumption gives
 \begin{align}
-	\expect{f(\natp_{t+1})} 
-	&\leq \expect{f(\natp_{t+1} ; X_{t+1})}  + \lr_t \sigma^2 \\
-	&\leq f(\natp_t) +  \lr_t \sigma^2 - \inv{\lr_t}\expect{\cS(\natp_{t+1} ; \natp_{t})}
+	\expect{f(\nat_{t+1})} 
+	&\leq \expect{f(\nat_{t+1} ; X_{t+1})}  + \lr_t \sigma^2 \\
+	&\leq f(\nat_t) +  \lr_t \sigma^2 - \inv{\lr_t}\expect{\cS(\nat_{t+1} ; \nat_{t})}
 \end{align}
@@
 \begin{align}
-	\sigma^2 \leq \inv{\lr_t^2}\expect{\cS(\natp_{t+1} ; \natp_{t})} \; .
+	\sigma^2 \leq \inv{\lr_t^2}\expect{\cS(\nat_{t+1} ; \nat_{t})} \; .
 \end{align}
@@
 \paragraph{Descent Lemma and Telescopic Sum.}
-Similarly, applying the three points lemma with $\natp^*$ yields
-\begin{align}
-	\lr_t(f(\natp_{t+1},X_{t+1}) - f(\natp^*)) +  \cB(\natp_{t+1} ; \natp_{t})
-	&\leq  \cB( \natp^*; \natp_{t})  - \cB( \natp^*; \natp_{t+1}) \; .
-\end{align}
-Ignoring the distance between iterates $\cB(\natp_{t+1} ; \natp_{t})
+Similarly, applying the three points lemma with $\nat^*$ yields
+\begin{align}
+	\lr_t(f(\nat_{t+1},X_{t+1}) - f(\nat^*)) +  \cB(\nat_{t+1} ; \nat_{t})
+	&\leq  \cB( \nat^*; \nat_{t})  - \cB( \nat^*; \nat_{t+1}) \; .
+\end{align}
+Ignoring the distance between iterates $\cB(\nat_{t+1} ; \nat_{t})
 $, taking the tower expectation, and telescoping the sum, we get
 \begin{align}
-	\sum_{t=0}^{t-1} \lr_t \expect{f(\natp_{t+1},X_{t+1}) - f(\natp^*)}
-	\leq \cB( \natp^*; \natp_{0})  \; .
+	\sum_{t=0}^{t-1} \lr_t \expect{f(\nat_{t+1},X_{t+1}) - f(\nat^*)}
+	\leq \cB( \nat^*; \nat_{0})  \; .
 \end{align}
@@
 \begin{align}
-	 \min_{1\leq t \leq T} \expect{ f(\natp_t)} - f(\natp^*)
-	\leq \frac{\cB( \natp^*; \natp_{0}) }{\sum_{t=0}^{T-1} \lr_t}
+	 \min_{1\leq t \leq T} \expect{ f(\nat_t)} - f(\nat^*)
+	\leq \frac{\cB( \nat^*; \nat_{0}) }{\sum_{t=0}^{T-1} \lr_t}
 	+ \frac{\sum_{t=0}^{T-1} \lr_t^2  }{\sum_{t=0}^{T-1} \lr_t} \sigma^2 \; .
@@
 \paragraph{Application to the MAP.}
-The variance assumption \eqref{eq:proximal_variance} translates to $\Cov(T(x_{n}) , \natp_{n+1}) \leq  \gamma_n \sigma^2$ in the exponential family MAP setting -- e.g.
+The variance assumption \eqref{eq:proximal_variance} translates to $\Cov(T(x_{n}) , \nat_{n+1}) \leq  \gamma_n \sigma^2$ in the exponential family MAP setting -- e.g.
 exactly the same assumption as in \citet{hanzely2018fastest}.
@@
 \begin{align}
-	\natp_{t+1} 
-	&:= \argmin_\natp \lr \linear_f(\natp; \natp_t) + \bregman(\natp ; \natp_t)  \\
-	& = \nabla \conj (\nabla \logpart(\natp_t) - \lr \nabla f(\natp_t))
-\end{align}
-where $\linear_f(\natp; \natp_t) = f(\natp_t) + \langle \nabla f(\natp_t), \natp - \natp_t \rangle$ is the linear approximation of $f$ in $\natp_t$ evaluated at $\natp$.
-Solving this problem require solving problems of the form $\argmin_\natp - \langle c, \natp \rangle + \logpart(\natp)$, eg computing the convex conjugate of $\logpart$.
-Note that finding $\natp_*$ is done with 1 step of mirror descent with step-size 1.
+	\nat_{t+1} 
+	&:= \argmin_\nat \lr \linear_f(\nat; \nat_t) + \bregman(\nat ; \nat_t)  \\
+	& = \nabla \conj (\nabla \logpart(\nat_t) - \lr \nabla f(\nat_t))
+\end{align}
+where $\linear_f(\nat; \nat_t) = f(\nat_t) + \langle \nabla f(\nat_t), \nat - \nat_t \rangle$ is the linear approximation of $f$ in $\nat_t$ evaluated at $\nat$.
+Solving this problem require solving problems of the form $\argmin_\nat - \langle c, \nat \rangle + \logpart(\nat)$, eg computing the convex conjugate of $\logpart$.
+Note that finding $\nat_*$ is done with 1 step of mirror descent with step-size 1.
 Indeed plugging in definitions of $f$ and $\mu$, and assuming constant step-size $\lr$ yields
@@
 \paragraph{Stochastic Mirror Descent}
-Let $f(\natp, x) = -\log p(x|\natp)$.
-MAP iterations can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\natp_t, x_{t+1}) = \logpart(\natp_t) - T(x_{t+1})$ as a stochastic estimate  of $\nabla f(\natp_t)$.
+Let $f(\nat, x) = -\log p(x|\nat)$.
+MAP iterations can be cast as stochastic mirror descent (SMD),  with $g_t = \nabla f(\nat_t, x_{t+1}) = \logpart(\nat_t) - T(x_{t+1})$ as a stochastic estimate  of $\nabla f(\nat_t)$.
 Resuming from the Stochastic Bregman Proximal Point formula~\eqref{eq:SBPP}, we write $\MAPt$ for the n-th MAP iterate, and do some Bregman manipulations
 \begin{align}
-    \hat \natp_{n+1}
-    &= \argmin_\natp 
-    - \langle T(x_{t+1}), \natp \rangle  + \logpart(\natp) + (n_0 + n) \bregman(\natp ; \natp_n) \\
-    %&(n_0+n)\left(\logpart(\natp) - \logpart(\natp_n) - \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle \right)  \\
-    &= \argmin_\natp - \langle T(x_{t+1}), \natp \rangle   + \logpart(\natp_n) + \langle \nabla \logpart(\natp_n)  , \natp - \natp_n \rangle + (n_0+n + 1) \bregman(\natp ; \natp_n) \\
-    &= \argmin_\natp \linear_f(\natp;\natp_n, x_{t+1}) + (n_0 + n + 1) \bregman(\natp ; \natp_n)
-\end{align}
-where $\linear_f(\natp;\natp_n, x_{t+1})$ is the stochastic linearization of $f$ at $\natp_n$ evaluated at $\natp$ with randomness coming from $x_{t+1}$.
+    \hat \nat_{n+1}
+    &= \argmin_\nat 
+    - \langle T(x_{t+1}), \nat \rangle  + \logpart(\nat) + (n_0 + n) \bregman(\nat ; \nat_n) \\
+    %&(n_0+n)\left(\logpart(\nat) - \logpart(\nat_n) - \langle \nabla \logpart(\nat_n)  , \nat - \nat_n \rangle \right)  \\
+    &= \argmin_\nat - \langle T(x_{t+1}), \nat \rangle   + \logpart(\nat_n) + \langle \nabla \logpart(\nat_n)  , \nat - \nat_n \rangle + (n_0+n + 1) \bregman(\nat ; \nat_n) \\
+    &= \argmin_\nat \linear_f(\nat;\nat_n, x_{t+1}) + (n_0 + n + 1) \bregman(\nat ; \nat_n)
+\end{align}
+where $\linear_f(\nat;\nat_n, x_{t+1})$ is the stochastic linearization of $f$ at $\nat_n$ evaluated at $\nat$ with randomness coming from $x_{t+1}$.
 This is the formula for stochastic mirror descent (SMD) applied to $f$ with mirror map $\logpart$ and step-size $\gamma_n = \inv{n_0 + n + 1}$.
@@
 Let us have a look at their variance assumption \citep[Assumption 5.1]{hanzely2018fastest}.
-Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\natp_{t+1}$ the next iterate.
-Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the next iterate $\natp_{t+1} = \nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t g_t )$ that should hold for all time steps
+Let $g_t$ be the random gradient at step $t$  (coming from data point $x_t$), and $\nat_{t+1}$ the next iterate.
+Then the variance bound $\sigma^2$ is an upper bound on the covariance between the gradient update  $-g_t$ and the next iterate $\nat_{t+1} = \nabla \logpart^*(\nabla \logpart(\nat_t) - \lr_t g_t )$ that should hold for all time steps
 \begin{align}
 	\lr_t \sigma^2 \geq 
-	&\Cov(- g_t,\natp_{t+1} | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
-	 &= \expect{\langle \nabla f(\natp_t) - g_t , \natp_{t+1}  \rangle}
+	&\Cov(- g_t,\nat_{t+1} | X_{1 \dots t})  \label{eq:hanzely_hypothesis} \\
+	 &= \expect{\langle \nabla f(\nat_t) - g_t , \nat_{t+1}  \rangle}
 \end{align}
@@
 Remark that when we plug in $\conj = \|.\|^2$, we recover the gradient variance typical of SGD. 
-If we introduce $\bar \natp_{t+1}=\nabla \logpart^*(\nabla \logpart(\natp_t) - \lr_t  \nabla f(\natp_t) )$ the theoretical output of a deterministic Bregman gradient step, then following Lemma~\ref{lem:covariance} we can rewrite this covariance as the expectation of symmetrized Bregman between stochastic and deterministic iterates $\expect[X_t]{\bregman(\natp_{t+1} ; \bar \natp_{t+1} ) + \bregman(\bar \natp_{t+1} ; \natp_{t+1} ) }$.
+If we introduce $\bar \nat_{t+1}=\nabla \logpart^*(\nabla \logpart(\nat_t) - \lr_t  \nabla f(\nat_t) )$ the theoretical output of a deterministic Bregman gradient step, then following Lemma~\ref{lem:covariance} we can rewrite this covariance as the expectation of symmetrized Bregman between stochastic and deterministic iterates $\expect[X_t]{\bregman(\nat_{t+1} ; \bar \nat_{t+1} ) + \bregman(\bar \nat_{t+1} ; \nat_{t+1} ) }$.
 This view is useful to understand that we are measuring the variance of each iteration, which is necessarily positive.
@@
 \begin{align}
-	\nabla f(\natp_t) - g_t =  T(X_{t+1})	- \E[T(X)]
-\end{align}
-so that $\sigma$ is really a bound on the covariance of the sufficient statistics $T(X_{t+1})$ with the next iterate $\natp_{t+1}$, which itself depends on $T(X_{t+1})$ via a coefficient $\lr_t$ and a non-linearity $\nabla\conj$.
-\begin{align}
-	&\Cov( T(X_{t+1}), \natp_{t+1} | X_{1\dots t})
+	\nabla f(\nat_t) - g_t =  T(X_{t+1})	- \E[T(X)]
+\end{align}
+so that $\sigma$ is really a bound on the covariance of the sufficient statistics $T(X_{t+1})$ with the next iterate $\nat_{t+1}$, which itself depends on $T(X_{t+1})$ via a coefficient $\lr_t$ and a non-linearity $\nabla\conj$.
+\begin{align}
+	&\Cov( T(X_{t+1}), \nat_{t+1} | X_{1\dots t})
 	\leq \lr_t\sigma^2 
@@
 so we recover a $O(n^{-1})$ upper bound on the variance term and we are still looking for the $O(n^{-2})$ rate for the bias term.
-It is simpler to consider the primal pivot, but still hard to say anything about $\bregmanconj(\mu^* ; \tilde \mu_n) = \bregman(\E[\MAPt]; \natp^*)$ without further assumptions.
+It is simpler to consider the primal pivot, but still hard to say anything about $\bregmanconj(\mu^* ; \tilde \mu_n) = \bregman(\E[\MAPt]; \nat^*)$ without further assumptions.
 
@@
 	2 \lr^2 \sigma^2
-	\geq \expect[X]{\bregmanconj(\mu_t - 2 \lr_t \nabla f(\natp^*, X) ; \mu_t ) } \; .
+	\geq \expect[X]{\bregmanconj(\mu_t - 2 \lr_t \nabla f(\nat^*, X) ; \mu_t ) } \; .
 	\label{eq:dragomir_hypothesis}
@@
 \begin{align}
-	f(\natp^*) - \E_{X\sim\natp^*}[\min_\natp f(\natp ; X)] < \infty
+	f(\nat^*) - \E_{X\sim\nat^*}[\min_\nat f(\nat ; X)] < \infty
 \end{align}
@@
 \begin{align}
-\forall X, \min_\natp f(\natp ; X) >	 - \infty \iff \max_\natp p(X|\natp) < \infty
+\forall X, \min_\nat f(\nat ; X) >	 - \infty \iff \max_\nat p(X|\nat) < \infty
 \end{align}
@@
 \end{align}
-where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\natp^t$ and $\natp^*$ ; and $\lr_t \leq \inv{2L}$.
+where $B_t$ is some measure of suboptimality, in this case the $\ell^2$ distance between $\nat^t$ and $\nat^*$ ; and $\lr_t \leq \inv{2L}$.
 On the other hand, the problem of most Bregman analysis \citep{hanzely2018fastest,dorazio2021stochastic} is that they do not get this $\lr^2$ in front the variance.
@@
 Note that this equivalence holds only if sufficient statistics are within $\dom \conj  = \mathrm{Im} (\nabla\logpart)$, a point ignored by authors yet critical as it does not hold for multivariate normals with unknown covariance.
-\citet{raskutti2015information} highlight the equivalence between mirror descent in $\natp$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.
+\citet{raskutti2015information} highlight the equivalence between mirror descent in $\nat$ and natural gradient descent in $\meanp$ (or vice-versa) for the exponential family.
 Then they use the optimality result by Shunichi Amari on Natural Gradient Descent to show that Mirror Descent iterates reach the Cramer-Rao lower bound.
@@
 Although this hypothesis is far from intuitive, \citet{lu2019relative} obtained convergence rates for mirror descent that are very similar to typical SGD analysis.
-In this work we are concerned with strictly convex differentiable functions $\logpart(\natp)$, to ensure a clean bijection between natural and mean parameters.
+In this work we are concerned with strictly convex differentiable functions $\logpart(\nat)$, to ensure a clean bijection between natural and mean parameters.
 Consequently, we do not further investigate relative continuity.
@@
 	&=  \frac{m}{\sigma^2} X -\inv{2 \sigma^2} X^2 - \frac{m^2}{2 \sigma^2} - \half \log \sigma^2 \\
-	&=  \natp_1 X + \natp_2 X^2  - \logpart(\natp) \; .
+	&=  \nat_1 X + \nat_2 X^2  - \logpart(\nat) \; .
 \end{align}
@@
 \begin{align}
-	\natp_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
-	\natp_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
+	\nat_1 &= \frac{m}{\sigma^2} = \frac{\mu_1}{\mu_2 - \mu_1^2} \\
+	\nat_2 &= -\inv{2 \sigma^2} =  \frac{ - 1}{2(\mu_2 - \mu_1^2)}  < 0
 \end{align}
@@
 	\logpart(m, \sigma) &= \frac{m^2}{2 \sigma^2} + \half \log \sigma^2 \\
-	\logpart(\natp) &= \frac{\natp_1^2}{-4\natp_2} - \half \log(-2\natp_2) \; .
-\end{align}
-We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\natp)$
-\begin{align}
-	\mu_1 &= \frac{\natp_1}{-2\natp_2}  \\
-	\mu_2 & = \left( \frac{\natp_1}{-2\natp_2} \right)^2 + \inv{-2 \natp_2} \; .
+	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \; .
+\end{align}
+We can perform a sanity check by deriving $\logpart$ and verifying that its gradient gives $\mu(\nat)$
+\begin{align}
+	\mu_1 &= \frac{\nat_1}{-2\nat_2}  \\
+	\mu_2 & = \left( \frac{\nat_1}{-2\nat_2} \right)^2 + \inv{-2 \nat_2} \; .
 \end{align}
@@
 \begin{align}
-	\conj (\mu) &= \mu_1 \natp_1 + \mu_2 \natp_2  - \logpart(\natp(\mu)) \\
+	\conj (\mu) &= \mu_1 \nat_1 + \mu_2 \nat_2  - \logpart(\nat(\mu)) \\
 	& = \frac{\mu_1^2}{\mu_2 - \mu_1^2}  - \frac{\mu_2}{2(\mu_2 - \mu_1^2)} - \frac{\mu_1^2}{2(\mu_2 - \mu_1^2)}  - \half \log(\mu_2 - \mu_1^2) \\
@@
 Note that the strong convexity constant $m$ may be zero.
-Recall that $\ell_f(\natp, \natp^{t}) = f(\natp^t) + \langle \nabla f (\natp^t) ; \natp - \natp^t \rangle $ is the linearization of $f$ at $\natp^t$ evaluated in $\natp$.
-\begin{align}
-	f(\natp^{t+1}) 
-	&\leq \ell_f(\natp^{t+1}, \natp^{t})  + L \bregman(\natp^{t+1}; \natp^{t})
+Recall that $\ell_f(\nat, \nat^{t}) = f(\nat^t) + \langle \nabla f (\nat^t) ; \nat - \nat^t \rangle $ is the linearization of $f$ at $\nat^t$ evaluated in $\nat$.
+\begin{align}
+	f(\nat^{t+1}) 
+	&\leq \ell_f(\nat^{t+1}, \nat^{t})  + L \bregman(\nat^{t+1}; \nat^{t})
 	& \text{(Relative $L$-smoothness)} \\
-	& \leq \ell_f(\natp^{*}, \natp^{t})  + L \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
-	& \text{(3 Points Lemma w. $x=\natp^*$)} \\
-	& \leq f(\natp^*) + (L - m) \bregman(\natp^{*}; \natp^{t}) -L \bregman(\natp^{*}; \natp^{t+1}) 
+	& \leq \ell_f(\nat^{*}, \nat^{t})  + L \bregman(\nat^{*}; \nat^{t}) -L \bregman(\nat^{*}; \nat^{t+1}) 
+	& \text{(3 Points Lemma w. $x=\nat^*$)} \\
+	& \leq f(\nat^*) + (L - m) \bregman(\nat^{*}; \nat^{t}) -L \bregman(\nat^{*}; \nat^{t+1}) 
 	& \text{(Relative $m$-strong convexity)}
@@
 \paragraph{Monotonicity.}
-Applying the 3 points lemma with $x=\natp_t$, 
-\begin{align}
-	f(\natp^{t+1}) \leq  f(\natp^t) -L (\bregman(\natp^{t}, \natp^{t+1}) + \bregman(\natp^{t+1} , \natp^{t} ))
+Applying the 3 points lemma with $x=\nat_t$, 
+\begin{align}
+	f(\nat^{t+1}) \leq  f(\nat^t) -L (\bregman(\nat^{t}, \nat^{t+1}) + \bregman(\nat^{t+1} , \nat^{t} ))
 \end{align}
@@
 \begin{align}
-	\sum_{t=0}^{T-1} (f(\natp^{t+1}) - f(\natp^*)) 
-	&\leq L \sum_t \bregman(\natp^{*}; \natp^{t}) - \bregman(\natp^{*}); \natp^{t+1}) \\
-	&= L \bregman(\natp^{*}; \natp^{0}) - L\bregman(\natp^{*}); \natp^{T})\\
-	&\leq L \bregman(\natp^{*}; \natp^{0}) \; ,
-\end{align}
-from which we can conclude thanks to the monotonicity of $f(\natp^t)$
-\begin{align}
-	f(\natp^{T}) - f(\natp^*)
-	\leq  \frac{L}{T} \bregman(\natp^{*}; \natp^{0}) \; ,
+	\sum_{t=0}^{T-1} (f(\nat^{t+1}) - f(\nat^*)) 
+	&\leq L \sum_t \bregman(\nat^{*}; \nat^{t}) - \bregman(\nat^{*}); \nat^{t+1}) \\
+	&= L \bregman(\nat^{*}; \nat^{0}) - L\bregman(\nat^{*}); \nat^{T})\\
+	&\leq L \bregman(\nat^{*}; \nat^{0}) \; ,
+\end{align}
+from which we can conclude thanks to the monotonicity of $f(\nat^t)$
+\begin{align}
+	f(\nat^{T}) - f(\nat^*)
+	\leq  \frac{L}{T} \bregman(\nat^{*}; \nat^{0}) \; ,
 \end{align}
@@
 \begin{align}
-	\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\natp^{t+1}) - f(\natp^*)) 
-	&\leq \sum_{t=0}^{T-1} \frac{( L-m)^{T-t}}{L^{T-t -1}} \bregman(\natp^{*}; \natp^{t}) 
-	- \frac{( L-m)^{T-t-1}}{L^{T-t-2}} \bregman(\natp^{*}); \natp^{t+1}) \\
-	&= \frac{( L-m)^{T}}{L^{T-1}}  \bregman(\natp^{*}; \natp^{0}) 
-	- L \bregman(\natp^{*}); \natp^{T}) \\
-	&\leq L \left( \frac{L-m}{L} \right)^T \bregman(\natp^{*}; \natp^{0}) \; .
-\end{align}
-Then using the monotonicity of $f(\natp^t)$, we end up with the geometric sum of $\frac{L-m}{L} < 1$
-\begin{align}
-		\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\natp^{t+1}) - f(\natp^*)) 
-		&\geq (f(\natp^{T}) - f(\natp^*)) \sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} \\
-		&\geq (f(\natp^{T}) - f(\natp^*)) \inv{1 - \frac{L-m}{L}} \\
-		&= (f(\natp^{T}) - f(\natp^*)) \frac{L}{m} \; .
+	\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\nat^{t+1}) - f(\nat^*)) 
+	&\leq \sum_{t=0}^{T-1} \frac{( L-m)^{T-t}}{L^{T-t -1}} \bregman(\nat^{*}; \nat^{t}) 
+	- \frac{( L-m)^{T-t-1}}{L^{T-t-2}} \bregman(\nat^{*}); \nat^{t+1}) \\
+	&= \frac{( L-m)^{T}}{L^{T-1}}  \bregman(\nat^{*}; \nat^{0}) 
+	- L \bregman(\nat^{*}); \nat^{T}) \\
+	&\leq L \left( \frac{L-m}{L} \right)^T \bregman(\nat^{*}; \nat^{0}) \; .
+\end{align}
+Then using the monotonicity of $f(\nat^t)$, we end up with the geometric sum of $\frac{L-m}{L} < 1$
+\begin{align}
+		\sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} (f(\nat^{t+1}) - f(\nat^*)) 
+		&\geq (f(\nat^{T}) - f(\nat^*)) \sum_{t=0}^{T-1} \left( \frac{L-m}{L} \right)^{T-1-t} \\
+		&\geq (f(\nat^{T}) - f(\nat^*)) \inv{1 - \frac{L-m}{L}} \\
+		&= (f(\nat^{T}) - f(\nat^*)) \frac{L}{m} \; .
 \end{align}
@@
 \begin{align}
-	f(\natp^{T}) - f(\natp^*)
-	\leq m  \left( 1 - \frac{m}{L} \right)^T \bregman(\natp^{*}; \natp^{0}) 
+	f(\nat^{T}) - f(\nat^*)
+	\leq m  \left( 1 - \frac{m}{L} \right)^T \bregman(\nat^{*}; \nat^{0}) 
 \end{align}
@@
 In statistics, there is this famous theorem known as the Cramer-Rao lower bound.
-In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\natp$ is lower bounded by the inverse Fisher information matrix, divided by $n$
-\begin{align}
-    \Cov(\hat \natp_n)  
-    &\succeq \inv{n} \cI(\natp_*)^{-1} \\
+In its simplest form, it states that the covariance of an unbiased estimator of a distribution's parameter $\nat$ is lower bounded by the inverse Fisher information matrix, divided by $n$
+\begin{align}
+    \Cov(\hat \nat_n)  
+    &\succeq \inv{n} \cI(\nat_*)^{-1} \\
     \iff
-    \expect[\cD\sim \natp_*]{(\hat \natp_n - \natp_*)(\hat \natp_n - \natp_*)^T}
-    &\succeq \inv{n} \expect[X\sim \natp_*]{-\nabla^2 \log p(X|\natp_*)}^{-1} \; .
+    \expect[\cD\sim \nat_*]{(\hat \nat_n - \nat_*)(\hat \nat_n - \nat_*)^T}
+    &\succeq \inv{n} \expect[X\sim \nat_*]{-\nabla^2 \log p(X|\nat_*)}^{-1} \; .
 \end{align}
@@
 
-Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\natp)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
-For the estimator $\hat \natp_n$, this lower bound is not trivial.
+Note that in the case of an exponential family, the lower bound is always reached for the MLE estimator $\hat \mu_n$, because $\cI(\mu) = \nabla^2 \conj(\mu) = \nabla^2 \logpart(\nat)^{-1} = \Cov(T(X))^{-1}$, where the first step takes some calculus.
+For the estimator $\hat \nat_n$, this lower bound is not trivial.
 
@@
 
-A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\natp$ is written as
-\begin{align}
-    p(\natp) 
-    \propto \exp(-n_0 \bregman(\natp ; \natp_0)) 
-    \propto \exp(-n_0 \logpart(\natp) + \langle n_0 \mu_0 ; \natp \rangle) \; .
-\end{align}
-We see that $p(\natp)$ is part of the exponential family, with sufficient statistic $(\natp, \logpart(\natp))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
-When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $p(\natp | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
+A conjugate prior for the exponential family with sufficient statistic $T(X)$ and natural parameter $\nat$ is written as
+\begin{align}
+    p(\nat) 
+    \propto \exp(-n_0 \bregman(\nat ; \nat_0)) 
+    \propto \exp(-n_0 \logpart(\nat) + \langle n_0 \mu_0 ; \nat \rangle) \; .
+\end{align}
+We see that $p(\nat)$ is part of the exponential family, with sufficient statistic $(\nat, \logpart(\nat))$ and natural parameters $(n_0 \mu_0 , -n_0)$.
+When we update this prior with Bayes formula on dataset $\cD=(x_1,\dots,x_n)$, we get a posterior $p(\nat | \cD)$ with natural parameters $(n_0 \mu_0 + \sum_i T(x_i) , -(n_0 + n))$.
 This is well detailed in \citet{agarwal2010geometric}.
@@
 \begin{align}
-    \expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+    \expect[\cD\sim\nat_*]{\expect[p(\nat|\cD)]{\bregman(\nat ; \nat_*)}}
 \end{align}
@@
 	\begin{align}
-		\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)} = \bregman(\hat \natp_n ; \natp_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
+		\expect[p(\nat|\cD)]{\bregman(\nat ; \nat_*)} = \bregman(\hat \nat_n ; \nat_*) + \frac{2}{n} \frac{\mu_*}{\hat \mu_n} + \log\frac{n}{2} - \psi(1 + \frac{n}{2})
 	\end{align}
@@
 	\begin{align}
-		\expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		\expect[\cD\sim\nat_*]{\expect[p(\nat|\cD)]{\bregman(\nat ; \nat_*)}}
 		= \frac{2}{n} + \frac{2}{n(n-2)} + \psi(\half[n]) - \psi(1+ \half[n])
@@
 		\inv{2n} + \frac{4}{n(n-2)}
-		\leq \expect[\cD\sim\natp_*]{\expect[p(\natp|\cD)]{\bregman(\natp ; \natp_*)}}
+		\leq \expect[\cD\sim\nat_*]{\expect[p(\nat|\cD)]{\bregman(\nat ; \nat_*)}}
 		\leq 
@@
 	\begin{proof}
-	The ratio to the optimum  $\frac{\mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$ divided by $n$.
+	The ratio to the optimum  $\frac{n \mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$.
 Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
 	\begin{align}
-		\expect{\frac{\mu_*}{ \mu_n}  - 1} 
+		\expect{n \frac{\mu_*}{n \mu_n}  - 1} 
 		= \expect{\frac{n}{\chi^2(n)} - 1} =
@@
 	\end{align}
-	There is also a closed form solution for the expected logarithm of a Chi-squared\citep{pav2015moments}
+	There is also a closed form solution for the expected logarithm of a Chi-squared, because a Chi-squared is a special case of Gamma distributions, whose sufficient statistic is $(X,\log(X))$.
+For non-central chi-squares, one could also look at \citep{pav2015moments}.
 	\begin{align}
@@
 	\end{proof}
+
 Using the \href{https://en.wikipedia.org/wiki/Digamma_function#Recurrence_formula_and_characterization}{Stirling series}, we can also get an asymptotic approximation
@@
 		\leq \expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\natp^*}} 
+		= \expect{\frac{\MAPt}{\nat^*}} 
 		\leq \begin{cases}
@@
 		\expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\natp^*}} 
+		= \expect{\frac{\MAPt}{\nat^*}} 
 		&= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  \\
@@
 	= \nabla \conj(\mu^*) + \nabla^2\conj(\mu^*)(\mu - \mu^*) + O(\norm{\mu - \mu^*}^2) \\
-	\implies &\hat \natp_n 
-	= \natp^* + \Sigma^{-1} (\hat \mu_n - \mu^*)  + O(\E[\norm{\hat \mu_n - \mu^*}^2]) \\
-	\implies &\Cov(\hat \natp_n ; T(X) ) 
+	\implies &\hat \nat_n 
+	= \nat^* + \Sigma^{-1} (\hat \mu_n - \mu^*)  + O(\E[\norm{\hat \mu_n - \mu^*}^2]) \\
+	\implies &\Cov(\hat \nat_n ; T(X) ) 
 	= \inv{n+n_0} \E[ \norm{T(X) - \mu^*}^2_{\Sigma^{-1}}]   + O(n^{-2}) \\
-	\implies &\Cov(\hat \natp_n ; T(X) ) 
+	\implies &\Cov(\hat \nat_n ; T(X) ) 
 	 = \frac{d}{n+n_0} + O(n^{-2}) \; .
@@
 		\begin{align}
-			\Cov(\hat \natp_n ; \hat \mu_n )  
+			\Cov(\hat \nat_n ; \hat \mu_n )  
 			= \E[\MAPt \MAPm] - \E[\MAPt] \E[\MAPm] 
@@
 	\end{align}
-	where we used the fact that $\natp \mu = -\half$ for all valid pairs $(\natp, \mu)$. 
+	where we used the fact that $\nat \mu = -\half$ for all valid pairs $(\nat, \mu)$. 
 	Plugging-in \eqref{eq:theta_expectation_bound} yield
@@
 		-\half + \frac{\mu_n}{2 \mu_n}
-		&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
+		&\leq \Cov(\hat \nat_n ; \hat \mu_n ) 
 		\leq -\half + \half \frac{a + n}{a + n - 2} \\
-		\iff 0&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
+		\iff 0&\leq \Cov(\hat \nat_n ; \hat \mu_n ) 
 		\leq \half \frac{a+n - (a + n -2)}{a + n - 2} \\
-		\iff 0&\leq \Cov(\hat \natp_n ; \hat \mu_n ) 
+		\iff 0&\leq \Cov(\hat \nat_n ; \hat \mu_n ) 
 		\leq \inv{a +n - 2}
@@
 \subsection{Mixed Bias-Variance Term}
-The mixed bias variance term is $ \frac{n_0}{n+n_0} \langle  {\mu_* - \mu_0} ; {\natp_n - \E[\hat \natp_n]} \rangle $.
+The mixed bias variance term is $ \frac{n_0}{n+n_0} \langle  {\mu_* - \mu_0} ; {\nat_n - \E[\hat \nat_n]} \rangle $.
 Note that it might be negative !
@@
 \begin{align}
-	\frac{n_0}{n+n_0} \langle  \bias{\mu^* - \mu_0} ; \variance{\natp_n - \E[\hat \natp_n]} \rangle
+	\frac{n_0}{n+n_0} \langle  \bias{\mu^* - \mu_0} ; \variance{\nat_n - \E[\hat \nat_n]} \rangle
 	 = \frac{n_0}{(n+n_0)^3} (\mu^* - \mu_0)^\top \nabla^3 \conj (\mu^*) \Sigma + O(n^{-4}) = O(n^{-3}) \; .
@@
 		0
-		\leq \natp_n - \E[\MAPt]
+		\leq \nat_n - \E[\MAPt]
 		\leq \frac{n_0 +n}{2 \mu^*} \left ( \inv{a+n-2} - \inv{a+n} \right)
@@
 	Remark how this asymptotic class differs from the $O(n^{-2})$ I expected. Is this bound less tight than I thought, or are my calculus wrong ? 
-	Also remark that $\natp_n$ is always greater than $\E[\MAPt]$, which is a good sanity-check because $\nabla \conj(\mu) = - \inv{2\mu}$ is a concave function.
+	Also remark that $\nat_n$ is always greater than $\E[\MAPt]$, which is a good sanity-check because $\nabla \conj(\mu) = - \inv{2\mu}$ is a concave function.
 
@@
 	\begin{align}
-		\frac{n_0}{n+n_0} \langle  \mu^* - \mu_0 ; \natp_n - \E[\hat \natp_n] \rangle
+		\frac{n_0}{n+n_0} \langle  \mu^* - \mu_0 ; \nat_n - \E[\hat \nat_n] \rangle
 		\leq \begin{cases}

### 2021-10-18 aistats.tex
Add proof for gaussian MLE 1d

@@
 
+Let us recall the form of the divergence with different parametrizations
+\begin{align}
+	\KL( \sigma_*^2 ; \sigma_n^2 ) 
+	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
+	\bregmanconj( \mu_*; \mu_n) 
+	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
+	\bregman( \nat_n; \nat_* ) 
+	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
+\end{align}
 Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
@@
 
-For MAP, we get an upper bound thanks to
-\begin{align}
+To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focused on the symmetrized Bregman instead
+\aligns{
 	\label{eq:log_bound}
-	\phi(z) \leq \phi(z) + \phi(\inv{z}) =  \half (z + \inv{z}) - 1 \; .
-\end{align}
+	\phi(z) \leq \phi(z) + \phi(\inv{z}) 
+	&=  \half \paren{z -1  + \inv{z} -1} 
+	= \frac{(z-1)^2}{z}
+}
+\aligns{
+	\bregmanconj( \mu_*; \mu_n) 
+	&\leq \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \\ 
+	=: \cS_{\conj} (\mu_*, \mu_n )  
+	&= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 }
+}
+
+\subsection{Proof of MLE Upper Bound}
+
+	\begin{theorem}[MLE Upper Bound]
+	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
+	Its expected suboptimality is infinite when $n\leq 2$ and otherwise
+	\begin{align}
+		 \expect{\bregmanconj( \mu_*; \mu_n) }
+			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
+	\end{align}
+	\end{theorem}
+
+	\begin{proof}
+	The ratio to the optimum  $\frac{n \mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$.
+Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
+	\begin{align}
+		\expect{n \frac{\mu_*}{n \mu_n}  - 1} 
+		= \expect{\frac{n}{\chi^2(n)} - 1} =
+		\begin{cases}
+			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
+			+\infty \  \text{ otherwise. }
+		\end{cases}
+	\end{align}
+	There is also a closed form solution for the expected logarithm of a Chi-squared, because a Chi-squared is a special case of Gamma distributions, whose sufficient statistic is $(X,\log(X))$.
+For non-central chi-squares, one could also look at \citep{pav2015moments}.
+	\begin{align}
+		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
+	\end{align}
+	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
+	Consequently the suboptimality of the MLE has a closed form solution
+	\begin{align}
+	\expect{\bregmanconj( \mu_*; \mu_n) }
+	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
+	& =
+	\begin{cases}
+		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
+			+\infty \  \text{ otherwise. }
+	\end{cases}
+	\end{align}
+	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
+	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
+	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
+	\begin{align}
+		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
+	\end{align}
+	  For $n\geq 3$,
+	\begin{align}
+		\inv{n-2} - \inv{n}
+		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
+		\leq \inv{n-2} - \inv{2 n} \\
+		\iff
+			\frac{2}{n(n-2)}
+			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
+			\leq \inv{2n} +\frac{2}{n(n-2)}
+	\end{align}
+	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
+	\end{proof}
 

### 2021-10-18 aistats.tex
Enhance multivariate MLE proof and add it to aistats

@@
 
+\subsection{Multivariate MLE}
+In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, the mean parameter $\mu_*$ becomes a positive definite covariance matrix of order $d$, with $p = \frac{d(d+1)}{2} $ degrees of freedom.
+Its entropy is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, with gradient negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$, and associated Bregman divergence
+\alignn{
+	\bregmanconj(\mu_*; \hat \mu_n) 
+	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
+}
+Thanks to the linearity of the trace, the expectation becomes
+\alignn{
+	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
+	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
+	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
+}
+
+When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
+then the mean parameter "ratio" is
+\aligns{
+	\mV = n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top
+}
+but $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
+is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$. 
+We know formulas for the expectation of an inverse Wishart and the log-determinant of a Wishart (thanks to $\log\det$ being a sufficient statistics and $\nabla\logpart$)
+\aligns{
+	\E[\mV] &= n \mI \\
+	\forall n > d+1,\E[\mV^{-1}] &= \frac{\mI}{n - d - 1} 
+	\implies \Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}  \\
+	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
+}
+where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
+Putting it all together, we get the following closed form for the expectation of the KL
+ \begin{align}
+ 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
+ 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
+ \end{align}
+
+The complicated term is the $\psi_d$ function.
+We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
+\begin{align}
+	\psi_d(\half[n]) - d \log(\half[n]) 
+	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
+	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
+	\label{eq:psid_bound}
+\end{align}
+We can bound the harmonic sum thanks to the typical bound
+\aligns{
+	\inv{2n +1}
+	\leq H_n - \log(n) - \gamma \leq 
+	\inv{2n - 1}
+}
+where $\gamma$ is the Euler constant. Then
+\begin{align}
+	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
+	&= H_{n-d} - H_n \\
+	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
+	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
+\end{align}
+Plugging this back into~\eqref{eq:psid_bound} yields
+\begin{align}
+	\psi_d(\half[n]) - d \log(\half[n]) 
+	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
+	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1} 
+\end{align}
+where we emphasized that the sum runs up to $d$.
+Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
+\begin{align}
+	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
+	\leq \sum_{i=0}^{d} - \frac{i}{n}
+	 = -\frac{d(d+1)}{2n}
+\end{align}
+which yields the final rate
+\begin{align}
+ 	\boxed{
+ 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
+ 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} \;.
+ 	}
+\end{align}
+where $p = d(d+1)/2$.
+This rate is asymptotically tight.
+
 \section{PROOFS FOR GAUSSIAN}

### 2021-10-18 main.tex
Enhance multivariate MLE proof and add it to aistats

@@
 \subsection{Multivariate MLE}
-In higher dimensions $d$
+In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, the mean parameter $\mu_*$ becomes a positive definite covariance matrix of order $d$, with $p = \frac{d(d+1)}{2} $ degrees of freedom.
+Its entropy is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, with gradient negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$, and associated Bregman divergence
+\alignn{
+	\bregmanconj(\mu_*; \hat \mu_n) 
+	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
+}
+Thanks to the linearity of the trace, the expectation becomes
+\alignn{
+	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
+	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
+	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
+}
+
+When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
+then the mean parameter "ratio" is
+\aligns{
+	\mV = n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top
+}
+but $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
+is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$. 
+We know formulas for the expectation of an inverse Wishart and the log-determinant of a Wishart (thanks to $\log\det$ being a sufficient statistics and $\nabla\logpart$)
+\aligns{
+	\E[\mV] &= n \mI \\
+	\forall n > d+1,\E[\mV^{-1}] &= \frac{\mI}{n - d - 1} 
+	\implies \Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}  \\
+	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
+}
+where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
+Putting it all together, we get the following closed form for the expectation of the KL
  \begin{align}
@@
  \end{align}
-where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
-Using the same upper bound as in the univariate case
+
+The complicated term is the $\psi_d$ function.
+We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
 \begin{align}
 	\psi_d(\half[n]) - d \log(\half[n]) 
-	\leq \sum_{i=0}^{d-1} \log(1 - \frac{i}{n}) - \frac{1}{n-i} \; .
-\end{align}
-We can bound the harmonic sum with typical bounds
+	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
+	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
+	\label{eq:psid_bound}
+\end{align}
+We can bound the harmonic sum thanks to the typical bound
+\aligns{
+	\inv{2n +1}
+	\leq H_n - \log(n) - \gamma \leq 
+	\inv{2n - 1}
+}
+where $\gamma$ is the Euler constant. Then
 \begin{align}
@@
 \end{align}
-which yields 
+Plugging this back into~\eqref{eq:psid_bound} yields
 \begin{align}
@@
 	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
-	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}  \; .
-\end{align}
-From there we have two options. 
-Either we use a concave tangent bound on the logarithms to get
+	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1} 
+\end{align}
+where we emphasized that the sum runs up to $d$.
+Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
 \begin{align}
@@
 \begin{align}
+ 	\boxed{
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	\leq \frac{d(d+1)}{4n} + \frac{d(d+1)^2 + \half[d+1]}{2 n (n - d -1)} \;.
-\end{align}
-Note that this rate is not tight for $d>1$, but it is tight for $d=1$, and we almost recover the previous rate.
-
-Another option, potentially tighter than the tangent bound, is to bound the sum of logarithms with an integral, using the monotonicity of the logarithm $\log(1 - \frac{i}{n}) \leq \log(1- \frac{x}{n}) , \forall x \in [i - 1, i]$
-\begin{align}
-	\sum_{i=1}^{d} \log(1 - \frac{i}{n})
-	&\leq \int_{0}^{d} \log(1- \frac{x}{n}) dx \\
-	& = n \int_{1 - \frac{d}{n}}^{1} \log(y) dy \\
-	& = n [ y (\log(y) - 1) ]_{1 - \frac{d}{n}}^{1} \\
-	& =  - n - (n -d) (\log ( 1 - \frac{d}{n}) - 1) \\
-	& = - (n -d) \log ( 1 - \frac{d}{n}) - d \; .
-\end{align}
-Unfortunately, it is hard to get a tractable and non-vacuous bound for this guy.
-At $n=d$ it is worth $-d$, and then it grows up to $0$.
-and we can bound it using some exponents of $1 - \frac{d}{n}$, but that is not really helpful to get an easy formula.
-Actually looking at \href{https://www.desmos.com/calculator/q4sqk73tld}{graphics}, the first upper bound turns out to be tighter than the second one.
+ 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} \;.
+ 	}
+\end{align}
+where $p = d(d+1)/2$.
+This rate is asymptotically tight.
 

### 2021-10-19 aistats.tex
Add condition for multivariate
Improve definitions for multivariate MLE

@@
 \subsection{Multivariate MLE}
-In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, the mean parameter $\mu_*$ becomes a positive definite covariance matrix of order $d$, with $p = \frac{d(d+1)}{2} $ degrees of freedom.
-Its entropy is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, with gradient negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$, and associated Bregman divergence
+In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, and the mean parameter $\mu_*$ is a symmetric positive definite covariance matrix of order $d$, living in dimension $p = \frac{d(d+1)}{2} $.
+Note that we switched the role of $d$ from $\dim(T(x))=d$ to $\dim(x) =d$ and $\dim(T(x))=p$.
+
+The entropy of $X$ is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, whose gradient is the negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$.
+The associated Bregman divergence is
 \alignn{
@@
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} \;.
+ 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
+ 	\ , \forall n >d+1 \;.
  	}

### 2021-10-19 aistats.tex
Formalizer  theorem for multivariate MLE

@@
 \begin{align}
-    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \\
+    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \label{eq:mirror-map} \\
     \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
@@
 	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
-	Its expected suboptimality is infinite when $n\leq 2$ and otherwise
+	The expected KL between $\mu_*$ and $\hat\mu_n$	is infinite when $n\leq 2$ and otherwise upper bounded by
 	\begin{align}
@@
 
+\begin{theorem}[Multivariate MLE Upper Bound]
+	The MLE of  the covariance matrix of $X_i\sim\cN(0,\mu_*)$ is $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top \in\real^{d\times d}$ with $p = \frac{d(d+1)}{2} $ degrees of freedom. 
+	The expected KL between $\mu_*$ and $\hat\mu_n$ is infinite when $n\leq d+1$ and otherwise upper bounded by
+	\alignn{
+	 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
+ 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
+ 	\ , \forall n >d+1 \;.
+}
+This bound is asymptotically tight.
+\end{theorem}
+\begin{proof}
 The entropy of $X$ is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, whose gradient is the negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$.
@@
 is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$.
-We know formulas for the expectation of an inverse Wishart and the log-determinant of a Wishart (thanks to $\log\det$ being a sufficient statistics and $\nabla\logpart$)
-\aligns{
-	\E[\mV] &= n \mI \\
-	\forall n > d+1,\E[\mV^{-1}] &= \frac{\mI}{n - d - 1} 
-	\implies \Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}  \\
+Recall that $\E[\mV] = n \mI$.
+Thanks to $\log\det$ being a sufficient statistic of the Wishart, and the natural to mean parameter formula~\eqref{eq:mirror-map}, we have a closed form for the expected log-determinant of a Wishart 
+\aligns{
 	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
@@
 where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
+The expectation of an inverse Wishart  is straightforward to compute from the density and the log-partition function
+\begin{equation}
+	\E[\mV^{-1}] = \begin{cases}
+		+\infty \text{ if } n\leq d+1 \\
+		 \frac{\mI}{n - d - 1} \text{ otherwise,} 
+	\end{cases}
+\end{equation}
+which proves the infinite part of the statement.
+We now consider only $n>d+1$.
+We deduce that $\Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}$.
 Putting it all together, we get the following closed form for the expectation of the KL
@@
 \begin{align}
- 	\boxed{
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
@@
  	\ , \forall n >d+1 \;.
- 	}
 \end{align}
 where $p = d(d+1)/2$.
-This rate is asymptotically tight.
+\end{proof}
 

### 2021-10-19 aistats.tex
Polish lemma on expected MAP theta

@@
 
+\subsection{Bounds on the Expected MAP Natural Parameter}
+Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
+We introduce the notation $(z)_+ = \max(0,z)$.
+\begin{lemma}[Expected MAP natural parameter]
+	Define $a=n_0 \frac{\mu_0}{\mu^*} $ a variable characterizing the importance of the prior relative to the true parameters.
+	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+\begin{equation}
+		\frac{n_0+n}{a+n}
+		=\frac{\mu^*}{\mu_n}
+		\leq \expect{\frac{\mu^*}{\MAPm}} 
+		= \expect{\frac{\MAPt}{\nat^*}} 
+		\leq
+		\frac{n_0 +n}{a+(n-2)_+}
+		\label{eq:lemma_nat_bound}
+	\end{equation}
+\end{lemma}
+
+\begin{proof}
+	The lower bound can be readily obtained with the Jensen inequality applied on the convex function $x\mapsto\inv{x}$ for $x>0$.
+	The upper bound requires more work.
+	To start, let us plug in the definition of $\MAPm$
+\alignn{
+		\expect{\frac{\mu^*}{\MAPm}} 
+		= \expect{\frac{\MAPt}{\nat^*}} 
+		&= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  \\
+		&= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
+		&= (n_0 +n) \expect{\inv{a + \chi^2(n)}}
+}
+	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$. The expectation reads
+\alignn{
+ 	\expect{\inv{a + \chi^2(n)}}  
+	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx \\
+}
+	We reformulate this expectation with the trick
+\alignn{
+		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
+		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
+		\implies & \boxed{\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,}
+} 
+	to get two integrals instead of one
+\alignn{
+		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
+		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
+		&= \int_0^\infty dx\  x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt\  e^{-\half[(a+x)t]}  \\
+		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
+}
+	where we used Fubini to switch integrals, without further justification.
+	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
+	\begin{align}
+		\expect{\inv{a + \chi^2(n)}}  
+		&= \frac{e^{\half[a]} }{2^{\half[n]+1}\Gamma(\half[n]) }\int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty 2\frac{dy}{t} \  (2\frac{y}{t})^{\frac{n}{2}-1} e^{-y} \\
+		&= \frac{e^{\half[a]} }{2\Gamma(\half[n]) } 
+		\underbrace{\int_1^\infty  \frac{e^{-\half[a t]}}{t^{\frac{n}{2}}}  dt}_{E_{\half[n]}(\half[a])  } \ 
+		\underbrace{\int_0^\infty  y^{\frac{n}{2}-1} e^{-y} dy}_{\Gamma(\half[n])}
+	\end{align}
+	so we finally get  the formula valid for all $n\geq 1$
+	\begin{align}
+		\boxed{\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])}
+		\label{eq:hard_expectation}
+	\end{align}
+	where the {generalized exponential integral function} is defined as 
+	\begin{align}
+		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
+	\end{align}
+	Now our goal is to bound this function with simpler functions.
+	Fortunately, mathematicians have been working on these integrals for decades.
+	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
+	\begin{align}
+		e^x E_k(x) 
+		&\leq \inv{x + k - 1} \\
+		\iff \expect{\inv{a + \chi^2(n)}}
+		&\leq \inv{a + n - 2} \\
+		\iff \expect{\frac{\mu^*}{\MAPm}} 
+		&\leq \frac{n +n_0}{a + n - 2}
+	\end{align}
+We are left with a special case when  $n=1$, eg $k=\half$. 
+Then we can use the trivial bound 
+\begin{align}
+	\expect{\inv{a + X^2}} < \inv{a}
+	\label{eq:trivial_bound}
+\end{align}
+to conclude the proof.
+\end{proof}
+
 \section{PROOFS FOR GAUSSIAN}

### 2021-10-19 main.tex
Add condition for multivariate
Polish lemma on expected MAP theta

@@
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} \;.
+ 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)}
+ 	\ , \forall n >d+1 \;.
  	}
@@
 
-\subsection{Proof of Expected MAP Natural Parameter}
-Let us introduce a variable characterizing the importance of the prior
-\begin{important}
-	\begin{align}
-			a=n_0 \frac{\mu_0}{\mu^*} \; .
-	\end{align}
-\end{important}
+\subsection{Bounds on the Expected MAP Natural Parameter}
+Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
+We introduce the notation $(z)_+ = \max(0,z)$.
 \begin{lemma}[Expected MAP natural parameter]
+	Define $a=n_0 \frac{\mu_0}{\mu^*} $ a variable characterizing the importance of the prior relative to the true parameters.
 	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
 \begin{equation}
-		\frac{\mu^*}{\mu_n}
+		\frac{n_0+n}{a+n}
+		=\frac{\mu^*}{\mu_n}
 		\leq \expect{\frac{\mu^*}{\MAPm}} 
 		= \expect{\frac{\MAPt}{\nat^*}} 
-		\leq \begin{cases}
-		\frac{n_0 +n}{a+n-2} \text{ when } n\geq 2,\\
-		\frac{n_0 +1}{a} \text{ when } n=1.
-		\end{cases}
-		\label{eq:lemma_natp_bound}
+		\leq
+		\frac{n_0 +n}{a+(n-2)_+}
+		\label{eq:lemma_nat_bound}
 	\end{equation}
@@
 	The lower bound can be readily obtained with the Jensen inequality applied on the convex function $x\mapsto\inv{x}$ for $x>0$.
-	The upper bound requires much more work.
+	The upper bound requires more work.
 	To start, let us plug in the definition of $\MAPm$
-	\begin{align}
+\alignn{
 		\expect{\frac{\mu^*}{\MAPm}} 
@@
 		&= (n_0 +n) \expect{\inv{a + \chi^2(n)}}
-	\end{align}
-	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$.
+}
+	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$. The expectation reads
+\alignn{
+ 	\expect{\inv{a + \chi^2(n)}}  
+	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx \\
+}
 	We reformulate this expectation with the trick
-	\begin{align}
+\alignn{
 		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
 		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
-		\implies &\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,
-	\end{align} 
+		\implies & \boxed{\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,}
+} 
 	to get two integrals instead of one
-	\begin{align}
+\alignn{
 		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
@@
 		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
-	\end{align}
+}
 	where we used Fubini to switch integrals, without further justification.
@@
 	Fortunately, mathematicians have been working on these integrals for decades.
-	When $n=2$, eg $k=1$, we have the simple exponential integral function which \href{https://dlmf.nist.gov/6.8.E1}{verifies the bound}
-	\begin{align}
-		\half \log(1 + \frac{2}{x}) 
-		\leq e^x E_1(x) 
-		\leq \log(1+\inv{x}) \leq \inv{x} \; .
-	\end{align}
-	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound}
-	\begin{align}
-		\inv{x+k}
-		\leq e^x E_k(x) 
-		\leq \inv{x + k - 1}
-	\end{align}
-	so that 
-	\begin{align}
-		\inv{a+n}
-		&\leq \expect{\inv{a + \chi^2(n)}}
-		\leq \inv{a + n - 2} \\
-		\iff \frac{n +n_0}{a + n}	
-		&\leq  \expect{\frac{\mu^*}{\MAPm}} 
-		\leq \frac{n +n_0}{a + n - 2}	\\
-		\iff \frac{\mu^*}{\mu_n}
-		&\leq \expect{\frac{\mu^*}{\MAPm}} 
-		\leq \frac{1}{\frac{\mu_n}{\mu^*} - \frac{2}{n + n_0}}  
-	\end{align}
-	where we used $\frac{a + n}{n +n_0} = \frac{\mu_n}{\mu^*}$.
-
+	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
+	\begin{align}
+		e^x E_k(x) 
+		&\leq \inv{x + k - 1} \\
+		\iff \expect{\inv{a + \chi^2(n)}}
+		&\leq \inv{a + n - 2} \\
+		\iff \expect{\frac{\mu^*}{\MAPm}} 
+		&\leq \frac{n +n_0}{a + n - 2}
+	\end{align}
 We are left with a special case when  $n=1$, eg $k=\half$. 
@@
 \end{align}
-which is what we are using in the lemma,
-or we can get a stronger result with further derivations.
-The integral can be written with \href{https://dlmf.nist.gov/7.8.E1}{Mill's ratio} $M(x) = e^{x^2} \int_x^\infty e^{-t^2} dt$, by doing the change of variable $t= x \sqrt{u}$, $dt=\frac{x}{2} \frac{du}{\sqrt{u}}$
+to conclude the proof.
+\end{proof}
+\paragraph{Refinement.}
+In the interesting few sample  case $n=1$, 
+on can get a stronger result with further derivations.
+The integral can be written with \href{https://dlmf.nist.gov/7.8.E1}{Mill's ratio} $M(x) = e^{x^2} \int_x^\infty e^{-t^2} dt$, 
+by doing the change of variable $t= x \sqrt{u}$, $dt=\frac{x}{2} \frac{du}{\sqrt{u}}$ 
 \begin{align}
@@
 \end{align}
-Then using \href{https://dlmf.nist.gov/7.8.E8}{another bound from the DLMF}, we get that 
-\begin{align}
-	\inv{a+1}
-	\leq \expect{\inv{a + X^2}} 
-	< \inv{a}\left( 1 - \inv{a+3} \right) \; .
-\end{align}
-\end{proof}
+Then using \href{https://dlmf.nist.gov/7.8.E8}{another bound} \citep[Eq.~7.8.8]{DLMF} , we obtain
+\begin{align}
+	\expect{\inv{a + X^2}} 
+	< \inv{a}\left( 1 - \inv{a+3} \right) 
+\end{align}
+which can be up to $2/3$ lower than the previous bound if $a$ is small.
 
@@
 There remains the more problematic term with the expectation of the inverse mean parameter.
-Fortunately, we derived the bound \eqref{eq:lemma_natp_bound} for this purpose.
+Fortunately, we derived the bound \eqref{eq:lemma_nat_bound} for this purpose.
 
@@
 
-\textbf{When $n=1$} the bound~\eqref{eq:lemma_natp_bound}  on the expected natural parameter gives
+\textbf{When $n=1$} the bound~\eqref{eq:lemma_nat_bound}  on the expected natural parameter gives
 \begin{align}

### 2021-10-19 aistats.tex
Simplify multivariate bound

@@
 	 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
+ 	\leq \frac{p}{2n} + \frac{p(d+2)}{n (n - d -1)} 
  	\ , \forall n >d+1 \;.
@@
 When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
-then the mean parameter "ratio" is
+then we define the mean parameter "ratio" as
 \aligns{
-	\mV = n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top
-}
-but $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
+	\mV &:= n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top 
+}
+such that $\Tr(\mu_* \hat \mu_n^{-1})  = n \Tr(\mV^{-1})$.
+But $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
 is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$.
@@
 We now consider only $n>d+1$.
-We deduce that $\Tr(\E[\mV^{-1}]) = \frac{d}{n - d - 1}$.
+We deduce that 
+\alignn{
+	\Tr(\expect{\mu_* \hat \mu_n^{-1}} - d
+	= n\Tr(\E[\mV^{-1}]) - d 
+	= \frac{nd}{n - d - 1} -d
+	= \frac{d (d+1}{n -d -1}
+	= \frac{2p}{n -d -1}
+}
 Putting it all together, we get the following closed form for the expectation of the KL
@@
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	= \half \left ( \frac{d(d+1)}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
+ 	= \half \left ( \frac{2p}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
  \end{align}
@@
 	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
-	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1}
+	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)}
 \end{align}
@@
 	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
-	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1} 
+	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)} 
 \end{align}
@@
 	\leq \sum_{i=0}^{d} - \frac{i}{n}
-	 = -\frac{d(d+1)}{2n}
-\end{align}
-which yields the final rate
+	 = -\frac{d(d+1)}{2n} = -\frac{p}{n} \; .
+\end{align}
+Putting it all together yields
 \begin{align}
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	\leq \frac{p}{2n} + \frac{(d+1) \paren{p + \inv{4}}}{n (n - d -1)} 
+ 	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{(d + 1)}{(2n+1)(2(n - d) - 1)} \\
+ 	&\leq p \frac{n + d +1}{2 n (n-d-1)}  + \frac{(d + 1)}{2n.2 (n - d - 1)}\\
+ 	&= \frac{p}{2n} + \frac{2 p(d+1)}{2 n (n-d-1)} + \frac{p/d}{2n (n - d - 1)} \\
+ 	&\leq \frac{p}{2n} + \frac{p(d+2)}{n (n - d -1)} 
  	\ , \forall n >d+1 \;.
 \end{align}
-where $p = d(d+1)/2$.
+where $p = d(d+1)/2$ and the last inequality used $\inv{2d}\leq 1$. 
 \end{proof}

### 2021-10-19 main.tex
Add alternative MAP rate in main.tex

@@
 	\expect{\bregmanconj(\mu^*; \MAPm)} 
-	& = \half \left(\frac{n_0 - a + 2}{a +n - 2}  + \frac{a - n_0}{n_0+n} \right )\\
-	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
-	&\leq \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
+	&\leq \half \left(\frac{n_0 - a + 2}{a +n - 2}  + \frac{a - n_0}{n_0+n} \right )\\
+	&= \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
+	&= \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
 	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
@@
 \end{align}
+Or using a slightly different derivation, we get slightly different bias and variance
+\begin{align}
+	\expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq \half \left(\frac{n_0 - a + 2}{a +n - 2}  + \frac{a - n_0 -2 +2}{n_0+n} \right )\\
+	&= \frac{1}{n_0 + n} + \frac{n_0+ 2 - a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
+	&= \frac{1}{n_0 + n} + \frac{n_0+ 2 - a}{2}  \frac{n_0 + 2 - a}{(a+n-2)(n_0+n)} \\
+	&= \frac{1}{n_0 + n} + \frac{(n_0 + 2 - a)^2}{2(a+n-2)(n_0+n)} \\
+	&= \frac{1}{n_0 + n} + \frac{\half \paren{1 + \frac{2}{n_0} - \frac{\mu_0}{\mu^*}}^2 }{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
+\end{align}
+We will stick with the first one, because we can get a uniform bias over $n=1$ and $n=2$, but this latter one is slightly tighter, and it has the advantage that the $1/n$ term does not depend on $\mu_0$, which is pleasant if we wish to interpret it as a variance term.
 

### 2021-10-19 aistats.tex
Add TODOs in appendix

@@
 These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
-\citep[p.177]{nesterov2003introductory}
+\citep[p.177]{nesterov2003introductory} % example 4.1.1.4
 , but $\logpart$ is not (see App.~\ref{app:gaussian}).
@@
 
-To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focused on the symmetrized Bregman instead
+To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we will focus on the symmetrized Bregman instead
 \aligns{
@@
 		\leq
-		\frac{n_0 +n}{a+(n-2)_+}
+		\frac{n_0 +n}{a+(n-2)_+} \; , \forall n\geq 0 \; .
 		\label{eq:lemma_nat_bound}
@@
 
-\section{PROOFS FOR GAUSSIAN}
+\subsection{Proof of MAP Bound}
+
+\begin{theorem}[MAP Bound]
+The expected symmetrized Bregman of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
+ \begin{align}
+	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
+	\leq \begin{cases}
+		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
+		\inv{2(n_0+1)}  +  b_1 \ \text{ if }\ n=1,\\
+		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{ if }\ n\geq 2
+	\end{cases}\\
+	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
+\end{align}
+\end{theorem}
+
+\begin{proof}
+When $n=0$, the inequality is an equality. 
+Wen $n>0$, Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound},
+\begin{align}
+	2\expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq \expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1 \; .
+\end{align}
+We readily get 
+\begin{align}
+	 \expect{\frac{\MAPm}{\mu^*}} - 1 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
+\end{align}
+where recall that $a=n_0\frac{\mu_0}{\mu^*}$. 
+There remains the more problematic term with the expectation of the inverse mean parameter.
+Fortunately, we derived the bound \eqref{eq:lemma_nat_bound} for this purpose.
+
+\textbf{When $n\geq 2$}, we get 
+\begin{align}
+	\expect{\frac{\mu^*}{\MAPm}} - 1 
+	\leq \frac{n_0 + n}{a +n - 2} -1
+	= \frac{n_0 - a + 2}{a +n - 2}
+	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
+\end{align}
+so putting it all together we get 
+\begin{align}
+	\expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq \half \left(\frac{n_0 - a + 2}{a +n - 2}  + \frac{a - n_0}{n_0+n} \right )\\
+	&= \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
+	&= \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
+	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
+	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
+	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
+\end{align}
+
+\textbf{When $n=1$} the bound~\eqref{eq:lemma_nat_bound}  on the expected natural parameter gives
+\begin{align}
+	\expect{\frac{\mu^*}{\MAPm}} - 1 
+	\leq \frac{n_0 + 1}{a} -1
+	= \frac{n_0 +1 - a}{a}
+\end{align}
+so putting it all together we get
+\begin{align}
+	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
+	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 +1 - a }{a} \\
+	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
+	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
+	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{\frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
+\end{align}
+so we do recover the same bias variance structure as when $n\geq2$.
+\end{proof}
+
+\subsection{Complementary Plots}
+
+\rlp{Show that the optimal mu0 is larger than mu* for small n0.}
+
+\section{COMPLEMENTS ON GAUSSIANS}
 \label{app:gaussian}
+
+\rlp{prove that $A$ is not self -concordant, while, $A^*$ is thanks to Nesterov.}
+Maybe plot both of them.
+Also plot the mirror maps.
+And plot gaussian trajectories.
 
@@
 
+\rlp{simple stuff}
+
 \section{SELF-CONCORDANCE}
 \label{app:self-concordant}
+\TODO{Damien : review self-concordance + prove proposition 1 + graph to explain $x<0.21$.}
+
+\rlp{Add connection duality / self-concordant}
 
@@
 
+\rlp{prove that the bias decreases like $1/n^2$ for gaussian variance and gaussian mean.}
+
+\rlp{prove that the expectation of Hanzely is the symmetrized bias, e.g.  an upper bound on the bias.}
+
 \section{REVIEW OF SMD}
 \label{app:SMD}
+
+\rlp{Introduce SMD, in primal and dual formulations.}
 

### 2021-10-20 aistats.tex
Polish proof of MAP bound

@@
 	\KL( \sigma_*^2 ; \sigma_n^2 ) 
-	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right) \\
+	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right)
+	= \phi\paren{\frac{\sigma_*^2}{ \sigma_n^2}} \\
 	\bregmanconj( \mu_*; \mu_n) 
-	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) \\
+	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) 
+	= \phi\paren{\frac{\mu_*}{ \mu_n}} \\
 	\bregman( \nat_n; \nat_* ) 
-	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
-\end{align}
-Intuitively, the divergence measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
+	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right)
+	= \phi\paren{\frac{ \nat_n}{\nat_*}}
+\end{align}
+where all these lines are equal, and $\phi$ measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
 \begin{align}
@@
 \end{align}
-illustrated in Figure~\ref{fig:phi}.
+illustrated in orange in Figure~\ref{fig:phi}.
+To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we will focus on the symmetrized Bregman instead
+\alignn{
+	\bregmanconj( \mu_*; \mu_n) 
+	\leq \cS_{\conj} (\mu_*, \mu_n ) 
+	:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) 
+	= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 }
+	\label{eq:symmetrized_bregman}
+}
+which can also be expressed with $\phi$ as follows:
+\aligns{
+	\label{eq:log_bound}
+	\phi(z) \leq \phi(z) + \phi(\inv{z}) 
+	&=  \half \paren{z -1  + \inv{z} -1} 
+	= \frac{(z-1)^2}{z} \; .
+}
+See the grey upper bound in Figure~\ref{fig:phi}.
+
 \begin{figure}[ht]
@@
 \end{figure}
-
-To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we will focus on the symmetrized Bregman instead
-\aligns{
-	\label{eq:log_bound}
-	\phi(z) \leq \phi(z) + \phi(\inv{z}) 
-	&=  \half \paren{z -1  + \inv{z} -1} 
-	= \frac{(z-1)^2}{z}
-}
-\aligns{
-	\bregmanconj( \mu_*; \mu_n) 
-	&\leq \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \\ 
-	=: \cS_{\conj} (\mu_*, \mu_n )  
-	&= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 }
-}
 
@@
 
+We did not find a closed form or an upper bound for the expected logarithm of the MAP $\expect{\log\frac{\MAPm}{\mu^*}}$.
+Consequently, we derived a bound for the symmetrized Bregman~\eqref{eq:symmetrized_bregman} instead. This bound is not asymptotically  tight for the Bregman by a factor $2$.
+
 \begin{theorem}[MAP Bound]
-The expected symmetrized Bregman of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
- \begin{align}
-	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
+The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
+ \begin{align}
+	& \expect{\cS_\conj( \m^*; \hat \m_n^\mathrm{MAP})}
 	\leq \begin{cases}
-		\bregmanconj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
+		\cS_\conj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
 		\inv{2(n_0+1)}  +  b_1 \ \text{ if }\ n=1,\\
@@
 When $n=0$, the inequality is an equality. 
-Wen $n>0$, Wolfram finds no closed form for $\expect{\log\frac{\MAPm}{\mu^*}}$ so we focus on the simple log-bound given by \eqref{eq:log_bound},
-\begin{align}
-	2\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1 \; .
-\end{align}
-We readily get 
-\begin{align}
-	 \expect{\frac{\MAPm}{\mu^*}} - 1 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
-\end{align}
-where recall that $a=n_0\frac{\mu_0}{\mu^*}$. 
+For $n>0$, we expand the symmetrized Bregman~\eqref{eq:symmetrized_bregman} to get
+\begin{align}
+	\expect{\cS_\conj(\mu^*; \MAPm)} 
+	&\leq \half \paren{\expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1} \; .
+\end{align}
+The expectation of $\MAPm$ is straightforward
+\begin{align}
+	 \expect{\frac{\MAPm}{\mu^*}} - 1 
+	 = \frac{n_0 \mu_0 + n \mu^*}{(n_0+n)\mu^*} - 1 
+	 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
+	 \quadtext{where}
+	 a:=n_0\frac{\mu_0}{\mu^*} \; .
+\end{align}
 There remains the more problematic term with the expectation of the inverse mean parameter.
@@
 	\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \half \left(\frac{n_0 - a + 2}{a +n - 2}  + \frac{a - n_0}{n_0+n} \right )\\
-	&= \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
+	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
 	&= \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
@@
 \end{align}
-so we do recover the same bias variance structure as when $n\geq2$.
+so we  recover the same bias term as when $n\geq2$.
 \end{proof}

### 2021-10-20 aistats.tex
Add new figure for the intro after submission

@@
 
-%In particular, no rates hold in the few samples regime  -- e.g. after seeing 5 samples from a gaussian, we do not know how many bits away from the true distribution we should expect our model to be.
-%No rates hold either when the loss is ill-behaved : infinite domain, not smooth, not self-concordant
+% TODO
+%\begin{figure}[t]
+%	\centering
+%\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
+%	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
+%		Solid curve  are average over $10^5$ trials.
+%		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
+%		Shaded areas are 90\% confidence interval.
+%		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
+%		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
+%		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
+%	}
+%	\label{fig:curves}
+%\end{figure}
 
@@
 With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
-We thus consider the fundamental problem:
+We thus consider the fundamental problem: %TODO thus
 %\begin{center}
@@
 or close to quadratic (\S\ref{ssec:quadratic}).
-However, a  general solution for finite $n$ remains elusive.
+However, a  general solution for finite $n$ remains elusive. %TODO add link to figure 1
 In this paper, we review partial solutions and give ideas on how to solve the problem.
@@
 These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
-\citep[p.177]{nesterov2003introductory} % example 4.1.1.4
+\citep[p.177]{nesterov2003introductory} % TODO example 4.1.1.4
 , but $\logpart$ is not (see App.~\ref{app:gaussian}).

### 2021-10-20 aistats.tex
Add section on optimal n0

@@
 
-\subsection{Complementary Plots}
-
-\rlp{Show that the optimal mu0 is larger than mu* for small n0.}
+\subsection{Optimal $\mu_0$ is larger than $\mu^*$ for small $n_0$ and small $n$}
+
+The upper bound~\eqref{eq:MAP_rate} has a bias term that is $0$ when $\frac{mu_0}{\mu^*} = 1 + \inv{n_0}$, e.g.
+for large values of $n_0$, it is $\mu_0=\mu^*$ is the best prior, but for small $n_0$, one better sets larger values for $\mu_0$.
+In Figure~\ref{fig:optimal_n0}, we observe this behavior numerically.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=\textwidth]{figs/optimal_n0.pdf}
+	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $\mu_0, n_0$ spanning $[10^{-2}, 10^2]$.
+We observe that the line $\frac{\mu_0}{\mu^*} = 1 + \inv{n_0}$ coincides with the bottom valley of this landscape.
+	}
+	\label{fig:optimal_n0}
+\end{figure}
 

### 2021-10-20 aistats.tex
Prove gaussian logpart is not self-concordant

@@
 \alignn{
-	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\
+	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\  % TODO remove factor 2
 	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
@@
 These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
-\citep[p.177]{nesterov2003introductory} % TODO example 4.1.1.4
+\citep[p.177]{nesterov2003introductory} % TODO add citation to example 4.1.1.4
 , but $\logpart$ is not (see App.~\ref{app:gaussian}).
@@
 
+In this section, we prove all the results mentioned in \S\ref{ssec:gaussian-variance}, and add some context and experimental observations.
+
 Let us recall the form of the divergence with different parametrizations
@@
 	\includegraphics[width=\textwidth]{figs/optimal_n0.pdf}
-	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $\mu_0, n_0$ spanning $[10^{-2}, 10^2]$.
-We observe that the line $\frac{\mu_0}{\mu^*} = 1 + \inv{n_0}$ coincides with the bottom valley of this landscape.
+	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $n_0, \mu_0$ spanning $[10^{-2}, 10^2]$.
+The expectation was estimated with $10^4$ draws for each value of $n_0, \mu_0$ We observe that the line $\frac{\mu_0}{\mu^*} = 1 + \inv{n_0}$ coincides with the bottom valley of this landscape.
 	}
@@
 
-\rlp{prove that $A$ is not self -concordant, while, $A^*$ is thanks to Nesterov.}
-Maybe plot both of them.
-Also plot the mirror maps.
-And plot gaussian trajectories.
+In this section, we prove that for a Gaussian, the entropy is self-concordant while the log-partition function is not.
+We also give complementary illustrations on the Gaussian mirror map and MLE.
+
+The Gaussian log-partition function and entropy are up to constants
+\alignn{
+	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
+	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; ,
+}
+where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
+These functions and their domains they are represented in Figure \TODO{}.
+
+\begin{definition}[Self-concordance]
+A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}}$
+\end{definition}
+
+\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
+A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(x)} \leq C \norm{v}_2 g''(x)$ for some $C\geq 0$
+\end{definition}
+
+\paragraph{Entropy is Self-Concordant,}
+\citet[p.177]{nesterov2003introductory} proves that logarithmic barriers for a second order region are self-concordant.
+Adopting their notation, $\conj$ fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = \begin{pmatrix} 1 \\ 0 \end{pmatrix}$.
+
+\paragraph{Log-partition is not Self-Concordant}
+Let us fix a straight line with $\nat_1 \neq 0$, e.g. $g(\nat_2) := A(\nat_1,\nat_2)$. Then
+\alignn{
+g'(\nat_2) = \frac{\partial A}{\partial\nat_2}(\nat) 
+&= +\frac{\nat_1^2}{4 \nat_2^2} + \frac{1}{2\nat_2} \\
+g''(\nat_2) = \frac{\partial^2 A}{(\partial\nat_2)^2}(\nat) 
+&= -\frac{\nat_1^2}{2 \nat_2^3} - \frac{1}{2\nat_2^2} 
+=  \frac{1}{2\nat_2^2}\paren{\frac{\nat_1^2}{-\nat_2^2} - 1} \\ 
+g'''(\nat_2) = \frac{\partial^3 A}{(\partial\nat_2)^3}(\nat) 
+&= +\frac{3\nat_1^2}{2 \nat_2^4} + \frac{1}{\nat_2^3}
+= \frac{1}{2\nat_2^3}\paren{2 - 3 \frac{\nat_1^2}{- \nat_2^2}}
+}
+Setting $\nat_2 = - \nat_1^2$ yields $g''(\nat_2) = 0$ and $g'''(\nat_2) \neq 0$, meaning the second derivative cannot dominate the third derivative for any choice of constant.
+The log-partition is neither self concordant nor pseudo-self-concordant.
 

### 2021-10-20 aistats.tex
Add asymptotic derivation

@@
 
-\subsection{Optimal $\mu_0$ is larger than $\mu^*$ for small $n_0$ and small $n$}
-
-The upper bound~\eqref{eq:MAP_rate} has a bias term that is $0$ when $\frac{mu_0}{\mu^*} = 1 + \inv{n_0}$, e.g.
+\subsection{On the Choice of a Prior}
+The optimal $\mu_0$ is larger than $\mu^*$ for small $n_0$ and small $n$.
+Indeed, the upper bound~\eqref{eq:MAP_rate} has a bias term that is $0$ when $\frac{mu_0}{\mu^*} = 1 + \inv{n_0}$, e.g.
 for large values of $n_0$, it is $\mu_0=\mu^*$ is the best prior, but for small $n_0$, one better sets larger values for $\mu_0$.
@@
 \label{app:asymptote}
-
-\rlp{simple stuff}
+In this section we fill-in the lines of \S\ref{ssec:asymptote} to prove Equation~\eqref{eq:asymptote}.
+Approximating $\conj$ with a second order Taylor expansion yields
+\aligns{
+    \bregmanconj(\m^* ; \m)
+    &= \frac{1}{2}\norm{\m - \m^*}^2_{\mF}
+    + O(\norm{\m - \m^*}^3),
+}
+where the norm is induced by the matrix
+\aligns{
+    \mF
+    :=\nabla^2\conj(\m^*)
+    = \nabla^2\logpart(\nat^*)^{-1}
+    = \Cov_{\nat^*}(T)^{-1} 
+}
+where the second equality is a general property of convex conjugates.
+Plugging the MLE~\eqref{eq:defMLE} into this quadratic and expanding it yields
+\aligns{
+	\E \half \norm{\inv{n}  \sum_i T_i - \m^* }_\mF^2 
+	&=\inv{2n^2} \sum_i \E \norm{T_i - \m^* }_\mF^2 
+	+ \inv{2n^2} \sum_{i\neq j}\expect{T_i - \m^*}^\top \mF \overbrace{\expect{T_j - \m^*}}^{0} ]\\
+	&=\inv{2n}  \E \norm{T_1 - \m^* }_\mF^2 \\
+	& = \inv{2n} \Tr(\mF\ \expect{(T_1 - \m^*) (T_1 - \m^*)^\top}) \\
+	&= \inv{2 n} \Tr(\mF \Cov_{\nat^*}(T)) = \frac{d}{2n} 
+}
+where on the first line we used independence of samples, on the second line we used the fact that samples are identically distributed, and on the third line we used the  trace trick along with the linearity of the trace $\Tr$.
+On the way, this also proves that for the MLE $\norm{\m - \m^*}^3 \in O(n^{-\half[3]})$.
+This yield the final rate for the MLE
+\begin{align}
+	\E \bregmanconj \paren{\E [T(X)] ; \inv{n}  \sum_i T_i}
+	= \frac{d}{2n} + O(n^{- \frac{3}{2}} ) \; .
+\end{align}
+For the MAP~\eqref{eq:defMAP}, the quadratic decomposes into bias and variance
+\begin{align}
+	\E \half \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
+	&= \frac{n d}{2(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \half \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
+	&= \frac{d}{2n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
+\end{align}
+This $O(n^{-2})$ term is dominated by the $O(n^{-\half[3]})$ term from the quadratic approximation of the Bregman, yielding the same first order rate as for the MLE
+\begin{align}
+	\E \bregmanconj \paren{\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }
+	= \frac{d}{2n} + O(n^{- \frac{3}{2}} ) \; .
+\end{align}
 

### 2021-10-20 aistats.tex
Move self-concordance definition

@@
 
-The Gaussian log-partition function and entropy are up to constants
+The Gaussian log-partition function and entropy are, up to constants,
 \alignn{
@@
 These functions and their domains they are represented in Figure \TODO{}.
-
-\begin{definition}[Self-concordance]
-A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}}$
-\end{definition}
-
-\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
-A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(x)} \leq C \norm{v}_2 g''(x)$ for some $C\geq 0$
-\end{definition}
+We provide definitions of self-concordance in \cref{app:self-concordant}.
 
@@
 }
-Setting $\nat_2 = - \nat_1^2$ yields $g''(\nat_2) = 0$ and $g'''(\nat_2) \neq 0$, meaning the second derivative cannot dominate the third derivative for any choice of constant.
+Setting $\nat_2 = - \nat_1^2$ yields $g''(\nat_2) = 0$ and $g'''(\nat_2) \neq 0$, meaning the second derivative cannot dominate the third derivative for any choice of constant or exponent.
 The log-partition is neither self concordant nor pseudo-self-concordant.
@@
 \rlp{Add connection duality / self-concordant}
+
+\begin{definition}[Self-concordance]
+A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}}$
+\end{definition}
+
+\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
+A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(x)} \leq C \norm{v}_2 g''(x)$ for some $C\geq 0$
+\end{definition}
 

### 2021-10-20 aistats.tex
Add connection self-concordance - duality

@@
 \label{app:self-concordant}
+
+In this section, we define self-concordance, we prove proposition~\ref{prop:selfConcordant}, and we prove that a 1 dimensional function is self-concordant if and only if its convex conjugate is self-concordant as well.
+
 \TODO{Damien : review self-concordance + prove proposition 1 + graph to explain $x<0.21$.}
-
-\rlp{Add connection duality / self-concordant}
 
@@
 \end{definition}
+
+\begin{proposition}
+	\label{prop:conjugateConcordance}
+	Let $f:\real \rightarrow \real$ be a self-concordant function of Legendre type. Then its convex conjugate $f^*$ is also self-concordant.
+\end{proposition}
+\begin{proof}
+When $f$ is of Legendre type then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ has inverse gadients
+\begin{align}
+    \nabla f \circ \nabla f^*  &= \Id \\
+     \nabla f^* \circ \nabla f &= \Id
+\end{align}
+where $\Id$ is the identity function on the relevant domain. 
+Deriving this equality proves that some Hessians are inverse of each other 
+\begin{align}
+    \nabla^2 f(x) \nabla^2 f^*(x^*) = \mI
+    \label{eq:conjugateHessian}
+\end{align}
+where $(x,x^*)$ are conjugate points -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
+\TODO{Why are Hessians not 0?}
+In 1 dimension, this inequality is more simply written 
+\aligns{f''(x) f^{* \prime \prime}(x^*) = 1}
+Until now, this should be familiar to the reader.
+The interesting surprise is that self-concordance appears when we derive again this equality. 
+\begin{align}
+    &1= f''(x) f^{* \prime \prime}(f'(x)), \forall x \\
+    \implies
+    &0 = f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))\\
+    \iff
+    &0 = f'''(x)f^{* \prime \prime}(x^*)  + f''(x)^2 f^{* \prime \prime \prime}(x^*)\\
+    \implies
+    &0 = \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  f^{* \prime \prime}(x^*)^{-\half}
+    + f''(x)^{\half} \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}} \\
+    \implies
+    &0= \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}}
+\end{align}
+where to reach the last line, we used~\eqref{eq:conjugateHessian} again.
+We see that for a pair of conjugate functions, the self-concordance ratios are opposite of each other, which concludes the proof.
+\end{proof}
 

### 2021-10-21 aistats.tex
wrote proof of prop 1, and added figures to plot omega

@@
 
+\subsection{Properties of Self-concordant functions}
+
+We quickly reviews some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}.
+We start with a few notations.
+Let $A^*$ be a self-concordant function.
+Then, we write
+\begin{itemize}
+	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
+	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.
+\end{itemize}
+Note that $\omega^*(t)$ is positive, convex and monotonically increasing for $t\in[0,1]$, and that $\omega^*(t) \leq t^2, \forall t\leq 0.68$.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega.eps}
+	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega_star.eps}
+	\caption{(left) Graph of the distance function $\omega$ and its dual $\omega^*$.
+(right) Graph of the dual of distance function $\omega^*$ evaluated at $\frac{t}{1-t}$, compared with $t^2$.
+The two curves cross each other at $t\approx 0.21$.}
+	\label{fig:omega}
+\end{figure}
+
+We now present two important results. The first one shows how to convert the local norm $\|y-x\|_y$ using $\|y-x\|_x$.
+\begin{proposition}\label{prop:conversion_norm}
+	(Conversion of norms, \citep[Theorem 4.1.5]{nesterov2003introductory}) For any $x,\,y\in\dom A^*$, if $\|y-x\|_x<1$, then
+	\[
+		\|y-x\|_y \leq \frac{\|y-x\|_x}{1-\|y-x\|_x}.
+	\]
+\end{proposition}
+
+The next result show that, if $y$ is sufficiently close to $x$, then we can bound the Bregman divergence of $A^*$ using the distance function $\omega^*$ and local norms.
+\begin{proposition}(Upper bound of self-concordant functions \citep[Theorem 4.1.8]{nesterov2003introductory}) \label{prop:upper_bound_self_concordance}
+	For any $x,\,y\in\dom A^*$, if $\|y-x\|_x<1$, then
+	\[
+		\bregmanconj(y,x) \leq \omega^*(\|y-x\|_x).
+	\]
+\end{proposition}
+
+We are now ready to prove Proposition~\ref{prop:selfConcordant}.
+
+\subsection{Proof of Proposition~\ref{prop:selfConcordant}.}
+
+We start with Proposition~\ref{prop:upper_bound_self_concordance}, evaluated at $y=\meanp^*$ and $x=\meanp$:
+\[
+	\bregmanconj(\meanp^*,\meanp) \leq \omega^*(\|\meanp^*-\meanp\|_{\meanp}).
+\]
+This hold if $\|\meanp^*-\meanp\|_{\meanp}<1$.
+Since $\omega^*$ is monotonically increasing, we can replace $\|\meanp^*-\meanp\|_{\meanp}$ by its upper bound from Proposition~\ref{prop:conversion_norm},
+\[
+	\omega^*(\|\meanp^*-\meanp\|_{\meanp}) \leq \omega^*\left(\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}}\right),
+\]
+under the conditions that $\|\meanp^*-\meanp\|_{\meanp^*}<1$ (to satisfy the assumption of Proposition~\ref{prop:conversion_norm}) and $\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}} <1$ (to ensure that $\|\meanp^*-\meanp\|_{\meanp} <1$).
+Those two conditions holds if $\|\meanp^*-\meanp\|_{\meanp^*} <0.5$.
+Now, we use the bound (see figure \ref{fig:omega})
+\[
+	\omega^*\left(\frac{t}{1-t}\right) \leq t^2, \quad 0\leq t \leq 0.21,
+\]
+and replace $t$ by $\|\meanp^*-\meanp\|_{\meanp^*}$. This finally gives the sequence of inequalities
+\[
+	\bregmanconj(\meanp^*,\meanp) \leq \omega^*(\|\meanp^*-\meanp\|_{\meanp}) \leq \omega^*\left(\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}}\right) \leq \|\meanp^*-\meanp\|_{\meanp^*}^2,
+\]
+that holds while $ \|\meanp^*-\meanp\|_{\meanp^*}<0.21$, which is the desired result.
+
+%\begin{proposition}
+%	Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
+%	\aligns{
+%		\norm{\m^*-\m}_{\mF} < 0.21
+%		\implies
+%		\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
+%	}
+%\end{proposition}
+%\begin{proof}
+%	
+%\end{proof}
+
 \section{BIAS-VARIANCE}

### 2021-10-21 aistats.tex
minor changes
Proofreading appendix A, notes on B+C

--- 2021-10-20 aistats.tex
@@
 
-In this section, we prove all the results mentioned in \S\ref{ssec:gaussian-variance}, and add some context and experimental observations.
-
-Let us recall the form of the divergence with different parametrizations
+In this section, we prove the results mentioned in \S\ref{ssec:gaussian-variance},
+and add some context and experimental observations.
+Let us recall the form of the divergence with different parametrizations.
+For a Gaussian with known mean $0$ and covariance $\sigma^2$, 
+natural and mean parameters $\theta, \m$;
 \begin{align}
@@
 	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right)
-	= \phi\paren{\frac{\sigma_*^2}{ \sigma_n^2}} \\
-	\bregmanconj( \mu_*; \mu_n) 
-	&= \half \left ( \frac{\mu_*}{ \mu_n} - 1 - \log  \frac{\mu_*}{ \mu_n} \right) 
-	= \phi\paren{\frac{\mu_*}{ \mu_n}} \\
+	= \phi\paren{\frac{\sigma_*^2}{ \sigma_n^2}},
+	\\
+	\bregmanconj(\m_*; \m_n) 
+	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) 
+	= \phi\paren{\frac{\m_*}{ \m_n}},
+	\\
 	\bregman( \nat_n; \nat_* ) 
 	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right)
-	= \phi\paren{\frac{ \nat_n}{\nat_*}}
+	= \phi\paren{\frac{ \nat_n}{\nat_*}},
 \end{align}
@@
 \begin{align}
-	\phi(z) := \half (z - 1 - \log(z))
+	\phi(z) := \half (z - 1 - \log(z)),
 \end{align}
 illustrated in orange in Figure~\ref{fig:phi}.
-To get an upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we will focus on the symmetrized Bregman instead
+To upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focus on the symmetrized Bregman instead
 \alignn{
@@
 	:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) 
-	= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 }
+	= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 },
 	\label{eq:symmetrized_bregman}
@@
 	\label{eq:log_bound}
-	\phi(z) \leq \phi(z) + \phi(\inv{z}) 
+	\phi(z) \leq \phi(z) + \phi\paren{\inv{z}}
 	&=  \half \paren{z -1  + \inv{z} -1} 
@@
 	\includegraphics[width=.4\textwidth]{phi.pdf}
-	\caption{$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. It is a barrier near $0$. As a result, it admits another upper bound (in grey) despite being poorly approximated by quadratics.}
+	\caption{%
+	$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. 
+	It is a barrier near $0$, and admits an upper bound (in grey) despite being poorly approximated by quadratics.
+	% fdk: Text said that it admits an the grey upper bound because it is a barrier. Don't follow the logical implication.
+	}
 	\label{fig:phi}
@@
 
-\subsection{Proof of MLE Upper Bound}
+\subsection{Upper bound for the MLE of the variance}
 
@@
 	\begin{align}
-		 \expect{\bregmanconj( \mu_*; \mu_n) }
+		 \expect{\bregmanconj( \mu_*; \hat \mu_n) }
 			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
@@
 	\begin{proof}
-	The ratio to the optimum  $\frac{n \mu_n}{\mu_*}$follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom $\chi^2(n)$.
-Its inverse $\frac{\mu_*}{ \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
-	\begin{align}
-		\expect{n \frac{\mu_*}{n \mu_n}  - 1} 
+	As $\hat \mu_n = \inv{n} \sum_i X_i^2$ 
+	with $X_i \sim \cN(0, \mu_*)$, the ratio to the optimum $n \frac{\hat \mu_n}{\mu_*}$
+	follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom, $\chi^2(n)$.
+Its inverse $\frac{\mu_*}{\hat \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
+	\begin{align}
+		\expect[\hat \mu_n]{\frac{\mu_*}{\hat \mu_n}  - 1}  
+		= \expect[\mu_n]{n \frac{\mu_*}{n \hat \mu_n}  - 1} 
 		= \expect{\frac{n}{\chi^2(n)} - 1} =
@@
 	\end{align}
-	There is also a closed form solution for the expected logarithm of a Chi-squared, because a Chi-squared is a special case of Gamma distributions, whose sufficient statistic is $(X,\log(X))$.
-For non-central chi-squares, one could also look at \citep{pav2015moments}.
-	\begin{align}
-		\expect{\log \frac{\mu_n}{\mu_*}} = \psi(\half[n]) - \log(\half[n])
+	There is also a closed form solution for the expected logarithm of a Chi-squared, 
+	because a Chi-squared is a special case of Gamma distributions, whose sufficient statistic is $(X, \log(X))$,
+	\fdk{I don't follow why this is ``because it is a Gamma with sufficient statistics $(X,\log(X))$''?}
+	\begin{align}
+		\expect{\log \frac{\hat \mu_n}{\mu_*}} = \psi\paren{\half[n]} - \log\paren{\half[n]}
 	\end{align}
 	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
+	Closed forms also exist for non-central chi-squares, see for example the note of \citep{pav2015moments}.
 	Consequently the suboptimality of the MLE has a closed form solution
 	\begin{align}
-	\expect{\bregmanconj( \mu_*; \mu_n) }
-	&= \half \expect{\frac{\mu_*}{ \mu_n} - 1 + \log \left(\frac{\mu_n}{\mu_*} \right) } \\
+	\expect{\bregmanconj( \mu_*; \hat \mu_n) }
+	&= \half \expect{\frac{\mu_*}{ \hat \mu_n} - 1 + \log \left(\frac{\hat \mu_n}{\mu_*} \right) },
+	\\
 	& =
@@
 	\end{align}
-	It is surprising that we need more than $3$ samples for the loss to have a bounded expectation. 
-	It is also very pleasant that the optimal value $\mu_*$ does not appear in this rate, so that the convergence rate is independent of the actual solution.
-	When the expectation is finite, we can make its formula intelligible thanks to \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function}
-	\begin{align}
-		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; .
-	\end{align}
-	  For $n\geq 3$,
+	Surprisingly, we need $3$ samples or more for the expected loss to be bounded and
+	the optimal value $\mu_*$ does not appear in this rate, 
+	so that the convergence rate is independent of the solution.
+	When the expectation is finite, we can get a more interpretable formula using known \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function},
+	\begin{align}
+		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; ,
+	\end{align}
+	giving, for $n\geq 3$,
 	\begin{align}
 		\inv{n-2} - \inv{n}
-		&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
-		\leq \inv{n-2} - \inv{2 n} \\
+		&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
+		\leq \inv{n-2} - \inv{2 n},
+		\\
 		\iff
 			\frac{2}{n(n-2)}
-			&\leq \expect{\bregmanconj( \mu_*; \mu_n) }
-			\leq \inv{2n} +\frac{2}{n(n-2)}
-	\end{align}
-	so we get a $O(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
+			&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
+			\leq \inv{2n} +\frac{2}{n(n-2)},
+	\end{align}
+	so we get a $\Omega(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
 	\end{proof}
@@
 \subsection{Multivariate MLE}
-In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, and the mean parameter $\mu_*$ is a symmetric positive definite covariance matrix of order $d$, living in dimension $p = \frac{d(d+1)}{2} $.
+In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, and the mean parameter $\mu_*$ is a symmetric positive definite covariance matrix of order $d$, living in dimension $p = \frac{d(d+1)}{2}$.
 Note that we switched the role of $d$ from $\dim(T(x))=d$ to $\dim(x) =d$ and $\dim(T(x))=p$.
+\fdk{This $\dim$ is not introduced? I would say that $\dim(T(X))$ is $d^2$ if $T(X) = XX^\top$. Slightly confused here.}
 
@@
 	The MLE of  the covariance matrix of $X_i\sim\cN(0,\mu_*)$ is $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top \in\real^{d\times d}$ with $p = \frac{d(d+1)}{2} $ degrees of freedom. 
-	The expected KL between $\mu_*$ and $\hat\mu_n$ is infinite when $n\leq d+1$ and otherwise upper bounded by
+	The expected KL divergence between $\mu_*$ and $\hat\mu_n$ 
+	is infinite when $n\leq d+1$ and otherwise upper bounded by
 	\alignn{
@@
 }
-
-When the estimator is the MLE $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
-then we define the mean parameter "ratio" as
+When the estimator is the MLE, $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
+then we define the mean parameter ``ratio'' as
 \aligns{
-	\mV &:= n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top 
+	\mV &:= n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top,
 }
@@
 \aligns{
-	\E[\log\det \mV] &= \psi_d(\half[n]) + d \log 2 \; ,
+	\E[\log\det \mV] &= \psi_d\paren{\half[n]} + d \log 2 \; ,
 }
@@
 which proves the infinite part of the statement.
-We now consider only $n>d+1$.
-We deduce that 
-\alignn{
-	\Tr(\expect{\mu_* \hat \mu_n^{-1}} - d
+Consider now the case $n>d+1$. Using
+\alignn{
+	\Tr(\expect{\mu_* \hat \mu_n^{-1}}) - d
 	= n\Tr(\E[\mV^{-1}]) - d 
 	= \frac{nd}{n - d - 1} -d
-	= \frac{d (d+1}{n -d -1}
-	= \frac{2p}{n -d -1}
-}
-Putting it all together, we get the following closed form for the expectation of the KL
- \begin{align}
+	= \frac{d (d+1)}{n -d -1}
+	= \frac{2p}{n -d -1},
+}
+and putting it all together, we get the following closed form for the expectation of the divergence
+ \begin{align}
+ 	\label{eq:where-we-put-it-back-together-multivariate-gaussian}
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	= \half \left ( \frac{2p}{n-d-1} + \psi_d(\half[n]) - d \log(\half[n])  \right )
- \end{align}
-
-The complicated term is the $\psi_d$ function.
-We can use the upper bound $\psi(x) - \log(x) \leq - \inv{2x}$ (like in the univariate case) to get
-\begin{align}
-	\psi_d(\half[n]) - d \log(\half[n]) 
-	&= \sum_{i=0}^{d-1} \paren{\psi ( \half[n - i]) - \log(\half[n]) } \\
-	&\leq \sum_{i=0}^{d-1} \paren{\log(1 - \frac{i}{n}) - \frac{1}{n-i}} \; .
+ 	= \half \left ( \frac{2p}{n-d-1} + \psi_d\paren{\half[n]} - d \log\paren{\half[n]}  \right ).
+ \end{align}
+To bound $\psi_d$, we can use the same bound as in the univariate case, $\psi(x) \leq \log(x) - \inv{2x}$, to get
+\begin{align}
+	\psi_d\paren{\half[n]} - d \log\paren{\half[n]}
+	&= \sum_{i=0}^{d-1} \paren{\psi \paren{\half[n - i]} - \log\paren{\half[n]} },
+	\\
+	&\leq \sum_{i=0}^{d-1} \paren{\log\paren{\half[n-i]} - \inv{n-i} - \log\paren{\half[n]} }
+	= \sum_{i=0}^{d-1} \paren{\log\paren{1 - \frac{i}{n}} - \frac{1}{n-i}} \; .
 	\label{eq:psid_bound}
 \end{align}
-We can bound the harmonic sum thanks to the typical bound
+We can bound sum of reciprocals $\sum_{i=0}^{d-1} \frac{1}{n-i}$ by the typical bound on 
+the harmonic sum $H_n = \sum_{k=1}^n \frac{1}{k}$,
 \aligns{
@@
 	\leq H_n - \log(n) - \gamma \leq 
-	\inv{2n - 1}
+	\inv{2n - 1},
 }
@@
 	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
-	&= H_{n-d} - H_n \\
-	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1}\\
-	&= \log(1-\frac{d}{n})  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)}
+	= H_{n-d} - H_n
+	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1},
+	\\
+	&= \log\paren{1 - \frac{d}{n}}  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)}.
 \end{align}
@@
 	\psi_d(\half[n]) - d \log(\half[n]) 
-	\leq \sum_{i=0}^{\textcolor{red}{d}} \log(1 - \frac{i}{n})
-	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)} 
-\end{align}
-where we emphasized that the sum runs up to $d$.
-Now to bound this sum of logarithms, we can use a concave tangent bound on $\log$ to get
-\begin{align}
-	\sum_{i=0}^{d} \log(1 - \frac{i}{n})
+	\leq \sum_{i=0}^{\textcolor{red}{d}} \log\paren{1 - \frac{i}{n}}
+	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)},
+\end{align}
+where the sum now goes up to $d$.
+To bound the remaining sum, we use $\log(1+x)\leq x$ (for $x > -1$) to get
+\begin{align}
+	\sum_{i=0}^{d} \log\paren{1 - \frac{i}{n}}
 	\leq \sum_{i=0}^{d} - \frac{i}{n}
@@
 \end{align}
-Putting it all together yields
+Putting those bounds together in 
+\cref{eq:where-we-put-it-back-together-multivariate-gaussian}
+and reorganizing yields
 \begin{align}
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{(d + 1)}{(2n+1)(2(n - d) - 1)} \\
- 	&\leq p \frac{n + d +1}{2 n (n-d-1)}  + \frac{(d + 1)}{2n.2 (n - d - 1)}\\
- 	&= \frac{p}{2n} + \frac{2 p(d+1)}{2 n (n-d-1)} + \frac{p/d}{2n (n - d - 1)} \\
+ 	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{(d + 1)}{(2n+1)(2(n - d) - 1)}, 
+ 	\intertext{\fdk{I don't follow the second term on the next line. Is it because it is $\leq$ rather than $=$?}}
+ 	&\leq p \frac{n + d +1}{2 n (n-d-1)}  + \frac{(d + 1)}{2n.2 (n - d - 1)}, \\
+ 	&= \frac{p}{2n} + \frac{2 p(d+1)}{2 n (n-d-1)} + \frac{p/d}{2n (n - d - 1)}, \\
+ 	\intertext{\fdk{Why is this an inequality and }}
  	&\leq \frac{p}{2n} + \frac{p(d+2)}{n (n - d -1)}
@@
 
-\subsection{Bounds on the Expected MAP Natural Parameter}
+\subsection{Bounding the expected natural parameters for the MAP}
 Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
 We introduce the notation $(z)_+ = \max(0,z)$.
-\begin{lemma}[Expected MAP natural parameter]
-	Define $a=n_0 \frac{\mu_0}{\mu^*} $ a variable characterizing the importance of the prior relative to the true parameters.
-	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded as
+\begin{lemma}[Expected MAP natural parameter]\label{lem:expected-map-natural-parameter-gaussian}
+	Define the variable $a = n_0 \frac{\mu_0}{\mu^*}$, which characterizes 
+	the importance of the prior relative to the true parameters.
+	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded by
 \begin{equation}
@@
 \begin{proof}
-	The lower bound can be readily obtained with the Jensen inequality applied on the convex function $x\mapsto\inv{x}$ for $x>0$.
+	The lower bound can be readily obtained by applying Jensen's inequality 
+	to the convex function $x\mapsto\inv{x}$ for $x>0$.
 	The upper bound requires more work.
@@
 		= \expect{\frac{\MAPt}{\nat^*}} 
-		&= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  \\
-		&= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} \\
-		&= (n_0 +n) \expect{\inv{a + \chi^2(n)}}
+		= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  
+		= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} 
+		= (n_0 +n) \expect{\inv{a + \chi^2(n)}},
 }
@@
  	\expect{\inv{a + \chi^2(n)}}  
-	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx \\
+	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx.
 }
 	We reformulate this expectation with the trick
+	\fdk{I don't follow from there.}
 \alignn{
@@
 	where we used Fubini to switch integrals, without further justification.
+	\fdk{Fubini's theorem would be a justification here, right? Are you trying to say ``assuming we can exchange the order of integration''?}
 	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
@@
 	Fortunately, mathematicians have been working on these integrals for decades.
+	\fdk{Was this a comment from your notes that passed its way here or do you want to go with this?}
 	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
@@
 We did not find a closed form or an upper bound for the expected logarithm of the MAP $\expect{\log\frac{\MAPm}{\mu^*}}$.
-Consequently, we derived a bound for the symmetrized Bregman~\eqref{eq:symmetrized_bregman} instead. This bound is not asymptotically  tight for the Bregman by a factor $2$.
+Consequently, we derived a bound for the symmetrized Bregman~\eqref{eq:symmetrized_bregman} instead. 
+This bound is asymptotically tight for the Bregman, up to a factor $2$.
 
@@
 	& \expect{\cS_\conj( \m^*; \hat \m_n^\mathrm{MAP})}
-	\leq \begin{cases}
-		\cS_\conj( \mu_*; \mu_0) \ \text{ if } \ n=0, \\
-		\inv{2(n_0+1)}  +  b_1 \ \text{ if }\ n=1,\\
-		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{ if }\ n\geq 2
-	\end{cases}\\
-	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
+	\leq 
+	\left\{\begin{array}{ll}
+		\cS_\conj( \mu_*; \mu_0) 					& \text{ if } \ n=0, \\
+		\inv{2(n_0+1)}  +  b_1 						& \text{ if }\ n=1,\\
+		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n  & \text{ if }\ n\geq 2,
+	\end{array}\right.
+	\quad 
+	\quad 
+	\text{where }
+	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . 
+	%\nonumber
 \end{align}
@@
 \end{align}
-There remains the more problematic term with the expectation of the inverse mean parameter.
-Fortunately, we derived the bound \eqref{eq:lemma_nat_bound} for this purpose.
+There remains the more problematic term with the expectation of the inverse mean parameter, 
+for which we use the bound derived in Lemma \ref{lem:expected-map-natural-parameter-gaussian}.
 
@@
 The log-partition is neither self concordant nor pseudo-self-concordant.
+
+\fdk{TODO: Double check the above, something doesn't sit quite right, for the following. How can we get $g'' = 0$ if $A$strictly convex?}
 
@@
 
-\rlp{Introduce SMD, in primal and dual formulations.}
-
+\rlp{Introduce SMD, in primal (minimization) and dual formulations.}
+
+\iffalse
 \section{TODO}
@@
 \end{itemize}
+\fi
 

### 2021-10-21 aistats.tex

@@
 
+\subsection{Properties of Self-concordant functions}
+
+We quickly reviews some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}.
+We start with a few notations.
+Let $A^*$ be a self-concordant function.
+Then, we write
+\begin{itemize}
+	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
+	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.
+\end{itemize}
+Note that $\omega^*(t)$ is positive, convex and monotonically increasing for $t\in[0,1]$.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega.eps}
+	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega_star.eps}
+	\caption{(left) Graph of the distance function $\omega$ and its dual $\omega^*$.
+(right) Graph of the dual of distance function $\omega^*$ evaluated at $\frac{t}{1-t}$, compared with $t^2$.
+The two curves cross each other at $t\approx 0.21$.}
+	\label{fig:omega}
+\end{figure}
+
+We now present two important results. The first one shows how to convert the local norm $\|y-x\|_y$ using $\|y-x\|_x$.
+\begin{proposition}\label{prop:conversion_norm}
+	(Conversion of norms, \citep[Theorem 4.1.5]{nesterov2003introductory}) For any $x,\,y\in\dom A^*$, if $\|y-x\|_x<1$, then
+	\[
+		\|y-x\|_y \leq \frac{\|y-x\|_x}{1-\|y-x\|_x}.
+	\]
+\end{proposition}
+
+The next result show that, if $y$ is sufficiently close to $x$, then we can bound the Bregman divergence of $A^*$ using the distance function $\omega^*$ and local norms.
+\begin{proposition}(Upper bound of self-concordant functions \citep[Theorem 4.1.8]{nesterov2003introductory}) \label{prop:upper_bound_self_concordance}
+	For any $x,\,y\in\dom A^*$, if $\|y-x\|_x<1$, then
+	\[
+		\bregmanconj(y,x) \leq \omega^*(\|y-x\|_x).
+	\]
+\end{proposition}
+
+We are now ready to prove Proposition~\ref{prop:selfConcordant}.
+
+\subsection{Proof of Proposition~\ref{prop:selfConcordant}.}
+
+We start with Proposition~\ref{prop:upper_bound_self_concordance}, evaluated at $y=\meanp^*$ and $x=\meanp$:
+\[
+	\bregmanconj(\meanp^*,\meanp) \leq \omega^*(\|\meanp^*-\meanp\|_{\meanp}).
+\]
+This hold if $\|\meanp^*-\meanp\|_{\meanp}<1$.
+Since $\omega^*$ is monotonically increasing, we can replace $\|\meanp^*-\meanp\|_{\meanp}$ by its upper bound from Proposition~\ref{prop:conversion_norm},
+\[
+	\omega^*(\|\meanp^*-\meanp\|_{\meanp}) \leq \omega^*\left(\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}}\right),
+\]
+under the conditions that $\|\meanp^*-\meanp\|_{\meanp^*}<1$ (to satisfy the assumption of Proposition~\ref{prop:conversion_norm}) and $\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}} <1$ (to ensure that $\|\meanp^*-\meanp\|_{\meanp} <1$).
+Those two conditions holds if $\|\meanp^*-\meanp\|_{\meanp^*} <0.5$.
+Now, we use the bound (see figure \ref{fig:omega})
+\[
+	\omega^*\left(\frac{t}{1-t}\right) \leq t^2, \quad 0\leq t \leq 0.21,
+\]
+and replace $t$ by $\|\meanp^*-\meanp\|_{\meanp^*}$. This finally gives the sequence of inequalities
+\[
+	\bregmanconj(\meanp^*,\meanp) \leq \omega^*(\|\meanp^*-\meanp\|_{\meanp}) \leq \omega^*\left(\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}}\right) \leq \|\meanp^*-\meanp\|_{\meanp^*}^2,
+\]
+that holds while $ \|\meanp^*-\meanp\|_{\meanp^*}<0.21$, which is the desired result.
+
+%\begin{proposition}
+%	Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
+%	\aligns{
+%		\norm{\m^*-\m}_{\mF} < 0.21
+%		\implies
+%		\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
+%	}
+%\end{proposition}
+%\begin{proof}
+%	
+%\end{proof}
+
 \section{BIAS-VARIANCE}

### 2021-10-21 aistats.tex
Go over Frederik's feedback
And spot the error in self-concordance of the log-partition of a Gaussian

@@
 
-\pdfsuppresswarningpagegroup=1 %removes the warning
+%\pdfsuppresswarningpagegroup=1 %removes the warning
 % This is a hack, the other way to remove is to convert all pdf into postscript or eps.
+% RLP: not sure what is the point of this. With my latex it creates an error, and I do not have warnings, so I comment it out for now.
 
@@
 
-\usepackage[utf8]{inputenc} % allow utf-8 input
+%\usepackage[utf8]{inputenc} % allow utf-8 input
 \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
@@
 	\end{align}
-	There is also a closed form solution for the expected logarithm of a Chi-squared, 
-	because a Chi-squared is a special case of Gamma distributions, whose sufficient statistic is $(X, \log(X))$,
-	\fdk{I don't follow why this is ``because it is a Gamma with sufficient statistics $(X,\log(X))$''?}
+	There is also a closed form solution for the expected logarithm of a Chi-squared.
+	Indeed a Chi-squared is a special case of Gamma distributions, whose sufficient statistic is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a Gamma.
+The result is
 	\begin{align}
@@
 \subsection{Multivariate MLE}
-In higher dimensions, $X_i \sim \cN(0, \mu_*)$, $X_i\in\real^d$, and the mean parameter $\mu_*$ is a symmetric positive definite covariance matrix of order $d$, living in dimension $p = \frac{d(d+1)}{2}$.
-Note that we switched the role of $d$ from $\dim(T(x))=d$ to $\dim(x) =d$ and $\dim(T(x))=p$.
-\fdk{This $\dim$ is not introduced? I would say that $\dim(T(X))$ is $d^2$ if $T(X) = XX^\top$. Slightly confused here.}
+In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ is a symmetric positive definite covariance matrix of order $d$, living in dimension $p = \frac{d(d+1)}{2}$.
+Note that we switched the role of $d$ from $\dim(T(\cX))=d$ to $\dim(\cX) =d$ and $\dim(T(\cX))=p$, where $\dim(\cZ)$ is the dimension of the smallest affine subspace containing $\cZ$.
 
@@
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	= \half \left ( \frac{2p}{n-d-1} + \psi_d\paren{\half[n]} - d \log\paren{\half[n]}  \right ).
+ 	= \frac{p}{n-d-1} + \half \left ( \psi_d\paren{\half[n]} - d \log\paren{\half[n]}  \right ).
  \end{align}
@@
 	\psi_d\paren{\half[n]} - d \log\paren{\half[n]}
-	&= \sum_{i=0}^{d-1} \paren{\psi \paren{\half[n - i]} - \log\paren{\half[n]} },
-	\\
-	&\leq \sum_{i=0}^{d-1} \paren{\log\paren{\half[n-i]} - \inv{n-i} - \log\paren{\half[n]} }
-	= \sum_{i=0}^{d-1} \paren{\log\paren{1 - \frac{i}{n}} - \frac{1}{n-i}} \; .
+	&= \sum_{i=0}^{d-1} \paren{\psi \paren{\half[n - i]} - \log\paren{\half[n]} } \\
+	&\leq \sum_{i=0}^{d-1} \paren{\log\paren{\half[n-i]} - \inv{n-i} - \log\paren{\half[n]} } \\
+	&= \sum_{i=0}^{d-1} \paren{\log\paren{1 - \frac{i}{n}} - \frac{1}{n-i}} \; .
 	\label{eq:psid_bound}
@@
 	= H_{n-d} - H_n
-	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1},
-	\\
-	&= \log\paren{1 - \frac{d}{n}}  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)}.
+	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1} \\
+	&= \log\paren{1 - \frac{d}{n}}  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)} \\
+	&< \log\paren{1 - \frac{d}{n}}  + \frac{d + 1}{2n(n - d - 1)} \; .
 \end{align}
@@
 	\leq \sum_{i=0}^{\textcolor{red}{d}} \log\paren{1 - \frac{i}{n}}
-	+ \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)},
+	+  \frac{d + 1}{2n(n - d - 1)} \; ,
 \end{align}
@@
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{(d + 1)}{(2n+1)(2(n - d) - 1)}, 
- 	\intertext{\fdk{I don't follow the second term on the next line. Is it because it is $\leq$ rather than $=$?}}
- 	&\leq p \frac{n + d +1}{2 n (n-d-1)}  + \frac{(d + 1)}{2n.2 (n - d - 1)}, \\
- 	&= \frac{p}{2n} + \frac{2 p(d+1)}{2 n (n-d-1)} + \frac{p/d}{2n (n - d - 1)}, \\
- 	\intertext{\fdk{Why is this an inequality and }}
+ 	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{d + 1}{4n(n - d - 1)}\\
+ 	&< p \frac{n + d +1}{2 n (n-d-1)}  + \frac{d(d + 1)/2}{2dn(n - d - 1)}\\
+ 	&= \frac{p}{2n}\frac{n-d-1}{n-d-1} + p\frac{2(d+1)}{2 n (n-d-1)} + \frac{p}{2dn (n - d - 1)}\\
+ 	&= \frac{p}{2n} + \frac{p(d+ 1 + \inv{2d})}{n (n-d-1)} \\
  	&\leq \frac{p}{2n} + \frac{p(d+2)}{n (n - d -1)}
@@
 		= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} 
-		= (n_0 +n) \expect{\inv{a + \chi^2(n)}},
+		= (n_0 +n) \expect{\inv{a + \chi^2(n)}} \; ,
 }
@@
  	\expect{\inv{a + \chi^2(n)}}  
-	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx.
-}
-	We reformulate this expectation with the trick
-	\fdk{I don't follow from there.}
+	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx \; .
+}
+	This $1/(a+x)$ is hard to deal with.
+	Fortunately, there is a trick to express it as an integral itself , with exponential terms that fit well with the $\chi^2$-density,
+	\fdk{I don't follow from there.} \rlp{do you have more detailed feedback ? I don't know how to improve it.}
 \alignn{
@@
 		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
-		\implies & \boxed{\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt \; ,}
-} 
-	to get two integrals instead of one
+		\implies & \boxed{\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt  \; .}
+} 
+We can then plug this integral into the expectation and apply Fubini's theorem to get a tractable integral on the inside
 \alignn{
@@
 }
-	where we used Fubini to switch integrals, without further justification.
-	\fdk{Fubini's theorem would be a justification here, right? Are you trying to say ``assuming we can exchange the order of integration''?}
+	assuming we can exchange the order of integration.
+	%\fdk{Fubini's theorem would be a justification here, right? Are you trying to say ``assuming we can exchange the order of integration''?}\rlp{yes exactly.}
 	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
@@
 	Fortunately, mathematicians have been working on these integrals for decades.
-	\fdk{Was this a comment from your notes that passed its way here or do you want to go with this?}
+	\fdk{Was this a comment from your notes that passed its way here or do you want to go with this?} \rlp{Yes I like this sentence.}
 	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
@@
 where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
-These functions and their domains they are represented in Figure \TODO{}.
+These functions and their domains are represented in Figure \TODO{}.
 We provide definitions of self-concordance in \cref{app:self-concordant}.
@@
 \paragraph{Entropy is Self-Concordant,}
-\citet[p.177]{nesterov2003introductory} proves that logarithmic barriers for a second order region are self-concordant.
-Adopting their notation, $\conj$ fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = \begin{pmatrix} 1 \\ 0 \end{pmatrix}$.
-
-\paragraph{Log-partition is not Self-Concordant}
-Let us fix a straight line with $\nat_1 \neq 0$, e.g. $g(\nat_2) := A(\nat_1,\nat_2)$. Then
-\alignn{
-g'(\nat_2) = \frac{\partial A}{\partial\nat_2}(\nat) 
-&= +\frac{\nat_1^2}{4 \nat_2^2} + \frac{1}{2\nat_2} \\
-g''(\nat_2) = \frac{\partial^2 A}{(\partial\nat_2)^2}(\nat) 
-&= -\frac{\nat_1^2}{2 \nat_2^3} - \frac{1}{2\nat_2^2} 
-=  \frac{1}{2\nat_2^2}\paren{\frac{\nat_1^2}{-\nat_2^2} - 1} \\ 
-g'''(\nat_2) = \frac{\partial^3 A}{(\partial\nat_2)^3}(\nat) 
-&= +\frac{3\nat_1^2}{2 \nat_2^4} + \frac{1}{\nat_2^3}
-= \frac{1}{2\nat_2^3}\paren{2 - 3 \frac{\nat_1^2}{- \nat_2^2}}
-}
-Setting $\nat_2 = - \nat_1^2$ yields $g''(\nat_2) = 0$ and $g'''(\nat_2) \neq 0$, meaning the second derivative cannot dominate the third derivative for any choice of constant or exponent.
+\citet[Example 4.1.1.4,   p.177]{nesterov2003introductory} proves that logarithmic barriers for a second order region are self-concordant.
+Adopting their notation, $\conj$ fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = ( 1 \ 0 )^\top$.
+
+\paragraph{Log-partition is Self-Concordant}
+Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(0 \ -1)^\top$.
+Then 
+\alignn{g(t) := 2A(w+tv) = \frac{a}{t} - \log(t)}
+is well defined for $t>0$ and $a=\frac{\nat_1^2}{2} \geq 0$.
+The logarithm is self-concordant for the same reason that the entropy is, but the inverse function $\inv{t}$ is not \citep[example 4.1.1.5]{nesterov2003introductory}.
+Linear combinations with positive coefficients  preserves self-concordance \citep[theorem 4.1.1]{nesterov2003introductory} -- e.g.
+self-concordance is a cone, so it seems unlikely that this sum of a self-concordant function and a not self-concordant function would be self-concordant.
+But let us verify it to be sure. The derivatives of $g$ are
+\alignn{
+g'(t) &= -\frac{a}{t^2} - \frac{1}{t} \; ,\\
+g''(t) &= \frac{2a}{t^3} + \inv{t^2} = \frac{1}{t^2}\paren{\frac{2a}{t} + 1} \; , \\
+g'''(t) &= -\frac{6a}{t^4} - \frac{2}{t^3} = \frac{-2}{t^3}\paren{\frac{3a}{t} + 1} \;.
+}
+Then the definition of self-concordance reads
+\alignn{
+\abs{g'''(t)} &\leq 2 g''(t)^{\frac{3}{2}} \\
+\iff \frac{3a}{t} + 1 &\leq \paren{\frac{2a}{t} + 1}^{\frac{3}{2}} \; .
+}
+Writing $y = \frac{a}{t} \geq 0$, and taking the square of this expression yields
+\alignn{
+ (3y + 1)^2 &\leq (2y + 1)^3 \\
+ \iff 9 y ^2 + 6y + 1 &\leq 8 y^3  + 12 y^2 + 6y +1 \\
+\iff 0 &\leq y^2(8y + 3)
+}
+which is true since $y\geq 0$.
+
 The log-partition is neither self concordant nor pseudo-self-concordant.
-
-\fdk{TODO: Double check the above, something doesn't sit quite right, for the following. How can we get $g'' = 0$ if $A$strictly convex?}
 
@@
 \begin{definition}[Self-concordance]
-A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}}$
+\label{def:self-concordance}
+A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
+\alignn{\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}}}
+for all feasible $t$.
 \end{definition}
@@
 \begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
-A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies $\abs{g'''(x)} \leq C \norm{v}_2 g''(x)$ for some $C\geq 0$
+A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
+\alignn{\abs{g'''(x)} \leq C \norm{v}_2 g''(x)}
+for all feasible $t$ and for some $C\geq 0$.
 \end{definition}

### 2021-10-21 aistats.tex
Add table of content for appendix only
Make norm matrices consistent
minor fix

@@
 
+% To make a table of contents only for the appendix
+\usepackage[toc,page,header]{appendix}
+\usepackage{minitoc}
+\renewcommand \thepart{}
+\renewcommand \partname{}
+
 \usepackage[capitalise]{cleveref}
@@
 \end{abstract}
+
+\doparttoc % Tell to minitoc to generate a toc for the parts
+\faketableofcontents % Run a fake tableofcontents command for the partocs
+%\part{} % Start the document part
+%\parttoc % Insert the document TOC
 
@@
 
+\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
+\part{Appendix} % Start the appendix part
+\parttoc % Insert the appendix TOC
+
 \section{PROOFS FOR GAUSSIAN VARIANCE}
@@
 }
-This bound is asymptotically tight.
 \end{theorem}
+We see from~\eqref{eq:asymptote} that this bound is asymptotically tight.
 \begin{proof}
@@
 which is true since $y\geq 0$.
-
-The log-partition is neither self concordant nor pseudo-self-concordant.
+The log-partition is self concordant.
 
@@
 \begin{align}
-	\E \half \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mSigma^{-1}}
-	&= \frac{n d}{2(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \half \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} \\
-	&= \frac{d}{2n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mSigma^{-1}} }{n^2} \right) \; .
+	\E \half \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mF}
+	&= \frac{n d}{2(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \half \norm{\mu^* -  \mu_0}^2_{\mF} \\
+	&= \frac{d}{2n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mF} }{n^2} \right) \; .
 \end{align}

### 2021-10-21 aistats.tex
Complete proof of self-concordance

@@
 \paragraph{Log-partition is Self-Concordant}
-Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(0 \ -1)^\top$.
-Then 
-\alignn{g(t) := 2A(w+tv) = \frac{a}{t} - \log(t)}
-is well defined for $t>0$ and $a=\frac{\nat_1^2}{2} \geq 0$.
+Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(\alpha \ \ -1)^\top$. 
+For varying $\nat_1, \alpha$, this covers the whole range of possible straight lines in the 2D plane, except lines of constant $\nat_2$ (horizontal).
+But on lines with $\nat_2=\cst$,   the log-partition function $A$ is simply a quadratic, so it is self-concordant.
+Then, up to constants,
+\alignn{
+g(t) := 2A(w+tv) 
+= \frac{(\nat_1 + t \alpha)^2}{2 t} - \log(t) + \cst
+= \frac{a}{t}  + b t - \log(t) + \cst
+}
+where $a=\frac{\nat_1^2}{2} \geq 0$, and $b = \frac{\alpha^2}{2}$.
+$g$ is well defined for $t>0$.
 The logarithm is self-concordant for the same reason that the entropy is, but the inverse function $\inv{t}$ is not \citep[example 4.1.1.5]{nesterov2003introductory}.
@@
 self-concordance is a cone, so it seems unlikely that this sum of a self-concordant function and a not self-concordant function would be self-concordant.
-But let us verify it to be sure. The derivatives of $g$ are
-\alignn{
-g'(t) &= -\frac{a}{t^2} - \frac{1}{t} \; ,\\
+So let us prove it.
+The derivatives of $g$ are
+\alignn{
+g'(t) &= -\frac{a}{t^2} - \frac{1}{t} + b \; ,\\
 g''(t) &= \frac{2a}{t^3} + \inv{t^2} = \frac{1}{t^2}\paren{\frac{2a}{t} + 1} \; , \\

### 2021-10-21 aistats.tex
Move duality at the end of self-concordance

@@
 
-\begin{proposition}
-	\label{prop:conjugateConcordance}
-	Let $f:\real \rightarrow \real$ be a self-concordant function of Legendre type. Then its convex conjugate $f^*$ is also self-concordant.
-\end{proposition}
-\begin{proof}
-When $f$ is of Legendre type then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ has inverse gadients
-\begin{align}
-    \nabla f \circ \nabla f^*  &= \Id \\
-     \nabla f^* \circ \nabla f &= \Id
-\end{align}
-where $\Id$ is the identity function on the relevant domain. 
-Deriving this equality proves that some Hessians are inverse of each other 
-\begin{align}
-    \nabla^2 f(x) \nabla^2 f^*(x^*) = \mI
-    \label{eq:conjugateHessian}
-\end{align}
-where $(x,x^*)$ are conjugate points -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
-\TODO{Why are Hessians not 0?}
-In 1 dimension, this inequality is more simply written 
-\aligns{f''(x) f^{* \prime \prime}(x^*) = 1}
-Until now, this should be familiar to the reader.
-The interesting surprise is that self-concordance appears when we derive again this equality. 
-\begin{align}
-    &1= f''(x) f^{* \prime \prime}(f'(x)), \forall x \\
-    \implies
-    &0 = f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))\\
-    \iff
-    &0 = f'''(x)f^{* \prime \prime}(x^*)  + f''(x)^2 f^{* \prime \prime \prime}(x^*)\\
-    \implies
-    &0 = \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  f^{* \prime \prime}(x^*)^{-\half}
-    + f''(x)^{\half} \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}} \\
-    \implies
-    &0= \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}}
-\end{align}
-where to reach the last line, we used~\eqref{eq:conjugateHessian} again.
-We see that for a pair of conjugate functions, the self-concordance ratios are opposite of each other, which concludes the proof.
-\end{proof}
-
 \subsection{Properties of Self-concordant functions}
@@
 %\end{proof}
+
+\subsection{Self-concordance and Convex Conjugates}
+\begin{proposition}
+	\label{prop:conjugateConcordance}
+	Let $f:\real \rightarrow \real$ be a self-concordant function of Legendre type. Then its convex conjugate $f^*$ is also self-concordant.
+\end{proposition}
+\begin{proof}
+When $f$ is of Legendre type then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ has inverse gadients
+\begin{align}
+    \nabla f \circ \nabla f^*  &= \Id \\
+     \nabla f^* \circ \nabla f &= \Id
+\end{align}
+where $\Id$ is the identity function on the relevant domain. 
+Deriving this equality proves that Hessians are inverse of each other 
+\begin{align}
+    \nabla^2 f(x) \nabla^2 f^*(x^*) = \mI
+    \label{eq:conjugateHessian}
+\end{align}
+when evaluated at conjugate pairs of points $(x,x^*)$ -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
+These Hessians are never $0$.
+In 1 dimension, this inequality is more simply written 
+\aligns{f''(x) f^{* \prime \prime}(x^*) = 1}
+Until now, this should be familiar to the reader.
+The interesting surprise is that self-concordance appears when we derive again this equality. 
+\begin{align}
+    &1= f''(x) f^{* \prime \prime}(f'(x)), \forall x \\
+    \implies
+    &0 = f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))\\
+    \iff
+    &0 = f'''(x)f^{* \prime \prime}(x^*)  + f''(x)^2 f^{* \prime \prime \prime}(x^*)\\
+    \implies
+    &0 = \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  f^{* \prime \prime}(x^*)^{-\half}
+    + f''(x)^{\half} \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}} \\
+    \implies
+    &0= \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}}
+\end{align}
+where to reach the last line, we used~\eqref{eq:conjugateHessian} again.
+We see that for a pair of conjugate functions, the self-concordance ratios are opposite of each other, which concludes the proof.
+\end{proof}
 

### 2021-10-21 aistats.tex
Clarification comments on A.1, A.2

@@
 	\begin{align}
-		\expect{\log \frac{\hat \mu_n}{\mu_*}} = \psi\paren{\half[n]} - \log\paren{\half[n]}
+		\expect{\log \frac{\hat \mu_n}{\mu_*}} = \psi\paren{\half[n]} - \log\paren{\half[n]},
 	\end{align}
@@
 Note that we switched the role of $d$ from $\dim(T(\cX))=d$ to $\dim(\cX) =d$ and $\dim(T(\cX))=p$, where $\dim(\cZ)$ is the dimension of the smallest affine subspace containing $\cZ$.
+
+\fdk{I'm worried talking about dimension might be confused with curved EF (less parameters than sufficient statistics). What do you think of
+%
+\\~\\
+%
+In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ 
+is a $d \times d$ symmetric, positive definite covariance matrix with $p = \frac{d(d+1)}{2}$ degrees of freedom.
+Note that here $d$ denotes the dimensionality of the data $X$, rather than the dimensionality of the parameters $\mu_*$.
+}
 
@@
  	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{d + 1}{4n(n - d - 1)}\\
+ 	\intertext{\fdk{Isn't this =?
+ 	For the first term, $\inv{n-d-1} - \inv{2n} = \frac{2n}{2n(n-d-1)} - \frac{n-d-1}{2n(n-d-1)} = \frac{n+d+1}{2n(n-d-1)}$ and the second is $\times d/d$?
+ 	}}
  	&< p \frac{n + d +1}{2 n (n-d-1)}  + \frac{d(d + 1)/2}{2dn(n - d - 1)}\\
- 	&= \frac{p}{2n}\frac{n-d-1}{n-d-1} + p\frac{2(d+1)}{2 n (n-d-1)} + \frac{p}{2dn (n - d - 1)}\\
+ 	&= p\frac{n-d-1}{2n(n-d-1)} + p\frac{2(d+1)}{2 n (n-d-1)} + \frac{p}{2dn (n - d - 1)}\\
  	&= \frac{p}{2n} + \frac{p(d+ 1 + \inv{2d})}{n (n-d-1)} \\
@@
 	Fortunately, mathematicians have been working on these integrals for decades.
-	\fdk{Was this a comment from your notes that passed its way here or do you want to go with this?} \rlp{Yes I like this sentence.}
 	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}

### 2021-10-21 aistats.tex
Add self-concordance erratum

--- 2021-10-21 aistats.tex
@@
 \renewcommand \partname{}
+
+% To make colored boxes
+\usepackage[framemethod=TikZ]{mdframed}
+\newmdenv[
+	roundcorner=10pt,
+	backgroundcolor=Red!30,
+]{important}
 
@@
 \part{Appendix} % Start the appendix part
+\begin{important}
+	{\bf Erratum:}  contrary to what is written in the main text, the log-partition of a Gaussian \emph{is} self-concordant. See \cref{app:gaussian} for details. 
+\end{important}
 \parttoc % Insert the appendix TOC
@@
 
-In this section, we prove that for a Gaussian, the entropy is self-concordant while the log-partition function is not.
-We also give complementary illustrations on the Gaussian mirror map and MLE.
+\begin{important}
+{\bf Erratum:} There was an error in our derivations. In the main text, we claimed that the log-partition of a Gaussian was not self-concordant. We will now prove that the opposite is true.
+\end{important}
+
+In this section, we prove that for a Gaussian, the entropy and log-partition functions are self-concordant.
+%We also give complementary illustrations on the Gaussian mirror map and MLE.
 
@@
 where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
-These functions and their domains are represented in Figure \TODO{}.
+%These functions and their domains are represented in Figure \TODO{}.
 We provide definitions of self-concordance in \cref{app:self-concordant}.
@@
 \paragraph{Log-partition is Self-Concordant}
-Adopting the notation of definition~\ref{def:self-concordance}, let us set $w=(\nat_1 \ 0)^\top$ and $v=(\alpha \ \ -1)^\top$. 
-For varying $\nat_1, \alpha$, this covers the whole range of possible straight lines in the 2D plane, except lines of constant $\nat_2$ (horizontal).
-But on lines with $\nat_2=\cst$,   the log-partition function $A$ is simply a quadratic, so it is self-concordant.
-Then, up to constants,
-\alignn{
-g(t) := 2A(w+tv) 
-= \frac{(\nat_1 + t \alpha)^2}{2 t} - \log(t) + \cst
-= \frac{a}{t}  + b t - \log(t) + \cst
-}
-where $a=\frac{\nat_1^2}{2} \geq 0$, and $b = \frac{\alpha^2}{2}$.
-$g$ is well defined for $t>0$.
-The logarithm is self-concordant for the same reason that the entropy is, but the inverse function $\inv{t}$ is not \citep[example 4.1.1.5]{nesterov2003introductory}.
-Linear combinations with positive coefficients  preserves self-concordance \citep[theorem 4.1.1]{nesterov2003introductory} -- e.g.
-self-concordance is a cone, so it seems unlikely that this sum of a self-concordant function and a not self-concordant function would be self-concordant.
-So let us prove it.
-The derivatives of $g$ are
-\alignn{
-g'(t) &= -\frac{a}{t^2} - \frac{1}{t} + b \; ,\\
-g''(t) &= \frac{2a}{t^3} + \inv{t^2} = \frac{1}{t^2}\paren{\frac{2a}{t} + 1} \; , \\
-g'''(t) &= -\frac{6a}{t^4} - \frac{2}{t^3} = \frac{-2}{t^3}\paren{\frac{3a}{t} + 1} \;.
-}
-Then the definition of self-concordance reads
-\alignn{
-\abs{g'''(t)} &\leq 2 g''(t)^{\frac{3}{2}} \\
-\iff \frac{3a}{t} + 1 &\leq \paren{\frac{2a}{t} + 1}^{\frac{3}{2}} \; .
-}
-Writing $y = \frac{a}{t} \geq 0$, and taking the square of this expression yields
-\alignn{
- (3y + 1)^2 &\leq (2y + 1)^3 \\
- \iff 9 y ^2 + 6y + 1 &\leq 8 y^3  + 12 y^2 + 6y +1 \\
-\iff 0 &\leq y^2(8y + 3)
-}
-which is true since $y\geq 0$.
-The log-partition is self concordant.
+As proved in \citet{nesterov1994interior} self-concordance is preserved by Fenchel conjugacy.
+Since $\conj$ is self-concordant, $\logpart$ as well.
+For a more accessible reference, see also \citet[Prop.~6]{sun2019generalized}.
 
@@
 
-\TODO{Damien : review self-concordance + prove proposition 1 + graph to explain $x<0.21$.}
-
 \begin{definition}[Self-concordance]
@@
 \end{definition}
+
+{\bf Correction:} In the main text, we claimed that the Fenchel conjugate of a 1 dimensional function is also self-conjugate. 
+Actually, this is also true in higher dimension, as proved by \citet{nesterov1994interior}. See \citet[Prop.~6]{sun2019generalized} for a more accessible reference.
 
@@
     &0 = f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))\\
-    \iff
+    \implies
     &0 = f'''(x)f^{* \prime \prime}(x^*)  + f''(x)^2 f^{* \prime \prime \prime}(x^*)\\

### 2021-10-21 aistats.tex
Notation issue in D?

@@
 \begin{itemize}
-	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
+	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ 
+	and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
+	\fdk{The second one should be 
+	$\sqrt{ \langle [\nabla^2A^*({\color{red} \nabla A(x)})]^{-1}\cdot,\, \cdot \rangle }$?
+	}
 	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.

### 2021-10-21 aistats.tex
Incorporate Frederik's comments

@@
 \subsection{Multivariate MLE}
-In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ is a symmetric positive definite covariance matrix of order $d$, living in dimension $p = \frac{d(d+1)}{2}$.
-Note that we switched the role of $d$ from $\dim(T(\cX))=d$ to $\dim(\cX) =d$ and $\dim(T(\cX))=p$, where $\dim(\cZ)$ is the dimension of the smallest affine subspace containing $\cZ$.
-
-\fdk{I'm worried talking about dimension might be confused with curved EF (less parameters than sufficient statistics). What do you think of
-%
-\\~\\
-%
+
 In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ 
@@
 Note that here $d$ denotes the dimensionality of the data $X$, rather than the dimensionality of the parameters $\mu_*$.
-}
 
@@
  	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	&\leq p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{d + 1}{4n(n - d - 1)}\\
+ 	&< p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{d + 1}{4n(n - d - 1)}\\
  	\intertext{\fdk{Isn't this =?
@@
  	}}
- 	&< p \frac{n + d +1}{2 n (n-d-1)}  + \frac{d(d + 1)/2}{2dn(n - d - 1)}\\
+ 	&= p \frac{n + d +1}{2 n (n-d-1)}  + \frac{d(d + 1)/2}{2dn(n - d - 1)}\\
  	&= p\frac{n-d-1}{2n(n-d-1)} + p\frac{2(d+1)}{2 n (n-d-1)} + \frac{p}{2dn (n - d - 1)}\\
@@
 \end{align}
-where $p = d(d+1)/2$ and the last inequality used $\inv{2d}\leq 1$. 
+where $p = \frac{d(d+1)}{2}$ and the last inequality used $\inv{2d}\leq 1$. 
 \end{proof}
@@
 	This $1/(a+x)$ is hard to deal with.
-	Fortunately, there is a trick to express it as an integral itself , with exponential terms that fit well with the $\chi^2$-density,
+	Fortunately, there is a trick to express it as an integral itself , with an exponential that fits well with the $\chi^2$-density,
 	\fdk{I don't follow from there.} \rlp{do you have more detailed feedback ? I don't know how to improve it.}
@@
 \label{def:self-concordance}
-A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
-\alignn{\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}}}
-for all feasible $t$.
+A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies for all feasible $t$
+\alignn{
+	\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}} \; .
+}
 \end{definition}
 
-\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
-A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
-\alignn{\abs{g'''(x)} \leq C \norm{v}_2 g''(x)}
-for all feasible $t$ and for some $C\geq 0$.
-\end{definition}
+%\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
+%A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
+%\alignn{\abs{g'''(x)} \leq C \norm{v}_2 g''(x)}
+%for all feasible $t$ and for some $C\geq 0$.
+%\end{definition}
 
@@
 	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ 
-	and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
-	\fdk{The second one should be 
-	$\sqrt{ \langle [\nabla^2A^*({\color{red} \nabla A(x)})]^{-1}\cdot,\, \cdot \rangle }$?
-	}
+	%and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
+%	\fdk{The second one should be 
+%	$\sqrt{ \langle [\nabla^2A^*({\color{red} \nabla A(x)})]^{-1}\cdot,\, \cdot \rangle }$?
+%	}
 	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.
@@
 that holds while $ \|\meanp^*-\meanp\|_{\meanp^*}<0.21$, which is the desired result.
-
-%\begin{proposition}
-%	Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
-%	\aligns{
-%		\norm{\m^*-\m}_{\mF} < 0.21
-%		\implies
-%		\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
-%	}
-%\end{proposition}
-%\begin{proof}
-%	
-%\end{proof}
-
-\subsection{Self-concordance and Convex Conjugates}
-\begin{proposition}
-	\label{prop:conjugateConcordance}
-	Let $f:\real \rightarrow \real$ be a self-concordant function of Legendre type. Then its convex conjugate $f^*$ is also self-concordant.
-\end{proposition}
-\begin{proof}
-When $f$ is of Legendre type then its convex conjugate $f^*(y) = \max_x \langle x, y \rangle - f(x)$ has inverse gadients
-\begin{align}
-    \nabla f \circ \nabla f^*  &= \Id \\
-     \nabla f^* \circ \nabla f &= \Id
-\end{align}
-where $\Id$ is the identity function on the relevant domain. 
-Deriving this equality proves that Hessians are inverse of each other 
-\begin{align}
-    \nabla^2 f(x) \nabla^2 f^*(x^*) = \mI
-    \label{eq:conjugateHessian}
-\end{align}
-when evaluated at conjugate pairs of points $(x,x^*)$ -- e.g. $x^*=\nabla f(x)$ and $x = \nabla f^*(x^*)$.
-These Hessians are never $0$.
-In 1 dimension, this inequality is more simply written 
-\aligns{f''(x) f^{* \prime \prime}(x^*) = 1}
-Until now, this should be familiar to the reader.
-The interesting surprise is that self-concordance appears when we derive again this equality. 
-\begin{align}
-    &1= f''(x) f^{* \prime \prime}(f'(x)), \forall x \\
-    \implies
-    &0 = f'''(x)f^{* \prime \prime}(f'(x))  + f''(x)^2 f^{* \prime \prime \prime}(f'(x))\\
-    \implies
-    &0 = f'''(x)f^{* \prime \prime}(x^*)  + f''(x)^2 f^{* \prime \prime \prime}(x^*)\\
-    \implies
-    &0 = \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  f^{* \prime \prime}(x^*)^{-\half}
-    + f''(x)^{\half} \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}} \\
-    \implies
-    &0= \frac{f'''(x)}{f''(x)^{\frac{3}{2}}}  + \frac{f^{* \prime \prime \prime}(x^*)}{f^{* \prime \prime}(x^*)^{\frac{3}{2}}}
-\end{align}
-where to reach the last line, we used~\eqref{eq:conjugateHessian} again.
-We see that for a pair of conjugate functions, the self-concordance ratios are opposite of each other, which concludes the proof.
-\end{proof}
 

### 2021-10-21 aistats.tex
Start typing proof of Hanzely expectation

@@
 \newcommand{\logpart}{A}
-\newcommand{\conj}{\logpart^*}
+\newcommand{\conj}{{\logpart^*}}
 \newcommand{\bregman}{\cB_\logpart}
@@
 	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
-	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:primal_pivot} from~\S\ref{ssec:bias-variance}.
+	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:bias-variance} from~\S\ref{ssec:bias-variance}.
 	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
@@
 	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
-	\label{eq:primal_pivot}
+	\label{eq:bias-variance}
 \end{equation}
@@
 However, this is not helpful, as verifying the assumption becomes as hard as the initial problem.
-For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper bound on the variance term of~\eqref{eq:primal_pivot} (cf App.~\ref{app:bias-variance}).
+For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper bound on the variance term of~\eqref{eq:bias-variance} (cf App.~\ref{app:bias-variance}).
 Confronted with this difficulty, we investigate an alternative definition of variance.
@@
 	Fortunately, there is a trick to express it as an integral itself , with an exponential that fits well with the $\chi^2$-density,
-	\fdk{I don't follow from there.} \rlp{do you have more detailed feedback ? I don't know how to improve it.}
 \alignn{
@@
 In this section, we prove that for a Gaussian, the entropy and log-partition functions are self-concordant.
-%We also give complementary illustrations on the Gaussian mirror map and MLE.
+% TODO We also give complementary illustrations on the Gaussian mirror map and MLE.
 
@@
 where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
-%These functions and their domains are represented in Figure \TODO{}.
+% TODO These functions and their domains are represented in Figure \TODO{}.
 We provide definitions of self-concordance in \cref{app:self-concordant}.
@@
 \label{app:bias-variance}
-
+In this section, we start from the notions of bias and variance introduced in \cref{eq:bias-variance}.
+First, we prove that the bias of the MLE of a Gaussian variance decreases in $O(\inv{n^2})$. 
+Then we prove that the expectation of \cref{eq:hanzely} is an upper bound on the variance term.
+
+\subsection{Bias of a Gaussian Variance MLE}
 \rlp{prove that the bias decreases like $1/n^2$ for gaussian variance and gaussian mean.}
 
-\rlp{prove that the expectation of Hanzely is the symmetrized bias, e.g.  an upper bound on the bias.}
+\subsection{Expectation of SMD's Variance Assumption}
+The first step is to notice the symmetrized Bregman can be expressed as an inner product between primal and dual parameters
+\alignn{
+	\cS_\conj (\mu , \bar\mu)
+	= \lin{ \nabla \conj(\mu) - \nabla \conj(\bar\mu) , \mu - \bar\mu}
+	= \lin{ \nat -\bar\nat , \mu - \bar\mu} \; .
+}
+Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_n = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_n =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
+This means that
+\alignn{
+	\expect{\cS_\conj (\mu , \E[\mu])} 
+	= \expect{\lin{\nat, \mu - \E[\mu]}}
+	=\Cov(\nat, \mu)
+	=\expect{\lin{\nat - \E[\nat],\mu}} 
+	= \expect{\cS_\logpart(\nat, \E[\nat])} \; .
+}
+where the last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
+Note that the middle covariance formulation is actually the one  used by \citet{hanzely2018fastest}.
+Now, \cref{eq:bias-variance} defines the variance as 
+\aligns{
+	\expect{\bregmanconj(\tilde \m_n , \hat\m_n} \;,
+	\expect{\bregman(\hat\nat_n , \E[\hat\nat_n]} \;,
+}
+where expectations are on all samples $X_1, \dots, X_n$.
+
+\alignn{
+	\expect[g]{\cS_\conj \paren{
+			\hat\m_{n} - \lr g(\hat \nat_{n}),
+			\hat\m_{n}   -\lr \nabla f(\hat \nat_{n})
+	}} &= \expect{\lin{
+		\hat \nat_{n+1}  - \nabla \conj(\hat\m_{n}   -\lr \nabla f(\hat \nat_{n}))\ , \
+	  	-\lr( g(\hat \nat_{n})    - \nabla f(\hat \nat_{n}) )
+	}} \\
+	&=\expect{\lin{
+		\hat \nat_{n+1}\ , \
+	  	-\lr( g(\hat \nat_{n})    - \nabla f(\hat \nat_{n}) )
+	}}\\
+	&=\Cov_g(\hat \nat_{n+1}, -g(\hat \nat_{n}))
+	= \Cov_g(\hat \nat_{n+1}, \hat\m_{n+1})
+}
+where the last two lines are due to $\E[g] = \nabla f$.
 

### 2021-10-21 aistats.tex
Nocite Dragomir

@@
 
+%\usepackage{microtype} % RLP tried it, no change.
 %\usepackage[utf8]{inputenc} % allow utf-8 input
@@
 
+\nocite{dragomir2021optimal}
 \newpage
@@
 \aligns{
-	\expect{\bregmanconj(\tilde \m_n , \hat\m_n} \;,
-	\expect{\bregman(\hat\nat_n , \E[\hat\nat_n]} \;,
+	\expect{\bregmanconj(\tilde \m_n , \hat\m_n}
+	=\expect{\bregman(\hat\nat_n , \E[\hat\nat_n]} \;,
 }
 where expectations are on all samples $X_1, \dots, X_n$.
-
-\alignn{
-	\expect[g]{\cS_\conj \paren{
-			\hat\m_{n} - \lr g(\hat \nat_{n}),
-			\hat\m_{n}   -\lr \nabla f(\hat \nat_{n})
-	}} &= \expect{\lin{
-		\hat \nat_{n+1}  - \nabla \conj(\hat\m_{n}   -\lr \nabla f(\hat \nat_{n}))\ , \
-	  	-\lr( g(\hat \nat_{n})    - \nabla f(\hat \nat_{n}) )
-	}} \\
-	&=\expect{\lin{
-		\hat \nat_{n+1}\ , \
-	  	-\lr( g(\hat \nat_{n})    - \nabla f(\hat \nat_{n}) )
-	}}\\
-	&=\Cov_g(\hat \nat_{n+1}, -g(\hat \nat_{n}))
-	= \Cov_g(\hat \nat_{n+1}, \hat\m_{n+1})
-}
-where the last two lines are due to $\E[g] = \nabla f$.
 

### 2021-10-21 aistats.tex
Update erratum

@@
 \begin{important}
-{\bf Erratum:} There was an error in our derivations. In the main text, we claimed that the log-partition of a Gaussian was not self-concordant. We will now prove that the opposite is true.
+	{\bf Erratum:}  
+	In the main text, we claimed that the log-partition of a Gaussian was not self-concordant. 
+	There was an error in our derivations and we will now prove the opposite.
+	However this does not change our story, and in particular our concern about the lack of results in the few sample regime remains valid.
 \end{important}

### 2021-10-21 aistats.tex
F: SMD

--- 2021-10-21 aistats.tex
@@
 
-%\usepackage{microtype} % RLP tried it, no change.
 %\usepackage[utf8]{inputenc} % allow utf-8 input
@@
 \newcommand{\logpart}{A}
-\newcommand{\conj}{{\logpart^*}}
+\newcommand{\conj}{\logpart^*}
 \newcommand{\bregman}{\cB_\logpart}
@@
 	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
-	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:bias-variance} from~\S\ref{ssec:bias-variance}.
+	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:primal_pivot} from~\S\ref{ssec:bias-variance}.
 	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
@@
 	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
-	\label{eq:bias-variance}
+	\label{eq:primal_pivot}
 \end{equation}
@@
 However, this is not helpful, as verifying the assumption becomes as hard as the initial problem.
-For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper bound on the variance term of~\eqref{eq:bias-variance} (cf App.~\ref{app:bias-variance}).
+For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper bound on the variance term of~\eqref{eq:primal_pivot} (cf App.~\ref{app:bias-variance}).
 Confronted with this difficulty, we investigate an alternative definition of variance.
@@
 
-\nocite{dragomir2021optimal}
 \newpage
@@
 	Fortunately, there is a trick to express it as an integral itself , with an exponential that fits well with the $\chi^2$-density,
+	\fdk{I don't follow from there.} \rlp{do you have more detailed feedback ? I don't know how to improve it.}
 \alignn{
@@
 In this section, we prove that for a Gaussian, the entropy and log-partition functions are self-concordant.
-% TODO We also give complementary illustrations on the Gaussian mirror map and MLE.
+%We also give complementary illustrations on the Gaussian mirror map and MLE.
 
@@
 where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
-% TODO These functions and their domains are represented in Figure \TODO{}.
+%These functions and their domains are represented in Figure \TODO{}.
 We provide definitions of self-concordance in \cref{app:self-concordant}.
@@
 \label{app:bias-variance}
-In this section, we start from the notions of bias and variance introduced in \cref{eq:bias-variance}.
-First, we prove that the bias of the MLE of a Gaussian variance decreases in $O(\inv{n^2})$. 
-Then we prove that the expectation of \cref{eq:hanzely} is an upper bound on the variance term.
-
-\subsection{Bias of a Gaussian Variance MLE}
+
 \rlp{prove that the bias decreases like $1/n^2$ for gaussian variance and gaussian mean.}
 
-\subsection{Expectation of SMD's Variance Assumption}
-The first step is to notice the symmetrized Bregman can be expressed as an inner product between primal and dual parameters
-\alignn{
-	\cS_\conj (\mu , \bar\mu)
-	= \lin{ \nabla \conj(\mu) - \nabla \conj(\bar\mu) , \mu - \bar\mu}
-	= \lin{ \nat -\bar\nat , \mu - \bar\mu} \; .
-}
-Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_n = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_n =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
-This means that
-\alignn{
-	\expect{\cS_\conj (\mu , \E[\mu])} 
-	= \expect{\lin{\nat, \mu - \E[\mu]}}
-	=\Cov(\nat, \mu)
-	=\expect{\lin{\nat - \E[\nat],\mu}} 
-	= \expect{\cS_\logpart(\nat, \E[\nat])} \; .
-}
-where the last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
-Note that the middle covariance formulation is actually the one  used by \citet{hanzely2018fastest}.
-Now, \cref{eq:bias-variance} defines the variance as 
-\aligns{
-	\expect{\bregmanconj(\tilde \m_n , \hat\m_n}
-	=\expect{\bregman(\hat\nat_n , \E[\hat\nat_n]} \;,
-}
-where expectations are on all samples $X_1, \dots, X_n$.
+\rlp{prove that the expectation of Hanzely is the symmetrized bias, e.g.  an upper bound on the bias.}
 
@@
 
-\rlp{Introduce SMD, in primal (minimization) and dual formulations.}
+We use this section to give more details on the (stochastic) mirror descent algorithm.
+One perspective on gradient descent with step-sizes, 
+the update $\theta_{n+1} = \theta_n - \gamma \nabla f(\theta_n)$ is to view it 
+as the minimization of the linear approximation of $f$ at $\theta_n$, 
+$f(\theta) \approx f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n}$
+and a quadratic penalty scaled by ${1}/{\gamma}$, 
+\alignn{
+	\theta_{n+1} = \arg\min_\theta f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n} + \frac{1}{\gamma} \frac{1}{2}\norm{\theta - \theta_n}^2.
+}
+Mirror descent generalizes the above, using the Bregman divergence induced by a (Legendre) function $A$ instead of the Euclidean norm, 
+\alignn{\label{eq:mirror-descent-primal}
+	\theta_{n+1} = \arg\min_\theta f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n} + \frac{1}{\gamma} {\theta - \theta_n}^2.
+}
+Mirror descent coincides with gradient descent if $A(\theta) = \frac{1}{2}\norm{\theta}^2$.
+As \cref{eq:mirror-descent-primal} is convex, the minimum is at a stationary point, 
+found by taking the derivative and setting to 0, 
+leading to the update $\theta_{n+1}$ satisfying
+\alignn{
+	\nabla f(\theta_n) + \frac{1}{\gamma}\paren{\nabla A(\theta_{n+1}) - \nabla A(\theta_n)} = 0
+	&&\implies&&
+	\nabla A(\theta_{n+1}) = \nabla A(\theta_n) - \gamma \nabla f(\theta_n).
+}
+Expressed with the dual parameters $\mu_{n} = \nabla A(\theta_{n})$, 
+$\mu_{n+1} = \mu_n - \gamma \nabla f(\theta_n)$.
+
+{\bf In our case,} where the objective function is the (negative) log-likelihood of an exponential family, 
+\alignn{
+	f(\theta) = A(\theta) - \lin{\frac{1}{n} \sum_{i=1}^n T(X_i), \theta},
+}
+mirror descent with a step-size of $1$ and the log-partition function $A$ as the reference function is exact;
+\alignn{\begin{aligned}
+	\mu_{n+1} 
+	= \mu_n - \nabla f(\theta_n)
+	= \mu_n - \paren{\nabla A(\theta_n) - \frac{1}{n}\sum_{i=1}^n T(x_i)}
+	= \frac{1}{n}\sum_{i=1}^n T(x_i).
+	\end{aligned}
+}
+In a stochastic, online version 
+where the linearization of the objective is obtained from iid samples, 
+a decreasing step-size of $\gamma_n = 1/n$ recovers the ``online'' estimate of the MLE.
+The case of $\mu_1 = T(x_1)$ follows from the above, 
+and in general, assuming it holds for $\mu_n$, 
+\alignn{\begin{aligned}
+	\mu_{n+1} &= \mu_n - \gamma_n g(\theta_n)
+	= \mu_n - \gamma_n (\mu_n - x_n)
+	\\
+	&= (1 - \frac{1}{n}) \mu_n + \frac{1}{n} T(x_n)
+	=
+	\frac{n-1}{n} \frac{1}{n-1}\sum_{i=1}^{n-1} T(x_i)
+	= 
+	\frac{1}{n} \sum_{i=1}^{n} T(x_i)
+	+ \frac{1}{n} T(x_n).
+\end{aligned}}
+The derivation in the main text gives the more general result, 
+of using step-sizes of the form $1/(n+n_0)$ to recover online MAP estimation 
+with a conjugate prior depending on $n_0$ and the initial estimate of the parameters $\theta_0$.
 

### 2021-10-21 aistats.tex
resolve comments, more verbose on Nesterov's notation for self-concordance of gaussian entropy

@@
 	It is a barrier near $0$, and admits an upper bound (in grey) despite being poorly approximated by quadratics.
-	% fdk: Text said that it admits an the grey upper bound because it is a barrier. Don't follow the logical implication.
 	}
@@
  	&< p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{d + 1}{4n(n - d - 1)}\\
- 	\intertext{\fdk{Isn't this =?
- 	For the first term, $\inv{n-d-1} - \inv{2n} = \frac{2n}{2n(n-d-1)} - \frac{n-d-1}{2n(n-d-1)} = \frac{n+d+1}{2n(n-d-1)}$ and the second is $\times d/d$?
- 	}}
  	&= p \frac{n + d +1}{2 n (n-d-1)}  + \frac{d(d + 1)/2}{2dn(n - d - 1)}\\
@@
 	This $1/(a+x)$ is hard to deal with.
-	Fortunately, there is a trick to express it as an integral itself , with an exponential that fits well with the $\chi^2$-density,
+	Fortunately, there is a trick to express it as an integral itself, with an exponential that fits well with the $\chi^2$-density,
 \alignn{
@@
 	assuming we can exchange the order of integration.
-	%\fdk{Fubini's theorem would be a justification here, right? Are you trying to say ``assuming we can exchange the order of integration''?}\rlp{yes exactly.}
 	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
@@
 		\iff \expect{\frac{\mu^*}{\MAPm}} 
-		&\leq \frac{n +n_0}{a + n - 2}
+		&\leq \frac{n +n_0}{a + n - 2}.
 	\end{align}
@@
 \begin{align}
-	\expect{\inv{a + X^2}} < \inv{a}
+	\expect{\inv{a + X^2}} < \inv{a}.
 	\label{eq:trivial_bound}
@@
 \paragraph{Entropy is Self-Concordant,}
-\citet[Example 4.1.1.4,   p.177]{nesterov2003introductory} proves that logarithmic barriers for a second order region are self-concordant.
-Adopting their notation, $\conj$ fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = ( 1 \ 0 )^\top$.
+\citet[Example 4.1.1.4,   p.177]{nesterov2003introductory} proves that logarithmic barriers for second order regions are self-concordant, 
+that is functions of the form 
+\alignn{
+	f(\theta) = -\log\paren{\alpha + \lin{a, \theta} - \frac{1}{2}\lin{\mA \theta, \theta}}
+	\text{ on }
+	\left\{\theta\in\mathbb{R}^n \cond \alpha + \lin{a, \theta} - \frac{1}{2}\lin{\mA \theta, \theta} > 0\right\}.
+}
+The entropy fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = ( 0 \ 1 )^\top$.
 
@@
 	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ 
-	%and its dual local norm $\|\cdot\|_x^*= \sqrt{ \langle [\nabla^2A^*(x)]^{-1}\cdot,\, \cdot \rangle }$,
-%	\fdk{The second one should be 
-%	$\sqrt{ \langle [\nabla^2A^*({\color{red} \nabla A(x)})]^{-1}\cdot,\, \cdot \rangle }$?
-%	}
 	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.

### 2021-10-21 aistats.tex
typo fix

@@
 	&=\inv{2n^2} \sum_i \E \norm{T_i - \m^* }_\mF^2 
-	+ \inv{2n^2} \sum_{i\neq j}\expect{T_i - \m^*}^\top \mF \overbrace{\expect{T_j - \m^*}}^{0} ]\\
+	+ \inv{2n^2} \sum_{i\neq j}\expect{T_i - \m^*}^\top \mF \overbrace{\expect{T_j - \m^*}}^{0}\\
 	&=\inv{2n}  \E \norm{T_1 - \m^* }_\mF^2 \\

### 2021-10-21 aistats.tex
Complete connection Hanzely - variance

--- 2021-10-21 aistats.tex
@@
 	However this does not change our story, and in particular our concern about the lack of results in the few sample regime remains valid.
+	%TODO ostroskii and bach are using a subgaussian assumption which does not hold for a Gaussian, so there might still be work to do. But also check anastasiou to be sure.
 \end{important}
@@
 First, we prove that the bias of the MLE of a Gaussian variance decreases in $O(\inv{n^2})$. 
-Then we prove that the expectation of \cref{eq:hanzely} is an upper bound on the variance term.
+Then we prove that assuming~\eqref{eq:hanzely} holds, whether uniformly or in expectation, yields a convergence rate on the variance term.
 
@@
 }
-Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_n = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_n =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
+Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_{n+1} = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_{n+1} =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
 This means that
@@
 Now, \cref{eq:bias-variance} defines the variance as 
-\aligns{
-	\expect{\bregmanconj(\tilde \m_n , \hat\m_n}
-	=\expect{\bregman(\hat\nat_n , \E[\hat\nat_n]} \;,
-}
-where expectations are on all samples $X_1, \dots, X_n$.
+\alignn{
+	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
+	=\expect[1:n]{\bregman(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
+	\leq \expect[1:n]{\cS_A(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
+	=  \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }\;,
+	\label{eq:bregman-covariance}
+}
+where expectations $\E_{1:n}$ are on all samples $X_1, \dots, X_n$, whereas~\eqref{eq:hanzely} is written with 
+\alignn{
+	\expect[n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) } \leq \lr^2 C \; ,
+}
+where the expectation $\E_{n}$ is taken over only the last sample $X_n$, and the bound should hold uniformly over all $\hat\mu_{n-1}$.
+Taking the expectation over $\hat\mu_{n-1}$ instead yields
+\alignn{
+	\expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \leq \lr^2 C \;.
+}
+The only difference with the RHS of \cref{eq:bregman-covariance} is in the inner expectation.
+To overcome this difference, we need to plug in the form of $\hat \mu_n = \frac{n_0\mu_0 + \sum_i T_i}{n_0 +n}$. Notice that 
+\alignn{
+	&\MAPm - \E_n[\MAPm] = \frac{T_n - \m^*}{n_0+n} \\
+	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }
+	= \inv{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} }
+}
+while
+\alignn{
+	&\MAPm - \E_{1:n}[\MAPm] = \frac{\smallsum_i (T_i - \m^*)}{n_0+n} \\
+	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }
+	=\frac{n}{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} } \; .
+}
+In the end we get that the variance is dominated by $n$ times the expectation of \cref{eq:hanzely}
+\alignn{
+	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
+	\leq n \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \; .	
+}
+If assumption~\eqref{eq:hanzely} holds, then
+\alignn{
+\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
+\leq n \lr_n^2 C \in O\paren{\inv{n}} \; ,
+}
+where we assumed $\lr_n  \in O\paren{\inv{n}}$.
+In conclusion, assuming~\eqref{eq:hanzely} holds uniformly or in expectation immediately implies a $O\paren{\inv{n}}$ convergence rate on the variance.
 
@@
 
-We use this section to give more details on the (stochastic) mirror descent algorithm.
-One perspective on gradient descent with step-sizes, 
-the update $\theta_{n+1} = \theta_n - \gamma \nabla f(\theta_n)$ is to view it 
-as the minimization of the linear approximation of $f$ at $\theta_n$, 
-$f(\theta) \approx f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n}$
-and a quadratic penalty scaled by ${1}/{\gamma}$, 
-\alignn{
-	\theta_{n+1} = \arg\min_\theta f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n} + \frac{1}{\gamma} \frac{1}{2}\norm{\theta - \theta_n}^2.
-}
-Mirror descent generalizes the above, using the Bregman divergence induced by a (Legendre) function $A$ instead of the Euclidean norm, 
-\alignn{\label{eq:mirror-descent-primal}
-	\theta_{n+1} = \arg\min_\theta f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n} + \frac{1}{\gamma} {\theta - \theta_n}^2.
-}
-Mirror descent coincides with gradient descent if $A(\theta) = \frac{1}{2}\norm{\theta}^2$.
-As \cref{eq:mirror-descent-primal} is convex, the minimum is at a stationary point, 
-found by taking the derivative and setting to 0, 
-leading to the update $\theta_{n+1}$ satisfying
-\alignn{
-	\nabla f(\theta_n) + \frac{1}{\gamma}\paren{\nabla A(\theta_{n+1}) - \nabla A(\theta_n)} = 0
-	&&\implies&&
-	\nabla A(\theta_{n+1}) = \nabla A(\theta_n) - \gamma \nabla f(\theta_n).
-}
-Expressed with the dual parameters $\mu_{n} = \nabla A(\theta_{n})$, 
-$\mu_{n+1} = \mu_n - \gamma \nabla f(\theta_n)$.
-
-{\bf In our case,} where the objective function is the (negative) log-likelihood of an exponential family, 
-\alignn{
-	f(\theta) = A(\theta) - \lin{\frac{1}{n} \sum_{i=1}^n T(X_i), \theta},
-}
-mirror descent with a step-size of $1$ and the log-partition function $A$ as the reference function is exact;
-\alignn{\begin{aligned}
-	\mu_{n+1} 
-	= \mu_n - \nabla f(\theta_n)
-	= \mu_n - \paren{\nabla A(\theta_n) - \frac{1}{n}\sum_{i=1}^n T(x_i)}
-	= \frac{1}{n}\sum_{i=1}^n T(x_i).
-	\end{aligned}
-}
-In a stochastic, online version 
-where the linearization of the objective is obtained from iid samples, 
-a decreasing step-size of $\gamma_n = 1/n$ recovers the ``online'' estimate of the MLE.
-The case of $\mu_1 = T(x_1)$ follows from the above, 
-and in general, assuming it holds for $\mu_n$, 
-\alignn{\begin{aligned}
-	\mu_{n+1} &= \mu_n - \gamma_n g(\theta_n)
-	= \mu_n - \gamma_n (\mu_n - x_n)
-	\\
-	&= (1 - \frac{1}{n}) \mu_n + \frac{1}{n} T(x_n)
-	=
-	\frac{n-1}{n} \frac{1}{n-1}\sum_{i=1}^{n-1} T(x_i)
-	= 
-	\frac{1}{n} \sum_{i=1}^{n} T(x_i)
-	+ \frac{1}{n} T(x_n).
-\end{aligned}}
-The derivation in the main text gives the more general result, 
-of using step-sizes of the form $1/(n+n_0)$ to recover online MAP estimation 
-with a conjugate prior depending on $n_0$ and the initial estimate of the parameters $\theta_0$.
+\rlp{Introduce SMD, in primal (minimization) and dual formulations.}
 

### 2021-10-21 aistats.tex
Add gaussian variance MLE

@@
 \subsection{Bias of a Gaussian Variance MLE}
-\rlp{prove that the bias decreases like $1/n^2$ for gaussian variance and gaussian mean.}
+For a Gaussian variance model, the MLE follows a scaled $\chi^2(n)$ distribution. 
+This means that
+\alignn{
+	\frac{\m^*}{\tilde \m_n} 
+	= \frac{\tilde\nat_n}{\nat^*}
+	=\expect{\frac{\hat\nat_n}{\nat^*}}
+	=\expect{\frac{\m^*}{\hat \m_n} }
+	=\expect{\frac{n}{\chi^2(n)}}
+	= \frac{n}{n-2}\; .
+}
+Consequently
+\alignn{
+	\bregmanconj(\m^*, \tilde \m_n) 
+	&= \half \paren{\frac{n}{n-2} -1 + \log\frac{n-2}{n}}\\
+	&= \frac{1}{n-2} + \half \log\paren{1 - \frac{2}{n}}\\
+	&\leq \inv{n-2} - \inv{n} 
+	= \frac{2}{n(n-2)}
+}
+so the bias of the MLE of a Gaussian Variance decreases like $O\paren{\inv{n^2}}$.
 
@@
 \alignn{\label{eq:mirror-descent-primal}
-	\theta_{n+1} = \arg\min_\theta f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n} + \frac{1}{\gamma} {\theta - \theta_n}^2.
+	\theta_{n+1} = \arg\min_\theta f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n} + \frac{1}{\gamma} \bregman(\theta, \theta_n).
 }

### 2021-10-22 aistats.tex
Be more gentle on the covariance
Quick typo pass on Appendix + clarify erratum. RESUBMITTED

@@
 \begin{important}
-	{\bf Erratum:}  contrary to what is written in the main text, the log-partition of a Gaussian \emph{is} self-concordant. See \cref{app:gaussian} for details. 
+	{\bf Erratum:} contrary to what is written in the main text in the last paragraph of Section~\ref{ssec:gaussian} and~\ref{ssec:local-quadratic}, the log-partition of a full Gaussian \emph{is} self-concordant.
+See \cref{app:gaussian} for details.
+This does not change the narrative of the paper though.
 \end{important}
@@
 	{\bf Erratum:}  
-	In the main text, we claimed that the log-partition of a Gaussian was not self-concordant. 
-	There was an error in our derivations and we will now prove the opposite.
-	However this does not change our story, and in particular our concern about the lack of results in the few sample regime remains valid.
+	In the main text, we claimed by mistake in the last paragraph of Section~\ref{ssec:gaussian} and~\ref{ssec:local-quadratic} that the log-partition of a full Gaussian is not self-concordant. 
+	There was an error in our derivations and we will now prove the opposite (using classical results).
+	However, this does not change our main story, and in particular, the point that we made in the last paragraph of Section~\ref{ssec:local-quadratic} that there is a lack of results in the few sample regime remains valid.
 	%TODO ostroskii and bach are using a subgaussian assumption which does not hold for a Gaussian, so there might still be work to do. But also check anastasiou to be sure.
@@
 	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
-	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; ,
+	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; , \label{eq:app-entropy}
 }
@@
 
-\paragraph{Entropy is Self-Concordant,}
+\paragraph{Entropy is Self-Concordant.}
 \citet[Example 4.1.1.4,   p.177]{nesterov2003introductory} proves that logarithmic barriers for second order regions are self-concordant, 
@@
 }
-The entropy fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = ( 0 \ 1 )^\top$.
-
-\paragraph{Log-partition is Self-Concordant}
-As proved in \citet{nesterov1994interior} self-concordance is preserved by Fenchel conjugacy.
-Since $\conj$ is self-concordant, $\logpart$ as well.
+The entropy~\eqref{eq:app-entropy} fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = ( 0 \ 1 )^\top$.
+
+\paragraph{Log-partition is Self-Concordant.}
+As proved in \citet{nesterov1994interior}, self-concordance is preserved by Fenchel conjugacy.
+Since $\conj$ is self-concordant, $\logpart$ is as well.
 For a more accessible reference, see also \citet[Prop.~6]{sun2019generalized}.
@@
 
-In this section, we define self-concordance, we prove proposition~\ref{prop:selfConcordant}, and we prove that a 1 dimensional function is self-concordant if and only if its convex conjugate is self-concordant as well.
+In this section, we define self-concordance and we prove Proposition~\ref{prop:selfConcordant}.
 
@@
 
-{\bf Correction:} In the main text, we claimed that the Fenchel conjugate of a 1 dimensional function is also self-conjugate. 
-Actually, this is also true in higher dimension, as proved by \citet{nesterov1994interior}. See \citet[Prop.~6]{sun2019generalized} for a more accessible reference.
+{\bf Clarification:} In the main text in Section~\ref{ssec:local-quadratic}, we claimed that the Fenchel conjugate of a 1-dimensional function is also self-concordant. 
+Actually, this is also true in higher dimensions, as proved by \citet{nesterov1994interior}. See \citet[Prop.~6]{sun2019generalized} for a more accessible reference.
 
@@
 
-We quickly reviews some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}.
-We start with a few notations.
+We quickly review some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}.
+We start with a some notation.
 Let $A^*$ be a self-concordant function.
@@
 
-The next result show that, if $y$ is sufficiently close to $x$, then we can bound the Bregman divergence of $A^*$ using the distance function $\omega^*$ and local norms.
+The next result shows that, if $y$ is sufficiently close to $x$, then we can bound the Bregman divergence of $A^*$ using the distance function $\omega^*$ and local norms.
 \begin{proposition}(Upper bound of self-concordant functions \citep[Theorem 4.1.8]{nesterov2003introductory}) \label{prop:upper_bound_self_concordance}
@@
 	= \expect{\lin{\nat, \mu - \E[\mu]}}
-	=\Cov(\nat, \mu)
+	= \underbrace{\expect{\lin{\nat, \mu}} -\lin{\expect{\nat}, \expect{\mu} } }_{\Cov(\nat, \mu)}
 	=\expect{\lin{\nat - \E[\nat],\mu}} 

### 2021-10-22 aistats.tex
correction of typos and formulation in appendices

--- 2021-10-21 aistats.tex
@@
     = \nabla^2\logpart(\nat^*)^{-1}
-    = \Cov_{\nat^*}(T)^{-1}
+    = \Cov_{\nat^*}(T)^{-1},
 }
@@
 	& = \inv{2n} \Tr(\mF\ \expect{(T_1 - \m^*) (T_1 - \m^*)^\top}) \\
-	&= \inv{2 n} \Tr(\mF \Cov_{\nat^*}(T)) = \frac{d}{2n} 
+	&= \inv{2 n} \Tr(\mF \Cov_{\nat^*}(T)) = \frac{d}{2n} ,
 }
@@
 \end{align}
-For the MAP~\eqref{eq:defMAP}, the quadratic decomposes into bias and variance
+For the MAP~\eqref{eq:defMAP}, the quadratic decomposes into bias and variance:
 \begin{align}
@@
 First, we prove that the bias of the MLE of a Gaussian variance decreases in $O(\inv{n^2})$. 
-Then we prove that assuming~\eqref{eq:hanzely} holds, whether uniformly or in expectation, yields a convergence rate on the variance term.
+Then we prove that assuming~\eqref{eq:hanzely} holds, whether uniformly or in expectation, yields to a convergence rate on the variance term.
 
@@
 }
-Consequently
+Consequently,
 \alignn{
@@
 	&\leq \inv{n-2} - \inv{n} 
-	= \frac{2}{n(n-2)}
+	= \frac{2}{n(n-2)},
 }
@@
 }
-where the last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
+The last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
 Note that the middle covariance formulation is actually the one  used by \citet{hanzely2018fastest}.
@@
 where the expectation $\E_{n}$ is taken over only the last sample $X_n$, and the bound should hold uniformly over all $\hat\mu_{n-1}$.
-Taking the expectation over $\hat\mu_{n-1}$ instead yields
+Taking the expectation over $\hat\mu_{n-1}$ instead gives
 \alignn{
@@
 	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }
-	= \inv{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} }
+	= \inv{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} },
 }
@@
 }
-In the end we get that the variance is dominated by $n$ times the expectation of \cref{eq:hanzely}
+In the end, we get that the variance is dominated by $n$ times the expectation of \cref{eq:hanzely}:
 \alignn{
@@
 }
-If assumption~\eqref{eq:hanzely} holds, then
+If assumption~\eqref{eq:hanzely} holds, then we have
 \alignn{
@@
 We use this section to give more details on the (stochastic) mirror descent algorithm.
-One perspective on gradient descent with step-sizes, 
-the update $\theta_{n+1} = \theta_n - \gamma \nabla f(\theta_n)$ is to view it 
-as the minimization of the linear approximation of $f$ at $\theta_n$, 
-$f(\theta) \approx f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n}$
-and a quadratic penalty scaled by ${1}/{\gamma}$, 
+We start with gradient descent with step-sizes $\gamma$. 
+the update $\theta_{n+1} = \theta_n - \gamma \nabla f(\theta_n)$ can be viewed as the minimization of the linear approximation of $f$ at $\theta_n$
+$f(\theta) \approx f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n}$, 
+alongside with quadratic penalty scaled by ${1}/{\gamma}$:
 \alignn{
@@
 }
-Mirror descent generalizes the above, using the Bregman divergence induced by a (Legendre) function $A$ instead of the Euclidean norm, 
+Mirror descent generalizes the above, using the Bregman divergence induced by a (Legendre) function $A$ instead of the Euclidean norm as follow, 
 \alignn{\label{eq:mirror-descent-primal}
@@
 }
-Expressed with the dual parameters $\mu_{n} = \nabla A(\theta_{n})$, 
+Expressed with the dual parameters, we obtain $\mu_{n} = \nabla A(\theta_{n})$, 
 $\mu_{n+1} = \mu_n - \gamma \nabla f(\theta_n)$.
 
-{\bf In our case,} where the objective function is the (negative) log-likelihood of an exponential family, 
-\alignn{
-	f(\theta) = A(\theta) - \lin{\frac{1}{n} \sum_{i=1}^n T(X_i), \theta},
-}
-mirror descent with a step-size of $1$ and the log-partition function $A$ as the reference function is exact;
+{\bf In our case,} where the objective function is the (negative) log-likelihood of an exponential family, we have
+\alignn{
+	f(\theta) = A(\theta) - \lin{\frac{1}{n} \sum_{i=1}^n T(X_i), \theta}.
+}
+Using Mirror descent with a step-size of $1$ and the log-partition function $A$ as the reference function gives
 \alignn{\begin{aligned}

### 2021-10-22 aistats.tex
Add Gaussian figures

@@
 With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
-We thus consider the fundamental problem: %TODO thus
+We thus consider the fundamental problem: %TODO thus repetition
 %\begin{center}
@@
 In this section, we prove that for a Gaussian, the entropy and log-partition functions are self-concordant.
-% TODO We also give complementary illustrations on the Gaussian mirror map and MLE.
+We also provide complementary illustrations of 
+these functions in \cref{fig:logpart-entropy}, 
+their gradients (e.g. the mirror maps) in \cref{fig:mirrormaps},
+and paths taken by MLE and MAP in \cref{fig:gaussian-paths} .
 
@@
 \alignn{
-	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
+	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \label{eq:app-logpart} \\
 	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; , \label{eq:app-entropy}
@@
 where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
-% TODO These functions and their domains are represented in Figure \TODO{}.
 We provide definitions of self-concordance in \cref{app:self-concordant}.
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=.45\textwidth]{figs/2d/logpart.pdf}
+	\includegraphics[width=.45\textwidth]{figs/2d/entropy.pdf}
+	\caption{
+	(left) Contours of the log-partition function~\eqref{eq:app-logpart}.
+	(right) Contours of the entropy~\eqref{eq:app-entropy}.
+	}
+	\label{fig:logpart-entropy}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=.9\textwidth]{figs/2d/mirrormap.pdf}
+	\caption{
+	(top) Grid deformation produce by $\nabla \logpart(\nat)$.
+	(bottom) Grid deformation produced by $\nabla\conj(\m)$.
+	}
+	\label{fig:mirrormaps}
+\end{figure}
+
+\begin{figure}[ht]
+	\centering
+	\includegraphics[width=.9\textwidth]{figs/2d/gaussian_paths.pdf}
+	\caption{
+		Paths taken by MLE (blue) and MAP (orange) on top of contours for $\KL(\nat^* || \nat)$. 
+		We set $\mu^*=(0, 1), \mu_0 = (1, 2)$, and $n_0=4$, and $n$ varies from $1$ to $20$.
+		In green, red and purple, we represent the paths respectively taken by the MAP dual expectation  $(\bar\nat_n, \bar \mu_n)$, MAP primal expectation $(\tilde\nat_n,\tilde\mu_n)$, and MLE primal expectation.
+Recall that the MLE dual expectation is $\mu^*$ itself.
+	}
+	\label{fig:gaussian-paths}
+\end{figure}
 
@@
 First, we prove that the bias of the MLE of a Gaussian variance decreases in $O(\inv{n^2})$. 
-Then we prove that assuming~\eqref{eq:hanzely} holds, whether uniformly or in expectation, yields to a convergence rate on the variance term.
+Then we prove that assuming~\eqref{eq:hanzely} holds, whether uniformly or in expectation, yields a convergence rate on the variance term.
 
@@
 Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_{n+1} = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_{n+1} =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
-This means that
+For such a pair of variables, the expectation of the symmetrized Bregman corresponds to a covariance between primal and dual parameters
 \alignn{

### 2021-10-22 aistats.tex
Add figure for MAP upper bound

@@
 
+\begin{figure}[ht]
+	\centering
+\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
+	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
+		Solid curve  are average over $10^5$ trials.
+		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, tight up to a factor 2).
+		Shaded areas are 90\% confidence interval.
+		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
+		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
+	}
+	\label{fig:MAP_rate}
+\end{figure}
+
 In this section, we prove the results mentioned in \S\ref{ssec:gaussian-variance},
 and add some context and experimental observations.
+In particular, we show the bound~\eqref{eq:MAP_rate} in \cref{fig:MAP_rate}.
 Let us recall the form of the divergence with different parametrizations.

### 2021-10-26 aistats.tex
Change fig 1 and minor fixes

@@
 
-% TODO
-%\begin{figure}[t]
-%	\centering
-%\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
-%	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
-%		Solid curve  are average over $10^5$ trials.
-%		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
-%		Shaded areas are 90\% confidence interval.
-%		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
-%		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
-%		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
-%	}
-%	\label{fig:curves}
-%\end{figure}
-
 \begin{figure}[t]
 	\centering
-\includegraphics[width=.4\textwidth]{fewsamples.pdf}
-	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance MLE (\S\ref{ssec:gaussian-variance}) against number of samples $n$.
-The bold curve plots the average over 10000 trials,  dark shaded area is 90\% confidence interval, and light shade is a min-max interval.
-		The expected value is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}, while the variance remains large.
-		We wish to find similar upper bounds for a variety of exponential families.
+\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
+	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
+		Solid curve  are average over $10^5$ trials.
+		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
+		Shaded areas are 90\% confidence interval.
+		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
+		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
+		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
 	}
@@
 has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
-With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
-We thus consider the fundamental problem: %TODO thus repetition
-%\begin{center}
-%	\framebox{
-%		\parbox[c][]{0.9\linewidth}{
-%			\begin{center}
-%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
-%		\end{center}
-%		}
-%	}
-%\end{center}
+With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research,
+and raising the fundamental problem:
 \begin{equation}
@@
 \end{equation}
-% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.
-
-There are already general asymptotic results (\S\ref{ssec:asymptote}), and a finite $n$ result when $\logpart$ is quadratic (e.g., $X$ is Gaussian with known variance)
+
+There are already general asymptotic results (\S\ref{ssec:asymptote} and Fig.~\ref{fig:curves}), and a finite $n$ result when $\logpart$ is quadratic (e.g., $X$ is Gaussian with known variance)
 or close to quadratic (\S\ref{ssec:quadratic}).
-However, a  general solution for finite $n$ remains elusive. %TODO add link to figure 1
+However, a  general solution for finite $n$ remains elusive. 
 In this paper, we review partial solutions and give ideas on how to solve the problem.
@@
 \alignn{
-	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-2\nat_2) \\  % TODO remove factor 2
+	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\ 
 	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
@@
 These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
-\citep[p.177]{nesterov2003introductory} % TODO add citation to example 4.1.1.4
+\citep[p.177, example 4.1.1.4]{nesterov2003introductory}
 , but $\logpart$ is not (see App.~\ref{app:gaussian}).
@@
 
-\begin{figure}[ht]
-	\centering
-\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
-	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
-		Solid curve  are average over $10^5$ trials.
-		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, tight up to a factor 2).
-		Shaded areas are 90\% confidence interval.
-		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
-		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
-	}
-	\label{fig:MAP_rate}
-\end{figure}
-
 In this section, we prove the results mentioned in \S\ref{ssec:gaussian-variance},
 and add some context and experimental observations.
-In particular, we show the bound~\eqref{eq:MAP_rate} in \cref{fig:MAP_rate}.
 Let us recall the form of the divergence with different parametrizations.

### 2021-10-27 aistats.tex
Minimal change to incorporate the caveat

@@
 
-These functions are neither smooth, nor strongly convex, and $\conj$ is self-concordant
-\citep[p.177, example 4.1.1.4]{nesterov2003introductory}
-, but $\logpart$ is not (see App.~\ref{app:gaussian}).
+These functions are neither smooth, nor strongly convex, but they are self-concordant, since $\conj$ is  the logarithmic barrier of a quadratic domain
+\citep[p.177, example 4.1.1.4]{nesterov2003introductory}, and self-concordance is preserved by convex-conjugacy~\citep{nesterov1994interior} -- see more details in App.~\ref{app:gaussian}.
 We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.
@@
 
-The entropy $\conj$ is self-concordant for several common families.
-For instance Gaussians (\S\ref{ssec:gaussian}), or when $T$ is 1-dimensional and $\logpart$ is self-concordant (App.~\ref{app:self-concordant}),
-as is the case with:
+The entropy $\conj$ is self-concordant for several common families such as Gaussians (\S\ref{ssec:gaussian}). %or gamma ?
+Since self-concordance is preserved by convex-conjugacy~\citep{nesterov1994interior}, we can add to the list:
 exponential distributions,
@@
 Pareto with known minimum value,
-or Weibull with known shape $k$.
+Weibull with known shape $k$.
 The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g., categorical and Dirichlet distributions.
@@
 
-More generally, these results answer~\eqref{problem} only \emph{partially} for the following reasons.
-First, the present proposition and these related works all give \textit{large sample} results,  that hold when $n\geq N$ for some constant $N$. 
-None of them apply to small $n$.
-Second, they do not apply to some standard exponential families like the full Gaussian example that we saw in \S\ref{ssec:gaussian}  for which $f$ is \emph{not} self-concordant. 
-A full solution to~\eqref{problem} would address these caveats.
+More generally, the present proposition and these related works answer~\eqref{problem} only \emph{partially}, as they all give \textit{large sample} results, that hold when $n\geq N$ for some constant $N$.
+A full solution to~\eqref{problem} would apply to small $n$.
 Informed by the properties that we have seen so far, we next investigate a general decomposition of the Bregman that could guide us towards a solution.
@@
 \part{Appendix} % Start the appendix part
-\begin{important}
-	{\bf Erratum:} contrary to what is written in the main text in the last paragraph of Section~\ref{ssec:gaussian} and~\ref{ssec:local-quadratic}, the log-partition of a full Gaussian \emph{is} self-concordant.
-See \cref{app:gaussian} for details.
-This does not change the narrative of the paper though.
-\end{important}
 \parttoc % Insert the appendix TOC
@@
 
-\begin{important}
-	{\bf Erratum:}  
-	In the main text, we claimed by mistake in the last paragraph of Section~\ref{ssec:gaussian} and~\ref{ssec:local-quadratic} that the log-partition of a full Gaussian is not self-concordant. 
-	There was an error in our derivations and we will now prove the opposite (using classical results).
-	However, this does not change our main story, and in particular, the point that we made in the last paragraph of Section~\ref{ssec:local-quadratic} that there is a lack of results in the few sample regime remains valid.
 %TODO ostroskii and bach are using a subgaussian assumption which does not hold for a Gaussian, so there might still be work to do. But also check anastasiou to be sure.
-\end{important}
 

### 2021-10-28 main.tex
Scribble gaussian concentration bound

@@
 \]
-Assume $\|\mu^*-\mu_n\|_{\mu^*} < 0.21$, then
+\begin{proposition}
+Assume $\conj$ is self-concordant and  $\|\mu^*-\mu\|_{\mu^*} < 0.21$, then
 \[
-    \bregmanconj(\mu^*,\mu_n) \leq \|\mu^*-\mu_n\|_{\mu^*}^2
+    \bregmanconj(\mu^*,\mu) \leq \|\mu^*-\mu\|_{\mu^*}^2 \; .
 \]
-so we have a local quadratic behavior. To put into euclide's world. Problem is it requires global self-concordance.
-
-\paragraph{Discarded}
-However,
-\begin{align*}
-    \|\mu^*-\mu_n\|^2_{\mu_n} & = \left\|\frac{n+n_0-1}{n+n_0}(\mu^*-\mu_{n-1}) + \frac{1}{n+n_0}(\mu^*-T_n)\right\|^2_{\mu_n},\\
-    & = \left\|\frac{n+n_0-1}{n+n_0}(\mu^*-\mu_{n-1})\right\|_{\mu_n}^2 +  \left\|\frac{1}{n+n_0}(\mu^*-T_n)\right\|^2_{\mu_n} + \frac{2(n+n_0+1)}{(n+n_0)^2}\langle \nabla^2 [A(\mu_n)](\mu^*-\mu_{n-1}),\,(\mu^*-T_n) \rangle,\\
-\end{align*}
-and
-\begin{align*}
-    \|\mu^*-\mu_n\|_{\mu_n} & \leq  \frac{n+n_0-1}{n+n_0}\left\|\mu^*-\mu_{n-1}\right\|_{\mu_n} + \frac{1}{n+n_0}\left\|\mu^*-T_n\right\|_{\mu_n} .
-\end{align*}
-Moreover, by the bounded Hessian change,
-\[
-    \left\|\mu^*-\mu_{n-1}\right\|_{\mu_n} \leq \frac{1}{1-\left\|\mu_n-\mu_{n-1}\right\|_{\mu_{n-1}}}\left\|\mu^*-\mu_{n-1}\right\|_{\mu_{n-1}}
-\]
+\end{proposition}
+so we have a local quadratic behavior. We can most likely get a lower bound as well -- e.g. a quadratic sandwich.
+We would like to use a concentration bound, either on $\hat \mu_n$ around $\bar \mu_n$ or $\tilde \mu_n$ in norm, or on $\|\mu^*- \hat \mu_n\|_{\mu^*}$ above $\frac{d}{2n} + \text{bias}$.
+Then we could get a high probability bound on $\bregmanconj(\mu^*,\mu) $ after seeing a given number of samples.
+
+Concentration bounds with Mahalanobis bounds for vectors that are not sub-gaussian are not easy to find.
+In the Gaussian example, $T(x)=(x, x^2)$, and $x^2$ follows a gamma distribution.
+The framework of sub-exponential or sub-gamma variables may be more suited.
+Let us use a union bound to get something easier to deal with. But first we want an $\infty$ norm.
+Straightforward approach does not work
+\alignn{
+	\norm{\mu}_{\mM^\top \mM} = \norm{\mM \mu}_2
+	\geq \norm{\mu}_\infty \min_{\norm{\nu}_\infty \geq 1} \norm{\mM \nu}_2
+}
+as this minimization is non-convex and impossible to deal with.
+Using the operator norm of the matrix inverse is better
+\alignn{
+	\norm{\mu}_\infty = \norm{\mM^{-1}\mM \mu}_\infty
+	&\leq \norm{\mM^{-1}}_{2\rightarrow\infty} \norm{\mM\mu}_2
+	= \norm{\mM^{-1}}_{2\rightarrow\infty} \norm{\mu}_{\mM^\top\mM} \\
+	\implies
+	\norm{\mF^{-\half}}_{2\rightarrow\infty}^{-1} \norm{\mu}_\infty 
+	&\leq \norm{\mu}_\mF
+}
+Now denote $c:= \norm{\mF^{-\half}}_{2\rightarrow\infty}$ where the norm $2\rightarrow \infty$ is easy to compute as the maximum $\ell^2$ norm of a row.
+We are ready to enter the realm of probabilities
+\alignn{
+	\Prob(\norm{\mu}_\mF >t)
+	&\leq \Prob\paren{\norm{\mu}_\infty >c t}
+	= \Prob\paren{\bigcup_i \{\mu_i >c t \} \cup \{-\mu_i >c t \} }
+	\leq \sum_i \Prob(\mu_i > c t) + \Prob(\mu_i < - c t)
+}
+where the last inequality is a \textbf{union bound}.
+For MAP, 
+\alignn{
+	\mu =  \frac{n_0 (\mu_0 - \mu^*) + \sum_j ( T_j - \mu^*) }{n_0 + n} 
+}
+which means that $\mu_i >ct$ rewrites into
+\alignn{
+	\sum_j ( T_{j,i} - \mu_i^*) > (n_0 + n)\paren{ct - n_0 (\mu_{0,i} - \mu_i^*) }
+}
+and bounding the deviation of a sum of real valued random variables with zero mean is what concentration bounds typically do.
+Assuming $T_{j,i}$ is sub-gamma, we may obtain some nice results.
+We just need to assess the value of sub-gamma parameters if the data is really gaussian.
 

### 2021-10-29 aistats.tex
No break bibliography
Add acknowledgements

@@
 
-\nocite{dragomir2021optimal}
-\newpage
+\subsubsection*{Acknowledgments}
+We express our gratitude to Reza Babanezhad, Radu Dragomir and Hadrien Hendrikx for their insights on stochastic mirror descent. 
+This work was partially supported by the Canada CIFAR AI Chair Program. 
+Simon Lacoste-Julien is a CIFAR Associate Fellow in the Learning in Machines \& Brains program.
+
 \bibliographystyle{apalike}

### 2021-10-29 aistats.tex
Update aistats.tex
Introduce Gamma distributions in examples

@@
 Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights an explicit $O(\frac{v}{n} + \frac{b}{n^2})$ variance-bias decomposition.
-This inequality was derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus was more tractable.
+This inequality is derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus is more tractable.
 This explains why the variance term is twice larger than the asymptote~\eqref{eq:asymptote}.
@@
 This correlates well with our numerical observations (cf App.~\ref{app:gaussian-variance}).
+
+Note that if $X\sim \cN(0,\sigma^2)$, then $X^2 \sim \Gamma\paren{\half, \inv{2\sigma^2}}$ in the shape-rate parametrization of Gamma distributions.
+In fact the bounds above can be generalized to any distribution $\Gamma\paren{\alpha, \beta}$ with known shape $\alpha$, but we leave it to \cref{app:gaussian-variance} for the sake of clarity.
 
@@
 
-The entropy $\conj$ is self-concordant for several common families such as Gaussians (\S\ref{ssec:gaussian}). %or gamma ?
-Since self-concordance is preserved by convex-conjugacy~\citep{nesterov1994interior}, we can add to the list:
+The entropy $\conj$ is self-concordant for several common families such as Gaussians (\S\ref{ssec:gaussian}), and all families with $\logpart \approx -\log$ such as
 exponential distributions,
@@
 Pareto with known minimum value,
-Weibull with known shape $k$.
+or Weibull with known shape $k$.
 The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g., categorical and Dirichlet distributions.

### 2021-10-29 aistats.tex
Introduce gamma in appendix A

@@
 \usepackage{array}
+\usepackage{wrapfig}
 
@@
 and add some context and experimental observations.
-Let us recall the form of the divergence with different parametrizations.
-For a Gaussian with known mean $0$ and covariance $\sigma^2$, 
-natural and mean parameters $\theta, \m$;
-\begin{align}
-	\KL( \sigma_*^2 ; \sigma_n^2 ) 
-	&= \half \left ( \frac{\sigma_*^2}{ \sigma_n^2} - 1 - \log \frac{\sigma_*^2}{ \sigma_n^2} \right)
-	= \phi\paren{\frac{\sigma_*^2}{ \sigma_n^2}},
-	\\
+As mentioned in the main text, the centered gaussian $\cN(0,\sigma^2)$ has sufficient statistic $x^2$ following a Gamma distribution $\Gamma\paren{\half, \inv{2 \sigma^2}}$. 
+In this section we derive results for samples  from a general gamma distribution $X \sim\Gamma(\alpha,\beta)$ with known shape parameter $\alpha$, but unknown rate parameter $\beta$.
+Results for the Gaussian follow by taking $\alpha=\half$.
+
+The density of $\Gamma(\alpha,\beta)$ reads
+\alignn{
+	p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} \; .
+}
+When $\alpha$ is known, it can be cast as an exponential family~\eqref{eq:def_expfamily} with sufficient statistic $T(x)=x$, domain $\cX = \real_+$ and base measure $\nu(x)\propto x^{\alpha-1}$.
+Then the natural parameter is $\nat = -\beta < 0$ and the log-partition function is
+\alignn{
+	\logpart (\nat) = -\alpha \log(-\nat)  + \log\Gamma(\alpha) \; .
+}
+From there we find that the mean parameter is $\mu = \frac{\alpha}{-\nat} >0$ and the entropy has the same form as the log-partition $\conj(\mu) = - \alpha \log(\mu) + \cst$.
+This means that the primal and dual Bregman divergences have the same form as well
+\begin{align}
 	\bregmanconj(\m_*; \m_n) 
-	&= \half \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) 
-	= \phi\paren{\frac{\m_*}{ \m_n}},
+	&= \alpha \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) 
+	= \alpha \phi\paren{\frac{\m_*}{ \m_n}},
 	\\
 	\bregman( \nat_n; \nat_* ) 
-	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right)
-	= \phi\paren{\frac{ \nat_n}{\nat_*}},
-\end{align}
-where all these lines are equal, and $\phi$ measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
-\begin{align}
-	\phi(z) := \half (z - 1 - \log(z)),
+	&=  \alpha \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right)
+	= \alpha \phi\paren{\frac{ \nat_n}{\nat_*}},
+\end{align}
+\begin{wrapfigure}[14]{r}{0.4\textwidth}
+	\centering
+	\includegraphics[width=.38\textwidth]{phi.pdf}
+	\caption{%
+	Illustration of $\phi(z)$ (orange) and its upper bound $\phi(z) + \phi\paren{z^{-1}}$ (grey).
+	They both are barriers near $0$.
+	}
+	\label{fig:phi}
+\end{wrapfigure}
+where these 2 lines are equal, and $\phi$ measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
+\begin{align}
+	\phi(z) := z - 1 - \log(z),
 \end{align}
@@
 	\leq \cS_{\conj} (\mu_*, \mu_n )
-	:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) 
-	= \half \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 },
+	&:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \nonumber\\
+	&=  \alpha \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 },
 	\label{eq:symmetrized_bregman}
@@
 	\label{eq:log_bound}
-	\phi(z) \leq \phi(z) + \phi\paren{\inv{z}}
-	&=  \half \paren{z -1  + \inv{z} -1} 
-	= \frac{(z-1)^2}{z} \; .
-}
-See the grey upper bound in Figure~\ref{fig:phi}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.4\textwidth]{phi.pdf}
-	\caption{%
-	$\phi(z)$ is the Bregman divergence induced by $-\log(z)$. 
-	It is a barrier near $0$, and admits an upper bound (in grey) despite being poorly approximated by quadratics.
-	}
-	\label{fig:phi}
-\end{figure}
+	\phi(z) \leq \phi(z) + \phi\paren{z^{-1}}
+	&=  z -1  + \inv{z} -1
+	= \frac{(z-1)^2}{z} \; ,
+}
+and as illustrated by the grey upper bound in Figure~\ref{fig:phi}.
 

### 2021-10-29 aistats.tex
Express MLE upper bound for Gamma

@@
 In this section we derive results for samples  from a general gamma distribution $X \sim\Gamma(\alpha,\beta)$ with known shape parameter $\alpha$, but unknown rate parameter $\beta$.
-Results for the Gaussian follow by taking $\alpha=\half$.
+Results for the Gaussian follow by taking $\alpha=\half$,
+ and we also immediately have results for exponential distributions by taking $\alpha=1$.
 
@@
 	\begin{theorem}[MLE Upper Bound]
-	The MLE of $\cN(0,\mu_*)$ is $\mu_n = \inv{n} \sum_i X_i^2 $.
-	The expected KL between $\mu_*$ and $\hat\mu_n$	is infinite when $n\leq 2$ and otherwise upper bounded by
+	For a Gamma $\Gamma(\alpha, \beta)$ with known shape $\alpha$,
+	the expected KL between $\mu_*$ and the MLE $\hat\mu_n$	is infinite when $\alpha n\leq 1$ and otherwise upper bounded by
 	\begin{align}
 		 \expect{\bregmanconj( \mu_*; \hat \mu_n) }
-			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
+			\leq \inv{2n} + \frac{1}{n(n \alpha-1)}  \; .
 	\end{align}
@@
 	\begin{proof}
-	As $\hat \mu_n = \inv{n} \sum_i X_i^2$ 
-	with $X_i \sim \cN(0, \mu_*)$, the ratio to the optimum $n \frac{\hat \mu_n}{\mu_*}$
-	follows a \href{https://en.wikipedia.org/wiki/Chi-square_distribution}{Chi-square distribution} with $n$ degrees of freedom, $\chi^2(n)$.
-Its inverse $\frac{\mu_*}{\hat \mu_n} $ follows an \href{https://en.wikipedia.org/wiki/Inverse-chi-squared_distribution}{inverse Chi-square distribution} with expectation
-	\begin{align}
-		\expect[\hat \mu_n]{\frac{\mu_*}{\hat \mu_n}  - 1}  
-		= \expect[\mu_n]{n \frac{\mu_*}{n \hat \mu_n}  - 1} 
-		= \expect{\frac{n}{\chi^2(n)} - 1} =
-		\begin{cases}
-			\frac{n}{n-2} -1 = \frac{2}{n-2} \ \text{ if } n>2, \\
+	The MLE is a scaled sum of Gammas $\hat \mu_n = \inv{n} \sum_i X_i$.
+	As such it is also a Gamma with parameter $\Gamma(n \alpha, n \beta)$ and expectation $\frac{n\alpha}{n\beta} = \frac{\alpha}{\beta} = \mu_*$.
+	If we consider the ratio $\frac{\MAPm}{\mu_*}$, it is also a Gamma with parameter $\Gamma(n \alpha, n \alpha)$ 
+	Its inverse follows an inverse gamma distribution with expectation
+	\begin{equation}
+		\expect{\frac{\mu_*}{\hat \mu_n}} 
+		= \begin{cases}
+			\frac{n\alpha}{n \alpha -1 }\ \text{ if } n\alpha > 1, \\
 			+\infty \  \text{ otherwise. }
 		\end{cases}
-	\end{align}
-	There is also a closed form solution for the expected logarithm of a Chi-squared.
-	Indeed a Chi-squared is a special case of Gamma distributions, whose sufficient statistic is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a Gamma.
-The result is
-	\begin{align}
-		\expect{\log \frac{\hat \mu_n}{\mu_*}} = \psi\paren{\half[n]} - \log\paren{\half[n]},
+	\end{equation}
+	which implies that for $n\alpha>1$, 
+	\begin{align}
+		\expect{\frac{\mu_*}{\hat \mu_n}} -1  
+		= \frac{n \alpha}{n\alpha -1 } - 1
+		= \frac{1}{n \alpha -1}
+	\end{align}
+	There is also a closed form solution for the expected logarithm of a Gamma.
+	Indeed a the sufficient statistic of a Gamma is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a Gamma to get
+	\begin{align}
+		\expect{\log \frac{\hat \mu_n}{\mu_*}} 
+		= \psi\paren{n \alpha} - \log\paren{n \alpha},
 	\end{align}
 	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
-	Closed forms also exist for non-central chi-squares, see for example the note of \citep{pav2015moments}.
+	%Closed forms also exist for non-central chi-squares, see for example the note of \citep{pav2015moments}.
 	Consequently the suboptimality of the MLE has a closed form solution
@@
 	\expect{\bregmanconj( \mu_*; \hat \mu_n) }
-	&= \half \expect{\frac{\mu_*}{ \hat \mu_n} - 1 + \log \left(\frac{\hat \mu_n}{\mu_*} \right) },
+	&= \alpha \expect{\frac{\mu_*}{ \hat \mu_n} - 1 + \log \left(\frac{\hat \mu_n}{\mu_*} \right) },
 	\\
@@
 	\begin{cases}
-		\half \left ( \frac{2}{n-2} +\psi(\half[n]) - \log(\half[n]) \right )  \ \text{ if } n>2, \\
+		\alpha \paren{ \frac{1}{n\alpha - 1} + \psi\paren{n \alpha} - \log\paren{n \alpha} }, \ \text{ if } n\alpha>1, \\
 			+\infty \  \text{ otherwise. }
@@
 	\end{align}
-	Surprisingly, we need $3$ samples or more for the expected loss to be bounded and
-	the optimal value $\mu_*$ does not appear in this rate, 
-	so that the convergence rate is independent of the solution.
+	Surprisingly, for a gaussian variance where $\alpha = \half$, we need $3$ samples or more for the expected loss to be bounded.
 	When the expectation is finite, we can get a more interpretable formula using known \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function},
 	\begin{align}
-		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x}	\; ,
-	\end{align}
-	giving, for $n\geq 3$,
-	\begin{align}
-		\inv{n-2} - \inv{n}
+		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x} = - \inv{x} + \inv{2x}	\; ,
+	\end{align}
+	giving, for $n \alpha >1 $,
+	\begin{align}
+		\frac{\alpha}{n\alpha-1} - \frac{\alpha}{n\alpha}
 		&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
-		\leq \inv{n-2} - \inv{2 n},
+		\leq \frac{\alpha}{n\alpha-1} - \frac{\alpha}{n\alpha} + \frac{\alpha}{2n\alpha}
 		\\
 		\iff
-			\frac{2}{n(n-2)}
+			\frac{1}{n(n \alpha-1)}
 			&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
-			\leq \inv{2n} +\frac{2}{n(n-2)},
+			\leq \inv{2n} + \frac{1}{n(n \alpha-1)} \; ,
 	\end{align}

### 2021-10-29 aistats.tex
Derive gamma bound for natural MAP

@@
 
+For the sake of simplicity, in higher dimension we focus on Gaussian covariance estimation and avoid the general Wishart discussion. 
 In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ 
@@
 	the importance of the prior relative to the true parameters.
-	The expectation of the natural parameter of the MAP of $\cN(0,\mu_*)$ is bounded by
+	The expectation of the natural parameter of a MAP of $\Gamma(\alpha,\beta)$ is bounded by
 \begin{equation}
@@
 		\leq
-		\frac{n_0 +n}{a+(n-2)_+} \; , \forall n\geq 0 \; .
+		\frac{n_0 +n}{a+(n-\inv{\alpha} )_+} \; , \forall n\geq 0 \; .
 		\label{eq:lemma_nat_bound}
@@
 		= \expect{\frac{\MAPt}{\nat^*}} 
-		= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i^2}}  
-		= (n_0 + n) \expect{\inv{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i^2}{ \mu^*} }} 
-		= (n_0 +n) \expect{\inv{a + \chi^2(n)}} \; ,
-}
-	where $\chi^2(n) = \sum_i \frac{X_i^2}{ \mu^*}$ is a chi-square random variable of degree $n$ and $a=n_0 \frac{\mu_0}{\mu^*}$. The expectation reads
-\alignn{
- 	\expect{\inv{a + \chi^2(n)}}  
-	= \int_0^\infty \inv{a+x} \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{2^{\half[n]}\Gamma(\half[n])} dx \; .
-}
-	This $1/(a+x)$ is hard to deal with.
-	Fortunately, there is a trick to express it as an integral itself, with an exponential that fits well with the $\chi^2$-density,
-\alignn{
-		&\int_1^\infty e^{-\frac{zt}{2}} dt = \frac{2}{z} e^{-\half[z]} \\
-		\implies &\inv{z} = \half e^{\half[z]} \int_1^\infty e^{-\frac{zt}{2}} dt \\
-		\implies & \boxed{\inv{a+x} = \half e^{\frac{a+x}{2}} \int_1^\infty e^{- \frac{(a+x)t}{2}}dt  \; .}
-} 
-We can then plug this integral into the expectation and apply Fubini's theorem to get a tractable integral on the inside
-\alignn{
-		2^{\half[n]}\Gamma(\half[n]) \expect{\inv{a + \chi^2(n)}}  
-		&= \int_0^\infty \frac{x^{\frac{n}{2}-1}e^{-\half[x]} }{a+x} dx \\
-		&= \int_0^\infty dx\  x^{\frac{n}{2}-1}e^{-\half[x]} \half e^{\half[a+x]} \int_1^\infty  dt\  e^{-\half[(a+x)t]}  \\
-		&= \half e^{\half[a]} \int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty dx \  x^{\frac{n}{2}-1} e^{-\half[x t]}  \; ,
-}
-	assuming we can exchange the order of integration.
-	Now with the change of variable $x = 2\frac{y}{t}$, eg $dx = 2\frac{dy}{t}$, the inner integral becomes
-	\begin{align}
-		\expect{\inv{a + \chi^2(n)}}  
-		&= \frac{e^{\half[a]} }{2^{\half[n]+1}\Gamma(\half[n]) }\int_1^\infty dt \ e^{-\half[a t]}   \int_0^\infty 2\frac{dy}{t} \  (2\frac{y}{t})^{\frac{n}{2}-1} e^{-y} \\
-		&= \frac{e^{\half[a]} }{2\Gamma(\half[n]) } 
-		\underbrace{\int_1^\infty  \frac{e^{-\half[a t]}}{t^{\frac{n}{2}}}  dt}_{E_{\half[n]}(\half[a])  } \ 
-		\underbrace{\int_0^\infty  y^{\frac{n}{2}-1} e^{-y} dy}_{\Gamma(\half[n])}
-	\end{align}
-	so we finally get  the formula valid for all $n\geq 1$
-	\begin{align}
-		\boxed{\expect{\inv{a + \chi^2(n)}}  = \half e^{\frac{a}{2}} E_{\half[n]}(\half[a])}
-		\label{eq:hard_expectation}
-	\end{align}
-	where the {generalized exponential integral function} is defined as 
-	\begin{align}
-		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; .
-	\end{align}
-	Now our goal is to bound this function with simpler functions.
+		= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i}}  
+		= \expect{\frac{n_0+n}{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i}{ \mu^*} }} 
+		= \expect{\frac{n_0+n}{a + \Gamma(n\alpha, \alpha)}} \; ,
+}
+	where  $\sum_i \frac{X_i^2}{ \mu^*} \sim \Gamma(n\alpha, \alpha)$ is a Gamma random variable and $a=n_0 \frac{\mu_0}{\mu^*}$. 
+	Further note that 
+\alignn{
+	\expect{\frac{n_0+n}{a + \Gamma(n\alpha, \alpha)}} 
+	= \frac{n_0 + n}{a} \expect{\frac{1}{1 + \Gamma(n \alpha , a \alpha) }} \; .
+}
+This kind of integrals can be expressed with {generalized exponential integral functions}
+	\begin{align}
+		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; ,
+	\end{align}
+with \href{http://dlmf.nist.gov/8.19.E4}{the formula} \citep[Eq.~8.19.4]{DLMF}
+\alignn{
+	\expect{\frac{1}{1 + \Gamma(\alpha , \beta) }}
+	= \frac{\beta^{\alpha}}{\Gamma(\alpha)} \int_0^\infty \frac{x^{\alpha-1}e^{-\beta x} }{1+x}  dx 
+	= \beta e^{\beta} E_\alpha (\beta) \; .
+}
+Overall we get
+\alignn{
+\expect{\frac{\mu^*}{\MAPm}}  = (n_0 +n) \alpha e^{a \alpha} E_{n\alpha} ( a \alpha) 
+}
+	Now our goal is to bound this generalized exponential integral with simpler functions.
 	Fortunately, mathematicians have been working on these integrals for decades.
-	For $n \geq 2$, eg $k\geq 1$, we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
+	For instance , we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
 	\begin{align}
 		e^x E_k(x) 
-		&\leq \inv{x + k - 1} \\
-		\iff \expect{\inv{a + \chi^2(n)}}
-		&\leq \inv{a + n - 2} \\
+		&\leq \inv{x + k - 1} , \ \forall k>1\\
 		\iff \expect{\frac{\mu^*}{\MAPm}} 
-		&\leq \frac{n +n_0}{a + n - 2}.
-	\end{align}
-We are left with a special case when  $n=1$, eg $k=\half$. 
+		&\leq \frac{(n_0 +n)\alpha}{a\alpha + n\alpha - 1}
+		= \frac{n_0 +n}{a + n - \inv{\alpha}} , \ \forall n >\inv{\alpha} \; .
+	\end{align}
+We are left with a special case when  $n\leq \alpha$. 
 Then we can use the trivial bound 
@@
 \end{proof}
+
+For small $n$. a more subtle alternative uses \href{http://dlmf.nist.gov/8.19.E12}{the recurrence relationship} \cite[Eq.~8.19.12]{DLMF}
+\alignn{
+	\alpha E_{\alpha+1}(\beta) + \beta E_\alpha(\beta) 
+	= e^{-\beta} 
+}
+and combines it with the inequality \citep[Eq.~8.19.21]{DLMF} to get
+\alignn{
+	\beta e^\beta E_\alpha(\beta)
+ 	= 1 - \alpha e^\beta E_{\alpha+1}(\beta)
+ 	\leq 1 - \frac{\alpha}{\alpha + 1 + \beta}
+ 	 = \frac{\beta + 1}{\alpha + \beta + 1} \; .
+}
+Back to the MAP, we get
+\alignn{
+	\expect{\frac{\mu^*}{\MAPm}} 
+	\leq \frac{n_0 + n}{a} \frac{a\alpha + 1}{n \alpha + a \alpha +1}
+	= \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}}
+}
 

### 2021-10-30 aistats.tex
Write proof for Gamma MAP bound

@@
 Note that if $X\sim \cN(0,\sigma^2)$, then $X^2 \sim \Gamma\paren{\half, \inv{2\sigma^2}}$ in the shape-rate parametrization of Gamma distributions.
-In fact the bounds above can be generalized to any distribution $\Gamma\paren{\alpha, \beta}$ with known shape $\alpha$, but we leave it to \cref{app:gaussian-variance} for the sake of clarity.
+In fact the bounds above can be generalized to any distribution $\Gamma\paren{\alpha, \beta}$ with known shape $\alpha$.
+This generalization encompasses exponential distribution when $\alpha=1$, as another important special case.
+ We postpone these rates to \cref{app:gaussian-variance} for the sake of clarity.
 
@@
 	= \beta e^{\beta} E_\alpha (\beta) \; .
+	\label{eq:expect1+gamma}
 }
@@
 
-For small $n$. a more subtle alternative uses \href{http://dlmf.nist.gov/8.19.E12}{the recurrence relationship} \cite[Eq.~8.19.12]{DLMF}
+When $n<\inv{\alpha}$, it is possible to get a much tighter bound by exploiting \href{http://dlmf.nist.gov/8.19.E12}{the recurrence relationship} \cite[Eq.~8.19.12]{DLMF}
 \alignn{
@@
 }
-and combines it with the inequality \citep[Eq.~8.19.21]{DLMF} to get
+and combining it with the inequality \citep[Eq.~8.19.21]{DLMF} to get
 \alignn{
@@
 }
-Back to the MAP, we get
+Plugging this inequality back into~\ref{eq:expect1+gamma}, we get the following upper bound for the MAP:
 \alignn{
@@
 	\leq \frac{n_0 + n}{a} \frac{a\alpha + 1}{n \alpha + a \alpha +1}
-	= \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}}
-}
+	= \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}} \; .
+}
+Unfortunately, this formula does not yield an elegant convergence rate, so we keep it out of the lemma.
 
@@
 
+There are several ways to write down the convergence rate.
+The Gaussian variance example can be written in a particularly simple form, so we give  it a special treatment in \S\ref{app:proof-MAP-gaussian}, corresponding to the theorem displayed in the main text.
+We make a more general statement about Gamma distributions in \S\ref{app:proof-MAP-gamma}.
+
+\subsubsection{Gamma with known shape}
+\label{app:proof-MAP-gamma}
+
 \begin{theorem}[MAP Bound]
-The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
+The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP of $\Gamma(\alpha, \frac{\alpha}{\mu^*})$ with prior hyper-parameters $(n_0,\mu_0)$ is upper bounded as
+\alignn{
+    \forall n\geq \inv{\alpha},\quad \quad
+	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
+	\leq\frac{1}{n_0+n} + 
+	\frac{\alpha\paren{\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} - 1 }^2 }{\paren{1+\frac{n}{n_0}}\paren{\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} + \frac{n}{n_0}}}
+}
+\end{theorem} 
+Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$. Expressions for $n\alpha < 1$ are less elegant so we do not write them down here.
+
+\begin{proof}
+We expand the symmetrized Bregman~\eqref{eq:symmetrized_bregman} to get
+\begin{align}
+	\expect{\cS_\conj(\mu^*; \MAPm)} 
+	&\leq \alpha \paren{\expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1} \; .
+\end{align}
+The expectation of $\MAPm$ is straightforward
+\begin{align}
+	 \expect{\frac{\MAPm}{\mu^*}} - 1 
+	 = \frac{n_0 \mu_0 + n \mu^*}{(n_0+n)\mu^*} - 1 
+	 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
+	 \quadtext{where}
+	 a:=n_0\frac{\mu_0}{\mu^*} \; .
+\end{align}
+There remains the more problematic term with the expectation of the inverse mean parameter, 
+for which we use the bound derived in Lemma \ref{lem:expected-map-natural-parameter-gaussian}  when $n\alpha \geq 1$
+\begin{align}
+	\expect{\frac{\mu^*}{\MAPm}} - 1 
+	\leq \frac{n_0 + n}{a +n - \inv{\alpha}} -1
+	= \frac{n_0 - a + \inv{\alpha}}{a +n - \inv{\alpha}}
+\end{align}
+so putting it all together we get 
+\begin{align}
+	\expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq 
+	\alpha \paren{\frac{n_0 - a + \inv{\alpha}}{a +n - \inv{\alpha}} 
+	 + \frac{a - n_0 \pm \inv{\alpha}}{n_0+n}} \\
+	 &=
+	 \alpha \paren{n_0 + \inv{\alpha} -a}  
+	 \left( \inv{a+n-\inv{\alpha}} - \inv{n_0+n} \right)
+	 + \inv{n_0 + n} \\
+	 &= \inv{n_0 + n} + \alpha \frac{(n_0 + \inv{\alpha} - a)^2}{(n_0 + n)(a+ n - \inv{\alpha} )} \\
+	&= \inv{n_0 + n} + \alpha \frac{(1 + \inv{\alpha n_0} - \frac{\mu_0}{\mu^*})^2}{ (1 + \frac{n}{n_0} ) (\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} +\frac{n}{n_0})} \; .
+\end{align}
+\end{proof}
+
+\subsubsection{Gaussian Variance}
+\label{app:proof-MAP-gaussian}
+
+\begin{theorem}[MAP Bound]
+The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP of $\cN(  0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
  \begin{align}
@@
 When $n=0$, the inequality is an equality. 
-For $n>0$, we expand the symmetrized Bregman~\eqref{eq:symmetrized_bregman} to get
+For $n>0$, we expand the symmetrized Bregman~\eqref{eq:symmetrized_bregman} with $\alpha=\half$ to get
 \begin{align}

### 2021-11-02 aistats.tex
Emphasize that SMD converges in the MAP example

@@
 
-As we saw in \S\ref{sec:problem}, MAP can be interpreted as stochastic mirror descent (SMD).
-This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform other designs and analyses of SMD.
+As we saw in \S\ref{sec:problem}, MAP  is a special case of stochastic mirror descent (SMD).
+This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from MAP may inform other designs and analyses of SMD.
+In particular, we know that MAP converges asymptotically as $O(n^{-1})$, so we hope to find a convergence rate for SMD that could capture this behavior. 
 We first review the assumptions of relative smoothness, helpful to deal with non-smooth functions, before investigating recent analyses of SMD with the MAP.

### 2021-11-04 aistats.tex
typo + finish ack.

@@
 We express our gratitude to Reza Babanezhad, Radu Dragomir and Hadrien Hendrikx for their insights on stochastic mirror descent. 
-This work was partially supported by the Canada CIFAR AI Chair Program. 
+This work was partially supported by the NSERC Discovery Grant RGPIN2017-06936 and the Canada CIFAR AI Chair Program. 
 Simon Lacoste-Julien is a CIFAR Associate Fellow in the Learning in Machines \& Brains program.
@@
 illustrated in orange in Figure~\ref{fig:phi}.
-To upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focus on the symmetrized Bregman instead
+To derive the upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focus on the symmetrized Bregman instead
 \alignn{

### 2021-11-09 aistats.tex
Add meeting notes

--- 2021-11-02 aistats.tex
@@
 
-\usepackage{aistats/aistats2022}
+\usepackage[accepted]{aistats/aistats2022}
 % If your paper is accepted, change the options for the package
@@
 
-\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner \And  Damien Scieur \And Simon Lacoste-Julien }
-
-\aistatsaddress{ Mila \And  UBC \And SAIT SAIL \And Mila}
+\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner \And  Damien Scieur \And Simon Lacoste-Julien \TODO{s} }
+
+\aistatsaddress{ 
+	Mila, Université de Montreal 
+	\And  University of \\ British Columbia
+	\And SAIT AI Lab, Montreal 
+	\And Mila, Université de Montreal \\
+	Canada CIFAR AI Chair \\
+	SAIT AI Lab, Montreal}
 ]
@@
 
-As we saw in \S\ref{sec:problem}, MAP  is a special case of stochastic mirror descent (SMD).
+As we saw in \S\ref{sec:problem}, MAP  can be interpreted as stochastic mirror descent (SMD).
 This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from MAP may inform other designs and analyses of SMD.
@@
 and add some context and experimental observations.
-As mentioned in the main text, the centered gaussian $\cN(0,\sigma^2)$ has sufficient statistic $x^2$ following a Gamma distribution $\Gamma\paren{\half, \inv{2 \sigma^2}}$. 
-In this section we derive results for samples  from a general gamma distribution $X \sim\Gamma(\alpha,\beta)$ with known shape parameter $\alpha$, but unknown rate parameter $\beta$.
+As mentioned in the main text, the centered gaussian $\cN(0,\sigma^2)$ has sufficient statistic $T(X)=X^2$ which follows a gamma distribution $\Gamma\paren{\half, \inv{2 \sigma^2}}$. 
+MLE and MAP only depend on $T(X)$, not $X$, so their performance only distributions of $T(X)$. 
+In this section we derive results for samples  from a general gamma distribution $X \sim \Gamma(\alpha,\beta)$ with known shape parameter $\alpha$, but unknown rate parameter $\beta$.
 Results for the Gaussian follow by taking $\alpha=\half$,
@@
 
+\begin{theorem}[MLE Upper Bound]
+	Consider an exponential family such that $T(X)$ is a gamma $\Gamma(\alpha, \beta)$ with known shape $\alpha$.
+	the expected KL between $\mu_*$ and the MLE $\hat\mu_n$	is infinite when $\alpha n\leq 1$ and otherwise upper bounded by
+	\begin{align}
+		 \expect{\bregmanconj( \mu_*; \hat \mu_n) }
+			\leq \inv{2n} + \frac{1}{n(n \alpha-1)}  \; .
+	\end{align}
+\end{theorem}
+
+This theorem applies to Gaussian variance since 
+\TODO{explain how this corresponds to theorem 4.1}
+
+\subsection{Gamma distribution}
 The density of $\Gamma(\alpha,\beta)$ reads
@@
 	&=  \alpha \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 },
-	\label{eq:symmetrized_bregman}
-}
-which can also be expressed with $\phi$ as follows:
+	\label{eq:symmetrized_bregman} \\
+	&= \alpha\phi\paren{\frac{\mu_*}{\mu_n}} + \alpha\phi\paren{\frac{\mu_n}{\mu_*}}
+}
+Writing $z=\frac{\mu_*}{\mu_n}$ this is equivalent to
 \aligns{
@@
 
-\subsection{Upper bound for the MLE of the variance}
-
-	\begin{theorem}[MLE Upper Bound]
-	For a Gamma $\Gamma(\alpha, \beta)$ with known shape $\alpha$,
-	the expected KL between $\mu_*$ and the MLE $\hat\mu_n$	is infinite when $\alpha n\leq 1$ and otherwise upper bounded by
-	\begin{align}
-		 \expect{\bregmanconj( \mu_*; \hat \mu_n) }
-			\leq \inv{2n} + \frac{1}{n(n \alpha-1)}  \; .
-	\end{align}
-	\end{theorem}
+\subsection{Proof for the MLE}
 
 	\begin{proof}
-	The MLE is a scaled sum of Gammas $\hat \mu_n = \inv{n} \sum_i X_i$.
+	For gamma distributions, since $T(X)=X$ the MLE is a scaled sum of gammas $\hat \mu_n = \inv{n} \sum_i X_i$.
 	As such it is also a Gamma with parameter $\Gamma(n \alpha, n \beta)$ and expectation $\frac{n\alpha}{n\beta} = \frac{\alpha}{\beta} = \mu_*$.
@@
 }
-Plugging this inequality back into~\ref{eq:expect1+gamma}, we get the following upper bound for the MAP:
+Plugging this inequality back into~\eqref{eq:expect1+gamma}, we get the following upper bound for the MAP:
 \alignn{
@@
 
+\subsubsection{Gaussian Variance}
+\label{app:proof-MAP-gaussian}
+
+\begin{theorem}[MAP Bound]
+The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP of $\cN(  0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
+ \begin{align}
+	& \expect{\cS_\conj( \m^*; \hat \m_n^\mathrm{MAP})}
+	\leq 
+	\left\{\begin{array}{ll}
+		\cS_\conj( \mu_*; \mu_0) 					& \text{ if } \ n=0, \\
+		\inv{2(n_0+1)}  +  b_1 						& \text{ if }\ n=1,\\
+		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n  & \text{ if }\ n\geq 2,
+	\end{array}\right.
+	\quad 
+	\quad 
+	\text{where }
+	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \in O\paren{\frac{n_0^2}{n^2}} \; . 
+	%\nonumber
+\end{align}
+\end{theorem}
+
+\begin{proof}
+When $n=0$, the inequality is an equality. 
+For $n>0$, we expand the symmetrized Bregman~\eqref{eq:symmetrized_bregman} with $\alpha=\half$ to get
+\begin{align}
+	\expect{\cS_\conj(\mu^*; \MAPm)} 
+	&\leq \half \paren{\expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1} \; .
+\end{align}
+The expectation of $\MAPm$ is straightforward
+\begin{align}
+	 \expect{\frac{\MAPm}{\mu^*}} - 1 
+	 = \frac{n_0 \mu_0 + n \mu^*}{(n_0+n)\mu^*} - 1 
+	 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
+	 \quadtext{where}
+	 a:=n_0\frac{\mu_0}{\mu^*} \; .
+\end{align}
+There remains the more problematic term with the expectation of the inverse mean parameter, 
+for which we use the bound derived in Lemma \ref{lem:expected-map-natural-parameter-gaussian}.
+
+\textbf{When $n\geq 2$}, we get 
+\begin{align}
+	\expect{\frac{\mu^*}{\MAPm}} - 1 
+	\leq \frac{n_0 + n}{a +n - 2} -1
+	= \frac{n_0 - a + 2}{a +n - 2}
+	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
+\end{align}
+so putting it all together we get 
+\begin{align}
+	\expect{\bregmanconj(\mu^*; \MAPm)} 
+	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
+	&= \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
+	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
+	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
+	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
+\end{align}
+
+\textbf{When $n=1$} the bound~\eqref{eq:lemma_nat_bound}  on the expected natural parameter gives
+\begin{align}
+	\expect{\frac{\mu^*}{\MAPm}} - 1 
+	\leq \frac{n_0 + 1}{a} -1
+	= \frac{n_0 +1 - a}{a}
+\end{align}
+so putting it all together we get
+\begin{align}
+	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
+	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 +1 - a }{a} \\
+	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
+	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
+	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{\frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
+\end{align}
+so we  recover the same bias term as when $n\geq2$.
+\end{proof}
+
 \subsubsection{Gamma with known shape}
@@
 \begin{theorem}[MAP Bound]
-The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP of $\Gamma(\alpha, \frac{\alpha}{\mu^*})$ with prior hyper-parameters $(n_0,\mu_0)$ is upper bounded as
+Consider an exponential distribution with sufficient statistics coming from a gamma distribution $\Gamma(\alpha, \beta^*)$ with mean parameter $\mu^* = \frac{\alpha}{\beta^*}$.
+The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP  with prior hyper-parameters $(n_0,\mu_0)$ is upper bounded as
 \alignn{
@@
 	\leq\frac{1}{n_0+n} + 
-	\frac{\alpha\paren{\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} - 1 }^2 }{\paren{1+\frac{n}{n_0}}\paren{\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} + \frac{n}{n_0}}}
+	\frac{\alpha\paren{\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} - 1 }^2 }{\paren{1+\frac{n}{n_0}}\paren{\frac{\mu_0}{\mu^*} +  \frac{n - \inv{\alpha}}{n_0}}}
 }
 \end{theorem} 
-Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$. Expressions for $n\alpha < 1$ are less elegant so we do not write them down here.
+Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.
+Expressions for $n\alpha < 1$ are less elegant so we do not write them down here. 
+\TODO{try simplification with wolfram alpha}.
 
@@
 \end{align}
-\end{proof}
-
-\subsubsection{Gaussian Variance}
-\label{app:proof-MAP-gaussian}
-
-\begin{theorem}[MAP Bound]
-The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP of $\cN(  0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
- \begin{align}
-	& \expect{\cS_\conj( \m^*; \hat \m_n^\mathrm{MAP})}
-	\leq 
-	\left\{\begin{array}{ll}
-		\cS_\conj( \mu_*; \mu_0) 					& \text{ if } \ n=0, \\
-		\inv{2(n_0+1)}  +  b_1 						& \text{ if }\ n=1,\\
-		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n  & \text{ if }\ n\geq 2,
-	\end{array}\right.
-	\quad 
-	\quad 
-	\text{where }
-	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \; . 
-	%\nonumber
-\end{align}
-\end{theorem}
-
-\begin{proof}
-When $n=0$, the inequality is an equality. 
-For $n>0$, we expand the symmetrized Bregman~\eqref{eq:symmetrized_bregman} with $\alpha=\half$ to get
-\begin{align}
-	\expect{\cS_\conj(\mu^*; \MAPm)} 
-	&\leq \half \paren{\expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1} \; .
-\end{align}
-The expectation of $\MAPm$ is straightforward
-\begin{align}
-	 \expect{\frac{\MAPm}{\mu^*}} - 1 
-	 = \frac{n_0 \mu_0 + n \mu^*}{(n_0+n)\mu^*} - 1 
-	 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
-	 \quadtext{where}
-	 a:=n_0\frac{\mu_0}{\mu^*} \; .
-\end{align}
-There remains the more problematic term with the expectation of the inverse mean parameter, 
-for which we use the bound derived in Lemma \ref{lem:expected-map-natural-parameter-gaussian}.
-
-\textbf{When $n\geq 2$}, we get 
-\begin{align}
-	\expect{\frac{\mu^*}{\MAPm}} - 1 
-	\leq \frac{n_0 + n}{a +n - 2} -1
-	= \frac{n_0 - a + 2}{a +n - 2}
-	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
-\end{align}
-so putting it all together we get 
-\begin{align}
-	\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
-	&= \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
-	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
-	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
-	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
-\end{align}
-
-\textbf{When $n=1$} the bound~\eqref{eq:lemma_nat_bound}  on the expected natural parameter gives
-\begin{align}
-	\expect{\frac{\mu^*}{\MAPm}} - 1 
-	\leq \frac{n_0 + 1}{a} -1
-	= \frac{n_0 +1 - a}{a}
-\end{align}
-so putting it all together we get
-\begin{align}
-	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
-	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 +1 - a }{a} \\
-	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
-	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
-	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{\frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
-\end{align}
-so we  recover the same bias term as when $n\geq2$.
 \end{proof}

### 2021-11-09 aistats.tex
Fix affiliations

@@
 %
-%\runningtitle{I use this title instead because the last one was very long}
+\runningtitle{Convergence Rates for the MAP of an Exponential Family and SMD -- an Open Problem}
 
@@
 
-\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner \And  Damien Scieur \And Simon Lacoste-Julien \TODO{s} }
-
+\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner}
 \aistatsaddress{ 
 	Mila, Université de Montreal 
-	\And  University of \\ British Columbia
-	\And SAIT AI Lab, Montreal 
+	\And  University of British Columbia
+	}
+
+\aistatsauthor{Damien Scieur \And Simon Lacoste-Julien}
+\aistatsaddress{SAIT AI Lab, Montreal 
 	\And Mila, Université de Montreal \\
-	Canada CIFAR AI Chair \\
-	SAIT AI Lab, Montreal}
+	SAIT AI Lab, Montreal\\
+	Canada CIFAR AI Chair
+}
+
 ]

### 2021-11-09 aistats.tex
Clean beginning of App.A with feedback from Simon/Damien

@@
 As mentioned in the main text, the centered gaussian $\cN(0,\sigma^2)$ has sufficient statistic $T(X)=X^2$ which follows a gamma distribution $\Gamma\paren{\half, \inv{2 \sigma^2}}$.
-MLE and MAP only depend on $T(X)$, not $X$, so their performance only distributions of $T(X)$. 
+
+In general, if $X$ is part of the exponential family,  $T(X)$ is part of the natural exponential family with the appropriate support and base measure, and with the same log-partition function as $X$, up to constants.
+MLE and MAP only depend on $T(X)$, not $X$, so their performance only depends on the distributions of $T(X)$.
+
 In this section we derive results for samples  from a general gamma distribution $X \sim \Gamma(\alpha,\beta)$ with known shape parameter $\alpha$, but unknown rate parameter $\beta$.
 Results for the Gaussian follow by taking $\alpha=\half$,
- and we also immediately have results for exponential distributions by taking $\alpha=1$.
+We also immediately get results for exponential distributions by taking $\alpha=1$.
+For instance for the MLE we derive the following theorem:
 
 \begin{theorem}[MLE Upper Bound]
+	\label{thm:gammaMLE}
 	Consider an exponential family such that $T(X)$ is a gamma $\Gamma(\alpha, \beta)$ with known shape $\alpha$.
@@
 
-This theorem applies to Gaussian variance since 
-\TODO{explain how this corresponds to theorem 4.1}
-
-\subsection{Gamma distribution}
+To obtain the result for Gaussian variance, theorem~\ref{thm:varianceMLE}, simply plug-in $\alpha=\half$.
+
+In this section, we review useful properties of the gamma distribution and associated Bregman divergence in \S\ref{app:gamma}. 
+Then we prove theorem~\ref{thm:gammaMLE} in \S\ref{app:gammaMLE}.
+Then we prove an extension of theorem~\ref{thm:varianceMLE} in \S\ref{app:multivariateMLE}.
+Then we prove a useful lemma about the expectation of the natural parameter of the MAP in \S\ref{app:nat-bound}.
+Finally we prove upper bounds for the MAP in \S\ref{app:MAP-bound}.
+
+\subsection{Gamma Distribution}
+\label{app:gamma}
 The density of $\Gamma(\alpha,\beta)$ reads
@@
 \subsection{Proof for the MLE}
+	\label{app:gammaMLE}
 
 	\begin{proof}
-	For gamma distributions, since $T(X)=X$ the MLE is a scaled sum of gammas $\hat \mu_n = \inv{n} \sum_i X_i$.
-	As such it is also a Gamma with parameter $\Gamma(n \alpha, n \beta)$ and expectation $\frac{n\alpha}{n\beta} = \frac{\alpha}{\beta} = \mu_*$.
-	If we consider the ratio $\frac{\MAPm}{\mu_*}$, it is also a Gamma with parameter $\Gamma(n \alpha, n \alpha)$ 
+	Since $T(X)$ follows a gamma distribution $\Gamma(\alpha, \beta)$, the MLE is a scaled sum of gammas $\hat \mu_n = \inv{n} \sum_i T(X_i)$.
+	As such it is also a gamma with parameter $\Gamma(n \alpha, n \beta)$ and expectation $\frac{n\alpha}{n\beta} = \frac{\alpha}{\beta} = \mu_*$.
+	If we consider the ratio $\frac{\MAPm}{\mu_*}$, it is also a gamma with parameter $\Gamma(n \alpha, n \alpha)$.
 	Its inverse follows an inverse gamma distribution with expectation
@@
 	\end{align}
-	There is also a closed form solution for the expected logarithm of a Gamma.
-	Indeed a the sufficient statistic of a Gamma is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a Gamma to get
+	There is also a closed form solution for the expected logarithm of a gamma.
+	Indeed a the sufficient statistic of a gamma is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a gamma to get
 	\begin{align}
@@
 \subsection{Multivariate MLE}
+\label{app:multivariateMLE}
 
@@
 \subsection{Bounding the expected natural parameters for the MAP}
+\label{app:nat-bound}
 Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
@@
 }
-	where  $\sum_i \frac{X_i^2}{ \mu^*} \sim \Gamma(n\alpha, \alpha)$ is a Gamma random variable and $a=n_0 \frac{\mu_0}{\mu^*}$. 
+	where  $\sum_i \frac{X_i^2}{ \mu^*} \sim \Gamma(n\alpha, \alpha)$ is a gamma random variable and $a=n_0 \frac{\mu_0}{\mu^*}$. 
 	Further note that 
@@
 \subsection{Proof of MAP Bound}
+\label{app:MAP-bound}
 
@@
 The Gaussian variance example can be written in a particularly simple form, so we give  it a special treatment in \S\ref{app:proof-MAP-gaussian}, corresponding to the theorem displayed in the main text.
-We make a more general statement about Gamma distributions in \S\ref{app:proof-MAP-gamma}.
+We make a more general statement about gamma distributions in \S\ref{app:proof-MAP-gamma}.
 
@@
 
-\subsubsection{Gamma with known shape}
+\subsubsection{Gamma with Known Shape}
 \label{app:proof-MAP-gamma}
@@
 Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.
-Expressions for $n\alpha < 1$ are less elegant so we do not write them down here. 
-\TODO{try simplification with wolfram alpha}.
+Expressions for $n\alpha < 1$ are less elegant. See below \TODO{Damien try this on Matlab?}
+
+\alignn{
+ 	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
+	\leq \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}} -1 + \frac{a - n_0}{n_0+n}
+	\quadtext{where}
+	 a:=n_0\frac{\mu_0}{\mu^*} \; .
+}
 

### 2021-11-09 aistats.tex
Add ref to prior choice subsection

@@
 Then we prove an extension of theorem~\ref{thm:varianceMLE} in \S\ref{app:multivariateMLE}.
-Then we prove a useful lemma about the expectation of the natural parameter of the MAP in \S\ref{app:nat-bound}.
-Finally we prove upper bounds for the MAP in \S\ref{app:MAP-bound}.
+Then we prove a useful lemma about the expectation of the natural parameter of the MAP in \S\ref{app:nat-bound}, in order to prove upper bounds for the MAP in \S\ref{app:MAP-bound}.
+Finally we numerically investigate the effect of prior hyper-parameters in \S\ref{app:prior-choice}.
 
@@
 Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.
-Expressions for $n\alpha < 1$ are less elegant. See below \TODO{Damien try this on Matlab?}
-
+Expressions for $n\alpha < 1$ are less elegant as visible below
 \alignn{
@@
 \subsection{On the Choice of a Prior}
+\label{app:prior-choice}
 The optimal $\mu_0$ is larger than $\mu^*$ for small $n_0$ and small $n$.

### 2021-11-12 aistats.tex
Use running authors because only first row of authors was appearing
corrected typos + updated affiliation

@@
 %
-%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}
+\runningauthor{R\'emi Le Priol, Frederik Kunstner, Damien Scieur, and Simon Lacoste-Julien}
 
@@
 \aistatsauthor{Damien Scieur \And Simon Lacoste-Julien}
-\aistatsaddress{SAIT AI Lab, Montreal 
+\aistatsaddress{Samsung, SAIT AI Lab, Montreal 
 	\And Mila, Université de Montreal \\
-	SAIT AI Lab, Montreal\\
+	Samsung, SAIT AI Lab, Montreal\\
 	Canada CIFAR AI Chair
@@
 }
-where $\logpart$ is the convex log-partition function and $\nat\in\Theta$ is the \textit{natural parameter}.
+where $\logpart$ is the convex log-partition function and \mbox{$\nat\in\Theta$} is the \textit{natural parameter}.
 This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} or variants of principal component analysis \citep{collins2001generalization}, among other applications.
@@
 
-In general, if $X$ is part of the exponential family,  $T(X)$ is part of the natural exponential family with the appropriate support and base measure, and with the same log-partition function as $X$, up to constants.
-MLE and MAP only depend on $T(X)$, not $X$, so their performance only depends on the distributions of $T(X)$.
+In general, if $X$ is part of the exponential family, then $T(X)$ is part of the natural exponential family with the appropriate support and base measure, with the same log-partition function as $X$ up to constants.
+MLE and MAP only depend on $T(X)$, not $X$, so their performance only depends on the distribution of $T(X)$.
 
@@
 
-To obtain the result for Gaussian variance, theorem~\ref{thm:varianceMLE}, simply plug-in $\alpha=\half$.
+To obtain the result for Gaussian variance (see Theorem~\ref{thm:varianceMLE}), it suffices to set $\alpha=\half$ in Theorem~\ref{thm:gammaMLE}.
 
@@
 Then we prove theorem~\ref{thm:gammaMLE} in \S\ref{app:gammaMLE}.
-Then we prove an extension of theorem~\ref{thm:varianceMLE} in \S\ref{app:multivariateMLE}.
-Then we prove a useful lemma about the expectation of the natural parameter of the MAP in \S\ref{app:nat-bound}, in order to prove upper bounds for the MAP in \S\ref{app:MAP-bound}.
+Then we prove an extension of theorem~\ref{thm:varianceMLE} in \S\ref{app:multivariateMLE}, and prove a useful lemma about the expectation of the natural parameter of the MAP in \S\ref{app:nat-bound}, in order to prove upper bounds for the MAP in \S\ref{app:MAP-bound}.
 Finally we numerically investigate the effect of prior hyper-parameters in \S\ref{app:prior-choice}.
@@
 	There is also a closed form solution for the expected logarithm of a gamma.
-	Indeed a the sufficient statistic of a gamma is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a gamma to get
+	Indeed, the sufficient statistic of a gamma is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a gamma to get
 	\begin{align}
@@
 Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.
-Expressions for $n\alpha < 1$ are less elegant as visible below
+Expressions for $n\alpha < 1$ are less elegant as shown below:
 \alignn{

### 2021-11-28 rebuttal-aistats/main_response.tex
Add author response template and scribbles

% !TEX root = aistats2022_author_response.tex

\ifenablecomments{
	\color{red}
	\large Comments are enabled.
	Uncomment {\tt \textbackslash  enablecommentsfalse} for the submission.
}\fi

\section*{Main criticisms}

\subsection*{It's an open problem paper}

R1: 
``This does not qualify as a complete research work.''

R2: 
``The major weakness is that this paper is not a "regular paper" in the usual sense. 
[...]
So I am not sure whether the very nature of the paper is suitable for aistats, or more suitable for other venues which have calls for open problems.''

\FDK{Paraphrasing SLJ;}
The main weakness noted by Reviewers 1 \& 2 
is that our submission aims to promote discussions around an open problem, 
rather than a ``regular paper'' (R2), 
and they are unsure whether it is suitable for the AISTATS audience.
We agree with this characterization of the submission. 
%and acknowledge that such papers are (perhaps wrongly) scarce in the major machine learning conferences.
Our submission to AISTATS was motivated by the emphasis, in the call for papers, on the promotion of the exchange of ideas between the communities working in statistics and optimization in the context of machine learning.

The people in optimization and statistics we talked to during our investigation of this problem where interested in knowing the answers to these questions; members of one community thought this problem must have already been solved by the other. This motivated our submission; to show the similarities between those questions, stimulate interest and get the help of the community to make progress on this hard problem.

\subsection*{The statement of the conjecture is not sufficiently precise}

R1: 
``Besides, for the open problems, it would be better to state the desired results in a concrete way.
[...]
This is a vague question, as there could be many forms of possible bounds, under various assumptions, and many of such results are essentially known.''

\FDK{This one is fair, and I'm not sure how to fix it. 
I would go with ``Under no additional assumption, other than steep exponential families, what is the generalization of the quadratic bias-variance bound?''. But this is still a question, rather than a conjecture about a concrete result that should hold.
}

\RLP{Re: ``and many of such results are essentially known'', what are those?
	\FDK{Any non-asymptotic result under additional assumptions fit the bill like the ones we cite by Bach et al. and others. I assume many special cases are much more well known in stats}
}

\subsection*{The coverage of related work needs improvement}

R2:
``some related work should be more thoroughly discussed. Currently the authors did a good job in reviewing the literature from optimization, but the counterpart on the statistical literature is less satisfactory''

\FDK{I think it's fair, but I don't know how to fix it (without a $\sim$3 weeks lit review and a new page)}

\subsection*{Additional examples, beyond the Gaussian case, should be considered}

R2:
``When it comes to specific examples, in addition to the Gaussian model discussed extensively in the paper, some other models (mainly categorical models) should also be discussed.
[...] for categorical examples such as the multinomial model studied in
~\\
Kamath, S., Orlitsky, A., Pichapati, D., \& Suresh, A. T. (2015, June). On learning distributions from their samples. In Conference on Learning Theory (pp. 1066-1100). PMLR.
''

R2:
``For example, the KL divergence computation in the multinomial model or a product Poisson model is also a classical problem, which should be discussed more thoroughly.''

R5:
``While the MLE/MAP
estimates for exponential families are well-known, I would have liked if the
paper could comment on the potential generalizability of the MLE/MAP bound
found for the Gaussian, to other exponential family distributions.''

\FDK{This is fair, and should be relatively easy to add. No internet so couldn't check Kamath et al. (2015), but this must be out there.
I haven't delved into Bernoulli bounds (I was only plotting the actual expected losses because you could get it in closed form), but I can look into it this week.
}

\section*{Minor comments}

\subsection*{Recommended litterature}

R1:
``In a more general stochastic optimization setup, [...]  bound of (nearly) the desired form (but not restricted to exponential families) [...]

Competing with the Empirical Risk Minimizer in a Single Pass (COLT 2015), Roy Frostig, Rong Ge, Sham M. Kakade, Aaron Sidford''

R3:
``other works use a similar mirror descent framework for the optimization of a parameter of an exponential family and might be relevant to cite:

The Information Geometry of Mirror Descent
Garvesh Raskutti, Sayan Mukherjee, 2013.

Conjugate-Computation Variational Inference : Converting Variational Inference in Non-Conjugate Models to Inferences in Conjugate Models. Mohammad Emtiyaz, Khan Wu Lin, 2017.''

\FDK{
Haven't read it, but I'd guess Frostig et al. (2015) is a generalization of SGD being asymptotically ``optimal'' on the test set, which wouldn't solve the problem. Good to cite though. 
For the other two, they are on, or use, mirror descent, so could be cited as related work.
But there's no convergence rate, whether stats nor optimization.
}

\subsection*{Other results to mention}

R2:
``The asymptotic rate d/2n is also a minimax lower bound of the KL divergence (not necessarily restricted to MLE/MAP). This is well-known in the information theory literature on the minimax prediction error/redundancy.''

\FDK{That should probably be added as a parenthetical somewhere}

### 2021-11-28 rebuttal-aistats/reviews_and_comments.tex
Add author response template and scribbles

% !TEX root = aistats2022_author_response.tex

\section*{Reviewer \#1 (Score: 4, Confidence: 4)}

\subsection*{1. Summary and Contributions}

This paper proposes a class of problems regarding the non-asymptotic rates of MLE/MAP in exponential families, especially in the small sample size regime. The convergence of stochastic mirror descent under natural parametrization is also discussed. Since AISTATS is a conference aimed at publishing complete research papers, my evaluation is based on the results proven, instead of the conjectures being made. I would encourage the authors to consider venues with an open problem session, for instance, COLT. Nevertheless, in the comments to follow, I will also remark on the open problems themselves.

\subsection*{2. Strengths}

It is true that the current non-asymptotic understanding into the class of classical parametric estimation problems is still lacking in some aspects, especially in the case involving small sample size, high dimensions, and optimization algorithms. In particular, for exponential families with natural parametrization, many relevant quantities are measured with respect to Bregman divergence, instead of the Euclidean distance. Taking KL divergence as the loss function, from Le Cam theory and Fisher's identity, we know that the $\frac{d}{2n}$ leading-order term is asymptotically efficient. It is therefore natural to ask for non-asymptotic bound matching its performance. Besides, Proposition 1 presents interesting results that relate weighted $\ell^2$ distance with Bregman divergence, under the self-concordance assumption. This could be useful for establishing desired bounds.

\subsection*{3. Weaknesses}

The paper only presented conjectures with discussion and partial results on toy examples. This does not qualify as a complete research work. Besides, for the open problems, it would be better to state the desired results in a concrete way. For example, Open Problem 1 simply asks to bound some expected Bregman divergences. This is a vague question, as there could be many forms of possible bounds, under various assumptions, and many of such results are essentially known. A more concrete way of stating such open problem would be in the form of ``under certain assumptions (self-concordance, smoothness of second and third order, etc), the expected Bregman divergence is upper bounded by $d / 2n + O (n^{-2})$''.

\subsection*{4. Correctness}

The proof of the partial results and toy examples are correct.

\subsection*{5. Clarity}

The conjectures need to be stated in a clearer way. (see ``weakness'' section)

\subsection*{6. Relation to prior work}

In a more general stochastic optimization setup, there are some works that obtain a bound of (nearly) the desired form (but not restricted to exponential families). It would be better to discuss the connection and differences:

Competing with the Empirical Risk Minimizer in a Single Pass (COLT 2015), Roy Frostig, Rong Ge, Sham M. Kakade, Aaron Sidford

\subsection*{7. Additional Comments}

(N/A)

\newpage

\section*{Reviewer \#2 (Score: 5, Confidence: 4)}

\subsection*{1. Summary and Contributions}

This paper proposed the following general problem: in an exponential family $P_\theta$, how to prove an expectation or high-probability upper bound of the $KL(P_\theta, P_{\hat\theta})$ where $\hat\theta$ is either the MLE or the MAP estimator based on a conjugate prior? The authors provided some basic properties and ideas towards this problem.

From the statistical perspective, the authors proved non-asymptotic upper bounds on the KL divergence under Gaussian models with unknown covariance, and argued the non-triviality for the full Gaussian model. For the general theory, the authors derived the asymptotic rate, upper bounds when the associated Bregman divergence is quadratic or locally quadratic, and a generic bias-variance tradeoff. The general analysis remains largely open.

From the optimization perspective, the authors observed the analogy between the MAP estimator and stochastic mirror descent (SMD). Specifically, the MAP estimator with different sample sizes could be formulated as a proper SMD update, and the final KL divergence is the sub optimality gap in the minimization of the negative log-likelihood function. Despite this similarity, in Section 6 the authors discussed why existing literature on SMD is not sufficient for solving this problem.

\subsection*{2. Strengths}

This paper presents a nice summary and effort of a fundamental problem in statistics. Both the statistical and optimization perspectives are interesting, and it is an outstanding question to see how different views are related to each other. Overall speaking this is a nice "open problem paper" which presents an open problem clearly and contains sufficient discussions.

\subsection*{3. Weaknesses}

The major weakness is that this paper is not a "regular paper" in the usual sense. There are few solid and formal theorems - many of them are easy consequences or straightforward computation in specific models. In contrast, there are a lot of intuitions and discussions on limitations of certain approaches or why some ideas fail. So I am not sure whether the very nature of the paper is suitable for aistats, or more suitable for other venues which have calls for open problems.

If I start to evaluate this work from the "open problem paper" perspective, I think some related work should be more thoroughly discussed. Currently the authors did a good job in reviewing the literature from optimization, but the counterpart on the statistical literature is less satisfactory - many papers are only mentioned, without concrete settings and results. A more detailed picture of what have been done should be presented to the reader to better grasp the state-of-the-art.

When it comes to specific examples, in addition to the Gaussian model discussed extensively in the paper, some other models (mainly categorical models) should also be discussed. For example, the KL divergence computation in the multinomial model or a product Poisson model is also a classical problem, which should be discussed more thoroughly.

\subsection*{4. Correctness}

I checked most proofs and they are correct.

\subsection*{5. Clarity}

The writing is mostly clear and I enjoyed reading it. A possible improvement is to discuss some related work in more detail, rather than the current short descriptions.

\subsection*{6. Relation to prior work}

As included in the weakness, I would recommend to add some papers for categorical examples such as the multinomial model studied in

Kamath, S., Orlitsky, A., Pichapati, D., \& Suresh, A. T. (2015, June). On learning distributions from their samples. In Conference on Learning Theory (pp. 1066-1100). PMLR.

\subsection*{7. Additional Comments}

The asymptotic rate d/2n is also a minimax lower bound of the KL divergence (not necessarily restricted to MLE/MAP). This is well-known in the information theory literature on the minimax prediction error/redundancy.

\newpage
\section*{Reviewer \#3 (Score: 7, Confidence: 3)}

\subsection*{1. Summary and Contributions}

This paper focuses on the problem of deriving rates of convergence for the maximum a posteriori estimator of an exponential family (e.g., gaussian with unknown mean and variance) parametrized by the natural parameter theta and the log partition function A, in the Kullback-Leibler divergence objective, given a set of samples. Existing non asymptotic results only cover the case of a gaussian with known variance.

Their starting point is to see the MAP for a finite number of points n as a stochastic mirror descent update on the parameter theta, where the mirror map is the gradient of the log partition function, mapping theta to the mean parameter of the distribution parametrized by theta; with a stepsize is close to 1/n and the objective function is the KL between $p_{\theta^*}$ and $p_\theta$.

The paper discusses several paths for deriving rates of convergence for MAP in the KL objective. They first obtain bounds for gaussian with zero mean and unknown variance. They highlight then that the case of unknown mean/variance is harder to study. They provide a first partial result, a quadratic bound on the KL (the quadratic case) in the large number of samples regimes. As an optimization problem (section 6), recent results from the optimization literature on stochastic mirror descent do not apply, even with relative smoothness, as these works assume boundedness of the gradient of the objective, which does not hold for the KL (expected - log) here.

\subsection*{2. Strengths}

The paper tackles an important problem, ie non asymptotic guarantees regarding MAP, that is surely of interest for the statistics/ML community. It provides an interesting viewpoint and novel (even if partial) results compared to what can be found in the literature. The discussion is regularly illustrated with examples of distributions that satisfy the assumptions or not (e.g. smoothness, quadratic behavior of the Bregman divergence), which is also novel to the best of my knowledge. The optimization section mostly tells why results from the optimization literature do not directly apply but I found this discussion informative.

\subsection*{3. Weaknesses}

The paper only provide partial results but it is clearly stated.

\subsection*{4. Correctness}

The results appear technically sound, I did not find any flaws.

\subsection*{5. Clarity}

The paper is clearly written and discusses well the novel contributions.

\subsection*{6. Relation to prior work}

other works use a similar mirror descent framework for the optimization of a parameter of an exponential family and might be relevant to cite:

The Information Geometry of Mirror Descent
Garvesh Raskutti, Sayan Mukherjee, 2013.

Conjugate-Computation Variational Inference : Converting Variational Inference in Non-Conjugate Models to Inferences in Conjugate Models. Mohammad Emtiyaz, Khan Wu Lin, 2017.

\subsection*{7. Additional Comments}

Minor comments:
Figure 2 caption : MLE /MAP in orange/blue seem to have been swapped.

Section 6.2.1 : why is assumption 32 weaker than 31? what is tilde(g) in 32?

\newpage
\section*{Reviewer \#5 (Score: 8, Confidence 3)}

\subsection*{1. Summary and Contributions}

This paper is concerned with the problem of bounding the suboptimality of the MLE
and MAP (under conjugate priors) estimators of exponential family
distributions in a non-asymptotic way. It also presents an optimization
perspective into the MAP estimate by reinterpreting it as a Bregman proximal
update for the expected negative log-likelihood using the log-partition
function as the Bregman generator. The two open problems presented in this
paper are closely related to these persepctives: the first is to upper bound
the expected suboptimality of the MLE and MAP estimators, and the second is to
derive convergence rates for the online mirror descent algorithm for barrier
functions. The paper presents results for the first problem in the case of
a zero-mean Gaussian, and explores various avenues for the case of
the full Gaussian. In particular, the second problem is approached from the
optimization perspective, using tools such as relative smoothness and
convexity to try deriving a convergence rate for the stochastic mirror descent
algorithm.

\subsection*{2. Strengths}

The paper uses well-established tools from convex analysis and information
geometry to facilitate a well-grounded analysis of the problems they presented
therein. The problem of establishing convergence rates for the mirror descent
algorithm has been well-studied in optimization literature. The paper does
discuss a couple of techniques that have been proposed in order to establish
favorable convergence results, but existing techniques do not suffice for the
case of barrier objectives. The significance of the paper is that it provides
insight into a deeper connection between exponential family MLE/MAP estimation
and the Bregman geometry. The paper does build on existing results from the
relationship between exponential families and bregman divergences, but the
connection to stochastic mirror descent would be of great interest for future
work in optimization and (online) parameter estimation. Most importantly, the
general one-to-one correspondence between the Bregman divergence and the
exponential family is used throughout (wherever possible) in order to keep the
analysis as generally applicable as possible (the bias variance decomposition
for example). The connections between self-concordance and the quadratic
approximation to the Bregman was also fairly illuminating.

\subsection*{3. Weaknesses}

The biggest weakness seems that the answers discussed to the problems posed in
the paper are only discussed for the Gaussian distribution. While the MLE/MAP
estimates for exponential families are well-known, I would have liked if the
paper could comment on the potential generalizability of the MLE/MAP bound
found for the Gaussian, to other exponential family distributions. On the
novelty side, the partial solutions discussed in Section 5 mostly utilize
results from optimization and the quadratic approximation the Bregman, which
are minimally novel. However, the application of these techniques are fairly standard.

\subsection*{4. Correctness}

The proofs for the MLE and MAP bounds mostly employ various techniques from statistics and analysis, and the methods are sound.

\subsection*{5. Clarity}

The paper was really well-written and made it easier to follow and understand
the big picture. I would put this as one of the paper's biggest strengths.

\subsection*{6. Relation to prior work}

The relation to prior work has been discussed quite thoroughly, especially
throughout sections 5 and 6. Some of these works are quite recent (2021), and
in particular, I think that this paper is not too far away from producing
a solution to their second problem (the work of D'Orazio (2021) proves linear
convergence under constant step size, which also handles barrier objectives).

\subsection*{7. Additional Comments}

Typos:

- in Figure 2, seems that the colors for the MAP and MLE are reversed to
what the caption suggests.

- Minor notation comments:The notation for the Bregman divergence should only use semicolon or comma
throughout (to separate its arguments). Proposition 1, 2nd column on Page 7, and first column on Page 8
use comma.

In the accepted version, some things that would make the paper better:

- Maybe add a discussion on the proofs for proposition 1 and 2, just talking
about the kind of arguments used to establish these bounds.

- A further discussion of stochastic mirror descent and its convergence.
I am almost certain there might be something already out there which might
help improve the discussion for problem 2.

\newpage
\section*{Reviewer \#6 (Score: 8, Confidence: 4)}

\subsection*{1. Summary and Contributions}

This paper brings forth some fundamental gaps in our understanding of the non-assymptotic behavior of the risk of MLE/MAP estimators of the exponential family. The authors provide relevant background, discuss assymptotic behaviors and present lack of rigorous results for the non-assymptotic regimes.

\subsection*{2. Strengths}

The paper is largely well-written, and is well-motivated. The presented results (or lack of) are very relevant and useful to both optimisation and statistics communities, and should spark several discussions at the conference. The results and also well discussed using relevant toy empirical examples, which help in the understanding.

\subsection*{3. Weaknesses}

None I can see. The paper should be accepted IMO. The authors are very upfront about the problems they consider as being open and (atleast so far, seemingly) hard to solve. Nevertheless, there are some results which are somewhat easier toy settings. I would suggest the authors emphasize this more, a lot of results in statistical literature are for simpler settings. This could motivate readers to work through the simpler settings and work on the harder settings.

\subsection*{4. Correctness}
Yes

\subsection*{5. Clarity}

Yes

\subsection*{6. Relation to prior work}

Yes

\subsection*{7. Additional Comments}

None

### 2021-11-30 rebuttal-aistats/main_response.tex
Start cleaning rebuttal

@@
 
-\section*{Main criticisms}
+We thank all the reviewers for their careful review and insightful comments. We start our rebuttal by answering the common concern, followed by specific comments for each reviewer.
 
-\subsection*{It's an open problem paper}
+\paragraph{Common answer: Is the work relevant for AISTATS?}
+We understand from the reviewers that “open problem papers”  are unusual, and the question of whether such an article should be accepted or not at AISTATS is legitimate. 
+When writing the paper, we aimed to submit it at a cross disciplinary conference, as the paper’s topic is at the intersection of statistics and optimization.
+After a careful review of possibilities, AISTATS seems to be a perfect match, given what is written in the call of papers:
+"AISTATS is an interdisciplinary gathering of researchers at the intersection of computer science, artificial intelligence, machine learning, statistics, and related areas.
+Since its inception in 1985, the primary goal of AISTATS has been to broaden research in these fields by promoting the exchange of ideas among them."
+We strongly believe our article will promote an exchange of ideas, as emphasized by some of the reviewers below:
+(R3) "It provides an interesting viewpoint and novel (even if partial) results compared to what can be found in the literature."
+(R5)  "The significance of the paper is that it provides insight into a deeper connection between exponential family MLE/MAP estimation and the Bregman geometry.
+[...]  The connections between self-concordance and the quadratic approximation to the Bregman was also fairly illuminating."
+(R6) "The presented results (or lack of) are very relevant and useful to both optimisation and statistics communities, and should spark several discussions at the conference."
 
-R1: 
-``This does not qualify as a complete research work.''
+Additionally, members of the optimization and statistics community we talked to during our investigation of this problem were 
+either convinced a result already exist in the other community, 
+either interested in knowing the answers to these questions.
+Motivated by these feedbacks, we decided to submit as is, to highlight the similarity between these questions, 
+and to stimulate interest and get the help of the community to make progress on this hard problem.
 
-R2: 
-``The major weakness is that this paper is not a "regular paper" in the usual sense. 
-[...]
-So I am not sure whether the very nature of the paper is suitable for aistats, or more suitable for other venues which have calls for open problems.''
+\paragraph{(R1) The statement of the conjecture should be more precise.}
+We know that asymptotically, the expected KL belongs to $\frac{d}{2 n} + O (n^{-\frac{3}{2}})$, so it is tempting to explicitly search for a non-asymptotic upper bound in this set. 
+Yet, we are still unaware of any general upper bound.
+In our opinion, a bound of the form $\frac{c}{n}$ with the wrong constant $c>\frac{d}{2}$, or even a bound $\frac{c}{\sqrt{n}}$ would be novel and interesting.
+Consequently, we decided to keep the question open-ended.  
+Additionally, about the statement: (R1) "many of such results are essentially known."
+We are very interested in knowing examples of such   results, as we believe they could help us improve the paper.
 
-\FDK{Paraphrasing SLJ;}
-The main weakness noted by Reviewers 1 \& 2 
-is that our submission aims to promote discussions around an open problem, 
-rather than a ``regular paper'' (R2), 
-and they are unsure whether it is suitable for the AISTATS audience.
-We agree with this characterization of the submission. 
-%and acknowledge that such papers are (perhaps wrongly) scarce in the major machine learning conferences.
-Our submission to AISTATS was motivated by the emphasis, in the call for papers, on the promotion of the exchange of ideas between the communities working in statistics and optimization in the context of machine learning.
-
-The people in optimization and statistics we talked to during our investigation of this problem where interested in knowing the answers to these questions; members of one community thought this problem must have already been solved by the other.
-This motivated our submission; to show the similarities between those questions, stimulate interest and get the help of the community to make progress on this hard problem.
-
-\subsection*{The statement of the conjecture is not sufficiently precise}
-
-R1: 
-``Besides, for the open problems, it would be better to state the desired results in a concrete way.
-[...]
-This is a vague question, as there could be many forms of possible bounds, under various assumptions, and many of such results are essentially known.''
-
-\FDK{This one is fair, and I'm not sure how to fix it. 
-I would go with ``Under no additional assumption, other than steep exponential families, what is the generalization of the quadratic bias-variance bound?''.
-But this is still a question, rather than a conjecture about a concrete result that should hold.
-}
-
-\RLP{Re: ``and many of such results are essentially known'', what are those?
-	\FDK{Any non-asymptotic result under additional assumptions fit the bill like the ones we cite by Bach et al. and others. I assume many special cases are much more well known in stats}
-}
-
-\subsection*{The coverage of related work needs improvement}
-
-R2:
+\paragraph{(R2) The coverage of related work needs improvement.}
 ``some related work should be more thoroughly discussed.
 Currently the authors did a good job in reviewing the literature from optimization, but the counterpart on the statistical literature is less satisfactory''
+\FDK{I think it's fair, but I don't know how to fix it (without a $\sim$3 weeks lit review and a new page)}
+\RLP{Let's not address it as is, and focus on the specific references that were given by R2 and R3. Also add \url{https://arxiv.org/pdf/1811.11419.pdf} ?}
 
-\FDK{I think it's fair, but I don't know how to fix it (without a $\sim$3 weeks lit review and a new page)}
-
-\subsection*{Additional examples, beyond the Gaussian case, should be considered}
-
+\paragraph{(R2 \& R5) Additional examples, beyond the Gaussian case, should be considered}
 R2:
@@
 }
+\RLP{We are going to investigate categoricals, and if we get satisfying upper bound we will include it in the paper.}
+
+\paragraph{(R5)}
+Recall that D'Orazio (2021) proves linear convergence only down to a variance ball around the optimum.
+
+\paragraph{(R6)} 
+Thank you for your very encouraging comment.
 

### 2021-12-03 aistats.tex
Fix typos spotted by reviewers

@@
 	\caption{
-	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MAP (blue) and MLE (orange) (\S\ref{ssec:gaussian} with $n=3$).
+	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MLE (blue) and MAP (orange) (\S\ref{ssec:gaussian} with $n=3$).
 	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
@@
 	\implies
-	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2.
+	\bregmanconj(\m^*; \m) \leq \norm{\m^*-\m}_{\mF}^2.
 }
@@
 \aligns{
-	\stgcvx \cB_{A}(x, y)
+	\stgcvx \cB_{A}(x ; y)
 	\leq
-	\cB_f(x,y)
+	\cB_f(x ; y)
 	\leq
-	\smooth \cB_A(x,y) \; .
+	\smooth \cB_A(x ; y) \; .
 }
@@
 		($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
-		($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^*,\hat\nat_n)$?
+		($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^* ; \hat\nat_n)$?
 		($\hat\nat_n$) is it for the last iterate or an average ?
@@
 \subsubsection{Analogs of the Variance}
-Let us introduce the symmetrized Bregman induced by $\conj$, written $\cS_\conj(\m_1, \m_2) = \bregmanconj(\m_1, \m_2) + \bregmanconj(\m_2, \m_1)$.
+Let us introduce the symmetrized Bregman induced by $\conj$, written $\cS_\conj(\m_1 ; \m_2) = \bregmanconj(\m_1 ; \m_2) + \bregmanconj(\m_2 ; \m_1)$.
 \citet{hanzely2018fastest} assume that the expectation of $\cS$ between stochastic and deterministic updates verifies
 \alignn{
-	\expect[g]{\cS_\conj \paren{
-			\hat\m_{n} - \lr g(\hat \nat_{n}),
+	\expect[g\!]{\cS_\conj \paren{
+			\hat\m_{n} - \lr g(\hat \nat_{n}) ;
 			\hat\m_{n}   -\lr \nabla f(\hat \nat_{n})
@@
 this definition recovers the variance of the stochastic gradient
-\[
-\expect[g\!\!]{\norm{\nabla f(\theta) - g(\theta)}^2}\leq C.
-\]
+\alignn{
+	\expect[g\!]{\norm{\nabla f(\theta) - g(\theta)}^2}\leq C \; .
+	\label{eq:gradient-var}
+}
 Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate on function values with $O(1/n)$ step-sizes and tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
 
-\citet{dragomir2021fast} define the weaker assumption
-\alignn{
-	\expect[\tilde g]{
+\citet{dragomir2021fast} define the assumption
+\alignn{
+	\expect[g\!]{
 		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
@@
 % or when $\logpart$ is $\stgcvx$-strongly-convex,
-we recover the variance of the gradients at the optimum
-\[
-\expect[g\!\!]{\norm{\nabla g(\theta^*)}^2}\leq C.
-\]
+we recover the variance of the gradients at the optimum, which is weaker than~\eqref{eq:gradient-var},
+\alignn{
+	\expect[g\!]{\norm{\nabla g(\theta^*)}^2}\leq C \; .
+	\label{eq:opt-gradient-var}
+}
+%Note that~\eqref{eq:opt-gradient-var} is weaker than~\eqref{eq:gradient-var}.
 Using their descent lemma \citep[Eq.
 (12)]{dragomir2021fast} with the $O(1/n)$ step-size used by \citet[Th.
-3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^*, \hat\nat_n)$.
+3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^* ; \hat\nat_n)$.
 
@@
 \alignn{
-	\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C,
+	\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C \; ,
 	\label{eq:dorazio}

### 2021-12-03 aistats.tex
Add citations requested by R2

@@
 }
-has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
+has led statisticians to study better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial}, 
+or more recently  $\chi^2$ distance \citep{kamath2015learning} or Hellinger distance \citep{baraud2017new}.
 With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research,
@@
 }
-Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, we have
+Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, we get
 \begin{align}
@@
 This asymptote is independent of the optimum $\meanp^*$ or $\mF$ for well-specified models.
-Next, we give another example for which we get a rate matching~\eqref{eq:asymptote}.
+Actually, $\frac{d}{2n}$ is also the minimax of the KL over all estimators, at least for categorical data \citep{braess2004bernstein, kamath2015learning}.
+Next, we focus on the quadratic example,  whose finite sample rate matches~\eqref{eq:asymptote}.
 

### 2021-12-03 rebuttal-aistats/main_response.tex
Address individual reviewers

@@
 
-We thank all the reviewers for their careful review and insightful comments. We start our rebuttal by answering the common concern, followed by specific comments for each reviewer.
+We thank all the reviewers for their careful review and insightful or encouraging comments. We start our rebuttal by answering the common concern, followed by specific comments for each reviewer.
 
-\paragraph{Common answer: Is the work relevant for AISTATS?}
+\paragraph{Common answer: Is the paper relevant to AISTATS?}
 We understand from the reviewers that “open problem papers”  are unusual, and the question of whether such an article should be accepted or not at AISTATS is legitimate. 
@@
 "AISTATS is an interdisciplinary gathering of researchers at the intersection of computer science, artificial intelligence, machine learning, statistics, and related areas.
-Since its inception in 1985, the primary goal of AISTATS has been to broaden research in these fields by promoting the exchange of ideas among them."
+Since its inception in 1985, the primary goal of AISTATS has been to broaden research in these fields by \emph{promoting the exchange of ideas} among them."
 We strongly believe our article will promote an exchange of ideas, as emphasized by some of the reviewers below:
-(R3) "It provides an interesting viewpoint and novel (even if partial) results compared to what can be found in the literature."
+(R3) "The paper tackles an important problem [...] that is surely of interest for the statistics/ML community."
 (R5)  "The significance of the paper is that it provides insight into a deeper connection between exponential family MLE/MAP estimation and the Bregman geometry.
@@
 Additionally, members of the optimization and statistics community we talked to during our investigation of this problem were 
-either convinced a result already exist in the other community, 
+either convinced a result already exist, 
 either interested in knowing the answers to these questions.
-Motivated by these feedbacks, we decided to submit as is, to highlight the similarity between these questions, 
-and to stimulate interest and get the help of the community to make progress on this hard problem.
+Motivated by these feedbacks, we decided to submit as is, 
+to stimulate interest and get the help of the community to make progress on this hard problem.
 
@@
 Yet, we are still unaware of any general upper bound.
-In our opinion, a bound of the form $\frac{c}{n}$ with the wrong constant $c>\frac{d}{2}$, or even a bound $\frac{c}{\sqrt{n}}$ would be novel and interesting.
+In our opinion, a bound of the form $\frac{c}{n}$ with a large constant $c>\frac{d}{2}$, or even a bound $\frac{c}{\sqrt{n}}$ would be novel and interesting.
 Consequently, we decided to keep the question open-ended.
+
 Additionally, about the statement: (R1) "many of such results are essentially known."
@@
 
-\paragraph{(R2) The coverage of related work needs improvement.}
-``some related work should be more thoroughly discussed.
-Currently the authors did a good job in reviewing the literature from optimization, but the counterpart on the statistical literature is less satisfactory''
-\FDK{I think it's fair, but I don't know how to fix it (without a $\sim$3 weeks lit review and a new page)}
-\RLP{Let's not address it as is, and focus on the specific references that were given by R2 and R3. Also add \url{https://arxiv.org/pdf/1811.11419.pdf} ?}
+We will also discuss \citet{frostig2015competing}, noting that they work with smooth and strongly convex losses, which is precisely the assumptions we aim to relax.
 
-\paragraph{(R2 \& R5) Additional examples, beyond the Gaussian case, should be considered}
-R2:
-``When it comes to specific examples, in addition to the Gaussian model discussed extensively in the paper, some other models (mainly categorical models) should also be discussed.
-[...] for categorical examples such as the multinomial model studied in
-~\\
-Kamath, S., Orlitsky, A., Pichapati, D., \& Suresh, A. T. (2015, June). On learning distributions from their samples. In Conference on Learning Theory (pp. 1066-1100). PMLR.
-''
+\paragraph{(R2) The statistical literature review should be more detailed.}
+We understand R2 is mostly talking about the papers mentioned in the locally quadratic subsection. 
+As requested, we will expand the description of these references to convey a sharper picture of the state-of-the-art
+We will also add the reference you provided that proves minimax bound for categorical distributions with $\ell^1, \ell^2$ and $\chi^2$ distances, or for all $f$-divergences on a subset of the simplex.
 
-R2:
-``For example, the KL divergence computation in the multinomial model or a product Poisson model is also a classical problem, which should be discussed more thoroughly.''
+\paragraph{(R3)} We will discuss the references you provided.
+In \S6.2.1, we misleadingly asserted that 32 is weaker than 31.
+While this is true for quadratics, it may not be true in general.
+(And tilde g was a typo.)
 
-R5:
-``While the MLE/MAP
-estimates for exponential families are well-known, I would have liked if the
-paper could comment on the potential generalizability of the MLE/MAP bound
-found for the Gaussian, to other exponential family distributions.''
-
-\FDK{This is fair, and should be relatively easy to add. No internet so couldn't check Kamath et al. (2015), but this must be out there.
-I haven't delved into Bernoulli bounds (I was only plotting the actual expected losses because you could get it in closed form), but I can look into it this week.
-}
-\RLP{We are going to investigate categoricals, and if we get satisfying upper bound we will include it in the paper.}
-
-\paragraph{(R5)}
-Recall that D'Orazio (2021) proves linear convergence only down to a variance ball around the optimum.
+\paragraph{(R5)} We would like to clarify that \citet{dorazio2021stochastic} proves linear convergence of SMD with constant step-size, but only up to a variance ball around the optimum.
+Critically, the size of this variance ball does not depend on the step-size, and there is no way to prove the algorithm really converges to the optimum.
 
@@
 Thank you for your very encouraging comment.
+
+\bibliographystyle{apalike}
+\bibliography{../references}
 
@@
 }
-
-\subsection*{Other results to mention}
-
-R2:
-``The asymptotic rate d/2n is also a minimax lower bound of the KL divergence (not necessarily restricted to MLE/MAP).
-This is well-known in the information theory literature on the minimax prediction error/redundancy.''
-
-\FDK{That should probably be added as a parenthetical somewhere}

### 2021-12-03 rebuttal-aistats/main_response.tex
Add citation by John Tukey

@@
 
-Additionally, members of the optimization and statistics community we talked to during our investigation of this problem were 
-either convinced a result already exist, 
-either interested in knowing the answers to these questions.
+In 1954, John Tukey  \nocite{tukey1954unsolved} wrote "Difficulties in identifying problems have delayed statistics far more than difficulties in solving problems.
+This seems likely to be the case in the future, too."
+Many members of the optimization and statistics community we talked to during our investigation of this problem thought: "Surely, this problem has been solved."
 Motivated by these feedbacks, we decided to submit as is, 
@@
 \bibliography{../references}
-
-\section*{Minor comments}
-
-\subsection*{Recommended litterature}
-
-R1:
-``In a more general stochastic optimization setup, [...]  bound of (nearly) the desired form (but not restricted to exponential families) [...]
-
-Competing with the Empirical Risk Minimizer in a Single Pass (COLT 2015), Roy Frostig, Rong Ge, Sham M. Kakade, Aaron Sidford''
-
-R3:
-``other works use a similar mirror descent framework for the optimization of a parameter of an exponential family and might be relevant to cite:
-
-The Information Geometry of Mirror Descent
-Garvesh Raskutti, Sayan Mukherjee, 2013.
-
-Conjugate-Computation Variational Inference : Converting Variational Inference in Non-Conjugate Models to Inferences in Conjugate Models. Mohammad Emtiyaz, Khan Wu Lin, 2017.''
-
-\FDK{
-Haven't read it, but I'd guess Frostig et al. (2015) is a generalization of SGD being asymptotically ``optimal'' on the test set, which wouldn't solve the problem. Good to cite though. 
-For the other two, they are on, or use, mirror descent, so could be cited as related work.
-But there's no convergence rate, whether stats nor optimization.
-}

### 2022-01-07 aistats.tex
Minor fixes

@@
 \usepackage{lmodern}
-\usepackage{hyperref}       % hyperlinks
 \usepackage{url}            % simple URL typesetting
@@
 %\usepackage{subfigure}
-\usepackage{subcaption}
-\usepackage{placeins}
+%\usepackage{subcaption}
+%\usepackage{placeins}
 \usepackage{hyperref}       % hyperlinks
@@
     &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
-    &\leq 2L \norm{\m^* - \m},
+    &\leq 2L \norm{\m^* - \m} \; ,
 \end{align}
@@
     \bregmanconj(\m^* ; \m)
-    \leq \frac{L}{2} \norm{\m^* - \m}^2,
+    \leq \frac{L}{2} \norm{\m^* - \m}^2 \; ,
 \end{align}
@@
 The expected KL of the categorical MLE is always infinite, as previously mentioned in \S\ref{sec:problem}.
-
-Overcoming this limitation,
-\citet{ostrovskii2021finite} characterize the number of samples needed to be upper bounded by a quadratic \emph{with high-probability}, for any parametric models with a self-concordant log-likelihood $f$.
+However, Proposition~\ref{prop:selfConcordant} may be used to prove high-probability, many samples, convergence rates for one dimensional (and possibly multivariate) normal distributions.
+
+This is the spirit of
+\citet{ostrovskii2021finite} which characterizes the number of samples needed to be upper bounded by a quadratic \emph{with high-probability}, for any parametric models with a self-concordant log-likelihood $f$.
 \citet{anastasiou2017bounds} obtains a similar flavor of result under other assumptions on the third derivative of $f$.
@@
 \alignn{
-	\bregmanconj( \mu_*; \mu_n) 
-	\leq \cS_{\conj} (\mu_*, \mu_n )
-	&:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \nonumber\\
-	&=  \alpha \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 },
-	\label{eq:symmetrized_bregman} \\
-	&= \alpha\phi\paren{\frac{\mu_*}{\mu_n}} + \alpha\phi\paren{\frac{\mu_n}{\mu_*}}
-}
+	\cS_{\conj} (\mu_*, \mu_n )
+	&:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \\
+	&= \alpha\phi\paren{\frac{\mu_*}{\mu_n}} + \alpha\phi\paren{\frac{\mu_n}{\mu_*}} \\
+		&=  \alpha \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 } \; ,
+	\label{eq:symmetrized_bregman}
+}
+which verifies $\bregmanconj( \mu_*; \mu_n) \leq \cS_{\conj} (\mu_*, \mu_n )$.
 Writing $z=\frac{\mu_*}{\mu_n}$ this is equivalent to
@@
 }
-and as illustrated by the grey upper bound in Figure~\ref{fig:phi}.
+which is illustrated by the grey upper bound in Figure~\ref{fig:phi}.
 

### 2022-01-26 aistats-submission/rebuttal-aistats/instructions.tex
created colt submission + cleanup files

% !TEX root = aistats2022_author_response.tex

You have until \textbf{Friday, December 3, 2021 (11:59AM UTC)} to (optionally) respond to the reviews. You must submit a single response that addresses all reviews (not one response per review). The author response is limited to a \textbf{single page} in PDF format, including all figures, tables, and references, and has to use the AISTATS ``author response'' style that accompanies this \texttt{tex}-file. You may not alter this style file; in particular, you may not change the paper size, font, font size, or margins. Moreover, author responses must not contain external links, and must be \textbf{anonymized}.

Please focus your response on either answering specific questions raised in the reviews or correcting any misunderstanding or factual errors in the reviews.

You can change your response as often as you like until the above deadline. Please note that \textbf{this deadline is strict} and we encourage you to submit your response early so as to avoid technical issues. Please be aware that the deadline is \textbf{11:59AM (\textit{not} PM)} and the time zone is \textbf{UTC}. You may find it useful to use an online time zone converter.\footnote{See, e.g., \url{https://www.timeanddate.com/worldclock/converter.html?iso=20211203T115900&p1=tz_hkt&p2=tz_jst&p3=tz_gmt&p4=1440&p5=tz_et&p6=tz_pt}}

To include a figure in your response, the following LaTeX code is a possible solution:

\begin{verbatim}
\begin{minipage}[b]{0.3\linewidth}
\includegraphics[width=\linewidth]{path_to_figure}
\captionof{figure}{figure_caption}
\end{minipage}
\end{verbatim}

### 2022-01-26 colt/abstract.tex
created colt submission + cleanup files

--- 2021-12-03 aistats.tex
@@
-\documentclass[twoside]{article}
-
-\usepackage{silence}
-% Disable all warnings issued by latex starting with "You have..."
-% This is a hack, there is no clean solution for that unless we install locally the package
-\WarningFilter{latex}{You have requested package}
-
-%\pdfsuppresswarningpagegroup=1 %removes the warning
-% This is a hack, the other way to remove is to convert all pdf into postscript or eps.
-% RLP: not sure what is the point of this. With my latex it creates an error, and I do not have warnings, so I comment it out for now.
-
-\usepackage[accepted]{aistats/aistats2022}
-% If your paper is accepted, change the options for the package
-% aistats2022 as follows:
-%\usepackage[accepted]{aistats2022}
-
-\setlength{\footskip}{3.31pt}
-
-% If you set papersize explicitly, activate the following three lines:
-%\special{papersize = 8.5in, 11in}
-%\setlength{\pdfpageheight}{11in}
-%\setlength{\pdfpagewidth}{8.5in}
-
-%\let\oldsection\section
-%\renewcommand{\section}[1]{\oldsection{\texorpdfstring{\uppercase{#1}}{#1}}}
-
-% If you use natbib package, activate the following three lines:
-\usepackage[round]{natbib}
-\renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsubsection*{\bibname}}
-
-%\usepackage{microtype} % RLP tried it, no change.
-%\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{lmodern}
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{microtype}      % microtypography
-\usepackage{graphicx}
-\graphicspath{{figs/}}
-%\usepackage{subfigure}
-\usepackage{subcaption}
-\usepackage{placeins}
-\usepackage{hyperref}       % hyperlinks
-\usepackage[dvipsnames]{xcolor}
-
-\usepackage{array}
-\usepackage{wrapfig}
-
-\hypersetup{ % SLJ: my standard paper setup...
-	pdftitle={MAP convergence},
-	pdfkeywords={},
-	pdfborder=0 0 0,
-	pdfpagemode=UseNone,
-	colorlinks=true,
-	linkcolor=blue, %mydarkblue,
-	citecolor=blue, %mydarkblue,
-	filecolor=blue, %mydarkblue,
-	urlcolor=blue, %mydarkblue,
-	pdfview=FitH,
-	pdfauthor={Anonymous},
-	% draft,
-	final,
-}
-
-% To make a table of contents only for the appendix
-\usepackage[toc,page,header]{appendix}
-\usepackage{minitoc}
-\renewcommand \thepart{}
-\renewcommand \partname{}
-
-% To make colored boxes
-\usepackage[framemethod=TikZ]{mdframed}
-\newmdenv[
-	roundcorner=10pt,
-	backgroundcolor=Red!30,
-]{important}
-
-\usepackage[capitalise]{cleveref}
-\newcommand{\rlp}[1]{\textcolor{BrickRed}{(RLP:#1)}}
-\newcommand{\fdk}[1]{\textcolor{Periwinkle}{(fdk:#1)}}
-\newcommand{\TODO}[1]{\textcolor{cyan}{(TODO #1)}}
-\newcommand{\tocite}{\textcolor{purple}{(add citation)}}
-
-% my packages
-\usepackage{math_commands}
-% some custom math commands
-\newtheorem{proposition}{Proposition}
-\newtheorem{problem}{Open Problem}
-
-\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
-\newcommand*{\expecti}[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator
-
-\newcommand{\cond}{\,\vert\,}
-\newcommand{\logpart}{A}
-\newcommand{\conj}{{\logpart^*}}
-\newcommand{\bregman}{\cB_\logpart}
-\newcommand{\bregmanconj}{\cB_{\logpart^*}}
-\newcommand{\nat}{\theta}
-\newcommand{\m}{\mu}
-\newcommand{\meanp}{\m}
-\newcommand{\decrement}{D}
-\newcommand{\linear}{\ell} % linearization of a function
-\newcommand{\lr}{\gamma} % learning rate, or step-size
-\newcommand{\lin}[1]{\left\langle#1\right\rangle}
-
-\newcommand{\MAPm}{\hat \m_n}
-\newcommand{\MAPt}{\hat \nat_n}
-\DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}
-
-\newcommand{\stgcvx}{\alpha} % strong convexity
-\newcommand{\smooth}{\beta} % smoothness
-
-\begin{document}
-
-% If your paper is accepted and the title of your paper is very long,
-% the style will print as headings an error message. Use the following
-% command to supply a shorter title of your paper so that it can be
-% used as headings.
-%
-\runningtitle{Convergence Rates for the MAP of an Exponential Family and SMD -- an Open Problem}
-
-% If your paper is accepted and the number of authors is large, the
-% style will print as headings an error message. Use the following
-% command to supply a shorter version of the authors names so that
-% they can be used as headings (for example, use only the surnames)
-%
-\runningauthor{R\'emi Le Priol, Frederik Kunstner, Damien Scieur, and Simon Lacoste-Julien}
-
-\twocolumn[
-
-\aistatstitle{Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem}
-
-\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner}
-\aistatsaddress{ 
-	Mila, Université de Montreal 
-	\And  University of British Columbia
-	}
-
-\aistatsauthor{Damien Scieur \And Simon Lacoste-Julien}
-\aistatsaddress{Samsung, SAIT AI Lab, Montreal 
-	\And Mila, Université de Montreal \\
-	Samsung, SAIT AI Lab, Montreal\\
-	Canada CIFAR AI Chair
-}
-
-]
-
 \begin{abstract}
@@
 \end{abstract}
-
-\doparttoc % Tell to minitoc to generate a toc for the parts
-\faketableofcontents % Run a fake tableofcontents command for the partocs
-%\part{} % Start the document part
-%\parttoc % Insert the document TOC
-
-\begin{figure}[t]
-	\centering
-\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
-	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
-		Solid curve  are average over $10^5$ trials.
-		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
-		Shaded areas are 90\% confidence interval.
-		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
-		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
-		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
-	}
-	\label{fig:curves}
-\end{figure}
-
-\section{INTRODUCTION}
-\label{sec:motivation}
-
-\paragraph{Models}
-Exponential families are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
-Many standard random variables are exponential families: Gaussians, categorical, gamma, or Dirichlet, for example.
-They are flexible enough to model a variety of data sources $X$ and easy to describe with some sufficient statistics $T(X) \in \real^d$.
-They are particularly appreciated for their convex log-likelihood
-\alignn{
-f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
-\label{eq:defNLL}
-}
-where $\logpart$ is the convex log-partition function and \mbox{$\nat\in\Theta$} is the \textit{natural parameter}.
-This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} or variants of principal component analysis \citep{collins2001generalization}, among other applications.
-
-\paragraph{Estimators}
-In this paper, we consider the problem of estimating $\nat$ from a dataset $\mD = (X_1, \dots, X_n)$ of iid observations from $p_\theta$ in an exponential family.
-In this case, not only is $f$ convex, but it yields a simple condition for the maximum-likelihood estimate (MLE)
-\begin{align}
- \hat \mu_n^\text{MLE} = \nabla  \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
-	\label{eq:defMLE}
-\end{align}
-This rule is also known as moment matching.
-Given a specific conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
-In this paper, we will focus on analyzing MLE and MAP estimators.\footnote{A related analysis is present in the online-learning literature, but for different online estimators, which are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.}
-
-\paragraph{Statistical decision theory}
-To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
-We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
-{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$.
-This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
-In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
-{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
-For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
-but its lack of robustness to misspecification\footnote{
-For $p$ and $q$ continuous densities,
-$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
-%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
-}
-has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
-With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research,
-and raising the fundamental problem:
-\begin{equation}
-\boxed{
-\begin{aligned}
-	\textit{Find an upper}&\textit{ bound on the expected value of } \\
-	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
-\end{aligned}
-}
-\tag{$\star$}
-\label{problem}
-\end{equation}
-
-There are already general asymptotic results (\S\ref{ssec:asymptote} and Fig.~\ref{fig:curves}), and a finite $n$ result when $\logpart$ is quadratic (e.g., $X$ is Gaussian with known variance)
-or close to quadratic (\S\ref{ssec:quadratic}).
-However, a  general solution for finite $n$ remains elusive. 
-In this paper, we review partial solutions and give ideas on how to solve the problem.
-
-\paragraph{Optimization} 
-Stochastic optimization offers an interesting perspective on~\eqref{problem}.
-Consider the problem
-\alignn{
-	\min_{\theta\in \Theta} f(\theta)\,,
-	\label{eq:optimization_problem}
-}
-solved by $\nat^*\in \Theta$.
-Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
-\alignn{
-	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} } .
-	\label{eq:suboptimalityKL}
-}
-Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
-In particular, with exponential families, MAP is equivalent to stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
-Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we consider using existing convergence rates for SMD to get the upper bound we seek.
-Unfortunately, none of the current analyses apply, highlighting open problems for the analysis of SMD.
-
-\paragraph{Expected Outcomes}
-A solution to~\eqref{problem} can clarify the importance of the prior in MAP, in particular in the few sample regime. %, for instance for Gaussians $\cN(m,\sigma^2)$.
-Also, it could enable stochastic optimization to tackle a broad class of barrier objectives.\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}
-A good example is the generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
-It could also help assess the impact of alternative forms of regularization (prior) for these models.
-
-\paragraph{Contributions}
-After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
-\begin{itemize}
-	\itemsep0em
-	\item We provide an upper bound on the KL in the particular case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}), illustrating that tight rates are possible even though the current theory does not cover them.
-	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
-	%SLJ REMOVED per Slack discussion
-	%\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
-	\item By linking MAP and SMD, we show that modern analysis of SMD is yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
-\end{itemize}
-
-\paragraph{Notation}
-$X$ and $T=T(X)$ are random variables, $x$ is a sample, $n$ is the number of samples and $d= \dim(T)$.
-$\langle \cdot , \cdot \rangle$ is the Euclidean scalar product in $\real^d$.
-
-\section{TECHNICAL BACKGROUND}
-\label{sec:background}
-This section reviews the formalism of exponential families, their duality, a conjugate prior, and the corresponding MAP.
-We point the reader towards \citet[Chapter 3]{wainwright2008graphical} for a more detailed introduction.
-
-The density of an exponential family for a sample $x$ is
-\begin{equation}
-	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
-	 \label{eq:def_expfamily}
-\end{equation}
-where  $\nat$ is called natural (or primal) parameter.
-It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic,
-and 2) a base measure $\nu$ on $\cX$.
-Since the exponential is positive, $p$ has the same support as $\nu$.
-The log-partition function $\logpart$ acts as a normalization term, since
-\begin{align}
-    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \;.
-\end{align}
-This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding and multivariate normal distributions $\cX=\real^d, \nu$ Lebesgue and $T(X)=(X, X X^\top)$.
-
-For convenience, we focus on steep, regular exponential families with minimal statistic $T$ \citep{barndoffnielsen2014information}.
-Then $\logpart$ is a strictly convex function of Legendre type,
-and the set $\Theta = \{ \nat \cond \logpart(\nat) < \infty\}$ is open and convex.
-When explicit, we write the random variable $T = T(X)$.
-
-% TODO sort out relationships between : steep family, minimal statistics and Legendre type. What implies what ?
-
-{\bf Duality.}
-The log-partition function $\logpart$ verifies the two following identities:
-\begin{align}
-    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \label{eq:mirror-map} \\
-    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
-\end{align}
-where $\meanp$ is called the mean (or dual) parameter, which lives in the open convex set $\cM$ equal to the relative interior of the convex hull of $T(\cX)$.
-Given that $\logpart$ is strictly convex, its Hessian is positive definite, and its gradient $\nabla \logpart$ is a \textit{bijection} between natural parameters $\nat$ and mean parameters $\m$.
-We will write $\m$ or $\nat$ interchangeably depending on the context, being aware that both are linked and represent the same distribution.
-
-We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
-\aligns{
-	\conj(\m) =  \langle \m, \nat \rangle - \logpart(\nat)
-	=  \max_{\nat'\in\Theta}  \langle \m, \nat' \rangle - \logpart(\nat')\; ,
-}
-which is the common notion of \textit{entropy} in information theory.
-By Fenchel duality, its gradient is the inverse of the gradient of $\logpart$,  $\nabla\conj=\nabla\logpart^{-1}$, giving
-\aligns{
-	\nabla\conj \circ \nabla\logpart(\nat) = \nat, \quad \nabla\logpart\circ \nabla\conj(\meanp) = \meanp.
-}
-%\begin{figure}[t]
-%	\centering
-%	\includegraphics{duality}
-%	\caption{The gradient of the log-partition function and its dual, $(\nabla \logpart, \nabla \conj)$, form a bijection between natural and mean parameters $\nat, \meanp$.
-Figure copied from \citet{kunstner2020homeomorphic}.
-% TODO make an original one.
-%	}
-%	\label{fig:duality}
-%\end{figure}%
-%
-%Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
-{\bf The Bregman Divergence} induced by $\logpart$ measures the discrepancy between two parameters $\nat$ and $\nat_0$,
-\begin{align}
-    \bregman (\nat ; \nat_0)
-    & = \logpart(\nat) - \logpart(\nat_0)
-    - \langle \nabla \logpart(\nat_0)  , \nat - \nat_0 \rangle,
-    \label{eq:defBregman}
-\end{align}
-with $\nabla \logpart(\nat_0) = \expect[\nat_0]{T(X)} =: \meanp_0$ the mean parameter associated to $\nat_0$.
-In general, Bregman divergences are not symmetric, i.e., $\bregman (\nat ; \nat_0)\neq \bregman (\nat_0 ; \nat)$.
-%Finally, it is equal to the divergence of its convex conjugate with switched arguments,
-%\begin{align}
-%	\bregman (\nat ; \nat_0)
-%    = \bregmanconj ( \meanp_0 ; \meanp) \; .
-%    \label{eq:bregman_switch}
-%\end{align}
-
-{\bf A Conjugate Prior} for $p(X|\nat)$ is
-\begin{align}
-    p(\nat)
-    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \nonumber \\
-    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
-    \label{eq:def_prior}
-\end{align}
-where $n_0$ and $\nat_0$ are (hyper)parameters of the prior  \citep{agarwal2010geometric}.
-This is an exponential family with sufficient statistics $(\nat ,\logpart(\nat))$ and natural parameter $(n_0 \m_0, -n_0)$.
-Intuitively, $n_0$ is the number of fictive data points observed from a distribution with natural parameter $\nat_0$.
-
-{\bf Maximum A Posteriori (MAP).}
-Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
-Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
-\aligns{
-	p(\nat \cond \mD_n)
-    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
-%    \label{eq:joint_likelihood}
-}
-which reaches its maximum at $\MAPt^\text{MAP}$ such that
-\begin{align}
-    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
-    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; ,
-    \label{eq:defMAP}
-\end{align}
-where $T_i=T(X_i)$.
-When $n_0=0$ (no samples from the prior), we recover the MLE~\eqref{eq:defMLE}.
-We write $\MAPt$ for the MAP and view the MLE as a particular case.
-
-\section{PROBLEMS FORMULATION}
-\label{sec:problem}
-% Assume we observe a dataset $\mD$ drawn i.i.d. from a distribution $\cD$.
-% We further assume that our model is well-specified and there exist $\nat^*\in\Theta$ such that $\cD = p(\cdot \cond \nat)$.
-% As we will see, this is already an interesting setting.
-% We wish to estimate how well the MLE or a MAP model the true distribution.
-
-We are now ready to formalize the main problem of this paper. Assume we observe a dataset $\mD_n$ drawn i.i.d. from $p(\cdot \cond\nat^*)$, an exponential family distribution
-with parameters $\nat^*$.
-We wish to quantify how well the MLE or a MAP approximates the true distribution.
-
-A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
-In the well-specified setting, it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
-With exponential families, the KL is also a Bregman divergence:
-\alignn{
-	\KL(p_{\nat^*} || p_\nat)
-	 = \bregman(\nat ; \nat^*)
-	 = \bregmanconj(\m^* ; \m) \; .
-}
-The second equality is a general property of Bregman divergences and convex conjugates. % (Be careful about the order of arguments.)
-How does this quantity behave when $\hat \nat$ is the MLE or MAP?
- Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators when the loss is the KL divergence?
-This is our first problem.
-
-\begin{problem}[Upper-bounding MAP and MLE]
-Upper bound the following quantities:
-\begin{align}
-	\label{eq:bregmanMLE}
-	\text{MLE: } \quad &\expect[\mD_n]{\bregmanconj \left (\E_{\nat^*} [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
-	\label{eq:bregmanMAP}
-	\text{MAP: } \quad &\expect[\mD_n]{\bregmanconj \left (\E_{\nat^*} [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
-\end{align}
-where the expectation is on the data $\mD_n = (T_1, \dots, T_n)$.
-\end{problem}
-
-More explicitly, we want an upper bound that does not involve this expectation over the dataset.
-Surprisingly, we found no general solution to this seemingly simple problem, whether in the literature or by our means.
-In \S\ref{sec:example}, we provide results for special cases such as $\cN(0, \sigma^2)$,
-while in \S,\ref{sec:insights} we provide realistic conditions to obtain valid bounds after seeing a large number of samples.
-However, we have yet to find a solution encompassing both a broad range of exponential families
-and applicable to small sample sizes $n \lesssim d$.
-
-{\bf A Difficulty with the MLE.}
-While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite,
-for instance when estimating the covariance of a Gaussian when $n \leq d + 1$.
-Even worse, there is a non-zero probability never to sample one of the categories with categorical variables.
-In those cases the MLE gives zero weight to this category and $\KL(p_{\nat^*} \,\Vert\, p_\text{MLE} ) = +\infty$.
-Therefore, the expected KL~\eqref{eq:bregmanMLE} is infinite for any number of samples.
-Instead of taking the expectation, one might want to bound the risk in high probability
-without resorting to Markov inequality, as achieved by~\citep{ostrovskii2021finite},
- but this is a difficult endeavor.
-These examples make a case for regularized estimators such as MAP,
-for which we may find upper bounds.
-
-\paragraph{Optimization.}
-With exponential families, MAP can be linked to \emph{stochastic mirror descent (SMD)}, see App.~\ref{app:SMD}.
-More precisely, let us re-write~\eqref{eq:defMAP} as
-\alignn{
-\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
-\label{eq:mean_update}
-}
-where $\lr_n := \inv{n_0 + n}$.
-Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
-If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
-\alignn{
-	\nabla\conj(\hat \nat_{n})
-	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
-}
-which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
-and step-size schedule $\lr_n$, initialized at $\nat_0$.
-In this view, MLE forgets its (arbitrary) initialization after the first step with step size 1.
-The observation MAP $\in$ SMD brings us to our second problem.
-\begin{problem}[Convergence rate for SMD]
-Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(\meanp, \sigma^2)$.
-\end{problem}
-
-To address these problems, we start by investigating simple examples to provide solutions to Problem 1, getting insights into what is achievable.
-
-\section{ILLUSTRATING EXAMPLES}\label{sec:example}
-\subsection{Gaussian with Unknown Variance}\label{ssec:gaussian-variance}
-
-A non-trivial yet straightforward example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
-Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
-Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
-Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
-Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
-\aligns{
-	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
-	\conj(\m) = - \half \log(\m)  \; ,
-}
-up to constants.
-Both $A$ and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
-It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative gamma distribution.
-It also means $\bregman$ and $\bregmanconj$ have the same shape
-\begin{align}
-	\bregmanconj( \m^*; \m_n)
-	&= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
-%	\bregman( \nat_n; \nat_* )
-%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
-\end{align}
-In Theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
-All proofs for this section are in App.~\ref{app:gaussian-variance}.
-\begin{theorem}[MLE Bound]
-\label{thm:varianceMLE}
-	The MLE of $\cN(0,\m^*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
-	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
-	\begin{align}
-		 \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
-			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
-			\label{eq:MLE_rate}
-	\end{align}
-\end{theorem}
-This upper bound matches the asymptotic result~\eqref{eq:asymptote} that we derive in \S\ref{ssec:asymptote}.
-We illustrate its numerical behavior in Figure~\ref{fig:curves}.
-With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O\paren{\frac{d^2}{n} + \frac{d^3}{n(n-d-1)}}$.
-
-\begin{theorem}[MAP Bound]
-\label{thm:varianceMAP}
-The expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
- \begin{align}
-	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
-	\leq \begin{cases}
-		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
-		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
-	\end{cases}
-	\label{eq:MAP_rate}\\
-	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
-\end{align}
-\end{theorem}
-Anticipating on \S\ref{ssec:bias-variance}, this inequality highlights an explicit $O(\frac{v}{n} + \frac{b}{n^2})$ variance-bias decomposition.
-This inequality is derived with the symmetrized Bregman $\cB(a,b) + \cB(b,a)$ for which calculus is more tractable.
-This explains why the variance term is twice larger than the asymptote~\eqref{eq:asymptote}.
-Regarding the bias, it vanishes when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.
-This correlates well with our numerical observations (cf App.~\ref{app:gaussian-variance}).
-
-Note that if $X\sim \cN(0,\sigma^2)$, then $X^2 \sim \Gamma\paren{\half, \inv{2\sigma^2}}$ in the shape-rate parametrization of Gamma distributions.
-In fact the bounds above can be generalized to any distribution $\Gamma\paren{\alpha, \beta}$ with known shape $\alpha$.
-This generalization encompasses exponential distribution when $\alpha=1$, as another important special case.
- We postpone these rates to \cref{app:gaussian-variance} for the sake of clarity.
-
-\begin{figure*}[t]
-	\centering
-	\includegraphics[width=.8\textwidth]{figs/thales/numerical_schema_n=3.pdf}
-	\caption{
-	Primal and dual representations of a Gaussian $\cN(m,\sigma^2)$ MLE (blue) and MAP (orange) (\S\ref{ssec:gaussian} with $n=3$).
-	In dual space, MAP is a scaled version of the MLE~\eqref{eq:defMAP} with expectation $\E[\hat\m_n^\text{MAP}]=:\bar \m_n$ (light green), and MLE is unbiased $\E[\hat\m_n^\text{MLE}]=\m^*$, as illustrated by the parallels in the grey triangle.
-	In primal space, MAP has expectation $\tilde \nat_n$ (red), which intervenes in the bias-variance decomposition~\eqref{eq:bias-variance} from~\S\ref{ssec:bias-variance}.
-	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
-	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution.
-	}
-	\label{fig:thales}
-\end{figure*}
-
-\subsection{Full Gaussian (Non-Trivial)}
-\label{ssec:gaussian}
-Now that we have solved the case of $\cN(0,\sigma^2)$, consider the full Gaussian $\cN(m,\sigma^2)$, which offers a highly non-trivial example for Problem 1.
-Their log-likelihood reads $p(x) = -\frac{(x-m)^2}{2 \sigma^2} - \half \log(2\pi\sigma^2)$.
-With sufficient statistic $T(x)=(x, x^2)$,
-the mean parameters are $\m = \E[T(X)] = (m , m^2 + \sigma^2)$ belonging to the open set $\cM= \{(u,v) \cond u^2 < v\}$,
-and the natural parameters are $\nat= (\frac{m}{\sigma^2} , \frac{-1}{2\sigma^2}) \in \Theta = \real \times \real_-$.
-Examples of MAP and MLE  are represented in \cref{fig:thales} within $\cM$ and $\Theta$ delimited in grey.
-Given these parameters, log-partition and entropy are, up to constants,
-\alignn{
-	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\ 
-	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
-}
-
-These functions are neither smooth, nor strongly convex, but they are self-concordant, since $\conj$ is  the logarithmic barrier of a quadratic domain
-\citep[p.177, example 4.1.1.4]{nesterov2003introductory}, and self-concordance is preserved by convex-conjugacy~\citep{nesterov1994interior} -- see more details in App.~\ref{app:gaussian}.
-We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.
-
-\section{PARTIAL SOLUTIONS}
-\label{sec:insights}
-
-\subsection{Asymptotic Rate}
-\label{ssec:asymptote}
-As a reference point for any finite convergence rate, it is interesting to briefly review the classical asymptotic behavior of these quantities as $n \rightarrow +\infty$.
-Proofs are in App.~\ref{app:asymptote}, and \citet[\S1.1]{ostrovskii2021finite} offers a more comprehensive review.
-
-Statistics typically give results on $\nat$, but the MAP~\eqref{eq:defMAP} is more simply expressed with $\meanp$, so let us focus on $\bregmanconj$.
-Bregman divergences are locally quadratic, as seen via a second order Taylor expansion
-\alignn{
-    \textstyle \bregmanconj(\m^* ; \m)
-    &\textstyle = \frac{1}{2}\norm{\m^* - \m}^2_{\mF}
-    + O(\norm{\m - \m^*}^3),
-    \label{eq:bregmanTaylor}
-}
-where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the Hessian of the entropy at the optimum.
-It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
-\aligns{
-    \mF
-    :=\nabla^2\conj(\m^*)
-    = \nabla^2\logpart(\nat^*)^{-1}
-    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
-}
-Plugging the MLE~\eqref{eq:defMLE} or MAP~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, we have
-\begin{align}
-	\label{eq:asymptote}
-	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
-	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
-\end{align}
-Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible for large $n$.
-This asymptote is independent of the optimum $\meanp^*$ or $\mF$ for well-specified models.
-Next, we give another example for which we get a rate matching~\eqref{eq:asymptote}.
-
-\subsection{Quadratic Case}
-\label{ssec:quadratic}
-As another classical reference point, we consider the case $\logpart(\nat) = \half \norm{\nat}_2^2$.
-For instance, this is the log-partition of a Gaussian with known variance $I$,
-\[
-	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
-\]
-In this case, $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances since
-\begin{align}
-	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
-\end{align}
-Thanks to the independence of samples, we can break down the MLE into individual point's contributions:
-\begin{align}
-	\expect{\half \norm{\m^* -  \inv{n}  \smallsum_i T_i}_2^2}
-	=\frac{\Var(T)}{2n}
-	=\frac{d}{2n}.
-\end{align}
-Adding a reference mean $\m_0$ to get the MAP yields
-\begin{align}
-		\!\!\!\!\! \expect{\half \norm{\m^* -   \MAPm^\text{MAP}}_2^2} \!
-	&= \frac{n \Var(T) +  n_0^2 \norm{\m^* -  \m_0}^2}{2(n+n_0)^2}.\!
-	%\\ &= O\left(\frac{\Var(T)}{n} \right) + O\left(\frac{\norm{\m^* -  \m_0}^2}{n^2} \right)
-	\label{eq:MAP_quadratic}
-\end{align}
-We see here a variance term defining the $\frac{d}{2n}$ asymptote and a bias term in $O(n^{-2})$.
-However, this result does not generalize well to other families unless we make restrictive assumptions on $\conj$.
-% we can relate $\bregmanconj$ to a norm or a quadratic.
-% Can we generalize these results to other families?
-
-{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
-\begin{align}
-    \bregmanconj(\m^* ; \m)
-    &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
-    &\leq 2L \norm{\m^* - \m},
-\end{align}
-so $\bregmanconj$ is Lipschitz, and~\eqref{eq:MAP_quadratic} yields a $O(\inv{\sqrt{n}})$ rate, but no common exponential families verify the assumption.
-
-{\bf If $\conj$ is $L$-smooth}\footnote{$\conj$ is $L$-smooth iff $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
-\begin{align}
-    \bregmanconj(\m^* ; \m)
-    \leq \frac{L}{2} \norm{\m^* - \m}^2,
-\end{align}
-so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
-It is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
-Besides the Gaussian with known variance, the problem is that no standard exponential family has a \textit{globally} strongly convex log-partition function.
-The next section focuses on \textit{local} quadratic behavior, which is more realistic.
-
-\subsection{Locally Quadratic Case}
-\label{ssec:local-quadratic}
-From the Taylor expansion~\eqref{eq:bregmanTaylor},
-we know that all Bregman divergences are locally quadratic.
-Under some assumptions, such as self-concordance\footnote{
-In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
-} of $\conj$ \citep[Ch.~4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.~\ref{app:self-concordant}.
-\begin{proposition}
-	\label{prop:selfConcordant}
-Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
-Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
-\aligns{
-	\norm{\m^*-\m}_{\mF} < 0.21
-	\implies
-	\bregmanconj(\m^*; \m) \leq \norm{\m^*-\m}_{\mF}^2.
-}
-\end{proposition}
-To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
-For the MLE, the proof of~\eqref{eq:asymptote} from~\eqref{eq:bregmanTaylor} yields $\expecti{\norm{\m^*-\hat \m_n}^2_{\mF}} = \frac{d}{n}$ in general, so a sufficient condition is $n \geq 25 d$.
-For MAP, transforming~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 25d + 5 \norm{\m^* -  \m_0} - n_0$.
-This means that on average, we need $25$ times more samples than the dimension to reach the quadratic regime and ensure an upper-bound like~\eqref{eq:MAP_quadratic}.
-
-The entropy $\conj$ is self-concordant for several common families such as Gaussians (\S\ref{ssec:gaussian}), and all families with $\logpart \approx -\log$ such as
-exponential distributions,
-Laplace with known mean,
-Pareto with known minimum value,
-or Weibull with known shape $k$.
-The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g., categorical and Dirichlet distributions.
-Precisely, categorical variables illustrate that Proposition~\ref{prop:selfConcordant} does not imply directly a bound on the \emph{expected} Bregman.
-The expected KL of the categorical MLE is always infinite, as previously mentioned in \S\ref{sec:problem}.
-
-Overcoming this limitation,
-\citet{ostrovskii2021finite} characterize the number of samples needed to be upper bounded by a quadratic \emph{with high-probability}, for any parametric models with a self-concordant log-likelihood $f$.
-\citet{anastasiou2017bounds} obtains a similar flavor of result under other assumptions on the third derivative of $f$.
-More closely, in the world of exponential families, \citet{kakade2010learning} prove a result similar to Proposition~\ref{prop:selfConcordant} from a local bound on all higher-order moments of $\logpart$ in $\nat^*$.
-However, these results are expressed with quadratics in $\nat$, not $\m$, and they do not directly translate to convergence rates for the MAP, but they might with some more work.
-
-More generally, the present proposition and these related works answer~\eqref{problem} only \emph{partially}, as they all give \textit{large sample} results, that hold when $n\geq N$ for some constant $N$.
-A full solution to~\eqref{problem} would apply to small $n$.
-Informed by the properties that we have seen so far, we next investigate a general decomposition of the Bregman that could guide us towards a solution.
-
-\subsection{Bias-Variance Decomposition}
-\label{ssec:bias-variance}
-In both the quadratic~\eqref{eq:MAP_quadratic} and the Gaussian variance examples~\eqref{eq:MAP_rate}, the upper bound takes the form $O(\inv{n}) + O(\frac{\text{bias}}{n^2})$,
-giving us a flavor of what we would like as a general result for exponential families: a finite sample convergence rate, with variance and bias terms that reflect the important constants of the problem.
-Such a decomposition exists for any Bregman divergence \citep[Theorem 0.1]{pfau2013generalized}.
-\begin{theorem}[Bregman Bias-Variance Decomposition]
-	Let $\tilde \theta_n := \expecti{\hat \theta_n}$ be the expectation of the MAP in primal space, and $\tilde \m_n = \nabla \logpart(\tilde \theta_n )$ be its dual representation.
-The  expected Bregman decomposes into
-\begin{equation}
-	\expect{\bregmanconj(\m^* ; \hat \m_n)} = {\bregmanconj(\m^* ; \tilde \m_n)}
-	+ {\expect{\bregmanconj(\tilde \m_n ; \MAPm)}}
-	\label{eq:bias-variance}
-\end{equation}
-\end{theorem}
-We plot this decomposition for $\cN(\mu, \sigma^2)$ in Fig.~\ref{fig:gaussian_decomposition} , and we illustrate the primal mean $\tilde \nat_n$ in  Fig.~\ref{fig:thales}.
-
-\textbf{Remark:} In this decomposition, the primal expectation $\expecti{\hat \theta_n}$ is the reference point.
-An estimator will be unbiased if $\tilde \nat_n = \nat^*$.
-This is not true for the MLE, which is unbiased w.r.t. the dual parameter $\expecti{\hat \m_n}=\m^*$.
-
-We show in App.~\ref{app:bias-variance} that the bias decreases like ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{2}{n(n-2)}$ for Gaussian variance MLE, and
- ${\bregmanconj(\m^* ; \tilde \m_n)} \leq \frac{ \norm{\m^* - \m_0}^2 }{(1 + \frac{n}{n_0})^2 }$ for a quadratic MAP.
-These observations hint towards a general $O(1/n^2)$ upper bound for the bias, while the variance may be less dependent on the initialization $\nat_0$.
-
-\begin{figure}[t]
-	\centering
-	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
-	\caption{
-	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
-	}
-	\label{fig:gaussian_decomposition}
-\end{figure}
-
-In this section, we considered direct expansions of~\eqref{eq:bregmanMAP}.
-None of them could fully solve~\eqref{problem}.
-Next, we investigate whether an optimization approach could solve it.
-
-\section{AN OPTIMIZATION PROBLEM}
-\label{sec:optimization}
-
-As we saw in \S\ref{sec:problem}, MAP  can be interpreted as stochastic mirror descent (SMD).
-This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from MAP may inform other designs and analyses of SMD.
-In particular, we know that MAP converges asymptotically as $O(n^{-1})$, so we hope to find a convergence rate for SMD that could capture this behavior. 
-We first review the assumptions of relative smoothness, helpful to deal with non-smooth functions, before investigating recent analyses of SMD with the MAP.
-
-\subsection{Relative Smoothness}
-Mirror descent (MD) \citep{nemirovski1983problem,beck2003mirror}, also known as
-Bregman (proximal) gradient, relative gradient descent or NoLips,
-%Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
-and SMD \citep{nemirovski2009robust,ghadimi2012optimal}
-are typically encountered in non-smooth (online) optimization,
-under bounded (or Lipschitz) gradient assumption on the objective $f$
-and strong convexity assumption on the potential $\logpart$
-\citep[Th. 4.2(MD) \& Th. 6.3(SMD)]{bubeck2015convex}.
-In our case, these assumptions do not hold.
-For instance $\logpart = -\log$ is neither smooth nor strongly convex.
-
-Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
-\emph{relative} to a reference function $\logpart$, defined as
-\aligns{
-	\stgcvx \cB_{A}(x ; y)
-	\leq
-	\cB_f(x ; y)
-	\leq
-	\smooth \cB_A(x ; y) \; .
-}
-When $\logpart = \norm{\cdot}^2$, we recover the standard smoothness and gradient descent.
-These conditions ensure the linear convergence of MD with mirror map $\nabla A$
-\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively},
-even when $f$ is not smooth, and $\logpart$ not strongly convex.
-
-For exponential families, MAP perfectly fits into this framework, as
-\aligns{
-	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
-}
-is $1$-smooth and $1$-strongly convex relative to $A$.
-Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
-
-\subsection{Bounding the Randomness}
-
-To analyze stochastic algorithms, one also needs to quantify the randomness of stochastic gradients $g(\nat)$.
-While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, but they have so far been lacking concrete examples.
-We review these analyses in the light of the MAP and provide a summary in \cref{tbl:assumptions}.
-
-\begin{table}[t]
-	\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
-	\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
-	\caption{Summary of results for SMD
-		under relative smoothness and relative strong convexity assumptions.
-		Each row correspond to one analysis, and each columns answers one question.
-		($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
-		($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
-		($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^* ; \hat\nat_n)$?
-		($\hat\nat_n$) is it for the last iterate or an average ?
-		None of these analysis check all the boxes needed to address~\eqref{problem}.
-	}
-	\begin{center}
-		\begin{tabular}{lcccc}
-			%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
-			\toprule
-			Boundedness & $-\log$ &  $\lr_n \sim \inv{n}$ & $f$ & $\hat\nat_n$ \\
-			\midrule
-			%Strongly convex and bounded variance\newline
-			%\citep[e.g.,][]{??}
-			%&
-			%$
-			%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
-			%\quad
-			%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
-			%$
-			%\\
-			Variance on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
-			& \redxmark & \greencmark & \greencmark  & \redxmark
-			\\
-			Variance at $\theta^*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
-			& \redxmark & \greencmark & \redxmark  & \greencmark
-			\\
-			Optimization gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
-			& \greencmark & \redxmark & \redxmark & \greencmark
-			\\
-			\bottomrule
-		\end{tabular}
-	\end{center}
-	\label{tbl:assumptions}
-\end{table}
-
-%Before we proceed, let $f_X$ be a stochastic estimate of $f = \expecti{f_X}$, with stochastic gradients $g_X(\nat) =\nabla f_X(\nat)$.
-In our case $f_X(\nat) = - \log p(X\cond \nat)$ and $g_X(\nat) = \m - T(X)$.
-
-\subsubsection{Analogs of the Variance}
-Let us introduce the symmetrized Bregman induced by $\conj$, written $\cS_\conj(\m_1 ; \m_2) = \bregmanconj(\m_1 ; \m_2) + \bregmanconj(\m_2 ; \m_1)$.
-\citet{hanzely2018fastest} assume that the expectation of $\cS$ between stochastic and deterministic updates verifies
-\alignn{
-	\expect[g\!]{\cS_\conj \paren{
-			\hat\m_{n} - \lr g(\hat \nat_{n}) ;
-			\hat\m_{n}   -\lr \nabla f(\hat \nat_{n})
-	}} \leq \lr^2 C
-	\label{eq:hanzely}
-}
-for all possible iterates $\MAPt$, relevant step-sizes $\lr$ and for some constant $C$.
-When $A(\theta) = \frac{1}{2}\norm{\theta}^2$,
-%or when $A$ is $\stgcvx$-strongly convex,
-this definition recovers the variance of the stochastic gradient
-\alignn{
-	\expect[g\!]{\norm{\nabla f(\theta) - g(\theta)}^2}\leq C \; .
-	\label{eq:gradient-var}
-}
-Under this assumption, \citet[Lem.4.8]{hanzely2018fastest} prove a $O(1/n)$ convergence rate on function values with $O(1/n)$ step-sizes and tail averaging \citep{lacostejulien2012simpler} in primal space $\Theta$.
-
-\citet{dragomir2021fast} define the assumption
-\alignn{
-	\expect[g\!]{
-		\cB_{A^*}(\MAPm - 2\lr g(\theta_*), \MAPm)
-	} \leq 2 \lr^2 C \; .
-	\label{eq:dragomir}
-}
-When $A(\theta) = \frac{1}{2}\norm{\theta}^2$,
-% or when $\logpart$ is $\stgcvx$-strongly-convex,
-we recover the variance of the gradients at the optimum, which is weaker than~\eqref{eq:gradient-var},
-\alignn{
-	\expect[g\!]{\norm{\nabla g(\theta^*)}^2}\leq C \; .
-	\label{eq:opt-gradient-var}
-}
-%Note that~\eqref{eq:opt-gradient-var} is weaker than~\eqref{eq:gradient-var}.
-Using their descent lemma \citep[Eq.
-(12)]{dragomir2021fast} with the $O(1/n)$ step-size used by \citet[Th.
-3.2]{gower2019sgd} for SGD, we obtain a $O(1/n)$ convergence rate, on the Bregman with \emph{reversed} arguments $\bregman(\nat^* ; \hat\nat_n)$.
-
-These two analyses seem promising for~\eqref{problem}, but none of these assumptions hold in front of barrier objectives such as the $-\log$ from \S\ref{ssec:gaussian-variance}.
-Indeed, they both assume their bound holds uniformly for every possible iterate $\hat \nat_n$.
-Yet $\cN(0,\sigma^2)$ has a positive mass around $0$.
-This means that $\hat \m_n$ can get arbitrarily close from $0$, where the $-\log$ is unbounded, along with the associated  Bregman divergences~\eqref{eq:hanzely} and~\eqref{eq:dragomir}.
-In general, this uniform bound over $\hat \m_n$ cannot hold for \emph{barrier} objectives -- functions exploding to infinity in some finite point of space.
-
-Both of their proofs hold if we add an expectation over $\MAPm$ to their assumption.
-However, this is not helpful, as verifying the assumption becomes as hard as the initial problem.
-For instance, the expectation of~\eqref{eq:hanzely} over $\MAPm$ is an upper bound on the variance term of~\eqref{eq:bias-variance} (cf App.~\ref{app:bias-variance}).
-Confronted with this difficulty, we investigate an alternative definition of variance.
-
-\subsubsection{Bounded Optimality Gap}
-Inspired by \citet{loizou2021stochastic}, \citet{dorazio2021stochastic} explore the hypothesis
-\alignn{
-	\min_\nat f(\nat) - \expect[X]{\min_\nat f_X(\nat)} \leq C \; ,
-	\label{eq:dorazio}
-}
-where $f_X$ is a stochastic estimate of $f = \expecti{f_X}$. In our case $f_X(\nat) = - \log p(X\cond \nat)$.
-In other words, this lower bounds the expectation of the minimum of the stochastic estimates.
-For probabilistic models, such a bound is finite as soon as the model cannot give infinite density to any data point $x$.
-This holds, for instance, for discrete distributions because the probability mass is upper bounded by $1$; however, it rules out many families.
-In the case of normal distributions $\cN(m, \sigma^2)$, setting $m=x$ and $\sigma^2 \rightarrow 0$ gets $p_\nat (x) \rightarrow +\infty$,.
-%A similar behavior holds for Gamma, Beta, inverse Gaussians, log-normal, Gamma or inverse Gamma distributions.
-We have a similar behavior for gamma distribution with $\alpha = \beta x$ and $\beta \rightarrow +\infty$, or with the beta distribution with $\alpha=\beta \frac{x}{1-x}$ and $\beta \rightarrow +\infty$.
-Other counter-examples include inverse Gaussians, log-normal, gamma, inverse gamma.
-
-It is possible to overcome this limitation by treating batches of samples as single samples by averaging sufficient statistics, e.g.,. $Y = \{X_1, \dots, X_k\}$ and $T(Y) = \inv{k}\sum_i T(X_i)$.
-For instance, a multivariate normal of dimension $d$ cannot attribute infinite density to $d+1$ samples that are not in an affine subspace.
-
-Overall, \eqref{eq:dorazio} can partially handle barrier objectives, but it fails to account for the step-size $\lr_n = \inv{n_0+n}$, as \citet[Thm.1]{dorazio2021stochastic} only proves linear convergence to a variance ball of size $\frac{C}{\stgcvx}$ under constant step-size.
-This is in contrast with~\citet{dragomir2021fast} which can handle decreasing step-sizes but not barrier objectives.
-Proving convergence of stochastic mirror descent on barrier loss remains an open problem.
-
-\section{CONCLUSION}
-Despite the MLE and MAP estimators in the exponential family being classical and known in statistics for decades, we highlighted in this paper open problems to bound their frequentist risk (the expected KL) in a non-asymptotic way.
-We reviewed some partial results, such as a large sample analysis that describes how many samples are needed to ensure a locally quadratic regime~\citep{kakade2010learning, ostrovskii2021finite} for which rates are known.
-We also related this problem to the one of obtaining convergence rates in stochastic optimization, observing that MAP fits the framework of stochastic mirror descent with relative smoothness assumptions.
-Nevertheless, none of the current analyses of SMD hold for the MAP, even on a simple family such as $\cN(0,\sigma^2)$, thus revealing an area for progress in non-Euclidean optimization.
-In writing this paper, we hope to attract attention to this fundamental problem, leading to progress in both optimization and statistics.
-
-\subsubsection*{Acknowledgments}
-We express our gratitude to Reza Babanezhad, Radu Dragomir and Hadrien Hendrikx for their insights on stochastic mirror descent. 
-This work was partially supported by the NSERC Discovery Grant RGPIN2017-06936 and the Canada CIFAR AI Chair Program. 
-Simon Lacoste-Julien is a CIFAR Associate Fellow in the Learning in Machines \& Brains program.
-
-\bibliographystyle{apalike}
-\bibliography{references}
-
-\clearpage
-\appendix
-\onecolumn
-
-\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
-\part{Appendix} % Start the appendix part
-\parttoc % Insert the appendix TOC
-
-\section{PROOFS FOR GAUSSIAN VARIANCE}
-\label{app:gaussian-variance}
-
-In this section, we prove the results mentioned in \S\ref{ssec:gaussian-variance},
-and add some context and experimental observations.
-As mentioned in the main text, the centered gaussian $\cN(0,\sigma^2)$ has sufficient statistic $T(X)=X^2$ which follows a gamma distribution $\Gamma\paren{\half, \inv{2 \sigma^2}}$.
-
-In general, if $X$ is part of the exponential family, then $T(X)$ is part of the natural exponential family with the appropriate support and base measure, with the same log-partition function as $X$ up to constants.
-MLE and MAP only depend on $T(X)$, not $X$, so their performance only depends on the distribution of $T(X)$.
-
-In this section we derive results for samples  from a general gamma distribution $X \sim \Gamma(\alpha,\beta)$ with known shape parameter $\alpha$, but unknown rate parameter $\beta$.
-Results for the Gaussian follow by taking $\alpha=\half$,
-We also immediately get results for exponential distributions by taking $\alpha=1$.
-For instance for the MLE we derive the following theorem:
-
-\begin{theorem}[MLE Upper Bound]
-	\label{thm:gammaMLE}
-	Consider an exponential family such that $T(X)$ is a gamma $\Gamma(\alpha, \beta)$ with known shape $\alpha$.
-	the expected KL between $\mu_*$ and the MLE $\hat\mu_n$	is infinite when $\alpha n\leq 1$ and otherwise upper bounded by
-	\begin{align}
-		 \expect{\bregmanconj( \mu_*; \hat \mu_n) }
-			\leq \inv{2n} + \frac{1}{n(n \alpha-1)}  \; .
-	\end{align}
-\end{theorem}
-
-To obtain the result for Gaussian variance (see Theorem~\ref{thm:varianceMLE}), it suffices to set $\alpha=\half$ in Theorem~\ref{thm:gammaMLE}.
-
-In this section, we review useful properties of the gamma distribution and associated Bregman divergence in \S\ref{app:gamma}. 
-Then we prove theorem~\ref{thm:gammaMLE} in \S\ref{app:gammaMLE}.
-Then we prove an extension of theorem~\ref{thm:varianceMLE} in \S\ref{app:multivariateMLE}, and prove a useful lemma about the expectation of the natural parameter of the MAP in \S\ref{app:nat-bound}, in order to prove upper bounds for the MAP in \S\ref{app:MAP-bound}.
-Finally we numerically investigate the effect of prior hyper-parameters in \S\ref{app:prior-choice}.
-
-\subsection{Gamma Distribution}
-\label{app:gamma}
-The density of $\Gamma(\alpha,\beta)$ reads
-\alignn{
-	p(x) = \frac{\beta^\alpha}{\Gamma(\alpha)} x^{\alpha-1} e^{-\beta x} \; .
-}
-When $\alpha$ is known, it can be cast as an exponential family~\eqref{eq:def_expfamily} with sufficient statistic $T(x)=x$, domain $\cX = \real_+$ and base measure $\nu(x)\propto x^{\alpha-1}$.
-Then the natural parameter is $\nat = -\beta < 0$ and the log-partition function is
-\alignn{
-	\logpart (\nat) = -\alpha \log(-\nat)  + \log\Gamma(\alpha) \; .
-}
-From there we find that the mean parameter is $\mu = \frac{\alpha}{-\nat} >0$ and the entropy has the same form as the log-partition $\conj(\mu) = - \alpha \log(\mu) + \cst$.
-This means that the primal and dual Bregman divergences have the same form as well
-\begin{align}
-	\bregmanconj(\m_*; \m_n) 
-	&= \alpha \left ( \frac{\m_*}{ \m_n} - 1 - \log  \frac{\m_*}{ \m_n} \right) 
-	= \alpha \phi\paren{\frac{\m_*}{ \m_n}},
-	\\
-	\bregman( \nat_n; \nat_* ) 
-	&=  \alpha \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right)
-	= \alpha \phi\paren{\frac{ \nat_n}{\nat_*}},
-\end{align}
-\begin{wrapfigure}[14]{r}{0.4\textwidth}
-	\centering
-	\includegraphics[width=.38\textwidth]{phi.pdf}
-	\caption{%
-	Illustration of $\phi(z)$ (orange) and its upper bound $\phi(z) + \phi\paren{z^{-1}}$ (grey).
-	They both are barriers near $0$.
-	}
-	\label{fig:phi}
-\end{wrapfigure}
-where these 2 lines are equal, and $\phi$ measures the discrepancy between the ratio $\frac{ \nat_n}{\nat_*} =  \frac{\m_*}{ \m_n}  $ and $1$ via the function
-\begin{align}
-	\phi(z) := z - 1 - \log(z),
-\end{align}
-illustrated in orange in Figure~\ref{fig:phi}.
-To derive the upper bound for the MAP, due to the difficulty of finding a closed form for the expectation of the logarithm, we focus on the symmetrized Bregman instead
-\alignn{
-	\bregmanconj( \mu_*; \mu_n) 
-	\leq \cS_{\conj} (\mu_*, \mu_n )
-	&:= \bregmanconj( \mu_*; \mu_n)  + \bregmanconj( \mu_n; \mu_*) \nonumber\\
-	&=  \alpha \paren{\frac{ \mu_*}{\mu_n} -1 +\frac{ \mu_n}{\mu_*} - 1 },
-	\label{eq:symmetrized_bregman} \\
-	&= \alpha\phi\paren{\frac{\mu_*}{\mu_n}} + \alpha\phi\paren{\frac{\mu_n}{\mu_*}}
-}
-Writing $z=\frac{\mu_*}{\mu_n}$ this is equivalent to
-\aligns{
-	\label{eq:log_bound}
-	\phi(z) \leq \phi(z) + \phi\paren{z^{-1}}
-	&=  z -1  + \inv{z} -1
-	= \frac{(z-1)^2}{z} \; ,
-}
-and as illustrated by the grey upper bound in Figure~\ref{fig:phi}.
-
-\subsection{Proof for the MLE}
-	\label{app:gammaMLE}
-
-	\begin{proof}
-	Since $T(X)$ follows a gamma distribution $\Gamma(\alpha, \beta)$, the MLE is a scaled sum of gammas $\hat \mu_n = \inv{n} \sum_i T(X_i)$.
-	As such it is also a gamma with parameter $\Gamma(n \alpha, n \beta)$ and expectation $\frac{n\alpha}{n\beta} = \frac{\alpha}{\beta} = \mu_*$.
-	If we consider the ratio $\frac{\MAPm}{\mu_*}$, it is also a gamma with parameter $\Gamma(n \alpha, n \alpha)$.
-	Its inverse follows an inverse gamma distribution with expectation
-	\begin{equation}
-		\expect{\frac{\mu_*}{\hat \mu_n}} 
-		= \begin{cases}
-			\frac{n\alpha}{n \alpha -1 }\ \text{ if } n\alpha > 1, \\
-			+\infty \  \text{ otherwise. }
-		\end{cases}
-	\end{equation}
-	which implies that for $n\alpha>1$, 
-	\begin{align}
-		\expect{\frac{\mu_*}{\hat \mu_n}} -1  
-		= \frac{n \alpha}{n\alpha -1 } - 1
-		= \frac{1}{n \alpha -1}
-	\end{align}
-	There is also a closed form solution for the expected logarithm of a gamma.
-	Indeed, the sufficient statistic of a gamma is $(X, \log(X))$, so one can apply formula~\eqref{eq:mirror-map} on the log-partition of a gamma to get
-	\begin{align}
-		\expect{\log \frac{\hat \mu_n}{\mu_*}} 
-		= \psi\paren{n \alpha} - \log\paren{n \alpha},
-	\end{align}
-	where $\psi$ is the \href{https://en.wikipedia.org/wiki/Digamma_function}{digamma function.}
-	%Closed forms also exist for non-central chi-squares, see for example the note of \citep{pav2015moments}.
-	Consequently the suboptimality of the MLE has a closed form solution
-	\begin{align}
-	\expect{\bregmanconj( \mu_*; \hat \mu_n) }
-	&= \alpha \expect{\frac{\mu_*}{ \hat \mu_n} - 1 + \log \left(\frac{\hat \mu_n}{\mu_*} \right) },
-	\\
-	& =
-	\begin{cases}
-		\alpha \paren{ \frac{1}{n\alpha - 1} + \psi\paren{n \alpha} - \log\paren{n \alpha} }, \ \text{ if } n\alpha>1, \\
-			+\infty \  \text{ otherwise. }
-	\end{cases}
-	\end{align}
-	Surprisingly, for a gaussian variance where $\alpha = \half$, we need $3$ samples or more for the expected loss to be bounded.
-	When the expectation is finite, we can get a more interpretable formula using known \href{https://en.wikipedia.org/wiki/Digamma_function#Inequalities}{bounds on the digamma function},
-	\begin{align}
-		-\inv{x} \leq \psi(x) - \log(x) \leq -\inv{2x} = - \inv{x} + \inv{2x}	\; ,
-	\end{align}
-	giving, for $n \alpha >1 $,
-	\begin{align}
-		\frac{\alpha}{n\alpha-1} - \frac{\alpha}{n\alpha}
-		&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
-		\leq \frac{\alpha}{n\alpha-1} - \frac{\alpha}{n\alpha} + \frac{\alpha}{2n\alpha}
-		\\
-		\iff
-			\frac{1}{n(n \alpha-1)}
-			&\leq \expect{\bregmanconj( \mu_*; \hat \mu_n) }
-			\leq \inv{2n} + \frac{1}{n(n \alpha-1)} \; ,
-	\end{align}
-	so we get a $\Omega(n^{-2})$ lower bound and a $O(n^{-1}) + O(n^{-2})$ upper bound.
-	\end{proof}
-
-\subsection{Multivariate MLE}
-\label{app:multivariateMLE}
-
-For the sake of simplicity, in higher dimension we focus on Gaussian covariance estimation and avoid the general Wishart discussion. 
-In higher dimensions, $X \sim \cN(0, \mu_*)$, $X\in\real^d$, $T(X) = XX^\top$, and the mean parameter $\mu_*$ 
-is a $d \times d$ symmetric, positive definite covariance matrix with $p = \frac{d(d+1)}{2}$ degrees of freedom.
-Note that here $d$ denotes the dimensionality of the data $X$, rather than the dimensionality of the parameters $\mu_*$.
-
-\begin{theorem}[Multivariate MLE Upper Bound]
-	The MLE of  the covariance matrix of $X_i\sim\cN(0,\mu_*)$ is $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top \in\real^{d\times d}$ with $p = \frac{d(d+1)}{2} $ degrees of freedom. 
-	The expected KL divergence between $\mu_*$ and $\hat\mu_n$ 
-	is infinite when $n\leq d+1$ and otherwise upper bounded by
-	\alignn{
-	 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	\leq \frac{p}{2n} + \frac{p(d+2)}{n (n - d -1)} 
- 	\ , \forall n >d+1 \;.
-}
-\end{theorem}
-We see from~\eqref{eq:asymptote} that this bound is asymptotically tight.
-\begin{proof}
-The entropy of $X$ is a negative log-determinant $\conj(\mu) = -\log\det(\mu)$, whose gradient is the negative matrix inverse $\nabla\conj (\mu) = - \mu^{-1}$.
-The associated Bregman divergence is
-\alignn{
-	\bregmanconj(\mu_*; \hat \mu_n) 
-	= \half \paren{\Tr(\mu_* \hat \mu_n^{-1} ) - d  - \log\det(\mu_* \hat \mu_n^{-1}) } \; .
-}
-Thanks to the linearity of the trace, the expectation becomes
-\alignn{
-	\expect{\bregmanconj(\mu_*; \hat \mu_n) }
-	= \half \paren{\Tr(\expect{\mu_* \hat \mu_n^{-1}} ) - d   
-	+\expect{\log\det( \hat \mu_n \mu_*^{ -1}))} } \; .
-}
-When the estimator is the MLE, $\hat \mu_n = \inv{n} \sum_i X_i X_i^\top$, 
-then we define the mean parameter ``ratio'' as
-\aligns{
-	\mV &:= n \mu_*^{-\half}\hat \mu_n \mu_*^{-\half} = \sum_i (\mu_*^{-\half} X_i) (\mu_*^{-\half}X_i)^\top,
-}
-such that $\Tr(\mu_* \hat \mu_n^{-1})  = n \Tr(\mV^{-1})$.
-But $\mu_*^{-\half}X_i \sim \cN(0,\mI)$, so that $\mV$
-is sampled from a Wishart  $\cW(\mI, n)$, where $\mI$ stands for the identity matrix of order $d$.
-Recall that $\E[\mV] = n \mI$.
-Thanks to $\log\det$ being a sufficient statistic of the Wishart, and the natural to mean parameter formula~\eqref{eq:mirror-map}, we have a closed form for the expected log-determinant of a Wishart 
-\aligns{
-	\E[\log\det \mV] &= \psi_d\paren{\half[n]} + d \log 2 \; ,
-}
-where $\psi_d(\half[n]) = \sum_{i=0}^{d-1} \psi ( \half[n - i])$ is the multivariate digamma function.
-The expectation of an inverse Wishart  is straightforward to compute from the density and the log-partition function
-\begin{equation}
-	\E[\mV^{-1}] = \begin{cases}
-		+\infty \text{ if } n\leq d+1 \\
-		 \frac{\mI}{n - d - 1} \text{ otherwise,} 
-	\end{cases}
-\end{equation}
-which proves the infinite part of the statement.
-Consider now the case $n>d+1$. Using
-\alignn{
-	\Tr(\expect{\mu_* \hat \mu_n^{-1}}) - d
-	= n\Tr(\E[\mV^{-1}]) - d 
-	= \frac{nd}{n - d - 1} -d
-	= \frac{d (d+1)}{n -d -1}
-	= \frac{2p}{n -d -1},
-}
-and putting it all together, we get the following closed form for the expectation of the divergence
- \begin{align}
- 	\label{eq:where-we-put-it-back-together-multivariate-gaussian}
- 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	= \frac{p}{n-d-1} + \half \left ( \psi_d\paren{\half[n]} - d \log\paren{\half[n]}  \right ).
- \end{align}
-To bound $\psi_d$, we can use the same bound as in the univariate case, $\psi(x) \leq \log(x) - \inv{2x}$, to get
-\begin{align}
-	\psi_d\paren{\half[n]} - d \log\paren{\half[n]}
-	&= \sum_{i=0}^{d-1} \paren{\psi \paren{\half[n - i]} - \log\paren{\half[n]} } \\
-	&\leq \sum_{i=0}^{d-1} \paren{\log\paren{\half[n-i]} - \inv{n-i} - \log\paren{\half[n]} } \\
-	&= \sum_{i=0}^{d-1} \paren{\log\paren{1 - \frac{i}{n}} - \frac{1}{n-i}} \; .
-	\label{eq:psid_bound}
-\end{align}
-We can bound sum of reciprocals $\sum_{i=0}^{d-1} \frac{1}{n-i}$ by the typical bound on 
-the harmonic sum $H_n = \sum_{k=1}^n \frac{1}{k}$,
-\aligns{
-	\inv{2n +1}
-	\leq H_n - \log(n) - \gamma \leq 
-	\inv{2n - 1},
-}
-where $\gamma$ is the Euler constant. Then
-\begin{align}
-	- \sum_{i=0}^{d-1} \frac{1}{n-i} 
-	= H_{n-d} - H_n
-	&\leq \log(n-d) + \gamma + \inv{2(n-d) - 1} - \log(n) - \gamma - \inv{2n+1} \\
-	&= \log\paren{1 - \frac{d}{n}}  + \frac{2(d + 1)}{(2n+1)(2(n - d) - 1)} \\
-	&< \log\paren{1 - \frac{d}{n}}  + \frac{d + 1}{2n(n - d - 1)} \; .
-\end{align}
-Plugging this back into~\eqref{eq:psid_bound} yields
-\begin{align}
-	\psi_d(\half[n]) - d \log(\half[n]) 
-	\leq \sum_{i=0}^{\textcolor{red}{d}} \log\paren{1 - \frac{i}{n}}
-	+  \frac{d + 1}{2n(n - d - 1)} \; ,
-\end{align}
-where the sum now goes up to $d$.
-To bound the remaining sum, we use $\log(1+x)\leq x$ (for $x > -1$) to get
-\begin{align}
-	\sum_{i=0}^{d} \log\paren{1 - \frac{i}{n}}
-	\leq \sum_{i=0}^{d} - \frac{i}{n}
-	 = -\frac{d(d+1)}{2n} = -\frac{p}{n} \; .
-\end{align}
-Putting those bounds together in 
-\cref{eq:where-we-put-it-back-together-multivariate-gaussian}
-and reorganizing yields
-\begin{align}
- 	\expect{\bregmanconj( \mu_*; \hat \mu_n^\text{MLE(d)})}
- 	&< p \paren{ \inv{n-d-1} - \inv{2n}}  + \frac{d + 1}{4n(n - d - 1)}\\
- 	&= p \frac{n + d +1}{2 n (n-d-1)}  + \frac{d(d + 1)/2}{2dn(n - d - 1)}\\
- 	&= p\frac{n-d-1}{2n(n-d-1)} + p\frac{2(d+1)}{2 n (n-d-1)} + \frac{p}{2dn (n - d - 1)}\\
- 	&= \frac{p}{2n} + \frac{p(d+ 1 + \inv{2d})}{n (n-d-1)} \\
- 	&\leq \frac{p}{2n} + \frac{p(d+2)}{n (n - d -1)}
- 	\ , \forall n >d+1 \;.
-\end{align}
-where $p = \frac{d(d+1)}{2}$ and the last inequality used $\inv{2d}\leq 1$. 
-\end{proof}
-
-\subsection{Bounding the expected natural parameters for the MAP}
-\label{app:nat-bound}
-Before proving a convergence rate for the MAP, we need to bound the expectation of its inverse, hence the following lemma.
-We introduce the notation $(z)_+ = \max(0,z)$.
-\begin{lemma}[Expected MAP natural parameter]\label{lem:expected-map-natural-parameter-gaussian}
-	Define the variable $a = n_0 \frac{\mu_0}{\mu^*}$, which characterizes 
-	the importance of the prior relative to the true parameters.
-	The expectation of the natural parameter of a MAP of $\Gamma(\alpha,\beta)$ is bounded by
-\begin{equation}
-		\frac{n_0+n}{a+n}
-		=\frac{\mu^*}{\mu_n}
-		\leq \expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\nat^*}} 
-		\leq
-		\frac{n_0 +n}{a+(n-\inv{\alpha} )_+} \; , \forall n\geq 0 \; .
-		\label{eq:lemma_nat_bound}
-	\end{equation}
-\end{lemma}
-
-\begin{proof}
-	The lower bound can be readily obtained by applying Jensen's inequality 
-	to the convex function $x\mapsto\inv{x}$ for $x>0$.
-	The upper bound requires more work.
-	To start, let us plug in the definition of $\MAPm$
-\alignn{
-		\expect{\frac{\mu^*}{\MAPm}} 
-		= \expect{\frac{\MAPt}{\nat^*}} 
-		= \expect{\frac{(n_0+n) \mu^*}{n_0 \mu_0 + \sum_i X_i}}  
-		= \expect{\frac{n_0+n}{n_0 \frac{\mu_0}{\mu^*}+ \sum_i \frac{X_i}{ \mu^*} }} 
-		= \expect{\frac{n_0+n}{a + \Gamma(n\alpha, \alpha)}} \; ,
-}
-	where  $\sum_i \frac{X_i^2}{ \mu^*} \sim \Gamma(n\alpha, \alpha)$ is a gamma random variable and $a=n_0 \frac{\mu_0}{\mu^*}$. 
-	Further note that 
-\alignn{
-	\expect{\frac{n_0+n}{a + \Gamma(n\alpha, \alpha)}} 
-	= \frac{n_0 + n}{a} \expect{\frac{1}{1 + \Gamma(n \alpha , a \alpha) }} \; .
-}
-This kind of integrals can be expressed with {generalized exponential integral functions}
-	\begin{align}
-		E_k(z) = \int_1^\infty \frac{e^{-z t} }{t^k} dt \; ,
-	\end{align}
-with \href{http://dlmf.nist.gov/8.19.E4}{the formula} \citep[Eq.~8.19.4]{DLMF}
-\alignn{
-	\expect{\frac{1}{1 + \Gamma(\alpha , \beta) }}
-	= \frac{\beta^{\alpha}}{\Gamma(\alpha)} \int_0^\infty \frac{x^{\alpha-1}e^{-\beta x} }{1+x}  dx 
-	= \beta e^{\beta} E_\alpha (\beta) \; .
-	\label{eq:expect1+gamma}
-}
-Overall we get
-\alignn{
-\expect{\frac{\mu^*}{\MAPm}}  = (n_0 +n) \alpha e^{a \alpha} E_{n\alpha} ( a \alpha) 
-}
-	Now our goal is to bound this generalized exponential integral with simpler functions.
-	Fortunately, mathematicians have been working on these integrals for decades.
-	For instance , we have \href{https://dlmf.nist.gov/8.19.E21}{the general bound} \citep[Eq.~8.19.21]{DLMF}
-	\begin{align}
-		e^x E_k(x) 
-		&\leq \inv{x + k - 1} , \ \forall k>1\\
-		\iff \expect{\frac{\mu^*}{\MAPm}} 
-		&\leq \frac{(n_0 +n)\alpha}{a\alpha + n\alpha - 1}
-		= \frac{n_0 +n}{a + n - \inv{\alpha}} , \ \forall n >\inv{\alpha} \; .
-	\end{align}
-We are left with a special case when  $n\leq \alpha$. 
-Then we can use the trivial bound 
-\begin{align}
-	\expect{\inv{a + X^2}} < \inv{a}.
-	\label{eq:trivial_bound}
-\end{align}
-to conclude the proof.
-\end{proof}
-
-When $n<\inv{\alpha}$, it is possible to get a much tighter bound by exploiting \href{http://dlmf.nist.gov/8.19.E12}{the recurrence relationship} \cite[Eq.~8.19.12]{DLMF}
-\alignn{
-	\alpha E_{\alpha+1}(\beta) + \beta E_\alpha(\beta) 
-	= e^{-\beta} 
-}
-and combining it with the inequality \citep[Eq.~8.19.21]{DLMF} to get
-\alignn{
-	\beta e^\beta E_\alpha(\beta)
- 	= 1 - \alpha e^\beta E_{\alpha+1}(\beta)
- 	\leq 1 - \frac{\alpha}{\alpha + 1 + \beta}
- 	 = \frac{\beta + 1}{\alpha + \beta + 1} \; .
-}
-Plugging this inequality back into~\eqref{eq:expect1+gamma}, we get the following upper bound for the MAP:
-\alignn{
-	\expect{\frac{\mu^*}{\MAPm}} 
-	\leq \frac{n_0 + n}{a} \frac{a\alpha + 1}{n \alpha + a \alpha +1}
-	= \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}} \; .
-}
-Unfortunately, this formula does not yield an elegant convergence rate, so we keep it out of the lemma.
-
-\subsection{Proof of MAP Bound}
-\label{app:MAP-bound}
-
-We did not find a closed form or an upper bound for the expected logarithm of the MAP $\expect{\log\frac{\MAPm}{\mu^*}}$.
-Consequently, we derived a bound for the symmetrized Bregman~\eqref{eq:symmetrized_bregman} instead. 
-This bound is asymptotically tight for the Bregman, up to a factor $2$.
-
-There are several ways to write down the convergence rate.
-The Gaussian variance example can be written in a particularly simple form, so we give  it a special treatment in \S\ref{app:proof-MAP-gaussian}, corresponding to the theorem displayed in the main text.
-We make a more general statement about gamma distributions in \S\ref{app:proof-MAP-gamma}.
-
-\subsubsection{Gaussian Variance}
-\label{app:proof-MAP-gaussian}
-
-\begin{theorem}[MAP Bound]
-The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP of $\cN(  0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is upper bounded as
- \begin{align}
-	& \expect{\cS_\conj( \m^*; \hat \m_n^\mathrm{MAP})}
-	\leq 
-	\left\{\begin{array}{ll}
-		\cS_\conj( \mu_*; \mu_0) 					& \text{ if } \ n=0, \\
-		\inv{2(n_0+1)}  +  b_1 						& \text{ if }\ n=1,\\
-		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n  & \text{ if }\ n\geq 2,
-	\end{array}\right.
-	\quad 
-	\quad 
-	\text{where }
-	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{(n-2)_+}{n_0})(1 + \frac{n}{n_0} )} \in O\paren{\frac{n_0^2}{n^2}} \; . 
-	%\nonumber
-\end{align}
-\end{theorem}
-
-\begin{proof}
-When $n=0$, the inequality is an equality. 
-For $n>0$, we expand the symmetrized Bregman~\eqref{eq:symmetrized_bregman} with $\alpha=\half$ to get
-\begin{align}
-	\expect{\cS_\conj(\mu^*; \MAPm)} 
-	&\leq \half \paren{\expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1} \; .
-\end{align}
-The expectation of $\MAPm$ is straightforward
-\begin{align}
-	 \expect{\frac{\MAPm}{\mu^*}} - 1 
-	 = \frac{n_0 \mu_0 + n \mu^*}{(n_0+n)\mu^*} - 1 
-	 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
-	 \quadtext{where}
-	 a:=n_0\frac{\mu_0}{\mu^*} \; .
-\end{align}
-There remains the more problematic term with the expectation of the inverse mean parameter, 
-for which we use the bound derived in Lemma \ref{lem:expected-map-natural-parameter-gaussian}.
-
-\textbf{When $n\geq 2$}, we get 
-\begin{align}
-	\expect{\frac{\mu^*}{\MAPm}} - 1 
-	\leq \frac{n_0 + n}{a +n - 2} -1
-	= \frac{n_0 - a + 2}{a +n - 2}
-	 = \frac{2}{a +n - 2} + \frac{n_0 - a}{a +n - 2}
-\end{align}
-so putting it all together we get 
-\begin{align}
-	\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq \frac{1}{a+n-2} + \frac{n_0-a}{2}  \left( \inv{a+n-2} - \inv{n_0+n} \right)\\
-	&= \frac{1}{a+n-2} + \frac{(n_0-a + 1) - 1}{2}  \frac{(n_0 - a + 1) + 1}{(a+n-2)(n_0+n)} \\
-	&= \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2 - 1}{2(a+n-2)(n_0+n)} \\
-	&\leq \frac{1}{a+n-2} + \frac{(n_0 - a +1)^2}{2(a+n-2)(n_0 + n)} \\
-	&= \frac{1}{a+n-2} + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{2 (\frac{\mu_0}{\mu^*}+\frac{n-2}{n_0})(1 + \frac{n}{n_0} )} \; .
-\end{align}
-
-\textbf{When $n=1$} the bound~\eqref{eq:lemma_nat_bound}  on the expected natural parameter gives
-\begin{align}
-	\expect{\frac{\mu^*}{\MAPm}} - 1 
-	\leq \frac{n_0 + 1}{a} -1
-	= \frac{n_0 +1 - a}{a}
-\end{align}
-so putting it all together we get
-\begin{align}
-	2\expect{\bregmanconj(\mu^*; \hat \mu_1)} 
-	&\leq \frac{a - n_0 \pm 1}{n_0 + 1}  + \frac{n_0 +1 - a }{a} \\
-	& = \inv{n_0+1}  + ( a - n_0 - 1)(\inv{n_0+1} - \inv{a}) \\
-	& = \inv{n_0+1}  + \frac{(n_0 + 1 -a)^2}{a(n_0+1)}\\
-	&= \inv{n_0+1}  + \frac{(1 + \inv{n_0} - \frac{\mu_0}{\mu^*})^2}{\frac{\mu_0}{\mu^*}(1 + \inv{n_0} )} 
-\end{align}
-so we  recover the same bias term as when $n\geq2$.
-\end{proof}
-
-\subsubsection{Gamma with Known Shape}
-\label{app:proof-MAP-gamma}
-
-\begin{theorem}[MAP Bound]
-Consider an exponential distribution with sufficient statistics coming from a gamma distribution $\Gamma(\alpha, \beta^*)$ with mean parameter $\mu^* = \frac{\alpha}{\beta^*}$.
-The expected symmetrized Bregman~\eqref{eq:symmetrized_bregman} of the MAP  with prior hyper-parameters $(n_0,\mu_0)$ is upper bounded as
-\alignn{
-    \forall n\geq \inv{\alpha},\quad \quad
-	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
-	\leq\frac{1}{n_0+n} + 
-	\frac{\alpha\paren{\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} - 1 }^2 }{\paren{1+\frac{n}{n_0}}\paren{\frac{\mu_0}{\mu^*} +  \frac{n - \inv{\alpha}}{n_0}}}
-}
-\end{theorem} 
-Note that the second term vanishes when $\mu_0 = \mu^* \paren{1 + \inv{\alpha n_0}}$.
-Expressions for $n\alpha < 1$ are less elegant as shown below:
-\alignn{
- 	\expect{\cS_\conj(\mu^*, \hat \mu_n)}
-	\leq \frac{n_0 + n}{a + n +\inv{\alpha}} \paren{1 + \inv{a \alpha}} -1 + \frac{a - n_0}{n_0+n}
-	\quadtext{where}
-	 a:=n_0\frac{\mu_0}{\mu^*} \; .
-}
-
-\begin{proof}
-We expand the symmetrized Bregman~\eqref{eq:symmetrized_bregman} to get
-\begin{align}
-	\expect{\cS_\conj(\mu^*; \MAPm)} 
-	&\leq \alpha \paren{\expect{\frac{\mu^*}{\MAPm}} -1  + \expect{\frac{\MAPm}{\mu^*}} - 1} \; .
-\end{align}
-The expectation of $\MAPm$ is straightforward
-\begin{align}
-	 \expect{\frac{\MAPm}{\mu^*}} - 1 
-	 = \frac{n_0 \mu_0 + n \mu^*}{(n_0+n)\mu^*} - 1 
-	 = \frac{a+n}{n_0+n} - 1 = \frac{a - n_0}{n_0+n}
-	 \quadtext{where}
-	 a:=n_0\frac{\mu_0}{\mu^*} \; .
-\end{align}
-There remains the more problematic term with the expectation of the inverse mean parameter, 
-for which we use the bound derived in Lemma \ref{lem:expected-map-natural-parameter-gaussian}  when $n\alpha \geq 1$
-\begin{align}
-	\expect{\frac{\mu^*}{\MAPm}} - 1 
-	\leq \frac{n_0 + n}{a +n - \inv{\alpha}} -1
-	= \frac{n_0 - a + \inv{\alpha}}{a +n - \inv{\alpha}}
-\end{align}
-so putting it all together we get 
-\begin{align}
-	\expect{\bregmanconj(\mu^*; \MAPm)} 
-	&\leq 
-	\alpha \paren{\frac{n_0 - a + \inv{\alpha}}{a +n - \inv{\alpha}} 
-	 + \frac{a - n_0 \pm \inv{\alpha}}{n_0+n}} \\
-	 &=
-	 \alpha \paren{n_0 + \inv{\alpha} -a}  
-	 \left( \inv{a+n-\inv{\alpha}} - \inv{n_0+n} \right)
-	 + \inv{n_0 + n} \\
-	 &= \inv{n_0 + n} + \alpha \frac{(n_0 + \inv{\alpha} - a)^2}{(n_0 + n)(a+ n - \inv{\alpha} )} \\
-	&= \inv{n_0 + n} + \alpha \frac{(1 + \inv{\alpha n_0} - \frac{\mu_0}{\mu^*})^2}{ (1 + \frac{n}{n_0} ) (\frac{\mu_0}{\mu^*} - \inv{\alpha n_0} +\frac{n}{n_0})} \; .
-\end{align}
-\end{proof}
-
-\subsection{On the Choice of a Prior}
-\label{app:prior-choice}
-The optimal $\mu_0$ is larger than $\mu^*$ for small $n_0$ and small $n$.
-Indeed, the upper bound~\eqref{eq:MAP_rate} has a bias term that is $0$ when $\frac{mu_0}{\mu^*} = 1 + \inv{n_0}$, e.g.
-for large values of $n_0$, it is $\mu_0=\mu^*$ is the best prior, but for small $n_0$, one better sets larger values for $\mu_0$.
-In Figure~\ref{fig:optimal_n0}, we observe this behavior numerically.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=\textwidth]{figs/optimal_n0.pdf}
-	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $n_0, \mu_0$ spanning $[10^{-2}, 10^2]$.
-The expectation was estimated with $10^4$ draws for each value of $n_0, \mu_0$ We observe that the line $\frac{\mu_0}{\mu^*} = 1 + \inv{n_0}$ coincides with the bottom valley of this landscape.
-	}
-	\label{fig:optimal_n0}
-\end{figure}
-
-\section{COMPLEMENTS ON GAUSSIANS}
-\label{app:gaussian}
-
-%TODO ostroskii and bach are using a subgaussian assumption which does not hold for a Gaussian, so there might still be work to do. But also check anastasiou to be sure.
-
-In this section, we prove that for a Gaussian, the entropy and log-partition functions are self-concordant.
-We also provide complementary illustrations of 
-these functions in \cref{fig:logpart-entropy}, 
-their gradients (e.g. the mirror maps) in \cref{fig:mirrormaps},
-and paths taken by MLE and MAP in \cref{fig:gaussian-paths} .
-
-The Gaussian log-partition function and entropy are, up to constants,
-\alignn{
-	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \label{eq:app-logpart} \\
-	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; , \label{eq:app-entropy}
-}
-where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
-We provide definitions of self-concordance in \cref{app:self-concordant}.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.45\textwidth]{figs/2d/logpart.pdf}
-	\includegraphics[width=.45\textwidth]{figs/2d/entropy.pdf}
-	\caption{
-	(left) Contours of the log-partition function~\eqref{eq:app-logpart}.
-	(right) Contours of the entropy~\eqref{eq:app-entropy}.
-	}
-	\label{fig:logpart-entropy}
-\end{figure}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.9\textwidth]{figs/2d/mirrormap.pdf}
-	\caption{
-	(top) Grid deformation produce by $\nabla \logpart(\nat)$.
-	(bottom) Grid deformation produced by $\nabla\conj(\m)$.
-	}
-	\label{fig:mirrormaps}
-\end{figure}
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=.9\textwidth]{figs/2d/gaussian_paths.pdf}
-	\caption{
-		Paths taken by MLE (blue) and MAP (orange) on top of contours for $\KL(\nat^* || \nat)$. 
-		We set $\mu^*=(0, 1), \mu_0 = (1, 2)$, and $n_0=4$, and $n$ varies from $1$ to $20$.
-		In green, red and purple, we represent the paths respectively taken by the MAP dual expectation  $(\bar\nat_n, \bar \mu_n)$, MAP primal expectation $(\tilde\nat_n,\tilde\mu_n)$, and MLE primal expectation.
-Recall that the MLE dual expectation is $\mu^*$ itself.
-	}
-	\label{fig:gaussian-paths}
-\end{figure}
-
-\paragraph{Entropy is Self-Concordant.}
-\citet[Example 4.1.1.4,   p.177]{nesterov2003introductory} proves that logarithmic barriers for second order regions are self-concordant, 
-that is functions of the form 
-\alignn{
-	f(\theta) = -\log\paren{\alpha + \lin{a, \theta} - \frac{1}{2}\lin{\mA \theta, \theta}}
-	\text{ on }
-	\left\{\theta\in\mathbb{R}^n \cond \alpha + \lin{a, \theta} - \frac{1}{2}\lin{\mA \theta, \theta} > 0\right\}.
-}
-The entropy~\eqref{eq:app-entropy} fits into this definition with $\mA = \begin{pmatrix} 2 & 0 \\ 0 & 0 \end{pmatrix}$ and $a = ( 0 \ 1 )^\top$.
-
-\paragraph{Log-partition is Self-Concordant.}
-As proved in \citet{nesterov1994interior}, self-concordance is preserved by Fenchel conjugacy.
-Since $\conj$ is self-concordant, $\logpart$ is as well.
-For a more accessible reference, see also \citet[Prop.~6]{sun2019generalized}.
-
-\section{ASYMPTOTIC DERIVATION}
-\label{app:asymptote}
-In this section we fill-in the lines of \S\ref{ssec:asymptote} to prove Equation~\eqref{eq:asymptote}.
-Approximating $\conj$ with a second order Taylor expansion yields
-\aligns{
-    \bregmanconj(\m^* ; \m)
-    &= \frac{1}{2}\norm{\m - \m^*}^2_{\mF}
-    + O(\norm{\m - \m^*}^3),
-}
-where the norm is induced by the matrix
-\aligns{
-    \mF
-    :=\nabla^2\conj(\m^*)
-    = \nabla^2\logpart(\nat^*)^{-1}
-    = \Cov_{\nat^*}(T)^{-1},
-}
-where the second equality is a general property of convex conjugates.
-Plugging the MLE~\eqref{eq:defMLE} into this quadratic and expanding it yields
-\aligns{
-	\E \half \norm{\inv{n}  \sum_i T_i - \m^* }_\mF^2 
-	&=\inv{2n^2} \sum_i \E \norm{T_i - \m^* }_\mF^2 
-	+ \inv{2n^2} \sum_{i\neq j}\expect{T_i - \m^*}^\top \mF \overbrace{\expect{T_j - \m^*}}^{0}\\
-	&=\inv{2n}  \E \norm{T_1 - \m^* }_\mF^2 \\
-	& = \inv{2n} \Tr(\mF\ \expect{(T_1 - \m^*) (T_1 - \m^*)^\top}) \\
-	&= \inv{2 n} \Tr(\mF \Cov_{\nat^*}(T)) = \frac{d}{2n} ,
-}
-where on the first line we used independence of samples, on the second line we used the fact that samples are identically distributed, and on the third line we used the  trace trick along with the linearity of the trace $\Tr$.
-On the way, this also proves that for the MLE $\norm{\m - \m^*}^3 \in O(n^{-\half[3]})$.
-This yield the final rate for the MLE
-\begin{align}
-	\E \bregmanconj \paren{\E [T(X)] ; \inv{n}  \sum_i T_i}
-	= \frac{d}{2n} + O(n^{- \frac{3}{2}} ) \; .
-\end{align}
-For the MAP~\eqref{eq:defMAP}, the quadratic decomposes into bias and variance:
-\begin{align}
-	\E \half \norm{\mu^* -  \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }^2_{\mF}
-	&= \frac{n d}{2(n+n_0)^2}  +  \frac{n_0^2}{(n+n_0)^2} \half \norm{\mu^* -  \mu_0}^2_{\mF} \\
-	&= \frac{d}{2n} + O\left(\frac{1 + \norm{\mu^* -  \mu_0}^2_{\mF} }{n^2} \right) \; .
-\end{align}
-This $O(n^{-2})$ term is dominated by the $O(n^{-\half[3]})$ term from the quadratic approximation of the Bregman, yielding the same first order rate as for the MLE
-\begin{align}
-	\E \bregmanconj \paren{\E [T(X)] ; \frac{n_0 \mu_0 + \smallsum_i T(x_i)}{n_0+n} }
-	= \frac{d}{2n} + O(n^{- \frac{3}{2}} ) \; .
-\end{align}
-
-\section{SELF-CONCORDANCE}
-\label{app:self-concordant}
-
-In this section, we define self-concordance and we prove Proposition~\ref{prop:selfConcordant}.
-
-\begin{definition}[Self-concordance]
-\label{def:self-concordance}
-A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies for all feasible $t$
-\alignn{
-	\abs{g'''(t)} \leq 2 g''(t)^{\frac{3}{2}} \; .
-}
-\end{definition}
-
-%\begin{definition}[Pseudo Self-concordance] \citep{bach2010self} 
-%A convex function is $F:\real^p \rightarrow \real$ is self-concordant if it is differentiable $3$ times and if for all $w, v \in\real^p$ the function $g(t) = F(w+tv)$ satisfies 
-%\alignn{\abs{g'''(x)} \leq C \norm{v}_2 g''(x)}
-%for all feasible $t$ and for some $C\geq 0$.
-%\end{definition}
-
-{\bf Clarification:} In the main text in Section~\ref{ssec:local-quadratic}, we claimed that the Fenchel conjugate of a 1-dimensional function is also self-concordant. 
-Actually, this is also true in higher dimensions, as proved by \citet{nesterov1994interior}. See \citet[Prop.~6]{sun2019generalized} for a more accessible reference.
-
-\subsection{Properties of Self-concordant functions}
-
-We quickly review some important properties of self-concordant functions, introduced in \citep{nesterov2003introductory}.
-We start with a some notation.
-Let $A^*$ be a self-concordant function.
-Then, we write
-\begin{itemize}
-	\item the local norm $\|\cdot\|_x = \sqrt{ \langle \nabla^2A^*(x)\cdot,\, \cdot \rangle }$ 
-	\item the distance function $\omega(t) = t-\ln (1+t),\;t\geq 0$, and its dual $\;\omega^*(t) = -t-\ln(1-t)$ defined for $t\in [0,1]$.
-\end{itemize}
-Note that $\omega^*(t)$ is positive, convex and monotonically increasing for $t\in[0,1]$.
-
-\begin{figure}[ht]
-	\centering
-	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega.eps}
-	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega_star.eps}
-	\caption{(left) Graph of the distance function $\omega$ and its dual $\omega^*$.
-(right) Graph of the dual of distance function $\omega^*$ evaluated at $\frac{t}{1-t}$, compared with $t^2$.
-The two curves cross each other at $t\approx 0.21$.}
-	\label{fig:omega}
-\end{figure}
-
-We now present two important results. The first one shows how to convert the local norm $\|y-x\|_y$ using $\|y-x\|_x$.
-\begin{proposition}\label{prop:conversion_norm}
-	(Conversion of norms, \citep[Theorem 4.1.5]{nesterov2003introductory}) For any $x,\,y\in\dom A^*$, if $\|y-x\|_x<1$, then
-	\[
-		\|y-x\|_y \leq \frac{\|y-x\|_x}{1-\|y-x\|_x}.
-	\]
-\end{proposition}
-
-The next result shows that, if $y$ is sufficiently close to $x$, then we can bound the Bregman divergence of $A^*$ using the distance function $\omega^*$ and local norms.
-\begin{proposition}(Upper bound of self-concordant functions \citep[Theorem 4.1.8]{nesterov2003introductory}) \label{prop:upper_bound_self_concordance}
-	For any $x,\,y\in\dom A^*$, if $\|y-x\|_x<1$, then
-	\[
-		\bregmanconj(y,x) \leq \omega^*(\|y-x\|_x).
-	\]
-\end{proposition}
-
-We are now ready to prove Proposition~\ref{prop:selfConcordant}.
-
-\subsection{Proof of Proposition~\ref{prop:selfConcordant}.}
-
-We start with Proposition~\ref{prop:upper_bound_self_concordance}, evaluated at $y=\meanp^*$ and $x=\meanp$:
-\[
-	\bregmanconj(\meanp^*,\meanp) \leq \omega^*(\|\meanp^*-\meanp\|_{\meanp}).
-\]
-This hold if $\|\meanp^*-\meanp\|_{\meanp}<1$.
-Since $\omega^*$ is monotonically increasing, we can replace $\|\meanp^*-\meanp\|_{\meanp}$ by its upper bound from Proposition~\ref{prop:conversion_norm},
-\[
-	\omega^*(\|\meanp^*-\meanp\|_{\meanp}) \leq \omega^*\left(\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}}\right),
-\]
-under the conditions that $\|\meanp^*-\meanp\|_{\meanp^*}<1$ (to satisfy the assumption of Proposition~\ref{prop:conversion_norm}) and $\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}} <1$ (to ensure that $\|\meanp^*-\meanp\|_{\meanp} <1$).
-Those two conditions holds if $\|\meanp^*-\meanp\|_{\meanp^*} <0.5$.
-Now, we use the bound (see figure \ref{fig:omega})
-\[
-	\omega^*\left(\frac{t}{1-t}\right) \leq t^2, \quad 0\leq t \leq 0.21,
-\]
-and replace $t$ by $\|\meanp^*-\meanp\|_{\meanp^*}$. This finally gives the sequence of inequalities
-\[
-	\bregmanconj(\meanp^*,\meanp) \leq \omega^*(\|\meanp^*-\meanp\|_{\meanp}) \leq \omega^*\left(\frac{\|\meanp^*-\meanp\|_{\meanp^*}}{1-\|\meanp^*-\meanp\|_{\meanp^*}}\right) \leq \|\meanp^*-\meanp\|_{\meanp^*}^2,
-\]
-that holds while $ \|\meanp^*-\meanp\|_{\meanp^*}<0.21$, which is the desired result.
-
-\section{BIAS-VARIANCE}
-\label{app:bias-variance}
-In this section, we start from the notions of bias and variance introduced in \cref{eq:bias-variance}.
-First, we prove that the bias of the MLE of a Gaussian variance decreases in $O(\inv{n^2})$. 
-Then we prove that assuming~\eqref{eq:hanzely} holds, whether uniformly or in expectation, yields a convergence rate on the variance term.
-
-\subsection{Bias of a Gaussian Variance MLE}
-For a Gaussian variance model, the MLE follows a scaled $\chi^2(n)$ distribution. 
-This means that
-\alignn{
-	\frac{\m^*}{\tilde \m_n} 
-	= \frac{\tilde\nat_n}{\nat^*}
-	=\expect{\frac{\hat\nat_n}{\nat^*}}
-	=\expect{\frac{\m^*}{\hat \m_n} }
-	=\expect{\frac{n}{\chi^2(n)}}
-	= \frac{n}{n-2}\; .
-}
-Consequently,
-\alignn{
-	\bregmanconj(\m^*, \tilde \m_n) 
-	&= \half \paren{\frac{n}{n-2} -1 + \log\frac{n-2}{n}}\\
-	&= \frac{1}{n-2} + \half \log\paren{1 - \frac{2}{n}}\\
-	&\leq \inv{n-2} - \inv{n} 
-	= \frac{2}{n(n-2)},
-}
-so the bias of the MLE of a Gaussian Variance decreases like $O\paren{\inv{n^2}}$.
-
-\subsection{Expectation of SMD's Variance Assumption}
-The first step is to notice the symmetrized Bregman can be expressed as an inner product between primal and dual parameters
-\alignn{
-	\cS_\conj (\mu , \bar\mu)
-	= \lin{ \nabla \conj(\mu) - \nabla \conj(\bar\mu) , \mu - \bar\mu}
-	= \lin{ \nat -\bar\nat , \mu - \bar\mu} \; .
-}
-Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_{n+1} = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_{n+1} =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
-For such a pair of variables, the expectation of the symmetrized Bregman corresponds to a covariance between primal and dual parameters
-\alignn{
-	\expect{\cS_\conj (\mu , \E[\mu])} 
-	= \expect{\lin{\nat, \mu - \E[\mu]}}
-	= \underbrace{\expect{\lin{\nat, \mu}} -\lin{\expect{\nat}, \expect{\mu} } }_{\Cov(\nat, \mu)}
-	=\expect{\lin{\nat - \E[\nat],\mu}} 
-	= \expect{\cS_\logpart(\nat, \E[\nat])} \; .
-}
-The last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
-Note that the middle covariance formulation is actually the one  used by \citet{hanzely2018fastest}.
-Now, \cref{eq:bias-variance} defines the variance as 
-\alignn{
-	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
-	=\expect[1:n]{\bregman(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
-	\leq \expect[1:n]{\cS_A(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
-	=  \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }\;,
-	\label{eq:bregman-covariance}
-}
-where expectations $\E_{1:n}$ are on all samples $X_1, \dots, X_n$, whereas~\eqref{eq:hanzely} is written with 
-\alignn{
-	\expect[n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) } \leq \lr^2 C \; ,
-}
-where the expectation $\E_{n}$ is taken over only the last sample $X_n$, and the bound should hold uniformly over all $\hat\mu_{n-1}$.
-Taking the expectation over $\hat\mu_{n-1}$ instead gives
-\alignn{
-	\expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \leq \lr^2 C \;.
-}
-The only difference with the RHS of \cref{eq:bregman-covariance} is in the inner expectation.
-To overcome this difference, we need to plug in the form of $\hat \mu_n = \frac{n_0\mu_0 + \sum_i T_i}{n_0 +n}$. Notice that 
-\alignn{
-	&\MAPm - \E_n[\MAPm] = \frac{T_n - \m^*}{n_0+n} \\
-	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }
-	= \inv{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} },
-}
-while
-\alignn{
-	&\MAPm - \E_{1:n}[\MAPm] = \frac{\smallsum_i (T_i - \m^*)}{n_0+n} \\
-	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }
-	=\frac{n}{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} } \; .
-}
-In the end, we get that the variance is dominated by $n$ times the expectation of \cref{eq:hanzely}:
-\alignn{
-	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
-	\leq n \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \; .	
-}
-If assumption~\eqref{eq:hanzely} holds, then we have
-\alignn{
-\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
-\leq n \lr_n^2 C \in O\paren{\inv{n}} \; ,
-}
-where we assumed $\lr_n  \in O\paren{\inv{n}}$.
-In conclusion, assuming~\eqref{eq:hanzely} holds uniformly or in expectation immediately implies a $O\paren{\inv{n}}$ convergence rate on the variance.
-
-\section{REVIEW OF SMD}
-\label{app:SMD}
-
-We use this section to give more details on the (stochastic) mirror descent algorithm.
-We start with gradient descent with step-sizes $\gamma$. 
-the update $\theta_{n+1} = \theta_n - \gamma \nabla f(\theta_n)$ can be viewed as the minimization of the linear approximation of $f$ at $\theta_n$
-$f(\theta) \approx f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n}$, 
-alongside with quadratic penalty scaled by ${1}/{\gamma}$:
-\alignn{
-	\theta_{n+1} = \arg\min_\theta f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n} + \frac{1}{\gamma} \frac{1}{2}\norm{\theta - \theta_n}^2.
-}
-Mirror descent generalizes the above, using the Bregman divergence induced by a (Legendre) function $A$ instead of the Euclidean norm as follow, 
-\alignn{\label{eq:mirror-descent-primal}
-	\theta_{n+1} = \arg\min_\theta f(\theta_n) + \lin{\nabla f(\theta_n), \theta - \theta_n} + \frac{1}{\gamma} \bregman(\theta, \theta_n).
-}
-Mirror descent coincides with gradient descent if $A(\theta) = \frac{1}{2}\norm{\theta}^2$.
-As \cref{eq:mirror-descent-primal} is convex, the minimum is at a stationary point, 
-found by taking the derivative and setting to 0, 
-leading to the update $\theta_{n+1}$ satisfying
-\alignn{
-	\nabla f(\theta_n) + \frac{1}{\gamma}\paren{\nabla A(\theta_{n+1}) - \nabla A(\theta_n)} = 0
-	&&\implies&&
-	\nabla A(\theta_{n+1}) = \nabla A(\theta_n) - \gamma \nabla f(\theta_n).
-}
-Expressed with the dual parameters, we obtain $\mu_{n} = \nabla A(\theta_{n})$, 
-$\mu_{n+1} = \mu_n - \gamma \nabla f(\theta_n)$.
-
-{\bf In our case,} where the objective function is the (negative) log-likelihood of an exponential family, we have
-\alignn{
-	f(\theta) = A(\theta) - \lin{\frac{1}{n} \sum_{i=1}^n T(X_i), \theta}.
-}
-Using Mirror descent with a step-size of $1$ and the log-partition function $A$ as the reference function gives
-\alignn{\begin{aligned}
-	\mu_{n+1} 
-	= \mu_n - \nabla f(\theta_n)
-	= \mu_n - \paren{\nabla A(\theta_n) - \frac{1}{n}\sum_{i=1}^n T(x_i)}
-	= \frac{1}{n}\sum_{i=1}^n T(x_i).
-	\end{aligned}
-}
-In a stochastic, online version 
-where the linearization of the objective is obtained from iid samples, 
-a decreasing step-size of $\gamma_n = 1/n$ recovers the ``online'' estimate of the MLE.
-The case of $\mu_1 = T(x_1)$ follows from the above, 
-and in general, assuming it holds for $\mu_n$, 
-\alignn{\begin{aligned}
-	\mu_{n+1} &= \mu_n - \gamma_n g(\theta_n)
-	= \mu_n - \gamma_n (\mu_n - x_n)
-	\\
-	&= (1 - \frac{1}{n}) \mu_n + \frac{1}{n} T(x_n)
-	=
-	\frac{n-1}{n} \frac{1}{n-1}\sum_{i=1}^{n-1} T(x_i)
-	= 
-	\frac{1}{n} \sum_{i=1}^{n} T(x_i)
-	+ \frac{1}{n} T(x_n).
-\end{aligned}}
-The derivation in the main text gives the more general result, 
-of using step-sizes of the form $1/(n+n_0)$ to recover online MAP estimation 
-with a conjugate prior depending on $n_0$ and the initial estimate of the parameters $\theta_0$.
-
-\iffalse
-\section{TODO}
-\begin{enumerate}
-	\item new plots: add 2d trajectories to paper, with background level set, and new legend: primal/dual expectation, and clear explanation!
-	\item Incorporate related work by Frederik.
-	\item Technical Background: be precise and add a reference for the steep exponential family of Legendre type, which guarantees that our sets are open and have a solution in the middle.
-	References: \begin{itemize}
-		\item \href{https://www.jstor.org/stable/4616462?seq=1#metadata_info_tab_contents}{Existence of Maximum Likelihood Estimates for Multi-Dimensional Exponential Families},
-		\item The singly truncated normal distribution: A non-steep exponential family,
-		\item Statistical exponential families: A digest with flashcards,
-		\item Information and Exponential Families In Statistical Theory,
-		\item Fundamentals of Statistical Exponential Families with Applications in Statistical Decision Theory.
-	\end{itemize}
-\end{enumerate}
-
-\section{Blurbs/ideas?}
-
-\subsection{SGD blurb}
-\fdk{
-%
-There is a long line of work bounding the convergence rate of stochastic gradient descent.
-Beyond the results of \citet{robbins1951stochastic},
-we now have proofs on smooth, strongly convex problems and bounded variance
-(as well as other assumptions on the noise, reviewed later)
-a decreasing step-size and an averaging scheme gets
-a $1/t$ convergence rate,
-which matches the asymptotic rate of unbiased estimation,
-through clever averaging schemes \citep{rakhlin2012making,lacostejulien2012simpler}
-\\
-Those papers assume the stochastic gradients are bounded,
-but minor modifications also work for bounded variance instead.
-\\
-We do have extensions to other notions of bounded variance,
-for example, assuming that the stochastic gradients are gradients of a perturbed smooth function $f_i$
-and that the minimum of $f$ and the minima of the $f_i$
-are bounded, $\expect{\min_x f(x) - \min_x f_i(x)} \leq \sigma^2$,
-or that the gradient noise is bounded only at the minimum.
-For a review, see \citet{gower2019sgd}.
-\\
-However, for some problems, combining decreasing step-sizes and averaging is not necessary,
-even when the function is not strongly convex.
-This is the case for maximum likelihood estimation and matches the asymptotic rates,
-but holds more generally, for example, for linear and logistic regressions \citep{bach2013nonstronglyconvex,moulines2011non}.
-}
-
-\subsection{Poisson likelihood?}
-\fdk{
-\citet{bauschke2017descent} and \citet{hanzely2018fastest} both use the example of Poisson inverse problems/Poisson regression
-as examples.
-The simpler case of the MLE of a Poisson distribution is also unsolved, though.
-In this case, $h(x) = 1/x!$, $x \in \mathbb{N}$, $A(\theta) = e^\theta$ over $\theta \in \mathbb{R}$,
-$A^*(\m) = \m \log \m - \m$ over $\m \in \mathbb{R}_+$.
-It is neither strongly-convex nor self-concordant (at least according to the standard definition,
-although it satisfies generalized notions of self-concordance as $A'''(\theta) = A''(\theta)$).
-}
-
-\subsection{``simple'' open problem?}
-\fdk{
-A ``simple open problem'';
-assume we have a deterministic problem so that we know where the minimum is.
-Can we figure out a (deterministic) path from $\theta_0$ to $\theta_*$
-that has optimality decreasing as $1/n^2$,
-to mirror the bias term of gradient descent?
-}
-
-\subsection{Connection with acceleration?}
-\fdk{
-The problems on how to deal with Bregman divergences abound in optimization, beyond stochasticity.
-For example, we have not yet figured out the analog of Nesterov-type acceleration
-on relatively smooth and strongly convex problems
-to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$,
-or just from $1/T$ to $1/T^2$ in the (non-strongly) convex case.
-\citet{dragomir2021optimal}  shows that naïve application of Bregman updates can not achieve acceleration.
-The tools developed to make progress on one problem might help make progress on the other.
-}
-
-\newpage
-\section{Notation convention}
-We use
-\begin{itemize}
-	\item $\theta, \nat$ for natural parameters
-	\item $\m$ for mean parameters
-	\item $X_1,\ldots,X_n$ for data
-	\item $\cX$ for the data space
-	\item $T$ for the sufficient statistics
-	\item $\nu$ for the base measure
-	\item $A, A^*, \logpart, \conj$ for the log-partition and its dual
-	\item $\cB_A$ for the Bregman divergence induced by $A$
-	\item $n$ for the number of samples
-	\item $(\theta^*,m^*)$ for the optimum
-	\item $\gamma, \lr$ for the step-size
-	\item $\mu,\sigma^2$ for the parameters of a Gaussian
-	\item $\hat\theta_n$ for estimates
-	\item $\tilde\theta_n$ for $\expect{\hat\theta_n}$, which is biased
-	\item $(\stgcvx , \smooth)$ for strong-convexity and smoothness
-\end{itemize}
-\fi
-
- \end{document}

### 2022-01-26 colt/colt-stylefile/colt2022-sample.tex
created colt submission + cleanup files

\documentclass[anon,12pt]{colt2022} % Anonymized submission
%\documentclass[final,12pt]{colt2022} % Include author names

% The following packages will be automatically loaded:
% amsmath, amssymb, natbib, graphicx, url, algorithm2e

\title[Short Title]{Full Title of Article}
\usepackage{times}
% Use \Name{Author Name} to specify the name.
% If the surname contains spaces, enclose the surname
% in braces, e.g. \Name{John {Smith Jones}} similarly
% if the name has a "von" part, e.g \Name{Jane {de Winter}}.
% If the first letter in the forenames is a diacritic
% enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}

% Two authors with the same address
% \coltauthor{\Name{Author Name1} \Email{abc@sample.com}\and
%  \Name{Author Name2} \Email{xyz@sample.com}\\
%  \addr Address}

% Three or more authors with the same address:
% \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
%  \Name{Author Name2} \Email{an2@sample.com}\\
%  \Name{Author Name3} \Email{an3@sample.com}\\
%  \addr Address}

% Authors with different addresses:
\coltauthor{%
 \Name{Author Name1} \Email{abc@sample.com}\\
 \addr Address 1
 \AND
 \Name{Author Name2} \Email{xyz@sample.com}\\
 \addr Address 2%
}

\begin{document}

\maketitle

\begin{abstract}%
  This is a great paper and it has a concise abstract.%
\end{abstract}

\begin{keywords}%
  List of keywords%
\end{keywords}

\section{Introduction}

This is where the content of your paper goes.
\begin{itemize}
  \item Limit the main text (not counting references and appendices) to 12 PMLR-formatted pages, using this template. Please add any additional appendix to the same file after references - there is no page limit for the appendix.
  \item Include, either in the main text or the appendices, \emph{all} details, proofs and derivations required to substantiate the results.
  \item The contribution, novelty and significance of submissions will be judged primarily based on
\textit{the main text of 12 pages}. Thus, include enough details, and overview of key arguments, 
to convince the reviewers of the validity of result statements, and to ease parsing of technical material in the appendix.
  \item Use the \textbackslash documentclass[anon,12pt]\{colt2022\} option during submission process -- this automatically hides the author names listed under \textbackslash coltauthor. Submissions should NOT include author names or other identifying information in the main text or appendix. To the extent possible, you should avoid including directly identifying information in the text. You should still include all relevant references, discussion, and scientific content, even if this might provide significant hints as to the author identity. But you should generally refer to your own prior work in third person. Do not include acknowledgments in the submission. They can be added in the camera-ready version of accepted papers.

  Please note that while submissions must be anonymized, and author names are withheld from reviewers, they are known to the area chair overseeing the paper’s review.  The assigned area chair is allowed to reveal author names to a reviewer during the rebuttal period, upon the reviewer’s request, if they deem such information is needed in ensuring a proper review.  
  \item Use \textbackslash documentclass[final,12pt]\{colt2022\} only during camera-ready submission.
\end{itemize}

% Acknowledgments---Will not appear in anonymized version
\acks{We thank a bunch of people and funding agency.}

\bibliography{yourbibfile}

\appendix

\section{My Proof of Theorem 1}

This is a boring technical proof.

\section{My Proof of Theorem 2}

This is a complete version of a proof sketched in the main text.

\end{document}

### 2022-01-26 colt/paper_content_backup.tex
created colt submission + cleanup files

--- 2022-01-07 aistats.tex
@@
 This rule is also known as moment matching.
-Given a specific conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
-In this paper, we will focus on analyzing MLE and MAP estimators.\footnote{A related analysis is present in the online-learning literature, but for different online estimators, which are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.}
+Given a specific conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP). In this paper, we will focus on analyzing MLE and MAP estimators.
 
@@
 In this paper, we review partial solutions and give ideas on how to solve the problem.
+
+\paragraph{Online Learning.}
+Variants of this problem have been considered by the online learning community.
+For instance \citet[Th.~9-10]{kaufmann2021mixture} provides high-confidence upper bounds on~\eqref{problem}, but only for Gaussians with known variance or Gamma with known shape.
+Other works compare other online estimators to MLE or MAP \citep{azoury2001relative,dasgupta2007online}.
 

### 2022-01-27 aistats.tex
Remove deprecated comments from aistats

@@
 
-%\pdfsuppresswarningpagegroup=1 %removes the warning
-% This is a hack, the other way to remove is to convert all pdf into postscript or eps.
-% RLP: not sure what is the point of this. With my latex it creates an error, and I do not have warnings, so I comment it out for now.
-
 \usepackage[accepted]{aistats/aistats2022}
-% If your paper is accepted, change the options for the package
-% aistats2022 as follows:
-%\usepackage[accepted]{aistats2022}
-
 \setlength{\footskip}{3.31pt}
 
-% If you set papersize explicitly, activate the following three lines:
-%\special{papersize = 8.5in, 11in}
-%\setlength{\pdfpageheight}{11in}
-%\setlength{\pdfpagewidth}{8.5in}
-
-%\let\oldsection\section
-%\renewcommand{\section}[1]{\oldsection{\texorpdfstring{\uppercase{#1}}{#1}}}
-
-% If you use natbib package, activate the following three lines:
 \usepackage[round]{natbib}

### 2022-01-27 COLT2022/colt2022.tex
Add files for COLT. Many errors left

--- 2022-01-26 colt/paper_content_backup.tex
@@
-\documentclass[twoside]{article}
+%\documentclass[anon,12pt]{colt2022} % Anonymized submission
+\documentclass[final,12pt]{colt2022} % Include author names
+% The following packages are automatically loaded:
+% amsmath, amssymb, natbib, graphicx, url, algorithm2e
+\usepackage{times}
 
@@
 
-%\pdfsuppresswarningpagegroup=1 %removes the warning
-% This is a hack, the other way to remove is to convert all pdf into postscript or eps.
-% RLP: not sure what is the point of this. With my latex it creates an error, and I do not have warnings, so I comment it out for now.
-
-\usepackage[accepted]{aistats/aistats2022}
-% If your paper is accepted, change the options for the package
-% aistats2022 as follows:
-%\usepackage[accepted]{aistats2022}
-
-\setlength{\footskip}{3.31pt}
-
-% If you set papersize explicitly, activate the following three lines:
-%\special{papersize = 8.5in, 11in}
-%\setlength{\pdfpageheight}{11in}
-%\setlength{\pdfpagewidth}{8.5in}
-
-%\let\oldsection\section
-%\renewcommand{\section}[1]{\oldsection{\texorpdfstring{\uppercase{#1}}{#1}}}
-
-% If you use natbib package, activate the following three lines:
-\usepackage[round]{natbib}
-\renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsubsection*{\bibname}}
-
 %\usepackage{microtype} % RLP tried it, no change.
@@
 \usepackage{lmodern}
-\usepackage{url}            % simple URL typesetting
 \usepackage{booktabs}       % professional-quality tables
 \usepackage{microtype}      % microtypography
-\usepackage{graphicx}
-\graphicspath{{figs/}}
+\graphicspath{{../figs/}}
 %\usepackage{subfigure}
@@
 \usepackage{hyperref}       % hyperlinks
-\usepackage[dvipsnames]{xcolor}
+%\usepackage[dvipsnames]{xcolor}
 
@@
 % some custom math commands
-\newtheorem{proposition}{Proposition}
 \newtheorem{problem}{Open Problem}
@@
 
-\begin{document}
-
-% If your paper is accepted and the title of your paper is very long,
-% the style will print as headings an error message. Use the following
-% command to supply a shorter title of your paper so that it can be
-% used as headings.
-%
-\runningtitle{Convergence Rates for the MAP of an Exponential Family and SMD -- an Open Problem}
-
-% If your paper is accepted and the number of authors is large, the
-% style will print as headings an error message. Use the following
-% command to supply a shorter version of the authors names so that
-% they can be used as headings (for example, use only the surnames)
-%
-\runningauthor{R\'emi Le Priol, Frederik Kunstner, Damien Scieur, and Simon Lacoste-Julien}
-
-\twocolumn[
-
-\aistatstitle{Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem}
-
-\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner}
-\aistatsaddress{ 
-	Mila, Université de Montreal 
-	\And  University of British Columbia
-	}
-
-\aistatsauthor{Damien Scieur \And Simon Lacoste-Julien}
-\aistatsaddress{Samsung, SAIT AI Lab, Montreal 
-	\And Mila, Université de Montreal \\
+% Use \Name{Author Name} to specify the name.
+% If the surname contains spaces, enclose the surname
+% in braces, e.g. \Name{John {Smith Jones}} similarly
+% if the name has a "von" part, e.g \Name{Jane {de Winter}}.
+% If the first letter in the forenames is a diacritic
+% enclose the diacritic in braces, e.g. \Name{{\'E}louise Smith}
+
+% Two authors with the same address
+% \coltauthor{\Name{Author Name1} \Email{abc@sample.com}\and
+%  \Name{Author Name2} \Email{xyz@sample.com}\\
+%  \addr Address}
+
+% Three or more authors with the same address:
+% \coltauthor{\Name{Author Name1} \Email{an1@sample.com}\\
+%  \Name{Author Name2} \Email{an2@sample.com}\\
+%  \Name{Author Name3} \Email{an3@sample.com}\\
+%  \addr Address}
+
+% Authors with different addresses:
+\coltauthor{%
+ \Name{R\'emi {Le Priol}} \Email{remi.lp.17@gmail.com}\\
+ \addr Mila, Université de Montréal
+ \AND
+ \Name{Damien Scieur} \Email{damien.scieur@gmail.com}\\
+ \addr Samsung, SAIT AI Lab, Montreal
+ \AND
+ \Name{Frederik Kunstner} \Email{frederik.kunstner@gmail.com}\\
+ \addr University of British Columbia
+ \AND
+ \Name{Simon Lacoste-Julien} \Email{slacoste@iro.umontreal.ca}\\
+ \addr Mila, Université de Montreal \\
 	Samsung, SAIT AI Lab, Montreal\\
@@
 }
-
-]
+\title[%Short Title
+	Convergence Rates for the MAP of an Exponential Family and SMD -- an Open Problem
+	]{%Full Title of Article
+	Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem
+	}
+
+\begin{document}
+
+\maketitle
 
@@
 
+\begin{keywords}%
+  List of keywords%
+\end{keywords}
+
 \doparttoc % Tell to minitoc to generate a toc for the parts
@@
 
-\begin{figure}[t]
+\begin{figure}
 	\centering
-\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
+\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
 	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
@@
 	\centering
-	\includegraphics[width=.8\textwidth]{figs/thales/numerical_schema_n=3.pdf}
+	\includegraphics[width=.8\textwidth]{thales/numerical_schema_n=3.pdf}
 	\caption{
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
+	\includegraphics[width=.4\textwidth]{gaussians/new_linear_n0=1.pdf}
 	\caption{
@@
 
-\subsubsection*{Acknowledgments}
+\acks{ % Acknowledgments---Will not appear in anonymized version
 We express our gratitude to Reza Babanezhad, Radu Dragomir and Hadrien Hendrikx for their insights on stochastic mirror descent. 
@@
 Simon Lacoste-Julien is a CIFAR Associate Fellow in the Learning in Machines \& Brains program.
+}
 
 \bibliographystyle{apalike}
-\bibliography{references}
+\bibliography{../references}
 
@@
 \appendix
-\onecolumn
 
@@
 	\centering
-	\includegraphics[width=\textwidth]{figs/optimal_n0.pdf}
+	\includegraphics[width=\textwidth]{optimal_n0.pdf}
 	\caption{Contours of $\expect{\KL(\nat^*, \hat \nat_n)}$  for the Gaussian variance with $n\in\{1,5,10\}$, $\mu^*=1$ and $n_0, \mu_0$ spanning $[10^{-2}, 10^2]$.
@@
 	\centering
-	\includegraphics[width=.45\textwidth]{figs/2d/logpart.pdf}
-	\includegraphics[width=.45\textwidth]{figs/2d/entropy.pdf}
+	\includegraphics[width=.45\textwidth]{2d/logpart.pdf}
+	\includegraphics[width=.45\textwidth]{2d/entropy.pdf}
 	\caption{
@@
 	\centering
-	\includegraphics[width=.9\textwidth]{figs/2d/mirrormap.pdf}
+	\includegraphics[width=.9\textwidth]{2d/mirrormap.pdf}
 	\caption{
@@
 	\centering
-	\includegraphics[width=.9\textwidth]{figs/2d/gaussian_paths.pdf}
+	\includegraphics[width=.9\textwidth]{2d/gaussian_paths.pdf}
 	\caption{
@@
 	\centering
-	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega.eps}
-	\includegraphics[width=0.4\textwidth]{figs/self_concordance/omega_star.eps}
+	\includegraphics[width=0.4\textwidth]{self_concordance/omega.eps}
+	\includegraphics[width=0.4\textwidth]{self_concordance/omega_star.eps}
 	\caption{(left) Graph of the distance function $\omega$ and its dual $\omega^*$.

### 2022-01-27 COLT2022/colt2022.tex
Fix all errors and new warnings

@@
+\RequirePackage[dvipsnames]{xcolor}
 %\documentclass[anon,12pt]{colt2022} % Anonymized submission
@@
 \usepackage{hyperref}       % hyperlinks
-%\usepackage[dvipsnames]{xcolor}
 
@@
 We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
-{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$.
+\textbf{1)} Distance in parameter space $d(\nat,\nat^*)$.
 This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
 In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
-{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
+\textbf{ 2)} Distance between distributions, as studied in \emph{density estimation}.
 For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
@@
 
-{\bf Duality.}
+\textbf{ Duality.}
 The log-partition function $\logpart$ verifies the two following identities:
@@
 %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
-{\bf The Bregman Divergence} induced by $\logpart$ measures the discrepancy between two parameters $\nat$ and $\nat_0$,
+\textbf{ The Bregman Divergence} induced by $\logpart$ measures the discrepancy between two parameters $\nat$ and $\nat_0$,
 \begin{align}
@@
 
-{\bf A Conjugate Prior} for $p(X|\nat)$ is
+\textbf{ A Conjugate Prior} for $p(X|\nat)$ is
 \begin{align}
@@
 
-{\bf Maximum A Posteriori (MAP).}
+\textbf{ Maximum A Posteriori (MAP).}
 Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
@@
 
-{\bf A Difficulty with the MLE.}
+\textbf{ A Difficulty with the MLE.}
 While~\eqref{eq:bregmanMAP} is always finite, \eqref{eq:bregmanMLE} may be infinite,
@@
 
-{\bf If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
+\textbf{ If $\conj$ is $L$-Lipschitz} (e.g. $\logpart$ is defined within the $\ell^2$-ball of radius $L$), then
 \begin{align}
@@
 
-{\bf If $\conj$ is $L$-smooth}\footnote{$\conj$ is $L$-smooth iff $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
+\textbf{ If $\conj$ is $L$-smooth}\footnote{$\conj$ is $L$-smooth iff $\nabla\conj$ is $L$-Lipschitz.} (e.g. $\logpart$ is $\frac{1}{L}$-strongly convex \citep{kakade2009duality}), then
 \begin{align}
@@
 
-{\bf Clarification:} In the main text in Section~\ref{ssec:local-quadratic}, we claimed that the Fenchel conjugate of a 1-dimensional function is also self-concordant. 
+\textbf{ Clarification:} In the main text in Section~\ref{ssec:local-quadratic}, we claimed that the Fenchel conjugate of a 1-dimensional function is also self-concordant. 
 Actually, this is also true in higher dimensions, as proved by \citet{nesterov1994interior}. See \citet[Prop.~6]{sun2019generalized} for a more accessible reference.
@@
 
-{\bf In our case,} where the objective function is the (negative) log-likelihood of an exponential family, we have
+\textbf{ In our case,} where the objective function is the (negative) log-likelihood of an exponential family, we have
 \alignn{

### 2022-01-27 aistats-submission/rebuttal-aistats/discussion.tex

% !TEX root = aistats2022_author_response.tex

\SLJ{%
Yes, kudos, all the reviewers seem to have found the paper interesting and have appreciated the writing! :) The only reason for the 4 and 5 is that they think that it is not a "completed result" and are unsure whether this is suitable.
~\\~\\
As an AC in machine learning, I disagree, and indeed there is no such notion about this on the call for papers. And importantly, there is the emphasis on "promoting the exchange of ideas" about ML and statistics in the summary for call for papers, nowhere does it say that these ideas should be "complete results"...
~\\~\\
You could also mention in the rebuttal is that each time you talked to people in optimization or statistics, they were interested in knowing the answers to these questions, and so this is why we decided to publish as is, to stimulate interest in these directions and get the help of the community to solve this hard problem...
}

### 2022-01-27 colt/paper_content.tex

--- 2022-01-26 colt/paper_content_backup.tex
@@
-\documentclass[twoside]{article}
-
-\usepackage{silence}
-% Disable all warnings issued by latex starting with "You have..."
-% This is a hack, there is no clean solution for that unless we install locally the package
-\WarningFilter{latex}{You have requested package}
-
-%\pdfsuppresswarningpagegroup=1 %removes the warning
-% This is a hack, the other way to remove is to convert all pdf into postscript or eps.
-% RLP: not sure what is the point of this. With my latex it creates an error, and I do not have warnings, so I comment it out for now.
-
-\usepackage[accepted]{aistats/aistats2022}
-% If your paper is accepted, change the options for the package
-% aistats2022 as follows:
-%\usepackage[accepted]{aistats2022}
-
-\setlength{\footskip}{3.31pt}
-
-% If you set papersize explicitly, activate the following three lines:
-%\special{papersize = 8.5in, 11in}
-%\setlength{\pdfpageheight}{11in}
-%\setlength{\pdfpagewidth}{8.5in}
-
-%\let\oldsection\section
-%\renewcommand{\section}[1]{\oldsection{\texorpdfstring{\uppercase{#1}}{#1}}}
-
-% If you use natbib package, activate the following three lines:
-\usepackage[round]{natbib}
-\renewcommand{\bibname}{References}
-\renewcommand{\bibsection}{\subsubsection*{\bibname}}
-
-%\usepackage{microtype} % RLP tried it, no change.
-%\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{lmodern}
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{microtype}      % microtypography
-\usepackage{graphicx}
-\graphicspath{{figs/}}
-%\usepackage{subfigure}
-%\usepackage{subcaption}
-%\usepackage{placeins}
-\usepackage{hyperref}       % hyperlinks
-\usepackage[dvipsnames]{xcolor}
-
-\usepackage{array}
-\usepackage{wrapfig}
-
-\hypersetup{ % SLJ: my standard paper setup...
-	pdftitle={MAP convergence},
-	pdfkeywords={},
-	pdfborder=0 0 0,
-	pdfpagemode=UseNone,
-	colorlinks=true,
-	linkcolor=blue, %mydarkblue,
-	citecolor=blue, %mydarkblue,
-	filecolor=blue, %mydarkblue,
-	urlcolor=blue, %mydarkblue,
-	pdfview=FitH,
-	pdfauthor={Anonymous},
-	% draft,
-	final,
-}
-
-% To make a table of contents only for the appendix
-\usepackage[toc,page,header]{appendix}
-\usepackage{minitoc}
-\renewcommand \thepart{}
-\renewcommand \partname{}
-
-% To make colored boxes
-\usepackage[framemethod=TikZ]{mdframed}
-\newmdenv[
-	roundcorner=10pt,
-	backgroundcolor=Red!30,
-]{important}
-
-\usepackage[capitalise]{cleveref}
-\newcommand{\rlp}[1]{\textcolor{BrickRed}{(RLP:#1)}}
-\newcommand{\fdk}[1]{\textcolor{Periwinkle}{(fdk:#1)}}
-\newcommand{\TODO}[1]{\textcolor{cyan}{(TODO #1)}}
-\newcommand{\tocite}{\textcolor{purple}{(add citation)}}
-
-% my packages
-\usepackage{math_commands}
-% some custom math commands
-\newtheorem{proposition}{Proposition}
-\newtheorem{problem}{Open Problem}
-
-\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
-\newcommand*{\expecti}[2][]{\ensuremath{\mathbb{E}_{#1} [ #2 ] }} % expectation operator
-
-\newcommand{\cond}{\,\vert\,}
-\newcommand{\logpart}{A}
-\newcommand{\conj}{{\logpart^*}}
-\newcommand{\bregman}{\cB_\logpart}
-\newcommand{\bregmanconj}{\cB_{\logpart^*}}
-\newcommand{\nat}{\theta}
-\newcommand{\m}{\mu}
-\newcommand{\meanp}{\m}
-\newcommand{\decrement}{D}
-\newcommand{\linear}{\ell} % linearization of a function
-\newcommand{\lr}{\gamma} % learning rate, or step-size
-\newcommand{\lin}[1]{\left\langle#1\right\rangle}
-
-\newcommand{\MAPm}{\hat \m_n}
-\newcommand{\MAPt}{\hat \nat_n}
-\DeclareMathSymbol{\shortminus}{\mathbin}{AMSa}{"39}
-
-\newcommand{\stgcvx}{\alpha} % strong convexity
-\newcommand{\smooth}{\beta} % smoothness
-
-\begin{document}
-
-% If your paper is accepted and the title of your paper is very long,
-% the style will print as headings an error message. Use the following
-% command to supply a shorter title of your paper so that it can be
-% used as headings.
-%
-\runningtitle{Convergence Rates for the MAP of an Exponential Family and SMD -- an Open Problem}
-
-% If your paper is accepted and the number of authors is large, the
-% style will print as headings an error message. Use the following
-% command to supply a shorter version of the authors names so that
-% they can be used as headings (for example, use only the surnames)
-%
-\runningauthor{R\'emi Le Priol, Frederik Kunstner, Damien Scieur, and Simon Lacoste-Julien}
-
-\twocolumn[
-
-\aistatstitle{Convergence Rates for the MAP of an Exponential Family\\ and Stochastic Mirror Descent -- an Open Problem}
-
-\aistatsauthor{R\'emi Le Priol \And Frederik Kunstner}
-\aistatsaddress{ 
-	Mila, Université de Montreal 
-	\And  University of British Columbia
-	}
-
-\aistatsauthor{Damien Scieur \And Simon Lacoste-Julien}
-\aistatsaddress{Samsung, SAIT AI Lab, Montreal 
-	\And Mila, Université de Montreal \\
-	Samsung, SAIT AI Lab, Montreal\\
-	Canada CIFAR AI Chair
-}
-
-]
-
-\begin{abstract}
-We consider the problem of upper bounding the expected log-likelihood sub-optimality of the maximum likelihood estimate (MLE), or a conjugate maximum a posteriori (MAP) for an exponential family, in a non-asymptotic way.
-Surprisingly, we found no general solution to this problem in the literature. In particular, current theories do not hold for a Gaussian or in the interesting few samples regime.
-After exhibiting various facets of the problem, we show we can interpret the MAP as running stochastic mirror descent (SMD) on the log-likelihood.
-However, modern convergence results do not apply for standard examples of the exponential family, highlighting holes in the convergence literature.
-We believe solving this very fundamental problem may bring progress to both the statistics and optimization communities.
-\end{abstract}
-
-\doparttoc % Tell to minitoc to generate a toc for the parts
-\faketableofcontents % Run a fake tableofcontents command for the partocs
+%\doparttoc % Tell to minitoc to generate a toc for the parts
+%\faketableofcontents % Run a fake tableofcontents command for the partocs
 %\part{} % Start the document part
@@
 
-\begin{figure}[t]
+\begin{figure}[h]
 	\centering
-\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}
-	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
+\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
+	\caption{\vspace{-5ex}KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
 		Solid curve  are average over $10^5$ trials.
@@
 \end{figure}
+
+%\begin{figure}[h]
+%	\centering
+%	\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=.6\textwidth}}]{figure}[\FBwidth]
+%	{\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
+%			Solid curve  are average over $10^5$ trials.
+%			Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
+%			Shaded areas are 90\% confidence interval.
+%			The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
+%			MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
+%			We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.}\label{fig:curves}}
+%	{\includegraphics[width=.4\textwidth]{gammaMAPrate.pdf}}
+%\end{figure}
 
@@
 \boxed{
-\begin{aligned}
-	\textit{Find an upper}&\textit{ bound on the expected value of } \\
-	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
-\end{aligned}
+	\textit{Find an upper bound on the expected value of } \KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
 }
@@
 Stochastic optimization offers an interesting perspective on~\eqref{problem}.
-Consider the problem
-\alignn{
-	\min_{\theta\in \Theta} f(\theta)\,,
-	\label{eq:optimization_problem}
-}
-solved by $\nat^*\in \Theta$.
+Consider the problem $\min_{\theta\in \Theta} f(\theta)$, solved by $\nat^*\in \Theta$.
 Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
@@
 }
-Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
+Both MLE and MAP can be seen as stochastic algorithms that minimizes $f$.
 In particular, with exponential families, MAP is equivalent to stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
@@
 \end{align}
-This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding and multivariate normal distributions $\cX=\real^d, \nu$ Lebesgue and $T(X)=(X, X X^\top)$.
+This model encompasses categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding and multivariate normal distributions $\cX=\real^d, \nu$ Lebesgue and $T(X)=(X, X X^\top)$.
 
@@
 \begin{align}
-    \nabla\logpart(\nat) &=  \expect[p_\nat]{T(X)} =: \meanp, \label{eq:mirror-map} \\
-    \nabla^2 \logpart(\nat) &= \Cov_{p_\nat}[T(X)] > 0,
+    \nabla\logpart(\nat) =  \expect[p_\nat]{T(X)} =: \meanp,\qquad    \nabla^2 \logpart(\nat) = \Cov_{p_\nat}[T(X)] > 0,  \label{eq:mirror-map}
 \end{align}
@@
 
-We now introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
+We introduce the convex conjugate (the Fenchel-Legendre transform) of the log-partition function
 \aligns{
@@
 %Not super convinced about introducing that here? it comes naturally when considering the KL on exponential maps) %RLP it is useful to the conjugate prior.
-{\bf The Bregman Divergence} induced by $\logpart$ measures the discrepancy between two parameters $\nat$ and $\nat_0$,
+{\bf The Bregman Divergence} induced by $\logpart$ measures the discrepancy between $\nat$ and $\nat_0$,
 \begin{align}
@@
     p(\nat)
-    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \nonumber \\
-    &\propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
+    &\propto \exp( - n_0 \bregman(\nat ; \nat_0) ) \propto \exp(n_0 \langle \m_0, \nat \rangle - n_0 \logpart(\nat)),
     \label{eq:def_prior}
@@
 How does this quantity behave when $\hat \nat$ is the MLE or MAP?
- Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators when the loss is the KL divergence?
+In other words, what is the \emph{frequentist risk} of these estimators when the loss is the KL divergence?
 This is our first problem.
@@
 With exponential families, MAP can be linked to \emph{stochastic mirror descent (SMD)}, see App.~\ref{app:SMD}.
-More precisely, let us re-write~\eqref{eq:defMAP} as
-\alignn{
-\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
-\label{eq:mean_update}
-}
-where $\lr_n := \inv{n_0 + n}$.
+More precisely, let us re-write~\eqref{eq:defMAP} as $\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)$, where $\lr_n := \inv{n_0 + n}$.
 Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
-If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
-\alignn{
-	\nabla\conj(\hat \nat_{n})
-	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
-}
-which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
+If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then the update on $g_n(\nat)$ becomes $\nabla\conj(\hat \nat_{n})= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})$, which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
 and step-size schedule $\lr_n$, initialized at $\nat_0$.
@@
 \aligns{
-	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
+	\textstyle \logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
 	\conj(\m) = - \half \log(\m)  \; ,
@@
 \begin{align}
-	\bregmanconj( \m^*; \m_n)
-	&= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
+	\textstyle \bregmanconj( \m^*; \m_n)
+	= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
 %	\bregman( \nat_n; \nat_* )
@@
 	\begin{align}
-		 \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
+		 \textstyle  \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
 			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
@@
  \begin{align}
-	& \expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
+	\expect{\bregmanconj( \m^*; \hat \m_n^\mathrm{MAP})}
 	\leq \begin{cases}
 		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
-		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
+		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2,
 	\end{cases}
-	\label{eq:MAP_rate}\\
-	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
+	\label{eq:MAP_rate} \;\;
+	b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; .
 \end{align}
@@
 	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
-	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution.
+	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution. \vspace{-3ex}
 	}
@@
 \alignn{
-	\textstyle \logpart(\nat) &= \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\ 
-	\textstyle \conj(\m) &= \textstyle - \half \log (\mu_2 - \mu_1^2)
+	\textstyle \logpart(\nat) = \textstyle \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2), \qquad \textstyle \conj(\m) = \textstyle - \half \log (\mu_2 - \mu_1^2)
 }
@@
 These functions are neither smooth, nor strongly convex, but they are self-concordant, since $\conj$ is  the logarithmic barrier of a quadratic domain
-\citep[p.177, example 4.1.1.4]{nesterov2003introductory}, and self-concordance is preserved by convex-conjugacy~\citep{nesterov1994interior} -- see more details in App.~\ref{app:gaussian}.
+% \citep[p.177, example 4.1.1.4]{nesterov2003introductory}, and self-concordance is preserved by convex-conjugacy~\citep{nesterov1994interior} 
+-- see more details in App.~\ref{app:gaussian}.
 We now discuss the general problem and some ways to solve it via direct expansions of the Bregman divergence.
@@
 \end{align}
-Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible for large $n$.
+Both MLE and MAP have the same asymptote, as the prior $n_0 \meanp_0$ gets negligible for large $n$.
 This asymptote is independent of the optimum $\meanp^*$ or $\mF$ for well-specified models.
@@
 \[
-	\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
+	\textstyle\cX=\real^d,\quad \nu(dx) = \exp\paren{\textstyle \half[-\|x\|^2]} dx,\quad T(x)=x.
 \]
-In this case, $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances since
-\begin{align}
-	\bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2  \; .
-\end{align}
+In this case, $\conj(\meanp) = \half \norm{\meanp}_2^2$ as well, and both Bregman divergences are squared $\ell^2$ distances since $ \bregmanconj(\meanp^* ; \meanp) = \half \norm{\meanp^* -  \meanp }_2^2$.
 Thanks to the independence of samples, we can break down the MLE into individual point's contributions:
@@
 so $\bregmanconj$ is upper bounded by a quadratic, and we get~\eqref{eq:MAP_quadratic} as an upper bound.
-It is also possible to get (more complex) upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
+% It is also possible to get upper bounds under restricted notions of strong-convexity \citep{negahban2012unified}.
 Besides the Gaussian with known variance, the problem is that no standard exponential family has a \textit{globally} strongly convex log-partition function.
@@
 \label{ssec:local-quadratic}
-From the Taylor expansion~\eqref{eq:bregmanTaylor},
-we know that all Bregman divergences are locally quadratic.
-Under some assumptions, such as self-concordance\footnote{
-In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
-} of $\conj$ \citep[Ch.~4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in App.~\ref{app:self-concordant}.
+From~\eqref{eq:bregmanTaylor}, we know that Bregman divergences are locally quadratic.
+Under some assumptions, such as self-concordance of $\conj$ (in 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$ \citep[Ch.~4.1]{nesterov2003introductory}), we can quantify when this quadratic behavior kicks in.
+% Proofs for this subsection are in App.~\ref{app:self-concordant}.
 \begin{proposition}
@@
 Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
-Then\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
+Then \vspace{-3ex} % \footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
 \aligns{
@@
 Laplace with known mean,
-Pareto with known minimum value,
-or Weibull with known shape $k$.
+Pareto with known minimum value,...
+% or Weibull with known shape $k$.
 The entropy is also self-concordant when $T$ lives in a compact \citep{bubeck2015entropic} -- e.g., categorical and Dirichlet distributions.
@@
 The expected KL of the categorical MLE is always infinite, as previously mentioned in \S\ref{sec:problem}.
-However, Proposition~\ref{prop:selfConcordant} may be used to prove high-probability, many samples, convergence rates for one dimensional (and possibly multivariate) normal distributions.
+However, Proposition~\ref{prop:selfConcordant} may be used to prove high-probability, many samples, convergence rates for % one dimensional (and possibly multivariate) 
+normal distributions.
 
@@
 \citet{ostrovskii2021finite} which characterizes the number of samples needed to be upper bounded by a quadratic \emph{with high-probability}, for any parametric models with a self-concordant log-likelihood $f$.
-\citet{anastasiou2017bounds} obtains a similar flavor of result under other assumptions on the third derivative of $f$.
+\citet{anastasiou2017bounds} obtains a similar result under other assumptions on the third derivative of $f$.
 More closely, in the world of exponential families, \citet{kakade2010learning} prove a result similar to Proposition~\ref{prop:selfConcordant} from a local bound on all higher-order moments of $\logpart$ in $\nat^*$.
@@
 \textbf{Remark:} In this decomposition, the primal expectation $\expecti{\hat \theta_n}$ is the reference point.
-An estimator will be unbiased if $\tilde \nat_n = \nat^*$.
-This is not true for the MLE, which is unbiased w.r.t. the dual parameter $\expecti{\hat \m_n}=\m^*$.
+An estimator is unbiased if $\tilde \nat_n = \nat^*$.
+This is not true for the MLE, unbiased w.r.t. the dual parameter $\expecti{\hat \m_n}=\m^*$.
 
@@
 
+%\begin{figure}[t]
+%	\centering
+%	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
+%	\caption{
+%	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
+%	}
+%	\label{fig:gaussian_decomposition}
+%\end{figure}
+
 \begin{figure}[t]
-	\centering
-	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
-	\caption{
-	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
-	}
-	\label{fig:gaussian_decomposition}
+	\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=.5\textwidth}}]{figure}[\FBwidth]
+	{\caption{
+			Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with parameters $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
+		}\label{fig:gaussian_decomposition}}
+	{\includegraphics[width=.3\textwidth]{figs/gaussians/new_linear_n0=1.pdf}\vspace{-5ex}}
 \end{figure}
@@
 Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
-\emph{relative} to a reference function $\logpart$, defined as
-\aligns{
-	\stgcvx \cB_{A}(x ; y)
-	\leq
-	\cB_f(x ; y)
-	\leq
-	\smooth \cB_A(x ; y) \; .
-}
+\emph{relative} to a reference function $\logpart$, defined as $\stgcvx \cB_{A}(x ; y)\leq\cB_f(x ; y)\leq\smooth\cB_A(x ; y)$.
 When $\logpart = \norm{\cdot}^2$, we recover the standard smoothness and gradient descent.
@@
 
-For exponential families, MAP perfectly fits into this framework, as
-\aligns{
-	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
-}
-is $1$-smooth and $1$-strongly convex relative to $A$.
+For exponential families, MAP perfectly fits into this framework, as $f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}$ is $1$-smooth and $1$-strongly convex relative to $A$.
 Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
@@
 \begin{table}[t]
-	\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
-	\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
-	\caption{Summary of results for SMD
+	\caption{\vspace{-4ex} Summary of results for SMD
 		under relative smoothness and relative strong convexity assumptions.
@@
 		($\hat\nat_n$) is it for the last iterate or an average ?
-		None of these analysis check all the boxes needed to address~\eqref{problem}.
+		None of these analysis check all the boxes needed to address~\eqref{problem}. \vspace{-4ex}
 	}
@@
 % or when $\logpart$ is $\stgcvx$-strongly-convex,
-we recover the variance of the gradients at the optimum, which is weaker than~\eqref{eq:gradient-var},
-\alignn{
-	\expect[g\!]{\norm{\nabla g(\theta^*)}^2}\leq C \; .
-	\label{eq:opt-gradient-var}
-}
-%Note that~\eqref{eq:opt-gradient-var} is weaker than~\eqref{eq:gradient-var}.
+we recover the variance of the gradients at the optimum, which is weaker than~\eqref{eq:gradient-var}, since $\expect[g\!]{\norm{\nabla g(\theta^*)}^2}\leq C$.
 Using their descent lemma \citep[Eq.
@@
 
-\subsubsection*{Acknowledgments}
-We express our gratitude to Reza Babanezhad, Radu Dragomir and Hadrien Hendrikx for their insights on stochastic mirror descent. 
-This work was partially supported by the NSERC Discovery Grant RGPIN2017-06936 and the Canada CIFAR AI Chair Program. 
-Simon Lacoste-Julien is a CIFAR Associate Fellow in the Learning in Machines \& Brains program.
-
+% Delete this because its not anonymous
+%\subsubsection*{Acknowledgments}
+%We express our gratitude to Reza Babanezhad, Radu Dragomir and Hadrien Hendrikx for their insights on stochastic mirror descent. 
+%This work was partially supported by the NSERC Discovery Grant RGPIN2017-06936 and the Canada CIFAR AI Chair Program. 
+%Simon Lacoste-Julien is a CIFAR Associate Fellow in the Learning in Machines \& Brains program.
+
+\clearpage
 \bibliographystyle{apalike}
@@
 \appendix
-\onecolumn
-
-\addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
-\part{Appendix} % Start the appendix part
-\parttoc % Insert the appendix TOC
+
+% \addcontentsline{toc}{section}{Appendix} % Add the appendix text to the document TOC
+% \part{} % Start the appendix part
+
+% \parttoc % Insert the appendix TOC
+
+\renewcommand{\contentsname}{APPENDICES}
+\tableofcontents
 
@@
 \fi
-
- \end{document}

### 2022-02-03 aistats/rebuttal/main_response.tex
Enable aistats.tex to compile

--- 2021-12-03 rebuttal-aistats/main_response.tex
@@
 
-We thank all the reviewers for their careful review and insightful or encouraging comments. We start our rebuttal by answering the common concern, followed by specific comments for each reviewer.
+We thank all the reviewers for their careful review and insightful or encouraging comments. We start by answering the common concern, followed by specific comments for each reviewer.
 
@@
 
-In 1954, John Tukey  \nocite{tukey1954unsolved} wrote "Difficulties in identifying problems have delayed statistics far more than difficulties in solving problems.
+%In 1954, John Tukey  \nocite{tukey1954unsolved} wrote "Difficulties in identifying problems have delayed statistics far more than difficulties in solving problems.
 This seems likely to be the case in the future, too."
@@
 Yet, we are still unaware of any general upper bound.
-In our opinion, a bound of the form $\frac{c}{n}$ with a large constant $c>\frac{d}{2}$, or even a bound $\frac{c}{\sqrt{n}}$ would be novel and interesting.
+In our opinion, a bound of the form $\frac{c}{n}$ with a large constant $c>\frac{d}{2}$, or even a bound $\frac{c}{\sqrt{n}}$ valid for small $n$ would be novel and interesting.
 Consequently, we decided to keep the question open-ended.
@@
 Additionally, about the statement: (R1) "many of such results are essentially known."
-We are very interested in knowing examples of such   results, as we believe they could help us improve the paper.
-
-We will also discuss \citet{frostig2015competing}, noting that they work with smooth and strongly convex losses, which is precisely the assumptions we aim to relax.
+If the reviewer has in mind such results that were not mentioned in the paper and the reviews, we would really appreciate specific references, as they could help us improve the paper.
+We will also discuss \citet{frostig2015competing}, noting that they work with smooth and strongly convex losses, which are precisely the assumptions we are aiming to relax.
 

### 2022-02-03 colt/paper_content.tex
Compile on my computer

@@
 \boxed{
-	\textit{Find an upper bound on the expected value of } \KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
+	\textit{Find an upper bound on the expected value of}\ \ \KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
 }
@@
 
-\begin{figure*}[t]
+\begin{figure}[t]
 	\centering
@@
 	\label{fig:thales}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{figure}[t]
-	\floatbox[{\capbeside\thisfloatsetup{capbesideposition={right,top},capbesidewidth=.5\textwidth}}]{figure}[\FBwidth]
+\begin{figure}[t]\floatbox[{\capbeside\thisfloatsetup{capbesideposition={left,top},capbesidewidth=.3\textwidth}}]{figure}[\FBwidth]
 	{\caption{
@@
 		}\label{fig:gaussian_decomposition}}
-	{\includegraphics[width=.3\textwidth]{figs/gaussians/new_linear_n0=1.pdf}\vspace{-5ex}}
+	{\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}\vspace{-5ex}}
 \end{figure}

### 2022-02-03 COLT2022/colt2022.tex
Minor changes on my colt version

@@
 \RequirePackage[dvipsnames]{xcolor}
-%\documentclass[anon,12pt]{colt2022} % Anonymized submission
-\documentclass[final,12pt]{colt2022} % Include author names
+\documentclass[anon,12pt]{colt2022} % Anonymized submission
+%\documentclass[final,12pt]{colt2022} % Include author names
 % The following packages are automatically loaded:
@@
 \begin{keywords}%
-  List of keywords%
+    Exponential Familly, Estimator, Mirror Descent
 \end{keywords}
@@
 
-\begin{figure}
+\section{INTRODUCTION}
+\label{sec:motivation}
+
+\paragraph{Models}
+Exponential families are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
+Many standard random variables are exponential families: Gaussians, categorical, gamma, or Dirichlet, for example.
+They are flexible enough to model a variety of data sources $X$ and easy to describe with some sufficient statistics $T(X) \in \real^d$.
+They are particularly appreciated for their convex log-likelihood
+\alignn{
+f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
+\label{eq:defNLL}
+}
+where $\logpart$ is the convex log-partition function and \mbox{$\nat\in\Theta$} is the \textit{natural parameter}.
+This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} or variants of principal component analysis \citep{collins2001generalization}, among other applications.
+
+\begin{figure}[ht]
 	\centering
@@
 \end{figure}
-
-\section{INTRODUCTION}
-\label{sec:motivation}
-
-\paragraph{Models}
-Exponential families are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
-Many standard random variables are exponential families: Gaussians, categorical, gamma, or Dirichlet, for example.
-They are flexible enough to model a variety of data sources $X$ and easy to describe with some sufficient statistics $T(X) \in \real^d$.
-They are particularly appreciated for their convex log-likelihood
-\alignn{
-f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
-\label{eq:defNLL}
-}
-where $\logpart$ is the convex log-partition function and \mbox{$\nat\in\Theta$} is the \textit{natural parameter}.
-This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized} or variants of principal component analysis \citep{collins2001generalization}, among other applications.
 
@@
 \begin{aligned}
-	\textit{Find an upper}&\textit{ bound on the expected value of } \\
-	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
+	\textit{Find an upper bound on the expected value of } \ \
+	\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
 \end{aligned}
@@
 Consider the problem
-\alignn{
-	\min_{\theta\in \Theta} f(\theta)\,,
-	\label{eq:optimization_problem}
-}
+	$\min_{\theta\in \Theta} f(\theta)$,
 solved by $\nat^*\in \Theta$.
@@
 }
-Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
+Both MLE and MAP can be seen as stochastic algorithms minimizing $f$.
 In particular, with exponential families, MAP is equivalent to stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
@@
 \bibliographystyle{apalike}
-\bibliography{../references}
+\bibliography{references}
 

### 2022-02-03 colt/paper_content.tex
Fix table appearance and appendix
Cite  Chowdbury and move fig 1 to 2nd page
Problem doc is 13 pages

@@
 %\parttoc % Insert the document TOC
-
-\begin{figure}[h]
-	\centering
-\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
-	\caption{\vspace{-5ex}KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
-		Solid curve  are average over $10^5$ trials.
-		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
-		Shaded areas are 90\% confidence interval.
-		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
-		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
-		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
-	}
-	\label{fig:curves}
-\end{figure}
 
@@
 
+\begin{figure}[ht]
+	\centering
+\includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
+	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
+		Solid curve  are average over $10^5$ trials.
+		Dashed curves are upper bounds~\eqref{eq:MLE_rate} (blue) and~\eqref{eq:MAP_rate} (orange, not tight by a factor 2).
+		Shaded areas are 90\% confidence interval.
+		The MLE expected KL is infinite for $n=1$ and $n=2$, but for $n\geq3$ it quickly joins the upper bound~\eqref{eq:MLE_rate} and the $1/2n$ asymptote~\eqref{eq:asymptote}.
+		MAP's expected KL is always finite, and it has lower variance than MLE, but it is slower to join the asymptote.
+		We wish to find upper bounds similar to~\eqref{eq:MAP_rate} characterizing the relative importance of the prior and the few sample behavior of MAP for a variety of exponential families.
+	}
+	\label{fig:curves}
+\end{figure}
+
 \paragraph{Estimators}
@@
 For instance \citet[Th.~9-10]{kaufmann2021mixture} provides high-confidence upper bounds on~\eqref{problem}, but only for Gaussians with known variance or Gamma with known shape.
+Very recently, \citet{chowdhury2022bregman} expressed implicit confidence bounds for the MLE of exponential families with the KL divergence, which is a similar to~\eqref{problem} but different.
 Other works compare other online estimators to MLE or MAP \citep{azoury2001relative,dasgupta2007online}.
@@
 \begin{table}[t]
-	\caption{\vspace{-4ex} Summary of results for SMD
+	\caption{Summary of results for SMD
 		under relative smoothness and relative strong convexity assumptions.
@@
 		($\hat\nat_n$) is it for the last iterate or an average ?
-		None of these analysis check all the boxes needed to address~\eqref{problem}. \vspace{-4ex}
+		None of these analysis check all the boxes needed to address~\eqref{problem}.
 	}
@@
 \clearpage
-\appendix
 
@@
 
-\renewcommand{\contentsname}{APPENDICES}
+%\renewcommand{\contentsname}{APPENDICES}
 \tableofcontents
 
+\appendix
 \section{PROOFS FOR GAUSSIAN VARIANCE}
@@
 \end{align}
-\begin{wrapfigure}[14]{r}{0.4\textwidth}
+\begin{wrapfigure}[12]{r}{0.4\textwidth}
 	\centering
-	\includegraphics[width=.38\textwidth]{phi.pdf}
+	\includegraphics[width=\textwidth]{phi.pdf}
 	\caption{%
@@
  \begin{align}
-	& \expect{\cS_\conj( \m^*; \hat \m_n^\mathrm{MAP})}
+	\expect{\cS_\conj( \m^*; \hat \m_n^\mathrm{MAP})}
 	\leq 
@@
 		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n  & \text{ if }\ n\geq 2,
-	\end{array}\right.
-	\quad 
-	\quad 
+	\end{array}\right. \\
 	\text{where }

### 2022-02-10 colt/paper_content.tex
Remove some space but still 13 pages
Update paper_content.tex

@@
 \includegraphics[width=.6\textwidth]{gammaMAPrate.pdf}
+	\vspace{-3ex}
 	\caption{KL divergence~\eqref{eq:suboptimalityKL} for Gaussian variance (\S\ref{ssec:gaussian-variance}) MLE (blue) and MAP (orange) against number of samples $n$. 
@@
 and raising the fundamental problem:
-\begin{equation}
+\begin{equation}\setlength\fboxsep{2ex}
 \boxed{
-	\textit{Find an upper bound on the expected value of}\ \ \KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
+	\textit{Find an upper bound in expectation on}\ \ \KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; \textit{(or in high probability)}.
 }
@@
 	The hyperparameter of the prior $\nat_0$ controls the brown point's location while varying $n_0$ spans the long edges of the triangle and the red curve.
-	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution. \vspace{-3ex}
+	Large blurry circles in the background are other instances of MAP and MLE revealing their distribution.
 	}
@@
 \begin{align}
-    \bregmanconj(\m^* ; \m)
-    &\leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m} \\
-    &\leq 2L \norm{\m^* - \m} \; ,
+    \bregmanconj(\m^* ; \m)  \leq L \norm{\m^* - \m} + \norm{\nat} \norm{\m^* - \m}  2L \norm{\m^* - \m} \; ,
 \end{align}
@@
 Besides the Gaussian with known variance, the problem is that no standard exponential family has a \textit{globally} strongly convex log-partition function.
-The next section focuses on \textit{local} quadratic behavior, which is more realistic.
+The next section focuses on the more realistic \textit{local} quadratic behavior.
 
@@
 Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$.
-Then \vspace{-3ex} % \footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
+Then % \footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$.}
 \aligns{
@@
 \begin{table}[t]
-	\caption{Summary of results for SMD
-		under relative smoothness and relative strong convexity assumptions.
-		Each row correspond to one analysis, and each columns answers one question.
-		($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
-		($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
-		($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^* ; \hat\nat_n)$?
-		($\hat\nat_n$) is it for the last iterate or an average ?
-		None of these analysis check all the boxes needed to address~\eqref{problem}.
-	}
 	\begin{center}
@@
 	\end{center}
+	\vspace{-3ex}
+	\caption{Summary of results for SMD
+		under relative smoothness and relative strong convexity assumptions.
+		Each row correspond to one analysis, and each columns answers one question.
+		($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
+		($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
+		($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^* ; \hat\nat_n)$?
+		($\hat\nat_n$) is it for the last iterate or an average ?
+		None of these analysis check all the boxes needed to address~\eqref{problem}.
+	}
 	\label{tbl:assumptions}
